---
title: "Hooknet downstream processing"
author: "Daniel Schulz"
date: "2026-08-17"
output: html_document
---


In this script we will investigate the results from HooknetTLS applied to H&E images.

The tool has been run using this repo: [HooknetTLS](https://github.com/gorkemsolun/pathology-hooknet-tls)

It exports a raw folder that contains all identified TLS like and GC (Germinal Center) like structures.
We also exported a folder applying filtering as used in the original publication.

In order to inspect the annotations with QuPath the json files need to be transformed to geojsons.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

Convert a HookNet-TLS *_with_confidence.json export into a GeoJSON 
FeatureCollection, preserving the full per-polygon confidence histogram
as a feature property. Geojson can then be viewed in Qupath

```{r}
library(jsonlite)
```

# geojson function

```{r}


## Same convention as xml_to_geojson_adapted.py, so output stays QuPath-compatible.
GROUP_COLORS  <- list(tls = c(0, 255, 0), gc = c(255, 0, 0))
DEFAULT_COLOR <- c(128, 128, 128)


feature_for_item <- function(item) {
  coords <- item$coordinates
  ring   <- coords
  if (!identical(coords[[1]], coords[[length(coords)]])) {
    ring <- c(ring, coords[1])  # close the ring if the source didn't already
  }
  
  name  <- item$label$name
  color <- GROUP_COLORS[[tolower(name)]]
  if (is.null(color)) color <- DEFAULT_COLOR
  
  list(
    type = "Feature",
    id = "PathDetectionObject",
    geometry = list(
      type = "Polygon",
      coordinates = list(ring)
    ),
    properties = list(
      objectType = "annotation",
      index = item$index,
      classification = list(name = name, color = color),
      color = color,
      confidence_map = item$label$confidence_map
    )
  )
}



## Each item's coordinates form a single ring (no holes in this dataset).


conv_to_geojson <- function(json_file, outpath){
  
  raw <- fromJSON(json_file, simplifyVector = FALSE) 
  
  features <- lapply(raw, feature_for_item)
  
  geojson  <- list(type = "FeatureCollection", features = features)
  cur_name <- sub("\\.json$", ".geojson", basename(json_file))
  
  writeLines(toJSON(geojson, auto_unbox = TRUE, digits = NA, pretty = TRUE), file.path(outpath,cur_name))

}

```


# geojson for all raw json

run for all raw json files. this transforms the raw json files (detected TLS and GCs) to geojsons
```{r}
json_files <- list.files("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/HE/hooknetTLS/raw-20260724T081937Z-1-001/raw", full.names = TRUE)
unlink(json_files[grepl("IMMU-NSCLC", json_files)])
json_files <- list.files("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/HE/hooknetTLS/raw-20260724T081937Z-1-001/raw", full.names = TRUE, pattern = "*.json")

# transform to geojson
lapply(json_files, function(x) conv_to_geojson(x,outpath = "/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/HE/hooknetTLS/raw-20260724T081937Z-1-001/raw_geojson"))
```

# geojson for all filtered json

run for all filtered json files. this transforms the filtered json files (filtered as in the hooknetTLS paper) to geojsons
```{r}
# we will remove the second HE files
json_files <- list.files("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/HE/hooknetTLS/filtered-20260817T121553Z-1-001/filtered", full.names = TRUE)
unlink(json_files[grepl("IMMU-NSCLC", json_files)])
json_files <- list.files("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/HE/hooknetTLS/filtered-20260817T121553Z-1-001/filtered", full.names = TRUE, pattern = "*.json")

# generate geojson
lapply(json_files, function(x) conv_to_geojson(x,outpath = "/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/HE/hooknetTLS/filtered-20260817T121553Z-1-001/filtered_geojson"))
```


# Custom filtering
Filter a GeoJSON FeatureCollection produced by json_to_geojson.R.

Filters, operating directly on the GeoJSON so the result can be opened
straight in QuPath:
  1. Confidence filter: keep a feature only if at least `min_pct_above`
     of its pixels have a confidence bin >= `conf_threshold` (read back
     out of the confidence_map stored in properties by json_to_geojson.R).
  2. Size filter: drop "tls" features below `min_area_tls` and "gc"
     features below `min_area_gc` (area in the same squared units as the
     input coordinates, e.g. tiny high-confidence specks that are too
     small to plausibly be a real TLS/GC).
  3. Orphan-GC filter: a "gc" feature is kept only if its centroid falls
     inside at least one "tls" polygon (a real GC must sit inside a TLS).
     "tls" features themselves are never dropped by this filter.

Reading/writing GeoJSON is done with jsonlite (not sf::st_read/st_write)
so nested properties like confidence_map round-trip exactly as written --
sf is used only in-memory for the two spatial computations (centroid,
point-in-polygon).

```{r}
library(sf)

## % of a polygon's pixels with confidence >= conf_threshold, from a
## confidence_map list as parsed by fromJSON(simplifyVector = FALSE).
pct_pixels_above <- function(confidence_map, conf_threshold) {
  bins   <- as.numeric(names(confidence_map))
  counts <- as.numeric(unlist(confidence_map, use.names = FALSE))
  sum(counts[bins >= conf_threshold]) / sum(counts)
}

## Outer ring only (this dataset has no holes).
feature_to_sfg <- function(feature) {
  ring <- feature$geometry$coordinates[[1]]
  mat  <- t(vapply(ring, function(pt) c(pt[[1]], pt[[2]]), numeric(2)))
  st_polygon(list(mat))
}

filter_geojson_annotations <- function(geojson_path,
                                        out_path,
                                        conf_threshold,
                                        min_pct_above,
                                        min_area_tls = 0,
                                        min_area_gc = 0,
                                        discard_orphan_gc = TRUE,
                                        gc_check_against = c("filtered", "all")) {
  gc_check_against <- match.arg(gc_check_against)

  gj       <- fromJSON(geojson_path, simplifyVector = FALSE)
  features <- gj$features

  label_name <- vapply(features, function(f) f$properties$classification$name, character(1))
  pct_above  <- vapply(features, function(f) pct_pixels_above(f$properties$confidence_map, conf_threshold), numeric(1))

  sfgs <- lapply(features, feature_to_sfg)
  area <- vapply(sfgs, function(g) as.numeric(st_area(g)), numeric(1))

  min_area_by_class <- c(tls = min_area_tls, gc = min_area_gc)
  area_threshold     <- min_area_by_class[label_name]
  area_threshold[is.na(area_threshold)] <- 0  # classes other than tls/gc: no size filter

  keep <- (pct_above >= min_pct_above) & (area >= area_threshold)

  if (discard_orphan_gc) {
    gc_idx <- which(label_name == "gc" & keep)

    if (length(gc_idx) > 0) {
      tls_pool_idx <- which(label_name == "tls" &
                               (if (gc_check_against == "filtered") keep else TRUE))

      if (length(tls_pool_idx) == 0) {
        keep[gc_idx] <- FALSE
      } else {
        tls_geoms  <- st_sfc(sfgs[tls_pool_idx])
        centroids  <- st_sfc(lapply(sfgs[gc_idx], st_centroid))
        contained  <- lengths(suppressMessages(st_within(centroids, tls_geoms))) > 0
        keep[gc_idx] <- contained
      }
    }
  }

  out_gj <- list(type = "FeatureCollection", features = features[keep])
  
  cur_name <- sub("\\.json$", ".geojson", basename(geojson_path))
  
  writeLines(toJSON(out_gj, auto_unbox = TRUE, digits = NA, pretty = TRUE), file.path(out_path, cur_name))

  cat(sprintf("Kept %d / %d features -> %s\n", sum(keep), length(features), out_path))
  invisible(sum(keep))
}


# apply filtering
geojson_files <- list.files("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/HE/hooknetTLS/raw-20260724T081937Z-1-001/raw_geojson/", full.names = TRUE)

lapply(geojson_files, function(x) filter_geojson_annotations(geojson_path = x,
                                                      out_path = "/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/HE/hooknetTLS/custom_filtered",
                                                      conf_threshold = 100,
                                                      min_pct_above = 0.5,
                                                      min_area_tls = 35000,
                                                      min_area_gc = 5000,
                                                      discard_orphan_gc = TRUE)
                                                      )


```

# summarize data custom fileterd

```{r}
feature_to_sfg <- function(feature) {
  ring <- feature$geometry$coordinates[[1]]
  mat  <- t(vapply(ring, function(pt) c(pt[[1]], pt[[2]]), numeric(2)))
  st_polygon(list(mat))
}

files <- sort(list.files("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/HE/hooknetTLS/custom_filtered", pattern = "\\.geojson$", full.names = TRUE))
cat(sprintf("Found %d geojson files\n", length(files)))

rows <- list()

for (filepath in files) {
  sample   <- tools::file_path_sans_ext(basename(filepath))
  gj       <- fromJSON(filepath, simplifyVector = FALSE)
  features <- gj$features
  if (is.null(features)) features <- list()

  for (idx in seq_along(features)) {
    feature <- features[[idx]]

    name <- feature$properties$classification$name
    if (is.null(name)) name <- "unknown"

    coords <- feature$geometry$coordinates[[1]]
    poly   <- feature_to_sfg(feature)

    area      <- suppressWarnings(as.numeric(st_area(poly)))
    perimeter <- suppressWarnings(as.numeric(st_length(st_sfc(st_boundary(poly)))))
    n_vertices <- length(coords) - 1  # last point repeats the first

    rows[[length(rows) + 1]] <- data.frame(
      sample        = sample,
      annotation_id = idx,
      class         = name,
      area          = area,
      perimeter     = perimeter,
      n_vertices    = n_vertices,
      stringsAsFactors = FALSE
    )
  }
}

summary_df <- do.call(rbind, rows)

saveRDS(summary_df,"/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/HE/hooknetTLS/custom_filtered/annotation_summary.rds")
```

# summarize data standard filtered

```{r}
feature_to_sfg <- function(feature) {
  ring <- feature$geometry$coordinates[[1]]
  mat  <- t(vapply(ring, function(pt) c(pt[[1]], pt[[2]]), numeric(2)))
  st_polygon(list(mat))
}

files <- sort(list.files("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/HE/hooknetTLS/filtered-20260817T121553Z-1-001/filtered_geojson", pattern = "\\.geojson$", full.names = TRUE))
cat(sprintf("Found %d geojson files\n", length(files)))

rows <- list()

for (filepath in files) {
  sample   <- tools::file_path_sans_ext(basename(filepath))
  gj       <- fromJSON(filepath, simplifyVector = FALSE)
  features <- gj$features
  if (is.null(features)) features <- list()

  for (idx in seq_along(features)) {
    feature <- features[[idx]]

    name <- feature$properties$classification$name
    if (is.null(name)) name <- "unknown"

    coords <- feature$geometry$coordinates[[1]]
    poly   <- feature_to_sfg(feature)

    area      <- suppressWarnings(as.numeric(st_area(poly)))
    perimeter <- suppressWarnings(as.numeric(st_length(st_sfc(st_boundary(poly)))))
    n_vertices <- length(coords) - 1  # last point repeats the first

    rows[[length(rows) + 1]] <- data.frame(
      sample        = sample,
      annotation_id = idx , 
      class         = name,
      area          = area,
      perimeter     = perimeter,
      n_vertices    = n_vertices,
      stringsAsFactors = FALSE
    )
  }
}

summary_df <- do.call(rbind, rows)

saveRDS(summary_df,"/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/HE/hooknetTLS/filtered-20260817T121553Z-1-001/filtered_geojson/annotation_summary.rds")
```

