---
title: "clinical_data_preparation"
author: "Daniel Schulz"
date: "2026-07-03"
output: html_document
---

here we investigate the clinical data from the prospective advanced stage NSCLC cohort

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(survival)
#library(survivalAnalysis)
library(survminer)
library(ggplot2)
library(tidyverse)
```

# read data

## 1
the data consists of 5 sheets. we will read and clean each one
```{r}
traits <- read_csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/all_traits.csv")
dat1 <- readr::read_tsv("/mnt/immucan_volume/clinical_data/FINAL/LUNG-NSCLC2_curated_export_20260423_#_e1b3bae311355815dd66fbd785d61084/LUNG-NSCLC2_curated_export_20260423_clinical.tsv")
dat1$immucan_id <- dat1$patient_id
dplyr::rename(.data = dat1,Patient_id = patient_id)
all_pats <- unique(dat1$immucan_id)
```
## 2
```{r}
dat2 <- read_tsv("/mnt/immucan_volume/clinical_data/FINAL/LUNG-NSCLC2_curated_export_20260423_#_e1b3bae311355815dd66fbd785d61084/LUNG-NSCLC2_curated_export_20260423_samples.tsv")
colnames(dat2) <- c("Patient_id","Sample_id","Sampling_day","sample_type","Sample_origin","Sampling_site","Treatment_naive")
# rename column
dat2 <- dplyr::rename(dat2,immucan_id = Patient_id)

dat2 <- dat2 %>%
  filter(sample_type == "Fixed tissue")

# some patient ids are duplicated. check why
DT::datatable(dat2 %>%
  filter(immucan_id %in% dat2$immucan_id[duplicated(dat2$immucan_id)]))
```

## 3
treatment
note: we will get all the drugs from dat4 from the next chunk and then extract all the drugs from dat3 based on the names
```{r}
dat3 <- read_tsv("/mnt/immucan_volume/clinical_data/FINAL/LUNG-NSCLC2_curated_export_20260423_#_e1b3bae311355815dd66fbd785d61084/LUNG-NSCLC2_curated_export_20260423_lots.tsv")

# fix type in Patient_id
dat3 <- rename(dat3,immucan_id = patient_id)
dat3 <- rename(dat3,Regimen_details = medication_details)

# check if all patients are there. here we miss 5 patients.
missing_treatment <-all_pats[!all_pats %in% unique(dat3$immucan_id)]

dat3 <- dat3 %>%
  filter(LoT_number == 1)

dat3
table(dat3$clinical_setting)
length(unique(dat3$immucan_id))
```

## 4

```{r}
dat4 <- read_tsv("/mnt/immucan_volume/clinical_data/FINAL/LUNG-NSCLC2_curated_export_20260423_#_e1b3bae311355815dd66fbd785d61084/LUNG-NSCLC2_curated_export_20260423_medications.tsv")

DT::datatable(dat4)

dat4 <- rename(dat4,Drug = drug)
dat4 <- rename(dat4,Drug_Class = drug_class)
# fix type in Patient_id
dat4 <- rename(dat4,immucan_id = patient_id)

# check if all patients are there. here we miss 5 patients.
missing_treatment <- all_pats[!all_pats %in% unique(dat4$immucan_id)]

dat4 %>%
  filter(LoT_number == 1) %>%
  distinct(Drug)

dat4 %>%
  filter(grepl("Immuno", Drug))
# we will fix some drug names
dat4 <- dat4 %>%
  mutate(Drug = ifelse(Drug == "Ado-trastuzumab emtansine","Trastuzumab Emtansine", Drug))

chemodrugs <- dat4 %>%
  filter(Drug_Class == "Chemotherapy") %>%
  pull(Drug) %>%
  unique()
immunodrugs <- dat4 %>%
  filter(Drug_Class == "Immunotherapy") %>%
  pull(Drug) %>%
  unique()
TKIdrugs <- dat4 %>%
  filter(Drug_Class == "Tyrosine kinase inhibitor") %>%
  pull(Drug) %>%
  unique()
Otherdrugs <- dat4 %>%
  filter(Drug_Class == "Other") %>%
  pull(Drug) %>%
  unique()

# update dat3
dat3 <- dat3 %>%
  mutate(Chemotherapy_drugs = stringr::str_extract_all(dat3$Regimen_details, regex(paste(chemodrugs,collapse = "|"), ignore_case = TRUE)),
         Chemotherapy_drugs = sapply(Chemotherapy_drugs, function(x) paste(unique(x), collapse = ", ")),
         Immunotherapy_drugs = stringr::str_extract_all(dat3$Regimen_details, regex(paste(immunodrugs,collapse = "|"), ignore_case = TRUE)),
         Immunotherapy_drugs = sapply(Immunotherapy_drugs, function(x) paste(unique(x), collapse = ", ")),
         TKI_drugs = stringr::str_extract_all(dat3$Regimen_details, regex(paste(TKIdrugs,collapse = "|"), ignore_case = TRUE)),
         TKI_drugs = sapply(TKI_drugs, function(x) paste(unique(x), collapse = ", ")),
         Other_drugs = stringr::str_extract_all(dat3$Regimen_details, regex(paste(Otherdrugs,collapse = "|"), ignore_case = TRUE)),
         Other_drugs = sapply(Other_drugs, function(x) paste(unique(x), collapse = ", ")))

colnames(dat3) <- gsub(pattern = "medication",replacement = "Regimen", colnames(dat3))
length(unique(dat4$immucan_id))
```

## 5
```{r}
dat5 <- read_tsv("/mnt/immucan_volume/clinical_data/FINAL/LUNG-NSCLC2_curated_export_20260423_#_e1b3bae311355815dd66fbd785d61084/LUNG-NSCLC2_curated_export_20260423_progressions.tsv")
DT::datatable(dat5)

# fix type in Patient_id
dat5 <- rename(dat5,immucan_id = patient_id)
dat5 <- rename(dat5,Progression_number = progression_number)
unique(dat5$immucan_id)
# check if all patients are there. here we miss 5 patients.
missing_treatment <- all_pats[!all_pats %in% unique(dat5$immucan_id)]

dat5 <- dat5 %>%
  filter(Progression_number == 1)

table(dat5$regional)
```

```{r}
clin_dat <- plyr::join_all(dfs = list(dat1,dat2,dat3,dat5),by = "immucan_id")
cat("unique patients: ",length(unique(clin_dat$immucan_id)),"\n")
cat("rows in data: ",length(clin_dat$immucan_id),"\n")

clin_dat %>%
  filter(immucan_id == "LUNG-NSCLC2-0598") %>%
  select(Sample_id)
# list IMC patients
imc <- list.files("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/IMC/Panel_1/data/")
imc[grepl("LUNG-NSCLC2-0598", imc)]

clin_dat <- clin_dat %>%
  filter(Sample_id != "LUNG-NSCLC2-0598-FIXT-02")
```

# compare with old data
```{r}
traits
clin_dat
colnames(traits) <- paste0(colnames(traits), "_old")
traits$immucan_id <- traits$immucan_id_old
traits$immucan_id_old <- NULL
cur_dat <- left_join(clin_dat,traits ,by = "immucan_id")

cur_dat <- apply(cur_dat, 2,function(x) gsub(",",";",x))
cur_dat <- apply(cur_dat, 2,function(x) gsub("\\t",";",x))
cur_dat <- as.data.frame(cur_dat)
colorder <- c("immucan_id",sort(colnames(cur_dat[,-1])))
cur_dat <- cur_dat[,colorder]
write.csv(cur_dat, "/mnt/immucan_volume/clinical_data/Lung_NSCLC2/20260703_old_new_comparison.csv", row.names = FALSE, quote = FALSE)
```

```{r}
sel_dat <- cur_dat %>%
  select(immucan_id,smoking_status, smoking_status_old,stage_at_diagnosis, stage_old, type_of_adj_trt_old,first_line_old, Chemotherapy_drugs,  gender_old, sex,  neo_adjuvant_treatment_old,adjuvant_treatment_old,clinical_setting,Treatment_naive,time_to_event_old,last_FU_OS_days, censoring, censoring_old, Histology, simple_histology_old)
sel_dat <- sel_dat %>%
  mutate(stage_old = gsub("Stage ", "", stage_old))

sel_dat %>%
  mutate()

write.csv(sel_dat, "/mnt/immucan_volume/clinical_data/Lung_NSCLC2/20260703_old_new_comparison_selected.csv", row.names = FALSE, quote = FALSE)

library(ggalluvial)
sel_dat %>%
  ggalluvial::geom_alluvium()

```
```{r}

table(traits$never_vs_current_smoker_old)
table(traits$current_vs_Ex_smoker_old)
```

# generate more columns

```{r}
#censoring
clin_dat <- clin_dat %>%
  rename(OS_days = oS_days,
         Last_FU_OS_event = last_FU_OS_event) 
clin_dat <- clin_dat %>%
  mutate(censoring = ifelse(Last_FU_OS_event == "Dead",1,0),
         OS_days = as.numeric(OS_days))


# histology
clin_dat <- clin_dat %>%
  mutate(Histology = ifelse(grepl("adenocarcinoma|Adenocarcinoma", histopathologic_type),"LUAD", "Other"),
         Histology = ifelse(grepl("squamous cell carcinoma|Squamous cell carcinoma", histopathologic_type),"LUSC", Histology),
         Histology = ifelse(grepl("squamous cell carcinoma|Squamous cell carcinoma", histopathologic_type),"LUSC", Histology))

clin_dat <- clin_dat %>%
  rename(Number_of_regimens =number_of_Regimens)
# treatment
clin_dat <- clin_dat %>%
  mutate(treatment = ifelse(Regimen_Chemotherapy == "No" & Regimen_Immunotherapy == "No" & Regimen_TKI == "No"& Regimen_other_drug_treatments != "No", "Other","None"),
         treatment = ifelse(Regimen_Chemotherapy != "No" & Regimen_Immunotherapy == "No" & Regimen_TKI == "No"& Regimen_other_drug_treatments == "No", "Chemo",treatment),
         treatment = ifelse(Regimen_Chemotherapy == "No" & Regimen_Immunotherapy != "No" & Regimen_TKI == "No"& Regimen_other_drug_treatments == "No", "ICI",treatment),
         treatment = ifelse(Regimen_Chemotherapy == "No" & Regimen_Immunotherapy == "No" & Regimen_TKI != "No"& Regimen_other_drug_treatments == "No", "TKI",treatment),
         treatment = ifelse(Regimen_Chemotherapy != "No" & Regimen_Immunotherapy != "No" & Regimen_TKI == "No"& Regimen_other_drug_treatments == "No", "Chemo+ICI",treatment),
         treatment = ifelse(Regimen_Chemotherapy != "No" & Regimen_Immunotherapy != "No" & Regimen_TKI != "No"& Regimen_other_drug_treatments == "No", "Chemo+ICI+TKI",treatment),
         treatment = ifelse(Regimen_Chemotherapy != "No" & Regimen_Immunotherapy == "No" & Regimen_TKI != "No"& Regimen_other_drug_treatments == "No", "Chemo+TKI",treatment),
         treatment = ifelse(Regimen_Chemotherapy == "No" & Regimen_Immunotherapy != "No" & Regimen_TKI != "No"& Regimen_other_drug_treatments == "No", "ICI+TKI",treatment),
         treatment = ifelse(Regimen_Chemotherapy != "No" & Regimen_Immunotherapy == "No" & Regimen_TKI == "No"& Regimen_other_drug_treatments != "No", "Chemo+Other",treatment),
         treatment = ifelse(Regimen_Chemotherapy == "No" & Regimen_Immunotherapy != "No" & Regimen_TKI == "No"& Regimen_other_drug_treatments != "No", "ICI+Other",treatment),
         treatment = ifelse(Regimen_Chemotherapy == "No" & Regimen_Immunotherapy == "No" & Regimen_TKI != "No"& Regimen_other_drug_treatments != "No", "TKI+Other",treatment),
         treatment = ifelse(Regimen_Chemotherapy != "No" & Regimen_Immunotherapy != "No" & Regimen_TKI != "No"& Regimen_other_drug_treatments != "No", "Chemo+ICI+TKI+Other",treatment),
         treatment = ifelse(Regimen_Chemotherapy != "No" & Regimen_Immunotherapy != "No" & Regimen_TKI == "No"& Regimen_other_drug_treatments != "No", "Chemo+ICI+Other",treatment),
         treatment = ifelse(is.na(Number_of_regimens),"None", treatment))


# here we will ignore the "Other" drug treatments to obtain the broad treatment
clin_dat <- clin_dat %>%
  mutate(broad_treatment = ifelse(Regimen_Chemotherapy == "No" & Regimen_Immunotherapy == "No" & Regimen_TKI == "No"& Regimen_other_drug_treatments != "No", "Other","None"),
         broad_treatment = ifelse(Regimen_Chemotherapy == "No" & Regimen_Immunotherapy == "No" & Regimen_TKI != "No", "TKI",broad_treatment),
         broad_treatment = ifelse(Regimen_Chemotherapy == "No" & Regimen_Immunotherapy != "No" & Regimen_TKI == "No", "ICI",broad_treatment),
         broad_treatment = ifelse(Regimen_Chemotherapy != "No" & Regimen_Immunotherapy == "No" & Regimen_TKI == "No", "Chemo",broad_treatment),
         broad_treatment = ifelse(Regimen_Chemotherapy != "No" & Regimen_Immunotherapy != "No" & Regimen_TKI == "No", "Chemo+ICI",broad_treatment),
         broad_treatment = ifelse(Regimen_Chemotherapy != "No" & Regimen_Immunotherapy != "No" & Regimen_TKI != "No", "Chemo+ICI+TKI",broad_treatment),
         broad_treatment = ifelse(Regimen_Chemotherapy != "No" & Regimen_Immunotherapy == "No" & Regimen_TKI != "No", "Chemo+TKI",broad_treatment),
         broad_treatment = ifelse(Regimen_Chemotherapy == "No" & Regimen_Immunotherapy != "No" & Regimen_TKI != "No", "ICI+TKI",broad_treatment),
         broad_treatment = ifelse(is.na(Number_of_regimens),"None", broad_treatment))

# manual addition
clin_dat <- clin_dat %>%
  mutate(treatment = ifelse(immucan_id =="IMMU-NSCLC-0368", "Chemo+ICI", treatment),
         broad_treatment = ifelse(immucan_id =="IMMU-NSCLC-0368", "Chemo+ICI", broad_treatment))

# all TKI treatment will be called TKI irrespective of additional other treatments
clin_dat <- clin_dat %>%
  mutate(broad_treatment = ifelse(grepl("TKI", broad_treatment),"TKI",broad_treatment))

# recurrence: for some patients we have the recurrent samples. we will record this here
clin_dat <- clin_dat %>%
  mutate(Sample_type = ifelse(!is.na(TNM_at_recurrence_day),"recurrence_sample","primary_sample"))



# PFS censoring
clin_dat <- clin_dat %>%
  rename(PFS_event = pFS_event)
clin_dat <- clin_dat %>%
  rename(PFS_days = pFS_days)

clin_dat <- clin_dat %>%
  mutate(PFS_censoring = ifelse(PFS_event %in% c("alive_censored","lost to follow-up", "censored"), 0, 1),
         PFS_days =as.numeric(PFS_days))

# progression before 6 month -> primary resistance
# progression after 6 month -> secondary resistance
clin_dat <- clin_dat %>%
  mutate(resistance = ifelse(PFS_days < 365/2, "primary", "secondary"))

# long term survivors
clin_dat <- clin_dat %>%
  rename(OS_event = oS_event)
clin_dat <- clin_dat %>%
  mutate(long_survivor = ifelse(OS_days > 3*365 & OS_event == "Alive","yes", "no"))
```

# occult metastasis

select all stages below IIIB

date of diagnosis - date of progression < 150 days -> sample eligible
if > 150 days then if sample from original tumor -> exclude
if new sample -> keep
```{r}
# clin_dat %>%
#   filter(stage_at_diagnosis %in% c("IA1","IA3","IIB","IA2","IIIA","IB"))
#          
# test <-  clin_dat %>%
#   filter(stage_at_diagnosis %in% c("IA1","IA3","IIB","IA2","IIIA","IB"),
#          as.integer(PFS_days) > 150,
#          Sample_origin == "Primary tumor")
# table(test$broad_treatment)      
#          
#          
# exclude_ids <- clin_dat %>%
#   filter(stage_at_diagnosis %in% c("IA1","IA3","IIB","IA2","IIIA","IB"),
#          as.integer(PFS_days) > 150,
#          Sample_origin == "Primary tumor") %>%
#   pull(immucan_id)
# 
# clin_dat <- clin_dat %>%
#   filter(!immucan_id %in% exclude_ids)
```

# simple_stage
```{r}
clin_dat <- clin_dat %>%
  mutate(Stage = ifelse(stage_at_diagnosis %in% c("IB","IA2", "IA3"), "I", NA),
         Stage = ifelse(stage_at_diagnosis %in% c("IIB"), "II", Stage),
         Stage = ifelse(stage_at_diagnosis %in% c("IIIA","IIIB","IIIC"), "III", Stage),
         Stage = ifelse(stage_at_diagnosis %in% c("IVA","IVB"), "IV", Stage))
```

# age dichotomization
```{r}
clin_dat <- clin_dat %>%
  mutate(diagnosis_age = as.integer(diagnosis_age)) %>%
  mutate(Age_dicho = ifelse(diagnosis_age < mean(diagnosis_age,),"younger", "older"),
         young_patients = ifelse(diagnosis_age < 50,"<50",">50"))
```


# group sample origin
```{r}
clin_dat <- clin_dat %>%
  mutate(Tissue_origin = ifelse(!Sample_origin %in%  c("Distant metastases", "Primary tumor", "Regional lymph node metastases"),"Other", Sample_origin))
```


# duplicates
for some patients we actually have 2 samples. we still need to sort out which ones we will use.
```{r}
ids <- clin_dat[duplicated(clin_dat$immucan_id),] %>%
  pull(immucan_id)

# we will use the information from the sample selection to define which data to keep
clin_dat <- clin_dat %>%
  filter(Sample_id %in% keep_samp)
```

# driver mutations
many patients did not receive testing. we will add this information once we have it from the WES data
```{r}
clin_dat %>%
  mutate(driver_mutation = paste(EGFR_testing,PTEN_testing,ALK_testing, ROS1_testing, MET_testing,sep = "_")) %>%
  select(driver_mutation)
```

# clean up progression column names
```{r}
clin_dat <- clin_dat |>
  rename(local_progression = local,
         regional_progression = regional,
         distant_progression = distant,
         site_progression = site)
```


# clean up distant met location
```{r}
# fix one row
clin_dat <- clin_dat %>%
  mutate(Sample_origin = ifelse(Sample_origin == "ENDOBRONCHIAL BIOPSY FROM INTERMEDIATE BRONCHUS.", "Primary tumor",Sample_origin),
         Sample_origin = ifelse(Sample_origin == "FNA MEDIASTINAL LYMPH NODE", "Primary tumor",Sample_origin),
         Sample_origin = ifelse(Sample_origin == "Satellite/in-transit metastases", "Distant metastasis",Sample_origin),
         Sample_origin = ifelse(Sample_origin == "Local recurrence", "Regional lymph node metastases",Sample_origin),
         Sample_origin = ifelse(Sample_origin == "Distant metastasis", "Distant metastases",Sample_origin),
         Sample_origin = ifelse(Sample_origin == "PLEURAL METASTASE", "Distant metastases",Sample_origin),
         Sample_origin = ifelse(is.na(Sample_origin) & Sampling_site == "BRAIN, RIGHT CEREBELLUM", "Distant metastases",Sample_origin))

clin_dat <- clin_dat %>%
  mutate(Sample_location = ifelse(grepl("BRAIN|CEREBELLUM|CEREBELUM|CEREBELLAR|BULBAR|FRONTAL MASS|RIGHT PRECENTRAL LESION",Sampling_site),"Brain","Other"),
         Sample_location = ifelse(grepl("CERVICAL",Sampling_site),"Cervical_node",Sample_location),
         Sample_location = ifelse(grepl("LIVER|HEPATIC",Sampling_site),"Liver",Sample_location),
         Sample_location = ifelse(grepl("PLEURA",Sampling_site),"Pleura",Sample_location),
         Sample_location = ifelse(grepl("BONE|ULNA|OSTEOLITIC",Sampling_site),"Bone",Sample_location),
         Sample_location = ifelse(grepl("CUTANEOUS|SKIN|SUBCOUTANEOUS",Sampling_site),"Skin",Sample_location),
         Sample_location = ifelse(grepl("RIGHT ADRENAL|GLAND",Sampling_site),"Gland",Sample_location),
         Sample_location = ifelse(grepl("NODE",Sampling_site),"Lymph node",Sample_location),
         Sample_location = ifelse(Sample_origin == "Primary tumor","Lung",Sample_location),
         Sample_location = ifelse(Sample_origin == "Regional lymph node metastases","Lymph node",Sample_location))
clin_dat |>
  select(immucan_id,Sampling_site,Sample_origin, Sample_location) %>%
  filter(Sample_location == "Other")

table(clin_dat$Sample_location)
table(clin_dat$Sample_origin)
table(clin_dat$Sampling_site)
table(clin_dat$Sample_origin,clin_dat$Sample_location)
```


# clean up colnames
```{r}
colnames(clin_dat)
clin_dat <- clin_dat %>%
  rename(Sex = sex,
         Stage_at_diagnosis = stage_at_diagnosis,
         Young_patients = young_patients,
         Broad_treatment= broad_treatment)
```

# remove data
```{r}
clin_dat$sample_type <- NULL
clin_dat$T_category_yTNM <- NULL
clin_dat$N_category_yTNM <- NULL
clin_dat$M_category_yTNM <- NULL
clin_dat$yTNM_day <- NULL
clin_dat$HPV_test <- NULL
clin_dat$HER2_testing <- NULL
clin_dat$PTEN_testing <- NULL
clin_dat$Residual_cancer_burden <- NULL
clin_dat$commercial_use <- NULL
clin_dat$comment <- NULL
```


# uniqueness of data
```{r}
cat("unique patients: ",length(unique(clin_dat$immucan_id)), "\n")
cat("unique samples: ",length(unique(clin_dat$Sample_id)), "\n")
```

# export
```{r}
# export
write.table(clin_dat, "/mnt/immucan_volume/processed_data/NSCLC_cohort/data/clinical/20251205_clinical.txt",
            sep = "\t", row.names = FALSE)
```


# Driver addition from WES

```{r}
mut <- readRDS("/mnt/immucan_volume/processed_data/NSCLC_cohort/output/WES/Rout/selected_mutations.rds")

clin_dat <- left_join(clin_dat,mut,"immucan_id")
```

# MSI
```{r}
msi <- read.csv("/mnt/immucan_volume/processed_data/NSCLC_cohort/output/WES/Rout/NSCLC_msi.csv")
msi %>%
  filter(immucan_id %in% msi$immucan_id[duplicated(msi$immucan_id)])

clin_dat <- left_join(clin_dat,msi %>% select(immucan_id,MSI, msi_score), "immucan_id")
```

# TMB
```{r}
tmb <- read.csv("/mnt/immucan_volume/processed_data/NSCLC_cohort/output/WES/Rout/NSCLC_tmb.csv")

clin_dat <- left_join(clin_dat,tmb %>% select(immucan_id,TMB, TMB_score), "immucan_id")
```


# unique clinical data
note: there are patients from which we have more than one sample. We keep this information because we want to know the sample origin. However, for survival modeling, we need unique entries per patient. we will do this now.

```{r}
surv_table <- clin_dat %>%
  select(-Sample_id,Sampling_day,Sample_type, Sample_origin,Sampling_site, Tissue_origin)
```

# Save file
```{r}
write.table(surv_table, "/mnt/immucan_volume/processed_data/NSCLC_cohort/data/clinical/20251205_survival.txt",
            sep = "\t", row.names = FALSE)
```


# Curation
after generation of this file it was curated by Marie. We will merge with the curation

```{r}
curated <- as.data.frame(read_tsv("/mnt/immucan_volume/processed_data/NSCLC_cohort/data/clinical/20251205_survival_MM_curated.txt"))
surv_table <- as.data.frame(read_tsv("/mnt/immucan_volume/processed_data/NSCLC_cohort/data/clinical/20251205_survival.txt"))

for (i in colnames(surv_table)) {
  cat(tryCatch(identical(surv_table[i], curated[i])), i,"\n")
}

# only diagnosis stage changed

data.frame(old = surv_table$stage_at_registration,
           new = curated$stage_at_registration) %>%
  filter(old != new)

data.frame(old = surv_table$Stage_at_diagnosis,
           new = curated$Stage_at_diagnosis) %>%
  filter(!old == new)

curated %>% 
  filter(is.na(Stage)) %>%
  select(Stage, Stage_at_diagnosis, stage_at_registration)
## simple_stage
table(curated$Stage_at_diagnosis)
curated <- curated %>%
  mutate(Stage = ifelse(Stage_at_diagnosis %in% c("IB","IA2","IA1","IA", "IA3"), "I", NA),
         Stage = ifelse(Stage_at_diagnosis %in% c("IIA","IIB"), "II", Stage),
         Stage = ifelse(Stage_at_diagnosis %in% c("IIIA","IIIB","IIIC"), "III", Stage),
         Stage = ifelse(Stage_at_diagnosis %in% c("IVA","IVB", "IV"), "IV", Stage))
table(curated$Stage)

# save
write.table(curated, "/mnt/immucan_volume/processed_data/NSCLC_cohort/data/clinical/20251205_survival.txt",
            sep = "\t", row.names = FALSE)
```
