---
title: "mutation frequency investigation"
author: "Daniel Schulz"
date: "2026-01-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup, include = FALSE}
library(pander)
library(ggplot2)
theme_set(theme_minimal())
```
  
# Introduction

Whole exome sequencing data was processed by the "vegan-pipelin" for calling and annotation of genomic variants.  
  
In this notebook variants are filtered by following criteria:  
    - Only variants that passed "vegan" quality control are kept ("PASS" filter)
    - Only variants with a VAF of > 10% and a minimum of 3 reads covering the mutated allele are kept for analysis.
    - A selection of genes are taken from [this review](https://doi.org/10.1038/s41568-019-0179-8 (2019))
    - Only variants included in the COSMIC data base as annotated by the "vegan-pipeline" are kept for further analysis
    - Variants are further filtered to only include functionally relevant alterations for downstream analysis



# Libraries
```{r loading_libraries, message=FALSE}
library(data.table)
library(dplyr)
library(ComplexHeatmap)
library(tidyr)
library(ggtext)
```

# Variables and directories
```{r}
# The folowing firectories need to change according to the setup
mount_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/WES/data/"
path_output <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/WES/data/Rout/"


#following directories are set automatically
path_data_clin <- file.path( "/mnt/immucan_volume/clinical_data/Lung_NSCLC2/20230825/NSCLC2_for_R_update_2024.csv")
path_data_tmb <- file.path("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/WES/data/tmb")
path_vegan <-file.path("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/WES/data/tableReports") # results of main vegan run


# Define driver Genes
Genes_driver <- c("KRAS", "EGFR", "BRAF", "MET", "ALK", "ROS", "RET", "NTRK", "NRG1", "FGFR1", "FGFR2", "HRAS", "NRAS", "MAP2K1", "ERBB2", "ROS1", "PIK3CA", "NTRK1", "STK11")
Genes_sup <- c("TP53", "KEAP1","STK11","RB1","CDKN2A", "CDKN2B","NKX2", "NF1")

genes_sel <- c(Genes_driver, Genes_sup)
```


# Main Program with Plots and Results

## Load data

### Clinical

```{r clinical_data}
data_clin <- fread(path_data_clin)
data_clin
```

### VeganID to IMMUcanID
```{r}
vegan_lookup <- list.files(path_vegan, pattern = "design.csv", full.names = T)
vegan_lookup <- lapply(vegan_lookup, fread)
vegan_lookup <- do.call("rbind", vegan_lookup)
vegan_lookup[["immucan_id"]] <- gsub("-FIXT.*", "", vegan_lookup[["PAIR_ID"]])
vegan_lookup
```


### TMB

```{r}
data_tmb <- list.files(path_data_tmb, pattern = "tmb.txt", recursive = T, full.names = T)
data_tmb <- lapply(data_tmb, function(x){
  name <- gsub("_Mutect2.*", "", 
               gsub(".*tmb/", "", x)
               )
  data_tmp <- read.delim(x) %>%
    separate(col = 1, sep = "=", into = c("a", "b")) %>%
    filter(a == "TMB")
  
  data_tmp <- data_tmp[,"b",drop=T]
  data_tmp <- as.numeric(gsub(" ", "", data_tmp))

  
  names(data_tmp) <- name
  
  return(data_tmp)
  
  
})

data_tmb <-unlist(data_tmb)

data_tmb <- data.frame(TUMOR_ID = names(data_tmb),
           TMB = data_tmb)

data_tmb <- merge(data_tmb, vegan_lookup, all.x = T, by = "TUMOR_ID")

data_tmb
```

### Combine meta-data per sample
```{r}
data_meta <- merge(data_clin, data_tmb, by="immucan_id") # TMB data
data_meta
```

```{r}
data_mut <- list.files(path = path_vegan, 
                       pattern = "Mutect2_filtered_pass_norm_COSMIC_CancerHotspots_GnomAD_dbNSFP.tsv",
                       recursive = T,
                       full.names = T)


data_mut <- lapply(data_mut, function(path){
  
  name <- gsub("_.*", "", 
               gsub(".*tableReports/", "", path)
               )
  data_tmp <- fread(path)
  colnames(data_tmp) <- c(colnames(data_tmp)[1:14],"AD","DP","AF")
  data_tmp$TUMOR_ID <- name
  
  return(data_tmp)
  
})

data_mut <- do.call("rbind", data_mut)

data_mut <- merge(data_mut, data_meta, all.x = T, by = "TUMOR_ID")
data_mut <-  as.data.frame(data_mut)

# Format some column names
#colnames(data_mut) <- gsub("ANN\\[0\\]\\.", "", colnames(data_mut))
data_mut <- data_mut %>%
  transform(AF = as.numeric(AF),
            DP = as.numeric(DP))

hist(data_mut$AF)
```

# TP53
select all patients that do have TP53 mutation and those that do not
```{r}
TP53_pos <- data_mut %>%
  filter(Gene == "TP53") %>%
  pull(immucan_id)

data_mut %>%
  mutate(TP53_pos_patients = ifelse(immucan_id %in% TP53_pos,"TP53_mutated", "TP53_wildtype")) %>%
  ggplot(aes(x = TP53_pos_patients,y = DP))+
  geom_boxplot()+
  scale_y_log10()


data_mut %>%
  mutate(TP53_pos_patients = ifelse(immucan_id %in% TP53_pos,"TP53_mutated", "TP53_wildtype")) %>%
  ggplot(aes(x = TP53_pos_patients,y = DP))+
  geom_boxplot()+
  scale_y_log10()

data_mut %>%
  mutate(TP53_pos_patients = ifelse(immucan_id %in% TP53_pos,"TP53_mutated", "TP53_wildtype")) %>%
  ggplot(aes(x = TP53_pos_patients,y = DP))+
  geom_boxplot()+
  scale_y_log10()

```

Y220C mutation

```{r}
data_mut %>%
  filter(Gene == "TP53",
         grepl("220",HGVS_C)) 

NSCLC1 <- read.csv("/mnt/immucan_volume/processed_data/NSCLC_cohort/output/WES/Rout/mutations_all.csv")
NSCLC1 %>%
  filter(gene == "TP53")
         grepl("818",HGVS_C)) 
```

# PON files
all evidence points towards the fact that there are potentially more TP53 variations in the PONs and this will generate uncertainty for the variant caller.
We will check how many sites in the PONs overlap TP53 -> how many mutations were detected for TP53 in the PON?
Then repeat for KRAS.
If there are significantly more for TP53 then this supports that the idea that variants will not be called in the tumor if they are already in the PON.
```{r}
library(VariantAnnotation)
library(GenomicRanges)
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(org.Hs.eg.db)
pon_f <- readVcf("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/WES/data/PONs/pon_female.vcf.gz", genome = "hg38")
pon_m <-readVcf("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/WES/data/PONs/pon_male.vcf.gz", genome = "hg38")


txdb <- TxDb.Hsapiens.UCSC.hg38.knownGene
genes_gr <- genes(txdb)

tp53_id <- mapIds(org.Hs.eg.db, "TP53", "ENTREZID", "SYMBOL")
kras_id <- mapIds(org.Hs.eg.db, "KRAS", "ENTREZID", "SYMBOL")

tp53_gr <- genes_gr[names(genes_gr) == tp53_id]
kras_gr <- genes_gr[names(genes_gr) == kras_id]

c(
  female_tp53 = length(subsetByOverlaps(rowRanges(pon_f), tp53_gr)),
  male_tp53   = length(subsetByOverlaps(rowRanges(pon_m), tp53_gr)),
  female_kras = length(subsetByOverlaps(rowRanges(pon_f), kras_gr)),
  male_kras   = length(subsetByOverlaps(rowRanges(pon_m), kras_gr))
)

seqlevels(rowRanges(pon_f))[1:5]
seqlevels(tp53_gr)
seqlevels(pon_f)

length(rowRanges(pon_f))
length(rowRanges(pon_m))
table(seqnames(rowRanges(pon_f)))
table(seqnames(rowRanges(pon_m)))
```

