---
title: "MOFA bootstrapping"
author: "Daniel Schulz"
date: "2026-01-20"
output: html_document
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)

# Set path
if (Sys.info()["sysname"] == "Darwin"){
  mount_path <- "/Volumes/"
} else if (Sys.info()["sysname"] == "Linux"){
  mount_path <- "/mnt/"
} else {
  stop("Operating system is not supported")
}

# Load libraries
library(MOFA2)
library(reticulate)
# Potentially need to set the conda environment to use
#use_condaenv("MOFA")


library(dplyr)
library(readr)
library(stringr)


MOFA_INPUT_PATH <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/MOFA/"
CLINICAL_PATH <- "/mnt/immucan_volume/clinical_data/Lung_NSCLC2/20230825/NSCLC2_for_R_update_2024.csv"
out_folder <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/Revision"
```


# read in data

```{r}
# clinical data
clinical <- read.csv(CLINICAL_PATH) %>%
  dplyr::rename(Immucan_id = immucan_id) %>%
  mutate(Immucan_id = gsub("-", "_", Immucan_id))

# Helper function to z-score normalize
zscore <- function(mat) {
  t(scale(t(mat)))
}

# 1. Read data
cellohood <- read.csv(paste0(MOFA_INPUT_PATH, "/cellohood.csv"), row.names = 1)
densities <- read.csv(paste0(MOFA_INPUT_PATH, "/densities.csv"), row.names = 1)
expressions <- read.csv(paste0(MOFA_INPUT_PATH, "/expressions.csv"), row.names = 1)
interactions <- read.csv(paste0(MOFA_INPUT_PATH, "/interactions.csv"), row.names = 1)
binary <- read.csv(paste0(MOFA_INPUT_PATH, "/binary.csv"), row.names = 1)
RNA <- read.csv(paste0(MOFA_INPUT_PATH, "/RNA_named.csv"))

# 2. Set rownames from the first column (if applicable)
set_rownames <- function(df) {
  rownames(df) <- df$Immucan_id
  df$Immucan_id <- NULL
  return(df)
}

cellohood <- set_rownames(cellohood)
densities <- set_rownames(densities)
expressions <- set_rownames(expressions)
interactions <- set_rownames(interactions)
binary <- set_rownames(binary)
RNA <- set_rownames(RNA)
```

helper function
```{r}
preprocess_view <- function(df, 
                            log_transform = FALSE, 
                            arcsin_transform = FALSE, 
                            zscore_transform = TRUE,
                            filter_histology = FALSE) {
  mat <- as.matrix(df)
  
  if (filter_histology){
    mat <- mat[rownames(mat) %in% (clinical %>%
                 filter(simple_histology %in% c("Adenocarcinoma")) %>%
                 pull(Immucan_id)),]
  }
  
  # Optional log1p transform (for counts)
  if (log_transform) {
    mat <- log1p(mat)
  }
  
  # Optional arcsin sqrt transform (for proportions)
  if (arcsin_transform) {
    mat <- asin(sqrt(mat))
  }
  
  
  # Optional z-scoring
  if (zscore_transform) {
    mat <- scale(mat)
  }
  
  # Transpose to features x samples
  mat <- t(mat)
  return(mat)
}

```


# generate subsamples of patients
```{r}
all_pats <- Reduce(union,c(rownames(cellohood),rownames(densities),rownames(expressions),rownames(interactions),rownames(binary),rownames(RNA)))

set.seed(1235)
pats <- lapply(1:100, function(x){
  sample(all_pats,size = 0.8*length(all_pats), replace = FALSE)
})
names(pats) <- paste0("b",1:100)
```


```{r}
reticulate::use_condaenv("mofa_env")
b_list <- list()
for (i in paste0("b",1:100)) {

  filter_histology <- FALSE
  
  # we will filter each table by the selected sample of patients
  views <- list(
    Cellohood       = preprocess_view(cellohood[rownames(cellohood) %in% pats[[i]],], log_transform = TRUE, filter_histology=filter_histology),
    Densities       = preprocess_view(densities[ rownames(densities) %in% pats[[i]],], log_transform = TRUE, filter_histology=filter_histology),
    Interactions    = preprocess_view(interactions[ rownames(interactions) %in% pats[[i]],], filter_histology=filter_histology),
    Expressions     = preprocess_view(expressions[ rownames(expressions) %in% pats[[i]],], filter_histology=filter_histology),
    RNA             = preprocess_view(RNA[rownames(RNA) %in% pats[[i]],], filter_histology=filter_histology),
    BINARY = preprocess_view(binary[rownames(binary) %in% pats[[i]],], zscore_transform = FALSE, filter_histology=filter_histology)
  )
  
  
  # 4. Align samples across all views using union instead of intersection
  all_samples <- Reduce(union, lapply(views, colnames))
  
  # Fill in missing columns (samples) with NA
  views <- lapply(views, function(mat) {
    missing <- setdiff(all_samples, colnames(mat))
    if (length(missing) > 0) {
      empty_mat <- matrix(NA, nrow = nrow(mat), ncol = length(missing),
                          dimnames = list(rownames(mat), missing))
      mat <- cbind(mat, empty_mat)
    }
    # Reorder columns to match the union order
    mat[, all_samples]
  })
  
  # 5. Define likelihoods per view
  # Gaussian for continuous, Bernoulli for binary
  likelihoods <- c(
    "gaussian",   # Cellohood
    "gaussian",   # Densities
    "gaussian",   # Interactions
    "gaussian",   # Expressions (IMC)
    "gaussian",   # RNA
    "bernoulli"   # binary
  )
  
  # 6. Create MOFA object
  mofa <- create_mofa(views)
  
  # 7. Prepare options
  data_opts <- get_default_data_options(mofa)
  model_opts <- get_default_model_options(mofa)
  model_opts$num_factors <- 6
  training_opts <- get_default_training_options(mofa)
  training_opts$maxiter <- 1000
  
  # Assign likelihoods
  model_opts$likelihoods <- likelihoods
  
  
  # 8. Prepare and train MOFA
  mofa <- prepare_mofa(
    object = mofa,
    data_options = data_opts,
    model_options = model_opts,
    training_options = training_opts
  )
  
  # 9. Run MOFA
  mofa_trained <- run_mofa(mofa, use_basilisk = FALSE)
  b_list[[i]] <- MOFA2::get_factors(mofa_trained)
  #saveRDS(mofa_trained, model_path)
}

saveRDS(b_list,"/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/MOFA/bootstrapped_factor_list.rds")
```

# correlation of factors

we will use our original trained model with 100% of patients. factors can rotate
and thus do not necessarily have to rotate across runs. we will for each
bootstrap we will now correlate theeach factor with the original model and name
them based on the max absolute correlation. This will show if similar factors
where on average found across bootstrap runs.

```{r}
model <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/MOFA/mofa_V6_minor.rds")
org <- get_factors(model)

library(ggplot2)
library(tidyverse)

# we will check precisely how well factors where the same via diagonale
b_diag <- lapply(b_list, function(x){
  cur_org <- as.data.frame(org)[rownames(as.data.frame(x)),]
  colnames(cur_org) <- c(paste0("LF",c(1:6)))
  cur_bstrp <- as.data.frame(x)
  cor_dat <- cor(cur_org,cur_bstrp)
  
  diag_match <- abs(diag(cor_dat))
  return(diag_match)
})
cur_dat <- do.call(rbind,b_diag)
cur_dat <- as.data.frame(cur_dat)
colnames(cur_dat) <-  c(paste0("LF",c(1:6)))
cur_dat %>%
  pivot_longer(cols = everything()) %>%
  ggplot(aes(x = name, y = value))+
  geom_boxplot()+
  theme_bw()+
  ylab("Correlation LFs with bootstrapped LFs")+
  xlab("Factor correlation")

# we see that LF1, LF2 and L3 are mostly recovered with good median correlation.
# For the factors that explain less variance the correlation across bootstraps
# is small with the original factors

```

# Best matching original LF
Factor ordering is not necessarily stable in MOFA. Meaning, in a bootstrapped sample LF4 could resemble LF6 in the full data.
For this we will test if the maximum correlations between the original LFs per patient and the bootstrapped LFs.

```{r}
b_out <- lapply(b_list, function(x){
  cur_org <- as.data.frame(org)[rownames(as.data.frame(x)),]
  colnames(cur_org) <- c(paste0("LF",c(1:6)))
  cur_bstrp <- as.data.frame(x)
  cor_dat <- cor(cur_org,cur_bstrp)
  
  cur_cor <- apply(cor_dat, 1, function(x) (max(abs(x))))
  return(cur_cor)
})

# how robust
cur_dat <- do.call(rbind,b_out)
cur_dat %>%
  as.data.frame() %>%
  pivot_longer(cols = everything()) %>%
  ggplot(aes(x = name, y = value))+
  geom_boxplot()+
  theme_bw()+
  ylab("Correlation LFs with bootstrapped LFs")+
  xlab("LFs from all data")
```

# Best matching original LF
Factor ordering is not necessarily stable in MOFA. Meaning, in a bootstrapped sample LF4 could resemble LF6 in the full data.
For this we will test if the maximum correlations between the original LFs per patient and the bootstrapped LFs.

```{r}
b_out <- lapply(b_list, function(x){
  cur_org <- as.data.frame(org)[rownames(as.data.frame(x)),]
  colnames(cur_org) <- c(paste0("LF",c(1:6)))
  cur_bstrp <- as.data.frame(x)
  cor_dat <- cor(cur_org,cur_bstrp)
  cur_cor <- apply(cor_dat, 2, function(x) (max(abs(x))))
  return(cur_cor)
})

# how robust
cur_dat <- do.call(rbind,b_out)
cur_dat <- as.data.frame(cur_dat)
colnames(cur_dat) <-  c(paste0("LF",c(1:6)))

p1 <- cur_dat %>%
  as.data.frame() %>%
  pivot_longer(cols = everything()) %>%
  ggplot(aes(x = name, y = value))+
  geom_boxplot()+
  theme_bw()+
  ylab("max correlation with any original LF")+
  xlab("LFs from botstrapped data data")

p1
ggsave(plot = p1,
       filename = file.path(out_folder,"bootstraps_to_LF_correspondence.pdf"),
       device = "pdf",
       width = 5,
       height = 4)
```


# MOFA model factor correlation
the mofa factors do not correlate with each other


```{r}
org
cur_org <- as.data.frame(org)
colnames(cur_org) <- c(paste0("LF",c(1:6)))
pdf(file = file.path(out_folder,"LF_correlation.pdf"),
    width = 4,
    height = 4)
pheatmap::pheatmap(cor(cur_org,cur_org),cluster_rows = FALSE,cluster_cols = FALSE,main = "Latent factor correlation")
dev.off()
```


