---
title: "GO in smokers"
author: "Daniel Schulz"
date: "2026-02-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(ggpubr)
library(SingleCellExperiment)
library(edgeR)
library(cowplot)
library(ggrepel)
library(igraph)
library(ggraph)
library(ggrepel)
library(scales)
library(ggnewscale)

library(clusterProfiler)
library(org.Hs.eg.db)
library(dplyr)
```

```{r}
mount_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/"
out_folder <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/Figure_3_supp/"

if (!dir.exists(out_folder)) {
  dir.create(out_folder)
}
```

# load data
We will use the batch corrected protein coding gene counts to run differential expression analysis on
```{r}
RNA <- read.csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/RNASeq/Rout/tables/RNAseq/rawcounts_filtered_adj.txt", sep = "\t")

clin_dat <- read.csv("/mnt/immucan_volume/clinical_data/Lung_NSCLC2/20230825/NSCLC2_for_R_update_2024.csv")
traits <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/all_traits.csv"))


# variables
head(RNA)
```


# prepare data
```{r}
colnames(RNA) <- gsub("\\.","-",colnames(RNA))
colnames(RNA) <- str_split_i(colnames(RNA),"-FIXT",1)

# generate metadata
colnames(RNA) %in% traits$immucan_id
meta <- traits %>% filter(immucan_id %in% colnames(RNA))
```

# current vs never smoking all data

## differential abundance
```{r}
# rename
cur_meta <- meta %>%
  mutate(never_vs_current_smoker = ifelse(never_vs_current_smoker == "Non-smoker","Never", "Current"))
table(cur_meta$never_vs_current_smoker)
# subset RNA and metadata to contain only smokers and non-smokers
cur_meta <- cur_meta %>% filter(!is.na(never_vs_current_smoker))



group <- factor(cur_meta$never_vs_current_smoker, levels = c("Never", "Current"))
y <- DGEList(counts = RNA[,cur_meta$immucan_id], group = group)


keep <- filterByExpr(y, group = group)
y <- y[keep, , keep.lib.sizes = FALSE]

y <- calcNormFactors(y)
design <- model.matrix(~ group)
y <- estimateDisp(y, design)
fit <- glmQLFit(y, design)

qlf <- glmQLFTest(fit, coef = 2)
results <- topTags(qlf, n = Inf)$table

top_results <- topTags(qlf, n = 30)$table
head(results)
top_results
```

## get gene names
```{r}
# genecode annotations
annot <- read.csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/RNASeq/annotations/gencode.v29.annotation.genes.txt", sep = "\t")
#colnames(annot) <- c("gene_type","gene_id")
annot_type <- annot[,c("gene_id", "gene_type", "gene_name")]

results$gene_id <- rownames(results)
results <- left_join(results, annot_type,by = "gene_id")
top_results$gene_id <- rownames(top_results)
top_results <- left_join(top_results, annot_type,by = "gene_id")
```

## Vulcano plots
```{r}
ggplot(results,aes(x = logFC,y = -log10(PValue))) +
  geom_point()+
  theme_bw()+
  geom_label_repel(data = top_results,aes(x = logFC,y = -log10(PValue),label = gene_name))
```


## GO analysis
we will rank genes by the F st
```{r}
gene_list <- results$F * sign(results$logFC)
names(gene_list) <- results$gene_id
names(gene_list) <- str_split_i(names(gene_list),"\\.",1)
ranked_genes <- sort(gene_list, decreasing = TRUE)


# Keep only mapped genes# genes
ensembl2entrez <- bitr(
  names(ranked_genes),
  fromType = "ENSEMBL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
)

# Keep only mapped genes
ranked_genes <- ranked_genes[ensembl2entrez$ENSEMBL]



# gsea_res <- GSEA(
#   ranked_genes,
#   OrgDb = org.Hs.eg.db,
#   ont = "BP",          # Biological Process
#   keyType = "ENTREZID",
#   minGSSize = 10,
#   maxGSSize = 500,
#   pvalueCutoff = 0.05,
#   verbose = TRUE
# )
# 
# 
# pathway="GO_BP"
# term2gene <- read.delim("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/RNASeq/annotations/msigdbr/msigdbr_GO_BP_TERM2GENE.txt")
# set.seed(1234)
# res_GSEA <-    GSEA(geneList=ranked_genes,
#                     pvalueCutoff= 1,
#                     pAdjustMethod = "BH",
#                     minGSSize = 10,
#                     maxGSSize = 500,
#                     TERM2GENE = term2gene,
#                     seed = TRUE)
# #save(res_GSEA, file=paste0(out.tables, "GSEA_", pathway, "_signedP_", comp, ".RData"))
# res_GSEA_tb <- res_GSEA@result
# 
# 
# 
# dotplot(res_GSEA, showCategory=10, split=".sign") + facet_grid(.~.sign)

gse <- gseGO(geneList=ranked_genes, 
             ont ="ALL", 
             keyType = "ENSEMBL",
             minGSSize = 10, 
             maxGSSize = 500, 
             pvalueCutoff = 0.05, 
             verbose = TRUE, 
             OrgDb = org.Hs.eg.db, 
             pAdjustMethod = "none")

```

## dotplot
```{r, fig.height=10, fig.width=10}
dotplot(gse, showCategory=10, split=".sign") + facet_grid(.~.sign)
```


# current vs never smoking LUAD

## differential abundance
```{r}
# rename
cur_meta <- meta %>%
  mutate(never_vs_current_smoker = ifelse(never_vs_current_smoker == "Non-smoker","Never", "Current"))
table(cur_meta$never_vs_current_smoker)
# subset RNA and metadata to contain only smokers and non-smokers
cur_meta <- cur_meta %>% filter(!is.na(never_vs_current_smoker))

cur_meta <- cur_meta %>%
  filter(LUAD_vs_LUSC == "Adenocarcinoma")

group <- factor(cur_meta$never_vs_current_smoker, levels = c("Never", "Current"))
y <- DGEList(counts = RNA[,cur_meta$immucan_id], group = group)


keep <- filterByExpr(y, group = group)
y <- y[keep, , keep.lib.sizes = FALSE]

y <- calcNormFactors(y)
design <- model.matrix(~ group)
y <- estimateDisp(y, design)
fit <- glmQLFit(y, design)

qlf <- glmQLFTest(fit, coef = 2)
results <- topTags(qlf, n = Inf)$table

top_results <- topTags(qlf, n = 30)$table
head(results)
top_results
```

## get gene names
```{r}
# genecode annotations
annot <- read.csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/RNASeq/annotations/gencode.v29.annotation.genes.txt", sep = "\t")
#colnames(annot) <- c("gene_type","gene_id")
annot_type <- annot[,c("gene_id", "gene_type", "gene_name")]

results$gene_id <- rownames(results)
results <- left_join(results, annot_type,by = "gene_id")
top_results$gene_id <- rownames(top_results)
top_results <- left_join(top_results, annot_type,by = "gene_id")
inf_genes <- results %>%
  filter(gene_name %in% c("CXCL10", "CXCL9", "GZMB","GZMA","IL6","IRF1","STAT1","STAT3"))
```

## Vulcano plots
```{r}
ggplot(results,aes(x = logFC,y = -log10(PValue))) +
  geom_point()+
  theme_bw()+
  geom_label_repel(data = top_results,aes(x = logFC,y = -log10(PValue),label = gene_name))

ggplot(results,aes(x = logFC,y = -log10(PValue))) +
  geom_point()+
  theme_bw()+
  geom_label_repel(data = inf_genes,aes(x = logFC,y = -log10(PValue),label = gene_name), col = "red",nudge_x = 0.5)
```


## GO analysis
we will rank genes by the F st
```{r}
gene_list <- results$F * sign(results$logFC)
names(gene_list) <- results$gene_id
names(gene_list) <- str_split_i(names(gene_list),"\\.",1)
ranked_genes <- sort(gene_list, decreasing = TRUE)


# Keep only mapped genes# genes
ensembl2entrez <- bitr(
  names(ranked_genes),
  fromType = "ENSEMBL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
)

# Keep only mapped genes
ranked_genes <- ranked_genes[ensembl2entrez$ENSEMBL]


gse <- gseGO(geneList=ranked_genes, 
             ont ="ALL", 
             keyType = "ENSEMBL",
             minGSSize = 10, 
             maxGSSize = 500, 
             pvalueCutoff = 0.05, 
             verbose = TRUE, 
             OrgDb = org.Hs.eg.db, 
             pAdjustMethod = "none")

# hallmark genes
msig_hallmark <- msigdbr::msigdbr(
  species = "Homo sapiens",
  collection = "H")

gseH <- GSEA(
  ranked_genes,
  #OrgDb = org.Hs.eg.db,
  TERM2GENE = msig_hallmark[,c("gs_name","ensembl_gene")],
  minGSSize = 10,
  maxGSSize = 500,
  pvalueCutoff = 0.05,
  verbose = TRUE
)

```

## dotplot
```{r, fig.height=10, fig.width=10}
dotplot(gse, showCategory=10, split=".sign") + facet_grid(.~.sign)
dotplot(gseH, showCategory=10, split=".sign") + facet_grid(.~.sign)
```

## individual scatter plots for hallmark gene sets

```{r}
inflam_genes <- msig_hallmark %>%
  filter(gs_name == "HALLMARK_INFLAMMATORY_RESPONSE") %>%
  pull(gene_symbol)

tnf_genes <- msig_hallmark %>%
  filter(gs_name == "HALLMARK_TNFA_SIGNALING_VIA_NFKB") %>%
  pull(gene_symbol)

infg_genes <- msig_hallmark %>%
  filter(gs_name =="HALLMARK_INTERFERON_GAMMA_RESPONSE") %>%
  pull(gene_symbol)

il6_genes <- msig_hallmark %>%
  filter(gs_name =="HALLMARK_IL6_JAK_STAT3_SIGNALING") %>%
  pull(gene_symbol)


ggplot(results %>%
         filter(gene_name %in% inflam_genes),aes(x = logFC,y = -log10(PValue))) +
  geom_point()+
  theme_bw()+
  ggtitle("inflammatory hallmark")
ggplot(results %>%
         filter(gene_name %in% tnf_genes),aes(x = logFC,y = -log10(PValue))) +
  geom_point()+
  theme_bw()+
  ggtitle("tnf_genes hallmark")
ggplot(results %>%
         filter(gene_name %in% infg_genes),aes(x = logFC,y = -log10(PValue))) +
  geom_point()+
  theme_bw()+
  ggtitle("infg_genes hallmark")
ggplot(results %>%
         filter(gene_name %in% il6_genes),aes(x = logFC,y = -log10(PValue))) +
  geom_point()+
  theme_bw()+
  ggtitle("il6_genes hallmark")
```

## boxplots 

```{r}
RNA$gene_id <- rownames(RNA)
df <- left_join(RNA, annot_type,by = "gene_id")
df <- df %>%
  pivot_longer(cols = -c(gene_id, gene_type, gene_name), names_to = "immucan_id") %>%
  left_join(cur_meta, by= "immucan_id")
df <- df %>%
  mutate(hm_inflam = ifelse(gene_name %in% inflam_genes, TRUE, FALSE),
         hm_tnf = ifelse(gene_name %in% tnf_genes, TRUE, FALSE),
         hm_infg = ifelse(gene_name %in% inf_genes, TRUE, FALSE),
         hm_il6 = ifelse(gene_name %in% il6_genes, TRUE, FALSE))

```


# LUAD vs LUSC sanity check
## differential abundance
```{r}
# rename
cur_meta <- meta %>%
 filter(LUAD_vs_LUSC %in% c("Adenocarcinoma","Squamous cell carcinoma"))
table(cur_meta$LUAD_vs_LUSC)

cur_meta <- cur_meta %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma","LUAD","LUSC"))
group <- factor(cur_meta$LUAD_vs_LUSC, levels = c("LUAD", "LUSC"))
y <- DGEList(counts = RNA[,cur_meta$immucan_id], group = group)


keep <- filterByExpr(y, group = group)
y <- y[keep, , keep.lib.sizes = FALSE]

y <- calcNormFactors(y)
design <- model.matrix(~ group)
y <- estimateDisp(y, design)
fit <- glmQLFit(y, design)

qlf <- glmQLFTest(fit, coef = 2)
results <- topTags(qlf, n = Inf)$table

top_results <- topTags(qlf, n = 30)$table
head(results)
top_results
```

## get gene names
```{r}
# genecode annotations
annot <- read.csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/RNASeq/annotations/gencode.v29.annotation.genes.txt", sep = "\t")
#colnames(annot) <- c("gene_type","gene_id")
annot_type <- annot[,c("gene_id", "gene_type", "gene_name")]

results$gene_id <- rownames(results)
results <- left_join(results, annot_type,by = "gene_id")
top_results$gene_id <- rownames(top_results)
top_results <- left_join(top_results, annot_type,by = "gene_id")
```

## Vulcano plots
```{r}
ggplot(results,aes(x = logFC,y = -log10(PValue))) +
  geom_point()+
  theme_bw()+
  geom_label_repel(data = top_results,aes(x = logFC,y = -log10(PValue),label = gene_name))
```


## GO analysis
we will rank genes by the F st
```{r}
gene_list <- results$F * sign(results$logFC)
names(gene_list) <- results$gene_id
names(gene_list) <- str_split_i(names(gene_list),"\\.",1)
ranked_genes <- sort(gene_list, decreasing = TRUE)


# Keep only mapped genes# genes
ensembl2entrez <- bitr(
  names(ranked_genes),
  fromType = "ENSEMBL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
)

# Keep only mapped genes
ranked_genes <- ranked_genes[ensembl2entrez$ENSEMBL]



gse <- gseGO(geneList=ranked_genes, 
             ont ="ALL", 
             keyType = "ENSEMBL",
             minGSSize = 10, 
             maxGSSize = 500, 
             pvalueCutoff = 0.05, 
             verbose = TRUE, 
             OrgDb = org.Hs.eg.db, 
             pAdjustMethod = "none")

msig_hallmark <- msigdbr::msigdbr(
  species = "Homo sapiens",
  collection = "H")

gseH <- GSEA(
  ranked_genes,
  #OrgDb = org.Hs.eg.db,
  TERM2GENE = msig_hallmark[,c("gs_name","ensembl_gene")],
  minGSSize = 10,
  maxGSSize = 500,
  pvalueCutoff = 0.05,
  verbose = TRUE
)
```

## dotplot
```{r, fig.height=10, fig.width=10}
dotplot(gse, showCategory=10, split=".sign") + facet_grid(.~.sign)
dotplot(gseH, showCategory=10, split=".sign") + facet_grid(.~.sign)
```


# PDL1 sanity check
## differential abundance
```{r}
cur_meta <- meta %>%
 filter(TPS_score %in% c("positive","negative"))
group <- factor(cur_meta$TPS_score, levels = c("negative", "positive"))
y <- DGEList(counts = RNA[,cur_meta$immucan_id], group = group)

unique(cur_meta)

keep <- filterByExpr(y, group = group)
y <- y[keep, , keep.lib.sizes = FALSE]

y <- calcNormFactors(y)
design <- model.matrix(~ group)
y <- estimateDisp(y, design)
fit <- glmQLFit(y, design)

qlf <- glmQLFTest(fit, coef = 2)
results <- topTags(qlf, n = Inf)$table

top_results <- topTags(qlf, n = 30)$table
head(results)
top_results
```

## get gene names
```{r}
# genecode annotations
annot <- read.csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/RNASeq/annotations/gencode.v29.annotation.genes.txt", sep = "\t")
#colnames(annot) <- c("gene_type","gene_id")
annot_type <- annot[,c("gene_id", "gene_type", "gene_name")]

results$gene_id <- rownames(results)
results <- left_join(results, annot_type,by = "gene_id")
top_results$gene_id <- rownames(top_results)
top_results <- left_join(top_results, annot_type,by = "gene_id")
```

## Vulcano plots
```{r}
ggplot(results,aes(x = logFC,y = -log10(PValue))) +
  geom_point()+
  theme_bw()+
  geom_label_repel(data = top_results,aes(x = logFC,y = -log10(PValue),label = gene_name))
```


## GO analysis
we will rank genes by the F st
```{r}
gene_list <- results$F * sign(results$logFC)
names(gene_list) <- results$gene_id
names(gene_list) <- str_split_i(names(gene_list),"\\.",1)
ranked_genes <- sort(gene_list, decreasing = TRUE)


# Keep only mapped genes# genes
ensembl2entrez <- bitr(
  names(ranked_genes),
  fromType = "ENSEMBL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
)

# Keep only mapped genes
ranked_genes <- ranked_genes[ensembl2entrez$ENSEMBL]



gse <- gseGO(geneList=ranked_genes, 
             ont ="ALL", 
             keyType = "ENSEMBL",
             minGSSize = 10, 
             maxGSSize = 500, 
             pvalueCutoff = 0.05, 
             verbose = TRUE, 
             OrgDb = org.Hs.eg.db, 
             pAdjustMethod = "none")

msig_hallmark <- msigdbr::msigdbr(
  species = "Homo sapiens",
  collection = "H")

gseH <- GSEA(
  ranked_genes,
  #OrgDb = org.Hs.eg.db,
  TERM2GENE = msig_hallmark[,c("gs_name","ensembl_gene")],
  minGSSize = 10,
  maxGSSize = 500,
  pvalueCutoff = 0.05,
  verbose = TRUE
)
```

## dotplot
```{r, fig.height=10, fig.width=10}
dotplot(gse, showCategory=10, split=".sign") + facet_grid(.~.sign)
dotplot(gseH, showCategory=10, split=".sign") + facet_grid(.~.sign)
```


# tcga

## load clinical
```{r}
library(TCGAbiolinks)

query.clin <- GDCquery(
  project = "TCGA-LUAD",
  data.category = "Clinical",
  data.type = "Clinical Supplement",
  file.type = "xml"
)

GDCdownload(query.clin)

clinical <- GDCprepare_clinic(
  query.clin,
  clinical.info = "patient"
)

head(clinical)
clinical$tobacco_smoking_history

# Lifelong Non-smoker (less than 100 cigarettes smoked in Lifetime) = 1
# Current smoker (includes daily smokers and non-daily smokers or occasional smokers) = 2
# Current reformed smoker for > 15 years (greater than 15 years) = 3
# Current reformed smoker for ≤15 years (less than or equal to 15 years) = 4
# Current reformed smoker, duration not specified = 5
# Smoking History not documented = 7

table(clinical$tobacco_smoking_history)
```

## run diff analysis

```{r}
# # Load TCGA RNA data
tcga <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/MOFA/tcga_LUAD_LUSC.rds")

# Convert to SCE
sce <- as(tcga, "SingleCellExperiment")
sce$patient %in%
clinical$bcr_patient_barcode
sce <- sce[,sce$patient %in% clinical$bcr_patient_barcode]

### need to figure out which assay to use
assayNames(sce)
assay(sce, "tpm_unstrand")
RNA
```

# current vs former smoker
## differential abundance
```{r}
# rename
cur_meta <- meta %>%
  mutate(current_vs_Ex_smoker = ifelse(current_vs_Ex_smoker == "Ex-smoker","Former", "Current"))
table(cur_meta$current_vs_Ex_smoker)
# subset RNA and metadata to contain only smokers and non-smokers
cur_meta <- cur_meta %>% filter(!is.na(current_vs_Ex_smoker))



group <- factor(cur_meta$current_vs_Ex_smoker, levels = c("Current", "Former"))
y <- DGEList(counts = RNA[,cur_meta$immucan_id], group = group)


keep <- filterByExpr(y, group = group)
y <- y[keep, , keep.lib.sizes = FALSE]

y <- calcNormFactors(y)
design <- model.matrix(~ group)
y <- estimateDisp(y, design)
fit <- glmQLFit(y, design)

qlf <- glmQLFTest(fit, coef = 2)
results <- topTags(qlf, n = Inf)$table

top_results <- topTags(qlf, n = 30)$table
head(results)
top_results
```

## get gene names
```{r}
# genecode annotations
annot <- read.csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/RNASeq/annotations/gencode.v29.annotation.genes.txt", sep = "\t")
#colnames(annot) <- c("gene_type","gene_id")
annot_type <- annot[,c("gene_id", "gene_type", "gene_name")]

results$gene_id <- rownames(results)
results <- left_join(results, annot_type,by = "gene_id")
top_results$gene_id <- rownames(top_results)
top_results <- left_join(top_results, annot_type,by = "gene_id")
```

## Vulcano plots
```{r}
ggplot(results,aes(x = logFC,y = -log10(PValue))) +
  geom_point()+
  theme_bw()+
  geom_label_repel(data = top_results,aes(x = logFC,y = -log10(PValue),label = gene_name))
```


## GO analysis
we will rank genes by the F st
```{r}
gene_list <- results$F * sign(results$logFC)
names(gene_list) <- results$gene_id
names(gene_list) <- str_split_i(names(gene_list),"\\.",1)
ranked_genes <- sort(gene_list, decreasing = TRUE)


# Keep only mapped genes# genes
ensembl2entrez <- bitr(
  names(ranked_genes),
  fromType = "ENSEMBL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
)

# Keep only mapped genes
ranked_genes <- ranked_genes[ensembl2entrez$ENSEMBL]

gse <- gseGO(geneList=ranked_genes, 
             ont ="ALL", 
             keyType = "ENSEMBL",
             minGSSize = 10, 
             maxGSSize = 500, 
             pvalueCutoff = 0.05, 
             verbose = TRUE, 
             OrgDb = org.Hs.eg.db, 
             pAdjustMethod = "none")

```

## dotplot
```{r, fig.height=10, fig.width=10}
dotplot(gse, showCategory=10, split=".sign") + facet_grid(.~.sign)
```



# Trm signature
we will check if the T-resident memory signature is different between smokers and never smokers. this has been reported by [Weeden et al](https://doi.org/10.1016/j.ccell.2023.03.019)
the gene list was stored as a supplemental and is now located in the pdf `/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/RNASeq/TRM_signature_from_Corgnac_et_al.pdf` and we will construt the expression from this

```{r}
cur_RNA <- RNA
cur_RNA$gene_name <- annot[match(rownames(RNA), annot$gene_id),"gene_name"]
cur_RNA <- cur_RNA %>%
  filter(!duplicated(gene_name))
rownames(cur_RNA) <- cur_RNA$gene_name
cur_RNA$gene_name <- NULL
trm_sig <- data.frame(gene = c("HASPIN","GPR34","ITGAE","CXCR6","ACP5","LDLRAD4","JAML","ATP10D","IFNG","ABTB2","QPCT","XCL1","CD244","KCNN4","RACGAP1",
                               "MAPKAPK3","P2RY10","CCL5","PCED1B","VOPP1","ABI3","DOCK2",
                               "CDC25B","LITAF","NEDD4","BCL9L","STK38","RIPOR2","SAMHD1","FRMD4B","LEF1","PDE2A","ICAM2","RASA3","EOMES","KLF2","TCF7","FAM49A",
                               "KLF3","S1PR1","S1PR5"),
                      direction = c(rep("up",22),rep("down",19)))
rankData <- rankGenes(cur_RNA)

df <- simpleScore(rankData, downSet = trm_sig$gene[trm_sig$direction == "down"], upSet = trm_sig$gene[trm_sig$direction == "up"])

df$immucan_id <- rownames(df)
df <- left_join(df,clin_dat, "immucan_id")

df %>%
  filter(smoking_status %in% c("Current smoker","Non-smoker")) %>%
  ggplot(aes(x = smoking_status,TotalScore))+
  stat_compare_means(method = "wilcox")+
  geom_boxplot()+
  theme_bw()+
  ylab("Trm score")
df %>%
  ggplot(aes(x = smoking_status,UpScore))+
  geom_boxplot()
df %>%
  ggplot(aes(x = smoking_status,DownScore))+
  geom_boxplot()
```




```{r}
ranked_data <- rankGenes(sce)

top_n <- 10
p.list <- list()
gene_df <- data.frame()
for (f in seq(1,6,1)){
  
  # Extract weights for RNA view and Factor 1
  rna_weights <- get_weights(model, view = "RNA", factors = f)$RNA[,1]
  # filter for the ones present in the tcga data
  rna_weights <- rna_weights[names(rna_weights) %in% rownames(sce)]
  top_abs_features <- names(sort(abs(rna_weights), decreasing = TRUE))[1:top_n]
  
  signature_mofa_f1 <- list(
    up = top_abs_features[rna_weights[top_abs_features] > 0],
    down = top_abs_features[rna_weights[top_abs_features] < 0]
  )
  temp <- data.frame(Factor = rep(paste0("Factor",f),length(top_abs_features)),
                     gene = top_abs_features,
                     weight = rna_weights[top_abs_features])
  gene_df <- rbind(gene_df, temp)
  if (length(signature_mofa_f1$up) <= 2){
    print("WARNING: Score is computed inverted and is negated")
    score <- simpleScore(ranked_data, upSet = signature_mofa_f1$down)
    # Merge with clinical data
    survival_df <- data.frame(
      ID = rownames(colData(sce)),
      survival = sce$survival,
      days_to_death = sce$days_to_death,
      days_to_last_follow_up = sce$days_to_last_follow_up,
      score = -score$TotalScore,
      Histology = sce$histology
    )
  }else if(length(signature_mofa_f1$down) <= 2){
    score <- simpleScore(ranked_data, upSet = signature_mofa_f1$up)
    # Merge with clinical data
    survival_df <- data.frame(
      ID = rownames(colData(sce)),
      survival = sce$survival,
      days_to_death = sce$days_to_death,
      days_to_last_follow_up = sce$days_to_last_follow_up,
      score = score$TotalScore,
      Histology = sce$histology
    )
  }else{
    score <- simpleScore(ranked_data, downSet = signature_mofa_f1$down, upSet = signature_mofa_f1$up)
    # Merge with clinical data
    survival_df <- data.frame(
      ID = rownames(colData(sce)),
      survival = sce$survival,
      days_to_death = sce$days_to_death,
      days_to_last_follow_up = sce$days_to_last_follow_up,
      score = score$TotalScore,
      score_up = score$UpScore,
      score_down = score$DownScore,
      Histology = sce$histology
    )
  }
```










# current vs never smoker LUAD
## differential abundance
```{r}
# rename

# subset RNA and metadata to contain only smokers and non-smokers
cur_meta <- cur_meta %>% filter(!is.na(never_vs_Ex_smoker),
                                LUAD_vs_LUSC == "Adenocarcinoma")



group <- factor(cur_meta$never_vs_Ex_smoker, levels = c("Never", "Former"))
y <- DGEList(counts = RNA[,cur_meta$immucan_id], group = group)


keep <- filterByExpr(y, group = group)
y <- y[keep, , keep.lib.sizes = FALSE]

y <- calcNormFactors(y)
design <- model.matrix(~ group)
y <- estimateDisp(y, design)
fit <- glmQLFit(y, design)

qlf <- glmQLFTest(fit, coef = 2)
results <- topTags(qlf, n = Inf)$table

top_results <- topTags(qlf, n = 30)$table
head(results)
top_results
```

## get gene names
```{r}
# genecode annotations
annot <- read.csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/RNASeq/annotations/gencode.v29.annotation.genes.txt", sep = "\t")
#colnames(annot) <- c("gene_type","gene_id")
annot_type <- annot[,c("gene_id", "gene_type", "gene_name")]

results$gene_id <- rownames(results)
results <- left_join(results, annot_type,by = "gene_id")
top_results$gene_id <- rownames(top_results)
top_results <- left_join(top_results, annot_type,by = "gene_id")
inf_genes <- results %>%
  filter(gene_name %in% c("CXCL10", "CXCL9", "GZMB","GZMA","IL6","IRF1","STAT1","STAT3"))
```

## Vulcano plots
```{r}
ggplot(results,aes(x = logFC,y = -log10(PValue))) +
  geom_point()+
  theme_bw()+
  geom_label_repel(data = top_results,aes(x = logFC,y = -log10(PValue),label = gene_name))

ggplot(results,aes(x = logFC,y = -log10(PValue))) +
  geom_point()+
  theme_bw()+
  geom_label_repel(data = inf_genes,aes(x = logFC,y = -log10(PValue),label = gene_name), col = "red",nudge_x = 0.5)
```


## GO analysis
we will rank genes by the F st
```{r}
gene_list <- results$F * sign(results$logFC)
names(gene_list) <- results$gene_id
names(gene_list) <- str_split_i(names(gene_list),"\\.",1)
ranked_genes <- sort(gene_list, decreasing = TRUE)

# Keep only mapped genes# genes
ensembl2entrez <- bitr(
  names(ranked_genes),
  fromType = "ENSEMBL",
  toType = "ENTREZID",
  OrgDb = org.Hs.eg.db
)

# Keep only mapped genes
ranked_genes <- ranked_genes[ensembl2entrez$ENSEMBL]

gse <- gseGO(geneList=ranked_genes, 
             ont ="ALL", 
             keyType = "ENSEMBL",
             minGSSize = 10, 
             maxGSSize = 500, 
             pvalueCutoff = 0.05, 
             verbose = TRUE, 
             OrgDb = org.Hs.eg.db, 
             pAdjustMethod = "BH")

```

## dotplot
```{r, fig.height=10, fig.width=10}
dotplot(gse, showCategory=10, split=".sign") + facet_grid(.~.sign)
```


```{r}
table(traits$smoking_status,traits$LUAD_vs_LUSC)
```


# SBS based diff expression


```{r}
# Merge with mutational signatures
Lung_NSCLC2_filtered <-read.delim(paste0(mount_path,"data/WES/mutation_signatures/Lung_NSCLC2_filtered.tsv"), sep="\t")
names(Lung_NSCLC2_filtered)[names(Lung_NSCLC2_filtered)=="patient"] <- "immucan_id"
anno_WES <- merge(traits, Lung_NSCLC2_filtered, by="immucan_id", sort=F)

# Combine smoking signatures
anno_WES[,"SBS4_SBS92"] <- apply(anno_WES[,c("SBS4","SBS92")],1,sum)

# Dichotomise smoking signature
cutoff <- median(anno_WES$SBS4_SBS92)
anno_WES <- anno_WES %>%
      filter(!is.na(SBS4_SBS92))  %>%
      mutate(SBS4_SBS92_dicho = ifelse(SBS4_SBS92 > cutoff, "high","low"))

anno_WES$SBS4_SBS92_dicho

cur_dat <- read.table(paste0(mount_path,"data/RNASeq/CURIE/inflammation_quantification_NSCLC2/outputs/pathwayActivity/PathwayActivity_NSCLC2_mlm"))
my_comp <-  list( c("high", "low"))

cur_dat <- cur_dat %>%
  as.data.frame() %>%
  mutate(immucan_id = paste0("LUNG-",rownames(cur_dat))) %>%
  left_join(anno_WES,by = "immucan_id")

p <- cur_dat %>%
  pivot_longer(cols = c(JAK.STAT,NFkB,TNFa)) %>%
  filter(!is.na(SBS4_SBS92_dicho)) %>%
  ggplot(aes(x = SBS4_SBS92_dicho,y = value, fill = SBS4_SBS92_dicho))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(color = SBS4_SBS92_dicho),position = position_jitterdodge(jitter.width = 0.2), size= 1)+
  facet_wrap(.~name)+
  scale_fill_manual(values = alpha(c("blue", "lightblue", "cyan3"),alpha = 0.5))+
  scale_color_manual(values = c("blue", "lightblue", "cyan3"))+
  stat_compare_means(size = 5,label = "p.signif", comparisons = my_comp,method = "t.test")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5,hjust = 1, size = 16),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 14),
        legend.text = element_text(size = 14),
        strip.text.x = element_text(size = 14),
        legend.position = "bottom",
        legend.location = "plot")+
  ylab("Pathway score")

p

cur_dat %>%
  ggplot(aes(x = TNFa,y = SBS4_SBS92))+
  geom_point()
cor(cur_dat$TNFa,cur_dat$SBS4)
```

# inflammation signature for smoking within LUAD

```{r}
cur_dat <- read.table(paste0(mount_path,"data/RNASeq/CURIE/inflammation_quantification_NSCLC2/outputs/pathwayActivity/PathwayActivity_NSCLC2_mlm"))
my_comp <-  list( c("Current smoker", "Non-smoker"), c("Ex-smoker", "Non-smoker"), c("Current smoker", "Non-smoker") )

cur_dat <- cur_dat %>%
  as.data.frame() %>%
  mutate(immucan_id = paste0("LUNG-",rownames(cur_dat))) %>%
  left_join(traits,by = "immucan_id") %>%
  filter(LUAD_vs_LUSC == "Adenocarcinoma") 

cur_dat <- cur_dat %>%
  pivot_longer(cols = c(JAK.STAT,NFkB,TNFa))
p <- cur_dat %>%
  filter(!is.na(smoking_status)) %>%
  ggplot(aes(x = smoking_status,y = value, fill = smoking_status))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(color = smoking_status),position = position_jitterdodge(jitter.width = 0.2), size= 1)+
  facet_wrap(.~name)+
  scale_fill_manual(values = alpha(c("blue", "lightblue", "cyan3"),alpha = 0.5))+
  scale_color_manual(values = c("blue", "lightblue", "cyan3"))+
  stat_compare_means(size = 5,label = "p.signif", comparisons = my_comp,method = "t.test")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5,hjust = 1, size = 16),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 14),
        legend.text = element_text(size = 14),
        strip.text.x = element_text(size = 14),
        legend.position = "bottom",
        legend.location = "plot")+
  ylab("Pathway score")

p
```

# inflammation signature for LUAD vs LUSC

```{r}
cur_dat <- read.table(paste0(mount_path,"data/RNASeq/CURIE/inflammation_quantification_NSCLC2/outputs/pathwayActivity/PathwayActivity_NSCLC2_mlm"))
my_comp <-  list( c("Adenocarcinoma", "Squamous cell carcinoma"))

cur_dat <- cur_dat %>%
  as.data.frame() %>%
  mutate(immucan_id = paste0("LUNG-",rownames(cur_dat))) %>%
  left_join(traits,by = "immucan_id") %>%
  filter(LUAD_vs_LUSC %in% c("Adenocarcinoma", "Squamous cell carcinoma")) 

cur_dat <- cur_dat %>%
  pivot_longer(cols = c(JAK.STAT,NFkB,TNFa))
p <- cur_dat %>%
  filter(!is.na(LUAD_vs_LUSC)) %>%
  ggplot(aes(x = LUAD_vs_LUSC,y = value, fill = LUAD_vs_LUSC))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(color = LUAD_vs_LUSC),position = position_jitterdodge(jitter.width = 0.2), size= 1)+
  facet_wrap(.~name)+
  scale_fill_manual(values = alpha(c("blue", "lightblue", "cyan3"),alpha = 0.5))+
  scale_color_manual(values = c("blue", "lightblue", "cyan3"))+
  stat_compare_means(size = 5,label = "p.signif",comparisons = my_comp,method = "t.test")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5,hjust = 1, size = 16),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 14),
        legend.text = element_text(size = 14),
        strip.text.x = element_text(size = 14),
        legend.position = "bottom",
        legend.location = "plot")+
  ylab("Pathway score")+
  coord_cartesian(ylim = c(-3,5))

p
```

