---
title: "Revision_Daniel"
author: "Daniel Schulz"
date: "2026-01-19"
output: html_document
---

In this script we will investigate referee comments for revision


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(SingleCellExperiment)
library(tidyverse)
library(ggplot2)
library(ComplexHeatmap)#
library(pals)
library(dendextend)
```

# abundance of CNs across patients

How commen are the CNs?
Do we observe them in each patient or are there very patient specific CNs
Are some of the observed statistically significant differences based on outliers?

```{r, fig.height=20, fig.width=20}
# Create table with celltypes fractions
mount_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/"
cello_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/cello_dat.rds"))
cello_dens <- read.csv(paste0(mount_path,"Figures/scores_for_sharing/cellohood_density_cellbased_p1.csv"))

traits <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/all_traits.csv"))


cello_dat <- cello_dens %>%
  select(contains(c("immucan_id","CN")))

cur_df <- cello_dat[,-1]/rowSums(cello_dat[,-1])
cur_df$immucan_id <- cello_dat$immucan_id


dat <- left_join(cur_df,traits,"immucan_id")

# now we make a matix and cluster based on cell type fractions
matrixrownames <- dat$immucan_id 

# now we create a matrix from the data and cluster the data based on the cell fractions
hm_dat = as.matrix(dat[,grepl("CN",colnames(dat))])

rownames(hm_dat) <- as.character(matrixrownames)

# calculate distance and then cluster images based on cluster fraction
set.seed(1325)
dd <- dist((hm_dat), method = "euclidean")
hc <- hclust(dd, method = "ward.D2")
row_sorted <- hc$labels


mrownames <- traits$immucan_id
traits <- as.matrix(traits)
rownames(traits) <- mrownames
traits <- data.frame(traits[row_sorted,])

# generate the barplot. this is generated as the annotation for the heatmap of the Patient_ID that is generated below.
hm_dat <- hm_dat[row_sorted,]
colnames(hm_dat) <- paste0("CN_",colnames(hm_dat))
# col vector for CN_knn30_15_minor_celltypes
col_vec <- c(alphabet(n = 26),alphabet2(n = 10))
col_vector = structure(col_vec,
                       names= colnames(hm_dat))

# generate dendrogram
dend <- as.dendrogram(hc)
#dend <- color_branches(dend, h = 1, col = glasbey(n = 5), groupLabels = TRUE) # `color_branches()` returns a dendrogram object
#dend <- color_branches(dend, col = glasbey(n = 5),clusters = )

metaclusters <- cutree(hc,h=0.8)[order.dendrogram(dend)]
#define color vector 
col_vector_metacluster <- structure(glasbey(n=length(unique(metaclusters))),
                                    names = unique(metaclusters))

dend <- dend %>% branches_attr_by_clusters(metaclusters,values = col_vector_metacluster)


traits$metaclusters = metaclusters[traits$immucan_id]
# 

col_vector_Histology <- structure(c(glasbey()[6:11]), 
                                 names = c("Adenocarcinoma", "Squamous cell carcinoma","Adenosquamous carcinoma","LCNEC","Other","unknown"))

col_vector_simple_stage <- structure(c(glasbey()[12:15],glasbey()[16]), 
                                 names = c("I", "II", "III","IV","Unknown"))

col_vector_binary_stage <- structure(c("blue","red"), 
                                 names = c("high_stage","low_stage"))

col_vector_gender <- structure(c("purple","yellow"), 
                                 names = c("Male","Female"))

col_vector_Death <- structure(c("black","white"), 
                                 names = c(1,0))

col_vector_Smoker <- structure(glasbey()[5:7], 
                                 names = c("Current smoker","Ex-smoker","Non-smoker"))

ha <- rowAnnotation(Stage = traits[,c("simple_stage")],
                    Histology = traits[,c("simple_histology")],
                    Sex = traits[,"gender"],
                    Death = traits[,"censoring"],
                    Smoker = traits[,"smoking_status"],
                    `Cell Type Proportion` =
                      anno_barplot(hm_dat, 
                                   gp=gpar(fill=col_vector),
                                   bar_width = 1,
                                   height = unit(25,"cm"),
                                   width = unit(11,"cm"),
                                   show_row_names = FALSE),
                        col = list(Stage = col_vector_simple_stage,
                                   Histology = col_vector_Histology,
                                   Sex = col_vector_gender,
                                   Death = col_vector_Death,
                                   Smoker = col_vector_Smoker),
                        show_legend = FALSE)




# heatmap consisting of the patient_IDs. one color per patient
h1 = Heatmap(traits[,c("metaclusters")], 
             col = col_vector_metacluster, 
             width = unit(0.5, "cm"), 
             cluster_rows = dend,
             row_dend_width = unit(1, "cm"),
             height = unit(25, "cm"),
             show_heatmap_legend = FALSE, 
             heatmap_legend_param = list(title = "Stage"),
             row_names_gp = gpar(cex=3),
             show_row_names = TRUE,
             right_annotation =  ha,
             column_labels = "Metacluster")

h1
# add legend
lgd1 <- Legend(labels = names(col_vector), 
             title = "Cellohood 1", 
             legend_gp = gpar(fill = col_vector),
             ncol = 1)

lgd2 <- Legend(labels = names(col_vector_metacluster),
               legend_gp = gpar(fill = unname(col_vector_metacluster)),
               title = "Metacluster",
               ncol = 1)

lgd3 <- Legend(labels = names(col_vector_simple_stage), 
             legend_gp = gpar(fill = unname(col_vector_simple_stage)),
             title = "Stage",
             ncol = 1)

lgd4 <- Legend(labels = names(col_vector_Histology), 
             legend_gp = gpar(fill = unname(col_vector_Histology)),
             title = "Histology",
             ncol = 1)

lgd5 <- Legend(labels = names(col_vector_gender), 
             legend_gp = gpar(fill = unname(col_vector_gender)),
             title = "Sex",
             ncol = 1)

lgd6 <- Legend(labels = names(col_vector_Death), 
             legend_gp = gpar(fill = unname(col_vector_Death)),
             title = "Death",
             ncol = 1)


lgd7 <- Legend(labels = names(col_vector_Smoker), 
             legend_gp = gpar(fill = unname(col_vector_Smoker)),
             title = "Smoker",
             ncol = 1)

pd = packLegend(lgd1, lgd2, lgd3, lgd4,lgd5,lgd6 , lgd7, direction = "vertical", max_height = unit(15, "cm"),
    column_gap = unit(0.5, "cm"))

# plot the data
ht = grid.grabExpr(draw(h1))
grid.newpage()
pushViewport(viewport(x=unit(0.35,"npc"),y=unit(0.55,"npc"),angle = 270))
grid.draw(ht)
pushViewport(viewport(x = 0.5,y=0.9,angle = 90))
grid.draw(pd)
```

distribution of CNs per patient
```{r, fig.height=10, fig.width=  10}
cur_df %>%
  pivot_longer(cols = -immucan_id) %>%
  #filter(name == "CN_35__cello1") %>%
  ggplot(aes(x = value))+
  geom_histogram()+
  facet_wrap(.~name, scales = "free")
```

how many CNs are found at least in 10, 20, 30, 50 etc percent of patients
We only count the CNs if they have a presence of 1 %

```{r, fig.height=10}
plot_dat <- left_join(cur_df, traits,"immucan_id")
plot_dat <- plot_dat %>%
  #filter(simple_histology == "Adenocarcinoma") %>%
  select(contains(c("immucan_id","CN")))
p.list <- list()
for (i in c(seq(0.01,0.1,0.01))) {
  p.list[[as.character(i)]] <- plot_dat %>%
    
    summarise(across(.cols = -immucan_id,.fns = function(x) x > i)) %>%
    summarise(across(.cols = everything(),function(x) sum(x)/nrow(plot_dat)))

}

cur_dat <- do.call(rbind, p.list)

cur_dat %>%
  mutate(fraction = rownames(cur_dat)) %>%
  pivot_longer(-fraction) %>%
  #filter(name == "CN_5__cello1") %>%
  ggplot(aes(x = fraction, y = value))+
  geom_point()+
  facet_wrap(.~name)+
  geom_hline(yintercept = 0.5, col = "darkgreen")+
  geom_hline(yintercept = 0.1, col = "red")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  xlab(label = "Minimal fraction the spatial cluster needs to be present in a patient")+
  ylab("Fraction of patients with at least the minimal amount of the cluster")
```
LUSC
```{r, fig.height=10}
plot_dat <- left_join(cur_df, traits,"immucan_id")
plot_dat <- plot_dat %>%
  filter(simple_histology == "Squamous cell carcinoma") %>%
  select(contains(c("immucan_id","CN")))
p.list <- list()
for (i in c(seq(0.01,0.1,0.01))) {
  p.list[[as.character(i)]] <- plot_dat %>%
    
    summarise(across(.cols = -immucan_id,.fns = function(x) x > i)) %>%
    summarise(across(.cols = everything(),function(x) sum(x)/nrow(plot_dat)))

}

cur_dat <- do.call(rbind, p.list)

cur_dat %>%
  mutate(fraction = rownames(cur_dat)) %>%
  pivot_longer(-fraction) %>%
  #filter(name == "CN_5__cello1") %>%
  ggplot(aes(x = fraction, y = value))+
  geom_point()+
  facet_wrap(.~name)+
  geom_hline(yintercept = 0.5, col = "darkgreen")+
  geom_hline(yintercept = 0.1, col = "red")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  xlab(label = "Minimal fraction the spatial cluster needs to be present in a patient")+
  ylab("Fraction of patients with at least the minimal amount of the cluster")
```

box plots
```{r}
plot_dat <- left_join(cur_df, traits,"immucan_id")
w_test <- plot_dat %>%
  select(contains(c("immucan_id","CN"))) %>%
  pivot_longer(cols = -immucan_id) %>%
  left_join(traits,"immucan_id") %>%
  filter(simple_histology %in% c("Adenocarcinoma","Squamous cell carcinoma")) %>%
  group_by(name) %>%
  nest() %>%
  mutate(
    # Wilcoxon test
    test = map(data, ~ wilcox.test(value ~ simple_histology, data = .x)),
    summary = map(test, broom::tidy),
    
    # Compute medians for fold change
    medians = map(data, ~ .x %>%
                    group_by(simple_histology) %>%
                    summarise(median = median(value, na.rm = TRUE), .groups = "drop") %>%
                    pivot_wider(names_from = simple_histology, values_from = median)
                 ),
    
    # Compute fold change: TRUE / FALSE
    #fold_change = map_dbl(medians, ~ .x$Adenocarinoma / .x$`Squamous cell carcinoma`)
    LUAD_frac = map_dbl(medians, ~ .x$Adenocarcinoma),
    LUSC_frac = map_dbl(medians, ~ .x$`Squamous cell carcinoma`)
  ) %>%
  select(name, summary, LUAD_frac, LUSC_frac) %>%
  unnest(summary)

w_test <- w_test %>%
  mutate(test = "histology",
         sig = p.value < 0.05) 

w_test %>%
  filter(sig == TRUE)
```


```{r, fig.height=8, fig.width=15}
plot_dat %>%
  filter(!is.na(LUAD_vs_LUSC)) %>%
  pivot_longer(contains("CN")) %>%
  mutate(value = value) %>%
  ggplot(aes(x = name,y = value,fill = LUAD_vs_LUSC))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(color = LUAD_vs_LUSC),position = position_jitterdodge(jitter.width = 0.2), size = 0.2)+
  scale_fill_manual(values = alpha(c("red", "blue", "gray"),alpha = 0.6))+
  scale_color_manual(values = c("red", "blue", "gray"))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = .5, hjust = 1))+
  coord_cartesian(ylim = c(0,0.25))
```
# fractions and densities for EGFR

```{r, fig.height=8, fig.width=15}
plot_dat <- left_join(cur_df, traits,"immucan_id")
w_test <- plot_dat %>%
  select(contains(c("immucan_id","CN"))) %>%
  pivot_longer(cols = -immucan_id) %>%
  left_join(traits,"immucan_id") %>%
  filter(simple_histology %in% c("Adenocarcinoma")) %>%
  mutate(EGFR = ifelse(EGFR == TRUE, "mutated", "wildtype")) %>%
  group_by(name) %>%
  nest() %>%
  mutate(
    # Wilcoxon test
    test = map(data, ~ wilcox.test(value ~ EGFR, data = .x)),
    summary = map(test, broom::tidy),
    
    # Compute medians for fold change
    medians = map(data, ~ .x %>%
                    group_by(EGFR) %>%
                    summarise(median = median(value, na.rm = TRUE), .groups = "drop") %>%
                    pivot_wider(names_from = EGFR, values_from = median)
                 ),
    
    # Compute fold change: TRUE / FALSE
    #fold_change = map_dbl(medians, ~ .x$Adenocarinoma / .x$`Squamous cell carcinoma`)
    mutated = map_dbl(medians, ~ .x$mutated),
    wildtype = map_dbl(medians, ~ .x$wildtype)
  ) %>%
  select(name, summary, mutated, wildtype) %>%
  unnest(summary)

w_test <- w_test %>%
  ungroup() %>%
  mutate(test = "EGFR",
         sig = p.value < 0.05,
         FDR = p.adjust(p = p.value,method = "BH"),
         FDR = ifelse(FDR < 0.1, "FDR < 0.1",""),
         p.plot = ifelse(p.value < 0.001,"***",ifelse(p.value < 0.01, "**",ifelse(p.value < 0.05, "*","ns")))) 


plot_dat %>%
  filter(!is.na(EGFR)) %>%
  pivot_longer(contains("CN")) %>%
  ggplot(aes(x = name,y = value,fill = EGFR))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(color = EGFR),position = position_jitterdodge(jitter.width = 0.2), size = 0.2)+
  scale_fill_manual(values = alpha(c("red", "blue"),alpha = 0.6))+
  scale_color_manual(values = c("red", "blue"))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = .5, hjust = 1))+
  coord_cartesian(ylim = c(0,0.25))


# densities
cello_dens <- read.csv(paste0(mount_path,"Figures/scores_for_sharing/cellohood_density_cellbased_p1.csv"))

traits <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/all_traits.csv"))
dens <- left_join(cello_dens,traits,"immucan_id")


w_test <- dens %>%
  select(contains(c("immucan_id","CN"))) %>%
  pivot_longer(cols = -immucan_id) %>%
  left_join(traits,"immucan_id") %>%
  filter(simple_histology %in% c("Adenocarcinoma")) %>%
  mutate(EGFR = ifelse(EGFR == TRUE, "mutated", "wildtype")) %>%
  group_by(name) %>%
  nest() %>%
  mutate(
    # Wilcoxon test
    test = map(data, ~ wilcox.test(value ~ EGFR, data = .x)),
    summary = map(test, broom::tidy),
    
    # Compute medians for fold change
    medians = map(data, ~ .x %>%
                    group_by(EGFR) %>%
                    summarise(median = median(value, na.rm = TRUE), .groups = "drop") %>%
                    pivot_wider(names_from = EGFR, values_from = median)
                 ),
    
    # Compute fold change: TRUE / FALSE
    #fold_change = map_dbl(medians, ~ .x$Adenocarinoma / .x$`Squamous cell carcinoma`)
    mutated = map_dbl(medians, ~ .x$mutated),
    wildtype = map_dbl(medians, ~ .x$wildtype)
  ) %>%
  select(name, summary, mutated, wildtype) %>%
  unnest(summary)

w_test <- w_test %>%
  ungroup() %>%
  mutate(test = "EGFR",
         sig = p.value < 0.05,
         FDR = p.adjust(p = p.value,method = "BH"),
         FDR = ifelse(FDR < 0.1, "FDR < 0.1",""),
         p.plot = ifelse(p.value < 0.001,"***",ifelse(p.value < 0.01, "**",ifelse(p.value < 0.05, "*","ns")))) 

w_test %>%
  filter(sig == TRUE)


dens %>%
  filter(!is.na(EGFR),
         simple_histology == "Adenocarcinoma") %>%
  pivot_longer(cols = contains("CN")) %>%
  mutate(value = value + 0.1) %>%
  ggplot(aes(x = name,y = value,fill = EGFR))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(color = EGFR),position = position_jitterdodge(jitter.width = 0.2), size = 0.2)+
  scale_fill_manual(values = alpha(c("red", "blue"),alpha = 0.6))+
  scale_color_manual(values = c("red", "blue"))+
  geom_text(data = w_test, aes(x = name,y=10000,label = p.plot), inherit.aes = FALSE,size = 4)+
  geom_text(data = w_test, aes(x = name,y=14000,label = FDR), inherit.aes = FALSE,size = 2)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = .5, hjust = 1))+
  ylab("Log10 density [mm2]")+
  scale_y_log10()
```

patients with 10 cells per mm2
100 cells mm2 
```{r}
dens %>%
  pivot_longer(cols = contains("CN")) %>%
  rowwise() %>%
  mutate(min10 = value > 10,
         min50 = value > 50,
         min100 = value > 100) %>%
  group_by(name) %>%
  summarise(min_10 = sum(min10)/nrow(dens),
            min_50 = sum(min50)/nrow(dens),
            min_100 = sum(min100)/nrow(dens))
  
hist(log10(dens$CN_13), breaks = 100, main = "CN_13 density across samples")
abline(v = 1, col = "red", lwd = 2)
abline(v = 2, col = "blue", lwd = 2)

hist(log10(dens$CN_17), breaks = 100, main = "CN_17 density across samples")
abline(v = 1, col = "red", lwd = 2)
abline(v = 2, col = "blue", lwd = 2)
```

# densities for histology
```{r}
dens_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/densities_dat.rds"))
w_test <- dens_dat %>%
  pivot_longer(cols = -immucan_id) %>%
  left_join(traits,"immucan_id") %>%
  filter(simple_histology %in% c("Adenocarcinoma","Squamous cell carcinoma")) %>%
  group_by(name) %>%
  nest() %>%
  mutate(
    # Wilcoxon test
    test = map(data, ~ wilcox.test(value ~ simple_histology, data = .x)),
    summary = map(test, broom::tidy),
    
    # Compute medians for fold change
    medians = map(data, ~ .x %>%
                    group_by(simple_histology) %>%
                    summarise(median = median(value, na.rm = TRUE), .groups = "drop") %>%
                    pivot_wider(names_from = simple_histology, values_from = median)
                 ),
    
    # Compute fold change: TRUE / FALSE
    #fold_change = map_dbl(medians, ~ .x$Adenocarinoma / .x$`Squamous cell carcinoma`)
    LUAD_dens = map_dbl(medians, ~ .x$Adenocarcinoma),
    LUSC_dens = map_dbl(medians, ~ .x$`Squamous cell carcinoma`)
  ) %>%
  select(name, summary, LUAD_dens, LUSC_dens) %>%
  unnest(summary)

w_test <- w_test %>%
  mutate(test = "histology",
         sig = p.value < 0.05) 

sig_markers <- w_test %>%
  filter(sig == TRUE) %>%
  #filter(grepl("IF", name)) %>%
  pull(name)
```


```{r, fig.width=15}
histology <- dens_dat %>%
  pivot_longer(cols = -immucan_id) %>%
  left_join(traits,"immucan_id") %>%
  filter(simple_histology %in% c("Adenocarcinoma","Squamous cell carcinoma"),
         name %in% sig_markers) %>%
  ggplot(aes(x = name, y = value,fill = simple_histology))+
  geom_boxplot(outliers = FALSE)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  scale_y_log10()
```


# densities for smoking
```{r}
dens_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/densities_dat.rds"))
w_test <- dens_dat %>%
  pivot_longer(cols = -immucan_id) %>%
  left_join(traits,"immucan_id") %>%
  filter(smoking_status %in% c("Current smoker","Non-smoker")) %>%
  group_by(name) %>%
  nest() %>%
  mutate(
    # Wilcoxon test
    test = map(data, ~ wilcox.test(value ~ smoking_status, data = .x)),
    summary = map(test, broom::tidy),
    
    # Compute medians for fold change
    medians = map(data, ~ .x %>%
                    group_by(smoking_status) %>%
                    summarise(median = median(value, na.rm = TRUE), .groups = "drop") %>%
                    pivot_wider(names_from = smoking_status, values_from = median)
                 ),
    
    # Compute fold change: TRUE / FALSE
    #fold_change = map_dbl(medians, ~ .x$Adenocarinoma / .x$`Squamous cell carcinoma`)
    smoker = map_dbl(medians, ~ .x$`Current smoker`),
    non_smoker = map_dbl(medians, ~ .x$`Non-smoker`)
  ) %>%
  select(name, summary, smoker, non_smoker) %>%
  unnest(summary)

w_test <- w_test %>%
  mutate(test = "histology",
         sig = p.value < 0.05) 

sig_markers <- w_test %>%
  filter(sig == TRUE) %>%
  #filter(grepl("IF", name)) %>%
  pull(name)
```


```{r, fig.width=15}
smoking <- dens_dat %>%
  pivot_longer(cols = -immucan_id) %>%
  left_join(traits,"immucan_id") %>%
  filter(smoking_status %in% c("Current smoker","Non-smoker"),
         name %in% sig_markers) %>%
  ggplot(aes(x = name, y = value,fill = smoking_status))+
  geom_boxplot(outliers = FALSE)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  scale_y_log10()
```

# densities for TPS
```{r}
dens_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/densities_dat.rds"))
w_test <- dens_dat %>%
  pivot_longer(cols = -immucan_id) %>%
  left_join(traits,"immucan_id") %>%
  filter(TPS_score %in% c("positive","negative")) %>%
  group_by(name) %>%
  nest() %>%
  mutate(
    # Wilcoxon test
    test = map(data, ~ wilcox.test(value ~ TPS_score, data = .x)),
    summary = map(test, broom::tidy),
    
    # Compute medians for fold change
    medians = map(data, ~ .x %>%
                    group_by(TPS_score) %>%
                    summarise(median = median(value, na.rm = TRUE), .groups = "drop") %>%
                    pivot_wider(names_from = TPS_score, values_from = median)
                 ),
    
    # Compute fold change: TRUE / FALSE
    #fold_change = map_dbl(medians, ~ .x$Adenocarinoma / .x$`Squamous cell carcinoma`)
    positive = map_dbl(medians, ~ .x$positive),
    negative = map_dbl(medians, ~ .x$negative)
  ) %>%
  select(name, summary, positive, negative) %>%
  unnest(summary)

w_test <- w_test %>%
  mutate(test = "TPS",
         sig = p.value < 0.05) 

sig_markers <- w_test %>%
  filter(sig == TRUE) %>%
  #filter(grepl("IF", name)) %>%
  pull(name)
```


```{r, fig.width=15}
dens_dat %>%
  pivot_longer(cols = -immucan_id) %>%
  left_join(traits,"immucan_id") %>%
  filter(TPS_score %in% c("negative","positive"),
         name %in% sig_markers,
         grepl("IMC2", name)) %>%
  ggplot(aes(x = name, y = value,fill = TPS_score))+
  geom_boxplot(outliers = FALSE)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  scale_y_log10()
```



# densities for TPS
```{r}
dens_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/densities_dat.rds"))
w_test <- dens_dat %>%
  pivot_longer(cols = -immucan_id) %>%
  left_join(traits,"immucan_id") %>%
  filter(TPS_score %in% c("positive","negative")) %>%
  group_by(name) %>%
  nest() %>%
  mutate(
    # Wilcoxon test
    test = map(data, ~ wilcox.test(value ~ TPS_score, data = .x)),
    summary = map(test, broom::tidy),
    
    # Compute medians for fold change
    medians = map(data, ~ .x %>%
                    group_by(TPS_score) %>%
                    summarise(median = median(value, na.rm = TRUE), .groups = "drop") %>%
                    pivot_wider(names_from = TPS_score, values_from = median)
                 ),
    
    # Compute fold change: TRUE / FALSE
    #fold_change = map_dbl(medians, ~ .x$Adenocarinoma / .x$`Squamous cell carcinoma`)
    positive = map_dbl(medians, ~ .x$positive),
    negative = map_dbl(medians, ~ .x$negative)
  ) %>%
  select(name, summary, positive, negative) %>%
  unnest(summary)

w_test <- w_test %>%
  mutate(test = "TPS",
         sig = p.value < 0.05) 

sig_markers <- w_test %>%
  filter(sig == TRUE) %>%
  #filter(grepl("IF", name)) %>%
  pull(name)
```


```{r, fig.width=15}
TPS <- dens_dat %>%
  pivot_longer(cols = -immucan_id) %>%
  left_join(traits,"immucan_id") %>%
  filter(TPS_score %in% c("negative","positive"),
         name %in% sig_markers) %>%
  ggplot(aes(x = name, y = value,fill = TPS_score))+
  geom_boxplot(outliers = FALSE)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  scale_y_log10()
```



# densities for EGFR
```{r}
dens_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/densities_dat.rds"))
w_test <- dens_dat %>%
  pivot_longer(cols = -immucan_id) %>%
  left_join(traits,"immucan_id") %>%
  filter(EGFR %in% c(FALSE,TRUE),
         simple_histology == "Adenocarcinoma") %>%
  group_by(name) %>%
  nest() %>%
  mutate(
    # Wilcoxon test
    test = map(data, ~ wilcox.test(value ~ EGFR, data = .x)),
    summary = map(test, broom::tidy),
    
    # Compute medians for fold change
    medians = map(data, ~ .x %>%
                    group_by(EGFR) %>%
                    summarise(median = median(value, na.rm = TRUE), .groups = "drop") %>%
                    pivot_wider(names_from = EGFR, values_from = median, names_prefix = "EGFR_")
                 ),
    
    # Compute fold change: TRUE / FALSE
    #fold_change = map_dbl(medians, ~ .x$Adenocarinoma / .x$`Squamous cell carcinoma`)
    EGFR_TRUE = map_dbl(medians, ~ .x$EGFR_TRUE),
    EGFR_FALSE = map_dbl(medians, ~ .x$EGFR_FALSE)
  ) %>%
  select(name, summary, EGFR_TRUE, EGFR_FALSE) %>%
  unnest(summary)

w_test <- w_test %>%
  mutate(test = "EGFR",
         sig = p.value < 0.05) 

sig_markers <- w_test %>%
  filter(sig == TRUE) %>%
  #filter(grepl("IF", name)) %>%
  pull(name)
```


```{r, fig.width=15}
EGFR <- dens_dat %>%
  pivot_longer(cols = -immucan_id) %>%
  left_join(traits,"immucan_id") %>%
  filter(EGFR %in% c(FALSE,TRUE),
         simple_histology == "Adenocarcinoma",
         name %in% sig_markers) %>%
  ggplot(aes(x = name, y = value,fill = EGFR))+
  geom_boxplot(outliers = FALSE)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  scale_y_log10()
```


# densities for KRAS
```{r}
dens_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/densities_dat.rds"))
w_test <- dens_dat %>%
  pivot_longer(cols = -immucan_id) %>%
  left_join(traits,"immucan_id") %>%
  filter(KRAS %in% c(FALSE,TRUE),
         simple_histology == "Adenocarcinoma") %>%
  group_by(name) %>%
  nest() %>%
  mutate(
    # Wilcoxon test
    test = map(data, ~ wilcox.test(value ~ KRAS, data = .x)),
    summary = map(test, broom::tidy),
    
    # Compute medians for fold change
    medians = map(data, ~ .x %>%
                    group_by(KRAS) %>%
                    summarise(median = median(value, na.rm = TRUE), .groups = "drop") %>%
                    pivot_wider(names_from = KRAS, values_from = median, names_prefix = "KRAS_")
                 ),
    
    # Compute fold change: TRUE / FALSE
    #fold_change = map_dbl(medians, ~ .x$Adenocarinoma / .x$`Squamous cell carcinoma`)
    KRAS_TRUE = map_dbl(medians, ~ .x$KRAS_TRUE),
    KRAS_FALSE = map_dbl(medians, ~ .x$KRAS_FALSE)
  ) %>%
  select(name, summary, KRAS_TRUE, KRAS_FALSE) %>%
  unnest(summary)

w_test <- w_test %>%
  mutate(test = "KRAS",
         sig = p.value < 0.05) 

sig_markers <- w_test %>%
  filter(sig == TRUE) %>%
  #filter(grepl("IF", name)) %>%
  pull(name)
```


```{r, fig.width=15}
KRAS <- dens_dat %>%
  pivot_longer(cols = -immucan_id) %>%
  left_join(traits,"immucan_id") %>%
  filter(KRAS %in% c(FALSE,TRUE),
         simple_histology == "Adenocarcinoma") %>%
  filter(name %in% sig_markers) %>%
  #filter(grepl("PDL1",name)) %>%
  ggplot(aes(x = name, y = value,fill = KRAS))+
  geom_boxplot(outliers = FALSE)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  scale_y_log10()
```
# densities for TP53_LUAD
```{r}
dens_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/densities_dat.rds"))
w_test <- dens_dat %>%
  pivot_longer(cols = -immucan_id) %>%
  left_join(traits,"immucan_id") %>%
  filter(TP53 %in% c(FALSE,TRUE),
         simple_histology == "Adenocarcinoma") %>%
  group_by(name) %>%
  nest() %>%
  mutate(
    # Wilcoxon test
    test = map(data, ~ wilcox.test(value ~ TP53, data = .x)),
    summary = map(test, broom::tidy),
    
    # Compute medians for fold change
    medians = map(data, ~ .x %>%
                    group_by(TP53) %>%
                    summarise(median = median(value, na.rm = TRUE), .groups = "drop") %>%
                    pivot_wider(names_from = TP53, values_from = median, names_prefix = "TP53_")
                 ),
    
    # Compute fold change: TRUE / FALSE
    #fold_change = map_dbl(medians, ~ .x$Adenocarinoma / .x$`Squamous cell carcinoma`)
    TP53_TRUE = map_dbl(medians, ~ .x$TP53_TRUE),
    TP53_FALSE = map_dbl(medians, ~ .x$TP53_FALSE)
  ) %>%
  select(name, summary, TP53_TRUE, TP53_FALSE) %>%
  unnest(summary)

w_test <- w_test %>%
  mutate(test = "TP53",
         sig = p.value < 0.05) 

sig_markers <- w_test %>%
  filter(sig == TRUE) %>%
  #filter(grepl("IF", name)) %>%
  pull(name)
```


```{r, fig.width=15}
TP53_LUAD <- dens_dat %>%
  pivot_longer(cols = -immucan_id) %>%
  left_join(traits,"immucan_id") %>%
  filter(TP53 %in% c(FALSE,TRUE),
         simple_histology == "Adenocarcinoma") %>%
  filter(name %in% sig_markers) %>%
  #filter(grepl("PDL1",name)) %>%
  ggplot(aes(x = name, y = value,fill = TP53))+
  geom_boxplot(outliers = FALSE)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  scale_y_log10()
```


# densities for TP53_LUSC
```{r}
dens_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/densities_dat.rds"))
w_test <- dens_dat %>%
  pivot_longer(cols = -immucan_id) %>%
  left_join(traits,"immucan_id") %>%
  filter(TP53 %in% c(FALSE,TRUE),
         simple_histology == "Squamous cell carcinoma") %>%
  group_by(name) %>%
  nest() %>%
  mutate(
    # Wilcoxon test
    test = map(data, ~ wilcox.test(value ~ TP53, data = .x)),
    summary = map(test, broom::tidy),
    
    # Compute medians for fold change
    medians = map(data, ~ .x %>%
                    group_by(TP53) %>%
                    summarise(median = median(value, na.rm = TRUE), .groups = "drop") %>%
                    pivot_wider(names_from = TP53, values_from = median, names_prefix = "TP53_")
                 ),
    
    TP53_TRUE = map_dbl(medians, ~ .x$TP53_TRUE),
    TP53_FALSE = map_dbl(medians, ~ .x$TP53_FALSE)
  ) %>%
  select(name, summary, TP53_TRUE, TP53_FALSE) %>%
  unnest(summary)

w_test <- w_test %>%
  mutate(test = "TP53",
         sig = p.value < 0.05) 

sig_markers <- w_test %>%
  filter(sig == TRUE) %>%
  #filter(grepl("IF", name)) %>%
  pull(name)
```


```{r, fig.width=15}
TP53_LUSC <- dens_dat %>%
  pivot_longer(cols = -immucan_id) %>%
  left_join(traits,"immucan_id") %>%
  filter(TP53 %in% c(FALSE,TRUE),
         simple_histology == "Squamous cell carcinoma") %>%
  filter(name %in% sig_markers) %>%
  #filter(grepl("PDL1",name)) %>%
  ggplot(aes(x = name, y = value,fill = TP53))+
  geom_boxplot(outliers = FALSE)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  scale_y_log10()
```

#plots
```{r, fig.height=10, fig.width=20}
histology
smoking
TPS
EGFR
KRAS
TP53_LUAD
TP53_LUSC
```

