---
title: "feature association with survival"
author: "Daniel Schulz"
date: "2026-01-20"
output: html_document
---

In this script we will select all features that are associated with survival in a univariate fashion.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this script we will use the many features coming from the interactions counts of IMC and IF and run them through a LASSO to identify significant features for survival.
NOTE: running LASSO on the continuous data (transformed from caret, log or not-transformed) does not result in any significant association. However, dichotomizing the data reveals some hits which we will investigate further.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r}
library(SingleCellExperiment)
library(tidyverse)
library(imcRtools)
library(distances)
library(fastDummies)
library(survminer)
library(survivalAnalysis)
library(glmnet)
library(ComplexHeatmap)
library(caret)
library(survival)
library(broom)
library(survminer)
library(ggplot2)
library(scater)
```

data
```{r}
mount_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/"

dens_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/densities_dat.rds"))
cello_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/cello_dat.rds"))
IMC1 <- read.csv(paste0(mount_path,"Figures/scores_for_sharing/cell-pair_interactions_celltypes_minor_directional_P1.csv"))
IMC2 <- read.csv(paste0(mount_path,"Figures/scores_for_sharing/cell-pair_interactions_celltypes_minor_directional_P2.csv"))
IF1 <- read.csv(paste0(mount_path,"Figures/scores_for_sharing/cell-cell-pair_interactions_directional_IF1.csv")) 
IF2 <- read.csv(paste0(mount_path,"Figures/scores_for_sharing/cell-cell-pair_interactions_directional_IF2.csv"))
IF3 <- read.csv(paste0(mount_path,"Figures/scores_for_sharing/cell-cell-pair_interactions_directional_IF3.csv"))

# select the 1000 most variable genes - also the ones used for mofa
RNA <- read.csv(paste0(mount_path, "Figures/scores_for_sharing/MOFA/RNA_named.csv"))

traits <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/all_traits.csv"))
```

# merge data

Note, we lost 13 patients do to one of the modalities missing
```{r}
colnames(IMC1) <- c("immucan_id",paste0(colnames(IMC1[,-1]),"_IMC1"))
colnames(IMC2) <- c("immucan_id",paste0(colnames(IMC2[,-1]),"_IMC2"))
colnames(IF1) <- c("immucan_id",paste0(colnames(IF1[,-1]),"_IF1"))
colnames(IF2) <- c("immucan_id",paste0(colnames(IF2[,-1]),"_IF2"))
colnames(IF3) <- c("immucan_id",paste0(colnames(IF3[,-1]),"_IF3"))

dat <- plyr::join_all(list(IMC1, IMC2, IF1, IF2, IF3, dens_dat,cello_dat), by='immucan_id', type='left')
dim(dat)
```

# Filtering

```{r}
# we will remove columns with more than 25% of NAs
filt_dat <- apply(dat[,-1],2,function(x) sum(is.na(x)))
rm_cols <- names(filt_dat[filt_dat > 190/4])

dat <- dat %>%
  select(!all_of(rm_cols))

#we will now remove all columns with more thnan 50% of 0 after setting NAs to 0
dat <- dat %>%
  mutate(across(.cols = -immucan_id, ~ ifelse(is.na(.), 0,.)))
filt_dat <- apply(dat[,-1],2,function(x) sum(x == 0))

rm_cols <- names(filt_dat[filt_dat > 190/2])
dat <- dat %>%
  select(!all_of(rm_cols))


dim(dat)
```

# Univariate COX for continuous features - overall

```{r, fig.height=10}
cur_traits <-traits %>%
  select(immucan_id,TPS, CPS, TMB,TP53,EGFR, KRAS,STK11,CDKN2A,PIK3CA, censoring,time_to_event)
cur_traits <- cur_traits %>%
  mutate(across(.cols = c(TP53,EGFR, KRAS, STK11,CDKN2A,PIK3CA),.fns = ~ ifelse(. == TRUE,1,0)))
res <- left_join(dat,cur_traits,by = "immucan_id")


res <- res %>%
  select(immucan_id,censoring,time_to_event,everything())

p_list <- lapply(res[,-c(1:3)], function(x){
   coxph( Surv(time_to_event, censoring) ~  x ,data = res)
})

RNA <- RNA %>%
  rename(immucan_id = Immucan_id) %>%
  mutate(immucan_id = gsub("_","-",immucan_id))
RNA <- left_join(RNA, cur_traits %>% select(immucan_id,censoring, time_to_event),by = "immucan_id")
RNA <- RNA %>%
  select(immucan_id, censoring,time_to_event, everything())
p_list_RNA <- lapply(RNA[,-c(1:3)], function(x){
   coxph( Surv(time_to_event, censoring) ~ x ,data = RNA)
})
  
df <- data.frame(HR = unlist(lapply(p_list, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p_list, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p_list, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p_list, function(x){summary(x)$conf.int[, "upper .95"]}))) 
df <- df %>%
  mutate(feature = rownames(df))
df_RNA <- data.frame(HR = unlist(lapply(p_list_RNA, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p_list_RNA, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p_list_RNA, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p_list_RNA, function(x){summary(x)$conf.int[, "upper .95"]}))) 

df_RNA <- df_RNA %>%
  mutate(feature = paste0(rownames(df_RNA),"_RNA"))

# investigate only features that pass FDR here
out <- df %>%
  mutate(FDR = p.adjust(p = p.value,method = "BH")) %>%
  filter(FDR < 0.1)
out_RNA <- df_RNA %>%
  mutate(FDR = p.adjust(p = p.value,method = "BH")) %>%
  filter(FDR < 0.1)


res_out <- res %>%
  select(immucan_id,censoring, time_to_event,all_of(out$feature)) %>%
  mutate(across(.cols = all_of(out$feature),.fns = ~ as.factor(ntile(.,n = 2))))

p_list_final <- lapply(res_out[,-c(1:3)], function(x){
   coxph( Surv(time_to_event, censoring) ~  x ,data = res_out)
})

res_out_RNA <- RNA %>%
  select(immucan_id,censoring, time_to_event,all_of(out_RNA$feature)) %>%
  mutate(across(.cols = all_of(out_RNA$feature),.fns = ~ as.factor(ntile(.,n = 2))))

p_list_final <- lapply(res_out[,-c(1:3)], function(x){
   coxph( Surv(time_to_event, censoring) ~  x ,data = res_out)
})
p_list_final_RNA <- lapply(res_out_RNA[,-c(1:3)], function(x){
   coxph( Surv(time_to_event, censoring) ~  x ,data = res_out_RNA)
})
  
df_2 <- data.frame(HR = unlist(lapply(p_list_final, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p_list_final, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p_list_final, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p_list_final, function(x){summary(x)$conf.int[, "upper .95"]}))) 
df_2_RNA <- data.frame(HR = unlist(lapply(p_list_final_RNA, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p_list_final_RNA, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p_list_final_RNA, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p_list_final_RNA, function(x){summary(x)$conf.int[, "upper .95"]}))) 

df_2 <- rbind(df_2, df_2_RNA)

df_2 %>%
  mutate(feature = rownames(df_2),
         sig = as.factor(p.value < 0.05)) %>%
  ggplot(aes(x = HR,y = feature, col = sig))+
  geom_point(size = 2)+
  geom_errorbar(aes(xmin=lower_conf, xmax = upper_conf),lwd =1,width = 0.3)+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust= 1))+
  #scale_y_discrete(labels = function(x) str_split_i(x,"_",2))+
  scale_x_log10()+
  theme_bw()+
  geom_vline(xintercept = 1,linetype = "dashed")+
  geom_text(aes(label = paste0("p : ",round(p.value,digits = 3))),
            col = "black",
            vjust = -1,
            inherit.aes = TRUE)+
  scale_color_manual(values = c("TRUE" = "blue", "FALSE" = "orange"))+
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        legend.position = "none",
        axis.title.y = element_blank(),
        plot.title = element_text(hjust = 0.5))+
  ggtitle("HR for low infiltration ratio")
```

```{r}
feat_dat <- rbind(df, df_RNA)

feat_pval <- feat_dat %>%
  filter(p.value < 0.05) %>%
  mutate(modality = gsub("^.*_", "", feature),
         data_type = ifelse(grepl("\\.",feature),"interaction",""),
         data_type = ifelse(grepl("cello",feature),"CN",data_type),
         data_type = ifelse(grepl("density",feature),"density",data_type),
         data_type = ifelse(grepl("RNA",feature),"RNA",data_type),
         data_type = ifelse(grepl("TP53|EGFR|KRAS|STK11|PIK3CA|CDKN2A|TMB",feature),"WES",data_type),
         data_type = ifelse(grepl("TPS|CPS",feature),"PDL1",data_type))

feat_fdr <- feat_dat %>%
  mutate(FDR = p.adjust(p = p.value,method = "BH")) %>%
  filter(FDR < 0.1) %>%
  mutate(modality = gsub("^.*_", "", feature),
         data_type = ifelse(grepl("\\.",feature),"interaction",""),
         data_type = ifelse(grepl("cello",feature),"CN",data_type),
         data_type = ifelse(grepl("density",feature),"density",data_type),
         data_type = ifelse(grepl("RNA",feature),"RNA",data_type),
         data_type = ifelse(grepl("TP53|EGFR|KRAS|STK11|PIK3CA|CDKN2A|TMB",feature),"WES",data_type),
         data_type = ifelse(grepl("TPS|CPS",feature),"PDL1",data_type))

feat_abundance <- feat_dat %>%
  mutate(modality = gsub("^.*_", "", feature),
         data_type = ifelse(grepl("\\.",feature),"interaction",""),
         data_type = ifelse(grepl("cello",feature),"CN",data_type),
         data_type = ifelse(grepl("density",feature),"density",data_type),
         data_type = ifelse(grepl("RNA",feature),"RNA",data_type),
         data_type = ifelse(grepl("TP53|EGFR|KRAS|STK11|PIK3CA|CDKN2A|TMB",feature),"WES",data_type),
         data_type = ifelse(grepl("TPS|CPS",feature),"PDL1",data_type))

table(feat_pval$data_type)
table(feat_fdr$data_type)
table(feat_abundance$data_type)

plot_dat_pval <- table(feat_pval$data_type)/table(feat_abundance$data_type)[c(1:3,5,6)]
plot_dat_fdr <- table(feat_fdr$data_type)/table(feat_abundance$data_type)[c(3,5)]

df_RNA[(rownames(df_RNA) %in% c("ADGRF5","C1orf116","PHGDH","MYBL2","CENPF","KIF18B","BIRC5","FOXM1","TPX2","TROAP")),]
```

__we cannot find any of the ten genes found with LF1 from MOFA when just searching for gene expression associated with survival. so we need MOFA because it identifies data structure based on all modalities and not just RNA or others.__

# Univariate COX for continuous features - LUAD

```{r, fig.height=10}
cur_traits <-traits %>%
  select(immucan_id,TPS, CPS, TMB,TP53,EGFR, KRAS,STK11,CDKN2A,PIK3CA, censoring,time_to_event)
cur_traits <- cur_traits %>%
  mutate(across(.cols = c(TP53,EGFR, KRAS, STK11,CDKN2A,PIK3CA),.fns = ~ ifelse(. == TRUE,1,0)))
res <- left_join(dat,cur_traits,by = "immucan_id")

LUAD_samples <- traits %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  pull(immucan_id)

res <- res %>%
  select(immucan_id,censoring,time_to_event,everything()) %>%
  filter(immucan_id %in% LUAD_samples)


p_list <- lapply(res[,-c(1:3)], function(x){
   coxph( Surv(time_to_event, censoring) ~  x ,data = res)
})

  
df <- data.frame(HR = unlist(lapply(p_list, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p_list, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p_list, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p_list, function(x){summary(x)$conf.int[, "upper .95"]}))) 
  
out <- df %>%
  mutate(FDR = p.adjust(p = p.value,method = "BH")) %>%
  filter(FDR < 0.1)
out$feature <- rownames(out)

res_out <- res %>% 
  select(immucan_id,censoring, time_to_event,all_of(out$feature)) %>%
  mutate(across(.cols = all_of(out$feature),.fns = ~ as.factor(ntile(.,n = 2))))

p_list_final <- lapply(res_out[,-c(1:3)], function(x){
   coxph( Surv(time_to_event, censoring) ~  x ,data = res_out)
})

  
df_2 <- data.frame(HR = unlist(lapply(p_list_final, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p_list_final, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p_list_final, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p_list_final, function(x){summary(x)$conf.int[, "upper .95"]}))) 


df_2 %>%
  mutate(feature = rownames(df_2),
         sig = as.factor(p.value < 0.05)) %>%
  ggplot(aes(x = HR,y = feature, col = sig))+
  geom_point(size = 2)+
  geom_errorbar(aes(xmin=lower_conf, xmax = upper_conf),lwd =1,width = 0.3)+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust= 1))+
  #scale_y_discrete(labels = function(x) str_split_i(x,"_",2))+
  scale_x_log10()+
  theme_bw()+
  geom_vline(xintercept = 1,linetype = "dashed")+
  geom_text(aes(label = paste0("p : ",round(p.value,digits = 3))),
            col = "black",
            vjust = -1,
            inherit.aes = TRUE)+
  scale_color_manual(values = c("TRUE" = "blue", "FALSE" = "orange"))+
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        legend.position = "none",
        axis.title.y = element_blank(),
        plot.title = element_text(hjust = 0.5))+
  ggtitle("HR for low infiltration ratio")
```

# Univariate COX for continuous features - LUSC

```{r, fig.height=10}
cur_traits <-traits %>%
  select(immucan_id,TPS, CPS, TMB,TP53,EGFR, KRAS,STK11,CDKN2A,PIK3CA, censoring,time_to_event)
cur_traits <- cur_traits %>%
  mutate(across(.cols = c(TP53,EGFR, KRAS, STK11,CDKN2A,PIK3CA),.fns = ~ ifelse(. == TRUE,1,0)))
res <- left_join(dat,cur_traits,by = "immucan_id")

LUSC_samples <- traits %>%
  filter(simple_histology == "Squamous cell carcinoma") %>%
  pull(immucan_id)

res <- res %>%
  select(immucan_id,censoring,time_to_event,everything()) %>%
  filter(immucan_id %in% LUSC_samples)


p_list <- lapply(res[,-c(1:3)], function(x){
   coxph( Surv(time_to_event, censoring) ~  x ,data = res)
})

  
df <- data.frame(HR = unlist(lapply(p_list, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p_list, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p_list, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p_list, function(x){summary(x)$conf.int[, "upper .95"]}))) 
  
out <- df %>%
  mutate(FDR = p.adjust(p = p.value,method = "BH")) %>%
  filter(FDR < 0.1)
out$feature <- rownames(out)

res_out <- res %>% 
  select(immucan_id,censoring, time_to_event,all_of(out$feature)) %>%
  mutate(across(.cols = all_of(out$feature),.fns = ~ as.factor(ntile(.,n = 2))))

p_list_final <- lapply(res_out[,-c(1:3)], function(x){
   coxph( Surv(time_to_event, censoring) ~  x ,data = res_out)
})

  
df_2 <- data.frame(HR = unlist(lapply(p_list_final, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p_list_final, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p_list_final, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p_list_final, function(x){summary(x)$conf.int[, "upper .95"]}))) 


df_2 %>%
  mutate(feature = rownames(df_2),
         sig = as.factor(p.value < 0.05)) %>%
  ggplot(aes(x = HR,y = feature, col = sig))+
  geom_point(size = 2)+
  geom_errorbar(aes(xmin=lower_conf, xmax = upper_conf),lwd =1,width = 0.3)+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust= 1))+
  #scale_y_discrete(labels = function(x) str_split_i(x,"_",2))+
  scale_x_log10()+
  theme_bw()+
  geom_vline(xintercept = 1,linetype = "dashed")+
  geom_text(aes(label = paste0("p : ",round(p.value,digits = 3))),
            col = "black",
            vjust = -1,
            inherit.aes = TRUE)+
  scale_color_manual(values = c("TRUE" = "blue", "FALSE" = "orange"))+
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        legend.position = "none",
        axis.title.y = element_blank(),
        plot.title = element_text(hjust = 0.5))+
  ggtitle("HR for low infiltration ratio")
```


# Univariate COX for dichotomized features

```{r, fig.height=10}
cur_traits <- traits %>%
  select(immucan_id,TPS, CPS, TMB,TP53,EGFR, KRAS,STK11,CDKN2A,PIK3CA, censoring,time_to_event)
cur_traits <- cur_traits %>%
  mutate(across(.cols = c(TP53,EGFR, KRAS, STK11,CDKN2A,PIK3CA),.fns = ~ ifelse(. == TRUE,1,0)))
res <- left_join(dat,cur_traits,by = "immucan_id")

LUAD_samples <- traits %>%
  filter(simple_histology == "Squamous cell carcinoma") %>%
  pull(immucan_id)

res <- res %>%
  select(immucan_id,censoring,time_to_event,everything()) %>%
   mutate(across(.cols = -c(immucan_id,censoring, time_to_event),.fns = ~ as.factor(ntile(.,n = 2)))) %>%
  filter(immucan_id %in% LUAD_samples)



p_list <- lapply(res[,-c(1:3)], function(x){
   coxph( Surv(time_to_event, censoring) ~  x ,data = res)
})

  
df <- data.frame(HR = unlist(lapply(p_list, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p_list, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p_list, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p_list, function(x){summary(x)$conf.int[, "upper .95"]}))) 
  
df %>%
  filter(p.value < 0.05) %>%
  arrange(HR)
out <- df %>%
  mutate(FDR = p.adjust(p = p.value,method = "BH")) %>%
  arrange(FDR)
  filter(FDR < 0.1)
out$feature <- rownames(out)



df_2 %>%
  mutate(feature = rownames(df_2),
         sig = as.factor(p.value < 0.05)) %>%
  ggplot(aes(x = HR,y = feature, col = sig))+
  geom_point(size = 2)+
  geom_errorbar(aes(xmin=lower_conf, xmax = upper_conf),lwd =1,width = 0.3)+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust= 1))+
  #scale_y_discrete(labels = function(x) str_split_i(x,"_",2))+
  scale_x_log10()+
  theme_bw()+
  geom_vline(xintercept = 1,linetype = "dashed")+
  geom_text(aes(label = paste0("p : ",round(p.value,digits = 3))),
            col = "black",
            vjust = -1,
            inherit.aes = TRUE)+
  scale_color_manual(values = c("TRUE" = "blue", "FALSE" = "orange"))+
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        legend.position = "none",
        axis.title.y = element_blank(),
        plot.title = element_text(hjust = 0.5))+
  ggtitle("HR for low infiltration ratio")
```



# LASSO on dichotomozed data

```{r}
clin_dat <- read_csv("O:/projects/immucan/clinical_data/Lung_NSCLC2/20230825/NSCLC2_for_R_update_2024.csv")
df <- as.data.frame(apply(dat, 2, function(x) ntile(x,2)))
rownames(df) <- rownames(dat)


time <- clin_dat %>%
  filter(immucan_id %in% rownames(df)) %>%
  pull(time_to_event)
censoring <- clin_dat %>%
  filter(immucan_id %in% rownames(df)) %>%
  pull(censoring)


y <-  survival::Surv(time, censoring)

# generate folds for the cross validation
nfolds <- 5
set.seed(2135)
foldids <- sample(rep(1:nfolds, length.out = nrow(df)))  # Assign each observation to one of 5 folds


#determine lambda
set.seed(12354)
cv.fit <- cv.glmnet(as.matrix(df),y, family = "cox", alpha = 1, foldid = foldids)
lambda.min <- cv.fit$lambda.min
plot(cv.fit)
set.seed(2351)
fit <- glmnet(as.matrix(df), survival::Surv(time, censoring),lambda = lambda.min, family = "cox", alpha = 1)



# find coefficients at minimum lambda
Coefficients <- coef(fit)

Lasso_coefs <- as.data.frame(as.matrix(Coefficients))
Lasso_coefs <- Lasso_coefs %>%
  mutate(feature = rownames(Lasso_coefs),
         coef = s0,
         s0 = NULL) %>%
  filter(coef != 0,
         !grepl("smoking|simple_",feature))
sig_var_LASSO <- as.vector(Lasso_coefs$feature)

DT::datatable(Lasso_coefs)
```

# Test LASSO features in multivariate model {.tabset}
Note: the features extracted from LASSO are extracted based on a univariate model. Here we will build multivariate models and only keep those features that are significant in a multivariate model.
```{r,results='asis',fig.height=5, fig.width=21}
df_text <- as.data.frame(apply(df, 2, function(x) ifelse(x == 1,"low","high")))
df_text$immucan_id <- rownames(df)

# function to generate the header for the plotting with tabset
catHeader <- function(text = "", level = 3) {
    cat(paste0("\n\n", 
               paste(rep("#", level), collapse = ""), 
               " ", text, "\n"))
}

sig_var_LASSO_in_cox <- vector()

p.list <- list()
hr.list <- list()
factor.list <- list()
for (i in sig_var_LASSO) {

  plot_dat <- left_join(clin_dat,df_text,"immucan_id")

  plot_dat <- plot_dat %>%
    filter(!is.na(!!as.name(i)))

  
  plot_dat[,i] <- factor(pull(plot_dat,var = i),levels = c("low","high"))


  res <- plot_dat %>%
  analyse_multivariate(vars(time_to_event, censoring),
                       c("simple_stage","simple_histology","smoking_status","adjuvant_treatment","neo_adjuvant_treatment",i))



  pval <- res$summaryAsFrame %>%
    filter(grepl(i,factor.id)) %>%
    pull(p)
  if(round(pval,digits = 3) < 0.05 ) {
    sig_var_LASSO_in_cox <- c(sig_var_LASSO_in_cox,paste0(i))
  }
  
  p.list[[i]] <- forest_plot(res)
  hr.list[[i]] <- res$summaryAsFrame$HR
  factor.list[[i]] <- res$summaryAsFrame$factor.name
  p.list[[i]] <- list(p.list[[i]],i)
}

for(i in seq_along(p.list)){
    tmp <- p.list[[i]]
    # As you want to use tabset level here has to be lower than 
    # parent level (ie, parent is 1, so here you have to use 2)
    catHeader(tmp[[2]], 2)
    print(tmp[[1]])
}

```

#{}

# multivariate cox for top features

```{r, fig.height=10}
HRs <- data.frame(HRs = unlist(hr.list),
                  factor.ids = unlist(factor.list)) %>%
  filter(!grepl("smoking|simple|adjuvant",factor.ids))
top_hrs <- HRs %>%
  arrange(HRs) %>%
  pull(factor.ids)

#top_hrs <- top_hrs[c(1:2,(length(top_hrs)-1):length(top_hrs))]

clin_dat <- read_csv("O:/projects/immucan/clinical_data/Lung_NSCLC2/20230825/NSCLC2_for_R_update_2024.csv")

clin_dat <- df_text %>%
  select(immucan_id,top_hrs) %>%
  left_join(clin_dat,by = "immucan_id") %>%
  filter(simple_histology != "unknown")

time_to_event <- clin_dat %>%
  filter(immucan_id %in% df_text$immucan_id) %>%
  pull(time_to_event)
censoring <- clin_dat %>%
  filter(immucan_id %in% df_text$immucan_id) %>%
  pull(censoring)


y <- Surv(time_to_event, censoring)
p <- ggforest(coxph(formula(c(paste0("y ~",  "simple_stage +" , "simple_histology + " , "smoking_status +" , "adjuvant_treatment +" , "neo_adjuvant_treatment +" ), paste0(top_hrs,collapse = "+"))) ,data = as.data.frame(clin_dat)))
p
```

# clustering on selected features
```{r}
top_hrs <- HRs %>%
  arrange(HRs) %>%
  pull(factor.ids)

#top_hrs <- top_hrs[c(1:2,(length(top_hrs)-1):length(top_hrs))]
#top_hrs <- top_hrs[c(1:5,(length(top_hrs)-4):length(top_hrs))]


plot_dat <- dat %>%
  select(all_of(top_hrs))

set.seed(12345)
cur_clust <- kmeans(plot_dat,centers = 3)

plot_dat$kmeans_cluster <- cur_clust$cluster[rownames(plot_dat)]

clin_dat <- read_csv("O:/projects/immucan/clinical_data/Lung_NSCLC2/20230825/NSCLC2_for_R_update_2024.csv")
clin_dat$clusters <- as.factor(cur_clust$cluster[clin_dat$immucan_id])
clin_dat <- clin_dat %>%
  filter(simple_histology != "unknown")

p <- ggforest(coxph(Surv(time_to_event, censoring) ~ simple_stage + simple_histology + smoking_status + adjuvant_treatment + neo_adjuvant_treatment + clusters,data = as.data.frame(clin_dat)))
p
table(clin_dat$censoring,clin_dat$clusters)
```

## cluster heatmap
```{r}
library(pheatmap)
mat <- plot_dat %>%
  pivot_longer(cols = top_hrs) %>%
  group_by(name, kmeans_cluster, .drop = FALSE) %>%
  summarise(cluster_mean = mean(value)) %>%
  pivot_wider(id_cols = name,values_from = cluster_mean,names_from = kmeans_cluster)
mat_names <- mat$name
mat$name <- NULL
mat <- as.matrix(mat)
rownames(mat) <- mat_names

pheatmap(mat,color = viridis::viridis(100),cluster_rows = TRUE,clustering_method = "ward.D2")

pheatmap(mat,scale = "row",color = colorRampPalette(c("dark blue", "white", "dark red"))(100),cluster_rows = TRUE,clustering_method = "ward.D2")

most_different <- c("mregDC.CD4_IMC2","CD8_GB.Tumor_IF2","HLADR.Tumor_IMC1","B.Tumor_IF1","CD8.Tumor_IMC1")
```


## Kaplan Meier

```{r, fig.height=6}
clin_dat <- read_csv("O:/projects/immucan/clinical_data/Lung_NSCLC2/20230825/NSCLC2_for_R_update_2024.csv")

clin_dat$cluster <- cur_clust$cluster[clin_dat$immucan_id]

surv_fit <- survfit(Surv(time_to_event, censoring) ~ cluster, data = clin_dat)

ggsurv <- ggsurvplot(surv_fit, data = clin_dat, risk.table = TRUE, pval = TRUE)
ggsurv$plot <- ggsurv$plot + 
theme(legend.position = "none")
ggsurv
```
