---
title: "feature association with survival"
author: "Daniel Schulz"
date: "2026-01-20"
output: html_document
---

In this script we will select all features that are associated with survival in a univariate fashion.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this script we will use the many features coming from the interactions counts of IMC and IF and run them through a LASSO to identify significant features for survival.
NOTE: running LASSO on the continuous data (transformed from caret, log or not-transformed) does not result in any significant association. However, dichotomizing the data reveals some hits which we will investigate further.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r}
library(SingleCellExperiment)
library(tidyverse)
library(imcRtools)
library(distances)
library(fastDummies)
library(survminer)
library(survivalAnalysis)
library(glmnet)
library(ComplexHeatmap)
library(caret)
library(survival)
library(broom)
library(survminer)
library(ggplot2)
library(scater)

mount_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/"
```

data
```{r}
dens_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/densities_dat.rds"))
cello_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/cello_dat.rds"))
int_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/interaction_dat.rds"))
traits <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/all_traits.csv"))

traits <- traits %>%
  mutate(STK11_no_KRAS = STK11 & !KRAS, 
         KEAP1_no_KRAS = KEAP1 & !KRAS,
         KRAS_only = KRAS & !STK11 & !KEAP1,
         KRAS_co   = KRAS & (STK11 | KEAP1),
         KRAS_new = ifelse(KRAS_only, "KRAS_only", ifelse(KRAS_co, "KRAS_co", "Other")))

modules <- as.data.frame(readRDS( "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/patient_imaging_meta_cluster_scores_tpm.rds"))

modules <- modules  %>%
  mutate(immucan_id = rownames(modules))
rownames(modules) <- NULL

rna <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/rna_scores_tpm.rds")
rna <- as.data.frame(rna)
rna$immucan_id <- rownames(rna)
rownames(rna) <- NULL
colnames(rna) <- gsub(" ", "_", colnames(rna))
colnames(rna) <- gsub("-", "_", colnames(rna))

# select the 1000 most variable genes - also the ones used for mofa
RNA <- read.csv(paste0(mount_path, "Figures/scores_for_sharing/MOFA/RNA_named.csv"))
RNA <- RNA %>%
  mutate(Immucan_id = gsub("_","-",Immucan_id)) %>%
  rename(immucan_id = Immucan_id)
```

```{r}
library(org.Hs.eg.db)
library(AnnotationDbi)
library(msigdbr)
library(singscore)

rna_mat <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/rna_mat_tpm.rds")


# ── gene signatures ────────────────────────────────────────────────────
hallmark_sets <- msigdbr(species = "Homo sapiens", collection = "H") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  split(.$gs_name) %>%
  lapply(function(x) x$gene_symbol)

# Targeted Hallmark gene sets of interest
target_sets <- c(
  "HALLMARK_INFLAMMATORY_RESPONSE",
  "HALLMARK_INTERFERON_GAMMA_RESPONSE",
  "HALLMARK_INTERFERON_ALPHA_RESPONSE",
  "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
  "HALLMARK_IL6_JAK_STAT3_SIGNALING",
  "HALLMARK_IL2_STAT5_SIGNALING"
)
gs_list_hallmark <- hallmark_sets[intersect(target_sets, names(hallmark_sets))]

# ── 2. Custom gene sets ────────────────────────────────────────────────────────
# CXCL9/10/11 chemokine axis (T cell recruitment)
gs_list_custom <- list(
  CXCL9_10_11_axis = c("CXCL9", "CXCL10", "CXCL11", "CXCR3",
                       "CXCL13", "CXCR5"),
  T_cell_activation = c("CD8A", "CD8B", "GZMB", "PRF1", "IFNG",
                        "TNF", "IL2", "PDCD1", "HAVCR2", "LAG3"),
  type_I_IFN = c("IFNB1", "IRF7", "MX1", "OAS1", "OAS2", "OAS3",
                 "ISG15", "IFIT1", "IFIT2", "IFIT3", "RSAD2", "HERC5"),
  NF_kB_target  = c("NFKB1", "RELA", "RELB", "NFKB2", "REL",
                    "BCL2L1", "IL1B", "IL6", "CXCL8", "TNF", "ICAM1"),
  TCR_signaling = c("CD3D", "CD3E", "CD3G", "ZAP70", "LCK", "LAT",
                    "ITK", "PLCG1", "NFATC1", "NFATC2"),
  cytotoxic = c("GZMB", "GZMA", "GZMK", "PRF1", "NKG7", "IFNG", "KLRG1"),
  exhaustion = c("PDCD1","LAG3","HAVCR2","TX","CTLA4", "IDO1")
  )

all_gs <- c(gs_list_hallmark, gs_list_custom)

# compute score
rna_for_score <- t(rna_mat)

# Filter gene sets to genes present in the RNA-seq matrix
all_gs <- lapply(all_gs, function(g) intersect(g, rownames(rna_for_score)))
all_gs <- all_gs[sapply(all_gs, length) >= 4]



overlap <- sapply(all_gs, function(g) sum(g %in% rownames(rna_for_score)))
total   <- sapply(all_gs, length)
print(data.frame(found = overlap, total = total, pct = round(100 * overlap / total)))

# ── Score with singscore ──────────────────────────────────────────────────────
ranked <- rankGenes(rna_for_score)

scores_list <- lapply(names(all_gs), function(nm) {
  sc <- simpleScore(ranked, upSet = all_gs[[nm]], centerScore = TRUE)
  setNames(sc$TotalScore, rownames(sc))
})


inf_scores <- as.data.frame(do.call(cbind, scores_list))
colnames(inf_scores) <- names(all_gs)
rownames(inf_scores) <- colnames(rna_for_score)
inf_scores$immucan_id <- rownames(inf_scores)
rownames(inf_scores) <- NULL
```

# merge data

Note, we lost 13 patients do to one of the modalities missing
```{r}
dat <- plyr::join_all(list(dens_dat, int_dat,cello_dat, RNA, rna, inf_scores, modules), by='immucan_id', type='left')
dim(dat)
sum(colnames(dat) %in% colnames(RNA))
```



# investigate completeness
```{r}
raw <- dat  # pre-imputation, pre-filter

feat_qc <- tibble(name = setdiff(colnames(raw), "immucan_id")) %>%
  rowwise() %>%
  mutate(
    x        = list(raw[[name]]),
    n_obs    = sum(!is.na(x)),
    frac_na  = 1 - n_obs / nrow(raw),
    n_unique = dplyr::n_distinct(na.omit(x))
  ) %>%
  ungroup() %>% dplyr::select(-x) %>%
  mutate(
    drop_na        = frac_na > 0.5,
    drop_degenerate = n_unique < 2,      # constant where observed; breaks coxph
    keep = !(drop_na | drop_degenerate)
  )

feat_qc %>%
  filter(frac_na > 0.5)

feat_qc %>% count(drop_na, drop_degenerate)

dat_filt <- raw %>% dplyr::select(immucan_id, all_of(feat_qc$name[feat_qc$keep]))
dim(dat_filt)
```



# Univariate COX for continuous features - overall

we will use:
- imaging features
- binary scores
- 1000 most variable genes as used for MOFA
- img modules
- rna modules
- infl signatures

calculate HR for continuous values
then binarize the features that survived FDR correction
plot the HR for those features
```{r, fig.height=10}
cur_traits <-traits %>%
  dplyr::select(immucan_id,TPS, CPS, TMB,TP53,EGFR, KRAS, KEAP1, KRAS_only, KRAS_co,STK11,CDKN2A,PIK3CA, censoring,time_to_event)
cur_traits <- cur_traits %>%
  mutate(across(.cols = c(TP53,EGFR, KRAS, KRAS_co, KRAS_only, KEAP1, STK11,CDKN2A,PIK3CA),.fns = ~ ifelse(. == TRUE,1,0)))
cur_traits <- cur_traits %>%
  rename(gCDKN2A = CDKN2A)
res <- left_join(dat_filt,cur_traits,by = "immucan_id")

res_long<- res %>%
  pivot_longer(cols = -c(immucan_id, censoring, time_to_event)) %>%
  mutate(type = case_when(grepl("density", name) ~ "numeric",
                          name %in% colnames(int_dat) ~ "numeric",
                          grepl("cello", name) ~ "numeric",
                          name %in% colnames(rna) ~ "enrichment_score",
                          name %in% colnames(modules) ~ "z_score",
                          name %in% colnames(inf_scores) ~ "enrichment_score",
                          name %in% c("TP53","EGFR", "KRAS", "KRAS_co", "KRAS_only", "KEAP1",
                                      "STK11","gCDKN2A","PIK3CA") ~ "binary",
                          name %in% c("TPS", "CPS", "TMB") ~ "numeric",
                          name %in% colnames(RNA) ~ "numeric"))


stopifnot(!any(is.na(res_long$type)))
dim(res)
```

## all data LUAD 
```{r}
EVENT_VALUE <- 1
pats <- traits %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  pull(immucan_id)

luad <- res_long%>%
  filter(immucan_id %in% pats)
# one row per patient
surv_base <- luad %>%
  dplyr::distinct(immucan_id, time_to_event, censoring) %>%
  dplyr::mutate(status = as.integer(censoring == EVENT_VALUE)) %>%
  dplyr::filter(!is.na(time_to_event), time_to_event > 0, !is.na(status))

stopifnot(nrow(surv_base) == dplyr::n_distinct(surv_base$immucan_id))

# name -> type lookup, carried over from the long format
feat_type <- luad %>% dplyr::distinct(name, type)
stopifnot(nrow(feat_type) == dplyr::n_distinct(feat_type$name))  # no name in 2 blocks

# values as a patient x feature matrix, aligned to surv_base
wide <- luad %>%
  dplyr::select(immucan_id, name, value) %>%
  tidyr::pivot_wider(names_from = name, values_from = value)

stopifnot(!any(vapply(wide, is.list, logical(1))))  # list-cols = duplicated IDs

keep_ids  <- intersect(surv_base$immucan_id, wide$immucan_id)
surv_base <- surv_base %>% dplyr::filter(immucan_id %in% keep_ids) %>% dplyr::arrange(immucan_id)
wide      <- wide      %>% dplyr::filter(immucan_id %in% keep_ids) %>% dplyr::arrange(immucan_id)
stopifnot(identical(surv_base$immucan_id, wide$immucan_id))

feat_names <- setdiff(colnames(wide), "immucan_id")
c(patients = nrow(wide), features = length(feat_names), events = sum(surv_base$status))
```

### survival modeling
```{r}
library(survival)

MIN_EVENTS  <- 10          # events among patients with the feature observed
MIN_PREV    <- 10          # binary: smaller group size
covar_terms <- character(0)  # e.g. c("age","sex","stage","strata(centre)")

inv_norm <- function(x) {
  n <- sum(!is.na(x)); if (n < 3) return(rep(NA_real_, length(x)))
  qnorm((rank(x, na.last = "keep", ties.method = "average") - 0.5) / n)
}
tie_fraction <- function(x) { x <- x[!is.na(x)]; if (!length(x)) NA_real_ else max(table(x))/length(x) }

fit_one <- function(d, xcol, lab) {
  na_row <- tibble::tibble(scale = lab, hr = NA_real_, lo = NA_real_, hi = NA_real_,
                           coef = NA_real_, se = NA_real_, p_wald = NA_real_,
                           lr_chisq = NA_real_, p_lr = NA_real_,
                           cindex = NA_real_, p_ph = NA_real_)
  if (length(unique(stats::na.omit(d[[xcol]]))) < 2) return(na_row)

  f <- stats::as.formula(paste("Surv(time_to_event, status) ~",
                               paste(c(xcol, covar_terms), collapse = " + ")))
  fit <- try(coxph(f, data = d, ties = "efron"), silent = TRUE)
  if (inherits(fit, "try-error") || !all(is.finite(stats::coef(fit)))) return(na_row)

  s  <- summary(fit)
  cf <- s$coefficients[xcol, ]; ci <- s$conf.int[xcol, ]

  if (length(covar_terms)) {
    f0 <- stats::as.formula(paste("Surv(time_to_event, status) ~",
                                  paste(covar_terms, collapse = " + ")))
    fit0 <- try(coxph(f0, data = d, ties = "efron"), silent = TRUE)
    lr <- if (inherits(fit0, "try-error")) NA_real_ else 2*(fit$loglik[2] - fit0$loglik[2])
  } else {
    lr <- 2*(fit$loglik[2] - fit$loglik[1])
  }

  zph  <- try(cox.zph(fit), silent = TRUE)
  p_ph <- if (inherits(zph, "try-error") || !xcol %in% rownames(zph$table)) NA_real_
          else unname(zph$table[xcol, "p"])

  tibble::tibble(scale = lab,
                 hr = unname(ci["exp(coef)"]), lo = unname(ci["lower .95"]),
                 hi = unname(ci["upper .95"]), coef = unname(cf["coef"]),
                 se = unname(cf["se(coef)"]), p_wald = unname(cf["Pr(>|z|)"]),
                 lr_chisq = lr,
                 p_lr = if (is.na(lr)) NA_real_ else stats::pchisq(lr, 1, lower.tail = FALSE),
                 cindex = unname(concordance(fit)$concordance),
                 p_ph = p_ph)
}

fit_feature <- function(nm) {
  ty <- feat_type$type[match(nm, feat_type$name)]
  d  <- surv_base
  d$value <- wide[[nm]]
  d  <- d[!is.na(d$value), , drop = FALSE]

  n_obs <- nrow(d); n_ev <- sum(d$status)
  if (n_ev < MIN_EVENTS) return(NULL)

  if (ty == "binary") {
    tb <- table(d$value)
    if (length(tb) < 2 || min(tb) < MIN_PREV) return(NULL)
    d$x_rank <- d$value; d$x_sd <- d$value
    prev <- mean(d$value == 1); tf <- NA_real_; unit <- "present vs absent"
  } else {
    if (length(unique(d$value)) < 3) return(NULL)
    d$x_rank <- inv_norm(d$value)
    sdv <- stats::sd(d$value)
    d$x_sd <- if (sdv > 0) (d$value - mean(d$value))/sdv else NA_real_
    prev <- NA_real_; tf <- tie_fraction(d$value); unit <- "per 1 SD normal scores"
  }

  dplyr::bind_rows(fit_one(d, "x_rank", "rank_INT"),
                   fit_one(d, "x_sd",   "per_SD")) %>%
    dplyr::mutate(name = nm, type = ty, n = n_obs, n_events = n_ev,
                  prevalence = prev, tie_frac = tf, scale_unit = unit, .before = 1)
}

cox_res <- purrr::map_dfr(feat_names, fit_feature)

cox_res <- cox_res %>%
  dplyr::group_by(scale, type) %>%
  dplyr::mutate(fdr_within_type = stats::p.adjust(p_lr, method = "BH")) %>%
  dplyr::group_by(scale) %>%
  dplyr::mutate(fdr_global = stats::p.adjust(p_lr, method = "BH")) %>%
  dplyr::ungroup()

primary <- cox_res %>%
  dplyr::filter(scale == "rank_INT") %>%
  dplyr::arrange(dplyr::desc(abs(log(hr))))

cox_res %>% dplyr::filter(scale == "rank_INT") %>%
  dplyr::group_by(type) %>%
  dplyr::summarise(tested = dplyr::n(),
                   n_med = stats::median(n), ev_med = stats::median(n_events),
                   fdr = sum(fdr_within_type < 0.1, na.rm = TRUE),
                   raw05 = sum(p_lr < 0.05, na.rm = TRUE), .groups = "drop")

cox_res %>%
  arrange(p_lr) %>%
  filter(name == "Neutrophil_density__IF1")
cox_res %>%
  arrange(p_lr) %>%
  filter(name == "EGFR")
val_dat <- dat_filt %>% left_join(traits, "immucan_id")
coxph(Surv(time_to_event, censoring) ~ TP53, data = val_dat %>% filter(simple_histology == "Adenocarcinoma"))
coxph(Surv(time_to_event, censoring) ~ Cold_fibrotic, data = val_dat %>% filter(simple_histology == "Adenocarcinoma"))
cox_res %>%
  filter(name == "Cold_fibrotic")
colnames(cox_res)
```


## all data
```{r}
# one row per patient
surv_base <- res_long %>%
  dplyr::distinct(immucan_id, time_to_event, censoring) %>%
  dplyr::mutate(status = as.integer(censoring == EVENT_VALUE)) %>%
  dplyr::filter(!is.na(time_to_event), time_to_event > 0, !is.na(status))

stopifnot(nrow(surv_base) == dplyr::n_distinct(surv_base$immucan_id))

# name -> type lookup, carried over from the long format
feat_type <- res_long%>% dplyr::distinct(name, type)
stopifnot(nrow(feat_type) == dplyr::n_distinct(feat_type$name))  # no name in 2 blocks

# values as a patient x feature matrix, aligned to surv_base
wide <- res_long%>%
  dplyr::select(immucan_id, name, value) %>%
  tidyr::pivot_wider(names_from = name, values_from = value)

stopifnot(!any(vapply(wide, is.list, logical(1))))  # list-cols = duplicated IDs

keep_ids  <- intersect(surv_base$immucan_id, wide$immucan_id)
surv_base <- surv_base %>% dplyr::filter(immucan_id %in% keep_ids) %>% dplyr::arrange(immucan_id)
wide      <- wide      %>% dplyr::filter(immucan_id %in% keep_ids) %>% dplyr::arrange(immucan_id)
stopifnot(identical(surv_base$immucan_id, wide$immucan_id))

feat_names <- setdiff(colnames(wide), "immucan_id")
c(patients = nrow(wide), features = length(feat_names), events = sum(surv_base$status))
```

## survival modeling
```{r}
library(survival)

MIN_EVENTS  <- 10          # events among patients with the feature observed
MIN_PREV    <- 10          # binary: smaller group size
covar_terms <- character(0)  # e.g. c("age","sex","stage","strata(centre)")

inv_norm <- function(x) {
  n <- sum(!is.na(x)); if (n < 3) return(rep(NA_real_, length(x)))
  qnorm((rank(x, na.last = "keep", ties.method = "average") - 0.5) / n)
}
tie_fraction <- function(x) { x <- x[!is.na(x)]; if (!length(x)) NA_real_ else max(table(x))/length(x) }

fit_one <- function(d, xcol, lab) {
  na_row <- tibble::tibble(scale = lab, hr = NA_real_, lo = NA_real_, hi = NA_real_,
                           coef = NA_real_, se = NA_real_, p_wald = NA_real_,
                           lr_chisq = NA_real_, p_lr = NA_real_,
                           cindex = NA_real_, p_ph = NA_real_)
  if (length(unique(stats::na.omit(d[[xcol]]))) < 2) return(na_row)

  f <- stats::as.formula(paste("Surv(time_to_event, status) ~",
                               paste(c(xcol, covar_terms), collapse = " + ")))
  fit <- try(coxph(f, data = d, ties = "efron"), silent = TRUE)
  if (inherits(fit, "try-error") || !all(is.finite(stats::coef(fit)))) return(na_row)

  s  <- summary(fit)
  cf <- s$coefficients[xcol, ]; ci <- s$conf.int[xcol, ]

  if (length(covar_terms)) {
    f0 <- stats::as.formula(paste("Surv(time_to_event, status) ~",
                                  paste(covar_terms, collapse = " + ")))
    fit0 <- try(coxph(f0, data = d, ties = "efron"), silent = TRUE)
    lr <- if (inherits(fit0, "try-error")) NA_real_ else 2*(fit$loglik[2] - fit0$loglik[2])
  } else {
    lr <- 2*(fit$loglik[2] - fit$loglik[1])
  }

  zph  <- try(cox.zph(fit), silent = TRUE)
  p_ph <- if (inherits(zph, "try-error") || !xcol %in% rownames(zph$table)) NA_real_
          else unname(zph$table[xcol, "p"])

  tibble::tibble(scale = lab,
                 hr = unname(ci["exp(coef)"]), lo = unname(ci["lower .95"]),
                 hi = unname(ci["upper .95"]), coef = unname(cf["coef"]),
                 se = unname(cf["se(coef)"]), p_wald = unname(cf["Pr(>|z|)"]),
                 lr_chisq = lr,
                 p_lr = if (is.na(lr)) NA_real_ else stats::pchisq(lr, 1, lower.tail = FALSE),
                 cindex = unname(concordance(fit)$concordance),
                 p_ph = p_ph)
}

fit_feature <- function(nm) {
  ty <- feat_type$type[match(nm, feat_type$name)]
  d  <- surv_base
  d$value <- wide[[nm]]
  d  <- d[!is.na(d$value), , drop = FALSE]

  n_obs <- nrow(d); n_ev <- sum(d$status)
  if (n_ev < MIN_EVENTS) return(NULL)

  if (ty == "binary") {
    tb <- table(d$value)
    if (length(tb) < 2 || min(tb) < MIN_PREV) return(NULL)
    d$x_rank <- d$value; d$x_sd <- d$value
    prev <- mean(d$value == 1); tf <- NA_real_; unit <- "present vs absent"
  } else {
    if (length(unique(d$value)) < 3) return(NULL)
    d$x_rank <- inv_norm(d$value)
    sdv <- stats::sd(d$value)
    d$x_sd <- if (sdv > 0) (d$value - mean(d$value))/sdv else NA_real_
    prev <- NA_real_; tf <- tie_fraction(d$value); unit <- "per 1 SD normal scores"
  }

  dplyr::bind_rows(fit_one(d, "x_rank", "rank_INT"),
                   fit_one(d, "x_sd",   "per_SD")) %>%
    dplyr::mutate(name = nm, type = ty, n = n_obs, n_events = n_ev,
                  prevalence = prev, tie_frac = tf, scale_unit = unit, .before = 1)
}

cox_res <- purrr::map_dfr(feat_names, fit_feature)

cox_res <- cox_res %>%
  dplyr::group_by(scale, type) %>%
  dplyr::mutate(fdr_within_type = stats::p.adjust(p_lr, method = "BH")) %>%
  dplyr::group_by(scale) %>%
  dplyr::mutate(fdr_global = stats::p.adjust(p_lr, method = "BH")) %>%
  dplyr::ungroup()

primary <- cox_res %>%
  dplyr::filter(scale == "rank_INT") %>%
  dplyr::arrange(dplyr::desc(abs(log(hr))))

cox_res %>% dplyr::filter(scale == "rank_INT") %>%
  dplyr::group_by(type) %>%
  dplyr::summarise(tested = dplyr::n(),
                   n_med = stats::median(n), ev_med = stats::median(n_events),
                   fdr = sum(fdr_within_type < 0.1, na.rm = TRUE),
                   raw05 = sum(p_lr < 0.05, na.rm = TRUE), .groups = "drop")

cox_res %>%
  filter(name %in% colnames(modules))
```


# screen
```{r}
# ============================================================================
# QC / diagnostic plots for the univariate Cox screen
#
# Expects in the environment:
#   cox_res     - screen output (long over `scale`)
#   surv_base   - immucan_id, time_to_event, censoring, status  (1 row/patient)
#   wide        - immucan_id + one column per feature           (1 row/patient)
#   feat_type   - name, type
#   feat_names  - character vector of feature columns
#   fit_feature(), fit_one(), inv_norm(), tie_fraction()  from the screen chunk
#
# Message of this figure set: the screen is power-limited, not signal-free,
# and that is the argument for MOFA / module-level feature reduction.
# ============================================================================

suppressPackageStartupMessages({
  library(dplyr); library(tidyr); library(ggplot2); library(purrr); library(survival)
})

traits <- traits %>%
  mutate(simple_histology = ifelse(simple_histology == "Adenocarcinoma" ,"LUAD", simple_histology),
         simple_histology = ifelse(simple_histology == "Squamous cell carcinoma" ,"LUSC", simple_histology))

SUBTYPE_COL <- "simple_histology"   # column in `traits` holding LUAD / LUSC
N_PERM      <- 20            # permutation null for the subtype overlap (slow)
TOP_N       <- 20

theme_set(theme_bw(base_size = 10))
blk_cols <- c(numeric = "#4C72B0", enrichment_score = "#DD8452",
              z_score = "#55A868", binary = "#C44E52")

prim <- cox_res %>% filter(scale == "rank_INT")

# ===========================================================================
# 1. IS THERE ANY SIGNAL AT ALL?  p-value histogram per block
#    Flat = indistinguishable from null. This is the headline QC panel.
# ===========================================================================

p_hist <- ggplot(prim, aes(p_lr)) +
  geom_histogram(aes(fill = type), breaks = seq(0, 1, 0.05),
                 colour = "white", linewidth = 0.2, show.legend = FALSE) +
  geom_hline(data = prim %>% count(type) %>% mutate(exp = n * 0.05),
             aes(yintercept = exp), linetype = 2, colour = "grey30") +
  facet_wrap(~ type, scales = "free_y") +
  scale_fill_manual(values = blk_cols) +
  labs(x = "Likelihood-ratio p-value", y = "features",
       title = "P-value distribution per feature block",
       subtitle = "Dashed line = count expected under the global null")
p_hist
# ---- observed vs expected nominal hits ------------------------------------
enrich_tbl <- prim %>%
  group_by(type) %>%
  summarise(tested = n(),
            obs05  = sum(p_lr < 0.05, na.rm = TRUE),
            exp05  = 0.05 * tested,
            ratio  = obs05 / exp05,
            # one-sided binomial test: more nominal hits than chance?
            p_bin  = binom.test(obs05, tested, 0.05, alternative = "greater")$p.value,
            .groups = "drop")

p_enrich <- enrich_tbl %>%
  pivot_longer(c(obs05, exp05), names_to = "which", values_to = "count") %>%
  ggplot(aes(type, count, fill = which)) +
  geom_col(position = "dodge") +
  scale_fill_manual(values = c(obs05 = "#4C72B0", exp05 = "grey70"),
                    labels = c(obs05 = "observed", exp05 = "expected"), name = NULL) +
  labs(x = NULL, y = "features with p < 0.05",
       title = "Nominal hits vs chance expectation, by block")
p_enrich
# ===========================================================================
# 2. QQ PLOT -- deviation from uniform, with an inflation summary
# ===========================================================================

qq_dat <- prim %>%
  filter(!is.na(p_lr)) %>%
  group_by(type) %>%
  arrange(p_lr, .by_group = TRUE) %>%
  mutate(expected = -log10(ppoints(n())),
         observed = -log10(p_lr)) %>%
  ungroup()

lambda_tbl <- prim %>%
  filter(!is.na(lr_chisq)) %>%
  group_by(type) %>%
  summarise(lambda = median(lr_chisq, na.rm = TRUE) / qchisq(0.5, 1), .groups = "drop")

p_qq <- ggplot(qq_dat, aes(expected, observed, colour = type)) +
  geom_abline(slope = 1, intercept = 0, colour = "grey50", linetype = 2) +
  geom_point(size = 0.6, alpha = 0.6) +
  geom_text(data = lambda_tbl, inherit.aes = FALSE,
            aes(x = 0, y = Inf, label = sprintf("lambda = %.2f", lambda)),
            hjust = 0, vjust = 1.5, size = 3) +
  facet_wrap(~ type) +
  scale_colour_manual(values = blk_cols, guide = "none") +
  labs(x = expression(Expected~-log[10](p)), y = expression(Observed~-log[10](p)),
       title = "QQ plot against the uniform null",
       subtitle = "lambda > 1 indicates more signal (or confounding) than chance")
p_qq
# ===========================================================================
# 3. POWER -- what this design could have detected
#    This is the panel that turns "we found nothing" into a quantitative claim.
# ===========================================================================

n_ev    <- sum(surv_base$status)
n_tests <- nrow(prim)

detectable_hr <- function(d, alpha, power = 0.8) {
  exp((qnorm(1 - alpha / 2) + qnorm(power)) / sqrt(d))
}

pow_grid <- expand_grid(
  events = seq(10, 300, by = 5),
  thresh = c("nominal 0.05", sprintf("Bonferroni (%d tests)", n_tests))
) %>%
  mutate(alpha = ifelse(thresh == "nominal 0.05", 0.05, 0.05 / n_tests),
         hr    = detectable_hr(events, alpha))

p_power <- ggplot(pow_grid, aes(events, hr, colour = thresh)) +
  geom_line(linewidth = 0.8) +
  geom_vline(xintercept = n_ev, linetype = 2) +
  annotate("text", x = n_ev, y = Inf, label = sprintf("  this study: %d events", n_ev),
           hjust = 0, vjust = 1.5, size = 3) +
  scale_y_log10(breaks = c(1.2, 1.5, 2, 3, 5, 10)) +
  scale_colour_manual(values = c("#4C72B0", "#C44E52"), name = NULL) +
  labs(x = "number of events", y = "smallest detectable HR per SD (80% power)",
       title = "Detection limit of the screen",
       subtitle = "Effects below these curves cannot be recovered at this event count")
p_power
# observed effect sizes against the same limit
lim_nom  <- detectable_hr(n_ev, 0.05)
lim_bonf <- detectable_hr(n_ev, 0.05 / n_tests)

p_effect_dist <- ggplot(prim, aes(hr, fill = type)) +
  geom_histogram(bins = 60, colour = "white", linewidth = 0.15) +
  geom_vline(xintercept = c(1/lim_bonf, lim_bonf), colour = "#C44E52", linetype = 2) +
  geom_vline(xintercept = c(1/lim_nom, lim_nom), colour = "#4C72B0", linetype = 3) +
  scale_x_log10() +
  scale_fill_manual(values = blk_cols, name = NULL) +
  labs(x = "HR per SD (rank scale, log axis)", y = "features",
       title = "Observed effect sizes vs the detection limit",
       subtitle = "Dashed: Bonferroni limit. Dotted: nominal limit. Most of the distribution sits inside.")
p_effect_dist
# ===========================================================================
# 4. TECHNICAL DIAGNOSTICS -- are effect estimates being driven by artefacts?
# ===========================================================================

# 4a. SE is governed by events, not patients
p_se <- ggplot(prim, aes(n_events, se, colour = type)) +
  geom_point(size = 0.7, alpha = 0.6) +
  stat_function(fun = function(d) 1/sqrt(d), colour = "black", linetype = 2) +
  scale_colour_manual(values = blk_cols, name = NULL) +
  labs(x = "events among patients with the feature observed", y = "SE(log HR)",
       title = "Precision is set by event count",
       subtitle = "Dashed: theoretical 1/sqrt(events) for a standardised covariate")
p_se
# 4b. tie fraction -- features where the normal-score transform has a point mass
p_ties <- prim %>%
  filter(!is.na(tie_frac)) %>%
  ggplot(aes(tie_frac, abs(log(hr)), colour = type)) +
  geom_vline(xintercept = 0.3, linetype = 2, colour = "grey40") +
  geom_point(size = 0.7, alpha = 0.6) +
  scale_colour_manual(values = blk_cols, name = NULL) +
  labs(x = "fraction of observations in the largest tied block", y = "|log HR|",
       title = "Zero-inflation and effect size",
       subtitle = "Right of the line the rank transform is a point mass, not a continuum")
p_ties
# 4c. rank vs SD scale -- disagreement flags outlier-driven estimates
scale_cmp <- cox_res %>%
  filter(type != "binary") %>%
  dplyr::select(name, type, scale, coef) %>%
  pivot_wider(names_from = scale, values_from = coef) %>%
  filter(!is.na(rank_INT), !is.na(per_SD)) %>%
  mutate(discordant = sign(rank_INT) != sign(per_SD) |
           abs(rank_INT - per_SD) > 0.5 * pmax(abs(rank_INT), abs(per_SD)))

p_scale <- ggplot(scale_cmp, aes(per_SD, rank_INT, colour = discordant)) +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey50") +
  geom_hline(yintercept = 0, colour = "grey85") +
  geom_vline(xintercept = 0, colour = "grey85") +
  geom_point(size = 0.7, alpha = 0.6) +
  scale_colour_manual(values = c(`FALSE` = "grey55", `TRUE` = "#C44E52"),
                      labels = c("consistent", "discordant"), name = NULL) +
  labs(x = "log HR, per SD on native scale", y = "log HR, rank-based normal scores",
       title = "Effect estimates across transformations",
       subtitle = sprintf("%d/%d features discordant -- outlier-driven on the native scale",
                          sum(scale_cmp$discordant), nrow(scale_cmp)))
p_scale
# 4d. PH assumption
p_ph_hist <- prim %>%
  filter(!is.na(p_ph)) %>%
  ggplot(aes(p_ph)) +
  geom_histogram(breaks = seq(0, 1, 0.05), fill = "#8172B3", colour = "white",
                 linewidth = 0.2) +
  labs(x = "cox.zph p-value", y = "features",
       title = "Proportional hazards check",
       subtitle = "Underpowered at this event count; a flat distribution is weak reassurance")
p_ph_hist
# ===========================================================================
# 5. TOP-N FORESTS
# ===========================================================================

forest_of <- function(df, ttl, n = TOP_N) {
  d <- df %>%
    filter(!is.na(hr), is.finite(lo), is.finite(hi)) %>%
    slice_max(abs(log(hr)), n = n) %>%
    mutate(name = factor(name, levels = name[order(log(hr))]))
  ggplot(d, aes(hr, name, colour = type)) +
    geom_vline(xintercept = 1, linetype = 2, colour = "grey50") +
    geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0, linewidth = 0.4) +
    geom_point(size = 1.8) +
    geom_text(aes(x = hi, label = sprintf("  n=%d/%d ev", n, n_events)),
              hjust = 0, size = 2.4, colour = "grey35") +
    scale_x_log10() +
    scale_colour_manual(values = blk_cols, name = NULL) +
    coord_cartesian(clip = "off") +
    labs(x = "HR per SD (rank scale)", y = NULL, title = ttl,
         subtitle = "Ranked by |log HR|. No feature survives FDR; shown as hypothesis-generating only.") +
    theme(plot.margin = margin(5, 60, 5, 5))
}

p_top_all <- forest_of(prim, sprintf("Top %d features, full cohort", TOP_N))

p_top_all
```

```{r}
# ===========================================================================
# 6. SUBTYPE SPLIT -- with the null it has to be read against
# ===========================================================================

subtypes <- traits %>%
  dplyr::select(immucan_id, all_of(SUBTYPE_COL)) %>%
  rename(subtype = all_of(SUBTYPE_COL)) %>%
  filter(subtype %in% c("LUAD", "LUSC"))

sb_full <- surv_base   # keep a copy; run_screen rebinds the globals

#' Re-run the rank-scale screen on a subset of patients.
run_screen <- function(ids, features = feat_names) {
  keep <- surv_base$immucan_id %in% ids
  sb   <- surv_base[keep, , drop = FALSE]
  wd   <- wide[keep, , drop = FALSE]
  if (sum(sb$status) < 10) return(tibble())

  map_dfr(features, function(nm) {
    d <- sb; d$value <- wd[[nm]]
    d <- d[!is.na(d$value), , drop = FALSE]
    if (sum(d$status) < 8 || length(unique(d$value)) < 3) return(NULL)
    d$x <- if (length(unique(d$value)) == 2) d$value else inv_norm(d$value)
    fit <- try(coxph(Surv(time_to_event, status) ~ x, data = d, ties = "efron"),
               silent = TRUE)
    if (inherits(fit, "try-error") || !all(is.finite(coef(fit)))) return(NULL)
    s <- summary(fit)
    tibble(name = nm, coef = unname(s$coefficients["x", "coef"]),
           hr = unname(s$conf.int["x", "exp(coef)"]),
           lo = unname(s$conf.int["x", "lower .95"]),
           hi = unname(s$conf.int["x", "upper .95"]),
           p  = unname(s$coefficients["x", "Pr(>|z|)"]),
           n = nrow(d), n_events = sum(d$status))
  })
}

id_luad <- subtypes$immucan_id[subtypes$subtype == "LUAD"]
id_lusc <- subtypes$immucan_id[subtypes$subtype == "LUSC"]

message(sprintf("LUAD: %d patients / %d events | LUSC: %d patients / %d events",
                sum(surv_base$immucan_id %in% id_luad),
                sum(surv_base$status[surv_base$immucan_id %in% id_luad]),
                sum(surv_base$immucan_id %in% id_lusc),
                sum(surv_base$status[surv_base$immucan_id %in% id_lusc])))

res_luad <- run_screen(id_luad) %>% mutate(type = feat_type$type[match(name, feat_type$name)])
res_lusc <- run_screen(id_lusc) %>% mutate(type = feat_type$type[match(name, feat_type$name)])

p_top_luad <- forest_of(res_luad, sprintf("Top %d features, LUAD", TOP_N))
p_top_lusc <- forest_of(res_lusc, sprintf("Top %d features, LUSC", TOP_N))

# ---- 6a. rank-rank concordance --------------------------------------------
rr <- inner_join(select(res_luad, name, type, coef_luad = coef),
                 select(res_lusc, name, coef_lusc = coef), by = "name")

rho <- cor(rr$coef_luad, rr$coef_lusc, method = "spearman", use = "complete.obs")

p_rankrank <- ggplot(rr, aes(coef_luad, coef_lusc, colour = type)) +
  geom_hline(yintercept = 0, colour = "grey85") +
  geom_vline(xintercept = 0, colour = "grey85") +
  geom_abline(slope = 1, intercept = 0, linetype = 2, colour = "grey50") +
  geom_point(size = 0.6, alpha = 0.5) +
  scale_colour_manual(values = blk_cols, name = NULL) +
  labs(x = "log HR, LUAD", y = "log HR, LUSC",
       title = "Per-feature effects, LUAD vs LUSC",
       subtitle = sprintf("Spearman rho = %.3f. Scatter of this width is expected at these event counts.", rho))

# ---- 6b. formal interaction test -- the actual question --------------------
sub_map <- setNames(subtypes$subtype, subtypes$immucan_id)

interaction_test <- function(nm) {
  d <- surv_base
  d$value   <- wide[[nm]]
  d$subtype <- factor(sub_map[d$immucan_id], levels = c("LUAD", "LUSC"))
  d <- d[!is.na(d$value) & !is.na(d$subtype), , drop = FALSE]
  if (sum(d$status) < 15 || length(unique(d$value)) < 3) return(NULL)
  if (min(table(d$subtype)) < 15) return(NULL)
  d$x <- if (length(unique(d$value)) == 2) d$value else inv_norm(d$value)

  f_main <- try(coxph(Surv(time_to_event, status) ~ x + subtype, d, ties = "efron"), silent = TRUE)
  f_int  <- try(coxph(Surv(time_to_event, status) ~ x * subtype, d, ties = "efron"), silent = TRUE)
  if (inherits(f_main, "try-error") || inherits(f_int, "try-error")) return(NULL)
  if (!all(is.finite(coef(f_int)))) return(NULL)

  lr <- 2 * (f_int$loglik[2] - f_main$loglik[2])
  tibble(name = nm, lr_int = lr,
         p_int = pchisq(lr, df = 1, lower.tail = FALSE))
}

res_int <- map_dfr(feat_names, interaction_test) %>%
  mutate(type = feat_type$type[match(name, feat_type$name)],
         fdr_int = p.adjust(p_int, method = "BH"))

message(sprintf("\nFeature x subtype interactions: %d tested, %d nominal p<0.05 (%.0f expected), %d pass FDR",
                nrow(res_int), sum(res_int$p_int < 0.05, na.rm = TRUE),
                0.05 * nrow(res_int), sum(res_int$fdr_int < 0.05, na.rm = TRUE)))

p_int_hist <- ggplot(res_int, aes(p_int)) +
  geom_histogram(breaks = seq(0, 1, 0.05), fill = "#937860", colour = "white",
                 linewidth = 0.2) +
  geom_hline(yintercept = nrow(res_int) * 0.05, linetype = 2) +
  labs(x = "interaction p-value (feature x subtype)", y = "features",
       title = "Formal test of subtype-specific effects",
       subtitle = "Dashed = chance expectation. This is the test that supports or refutes a LUAD/LUSC difference.")

# ---- 6c. permutation null for top-list overlap -----------------------------
# How much overlap would two subgroups of these sizes share if subtype were
# irrelevant? If the real split matches the null, the non-overlap is a power
# artefact and cannot be cited as evidence of biological difference.

top_overlap <- function(a, b, n = TOP_N) {
  ta <- a %>% slice_max(abs(coef), n = n) %>% pull(name)
  tb <- b %>% slice_max(abs(coef), n = n) %>% pull(name)
  length(intersect(ta, tb))
}

obs_overlap <- top_overlap(res_luad, res_lusc)

perm_feats <- if (length(feat_names) > 400) sample(feat_names, 400) else feat_names
message(sprintf("\nPermutation null on %d features x %d replicates -- this is slow.",
                length(perm_feats), N_PERM))

obs_overlap_sub <- top_overlap(filter(res_luad, name %in% perm_feats),
                               filter(res_lusc, name %in% perm_feats))

n_a <- length(intersect(id_luad, surv_base$immucan_id))
perm_overlap <- map_dbl(seq_len(N_PERM), function(i) {
  ids <- sample(surv_base$immucan_id)
  a <- run_screen(ids[seq_len(n_a)],              perm_feats)
  b <- run_screen(ids[(n_a + 1):length(ids)],     perm_feats)
  if (!nrow(a) || !nrow(b)) return(NA_real_)
  top_overlap(a, b)
})

p_perm <- tibble(overlap = perm_overlap) %>%
  ggplot(aes(overlap)) +
  geom_histogram(binwidth = 1, fill = "grey70", colour = "white", linewidth = 0.2) +
  geom_vline(xintercept = obs_overlap_sub, colour = "#C44E52", linewidth = 0.9) +
  labs(x = sprintf("features shared between the two top-%d lists", TOP_N), y = "random splits",
       title = "Top-list overlap: real LUAD/LUSC split vs random splits",
       subtitle = sprintf("Red = observed (%d). Random splits of the same sizes: median %.0f. Overlapping distributions mean the split carries no information.",
                          obs_overlap_sub, median(perm_overlap, na.rm = TRUE)))

# ===========================================================================
# 7. ASSEMBLE
# ===========================================================================

qc_plots <- list(
  pval_hist     = p_hist,
  enrichment    = p_enrich,
  qq            = p_qq,
  power         = p_power,
  effect_dist   = p_effect_dist,
  se_vs_events  = p_se,
  ties          = p_ties,
  scale_cmp     = p_scale,
  ph            = p_ph_hist,
  top_all       = p_top_all,
  top_luad      = p_top_luad,
  top_lusc      = p_top_lusc,
  rankrank      = p_rankrank,
  interaction   = p_int_hist,
  perm_overlap  = p_perm
)

qc_tables <- list(
  enrichment  = enrich_tbl,
  lambda      = lambda_tbl,
  discordant  = filter(scale_cmp, discordant),
  luad        = res_luad,
  lusc        = res_lusc,
  interaction = res_int,
  limits      = tibble(events = n_ev, tests = n_tests,
                       hr_nominal = lim_nom, hr_bonferroni = lim_bonf)
)

print(enrich_tbl)
print(qc_tables$limits)
# invisible(lapply(qc_plots, print))
```

# older code

```{r}
select <- dplyr::select
filter <- dplyr::filter   # frequently masked by the same packages
rename <- dplyr::rename
# ============================================================================
# Univariate Cox screening across heterogeneous feature blocks
#
# Primary readout : rank-based inverse-normal transform (INT)  -> comparable HRs
# Sensitivity     : per-SD and per-IQR on the native (optionally logged) scale
# Diagnostics     : PH test, nonlinearity test, bootstrap rank stability,
#                   correlation clustering of hits
#
# Input: `res` in long format with columns
#        immucan_id, time_to_event, censoring, name, value, type
#        type in {numeric, enrichment_score, z_score, binary}
# ============================================================================

suppressPackageStartupMessages({
  library(survival)
  library(dplyr)
  library(tidyr)
  library(purrr)
  library(tibble)
  library(ggplot2)
  library(splines)
})

# ---------------------------------------------------------------------------
# 0. CONFIG -- set these deliberately
# ---------------------------------------------------------------------------

EVENT_VALUE  <- 1        # value of `censoring` that codes an EVENT. VERIFY THIS.
LOG_NUMERIC  <- TRUE     # log1p densities/intensities before the SD/IQR scales
                         # (no effect on the rank scale, which is monotone-invariant)
MIN_EVENTS   <- 10       # skip features with fewer usable events
MIN_PREV     <- 10       # for binary features: min patients in the smaller group
N_BOOT       <- 500      # bootstrap replicates for stability (0 to skip)
FDR_WITHIN_TYPE <- TRUE  # also compute BH within feature block

# Adjustment set. Leave NULL/empty for unadjusted screening.
# covar_df must contain immucan_id + the covariate columns.
covar_df    <- NULL
covar_terms <- character(0)   # e.g. c("age", "sex", "stage", "strata(centre)")

set.seed(1)

# ---------------------------------------------------------------------------
# 1. PRE-FLIGHT CHECKS
# ---------------------------------------------------------------------------

message("Unmatched features (type == NA):")
print(res %>% filter(is.na(type)) %>% distinct(name) %>% pull(name))

message("\ncensoring coding:")
print(table(res$censoring, useNA = "ifany"))

message("\nfeatures per type:")
print(res %>% filter(!is.na(type)) %>% distinct(name, type) %>% count(type))

# ---------------------------------------------------------------------------
# 2. TRANSFORMS
# ---------------------------------------------------------------------------

#' Blom / rankit inverse-normal transform. Ties averaged.
inv_norm <- function(x) {
  n <- sum(!is.na(x))
  if (n < 3) return(rep(NA_real_, length(x)))
  r <- rank(x, na.last = "keep", ties.method = "average")
  qnorm((r - 0.5) / n)
}

#' Fraction of observations sitting in the largest tied block.
tie_fraction <- function(x) {
  x <- x[!is.na(x)]
  if (!length(x)) return(NA_real_)
  max(table(x)) / length(x)
}

safe_sd_scale <- function(x) {
  s <- stats::sd(x, na.rm = TRUE)
  if (!is.finite(s) || s == 0) return(rep(NA_real_, length(x)))
  (x - mean(x, na.rm = TRUE)) / s
}

safe_iqr_scale <- function(x) {
  s <- stats::IQR(x, na.rm = TRUE)
  if (!is.finite(s) || s == 0) return(rep(NA_real_, length(x)))
  (x - stats::median(x, na.rm = TRUE)) / s
}

# ---------------------------------------------------------------------------
# 3. SURVIVAL BACKBONE
# ---------------------------------------------------------------------------

surv_base <- res %>%
  distinct(immucan_id, time_to_event, censoring) %>%
  mutate(status = as.integer(censoring == EVENT_VALUE)) %>%
  filter(!is.na(time_to_event), time_to_event > 0, !is.na(status))

stopifnot(nrow(surv_base) == n_distinct(surv_base$immucan_id))

message(sprintf("\n%d patients, %d events (%.1f%%)",
                nrow(surv_base), sum(surv_base$status),
                100 * mean(surv_base$status)))

if (!is.null(covar_df)) {
  surv_base <- left_join(surv_base, covar_df, by = "immucan_id")
}

feat_long <- res %>%
  filter(!is.na(type), !is.na(value)) %>%
  dplyr::select(immucan_id, name, type, value)

feat_split <- split(feat_long, feat_long$name)

# ---------------------------------------------------------------------------
# 4. CORE FITTER
# ---------------------------------------------------------------------------

#' Fit one Cox model on one scale and return a one-row tibble of statistics.
fit_one_scale <- function(d, xcol, scale_label) {

  na_row <- tibble(scale = scale_label, hr = NA_real_, lo = NA_real_, hi = NA_real_,
                   coef = NA_real_, se = NA_real_, z = NA_real_, p_wald = NA_real_,
                   lr_chisq = NA_real_, p_lr = NA_real_,
                   cindex = NA_real_, cindex_delta = NA_real_,
                   p_ph = NA_real_, p_nonlin = NA_real_)

  if (all(is.na(d[[xcol]])) || length(unique(na.omit(d[[xcol]]))) < 2) return(na_row)

  rhs_full <- paste(c(xcol, covar_terms), collapse = " + ")
  f_full   <- stats::as.formula(paste("Surv(time_to_event, status) ~", rhs_full))

  fit <- try(coxph(f_full, data = d, ties = "efron", x = TRUE), silent = TRUE)
  if (inherits(fit, "try-error") || any(!is.finite(coef(fit)))) return(na_row)

  s  <- summary(fit)
  cf <- s$coefficients[xcol, , drop = TRUE]
  ci <- s$conf.int[xcol, , drop = TRUE]

  # ---- likelihood-ratio test for the feature term only ----
  if (length(covar_terms)) {
    f_null   <- stats::as.formula(
      paste("Surv(time_to_event, status) ~", paste(covar_terms, collapse = " + ")))
    fit_null <- try(coxph(f_null, data = d, ties = "efron"), silent = TRUE)
    if (inherits(fit_null, "try-error")) {
      lr <- NA_real_; c_null <- NA_real_
    } else {
      lr     <- 2 * (fit$loglik[2] - fit_null$loglik[2])
      c_null <- unname(concordance(fit_null)$concordance)
    }
  } else {
    lr     <- 2 * (fit$loglik[2] - fit$loglik[1])
    c_null <- 0.5
  }
  p_lr <- if (is.na(lr)) NA_real_ else stats::pchisq(lr, df = 1, lower.tail = FALSE)

  c_full <- unname(concordance(fit)$concordance)

  # ---- proportional hazards ----
  zph <- try(cox.zph(fit), silent = TRUE)
  p_ph <- if (inherits(zph, "try-error")) NA_real_ else {
    tb <- zph$table
    if (xcol %in% rownames(tb)) unname(tb[xcol, "p"]) else NA_real_
  }

  # ---- nonlinearity: linear vs natural spline (3 df) ----
  p_nl <- NA_real_
  if (length(unique(na.omit(d[[xcol]]))) > 10) {
    f_sp <- stats::as.formula(
      paste("Surv(time_to_event, status) ~",
            paste(c(sprintf("ns(%s, df = 3)", xcol), covar_terms), collapse = " + ")))
    fit_sp <- try(coxph(f_sp, data = d, ties = "efron"), silent = TRUE)
    if (!inherits(fit_sp, "try-error")) {
      dev <- 2 * (fit_sp$loglik[2] - fit$loglik[2])
      if (is.finite(dev) && dev >= 0) p_nl <- stats::pchisq(dev, df = 2, lower.tail = FALSE)
    }
  }

  tibble(
    scale        = scale_label,
    hr           = unname(ci["exp(coef)"]),
    lo           = unname(ci["lower .95"]),
    hi           = unname(ci["upper .95"]),
    coef         = unname(cf["coef"]),
    se           = unname(cf["se(coef)"]),
    z            = unname(cf["z"]),
    p_wald       = unname(cf["Pr(>|z|)"]),
    lr_chisq     = lr,
    p_lr         = p_lr,
    cindex       = c_full,
    cindex_delta = c_full - c_null,
    p_ph         = p_ph,
    p_nonlin     = p_nl
  )
}

#' Assemble the per-feature model frame and fit all three scales.
fit_feature <- function(df) {

  nm <- df$name[1]
  ty <- df$type[1]

  d <- surv_base %>% inner_join(select(df, immucan_id, value), by = "immucan_id")
  if (nrow(d) < 3) return(NULL)

  n_obs    <- nrow(d)
  n_events <- sum(d$status)
  if (n_events < MIN_EVENTS) return(NULL)

  if (ty == "binary") {
    tb <- table(d$value)
    if (length(tb) < 2 || min(tb) < MIN_PREV) return(NULL)
    prev      <- mean(d$value == 1, na.rm = TRUE)
    ev_pos    <- sum(d$status[d$value == 1])
    ev_neg    <- sum(d$status[d$value == 0])
    tie_frac  <- NA_real_
    d$x_rank  <- d$value
    d$x_sd    <- d$value
    d$x_iqr   <- d$value
    unit      <- "present vs absent"
  } else {
    if (length(unique(d$value)) < 3) return(NULL)
    prev <- NA_real_; ev_pos <- NA_integer_; ev_neg <- NA_integer_
    tie_frac <- tie_fraction(d$value)
    v_nat <- if (LOG_NUMERIC && ty == "numeric" && min(d$value, na.rm = TRUE) >= 0) {
      log1p(d$value)
    } else d$value
    d$x_rank <- inv_norm(d$value)     # monotone: logging is irrelevant here
    d$x_sd   <- safe_sd_scale(v_nat)
    d$x_iqr  <- safe_iqr_scale(v_nat)
    unit     <- "per 1 SD of normal scores"
  }

  out <- bind_rows(
    fit_one_scale(d, "x_rank", "rank_INT"),
    fit_one_scale(d, "x_sd",   "per_SD"),
    fit_one_scale(d, "x_iqr",  "per_IQR")
  )

  out %>%
    mutate(name = nm, type = ty, n = n_obs, n_events = n_events,
           prevalence = prev, events_pos = ev_pos, events_neg = ev_neg,
           tie_frac = tie_frac, scale_unit = unit, .before = 1)
}

message("\nFitting ", length(feat_split), " features ...")
cox_res <- map_dfr(feat_split, fit_feature)

# ---------------------------------------------------------------------------
# 5. MULTIPLICITY
# ---------------------------------------------------------------------------

cox_res <- cox_res %>%
  group_by(scale) %>%
  mutate(fdr_global = p.adjust(p_lr, method = "BH")) %>%
  group_by(scale, type) %>%
  mutate(fdr_within_type = p.adjust(p_lr, method = "BH")) %>%
  ungroup()

# Primary table: rank scale, sorted by effect size not by p
primary <- cox_res %>%
  filter(scale == "rank_INT") %>%
  arrange(desc(abs(log(hr)))) %>%
  select(name, type, n, n_events, hr, lo, hi, lr_chisq, p_lr,
         fdr_global, fdr_within_type, cindex, cindex_delta,
         p_ph, p_nonlin, tie_frac, prevalence, scale_unit)

hits <- primary %>%
  filter(if (FDR_WITHIN_TYPE) fdr_within_type < 0.05 else fdr_global < 0.05)

message(sprintf("\n%d features pass FDR < 0.05 on the rank scale", nrow(hits)))
print(head(hits, 30), width = Inf)

# ---- agreement between scales: does the ranking survive the transform? ----
scale_agreement <- cox_res %>%
  filter(type != "binary") %>%
  select(name, scale, coef) %>%
  pivot_wider(names_from = scale, values_from = coef)

message("\nSpearman correlation of log-HR across scales:")
print(round(cor(scale_agreement[, -1], use = "pairwise.complete.obs",
                method = "spearman"), 3))

# Features where rank and SD scales disagree in sign or magnitude
# -> almost always outlier-driven on the native scale.
discordant <- scale_agreement %>%
  filter(!is.na(rank_INT), !is.na(per_SD),
         sign(rank_INT) != sign(per_SD) |
           abs(rank_INT - per_SD) > 0.5 * pmax(abs(rank_INT), abs(per_SD))) %>%
  arrange(desc(abs(per_SD)))

message(sprintf("\n%d features discordant between rank and SD scales (inspect these)",
                nrow(discordant)))

# ---------------------------------------------------------------------------
# 6. FLAGS ON THE HITS
# ---------------------------------------------------------------------------

flagged <- hits %>%
  mutate(
    flag_ph      = !is.na(p_ph) & p_ph < 0.05,          # time-varying effect
    flag_nonlin  = !is.na(p_nonlin) & p_nonlin < 0.05,  # not log-linear
    flag_ties    = !is.na(tie_frac) & tie_frac > 0.3,   # point mass, INT degraded
    flag_disc    = name %in% discordant$name
  )

message("\nFlag counts among hits:")
print(flagged %>% summarise(across(starts_with("flag_"), sum)))

# ---------------------------------------------------------------------------
# 7. BOOTSTRAP RANK STABILITY (on the rank scale, whole pipeline resampled)
# ---------------------------------------------------------------------------

boot_stability <- function(feature_names, B = N_BOOT) {

  if (B == 0 || !length(feature_names)) return(tibble())

  wide <- feat_long %>%
    filter(name %in% feature_names) %>%
    select(immucan_id, name, value) %>%
    pivot_wider(names_from = name, values_from = value)

  d0  <- surv_base %>% inner_join(wide, by = "immucan_id")
  ids <- seq_len(nrow(d0))

  reps <- map_dfr(seq_len(B), function(b) {
    idx <- sample(ids, replace = TRUE)
    db  <- d0[idx, , drop = FALSE]

    map_dfr(feature_names, function(fn) {
      x <- db[[fn]]
      if (length(unique(na.omit(x))) < 3) {
        return(tibble(boot = b, name = fn, coef = NA_real_, p = NA_real_))
      }
      db$xb <- if (length(unique(na.omit(x))) == 2) x else inv_norm(x)
      f <- stats::as.formula(
        paste("Surv(time_to_event, status) ~", paste(c("xb", covar_terms), collapse = " + ")))
      fit <- try(coxph(f, data = db, ties = "efron"), silent = TRUE)
      if (inherits(fit, "try-error")) {
        return(tibble(boot = b, name = fn, coef = NA_real_, p = NA_real_))
      }
      sc <- summary(fit)$coefficients
      tibble(boot = b, name = fn,
             coef = unname(sc["xb", "coef"]),
             p    = unname(sc["xb", "Pr(>|z|)"]))
    })
  })

  reps %>%
    group_by(boot) %>%
    mutate(rk = rank(-abs(coef), na.last = "keep")) %>%
    group_by(name) %>%
    summarise(
      sel_freq   = mean(p < 0.05, na.rm = TRUE),
      sign_cons  = max(mean(coef > 0, na.rm = TRUE), mean(coef < 0, na.rm = TRUE)),
      rank_med   = stats::median(rk, na.rm = TRUE),
      rank_q1    = stats::quantile(rk, 0.25, na.rm = TRUE),
      rank_q3    = stats::quantile(rk, 0.75, na.rm = TRUE),
      .groups    = "drop"
    ) %>%
    arrange(rank_med)
}

stab <- boot_stability(hits$name)
if (nrow(stab)) {
  message("\nBootstrap stability (rank scale):")
  print(head(stab, 30), width = Inf)
}

# ---------------------------------------------------------------------------
# 8. REDUNDANCY: are the hits independent signals?
# ---------------------------------------------------------------------------

hit_mat <- feat_long %>%
  filter(name %in% hits$name) %>%
  select(immucan_id, name, value) %>%
  pivot_wider(names_from = name, values_from = value) %>%
  column_to_rownames("immucan_id") %>%
  as.matrix()

hit_mat_int <- apply(hit_mat, 2, function(x) if (length(unique(na.omit(x))) > 2) inv_norm(x) else x)

if (ncol(hit_mat_int) > 2) {
  cmat <- cor(hit_mat_int, use = "pairwise.complete.obs", method = "spearman")
  hc   <- hclust(as.dist(1 - abs(cmat)), method = "average")
  # cut at |rho| = 0.7 -> features within a cluster are near-interchangeable
  clus <- cutree(hc, h = 1 - 0.7)

  cluster_tbl <- tibble(name = names(clus), cluster = clus) %>%
    left_join(select(hits, name, hr, lr_chisq, fdr_within_type), by = "name") %>%
    group_by(cluster) %>%
    arrange(desc(lr_chisq), .by_group = TRUE) %>%
    mutate(representative = row_number() == 1) %>%
    ungroup()

  message(sprintf("\n%d hits collapse into %d correlation clusters (|rho| > 0.7)",
                  nrow(cluster_tbl), n_distinct(cluster_tbl$cluster)))
  print(cluster_tbl %>% filter(representative), width = Inf)
} else {
  cluster_tbl <- tibble()
}

# ---------------------------------------------------------------------------
# 9. FOREST PLOT -- continuous and binary in separate panels (different units)
# ---------------------------------------------------------------------------

plot_dat <- hits %>%
  mutate(panel = ifelse(type == "binary", "Binary (present vs absent)",
                        "Continuous (per 1 SD of normal scores)"),
         name  = factor(name, levels = name[order(hr)])) %>%
  left_join(select(flagged, name, flag_ph, flag_nonlin), by = "name")

p_forest <- ggplot(plot_dat, aes(x = hr, y = name, colour = type)) +
  geom_vline(xintercept = 1, linetype = 2, colour = "grey50") +
  geom_errorbarh(aes(xmin = lo, xmax = hi), height = 0) +
  geom_point(aes(shape = flag_ph), size = 2) +
  scale_x_log10() +
  scale_shape_manual(values = c(`FALSE` = 16, `TRUE` = 1),
                     name = "PH violated") +
  facet_grid(panel ~ ., scales = "free_y", space = "free_y") +
  labs(x = "Hazard ratio (log scale)", y = NULL,
       title = "Univariate Cox screen, rank-based inverse-normal scale",
       subtitle = "Open circles: cox.zph p < 0.05 (effect not constant over time)") +
  theme_bw(base_size = 10) +
  theme(panel.grid.minor = element_blank())

print(p_forest)

# ---------------------------------------------------------------------------
# 10. OUTPUTS
# ---------------------------------------------------------------------------

results <- list(
  all        = cox_res,
  primary    = primary,
  hits       = flagged,
  discordant = discordant,
  stability  = stab,
  clusters   = cluster_tbl,
  forest     = p_forest
)

# write.csv(primary, "cox_univariate_rank_scale.csv", row.names = FALSE)
```





























###### old stuff
```
res <- res %>%
  dplyr::select(immucan_id,censoring,time_to_event,everything())

p_list <- lapply(res[,-c(1:3)], function(x){
   coxph( Surv(time_to_event, censoring) ~  x ,data = res)
})

df <- data.frame(HR = unlist(lapply(p_list, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p_list, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p_list, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p_list, function(x){summary(x)$conf.int[, "upper .95"]}))) 
df <- df %>%
  mutate(feature = rownames(df))

cur_RNA <- left_join(RNA, cur_traits %>% dplyr::select(immucan_id,censoring, time_to_event),by = "immucan_id")
cur_RNA <- cur_RNA %>%
  dplyr::select(immucan_id, censoring,time_to_event, everything())
p_list_RNA <- lapply(cur_RNA[,-c(1:3)], function(x){
   coxph( Surv(time_to_event, censoring) ~ x ,data = cur_RNA)
})

df_RNA <- data.frame(HR = unlist(lapply(p_list_RNA, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p_list_RNA, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p_list_RNA, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p_list_RNA, function(x){summary(x)$conf.int[, "upper .95"]}))) 

df_RNA <- df_RNA %>%
  mutate(feature = paste0(rownames(df_RNA),"_RNA"))

# investigate only features that pass FDR here
out <- df %>%
  mutate(FDR = p.adjust(p = p.value,method = "BH")) %>%
  filter(FDR < 0.1)
  filter(feature == "TAN")
out_RNA <- df_RNA %>%
  mutate(FDR = p.adjust(p = p.value,method = "BH")) %>%
  filter(FDR < 0.1)
# no RNA features survived FDR correction

out <- rbind(out, out_RNA)

## check HR for those features in dichotomized model
res_out <- res %>%
  dplyr::select(immucan_id,censoring, time_to_event,all_of(out$feature)) %>%
  mutate(across(.cols = all_of(out$feature),.fns = ~ as.factor(ntile(.,n = 2))))

p_list_final <- lapply(res_out[,-c(1:3)], function(x){
   coxph( Surv(time_to_event, censoring) ~  x ,data = res_out)
})


res_out_RNA <- cur_RNA %>%
  dplyr::select(immucan_id,censoring, time_to_event,all_of(out_RNA$feature)) %>%
  mutate(across(.cols = all_of(out_RNA$feature),.fns = ~ as.factor(ntile(.,n = 2))))

p_list_final_RNA <- lapply(res_out_RNA[,-c(1:3)], function(x){
   coxph( Surv(time_to_event, censoring) ~  x ,data = res_out_RNA)
})
  
df_2 <- data.frame(HR = unlist(lapply(p_list_final, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p_list_final, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p_list_final, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p_list_final, function(x){summary(x)$conf.int[, "upper .95"]}))) 
df_2_RNA <- data.frame(HR = unlist(lapply(p_list_final_RNA, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p_list_final_RNA, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p_list_final_RNA, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p_list_final_RNA, function(x){summary(x)$conf.int[, "upper .95"]}))) 

df_2 <- rbind(df_2, df_2_RNA)

out %>%
  mutate(feature = rownames(df_2),
         sig = as.factor(p.value < 0.05)) %>%
  ggplot(aes(x = HR,y = feature, col = sig))+
  geom_point(size = 2)+
  geom_errorbar(aes(xmin=lower_conf, xmax = upper_conf),lwd =1,width = 0.3)+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust= 1))+
  #scale_y_discrete(labels = function(x) str_split_i(x,"_",2))+
  scale_x_log10()+
  theme_bw()+
  geom_vline(xintercept = 1,linetype = "dashed")+
  geom_text(aes(label = paste0("p : ",round(p.value,digits = 3))),
            col = "black",
            vjust = -1,
            inherit.aes = TRUE)+
  scale_color_manual(values = c("TRUE" = "blue", "FALSE" = "orange"))+
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        legend.position = "none",
        axis.title.y = element_blank(),
        plot.title = element_text(hjust = 0.5))+
  ggtitle("HR for survival for binarized features")
```

## feature abundance
```{r}
feat_dat <- rbind(df, df_RNA)

feat_pval <- feat_dat %>%
  filter(p.value < 0.05) %>%
  mutate(modality = gsub("^.*_", "", feature),
         data_type = ifelse(grepl("\\.",feature),"interaction",""),
         data_type = ifelse(grepl("cello",feature),"CN",data_type),
         data_type = ifelse(grepl("density",feature),"density",data_type),
         data_type = ifelse(grepl("RNA",feature),"RNA",data_type),
         data_type = ifelse(grepl("TP53|EGFR|KRAS|STK11|PIK3CA|CDKN2A|TMB",feature),"WES",data_type),
         data_type = ifelse(grepl("TPS|CPS",feature),"PDL1",data_type))

feat_fdr <- feat_dat %>%
  mutate(FDR = p.adjust(p = p.value,method = "BH")) %>%
  filter(FDR < 0.1) %>%
  mutate(modality = gsub("^.*_", "", feature),
         data_type = ifelse(grepl("\\.",feature),"interaction",""),
         data_type = ifelse(grepl("cello",feature),"CN",data_type),
         data_type = ifelse(grepl("density",feature),"density",data_type),
         data_type = ifelse(grepl("RNA",feature),"RNA",data_type),
         data_type = ifelse(grepl("TP53|EGFR|KRAS|STK11|PIK3CA|CDKN2A|TMB",feature),"WES",data_type),
         data_type = ifelse(grepl("TPS|CPS",feature),"PDL1",data_type))

feat_abundance <- feat_dat %>%
  mutate(modality = gsub("^.*_", "", feature),
         data_type = ifelse(grepl("\\.",feature),"interaction",""),
         data_type = ifelse(grepl("cello",feature),"CN",data_type),
         data_type = ifelse(grepl("density",feature),"density",data_type),
         data_type = ifelse(grepl("RNA",feature),"RNA",data_type),
         data_type = ifelse(grepl("TP53|EGFR|KRAS|STK11|PIK3CA|CDKN2A|TMB",feature),"WES",data_type),
         data_type = ifelse(grepl("TPS|CPS",feature),"PDL1",data_type))

table(feat_pval$data_type)
table(feat_fdr$data_type)
table(feat_abundance$data_type)

plot_dat_pval <- table(feat_pval$data_type)/table(feat_abundance$data_type)[c(1:3,5,6)]
plot_dat_fdr <- table(feat_fdr$data_type)/table(feat_abundance$data_type)[c(3,5)]

df_RNA[(rownames(df_RNA) %in% c("ADGRF5","C1orf116","PHGDH","MYBL2","CENPF","KIF18B","BIRC5","FOXM1","TPX2","TROAP")),]
DT::datatable(as.data.frame(unclass(table(feat_pval$data_type))))
DT::datatable(as.data.frame(unclass(table(feat_fdr$data_type))))
DT::datatable(as.data.frame(unclass(table(feat_abundance$data_type))))
DT::datatable(as.data.frame(unclass(round(plot_dat_pval,digits = 3))))

```

__we cannot find any of the ten genes found with LF1 from MOFA when just searching for gene expression associated with survival. so we need MOFA because it identifies data structure based on all modalities and not just RNA or others.__

# Univariate COX for continuous features - LUAD

```{r, fig.height=10}
cur_traits <-traits %>%
  dplyr::select(immucan_id,TPS, CPS, TMB,TP53,EGFR, KRAS,STK11,CDKN2A,PIK3CA, censoring,time_to_event)
cur_traits <- cur_traits %>%
  mutate(across(.cols = c(TP53,EGFR, KRAS, STK11,CDKN2A,PIK3CA),.fns = ~ ifelse(. == TRUE,1,0)))
res <- left_join(dat,cur_traits,by = "immucan_id")

LUAD_samples <- traits %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  pull(immucan_id)

res <- res %>%
  dplyr::select(immucan_id,censoring,time_to_event,everything()) %>%
  filter(immucan_id %in% LUAD_samples)


p_list <- lapply(res[,-c(1:3)], function(x){
   coxph( Surv(time_to_event, censoring) ~  x ,data = res)
})

  
df <- data.frame(HR = unlist(lapply(p_list, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p_list, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p_list, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p_list, function(x){summary(x)$conf.int[, "upper .95"]}))) 
  
out <- df %>%
  mutate(FDR = p.adjust(p = p.value,method = "BH")) %>%
  filter(FDR < 0.1)
out$feature <- rownames(out)

res_out <- res %>% 
  dplyr::select(immucan_id,censoring, time_to_event,all_of(out$feature)) %>%
  mutate(across(.cols = all_of(out$feature),.fns = ~ as.factor(ntile(.,n = 2))))

p_list_final <- lapply(res_out[,-c(1:3)], function(x){
   coxph( Surv(time_to_event, censoring) ~  x ,data = res_out)
})

  
df_2 <- data.frame(HR = unlist(lapply(p_list_final, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p_list_final, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p_list_final, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p_list_final, function(x){summary(x)$conf.int[, "upper .95"]}))) 


df_2 %>%
  mutate(feature = rownames(df_2),
         sig = as.factor(p.value < 0.05)) %>%
  ggplot(aes(x = HR,y = feature, col = sig))+
  geom_point(size = 2)+
  geom_errorbar(aes(xmin=lower_conf, xmax = upper_conf),lwd =1,width = 0.3)+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust= 1))+
  #scale_y_discrete(labels = function(x) str_split_i(x,"_",2))+
  scale_x_log10()+
  theme_bw()+
  geom_vline(xintercept = 1,linetype = "dashed")+
  geom_text(aes(label = paste0("p : ",round(p.value,digits = 3))),
            col = "black",
            vjust = -1,
            inherit.aes = TRUE)+
  scale_color_manual(values = c("TRUE" = "blue", "FALSE" = "orange"))+
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        legend.position = "none",
        axis.title.y = element_blank(),
        plot.title = element_text(hjust = 0.5))+
  ggtitle("HR for surival")
```

# Univariate COX for continuous features - LUSC

no features LUSC
```{r, fig.height=10, eval=FALSE}
cur_traits <-traits %>%
  dplyr::select(immucan_id,TPS, CPS, TMB,TP53,EGFR, KRAS,STK11,CDKN2A,PIK3CA, censoring,time_to_event)
cur_traits <- cur_traits %>%
  mutate(across(.cols = c(TP53,EGFR, KRAS, STK11,CDKN2A,PIK3CA),.fns = ~ ifelse(. == TRUE,1,0)))
res <- left_join(dat,cur_traits,by = "immucan_id")

LUSC_samples <- traits %>%
  filter(simple_histology == "Squamous cell carcinoma") %>%
  pull(immucan_id)

res <- res %>%
  dplyr::select(immucan_id,censoring,time_to_event,everything()) %>%
  filter(immucan_id %in% LUSC_samples)


p_list <- lapply(res[,-c(1:3)], function(x){
   coxph( Surv(time_to_event, censoring) ~  x ,data = res)
})

  
df <- data.frame(HR = unlist(lapply(p_list, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p_list, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p_list, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p_list, function(x){summary(x)$conf.int[, "upper .95"]}))) 
  
out <- df %>%
  mutate(FDR = p.adjust(p = p.value,method = "BH")) %>%
  filter(FDR < 0.1)
out$feature <- rownames(out)

res_out <- res %>% 
  dplyr::select(immucan_id,censoring, time_to_event,all_of(out$feature)) %>%
  mutate(across(.cols = all_of(out$feature),.fns = ~ as.factor(ntile(.,n = 2))))

p_list_final <- lapply(res_out[,-c(1:3)], function(x){
   coxph( Surv(time_to_event, censoring) ~  x ,data = res_out)
})

  
df_2 <- data.frame(HR = unlist(lapply(p_list_final, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p_list_final, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p_list_final, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p_list_final, function(x){summary(x)$conf.int[, "upper .95"]}))) 


df_2 %>%
  mutate(feature = rownames(df_2),
         sig = as.factor(p.value < 0.05)) %>%
  ggplot(aes(x = HR,y = feature, col = sig))+
  geom_point(size = 2)+
  geom_errorbar(aes(xmin=lower_conf, xmax = upper_conf),lwd =1,width = 0.3)+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust= 1))+
  #scale_y_discrete(labels = function(x) str_split_i(x,"_",2))+
  scale_x_log10()+
  theme_bw()+
  geom_vline(xintercept = 1,linetype = "dashed")+
  geom_text(aes(label = paste0("p : ",round(p.value,digits = 3))),
            col = "black",
            vjust = -1,
            inherit.aes = TRUE)+
  scale_color_manual(values = c("TRUE" = "blue", "FALSE" = "orange"))+
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        legend.position = "none",
        axis.title.y = element_blank(),
        plot.title = element_text(hjust = 0.5))+
  ggtitle("HR for low infiltration ratio")
```


# Univariate COX for dichotomized features

```{r, fig.height=10}
cur_traits <- traits %>%
  dplyr::select(immucan_id,TPS, CPS, TMB,TP53,EGFR, KRAS,STK11,CDKN2A,PIK3CA, censoring,time_to_event)
cur_traits <- cur_traits %>%
  mutate(across(.cols = c(TP53,EGFR, KRAS, STK11,CDKN2A,PIK3CA),.fns = ~ ifelse(. == TRUE,1,0)))
res <- left_join(dat,cur_traits,by = "immucan_id")

res_out <- res %>%
  mutate(across(.cols =-c(immucan_id,censoring,time_to_event),.fns = ~ as.factor(ntile(.,n = 2))))

p_list_final <- lapply(res_out %>% dplyr::select(-immucan_id,-censoring, -time_to_event), function(x){
   coxph( Surv(time_to_event, censoring) ~  x ,data = res_out)
})

RNA <- read.csv(paste0(mount_path, "Figures/scores_for_sharing/MOFA/RNA_named.csv"))
RNA <- RNA %>%
  rename(immucan_id = Immucan_id) %>%
  mutate(immucan_id = gsub("_","-",immucan_id))
RNA <- left_join(RNA, cur_traits %>% dplyr::select(immucan_id,censoring, time_to_event),by = "immucan_id")
res_out_RNA <- RNA %>%
  mutate(across(.cols = -c(immucan_id,censoring,time_to_event),.fns = ~ as.factor(ntile(.,n = 2))))


p_list_final_RNA <- lapply(res_out_RNA %>% dplyr::select(-immucan_id,-censoring, -time_to_event), function(x){
   coxph( Surv(time_to_event, censoring) ~  x ,data = res_out_RNA)
})
  
df_2 <- data.frame(HR = unlist(lapply(p_list_final, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p_list_final, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p_list_final, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p_list_final, function(x){summary(x)$conf.int[, "upper .95"]}))) 
df_2_RNA <- data.frame(HR = unlist(lapply(p_list_final_RNA, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p_list_final_RNA, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p_list_final_RNA, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p_list_final_RNA, function(x){summary(x)$conf.int[, "upper .95"]}))) 

rownames(df_2_RNA) <- paste(rownames(df_2_RNA),"_RNA")
df_2 <- rbind(df_2, df_2_RNA)


# no features survive FDR correction


df_2 %>%
  mutate(feature = rownames(df_2),
         type = ifelse(grepl("density",feature),"density",""),
         type = ifelse(grepl("\\.",feature),"interaction",type),
         type = ifelse(grepl("RNA",feature),"RNA",type),
         type = ifelse(grepl("cello",feature),"CN",type),
         type = ifelse(grepl("",feature),"binary",type),
         tech = ifelse(grepl("IMC",feature),"IMC",""),
         tech = ifelse(grepl("IF1|IF2|IF3",feature),"mIF",tech),
         tech = ifelse(grepl("IF1|IF2|IF3",feature),"mIF",tech))




df_2 %>%
  mutate(feature = rownames(df_2),
         sig = as.factor(p.value < 0.05),
         effect = ifelse(HR < 1,"lower", "higher")) %>%
  filter(p.value < 0.01) %>%
  ggplot(aes(x = HR,y = feature, col = effect))+
  geom_point(size = 2)+
  geom_errorbar(aes(xmin=lower_conf, xmax = upper_conf),lwd =1,width = 0.3)+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust= 1))+
  #scale_y_discrete(labels = function(x) str_split_i(x,"_",2))+
  scale_x_log10()+
  theme_bw()+
  geom_vline(xintercept = 1,linetype = "dashed")+
  geom_text(aes(label = paste0("p : ",round(p.value,digits = 3))),
            col = "black",
            vjust = -1,
            inherit.aes = TRUE)+
  scale_color_manual(values = c("lower" = "blue", "higher" = "orange"))+
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        legend.position = "none",
        axis.title.y = element_blank(),
        plot.title = element_text(hjust = 0.5))+
  ggtitle("HR for survival for binarized features")


```

## feature abundance
```{r, fig.height=3, fig.width=3}
feat_dat <- df_2

feat_pval <- feat_dat %>%
  mutate(feature = rownames(df_2)) %>%
  filter(p.value < 0.05) %>%
  mutate(modality = gsub("^.*_", "", feature),
         data_type = ifelse(grepl("\\.",feature),"interaction",""),
         data_type = ifelse(grepl("cello",feature),"CN",data_type),
         data_type = ifelse(grepl("density",feature),"density",data_type),
         data_type = ifelse(grepl("RNA",feature),"RNA",data_type),
         data_type = ifelse(grepl("TP53|EGFR|KRAS|STK11|PIK3CA|CDKN2A|TMB",feature),"WES",data_type),
         data_type = ifelse(grepl("TPS|CPS",feature),"PDL1",data_type))

feat_abundance <- feat_dat %>%
  mutate(feature = rownames(df_2),
         modality = gsub("^.*_", "", feature),
         data_type = ifelse(grepl("\\.",feature),"interaction",""),
         data_type = ifelse(grepl("cello",feature),"CN",data_type),
         data_type = ifelse(grepl("density",feature),"density",data_type),
         data_type = ifelse(grepl("RNA",feature),"RNA",data_type),
         data_type = ifelse(grepl("TP53|EGFR|KRAS|STK11|PIK3CA|CDKN2A|TMB",feature),"WES",data_type),
         data_type = ifelse(grepl("TPS|CPS",feature),"PDL1",data_type))

table(feat_pval$data_type)
table(feat_abundance$data_type)

plot_dat_pval <- table(feat_pval$data_type)/table(feat_abundance$data_type)

df_RNA[(rownames(df_RNA) %in% c("ADGRF5","C1orf116","PHGDH","MYBL2","CENPF","KIF18B","BIRC5","FOXM1","TPX2","TROAP")),]
DT::datatable(as.data.frame(unclass(table(feat_pval$data_type))))
DT::datatable(as.data.frame(unclass(table(feat_abundance$data_type))))
DT::datatable(as.data.frame(unclass(round(plot_dat_pval,digits = 3))))

plot_dat <-as.data.frame(table(feat_pval$data_type))
plot_dat$total_count = as.data.frame(table(feat_abundance$data_type))$Freq

plot_dat %>%
  ggplot(aes(x = total_count,y = Freq))+
  geom_point()+
  theme_bw()+
  scale_x_log10()+
  scale_y_log10()
```





# LASSO on dichotomozed data

```{r}
clin_dat <- read_csv("O:/projects/immucan/clinical_data/Lung_NSCLC2/20230825/NSCLC2_for_R_update_2024.csv")
df <- as.data.frame(apply(dat, 2, function(x) ntile(x,2)))
rownames(df) <- rownames(dat)


time <- clin_dat %>%
  filter(immucan_id %in% rownames(df)) %>%
  pull(time_to_event)
censoring <- clin_dat %>%
  filter(immucan_id %in% rownames(df)) %>%
  pull(censoring)


y <-  survival::Surv(time, censoring)

# generate folds for the cross validation
nfolds <- 5
set.seed(2135)
foldids <- sample(rep(1:nfolds, length.out = nrow(df)))  # Assign each observation to one of 5 folds


#determine lambda
set.seed(12354)
cv.fit <- cv.glmnet(as.matrix(df),y, family = "cox", alpha = 1, foldid = foldids)
lambda.min <- cv.fit$lambda.min
plot(cv.fit)
set.seed(2351)
fit <- glmnet(as.matrix(df), survival::Surv(time, censoring),lambda = lambda.min, family = "cox", alpha = 1)



# find coefficients at minimum lambda
Coefficients <- coef(fit)

Lasso_coefs <- as.data.frame(as.matrix(Coefficients))
Lasso_coefs <- Lasso_coefs %>%
  mutate(feature = rownames(Lasso_coefs),
         coef = s0,
         s0 = NULL) %>%
  filter(coef != 0,
         !grepl("smoking|simple_",feature))
sig_var_LASSO <- as.vector(Lasso_coefs$feature)

DT::datatable(Lasso_coefs)
```

# Test LASSO features in multivariate model {.tabset}
Note: the features extracted from LASSO are extracted based on a univariate model. Here we will build multivariate models and only keep those features that are significant in a multivariate model.
```{r,results='asis',fig.height=5, fig.width=21}
df_text <- as.data.frame(apply(df, 2, function(x) ifelse(x == 1,"low","high")))
df_text$immucan_id <- rownames(df)

# function to generate the header for the plotting with tabset
catHeader <- function(text = "", level = 3) {
    cat(paste0("\n\n", 
               paste(rep("#", level), collapse = ""), 
               " ", text, "\n"))
}

sig_var_LASSO_in_cox <- vector()

p.list <- list()
hr.list <- list()
factor.list <- list()
for (i in sig_var_LASSO) {

  plot_dat <- left_join(clin_dat,df_text,"immucan_id")

  plot_dat <- plot_dat %>%
    filter(!is.na(!!as.name(i)))

  
  plot_dat[,i] <- factor(pull(plot_dat,var = i),levels = c("low","high"))


  res <- plot_dat %>%
  analyse_multivariate(vars(time_to_event, censoring),
                       c("simple_stage","simple_histology","smoking_status","adjuvant_treatment","neo_adjuvant_treatment",i))



  pval <- res$summaryAsFrame %>%
    filter(grepl(i,factor.id)) %>%
    pull(p)
  if(round(pval,digits = 3) < 0.05 ) {
    sig_var_LASSO_in_cox <- c(sig_var_LASSO_in_cox,paste0(i))
  }
  
  p.list[[i]] <- forest_plot(res)
  hr.list[[i]] <- res$summaryAsFrame$HR
  factor.list[[i]] <- res$summaryAsFrame$factor.name
  p.list[[i]] <- list(p.list[[i]],i)
}

for(i in seq_along(p.list)){
    tmp <- p.list[[i]]
    # As you want to use tabset level here has to be lower than 
    # parent level (ie, parent is 1, so here you have to use 2)
    catHeader(tmp[[2]], 2)
    print(tmp[[1]])
}

```

#{}

# multivariate cox for top features

```{r, fig.height=10}
HRs <- data.frame(HRs = unlist(hr.list),
                  factor.ids = unlist(factor.list)) %>%
  filter(!grepl("smoking|simple|adjuvant",factor.ids))
top_hrs <- HRs %>%
  arrange(HRs) %>%
  pull(factor.ids)

#top_hrs <- top_hrs[c(1:2,(length(top_hrs)-1):length(top_hrs))]

clin_dat <- read_csv("O:/projects/immucan/clinical_data/Lung_NSCLC2/20230825/NSCLC2_for_R_update_2024.csv")

clin_dat <- df_text %>%
  dplyr::select(immucan_id,top_hrs) %>%
  left_join(clin_dat,by = "immucan_id") %>%
  filter(simple_histology != "unknown")

time_to_event <- clin_dat %>%
  filter(immucan_id %in% df_text$immucan_id) %>%
  pull(time_to_event)
censoring <- clin_dat %>%
  filter(immucan_id %in% df_text$immucan_id) %>%
  pull(censoring)


y <- Surv(time_to_event, censoring)
p <- ggforest(coxph(formula(c(paste0("y ~",  "simple_stage +" , "simple_histology + " , "smoking_status +" , "adjuvant_treatment +" , "neo_adjuvant_treatment +" ), paste0(top_hrs,collapse = "+"))) ,data = as.data.frame(clin_dat)))
p
```

# clustering on selected features
```{r}
top_hrs <- HRs %>%
  arrange(HRs) %>%
  pull(factor.ids)

#top_hrs <- top_hrs[c(1:2,(length(top_hrs)-1):length(top_hrs))]
#top_hrs <- top_hrs[c(1:5,(length(top_hrs)-4):length(top_hrs))]


plot_dat <- dat %>%
  dplyr::select(all_of(top_hrs))

set.seed(12345)
cur_clust <- kmeans(plot_dat,centers = 3)

plot_dat$kmeans_cluster <- cur_clust$cluster[rownames(plot_dat)]

clin_dat <- read_csv("O:/projects/immucan/clinical_data/Lung_NSCLC2/20230825/NSCLC2_for_R_update_2024.csv")
clin_dat$clusters <- as.factor(cur_clust$cluster[clin_dat$immucan_id])
clin_dat <- clin_dat %>%
  filter(simple_histology != "unknown")

p <- ggforest(coxph(Surv(time_to_event, censoring) ~ simple_stage + simple_histology + smoking_status + adjuvant_treatment + neo_adjuvant_treatment + clusters,data = as.data.frame(clin_dat)))
p
table(clin_dat$censoring,clin_dat$clusters)
```

## cluster heatmap
```{r}
library(pheatmap)
mat <- plot_dat %>%
  pivot_longer(cols = top_hrs) %>%
  group_by(name, kmeans_cluster, .drop = FALSE) %>%
  summarise(cluster_mean = mean(value)) %>%
  pivot_wider(id_cols = name,values_from = cluster_mean,names_from = kmeans_cluster)
mat_names <- mat$name
mat$name <- NULL
mat <- as.matrix(mat)
rownames(mat) <- mat_names

pheatmap(mat,color = viridis::viridis(100),cluster_rows = TRUE,clustering_method = "ward.D2")

pheatmap(mat,scale = "row",color = colorRampPalette(c("dark blue", "white", "dark red"))(100),cluster_rows = TRUE,clustering_method = "ward.D2")

most_different <- c("mregDC.CD4_IMC2","CD8_GB.Tumor_IF2","HLADR.Tumor_IMC1","B.Tumor_IF1","CD8.Tumor_IMC1")
```


## Kaplan Meier

```{r, fig.height=6}
clin_dat <- read_csv("O:/projects/immucan/clinical_data/Lung_NSCLC2/20230825/NSCLC2_for_R_update_2024.csv")

clin_dat$cluster <- cur_clust$cluster[clin_dat$immucan_id]

surv_fit <- survfit(Surv(time_to_event, censoring) ~ cluster, data = clin_dat)

ggsurv <- ggsurvplot(surv_fit, data = clin_dat, risk.table = TRUE, pval = TRUE)
ggsurv$plot <- ggsurv$plot + 
theme(legend.position = "none")
ggsurv
```
