---
title: "imaging_module_generation"
author: "Daniel Schulz"
date: "2026-07-03"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# imaging modules

Cluster all imaging features by co-variation across patients.
exp_... and density features live on very different scales, so every
feature is z-scored first. The feature × feature Pearson correlation
matrix is then hierarchically clustered to reveal groups of co-varying
features. Each cluster is summarised as a mean z-score and correlated
with Bagaev RNA axes for biological annotation.

Outputs:
figures/feature_correlation_heatmap.pdf  — feature × feature r
figures/feature_cluster_heatmap.pdf       — features × patients
figures/cluster_rna_heatmap.pdf           — cluster × Bagaev r
data/processed/imaging_feature_clusters.csv  — feature → cluster table

```{r}
library(ComplexHeatmap)
library(circlize)
library(ggplot2)
library(dplyr)
library(tidyr)

OUT  <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing"

FIGURES <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/revised_figures/"

# ── Tuning ────────────────────────────────────────────────────────────────────
MIN_PATIENTS     <- 20     # drop features with fewer non-NA patients
HC_METHOD        <- "ward.D2"
N_FEAT_CLUSTERS  <- NULL   # NULL → TREE_HEIGHT auto-cut; integer overrides
TREE_HEIGHT      <- 0.8    # cut on (1-r)/2 scale: 0.4 ≈ r > 0.2 within cluster
MIN_CLUSTER_SIZE <- 3      # drop clusters smaller than this (noise / singletons)
N_META_CLUSTERS  <- 8      # number of meta-clusters (group imaging clusters by RNA profile)

# ── Load ──────────────────────────────────────────────────────────────────────
imaging_mat <- readRDS(file.path(OUT, "imaging_mat.rds"))
rna_scores  <- readRDS(file.path(OUT, "rna_scores.rds"))

cat("Imaging matrix:", nrow(imaging_mat), "patients ×", ncol(imaging_mat), "features\n")
cat("RNA scores    :", nrow(rna_scores),  "patients ×", ncol(rna_scores),  "axes\n")

# ── 1. Filter low-coverage features ──────────────────────────────────────────
n_obs    <- colSums(!is.na(imaging_mat))
keep     <- names(n_obs)[n_obs >= MIN_PATIENTS]
img_filt <- imaging_mat[, keep, drop = FALSE]
cat("Retained", length(keep), "/", ncol(imaging_mat),
    "features with ≥", MIN_PATIENTS, "patients\n")

```


# cluster imaging features

```{r}
img_z <- scale(img_filt)

# Remove any features that had zero variance (scale() returns NaN)
zero_var <- apply(img_z, 2, function(x) any(is.nan(x)))
if (any(zero_var)) {
  cat("Dropping", sum(zero_var), "zero-variance features\n")
  img_z <- img_z[, !zero_var, drop = FALSE]
}
cat("Final feature set for clustering:", ncol(img_z), "features\n")

# ── 3. Feature × feature Pearson correlation ─────────────────────────────────
feat_cor <- cor(img_z, use = "pairwise.complete.obs")

# ── 4. Hierarchical clustering of features ────────────────────────────────────
# (1 - r) / 2 maps r ∈ [-1, 1] → distance ∈ [0, 1], preserving sign:
# anti-correlated features (r = -1) get max distance.
feat_dist <- as.dist((1 - feat_cor) / 2)
feat_hc   <- hclust(feat_dist, method = HC_METHOD)

raw_cl <- if (!is.null(N_FEAT_CLUSTERS)) {
  cutree(feat_hc, k = N_FEAT_CLUSTERS)
} else {
  cutree(feat_hc, h = TREE_HEIGHT)
}

# Keep only clusters with ≥ MIN_CLUSTER_SIZE features
valid_cl  <- as.integer(names(which(table(raw_cl) >= MIN_CLUSTER_SIZE)))
cl_keep   <- raw_cl[raw_cl %in% valid_cl]

# Relabel clusters 1:K in left-to-right dendrogram order
dend_cl_order <- unique(raw_cl[feat_hc$labels[feat_hc$order]])
dend_cl_order <- dend_cl_order[dend_cl_order %in% valid_cl]
lbl_map       <- setNames(seq_along(dend_cl_order), as.character(dend_cl_order))
feat_cl       <- lbl_map[as.character(cl_keep)]
names(feat_cl) <- names(cl_keep)

K <- max(feat_cl)
n_dropped <- sum(!raw_cl %in% valid_cl)
cat("\nCut at height", TREE_HEIGHT, "→", K, "clusters",
    "(", n_dropped, "features in clusters <", MIN_CLUSTER_SIZE, "dropped )\n")
cat("Cluster sizes:\n"); print(sort(table(feat_cl), decreasing = TRUE))

# Within each cluster, identify the most central feature
# (highest mean |r| with other cluster members)
central_features <- sapply(1:K, function(cl) {
  feats <- names(feat_cl)[feat_cl == cl]
  if (length(feats) == 1) return(feats)
  sub   <- feat_cor[feats, feats, drop = FALSE]
  diag(sub) <- NA
  mean_r <- rowMeans(abs(sub), na.rm = TRUE)
  names(which.max(mean_r))
})
cat("\nMost central feature per cluster:\n")
for (cl in 1:K)
  cat(sprintf("  C%02d: %s\n", cl, central_features[cl]))
```


# mean expression across clusters

Here we will calculate the mean of z-scored features for the K initial clusters, NA-tolerant
```{r}
cluster_scores <- do.call(cbind, lapply(1:K, function(cl) {
  feats <- names(feat_cl)[feat_cl == cl]
  sub   <- img_z[rownames(imaging_mat), feats, drop = FALSE]
  if (length(feats) == 1) as.numeric(sub[, 1])
  else rowMeans(sub, na.rm = TRUE)
}))
colnames(cluster_scores) <- paste0("C", sprintf("%02d", 1:K))
rownames(cluster_scores) <- rownames(imaging_mat)
```

# meta clustering with RNA

Hierarchically cluster the K imaging clusters using their Bagaev RNA
correlation profiles as coordinates. Cutting this dendrogram at
N_META_CLUSTERS groups clusters with similar biological associations.

```{r}
# ── 6. Cluster × Bagaev RNA axis correlations ─────────────────────────────────
shared_pts <- intersect(rownames(cluster_scores), rownames(rna_scores))
cat("\nPatients with both imaging and RNA data:", length(shared_pts), "\n")

cl_rna_cor <- cor(
  cluster_scores[shared_pts, , drop = FALSE],
  rna_scores[shared_pts,     , drop = FALSE],
  use = "pairwise.complete.obs"
)

cl_dist  <- dist(cl_rna_cor,    method = "euclidean")
rna_dist <- dist(t(cl_rna_cor), method = "euclidean")
cl_hc2   <- hclust(cl_dist,  method = "ward.D2")
rna_hc2  <- hclust(rna_dist, method = "ward.D2")

meta_raw  <- cutree(cl_hc2, k = N_META_CLUSTERS)
meta_dend_order <- unique(meta_raw[cl_hc2$order])
meta_lbl_map    <- setNames(seq_along(meta_dend_order), as.character(meta_dend_order))
meta_cl         <- meta_lbl_map[as.character(meta_raw)]  # cluster label → meta-cluster
names(meta_cl)  <- paste0("C", sprintf("%02d", 1:K))

cat("\nMeta-cluster sizes (N =", N_META_CLUSTERS, "):\n")
print(table(meta_cl))

# Map meta-cluster assignment down to individual features
feat_meta <- meta_cl[paste0("C", sprintf("%02d", feat_cl))]
names(feat_meta) <- names(feat_cl)

meta_col <- setNames(
  colorRampPalette(c("#E64B35","#4DBBD5","#00A087","#3C5488",
                     "#F39B7F","#91D1C2","#8491B4","#DC0000"))(N_META_CLUSTERS),
  as.character(1:N_META_CLUSTERS))
```

# Figure of imaging clusters
```{r, fig.height=8, fig.width=9}

# Features in original co-variance dendrogram order (shows correlation blocks)
feat_sorted  <- names(feat_cl)[feat_hc$labels[feat_hc$order] %in% names(feat_cl)]
feat_sorted  <- feat_sorted[!is.na(feat_sorted)]
cl_sorted    <- feat_cl[feat_sorted]
meta_sorted  <- feat_meta[feat_sorted]
panel_sorted <- sub("_.*", "", feat_sorted)

cluster_col <- setNames(
  colorRampPalette(c("#3C5488","#E64B35","#00A087","#F39B7F","#8491B4",
                     "#91D1C2","#DC0000","#7E6148","#4DBBD5","#B09C85",
                     "#FFDC91","#80B1D3","#FB8072","#BC80BD","#CCEBC5"))(K),
  as.character(1:K))

panel_pal <- setNames(
  colorRampPalette(c("#66C2A5","#FC8D62","#8DA0CB","#E78AC3",
                     "#A6D854","#FFD92F","#E5C494","#B3B3B3"))(
    length(unique(panel_sorted))),
  sort(unique(panel_sorted)))

row_ann <- rowAnnotation(
  #Meta    = as.character(meta_sorted),
  Panel   = panel_sorted,
  Cluster = as.character(cl_sorted),
  col     = list(Meta    = meta_col,
                 Panel   = panel_pal,
                 Cluster = cluster_col),
  annotation_name_gp = gpar(fontsize = 10),
  show_annotation_name = TRUE,
  width = unit(9, "mm"))

cor_col <- colorRamp2(c(-1, 0, 1), c("#2166AC", "white", "#B2182B"))

ht_cor <- Heatmap(
  feat_cor[feat_sorted, feat_sorted],
  name              = "Pearson r",
  col               = cor_col,
  cluster_rows      = FALSE,
  row_title = NULL,
  cluster_columns   = FALSE,
  row_split         = factor(cl_sorted, levels = 1:K),
  column_split      = factor(cl_sorted, levels = 1:K),
  show_row_names    = FALSE,
  show_column_names = FALSE,
  left_annotation   = row_ann,
  column_title      = paste0("Imaging feature correlations  (",
                              length(feat_sorted), " features, ", K, " clusters)"),
  column_title_gp   = gpar(fontsize = 12),
  row_title_gp      = gpar(fontsize = 10),
  column_title_rot  = 0,
  heatmap_legend_param = list(title = "Pearson r")
)

pdf("/mnt/immucan_volume/processed_data/NSCLC2_cohort/revised_figures/Figure_2_supp/S2A_feature_correlation_heatmap.pdf",
    width = 14, height = 8)
draw(ht_cor)
dev.off()
cat("Feature correlation heatmap saved.\n")
```


```{r, fig.height=15, fig.width=10}
# Patient ordering: Ward.D2 HC on cluster scores (NA-tolerant mean scores)
cs_complete <- cluster_scores[rowSums(is.na(cluster_scores)) == 0, , drop = FALSE]
pat_hc      <- hclust(dist(cs_complete, method = "euclidean"), method = "ward.D2")
pat_ordered <- c(rownames(cs_complete)[pat_hc$order],
                 setdiff(rownames(imaging_mat), rownames(cs_complete)))

# Top annotation: all Bagaev RNA axes, scaled 0-1 for display
rna_ann_mat <- matrix(NA_real_,
                      nrow = length(pat_ordered),
                      ncol = ncol(rna_scores),
                      dimnames = list(pat_ordered, colnames(rna_scores)))
rna_shared <- intersect(pat_ordered, rownames(rna_scores))
rna_ann_mat[rna_shared, ] <- as.matrix(rna_scores[rna_shared, , drop = FALSE])

# Center each RNA axis for display
rna_ann_scaled <- t(scale(rna_ann_mat))

rna_col <- colorRamp2(c(-2, 0, 2), c("#4DBBD5", "white", "#E64B35"))

top_ann <- HeatmapAnnotation(
  df  = as.data.frame(t(rna_ann_scaled)),
  col = setNames(
    lapply(seq_len(nrow(rna_ann_scaled)), function(i) rna_col),
    rownames(rna_ann_scaled)),
  show_legend        = FALSE,
  show_annotation_name = TRUE,
  annotation_name_gp = gpar(fontsize = 6),
  height             = unit(3 * nrow(rna_ann_scaled), "mm")
)

# Re-order features by RNA meta-cluster (cl_hc2 order), preserving within-cluster
# dendrogram order so the feature correlation structure is still visible within blocks.
cl_in_rna_order  <- as.integer(sub("^C0*", "", rownames(cl_rna_cor)[cl_hc2$order]))
feat_sorted_meta <- unlist(lapply(cl_in_rna_order, function(cl) {
  cl_feats <- names(feat_cl)[feat_cl == cl]
  feat_sorted[feat_sorted %in% cl_feats]   # preserve original dendrogram order within cluster
}))
cl_sorted_meta   <- feat_cl[feat_sorted_meta]
meta_sorted_main <- feat_meta[feat_sorted_meta]
panel_sorted_meta <- sub("_.*", "", feat_sorted_meta)

plot_mat <- t(img_z[pat_ordered, feat_sorted_meta, drop = FALSE])
feat_col <- colorRamp2(c(-3, 0, 3), c("#2166AC", "white", "#B2182B"))

row_ann2 <- rowAnnotation(
  Meta  = as.character(meta_sorted_main),
  Panel = panel_sorted_meta,
  col   = list(Meta  = meta_col,
               Panel = panel_pal),
  annotation_name_gp = gpar(fontsize = 8),
  width = unit(6, "mm"),
  show_legend = FALSE)

# assign names to meta clusters
meta_cluster_names <-structure(c("Cold","Proliferating","Highly_inflamed","Thelper_MHCII_enriched","Inflamed_proliferating","Medium_inflamed","Neutrophil_enriched","Mixed"),
                               names = c(paste0("M",c(1:8))))

ht_main <- Heatmap(
  plot_mat,
  name               = "z-score",
  col                = feat_col,
  cluster_rows       = FALSE,
  cluster_columns    = FALSE,
  row_split          = factor(meta_sorted_main, levels = 1:N_META_CLUSTERS),
  show_row_names     = FALSE,
  show_column_names  = FALSE,
  top_annotation     = top_ann,
  left_annotation    = row_ann2,
  row_title          = meta_cluster_names[paste0("M", 1:N_META_CLUSTERS)],
  row_title_gp       = gpar(fontsize = 9, fontface = "bold"),
  column_title       = paste0("Imaging features — ", N_META_CLUSTERS,
                               " meta-clusters  (", length(pat_ordered), " patients)"),
  column_title_gp    = gpar(fontsize = 11),
  heatmap_legend_param = list(title = "z-score")
)

h_per_feature <- 1.2  # mm per feature row
fig_h <- max(8, length(feat_sorted) * h_per_feature / 25.4 + 3)  # inches

pdf(file.path(FIGURES, "feature_cluster_heatmap.pdf"),
    width = 16, height = fig_h)
draw(ht_main)
dev.off()
cat("Feature × patient heatmap saved.\n")

```


```{r, fig.height=8, fig.width=10}
# ── 9. Figure: cluster × Bagaev RNA axis correlation heatmap ─────────────────
# Rows (imaging clusters) ordered by cl_hc2, split by meta-cluster.
# RNA axes (columns) ordered by rna_hc2.
# cl_hc2 and rna_hc2 were computed in block 6b.

cl_rna_ordered  <- cl_rna_cor[cl_hc2$order, rna_hc2$order, drop = FALSE]
meta_cl_ordered <- meta_cl[rownames(cl_rna_ordered)]

row_ann_rna <- rowAnnotation(
  Meta = as.character(meta_cl_ordered),
  col  = list(Meta = meta_col),
  annotation_name_gp = gpar(fontsize = 8),
  width = unit(4, "mm"),
  show_legend = FALSE)

r_range <- max(abs(cl_rna_cor), na.rm = TRUE)
r_col   <- colorRamp2(c(-r_range, 0, r_range), c("#2166AC", "white", "#B2182B"))

ht_rna <- Heatmap(
  cl_rna_ordered,
  name              = "Pearson r",
  col               = r_col,
  cluster_rows      = FALSE,
  cluster_columns   = FALSE,
  row_split         = factor(meta_cl_ordered, levels = 1:N_META_CLUSTERS),
  row_names_gp      = gpar(fontsize = 12),
  column_names_gp   = gpar(fontsize = 12),
  row_title_rot = 0,
  column_names_rot  = 45,
  row_labels        = paste0(rownames(cl_rna_ordered), " (",
                              table(feat_cl)[sub("^C0*", "", rownames(cl_rna_ordered))],
                              ")"),
  left_annotation   = row_ann_rna,
  row_title         = meta_cluster_names[paste0("M", 1:N_META_CLUSTERS)],
  row_title_gp      = gpar(fontsize = 12, fontface = "bold"),
  column_title      = "Imaging feature clusters × Bagaev RNA axes",
  column_title_gp   = gpar(fontsize = 11),
  cell_fun          = function(j, i, x, y, width, height, fill) {
    r <- cl_rna_ordered[i, j]
    if (!is.na(r) && abs(r) >= 0.3)
      grid.text(sprintf("%.2f", r), x, y, gp = gpar(fontsize = 7))
  },
  heatmap_legend_param = list(
    title       = "Pearson r",
    title_gp    = gpar(fontsize = 14, fontface = "bold"),
    labels_gp   = gpar(fontsize = 12),
    legend_height = unit(6, "cm"),
    grid_width  = unit(0.6, "cm")
  )
)

pdf("/mnt/immucan_volume/processed_data/NSCLC2_cohort/revised_figures/Figure_2/2A_cluster_rna_heatmap.pdf",
    width  =12,
    height = 12)
draw(ht_rna)
dev.off()
cat("Cluster × RNA correlation heatmap saved.\n")

```


# feature export
```{r}
# Feature → meta-cluster assignment (all imaging features)
cluster_df <- data.frame(
  feature           = names(feat_cl),
  cluster           = as.integer(feat_cl),
  meta_cluster      = as.integer(feat_meta[names(feat_cl)]),
  meta_cluster_name = meta_cluster_names[paste0("M", feat_meta[names(feat_cl)])],
  panel             = sub("_.*", "", names(feat_cl)),
  central_feature   = names(feat_cl) %in% central_features,
  stringsAsFactors  = FALSE
)
cluster_df <- cluster_df[order(cluster_df$meta_cluster, cluster_df$cluster), ]
write.csv(cluster_df,
          file.path(OUT, "imaging_feature_clusters.csv"),
          row.names = FALSE)

# Per-patient meta-cluster scores (mean of z-scored features per meta-cluster)
meta_scores <- do.call(cbind, lapply(1:N_META_CLUSTERS, function(mc) {
  feats <- names(feat_meta)[feat_meta == mc]
  sub   <- img_z[rownames(imaging_mat), feats, drop = FALSE]
  if (length(feats) == 1) as.numeric(sub[, 1])
  else rowMeans(sub, na.rm = TRUE)
}))
colnames(meta_scores) <- meta_cluster_names[paste0("M", 1:N_META_CLUSTERS)]
rownames(meta_scores) <- rownames(imaging_mat)

# Set all-NA patients to NA (missing entire meta-cluster)
all_miss <- rowSums(!is.na(img_z[rownames(imaging_mat),
                                  names(feat_meta), drop = FALSE])) == 0
meta_scores[all_miss, ] <- NA

write.csv(
  data.frame(immucan_id = rownames(meta_scores), meta_scores, check.names = FALSE),
  file.path(OUT, "patient_imaging_meta_cluster_scores.csv"),
  row.names = FALSE)
saveRDS(meta_scores, file.path(OUT, "patient_imaging_meta_cluster_scores.rds"))

# Also save fine-grained cluster scores
#saveRDS(cluster_scores, file.path(OUT, "imaging_cluster_scores.rds"))

```

# Patient clustering based on imaging modules

```{r}
# Cluster patients by Ward.D2 HC on the 8 meta-cluster scores.

meta_complete  <- meta_scores[rowSums(is.na(meta_scores)) == 0, , drop = FALSE]
meta_z         <- scale(meta_complete)   # z-score across patients for display
meta_pat_hc    <- hclust(dist(meta_z, method = "euclidean"), method = "ward.D2")

# Silhouette sweep to suggest k
d_meta <- dist(meta_z, method = "euclidean")
sil_meta <- sapply(2:8, function(k) {
  km  <- kmeans(meta_z, centers = k, nstart = 50, iter.max = 100)
  mean(cluster::silhouette(km$cluster, d_meta)[, "sil_width"])
})
names(sil_meta) <- 2:8
best_k_meta <- as.integer(names(which.max(sil_meta)))
cat("\nMeta-cluster patient silhouette (k-means):", paste(round(sil_meta, 3), collapse=" "), "\n")
cat("Suggested k =", best_k_meta, "\n")

# we chose 6 patient clusters based on biological interpretation. the silhouette scores are very similar for 5, 6, 7, and 8 patient clusters
best_k_meta <- 6

set.seed(42)
km_meta   <- kmeans(meta_z, centers = best_k_meta, nstart = 50, iter.max = 100)
pat_cl    <- km_meta$cluster

# Order patients: by cluster, within-cluster by Ward.D2
pat_col_order <- unlist(lapply(sort(unique(pat_cl)), function(cl) {
  idx <- which(pat_cl == cl)
  if (length(idx) <= 2) return(idx)
  sub_hc <- hclust(dist(meta_z[idx, , drop = FALSE], method = "euclidean"),
                   method = "ward.D2")
  idx[sub_hc$order]
}))
pat_cl_ord <- factor(pat_cl[pat_col_order], levels = sort(unique(pat_cl)))

pat_col_pal <- setNames(
  colorRampPalette(c("#4DBBD5","#E64B35","#00A087","#3C5488",
                     "#F39B7F","#8491B4","#91D1C2","#DC0000"))(best_k_meta),
  as.character(1:best_k_meta))

meta_col_ann <- HeatmapAnnotation(
  Patient_cluster = as.character(pat_cl_ord),
  col = list(Patient_cluster = pat_col_pal),
  annotation_name_gp = gpar(fontsize = 9))

meta_feat_col <- colorRamp2(c(-2, 0, 2), c("#2166AC", "white", "#B2182B"))

ht_meta_pat <- Heatmap(
  t(meta_z)[, pat_col_order],
  name               = "z-score",
  col                = meta_feat_col,
  cluster_rows       = TRUE,
  cluster_columns    = FALSE,
  column_split       = pat_cl_ord,
  show_column_names  = FALSE,
  row_names_gp       = gpar(fontsize = 11, fontface = "bold"),
  row_names_side     = "left",
  top_annotation     = meta_col_ann,
  column_title_gp    = gpar(fontsize = 9),
  column_title       = paste0("Patient clusters on imaging meta-modules  (k=",
                               best_k_meta, ", n=", nrow(meta_complete), ")"),
  heatmap_legend_param = list(title = "z-score")
)

pdf(file.path(FIGURES, "patient_meta_cluster_heatmap.pdf"), width = 14, height = 5)
draw(ht_meta_pat)
dev.off()


# Save patient cluster assignments
write.csv(
  data.frame(immucan_id = rownames(meta_complete)[pat_col_order],
             patient_cluster = as.integer(pat_cl_ord),
             stringsAsFactors = FALSE),
  file.path(OUT, "patient_clusters_meta_imaging.csv"),
  row.names = FALSE)
```

# correlation with mofa LFs

```{r}
mofa <- readRDS( '/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/MOFA/mofa_V6_minor.rds')
lf_raw <- as.data.frame(get_factors(mofa))

lf_raw <- lf_raw %>%
  mutate(immucan_id = rownames(lf_raw),
         immucan_id = gsub("_","-",immucan_id))
colnames(lf_raw) <- gsub("group1\\.","",colnames(lf_raw))

meta_scores <- as.data.frame(meta_scores)
meta_scores$immucan_id <- rownames(meta_scores)
rownames(meta_scores) <- NULL

df <- left_join(meta_scores,lf_raw,"immucan_id")


cor(df %>%
      dplyr::select(-immucan_id), use = "pairwise.complete.obs") %>%
  as.data.frame() %>%
  dplyr::select(Factor1, Factor2, Factor3, Factor4, Factor5, Factor6) %>%
  dplyr::slice(1:8) %>%
  pheatmap(color = colorRampPalette(c("dark blue", "white", "dark red"))(100),
         breaks = seq(-1, 1, length.out = 101))
```

# survival association

```{r}
library(tidyverse)
library(survival)
meta_scores
traits <- readr::read_csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/all_traits.csv")
cur_dat <- left_join(meta_scores, traits,by= "immucan_id") %>%
  ungroup() %>%
  mutate(LUAD_vs_LUSC = ifelse(is.na(LUAD_vs_LUSC),"Other",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma","LUAD",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Squamous cell carcinoma","LUSC",LUAD_vs_LUSC))



p.list <- list()
for(j in c("LUAD", "LUSC", "Other")) {
  df <- cur_dat %>%
    filter(LUAD_vs_LUSC == j)
  
  surv_ob <- survival::Surv(df$time_to_event,df$censoring)
  for (i in colnames(meta_scores)[1:8]) {
    p.list[[paste0(j, "_", i)]] <- survival::coxph( as.formula(paste0("surv_ob ~ ", i)),
                                                    data = df) 
  }

}

df <- data.frame(HR =      unlist(lapply(p.list, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p.list, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "upper .95"]}))) 

df1 <- df %>% 
  mutate(hist = stringr::str_split_i(rownames(df),"_",1),
         feature = gsub("LUAD_|LUSC_|Other_","", rownames(df)))
  
df1

p <- df1 %>%
  ggplot(aes(x = HR,y = feature, col = hist))+
  geom_point(size = 2)+
  facet_wrap(.~hist)+
  geom_errorbar(aes(xmin=lower_conf, xmax = upper_conf),lwd =1,width = 0.3)+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust= 1))+
  #scale_y_discrete(labels = function(x) str_split_i(x,"_",2))+
  scale_x_log10()+
  theme_bw()+
  geom_vline(xintercept = 1,linetype = "dashed")+
  geom_text(aes(label = paste0("p : ",round(p.value,digits = 3))),
            col = "black",
            vjust = -1,
            inherit.aes = TRUE)+
  scale_color_manual(values = c("LUAD" = "blue", "LUSC" = "orange", "Other" = "gray40", "All" = "black"),name = "Histology")+
  theme(axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10))+
  labs(x = "HR",y = "Histology")

p
```

# cluster scores

```{r, fig.height=8, fig.width=6}
cluster_scores <- as.data.frame(cluster_scores)
cluster_scores <- cluster_scores %>%
  mutate(immucan_id = rownames(cluster_scores))
rownames(cluster_scores) <- NULL
cur_dat <- left_join(cluster_scores, traits,by= "immucan_id") %>%
  ungroup() %>%
  mutate(LUAD_vs_LUSC = ifelse(is.na(LUAD_vs_LUSC),"Other",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma","LUAD",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Squamous cell carcinoma","LUSC",LUAD_vs_LUSC))


p.list <- list()
for(j in c("LUAD", "LUSC", "Other")) {
  df <- cur_dat %>%
    filter(LUAD_vs_LUSC == j)
  
  surv_ob <- survival::Surv(df$time_to_event,df$censoring)
  for (i in colnames(cur_dat)[2:30]) {
    p.list[[paste0(j, "_", i)]] <- survival::coxph( as.formula(paste0("surv_ob ~ ", i)),
                                                    data = df) 
  }

}

df <- data.frame(HR =      unlist(lapply(p.list, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p.list, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "upper .95"]}))) 

df1 <- df %>% 
  mutate(hist = stringr::str_split_i(rownames(df),"_",1),
         feature = gsub("LUAD_|LUSC_|Other_","", rownames(df)))
  
df1

p <- df1 %>%
  ggplot(aes(x = HR,y = feature, col = hist))+
  geom_point(size = 2)+
  facet_wrap(.~hist)+
  geom_errorbar(aes(xmin=lower_conf, xmax = upper_conf),lwd =1,width = 0.3)+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust= 1))+
  #scale_y_discrete(labels = function(x) str_split_i(x,"_",2))+
  scale_x_log10()+
  theme_bw()+
  geom_vline(xintercept = 1,linetype = "dashed")+
  geom_text(aes(label = paste0("p : ",round(p.value,digits = 3))),
            col = "black",
            vjust = -1,
            inherit.aes = TRUE)+
  scale_color_manual(values = c("LUAD" = "blue", "LUSC" = "orange", "Other" = "gray40", "All" = "black"),name = "Histology")+
  theme(axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10))+
  labs(x = "HR",y = "Histology")

p
```

# patient cluster survival

```{r}
pcl <-  data.frame(immucan_id = rownames(meta_complete)[pat_col_order],
             patient_cluster = as.integer(pat_cl_ord),
             stringsAsFactors = FALSE)

andat <- data.frame(patient_cluster = 1:6,
                    pgroup = c("inflamed", "TAN","Thelper","mild_inflamed","proliferating","cold"))
pcl <- left_join(pcl, andat,by = "patient_cluster")
pcl <- left_join(pcl, traits, "immucan_id")

pcl <- pcl %>%
 ungroup() %>%
  mutate(LUAD_vs_LUSC = ifelse(is.na(LUAD_vs_LUSC),"Other",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma","LUAD",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Squamous cell carcinoma","LUSC",LUAD_vs_LUSC))

survminer::ggforest(coxph( Surv(time_to_event, censoring) ~ pgroup,
                data = pcl %>% filter(LUAD_vs_LUSC == "LUAD")))

```

