---
title: "11_sub-clustering"
author: "Daniel Schulz"
date: "5/16/2022"
output: html_document
---

Here we will subcluster some of the individual cell types to obtain more fine grained phenotypes.
we will sub-cluster the cell types only with cell-type relevant markers.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r}
mount_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/IMC/Panel_2/"
```


```{r, message=FALSE}
library(SingleCellExperiment)
library(imcRtools)
library(cytomapper)
library(tidyverse)
library(dittoSeq)
library(viridis)
library(doParallel)
library(BiocParallel)
library(BiocNeighbors)
library(scran)
library(bluster)
library(scater)
library(cowplot)
library(Rphenoannoy)
```

```{r}
sce <- readRDS(paste0(mount_path,"Rout/sce.rds"))
```


# cell type marker definition

```{r determine markers for clustering}
markers <- list()
# we will not subcluters "uncertain" cells as the merging will be very difficult. we may inspect this at a later time point again

#functional markers for T cells
markers[["Tcells"]] <- rownames(sce)[grepl(pattern = "CD7$|CD8|FOXP3|Bcl2|GATA3|PD1|NKG2A|CXCL13|EOMES|CD137|CD134|Tbet|GITR|Tim3|Ki67",rownames(sce))]
markers[["CD8"]] <- rownames(sce)[grepl(pattern = "CD7$|PD1|Bcl2|GATA3|PD1|NKG2A|CXCL13|EOMES|CD137|CD134|Tbet|GITR|Tim3|Ki67",rownames(sce))]
markers[["CD4"]] <- rownames(sce)[grepl(pattern = "CD7$|PD1|Bcl2|GATA3|PD1|NKG2A|CXCL13|EOMES|CD137|CD134|Tbet|GITR|Tim3|Ki67",rownames(sce))]
markers[["Treg"]] <- rownames(sce)[grepl(pattern = "CD7$|Bcl2|GATA3|PD1|NKG2A|CXCL13|EOMES|CD137|CD134|Tbet|GITR|Tim3|Ki67",rownames(sce))]


# NK markers
markers[["NK"]] <- rownames(sce)[grepl(pattern = "Tbet|Bcl2|GATA3|NKG2A|EOMES|GITR",rownames(sce))]


# B and BnT cell markers
markers[["B"]] <- rownames(sce)[grepl(pattern = "Bcl2|Ki67",rownames(sce))]
markers[["BnT"]] <- rownames(sce)[grepl(pattern = "CD7$|PD1|FOXP3|Bcl2|GATA3|CD8a|LAG3|PD1|NKG2A|CXCL13|EOMES|CD137|CD134|Tbet|GITR|Tim3|Ki67",rownames(sce))]


# myeloid markers
markers[["all_Myeloid"]] <- rownames(sce)[grepl(pattern = "CD204|CD209|CD11c|CD155|Arg1|MMP9|CD73|CD10|Tim3",rownames(sce))]
#markers[["MacCD204"]] <- rownames(sce)[grepl(pattern = "CD11c|Arg1|MMP9|CD73|CD10|Tim3",rownames(sce))]
#markers[["MacCD209"]] <- rownames(sce)[grepl(pattern = "CD11c|Arg1|MMP9|CD73|CD10|Tim3",rownames(sce))]
#markers[["MacCD204_CD209"]] <- rownames(sce)[grepl(pattern = "CD11c|Arg1|MMP9|CD73|CD10|Tim3",rownames(sce))]
# DCs
markers[["DC"]] <- rownames(sce)[grepl(pattern = "CD137|Tim3|CD20$",rownames(sce))]
markers[["mregDC"]] <- rownames(sce)[grepl(pattern = "CD11c|CD137|Tim3",rownames(sce))]


# MDSCs
markers[["all_MDSC"]] <- rownames(sce)[grepl(pattern = "CD11b|CD15$|CD66b|CD10|CXCL8|Arg1|MMP9",rownames(sce))]
#markers[["MDSC"]] <- rownames(sce)[grepl(pattern = "CD11b|CD10|CXCL8|Arg1|MMP9",rownames(sce))]
#markers[["PMN_MDSC"]] <- rownames(sce)[grepl(pattern = "CD11b|CD15$|CD66b|CD10|CXCL8|Arg1|MMP9",rownames(sce))]
#markers[["Neutrophil"]] <- rownames(sce)[grepl(pattern = "CD11b|CD15$|CD66b|CD10|CXCL8|Arg1|MMP9",rownames(sce))]


# stroma markers
markers[["all_stromal"]] <- rownames(sce)[grepl(pattern = "Podoplanin|CD10|MMP9|SMA|CD31|CD146",rownames(sce))]
#markers[["Vasculature"]] <- rownames(sce)[grepl(pattern = "CD31|CD146",rownames(sce))]
#markers[["vCAF"]] <- rownames(sce)[grepl(pattern = "SMA|CD146",rownames(sce))]
#markers[["Fibro"]] <- rownames(sce)[grepl(pattern = "Podoplanin|CD10|MMP9|SMA",rownames(sce))]


# tumor markers
markers[["Tumor"]] <- rownames(sce)[grepl(pattern = "panCK|Ecad|Arg1|EGFR|MMP9|p53|Lamp3|CD73|Bcl2|GATA3|CD155|CD10|CD56|CXCL8|Ki67|Podoplanin|IgG",rownames(sce))]

good_markers <- rownames(sce)[!grepl("DNA|Histone",rownames(sce))]
```


# cell type heatmap
```{r,fig.width=15, fig.height=8}
cur_sce <- sce[,order(sce$celltypes)]
cur_dat <- as_tibble(colData(cur_sce))
ids <- cur_dat %>%
  slice_sample(n = 500,replace = FALSE, by = "celltypes") %>%
  mutate(names = paste(SampleId,ROI,ObjectNumber,sep = "_")) %>%
  pull(names)
cur_sce <- cur_sce[,colnames(cur_sce) %in% ids]
dittoHeatmap(cur_sce,genes = good_markers, assay = "exprs",
            annot.by = c("celltypes"),
            cluster_rows = TRUE,
            show_colnames = FALSE,
            heatmap.colors = colorRampPalette(c("dark blue", "white", "dark red"))(100),
            breaks = seq(-3, 3, length.out = 101))
```

# cluster sweeps

```{r, include=FALSE, eval=FALSE}
# create necessary directory first
if (!dir.exists(paste0(mount_path, "Rout/subclustering"))) {
  dir.create(paste0(mount_path, "Rout/subclustering"))
}

celltypes <- list("CD4","CD8","Treg","all_Myeloid","DC","mregDC","B","BnT","all_MDSC","all_stromal", "NK")
celltypes

lapply(celltypes, function(x){
  if(x == "all_Myeloid"){
    cur_sce <- sce[,sce$celltypes %in% c("MacCD204","MacCD209","MacCD204_CD209")]
  } else if(x == "all_MDSC"){
    cur_sce <- sce[,sce$celltypes %in% c("MDSC","PMN_MDSC","Neutrophil")]
  } else if(x == "all_stromal"){
    cur_sce <- sce[,sce$celltypes %in% c("vCAF","Vasculature","Fibro")]
  } else {
    cur_sce <- sce[,sce$celltypes %in% x]
  }

  mat <- t(assay(cur_sce, "exprs")[markers[[x]],])  
  
  set.seed(12345)
  
  combinations <- clusterSweep(mat, 
                                   BLUSPARAM=SNNGraphParam(),
                                   k=c(30L, 40L, 50L, 60L, 70L, 80L), 
                                   type = c("rank", "jaccard"), 
                                   cluster.fun = "louvain",
                                   BPPARAM = MulticoreParam(RNGseed = 220427,workers = 10))
  
  saveRDS(combinations,paste0(mount_path,"Rout/subclustering/",x,".rds"))
  rm(cur_sce)
  gc()
})

```

# plot silhouette width and cluster purity
```{r}
# create necessary directory first
if (!dir.exists(paste0(mount_path, "Rout/subclustering/plots"))) {
  dir.create(paste0(mount_path, "Rout/subclustering/plots"))
}

files <- list.files(paste0(mount_path,"Rout/subclustering"), full.names = TRUE, pattern = "*.rds")

for (x in files){
  
  combination <- readRDS(x)
  
  cts <- gsub(".rds","",basename(x))
  if(cts == "all_Myeloid"){
    cur_sce <- sce[,sce$celltypes %in% c("MacCD204","MacCD209","MacCD204_CD209")]
  } else if(cts == "all_MDSC"){
    cur_sce <- sce[,sce$celltypes %in% c("MDSC","PMN_MDSC","Neutrophil")]
  } else if(cts == "all_stromal"){
    cur_sce <- sce[,sce$celltypes %in% c("vCAF","Vasculature","Fibro")]
  } else {
    cur_sce <- sce[,sce$celltypes %in% cts]
  }
  
  mat <- t(assay(cur_sce, "exprs")[markers[[cts]],])  
  
  sil <- vapply(as.list(combination$clusters), 
                function(x) mean(approxSilhouette(mat, x)$width), 
                0)
  pur <- vapply(as.list(combination$clusters), 
                function(x) mean(neighborPurity(mat, x)$purity), 
                0)
  
  sil_plot <- ggplot(data.frame(method = names(sil),
                                sil = sil)) +
    geom_point(aes(method, sil)) +
    theme_classic(base_size = 15) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab("Cluster parameter combination") +
    ylab("Average silhouette width")+
    ggtitle(cts)
  
  ggsave(filename = paste0(mount_path,"Rout/subclustering/plots/",cts,"_sil",".png"),sil_plot)

  pur_plot <- ggplot(data.frame(method = names(pur),
                                pur = pur)) +
    geom_point(aes(method, pur)) +
    theme_classic(base_size = 15) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab("Cluster parameter combination") +
    ylab("Average neighborhood purity")+
    ggtitle(cts)
  
  ggsave(filename = paste0(mount_path,"Rout/subclustering/plots/",cts,"_pure",".png"),pur_plot)
  
}
```

# parameter selection
based on the images saved above we will select max silhouette width and max purity.
```{r}
celltypes <- c("CD4","CD8","Treg","all_Myeloid","DC","mregDC","B","BnT","all_MDSC","all_stromal", "NK")
k <- c(70,70,80,80,60,70,80,80,50,60,60)
type <- c("rank","rank","rank","rank","rank","rank","jaccard","rank","rank","rank","rank")
param <- data.frame(celltype = celltypes,
                    k = k,
                    type = type)

# clustering
lapply(celltypes, function(x){

  if(x == "all_Myeloid"){
    cur_sce <- sce[,sce$celltypes %in% c("MacCD204","MacCD209","MacCD204_CD209")]
  } else if(x == "all_MDSC"){
    cur_sce <- sce[,sce$celltypes %in% c("MDSC","PMN_MDSC","Neutrophil")]
  } else if(x == "all_stromal"){
    cur_sce <- sce[,sce$celltypes %in% c("vCAF","Vasculature","Fibro")]
  } else {
    cur_sce <- sce[,sce$celltypes %in% x]
  }
  
  set.seed(220620)
  clusters <- clusterCells(cur_sce[markers[[x]],], 
                           assay.type = "exprs", 
                           BLUSPARAM = SNNGraphParam(k=param[param$celltype == x,"k"], 
                                                     cluster.fun = "louvain",
                                                     type = param[param$celltype == x,"type"],
                                                     BPPARAM = MulticoreParam(workers = 12)))


  cur_sce$cluster <- paste0(x,"_",clusters)
  saveRDS(cur_sce,paste0(mount_path,"Rout/subclustering/",x,"_sce.rds"))
  rm(cur_sce)
  gc()

})
```

# heatmaps scaled {.tabset}

```{r,results='asis',fig.height=4, fig.width=20}

# function to generate the header for the plotting with tabset
catHeader <- function(text = "", level = 2) {
    cat(paste0("\n\n", 
               paste(rep("#", level), collapse = ""), 
               " ", text, "\n"))
}

sces <- list.files(paste0(mount_path,"Rout/subclustering/"), full.names = TRUE, pattern= "*._sce.rds")

p.list <- list()
for (i in sces) {
  x <- readRDS(i)
  set.seed(1235)
  
  cur_sce <- x[,sample(colnames(x),ncol(x)/10, replace = FALSE)]
  name <- gsub("_sce.rds","",basename(i))
  
  
  p1 <- dittoHeatmap(cur_sce,genes = markers[[name]], assay = "exprs",
             annot.by = c("cluster", "celltypes"),
             cluster_rows = TRUE,
             show_colnames = FALSE,
             heatmap.colors = colorRampPalette(c("dark blue", "white", "dark red"))(100),
             breaks = seq(-3, 3, length.out = 101))

  
  p.list[[name]] <- (p1$gtable)

  p.list[[name]] <- list(p.list[[name]],name)
  
}

for(i in seq_along(p.list)){
    tmp <- p.list[[i]]
    # As you want to use tabset level here has to be lower than 
    # parent level (ie, parent is 1, so here you have to use 2)
    catHeader(tmp[[2]], 2)
    plot(tmp[[1]])
}

```

# {}

# heatmaps exprs {.tabset}

```{r,results='asis',fig.height=4, fig.width=20}

p.list <- list()
for (i in sces) {
  x <- readRDS(i)
  set.seed(1235)
  
  cur_sce <- x[,sample(colnames(x),ncol(x)/10, replace = FALSE)]
  name <- gsub("_sce.rds","",basename(i))
  
  p1 <- dittoHeatmap(cur_sce,genes = markers[[name]], assay = "exprs",
            annot.by = c("cluster", "celltypes"),
            cluster_rows = TRUE,
            show_colnames = FALSE,
            scale = "none",
            heatmap.colors = viridis(100))

  
  p.list[[name]] <- (p1$gtable)

  p.list[[name]] <- list(p.list[[name]],name)
  
}

for(i in seq_along(p.list)){
    tmp <- p.list[[i]]
    # As you want to use tabset level here has to be lower than 
    # parent level (ie, parent is 1, so here you have to use 2)
    catHeader(tmp[[2]], 2)
    plot(tmp[[1]])
}

```

# {}

__we will not sub-cluster B cells, DCs and mregDCs. there are only very few markers to sub-cluster those cell types and the clustering identifies too many clusters for only those few markers__

# B cell sub types 

```{r}
library(mclust)
library(mixtools)
B_sce <- readRDS(paste0(mount_path,"Rout/subclustering/B_sce.rds"))
B_sce$cluster <- B_sce$celltypes

# Ki67
Ki67 <- assay(B_sce,"exprs")["Ki67",]
fit <- Mclust(Ki67, G = 2)

# get cut point
post <- fit$z[, 1]
# Approximate cut point: where posteriors are equal (0.5)
idx <- which.min(abs(post - 0.5))
cut_point_mclust <- Ki67[idx]
cut_point_mclust

fit2 <- mixtools::normalmixEM(Ki67,k = 2)

# get cut point
post <- fit2$posterior[, 1]
# Approximate cut point: where posteriors are equal (0.5)
idx <- which.min(abs(post - 0.5))
cut_point_mixtools <- Ki67[idx]
cut_point_mixtools

p1 <- ggplot(data.frame(Ki67 = assay(B_sce, "exprs")["Ki67",],
                  positivity = fit$classification)) +
  geom_histogram(aes(x = Ki67, fill = positivity, group = positivity), binwidth = 0.01) +
  theme_minimal()+
  geom_vline(xintercept = mean(cut_point_mclust, cut_point_mixtools), color = "red")
p2 <- ggplot(data.frame(Ki67 = assay(B_sce, "exprs")["Ki67",],
                  positivity = apply(fit2$posterior, 1, which.max))) +
  geom_histogram(aes(x = Ki67, fill = positivity, group = positivity), binwidth = 0.01) +
  theme_minimal()+
  geom_vline(xintercept = mean(cut_point_mclust, cut_point_mixtools), color = "red")

p1 
p2

# Bcl2
Bcl2 <- assay(B_sce,"exprs")["Bcl2",]

fit <- Mclust(Bcl2, G = 2)
# get cut point
post <- fit$z[, 1]
# Approximate cut point: where posteriors are equal (0.5)
idx <- which.min(abs(post - 0.5))
cut_point_Bcl2_mclust <- Bcl2[idx]
cut_point_Bcl2_mclust

fit2 <- mixtools::normalmixEM(Bcl2,k = 2)
# get cut point
post <- fit2$posterior[, 1]
# Approximate cut point: where posteriors are equal (0.5)
idx <- which.min(abs(post - 0.5))
cut_point_Bcl2_mixtools <- Bcl2[idx]
cut_point_Bcl2_mixtools


p1 <- ggplot(data.frame(Bcl2 = assay(B_sce, "exprs")["Bcl2",],
                  positivity = fit$classification)) +
  geom_histogram(aes(x = Bcl2, fill = positivity, group = positivity), binwidth = 0.01) +
  theme_minimal()+
  geom_vline(xintercept = mean(cut_point_Bcl2_mclust, cut_point_Bcl2_mixtools), color = "red")
p2 <- ggplot(data.frame(Bcl2 = assay(B_sce, "exprs")["Bcl2",],
                  positivity = apply(fit2$posterior, 1, which.max))) +
  geom_histogram(aes(x = Bcl2, fill = positivity, group = positivity), binwidth = 0.01) +
  theme_minimal()+
  geom_vline(xintercept = mean(cut_point_Bcl2_mclust, cut_point_Bcl2_mixtools), color = "red")

p1 
p2

Bcl2_cutoff <- mean(cut_point_Bcl2_mclust,cut_point_Bcl2_mixtools)
Ki67_cutoff <-  mean(cut_point_mclust, cut_point_mixtools)

# we will choose a more conservative cut off of 1 then the mixture models propose
Ki67_cutoff <- 1
Bcl2_cutoff <- 0.75

B_sce[,assay(B_sce,"exprs")["Bcl2",] < Bcl2_cutoff  & assay(B_sce,"exprs")["Ki67",] > Ki67_cutoff]$cluster <- "B_Ki67"
B_sce[,assay(B_sce,"exprs")["Bcl2",] > Bcl2_cutoff & assay(B_sce,"exprs")["Ki67",] > Ki67_cutoff]$cluster <- "B_Ki67_Bcl2"
B_sce[,assay(B_sce,"exprs")["Bcl2",] > Bcl2_cutoff & assay(B_sce,"exprs")["Ki67",] < Ki67_cutoff]$cluster <- "B_Bcl2"
table(B_sce$cluster)
saveRDS(B_sce,paste0(mount_path,"Rout/subclustering/B_sce.rds"))
```

# DC sub types

```{r}
DC_sce <- readRDS(paste0(mount_path,"Rout/subclustering/DC_sce.rds"))
DC_sce$cluster <- DC_sce$celltypes

# CD20
CD20 <- assay(DC_sce,"exprs")["CD20",]

fit <- Mclust(CD20, G = 2)
# get cut point
post <- fit$z[, 1]
# Approximate cut point: where posteriors are equal (0.5)
idx <- which.min(abs(post - 0.5))
cut_point_mclust <- CD20[idx]
cut_point_mclust

fit2 <- mixtools::normalmixEM(CD20,k = 2)
# get cut point
post <- fit2$posterior[, 1]
# Approximate cut point: where posteriors are equal (0.5)
idx <- which.min(abs(post - 0.5))
cut_point_mixtools <- CD20[idx]
cut_point_mixtools

p1 <- ggplot(data.frame(CD20 = assay(DC_sce, "exprs")["CD20",],
                  positivity = fit$classification)) +
  geom_histogram(aes(x = CD20, fill = positivity, group = positivity), binwidth = 0.01) +
  theme_minimal()
p2 <- ggplot(data.frame(CD20 = assay(DC_sce, "exprs")["CD20",],
                  positivity = apply(fit2$posterior, 1, which.max))) +
  geom_histogram(aes(x = CD20, fill = positivity, group = positivity), binwidth = 0.01) +
  theme_minimal()

p1 
p2

# Tim3
Tim3 <- assay(DC_sce,"exprs")["Tim3",]

fit <- Mclust(Tim3, G = 2)
# get cut point
post <- fit$z[, 1]
# Approximate cut point: where posteriors are equal (0.5)
idx <- which.min(abs(post - 0.5))
cut_point_Tim3_mclust <- Tim3[idx]
cut_point_Tim3_mclust

fit2 <- mixtools::normalmixEM(Tim3,k = 2)
# get cut point
post <- fit2$posterior[, 1]
# Approximate cut point: where posteriors are equal (0.5)
idx <- which.min(abs(post - 0.5))
cut_point_Tim3_mixtools <- Tim3[idx]
cut_point_Tim3_mixtools

p1 <- ggplot(data.frame(Tim3 = assay(DC_sce, "exprs")["Tim3",],
                  positivity = fit$classification)) +
  geom_histogram(aes(x = Tim3, fill = positivity, group = positivity), binwidth = 0.01) +
  theme_minimal()
p2 <- ggplot(data.frame(Tim3 = assay(DC_sce, "exprs")["Tim3",],
                  positivity = apply(fit2$posterior, 1, which.max))) +
  geom_histogram(aes(x = Tim3, fill = positivity, group = positivity), binwidth = 0.01) +
  theme_minimal()

p1 
p2

CD20_cutoff <-  mean(cut_point_mclust, cut_point_mixtools)
Tim3_cutoff <- mean(cut_point_Tim3_mclust,cut_point_Tim3_mixtools)

# we will choose a more conservative cut off of 1 then the mixture models propose
CD20_cutoff <- 0.5
Tim3_cutoff <- 1

# we will choose a more conservative cut off of 1 then the mixture models propose
DC_sce[,assay(DC_sce,"exprs")["Tim3",] > Tim3_cutoff & assay(DC_sce,"exprs")["CD20",] < CD20_cutoff]$cluster <- "DC_Tim3"
DC_sce[,assay(DC_sce,"exprs")["Tim3",] > Tim3_cutoff & assay(DC_sce,"exprs")["CD20",] > CD20_cutoff]$cluster <- "DC_CD20_Tim3"
DC_sce[,assay(DC_sce,"exprs")["Tim3",] < Tim3_cutoff & assay(DC_sce,"exprs")["CD20",] > CD20_cutoff]$cluster <- "DC_CD20"
table(DC_sce$cluster)

# save file
saveRDS(DC_sce,paste0(mount_path,"Rout/subclustering/DC_sce.rds"))
```

# mregDC sub types

```{r}
library(mclust)
library(mixtools)
mregDC_sce <- readRDS(paste0(mount_path,"Rout/subclustering/mregDC_sce.rds"))
mregDC_sce$cluster <- mregDC_sce$celltypes

# CD11c
CD11c <- assay(mregDC_sce,"exprs")["CD11c",]

fit <- Mclust(CD11c, G = 2)
# get cut point
post <- fit$z[, 1]
# Approximate cut point: where posteriors are equal (0.5)
idx <- which.min(abs(post - 0.5))
cut_point_CD11c_mclust <- CD11c[idx]
cut_point_CD11c_mclust

fit2 <- mixtools::normalmixEM(CD11c,k = 2)
# get cut point
post <- fit2$posterior[, 1]
# Approximate cut point: where posteriors are equal (0.5)
idx <- which.min(abs(post - 0.5))
cut_point_CD11c_mixtools <- CD11c[idx]
cut_point_CD11c_mixtools


p1 <- ggplot(data.frame(CD11c = assay(mregDC_sce, "exprs")["CD11c",],
                  positivity = fit$classification)) +
  geom_histogram(aes(x = CD11c, fill = positivity, group = positivity), binwidth = 0.01) +
  theme_minimal()
p2 <- ggplot(data.frame(CD11c = assay(mregDC_sce, "exprs")["CD11c",],
                  positivity = apply(fit2$posterior, 1, which.max))) +
  geom_histogram(aes(x = CD11c, fill = positivity, group = positivity), binwidth = 0.01) +
  theme_minimal()

p1 
p2

# Tim3
Tim3 <- assay(mregDC_sce,"exprs")["Tim3",]

fit <- Mclust(Tim3, G = 2)
# get cut point
post <- fit$z[, 1]
# Approximate cut point: where posteriors are equal (0.5)
idx <- which.min(abs(post - 0.5))
cut_point_Tim3_mclust <- Tim3[idx]
cut_point_Tim3_mclust

fit2 <- mixtools::normalmixEM(Tim3,k = 2)
# get cut point
post <- fit2$posterior[, 1]
# Approximate cut point: where posteriors are equal (0.5)
idx <- which.min(abs(post - 0.5))
cut_point_Tim3_mixtools <- Tim3[idx]
cut_point_Tim3_mixtools

p1 <- ggplot(data.frame(Tim3 = assay(mregDC_sce, "exprs")["Tim3",],
                  positivity = fit$classification)) +
  geom_histogram(aes(x = Tim3, fill = positivity, group = positivity), binwidth = 0.01) +
  theme_minimal()
p2 <- ggplot(data.frame(Tim3 = assay(mregDC_sce, "exprs")["Tim3",],
                  positivity = apply(fit2$posterior, 1, which.max))) +
  geom_histogram(aes(x = Tim3, fill = positivity, group = positivity), binwidth = 0.01) +
  theme_minimal()

p1 
p2


CD11c_cutoff <-  mean(cut_point_CD11c_mclust, cut_point_CD11c_mixtools)
Tim3_cutoff <- mean(cut_point_Tim3_mclust,cut_point_Tim3_mixtools)

# we will choose a more conservative cut off of 1 then the mixture models propose
CD11c_cutoff <- 0.8
Tim3_cutoff <- 1



# we will choose a more conservative cut off of 1 then the mixture models propose
mregDC_sce[,assay(mregDC_sce,"exprs")["Tim3",] > Tim3_cutoff & assay(mregDC_sce,"exprs")["CD11c",] < CD11c_cutoff]$cluster <- "mregDC_Tim3"
mregDC_sce[,assay(mregDC_sce,"exprs")["Tim3",] > Tim3_cutoff & assay(mregDC_sce,"exprs")["CD11c",] > CD11c_cutoff]$cluster <- "mregDC_CD11c_Tim3"
mregDC_sce[,assay(mregDC_sce,"exprs")["Tim3",] < Tim3_cutoff & assay(mregDC_sce,"exprs")["CD11c",] > CD11c_cutoff]$cluster <- "mregDC_CD11c"
table(mregDC_sce$cluster)

# save file
saveRDS(mregDC_sce,paste0(mount_path,"Rout/subclustering/mregDC_sce.rds"))
```

# meta clustering CD4 and CD8
we will meta cluster CD4, CD8, myeloid cells, BnT and NK.
for Treg and MDSC I am not convinced they can be metaclustered


```{r}
cell_types <- c("8","9")
names(cell_types) <- c("CD4","CD8")

sces <- list.files(paste0(mount_path,"Rout/subclustering"),full.names = TRUE, pattern = "*._sce.rds")
for (i in c("CD4", "CD8")) {
  cur_sce <- readRDS(sces[grepl(i,sces)])
  name <- colnames(colData(cur_sce))[grepl("^cluster",colnames(colData(cur_sce)))]
  mean_sce <- aggregateAcrossCells(cur_sce,ids = cur_sce[[name]],statistics = "mean")
  assay(mean_sce,"exprs") <- asinh(counts(mean_sce))

  marker_names <- cell_types
  dittoHeatmap(mean_sce,genes = markers[[i]], assay = "exprs",
               annot.by = c( "cluster"),
               cluster_rows = TRUE,
               show_colnames = TRUE,
               cluster_cols = TRUE,
               clustering_method = "ward.D2",
               cutree_cols = cell_types[i],
               heatmap.colors = colorRampPalette(c("dark blue", "white", "dark red"))(100),
               breaks = seq(-3, 3, length.out = 101))
  
  dittoHeatmap(mean_sce,genes = markers[[i]], assay = "exprs",
               annot.by = c( "cluster"),
               cluster_rows = TRUE,
               show_colnames = TRUE,
               cluster_cols = TRUE,
               scale = "none",
               clustering_method = "ward.D2",
               cutree_cols = cell_types[i],
               heatmap.colors = viridis(100))
 
}
```

## assign meta cluster names
```{r}
CD4_vec <- c("CD4_PD1_CXCL13","CD4_activated","CD4_BCL2","CD4_CD7_BCL2","CD4_Ki67","CD4_Tim3","CD4_EOMES","CD4","CD4","CD4")
names(CD4_vec) <- paste0("CD4_",c(2,1,7,8,3,4,10,6,5,9))

CD8_vec <- c("CD8_BCL2","CD8","CD8","CD8_EOMES","CD8_TIM3","CD8_Ki67","CD8_exh","CD8_PD1_CXCL13","CD8_NKG2A","CD8_GATA3_BCL2")
names(CD8_vec) <- paste0("CD8_",c(10,5,7,9,1,2,3,4,6,8))
```


# cell types minor definition
we will substitute CD4, CD8 and will manually gate CD137+ Tregs with the subclusters
we will also fix a typo
```{r}
CD8_sce <- readRDS(paste0(mount_path,"Rout/subclustering/CD8_sce.rds"))
CD8_sce$meta_cluster <- CD8_vec[CD8_sce$cluster]

CD4_sce <- readRDS(paste0(mount_path,"Rout/subclustering/CD4_sce.rds"))
CD4_sce$meta_cluster <- CD4_vec[CD4_sce$cluster]


dittoSeq::dittoRidgePlot(sce[,sce$celltypes == "Treg"],"CD137", group.by = "celltypes",assay  = "exprs")

Treg_sce <- readRDS(paste0(mount_path,"Rout/subclustering/Treg_sce.rds"))

CD137 <- assay(Treg_sce,"exprs")["CD137",]

fit <- Mclust(CD137, G = 2)
# get cut point
post <- fit$z[, 1]
# Approximate cut point: where posteriors are equal (0.5)
idx <- which.min(abs(post - 0.5))
cut_point_CD137_mclust <- CD137[idx]
cut_point_CD137_mclust

fit2 <- mixtools::normalmixEM(CD137,k = 2)
# get cut point
post <- fit2$posterior[, 1]
# Approximate cut point: where posteriors are equal (0.5)
idx <- which.min(abs(post - 0.5))
cut_point_CD137_mixtools <- CD137[idx]
cut_point_CD137_mixtools

p1 <- ggplot(data.frame(CD137 = assay(Treg_sce, "exprs")["CD137",],
                  positivity = fit$classification)) +
  geom_histogram(aes(x = CD137, fill = positivity, group = positivity), binwidth = 0.01) +
  theme_minimal()
p2 <- ggplot(data.frame(CD137 = assay(Treg_sce, "exprs")["CD137",],
                  positivity = apply(fit2$posterior, 1, which.max))) +
  geom_histogram(aes(x = CD137, fill = positivity, group = positivity), binwidth = 0.01) +
  theme_minimal()

p1 
p2

# define cut off more conservative than gaussian mixture model
sce$CD137_pos <- assay(sce, "exprs")["CD137",] > 0.5

# merge CD8 and CD4 meta_cluster info
sce$meta_cluster <- sce$celltypes

#CD8
sce[,colnames(CD8_sce)]$meta_cluster <- CD8_sce$meta_cluster
#CD4
sce[,colnames(CD4_sce)]$meta_cluster <- CD4_sce$meta_cluster


x <- as.data.frame(colData(sce))

x <- x %>%
  mutate(celltypes_minor = celltypes,
         celltypes_minor = ifelse(celltypes == "Treg" & CD137_pos == TRUE,"Treg_CD137",celltypes_minor),
         celltypes_minor = ifelse(celltypes == "CD4",meta_cluster,celltypes_minor),
         celltypes_minor = ifelse(celltypes == "CD8",meta_cluster,celltypes_minor))
unique(x$celltypes_minor)

colData(sce) <- DataFrame(x)
```


# merge data

## add clustering to sce
```{r}
sces  <- list.files(paste0(mount_path,"Rout/subclustering"),
                           full.names = TRUE,
                           pattern = "*sce.rds")

sce$clusters <- sce$celltypes

for (i in sces){
  x <- readRDS(i)
  cur_var <- colnames(colData(x))[grep("cluster",colnames(colData(x)))]
  sce[,colnames(x)]$clusters <- x[[cur_var]]
}

unique(sce$clusters)

```

## add meta clustering to sce
```{r}
# add info to sce
sce$meta_cluster <- sce$celltypes
sce[,sce$celltypes == "Treg"]$meta_cluster <- Treg_vec[sce[,sce$celltypes == "Treg"]$clusters]
sce[,sce$celltypes == "NK"]$meta_cluster <- NK_vec[sce[,sce$celltypes == "NK"]$clusters]
sce[,sce$celltypes == "CD4"]$meta_cluster <- CD4_vec[sce[,sce$celltypes == "CD4"]$clusters]
sce[,sce$celltypes == "CD8"]$meta_cluster <- CD8_vec[sce[,sce$celltypes == "CD8"]$clusters]
sce[,sce$celltypes %in% c("MacCD204","MacCD209","MacCD204_CD209")]$meta_cluster <- myel_vec[sce[,sce$celltypes %in% c("MacCD204","MacCD209","MacCD204_CD209")]$clusters]
sce[,sce$celltypes == "BnT"]$meta_cluster <- BnT_vec[sce[,sce$celltypes == "BnT"]$clusters]
sce[,sce$celltypes == "Tumor"]$meta_cluster <- sce[,sce$celltypes == "Tumor"]$clusters 
sce[,sce$celltypes == "B"]$meta_cluster <- sce[,sce$celltypes == "B"]$clusters 

```

```{r, fig.height=20, fig.width=25}
set.seed(7496)
dittoHeatmap(sce[,sample(colnames(sce),20000,replace = FALSE)],
             genes = rownames(sce)[!grepl("DNA|Histone",rownames(sce))],
             assay = "exprs",
             annot.by = c("clusters"),
             cluster_rows = TRUE,
             scale = "none",
             show_colnames = FALSE,
             heatmap.colors = viridis(100))
```

```{r, fig.height=20, fig.width=40}
set.seed(7496)
dittoHeatmap(sce[,sample(colnames(sce),20000,replace = FALSE)],
             genes = rownames(sce)[!grepl("DNA|Histone",rownames(sce))],
             assay = "exprs",
             annot.by = c("meta_cluster"),
             cluster_rows = TRUE,
             show_colnames = FALSE,
             heatmap.colors = colorRampPalette(c("dark blue", "white", "dark red"))(100),
             breaks = seq(-3, 3, length.out = 101))
```

# UMAP
```{r, fig.height=15, fig.width=20}
sce <- runUMAP(sce,
               exprs_values = "exprs", 
               subset_row = rownames(sce)[!grepl("DNA|Histone",rownames(sce))],
               name = "UMAP",
               external_neighbors = TRUE, BNPARAM = AnnoyParam(),
               BPPARAM = MulticoreParam(workers = 12))

dittoDimPlot(sce,var = "celltypes",reduction.use = "UMAP",do.label = TRUE)+
  scale_color_manual(values = metadata(sce)$colour_vectors$cell_types)
```


# mean meta-cluster expression

```{r, fig.width=25, fig.height=10}
mean_sce <- aggregateAcrossCells(sce,ids = sce$celltypes_minor,statistics = "mean")
assay(mean_sce,"exprs") <- asinh(counts(mean_sce))

dittoHeatmap(mean_sce,
             genes = rownames(sce)[!grepl("DNA|Histone",rownames(sce))],
             assay = "exprs",
             #annot.by = c("meta_cluster"),
             cluster_rows = TRUE,
             cluster_cols = TRUE,
             show_colnames = TRUE,
             heatmap.colors = colorRampPalette(c("dark blue", "white", "dark red"))(100),
             breaks = seq(-4, 4, length.out = 101))
```




```{r}
table(sce$celltypes_minor)/ncol(sce)
```

# define colors

```{r, fig.height=10, fig.width=10}
cell_types <- sort(unique(sce2$celltypes_minor))

names(cell_types) <- cell_types
cell_types[grepl("CD8",cell_types)] <- c("mediumorchid2","darkorchid1","darkorchid2","darkorchid3","darkorchid4","mediumorchid", "mediumorchid3","mediumorchid4","orchid4")
cell_types[grepl("CD4",cell_types)] <- c("deeppink","deeppink1","deeppink2","deeppink3","magenta","magenta1","magenta2","magenta3")
cell_types[grepl("Mac",cell_types)] <- c("green","yellowgreen","greenyellow")
cell_types[grepl("DC",cell_types)] <- c("forestgreen","darkgreen")
cell_types[grepl("Tumor",cell_types)] <- c("sienna4")
cell_types["B"] <- "red3"
cell_types[grepl("BnT",cell_types)] <- c("indianred","maroon2")
cell_types["Igg"] <- "goldenrod2"
cell_types[grepl("Treg",cell_types)] <- c("hotpink","hotpink2")
cell_types[grepl("MDSC|Neutrophil",cell_types)] <- c("aquamarine4","blue1","deepskyblue")
cell_types["NK"] <- "yellow"
cell_types["vCAF"] <- "tomato"
cell_types["Fibro"] <- "chocolate1"
cell_types["SMA"] <- "goldenrod2"
cell_types["Vasculature"] <- "brown2"
cell_types["undefined"] <- "gray"

cell_types
pie(rep(1,length(cell_types)),col=cell_types, labels = names(cell_types))

metadata(sce2)$colour_vectors$celltypes_minor <- cell_types
```


# save results
```{r}
saveRDS(sce,paste0(mount_path,"Rout/sce.rds"))
```

