---
title: "Figure_1_supp"
author: "Daniel Schulz"
date: "2023-12-22"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggplot2)
library(ggnewscale)
```

```{r}
mount_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/"
out_folder <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/Figure_1_supp/"

if (!dir.exists(out_folder)) {
  dir.create(out_folder)
}
```

# S1A Labelled cells heatmap

## laod data
```{r}
sce <- readRDS("/mnt/immucan_volume/processed_data/Panel_2/cell_classification/classifier_v7/sce_selected.rds")
```

## heatmap
```{r heatmap-V1, fig.width=20, fig.height=15}
lab_sce <- sce[,sce$cell_labels != "unlabelled"]
agr_sce <- aggregateAcrossCells(lab_sce, 
                                ids = colData(lab_sce)[,c("sample_id", "cell_labels")], 
                                statistics = "mean")
assay(agr_sce, "exprs") <- asinh(counts(agr_sce))
colnames(agr_sce) <- paste0(agr_sce$sample_id, "_", 
                            agr_sce$cell_labels)

# Define markers that were used for gating
cur_markers <- c("Ecad", "CarbonicAnhydrase", "HLADR", 
                 "CD11c", "CD163", "CD303", "CD15", "MPO", "CD38",
                 "SMA", "PDGFRb", "CD20", "CD7", "CD3", "CD8a", "CD4", "FOXP3")


png(file =  paste0(out_folder,"S1A_labelled_cell_heatmap.png"),
    width = 900,
    height = 800,
    units = "px")
dittoSeq::dittoHeatmap(agr_sce,
                       genes = rownames(lab_sce)[!grepl("DNA|Histone", rownames(lab_sce))],
                       assay = "exprs",
                       annot.by = c( "cell_labels", "Indication"), 
                       cluster_rows = FALSE,
                       order.by = "cell_labels",
                       annotation_colors = list(cell_labels = metadata(sce)$colour_vectors$cell_types),
                       heatmap.colors = colorRampPalette(c("dark blue", "white", "dark red"))(100),
                       breaks = seq(-3, 3, length.out = 101))
dev.off()
```

# S1B classifier P2 results

## load data

```{r}
# load classifier
rfit <- readRDS("/mnt/immucan_volume/processed_data/Panel_2/cell_classification/classifier_v7/classifier_xgboost_weighted_filtered_igg_mregDC.rds")
```

## generate train and test split
note: this recycles code from the repo [here](https://github.com/ImmucanWP7/IMMUcan_IMC_panel_2_cell_type_classifier/tree/main)
```{r exclude-DCs}
lab_sce <- sce[,sce$cell_labels != "unlabelled"]
unlab_sce <- sce[,sce$cell_labels == "unlabelled"]

# We will filter out IgG cell labelled that are under 1 count
lab_sce <- lab_sce[,!((t(assay(lab_sce, "exprs"))[,"IgG"] < 1) & (lab_sce$cell_labels == "Igg"))]
# We will also relabel mregDC labelled cells that have high panCK or Ecad expression to tumor cells
lab_sce[,((t(assay(lab_sce, "exprs"))[,"panCK"] > 0.5) & (lab_sce$cell_labels == "mregDC")) | ((t(assay(lab_sce, "exprs"))[,"Ecad"] > 0.8) & (lab_sce$cell_labels == "mregDC"))]$cell_labels <- "Tumor"

table(lab_sce$cell_labels)
```

Next, we will first split the labelled data into training and test (validation) data at a ratio of 80/20 using images as grouping level.


```{r split-data-images-DCs}
library(caret)
set.seed(2928)
trainIndex <- groupKFold(group = factor(lab_sce$sample_id), k = 5)
train_sce <- lab_sce[,trainIndex$Fold5]
test_sce <- lab_sce[,-trainIndex$Fold5]

# cell type fraction
table(train_sce$cell_labels)/length(train_sce$cell_labels)
table(test_sce$cell_labels)/length(test_sce$cell_labels)

table(train_sce$BatchId)/length(train_sce$cell_labels)
table(test_sce$BatchId)/length(test_sce$cell_labels)

table(train_sce$Indication)/length(train_sce$cell_labels)
table(test_sce$Indication)/length(test_sce$cell_labels)
```

## test classifier performance
```{r}
# cell identifier
test_sce$cell_id <- paste0(test_sce$sample_id,"_",test_sce$ObjectNumber)

# extract expression matrix
cur_mat <- t(assay(test_sce, "exprs")[!grepl("DNA|Histone", rownames(test_sce)),])
# create dummy varaible for indication and batch
dummies <- dummyVars(SampleId ~ Indication, data = colData(test_sce))
all_dummies <- predict(dummies, newdata = colData(test_sce))
# concatenate with expression data
X_test <- cbind(cur_mat, all_dummies)

library(xgboost)
X_test <- xgb.DMatrix(data = X_test)
```

## Visualize predicted results
```{r, fig.height=8, fig.width=12}
#define some variables
num_classes <- length(unique(train_sce$cell_labels))
label_mapping <- levels(as.factor(test_sce$cell_labels))
names(label_mapping) <- as.character(0:(length(label_mapping) - 1))


prob_mat <- as.vector(predict(rfit, newdata = X_test))
prob_mat <- matrix(prob_mat, ncol = num_classes, byrow = TRUE)
y_pred <- as.factor(label_mapping[apply(prob_mat, 1, which.max)])
y_true <- as.factor(test_sce$cell_labels)

cm <- confusionMatrix(y_pred, y_true)

cm$table

p <- data.frame(cm$byClass) %>%
  mutate(class = sub("Class: ", "", rownames(cm$byClass))) %>%
  ggplot() +
  geom_point(aes(1 - Specificity, Sensitivity,
                 size = Detection.Rate,
                 fill = class),
             shape = 21) +
  scale_fill_manual(values = metadata(sce)$colour_vectors$cell_types) +
  theme_bw() +
  theme(legend.text = element_text(size = 12),
        axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.y = element_text(size = 12),
        axis.title.x = element_text(size = 12))+
  scale_size(range = c(1,12))+
  ylab("Sensitivity (TPR)") +
  xlab("1 - Specificity (FPR)")+ guides(fill = guide_legend(override.aes = list(size = 4)))

p
ggsave(plot = p,
       filename = "S1B_TPR_FPR_plot.pdf",
       device = "pdf",
       path = out_folder,
       width = 7,
       height = 5)
```


# S1C IF IMC correlation
## load data
```{r}
dens_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/densities_dat.rds"))

# remove columns that are positivity for each marker
dens_dat <- dens_dat[,!grepl("all", colnames(dens_dat))]
```

## correlation plot
```{r, fig.height=18, fig.width=16}
# correlation
cor <- psych::corr.test(dens_dat[,grepl("_IMC",colnames(dens_dat))],dens_dat[,grepl("__IF",colnames(dens_dat))],  method = "pearson",adjust = "BH")

# format correlation matrix
cur_dat <- as.data.frame(cor$r)
cur_dat$variable <- rownames(cur_dat)
dat_long <- reshape2::melt(cur_dat,id.vars="variable")
colnames(dat_long) <- c("cell_types_IMC","cell_types_mIF","correlation")

#get p-value matrix
p_dat <- as_tibble(cor$p)
p_dat$variable <- rownames(cur_dat)
pdat_long <- reshape2::melt(p_dat,id.vars="variable")
colnames(pdat_long) <- c("cell_types_IMC","cell_types_mIF","p_adj")

dat_long$p_adj <- pdat_long$p_adj

dat_long <- dat_long %>%
  mutate(sig = ifelse(p_adj <= 0.001 & p_adj > 0.0001,0.001,p_adj))

dat_long <- dat_long %>%
  mutate(sig = case_when(p_adj <= 0.0001 ~ "< 0.0001",
                         p_adj <= 0.001 & p_adj > 0.0001 ~ "< 0.001",
                         p_adj <= 0.01 & p_adj > 0.001 ~ "< 0.01",
                         p_adj <= 0.1 & p_adj > 0.01 ~ "< 0.1",
                         p_adj >0.1 ~ "ns"))


# plot
cor_filter <- 0.4
dat_long <- dat_long %>%
   mutate(cell_types_IMC = gsub("_density","",cell_types_IMC),
          cell_types_mIF = gsub("_density","",cell_types_mIF))


p <-  dat_long %>%
   filter(abs(correlation) > cor_filter) %>%
   ggplot(aes(x=factor(cell_types_IMC, levels=sort(unique(dat_long$cell_types_IMC))),y = factor(cell_types_mIF, levels=sort(unique(dat_long$cell_types_mIF))),col = correlation, size= -log10(p_adj)))+
    geom_point()+
    theme_bw()+
    scale_color_gradient2(low =  "blue",mid = "white",high = "red", name = expression("Pearson's" ~ rho))+
    #theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
    #      panel.background = element_blank(), axis.line = element_line(colour = "black"))+
    #theme_bw()+
    #scale_color_gradient(low = "white",   high = "red")+
    scale_size(range = c(1,3))+
    new_scale_colour() +
    geom_point(data = dat_long %>% filter(sig <0.01, abs(correlation) > cor_filter),shape = 1,aes(size = -log10(p_adj), col = sig), stroke =1)+
    scale_color_manual(values = c("TRUE" = "black","FALSE" = "white"), name = "FDR < 0.1")+
    theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 6),
          axis.text.y = element_text(size = 6),
          legend.text = element_text(size = 6),
          legend.title = element_text(size = 6),
          #axis.title.x = element_blank(),
          #axis.title.y = element_blank(),
          legend.position = "bottom")+
  labs(y = "IF cell types",x = "IMC cell types")
p


ggsave(plot = p, paste0(out_folder,"S1C_IMC_mIF_cell_type_correlation.pdf"),
       device = "pdf", 
       height = 7,
       width = 6.5)
```


# mean correlation of matche cell types
```{r, fig.height=18, fig.width=16}
celltype_dens_dat <- dens_dat %>%
  pivot_longer(-immucan_id) %>%
  mutate(celltype = str_split_i(name,"_",1),
         tech = str_split_i(name,"__",2)) %>%
  group_by(immucan_id,celltype, tech) %>%
  summarise(count = sum(value)) %>%
  pivot_wider(id_cols = immucan_id,values_from = (count),names_from = c(celltype,tech),names_sep = "__") %>%
  ungroup()


# correlation
cor <- psych::corr.test(celltype_dens_dat[,grepl("_IMC",colnames(celltype_dens_dat))],celltype_dens_dat[,grepl("__IF",colnames(celltype_dens_dat))],  method = "pearson",adjust = "BH")

# format correlation matrix
cur_dat <- as.data.frame(cor$r)
cur_dat$variable <- rownames(cur_dat)
dat_long <- reshape2::melt(cur_dat,id.vars="variable")
colnames(dat_long) <- c("cell_types_IMC","cell_types_mIF","correlation")

#get p-value matrix
p_dat <- as_tibble(cor$p)
p_dat$variable <- rownames(cur_dat)
pdat_long <- reshape2::melt(p_dat,id.vars="variable")
colnames(pdat_long) <- c("cell_types_IMC","cell_types_mIF","p_adj")

dat_long$p_adj <- pdat_long$p_adj

dat_long <- dat_long %>%
  mutate(sig = ifelse(p_adj <= 0.001 & p_adj > 0.0001,0.001,p_adj))

dat_long <- dat_long %>%
  mutate(sig = case_when(p_adj <= 0.0001 ~ "< 0.0001",
                         p_adj <= 0.001 & p_adj > 0.0001 ~ "< 0.001",
                         p_adj <= 0.01 & p_adj > 0.001 ~ "< 0.01",
                         p_adj <= 0.1 & p_adj > 0.01 ~ "< 0.1",
                         p_adj >0.1 ~ "ns"))


# plot
cor_filter <- 0.4
dat_long <- dat_long %>%
   mutate(cell_types_IMC = gsub("_density","",cell_types_IMC),
          cell_types_mIF = gsub("_density","",cell_types_mIF))


dat_long %>%
  mutate(IMC = str_split_i(cell_types_IMC,"_",1),
         IF = str_split_i(cell_types_mIF,"_",1),
          match = IMC == IF) %>%
  filter(match == TRUE) %>% arrange(correlation) %>%
  summarise(mean_cor = mean(abs(correlation)))

```

# S1D cellohood clusters heatmap panel 2
excluding TLS images
```{r, fig.height=10, fig.width=14}
library(ComplexHeatmap)
library(viridis)
library(circlize)

sce2 <- readRDS(paste0(mount_path,"data/IMC/Panel_2/Rout/sce.rds"))

# fraction data
cur_dat <- as.data.frame(colData(sce2))

frac_dat <- cur_dat %>%
  group_by(cellohood,.drop = FALSE) %>%
  dplyr::count(celltypes) %>%
  mutate(total_cells = sum(n),
         CN_freq = round(n/total_cells, digits = 2)*100) %>%
  pivot_wider(id_cols = cellohood,names_from = celltypes, values_from = CN_freq, values_fill = 0) %>%
  ungroup() %>%
  as.data.frame()

# generate data to merge with expression data which we can filter exp on later
merge_info <- frac_dat %>%
  pivot_longer(cols = -cellohood,names_to = "celltypes")

# generate matrix for heatmap
rownames(frac_dat) <- frac_dat$cellohood
frac_dat <- t(frac_dat[,-1])


# expression data
cur_dat <- as.data.frame(t(assay(sce2, "exprs")))
cur_dat$cellohood <- sce2$cellohood
cur_dat$celltypes <- sce2$celltypes
head(cur_dat)

cur_dat <- cur_dat %>%
  pivot_longer(-c(celltypes,cellohood)) %>%
  group_by(celltypes, cellohood,name,.drop = FALSE) %>%
  summarise(mean_exp = mean(value))

# merge with fraction data and filter only those expression that the respective cell type is present with at least 2% in the cluster
cur_dat <- cur_dat %>%
  left_join(merge_info,by = c("cellohood", "celltypes")) %>%
  filter(value > 2)


# fractions to merge
exp_dat <- cur_dat %>%
  mutate(type = paste0(celltypes,": ", name)) %>%
  filter(type %in% c("CD4: Tim3", "CD4: PD1","CD4: GITR", "CD8: PD1","CD8: Tim3", "CD8: CXCL13", "CD8: EOMES", "CD8: NKG2A","MacCD204: Arg1","MacCD204: CD11c","Treg: CD137","PMN_MDSC: MMP9","Tumor: p53","Tumor: EGFR","Tumor: Ki67","Tumor: CD56","Tumor: CD155","Tumor: BCL2", "Tumor: CD10","Tumor: CD73","Tumor: IgG", "Tumor: Podoplanin", "Fibro: Podoplanin", "Fibro: CD10" )) %>%
  pivot_wider(id_cols = cellohood,names_from = type, values_from = mean_exp) %>%
  as.data.frame()
rownames(exp_dat) <- exp_dat$cellohood
exp_dat <- exp_dat[,-1]
exp_dat_scaled <- t(scale(exp_dat))

# match column names of both matrices
exp_dat_scaled <- exp_dat_scaled[,colnames(frac_dat)]
# generate complex heatmap

#color functions
expression_col_fun <- colorRamp2(
  c(-3, 0, 3), 
  c("blue", "white", "red")
)
fraction_col_fun <- colorRamp2(seq(0, 50, length.out = 50), viridis(50))

# clustering 
col_clustering <- hclust(dist(t(frac_dat)))

ht1 <- Heatmap(
  frac_dat,
  name = "Fraction",
  col = fraction_col_fun,
  cluster_rows = TRUE,
  cluster_columns = col_clustering,
  show_row_names = TRUE,
  show_column_names = TRUE,
  rect_gp = gpar(col = "white", lwd = 1),
  cell_fun = function(j, i, x, y, width, height, fill) {
    value <- frac_dat[i, j]
    fill_col <- fraction_col_fun(value)  # get fill color from scale
    rgb_col <- col2rgb(fill_col) / 255   # normalize RGB
    lum <- 0.299 * rgb_col[1] + 0.587 * rgb_col[2] + 0.114 * rgb_col[3]  # luminance

    text_col <- if (lum < 0.5) "white" else "black"

    grid.text(
      label = round(value),
      x = x, y = y,
      gp = gpar(fontsize = 8, col = text_col)
    )
  }
)


ht2 <- Heatmap(
  exp_dat_scaled,
  name = "Z-scored\nExpression",
  rect_gp = gpar(col = "white", lwd = 1),
  col = expression_col_fun,
  
  cluster_rows = FALSE,  # Already clustered by ht1
  cluster_columns = col_clustering,
  show_row_names = TRUE,  # Avoid duplication
  show_column_names = TRUE,
  heatmap_legend_param = list(
    at = c(-3,-2,-1, 0, 1,2,3),  # control tick marks
    color_bar = "continuous"
  )
)

ht2 %v% ht1

pdf(paste0(out_folder,"S1D_cellohood_heatmap_panel_2.pdf"),
    width = 15,
    height = 10)
ht2 %v% ht1
dev.off()

cur_sce <- sce2[,sce2$celltypes %in% c("Fibro", "Tumor")]
plot_sce <- cur_sce[,sample(colnames(cur_sce),size = 10000,replace = FALSE)]
dittoSeq::dittoScatterPlot(plot_sce,x.var = "CD10", y.var = "Ecad",assay.x = "exprs", assay.y = "exprs", color.var = "immucan_id", legend.show = FALSE)
```
