---
title: "Figure S3"
author: "Daniel Schulz"
date: "2024-11-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggpubr)
library(SingleCellExperiment)
library(survivalAnalysis)
library(ggsurvfit)
library(survminer)
library(survival)
library(edgeR)
library(cowplot)
library(ggrepel)
library(igraph)
library(ggraph)
library(ggrepel)
library(scales)
library(ggnewscale)
```

```{r}
mount_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/"
out_folder <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/Figure_3_supp/"

if (!dir.exists(out_folder)) {
  dir.create(out_folder)
}
```

# load data
we will load the data we generated at the beginning of figure 2
```{r}
dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/abundances_dat.rds"))
dens_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/densities_dat.rds"))
cello_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/cello_dat.rds"))
clin_dat <- read.csv("/mnt/immucan_volume/clinical_data/Lung_NSCLC2/20230825/NSCLC2_for_R_update_2024.csv")
traits <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/all_traits.csv"))


# variables
data_type <- unique(str_split_i(string = colnames(dat[,-1]),pattern = "__",i = 2))
```

# S3A Kaplan Meier stage

```{r, fig.height=6}
cur_clin <- clin_dat
cur_clin$Stage <- cur_clin$simple_stage
cur_clin <- cur_clin %>%
  mutate(Stage = ifelse(Stage == "I","1",Stage),
         Stage = ifelse(Stage == "II","2",Stage),
         Stage = ifelse(Stage == "III","3",Stage),
         Stage = ifelse(Stage == "IV","4",Stage))
surv_fit <- survfit(Surv(time_to_event, censoring) ~ Stage, data = cur_clin)

p <- ggsurvplot(surv_fit, data = cur_clin, risk.table = TRUE, pval = TRUE,palette = c("palevioletred3","darkseagreen3","dodgerblue", "darkorange"))

p <- plot_grid(p$plot,p$table,ncol = 1, rel_heights = c(1,0.5))
p
ggsave(p,filename = paste0(out_folder,"S3A_Kaplan_Meier_CPS_score.pdf"),device = "pdf",width = 6, height = 5)
```

# S3B TCGA across stages adenocarcinoma
```{r}
tcga <- read.csv(paste0(mount_path,"data/RNAseq/luad_tcga_pan_can_atlas_2018/luad_tcga_pan_can_atlas_2018/xcell_between_samples_scores.csv"), sep = "\t")
tcga_patient <- read.csv(paste0(mount_path,"data/RNAseq/luad_tcga_pan_can_atlas_2018/luad_tcga_pan_can_atlas_2018/data_clinical_patient.txt"), sep = "\t")
tcga_clin <- read.csv(paste0(mount_path,"data/RNAseq/luad_tcga_pan_can_atlas_2018/luad_tcga_pan_can_atlas_2018/data_clinical_sample.txt"), sep = "\t")

tcga <- as.data.frame(t(tcga))
tcga <- tcga %>%
  mutate(PATIENT_ID = rownames(tcga)) %>%
  select(PATIENT_ID,everything())
rownames(tcga) <- NULL

tcga_patient <- tcga_patient %>%
  mutate(PATIENT_ID = gsub("-",".",PATIENT_ID))

tcga <- tcga %>%
  left_join(tcga_patient %>% select(PATIENT_ID,AJCC_PATHOLOGIC_TUMOR_STAGE),"PATIENT_ID") %>%
  mutate(stage = ifelse(AJCC_PATHOLOGIC_TUMOR_STAGE %in% c("STAGE I","STAGE IA","STAGE IB"),1,NA),
         stage = ifelse(AJCC_PATHOLOGIC_TUMOR_STAGE %in% c("STAGE II","STAGE IIA","STAGE IIB"),2,stage),
         stage = ifelse(AJCC_PATHOLOGIC_TUMOR_STAGE %in% c("STAGE IIIA","STAGE IIIB"),3,stage)) %>%
  filter(!is.na(stage)) %>%
  select(-AJCC_PATHOLOGIC_TUMOR_STAGE)

res <- tcga %>%
  pivot_longer(-c(PATIENT_ID, stage)) %>%
  mutate(stage = as.numeric(as.factor(stage))) %>%
  group_by(name) %>%
  nest() %>%
  mutate(model = map(data, ~ lm(value ~ stage, data = .x)),
         summary = map(model, broom::tidy)) %>%
  select(name, summary) %>%
  unnest(c(summary)) %>%
  filter(term == "stage") %>%
  ungroup() %>%
  mutate(FDR = p.adjust(p = p.value,method = "BH")) %>%
  filter(FDR < 0.1)

res <- res %>%
  mutate(p.plot = ifelse(p.value < 0.001,"***",ifelse(p.value < 0.01, "**","*")))

p <- tcga %>%
  pivot_longer(-c(PATIENT_ID, stage)) %>%
  mutate(stage = as.numeric(as.factor(stage))) %>%
  group_by(name) %>%
  filter(name %in% res$name) %>%
  ggplot(aes(x = name, y = value,fill=as.factor(stage)))+
  geom_boxplot()+
  geom_text(data = res, aes(x = name,y=0.8,label = p.plot), inherit.aes = FALSE,size = 8)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust =0.5,hjust = 1, size = 10),
        axis.title.x = element_blank())+
  scale_fill_manual(values = c("1" = "palevioletred3", "2" = "darkseagreen3", "3" = "dodgerblue"))+
  ggtitle(label = "Deconvoluted cell types across stages in TCGA LUAD")+
  labs(fill = "Stage")

p

ggsave(filename = paste0(out_folder,"S3B_Xcell_across_TCGA_stages_within_LUAD.pdf"),
       plot = p,
       device = "pdf",
       width = 10,
       height = 6)
```

# deconvolution of RNAseq
```{r, fig.width=25, fig.height=8}
dat <- read.csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/RNASeq/Rout/tables/RNAseq/xcell_results/xCell_TPM_norm_for_xCell_xCell_1129103125.txt", sep = "\t")
dat_t <- as.data.frame(t(dat[,-1]))
colnames(dat_t) <- dat$X
dat_t$immucan_id <- gsub("\\.","-",rownames(dat_t))
rownames(dat_t) <- NULL
unique(dat_t$immucan_id)
dat_t
dat <- left_join(dat_t,clin_dat[,c("immucan_id","simple_stage", "simple_histology")],"immucan_id")


sigs <- dat_t %>%
  pivot_longer(-immucan_id) %>%
  left_join(clin_dat[,c("immucan_id","simple_stage", "simple_histology")],by = "immucan_id") %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  select(immucan_id,name, value,simple_stage) %>%
  mutate(stage = ifelse(simple_stage == "I",1,NA),
         stage = ifelse(simple_stage == "II",2,stage),
         stage = ifelse(simple_stage == "III",3,stage),
         stage = ifelse(simple_stage == "IV",4,stage),
         stage = as.numeric(as.factor(stage))) %>%
  filter(stage != 4) %>%
  group_by(name) %>%
  nest() %>%
  mutate(model = map(data, ~ lm(value ~ stage, data = .x)),
         summary = map(model, broom::tidy)) %>%
  select(name, summary) %>%
  unnest(c(summary)) %>%
  filter(term == "stage") %>%
  mutate(tech = str_split_i(name,"__",2)) %>%
  group_by(tech) %>%
  mutate(FDR = p.adjust(p = p.value,method = "BH"),
         sig_p = ifelse(p.value < 0.05,TRUE, FALSE),
         sig_fdr = ifelse(FDR < 0.1,TRUE, FALSE))

summary_dat <- sigs %>%
  filter(sig_p == TRUE) %>%
  #filter(grepl("CD8_GATA3_BCL2|^CD8_density|CD8_Ki67_density|CD8_TCF7_density|Tumor_IDO1",name)) %>%
  mutate(y_position = 1000) %>%
  ungroup() %>%
  as.data.frame()

p2 <- dat_t %>%
  pivot_longer(-immucan_id) %>%
  mutate(tech = str_split_i(name,"__",2)) %>%
  left_join(clin_dat,"immucan_id") %>%
  filter(simple_histology == "Squamous cell carcinoma") %>%
  select(immucan_id,name, value,simple_stage, tech) %>%
  mutate(stage = ifelse(simple_stage == "I",1,NA),
         stage = ifelse(simple_stage == "II",2,stage),
         stage = ifelse(simple_stage == "III",3,stage),
         stage = ifelse(simple_stage == "IV",4,stage),
         stage = as.numeric(as.factor(stage)),
         name = gsub(pattern = "density_",replacement = "",x = name)) %>%
  #filter(name %in% summary_dat$name,
  #       stage != 4) %>%
  ggplot(aes(x = name,y = log10(value+1),fill=as.factor(stage))) +
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(col = as.factor(stage)),position = position_jitterdodge(jitter.width = 0.2),size= 0.5)+
  scale_fill_manual(values = alpha(c("1" = "palevioletred3", "2" = "darkseagreen3", "3" = "dodgerblue"),alpha = 0.6))+
  scale_color_manual(values = c("1" = "palevioletred3", "2" = "darkseagreen3", "3" = "dodgerblue"))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90,vjust= 0.5, hjust = 1),
        axis.title = element_blank())+
  guides(fill=guide_legend(title="Stage"),
         col=guide_legend(title="Stage"))+
  geom_text(data = summary_dat, aes(x = name,y=1,label = paste0("fdr: ",sig_fdr)), inherit.aes = FALSE)+
  geom_text(data = summary_dat, aes(x = name,y=1,label = paste0("p : ",round(p.value,digits = 3))), inherit.aes = FALSE)+
  ggtitle(label = "LUAD")
  
p2

dat_t %>%
  pivot_longer(-immucan_id) %>%
  mutate(tech = str_split_i(name,"__",2)) %>%
  left_join(clin_dat,"immucan_id") %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  distinct(immucan_id)
```




# diff abundance stages bubble plot
not in paper
here we will estimate the dispersion across the whole data set and then do the linear model of count comparison only for the groups of interest.
```{r, fig.height=8, fig.width=10}
condition <- c("I","I","II")
names(condition) <- c("stage_1_2","stage_1_3","stage_2_3")

p.list <- list()
for (i in c(data_type,"cello1","cello2")) {
    
    if(grepl("cello",i)){
      abundances <- cello_dat %>%
        select(contains(i))
      abundances <- t(abundances)
      colnames(abundances) <- cello_dat$immucan_id
    } else {
      abundances <- dat %>%
        select(contains(i)) %>%
        filter(complete.cases(.))
      abundances <- t(abundances)
      colnames(abundances) <- dat %>%
        select(contains(i),immucan_id) %>%
        filter(complete.cases(.)) %>%
        pull(immucan_id)
    }
    anno <- traits
    
    # remove patients without TMB info from the abundance data
    abundances <- abundances[,colnames(abundances) %in% anno$immucan_id]
  
  meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
  Size <- factor(meta[,"simple_stage"])
  meta$Size <- Size
  
  # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
  y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
  design <- model.matrix(~0+group,data = y.ab$samples)
  colnames(design) <- levels(y.ab$samples$group)
  # we do not fit a trend 
  y.ab <- estimateDisp(y.ab, design, trend.method = "none")
  y.ab$common.dispersion
  #plotBCV(y.ab,pch = 5)
  
  # we only compare the diff abundance for Adeno vs Sqam.
  fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
  
  
  fit_1_2 <- glmLRT(glmfit = fit.ab,contrast = c(-1,1,0,0))
  fit_1_3 <- glmLRT(glmfit = fit.ab,contrast = c(-1,0,1,0))
  fit_2_3 <- glmLRT(glmfit = fit.ab,contrast = c(0,-1,1,0))
  res_list <- list(fit_1_2,fit_1_3,fit_2_3)
  
  cur_list <- lapply(res_list,function(x){
    sig_dat <- x$table
    sig_dat$celltypes <- rownames(sig_dat)
    sig_dat$comparison <- gsub("Size","",x$comparison)
    return(sig_dat)
  })
  
  p.list[[i]] <- do.call(rbind,cur_list)
}   

cur_dat <- do.call(rbind,p.list)
cur_dat <- cur_dat %>%
  mutate(tech = gsub("_abundance","",str_split_i(celltypes,"__",2)),
         cond = ifelse(comparison == "-1*I 1*II","stage_2_over_1","stage_3_over_1"),
         cond = ifelse(comparison == "-1*II 1*III","stage_3_over_2",cond)) %>%
  group_by(tech) %>%
  mutate(FDR = p.adjust(PValue,method = "BH"),
         sig = ifelse(PValue < 0.05 & logFC > 0, "UP_trend",NA),
         sig = ifelse(FDR < 0.1 & logFC > 0, "UP",sig),
         sig = ifelse(PValue < 0.05 & logFC < 0, "DOWN_trend",sig),
         sig = ifelse(FDR < 0.1 & logFC < 0, "DOWN",sig),
         sig = ifelse(is.na(sig),"not sig", sig),
         ct_name = gsub("_density","",str_split_i(celltypes,"__",1)))

to_show <- cur_dat %>%
  filter(sig %in% c("DOWN", "UP")) %>%
  pull(ct_name)
# add the one from the linear model of densities to the plot
to_show <- c(to_show,"CD4","CD8","CD8_GATA3_BCL2","CD8_Ki67","tumor_B2M_HLADR","Tumor_IDO1")

cur_dat <- cur_dat %>%
  filter(ct_name %in% to_show) %>%
  mutate(celltypes = gsub("_abundance","",celltypes))

## spatial cluster renaming
cur_dat <- cur_dat %>%
  mutate(celltypes = gsub(pattern = "_cello",replacement = "IMC",x = celltypes))

# get max and min fold change values
max_up <- max(cur_dat$logFC)+1
min_down <- min(cur_dat$logFC)-1

# plot the data
p <- cur_dat %>%
  ggplot(aes(x=celltypes, y = cond,col = logFC, size= -log10(PValue)))+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(round(min_down),0,round(max_up))),
                         guide = "colorbar", limits=c(round(min_down),round(max_up)))+
  scale_size(range = c(1,4))+
  new_scale_colour() +
  geom_point(data = cur_dat,shape = 1,aes(size = -log10(PValue), col = sig), stroke =1)+
  scale_color_manual(values = c("UP" = "darkred","UP_trend" = "red","DOWN" = "darkblue","UP_trend" = "blue","not sig" = "gray"), name = "stats")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 10),
        axis.text.y = element_text(size = 10),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10),
        axis.title.y = element_blank(),
        plot.margin = unit(c(.1, .5, .5, .5), "cm"),
        legend.position = "bottom")+
  labs(x = "Feature")
p

# ggsave(filename = paste0(out_folder,"diff_abundance_stage.pdf"),
#        plot = p,
#        device = "pdf",
#        width = 7,
#        height =4)
```


## stages for LUAD  only
not in paper
here we will estimate the dispersion across the whole data set and then do the linear model of count comparison only for the groups of interest.
```{r, fig.height=8, fig.width=10}
condition <- c("I","I","II")
names(condition) <- c("stage_1_2","stage_1_3","stage_2_3")

p.list <- list()
for (i in c(data_type,"cello1","cello2")) {
    
    if(grepl("cello",i)){
      abundances <- cello_dat %>%
        select(contains(i))
      abundances <- t(abundances)
      colnames(abundances) <- cello_dat$immucan_id
    } else {
      abundances <- dat %>%
        select(contains(i)) %>%
        filter(complete.cases(.))
      abundances <- t(abundances)
      colnames(abundances) <- dat %>%
        select(contains(i),immucan_id) %>%
        filter(complete.cases(.)) %>%
        pull(immucan_id)
    }
    anno <- traits %>%
      filter(simple_histology == "Adenocarcinoma")
    
    # remove patients without TMB info from the abundance data
    abundances <- abundances[,colnames(abundances) %in% anno$immucan_id]
  
  meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
  Size <- factor(meta[,"simple_stage"])
  meta$Size <- Size
  
  # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
  y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
  design <- model.matrix(~0+group,data = y.ab$samples)
  colnames(design) <- levels(y.ab$samples$group)
  # we do not fit a trend 
  y.ab <- estimateDisp(y.ab, design, trend.method = "none")
  y.ab$common.dispersion
  #plotBCV(y.ab,pch = 5)
  
  # we only compare the diff abundance for Adeno vs Sqam.
  fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
  
  
  fit_1_2 <- glmLRT(glmfit = fit.ab,contrast = c(-1,1,0,0))
  fit_1_3 <- glmLRT(glmfit = fit.ab,contrast = c(-1,0,1,0))
  fit_2_3 <- glmLRT(glmfit = fit.ab,contrast = c(0,-1,1,0))
  res_list <- list(fit_1_2,fit_1_3,fit_2_3)
  
  cur_list <- lapply(res_list,function(x){
    sig_dat <- x$table
    sig_dat$celltypes <- rownames(sig_dat)
    sig_dat$comparison <- gsub("Size","",x$comparison)
    return(sig_dat)
  })
  
  p.list[[i]] <- do.call(rbind,cur_list)
}   

cur_dat <- do.call(rbind,p.list)
cur_dat <- cur_dat %>%
  mutate(tech = gsub("_abundance","",str_split_i(celltypes,"__",2)),
         cond = ifelse(comparison == "-1*I 1*II","stage_2_over_1","stage_3_over_1"),
         cond = ifelse(comparison == "-1*II 1*III","stage_3_over_2",cond)) %>%
  group_by(tech) %>%
  mutate(FDR = p.adjust(PValue,method = "BH"),
         sig_overall = ifelse(FDR < 0.1 ,TRUE,FALSE),
         sig = ifelse(PValue < 0.05 ,TRUE,FALSE),
         direction = ifelse(logFC > 0 ,"up", "down"),
         ct_name = gsub("_abundance","",str_split_i(celltypes,"__",1)))

to_show <- cur_dat %>%
  filter(sig == TRUE) %>%
  pull(ct_name)
# add the one from the linear model of densities to the plot
to_show <- c(to_show,"CD4","CD8","CD8_GATA3_BCL2","CD8_Ki67","Vasculature","Tumor_IDO1")

cur_dat <- cur_dat %>%
  filter(ct_name %in% to_show) %>%
  mutate(celltypes = gsub("_abundance","",celltypes))

## spatial cluster renaming
cur_dat <- cur_dat %>%
  mutate(celltypes = gsub(pattern = "_cello",replacement = "IMC",x = celltypes))

# get max and min fold change values
max_up <- max(cur_dat$logFC)+1
min_down <- min(cur_dat$logFC)-1

# plot the data
p <- cur_dat %>%
  ggplot(aes(x=celltypes, y = cond,col = logFC, size= -log10(PValue)))+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(round(min_down),0,round(max_up))),
                         guide = "colorbar", limits=c(round(min_down),round(max_up)))+
  scale_size(range = c(1,4))+
  new_scale_colour() +
  geom_point(data = cur_dat %>% filter(sig == TRUE),shape = 1,aes(size = -log10(PValue), col = sig), stroke =1)+
  scale_color_manual(values = c("TRUE" = "black","FALSE" = "white"), name = "p-value < 0.1")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 10),
        axis.text.y = element_text(size = 10),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10),
        axis.title.y = element_blank(),
        plot.margin = unit(c(.1, .5, .5, .5), "cm"),
        legend.position = "bottom")+
  labs(x = "Feature")
p
```

## stages for LUSC  only
here we will estimate the dispersion across the whole data set and then do the linear model of count comparison only for the groups of interest.
```{r, fig.height=8, fig.width=10}
condition <- c("I","I","II")
names(condition) <- c("stage_1_2","stage_1_3","stage_2_3")

p.list <- list()
for (i in c(data_type,"cello1","cello2")) {
    
    if(grepl("cello",i)){
      abundances <- cello_dat %>%
        select(contains(i))
      abundances <- t(abundances)
      colnames(abundances) <- cello_dat$immucan_id
    } else {
      abundances <- dat %>%
        select(contains(i)) %>%
        filter(complete.cases(.))
      abundances <- t(abundances)
      colnames(abundances) <- dat %>%
        select(contains(i),immucan_id) %>%
        filter(complete.cases(.)) %>%
        pull(immucan_id)
    }
    anno <- traits %>%
      filter(simple_histology == "Squamous cell carcinoma")
    
    # remove patients without TMB info from the abundance data
    abundances <- abundances[,colnames(abundances) %in% anno$immucan_id]
  
  meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
  Size <- factor(meta[,"simple_stage"])
  meta$Size <- Size
  
  # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
  y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
  design <- model.matrix(~0+group,data = y.ab$samples)
  colnames(design) <- levels(y.ab$samples$group)
  # we do not fit a trend 
  y.ab <- estimateDisp(y.ab, design, trend.method = "none")
  y.ab$common.dispersion
  #plotBCV(y.ab,pch = 5)
  
  # we only compare the diff abundance for Adeno vs Sqam.
  fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
  
  
  fit_1_2 <- glmLRT(glmfit = fit.ab,contrast = c(-1,1,0,0))
  fit_1_3 <- glmLRT(glmfit = fit.ab,contrast = c(-1,0,1,0))
  fit_2_3 <- glmLRT(glmfit = fit.ab,contrast = c(0,-1,1,0))
  res_list <- list(fit_1_2,fit_1_3,fit_2_3)
  
  cur_list <- lapply(res_list,function(x){
    sig_dat <- x$table
    sig_dat$celltypes <- rownames(sig_dat)
    sig_dat$comparison <- gsub("Size","",x$comparison)
    return(sig_dat)
  })
  
  p.list[[i]] <- do.call(rbind,cur_list)
}   

cur_dat <- do.call(rbind,p.list)
cur_dat <- cur_dat %>%
  mutate(tech = gsub("_abundance","",str_split_i(celltypes,"__",2)),
         cond = ifelse(comparison == "-1*I 1*II","stage_2_over_1","stage_3_over_1"),
         cond = ifelse(comparison == "-1*II 1*III","stage_3_over_2",cond)) %>%
  group_by(tech) %>%
  mutate(FDR = p.adjust(PValue,method = "BH"),
         sig_overall = ifelse(FDR < 0.1 ,TRUE,FALSE),
         sig = ifelse(PValue < 0.1 ,TRUE,FALSE),
         direction = ifelse(logFC > 0 ,"up", "down"),
         ct_name = gsub("_abundance","",str_split_i(celltypes,"__",1)))

to_show <- cur_dat %>%
  filter(sig_overall == TRUE) %>%
  pull(ct_name)
# add the one from the linear model of densities to the plot
to_show <- c(to_show,"CD4","CD8","CD8_GATA3_BCL2","CD8_Ki67","Vasculature","Tumor_IDO1")

cur_dat <- cur_dat %>%
  filter(ct_name %in% to_show) %>%
  mutate(celltypes = gsub("_abundance","",celltypes))

## spatial cluster renaming
cur_dat <- cur_dat %>%
  mutate(celltypes = gsub(pattern = "_cello",replacement = "IMC",x = celltypes))

# get max and min fold change values
max_up <- max(cur_dat$logFC)+1
min_down <- min(cur_dat$logFC)-1

# plot the data
p <- cur_dat %>%
  ggplot(aes(x=celltypes, y = cond,col = logFC, size= -log10(PValue)))+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(round(min_down),0,round(max_up))),
                         guide = "colorbar", limits=c(round(min_down),round(max_up)))+
  scale_size(range = c(1,4))+
  new_scale_colour() +
  geom_point(data = cur_dat %>% filter(sig == TRUE),shape = 1,aes(size = -log10(PValue), col = sig), stroke =1)+
  scale_color_manual(values = c("TRUE" = "black","FALSE" = "white"), name = "p-value < 0.1")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 10),
        axis.text.y = element_text(size = 10),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10),
        axis.title.y = element_blank(),
        plot.margin = unit(c(.1, .5, .5, .5), "cm"),
        legend.position = "bottom")+
  labs(x = "Feature")
p
```


# S3C linear regression for stage for LUSC

```{r, fig.width=10}
library(purrr)
library(broom)

sigs <- dens_dat %>%
  pivot_longer(-immucan_id) %>%
    left_join(clin_dat,"immucan_id") %>%
  filter(simple_histology == "Squamous cell carcinoma") %>%
  select(immucan_id,name, value,simple_stage) %>%
  mutate(stage = ifelse(simple_stage == "I",1,NA),
         stage = ifelse(simple_stage == "II",2,stage),
         stage = ifelse(simple_stage == "III",3,stage),
         stage = ifelse(simple_stage == "IV",4,stage),
         stage = as.numeric(as.factor(stage))) %>%
  filter(stage != 4) %>%
  group_by(name) %>%
  nest() %>%
  mutate(model = map(data, ~ lm(value ~ stage, data = .x)),
         summary = map(model, broom::tidy)) %>%
  select(name, summary) %>%
  unnest(c(summary)) %>%
  filter(term == "stage") %>%
  mutate(tech = str_split_i(name,"__",2)) %>%
  group_by(tech) %>%
  mutate(FDR = p.adjust(p = p.value,method = "BH"),
         sig_p = ifelse(p.value < 0.05,TRUE, FALSE),
         sig_fdr = ifelse(FDR < 0.1,TRUE, FALSE))

summary_dat <- sigs %>%
  filter(sig_p == TRUE) %>%
  #filter(grepl("CD8_GATA3_BCL2|^CD8_density|CD8_Ki67_density|CD8_TCF7_density|Tumor_IDO1",name)) %>%
  mutate(y_position = 1000,
         name = gsub(pattern = "density_",replacement = "",x = name)) %>%
  ungroup() %>%
  as.data.frame()

p2 <- dens_dat %>%
  pivot_longer(-immucan_id) %>%
  mutate(tech = str_split_i(name,"__",2)) %>%
  left_join(clin_dat,"immucan_id") %>%
  filter(simple_histology == "Squamous cell carcinoma") %>%
  select(immucan_id,name, value,simple_stage, tech) %>%
  mutate(stage = ifelse(simple_stage == "I",1,NA),
         stage = ifelse(simple_stage == "II",2,stage),
         stage = ifelse(simple_stage == "III",3,stage),
         stage = ifelse(simple_stage == "IV",4,stage),
         stage = as.numeric(as.factor(stage)),
         name = gsub(pattern = "density_",replacement = "",x = name)) %>%
  filter(name %in% summary_dat$name,
         stage != 4) %>%
  ggplot(aes(x = name,y = log10(value+1),fill=as.factor(stage))) +
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(col = as.factor(stage)),position = position_jitterdodge(jitter.width = 0.2),size= 0.5)+
  scale_fill_manual(values = alpha(c("1" = "palevioletred3", "2" = "darkseagreen3", "3" = "dodgerblue"),alpha = 0.6))+
  scale_color_manual(values = c("1" = "palevioletred3", "2" = "darkseagreen3", "3" = "dodgerblue"))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90,vjust= 0.5, hjust = 1),
        axis.title = element_blank())+
  guides(fill=guide_legend(title="Stage"),
         col=guide_legend(title="Stage"))+
  geom_text(data = summary_dat, aes(x = name,y=4.2,label = paste0("fdr: ",sig_fdr)), inherit.aes = FALSE,size = 3)+
  geom_text(data = summary_dat, aes(x = name,y=4.6,label = paste0("p : ",round(p.value,digits = 3))), inherit.aes = FALSE, size = 3)+
  coord_cartesian(ylim = c(0,4.6))+
  ggtitle(label = "LUSC")

p2

ggsave(filename = paste0(out_folder,"S3C_linear_model_cell_type_densitie_LUSC_for_stages.pdf"),
       plot = last_plot(),
       device = "pdf",
       width = 15,
       height = 3.75)
```

## major cell types changes
not in paper
```{r, fig.width=10}
library(purrr)
library(broom)
d1 <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/celltypes_density_dat_p1.csv"))
d2 <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/celltypes_density_dat_p2.csv")) 

colnames(d1) <- c("immucan_id",paste0(colnames(d1)[-1],"__IMC1"))
colnames(d2) <- c("immucan_id",paste0(colnames(d2)[-1],"__IMC2"))
d1
d2

dens_dat <- left_join(d1,d2,"immucan_id")

sigs <- dens_dat %>%
  pivot_longer(-immucan_id) %>%
    left_join(clin_dat,"immucan_id") %>%
  filter(simple_histology == "Squamous cell carcinoma") %>%
  select(immucan_id,name, value,simple_stage) %>%
  mutate(stage = ifelse(simple_stage == "I",1,NA),
         stage = ifelse(simple_stage == "II",2,stage),
         stage = ifelse(simple_stage == "III",3,stage),
         stage = ifelse(simple_stage == "IV",4,stage),
         stage = as.numeric(as.factor(stage))) %>%
  filter(stage != 4) %>%
  group_by(name) %>%
  nest() %>%
  mutate(model = map(data, ~ lm(value ~ stage, data = .x)),
         summary = map(model, broom::tidy)) %>%
  select(name, summary) %>%
  unnest(c(summary)) %>%
  filter(term == "stage") %>%
  mutate(tech = str_split_i(name,"__",2)) %>%
  group_by(tech) %>%
  mutate(FDR = p.adjust(p = p.value,method = "BH"),
         sig_p = ifelse(p.value < 0.05,TRUE, FALSE),
         sig_fdr = ifelse(FDR < 0.1,TRUE, FALSE))

summary_dat <- sigs %>%
  filter(sig_p == TRUE) %>%
  #filter(grepl("CD8_GATA3_BCL2|^CD8_density|CD8_Ki67_density|CD8_TCF7_density|Tumor_IDO1",name)) %>%
  mutate(y_position = 1000,
         name = gsub(pattern = "density_",replacement = "",x = name)) %>%
  ungroup() %>%
  as.data.frame()

p2 <- dens_dat %>%
  pivot_longer(-immucan_id) %>%
  mutate(tech = str_split_i(name,"__",2)) %>%
  left_join(clin_dat,"immucan_id") %>%
  filter(simple_histology == "Squamous cell carcinoma") %>%
  select(immucan_id,name, value,simple_stage, tech) %>%
  mutate(stage = ifelse(simple_stage == "I",1,NA),
         stage = ifelse(simple_stage == "II",2,stage),
         stage = ifelse(simple_stage == "III",3,stage),
         stage = ifelse(simple_stage == "IV",4,stage),
         stage = as.numeric(as.factor(stage)),
         name = gsub(pattern = "density_",replacement = "",x = name)) %>%
  filter(name %in% summary_dat$name,
         stage != 4) %>%
  ggplot(aes(x = name,y = log10(value+1),fill=as.factor(stage))) +
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(col = as.factor(stage)),position = position_jitterdodge(jitter.width = 0.2),size= 0.5)+
  scale_fill_manual(values = alpha(c("1" = "palevioletred3", "2" = "darkseagreen3", "3" = "dodgerblue"),alpha = 0.6))+
  scale_color_manual(values = c("1" = "palevioletred3", "2" = "darkseagreen3", "3" = "dodgerblue"))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90,vjust= 0.5, hjust = 1),
        axis.title = element_blank())+
  guides(fill=guide_legend(title="Stage"),
         col=guide_legend(title="Stage"))+
  geom_text(data = summary_dat, aes(x = name,y=4.2,label = paste0("fdr: ",sig_fdr)), inherit.aes = FALSE,size = 3)+
  geom_text(data = summary_dat, aes(x = name,y=4.6,label = paste0("p : ",round(p.value,digits = 3))), inherit.aes = FALSE, size = 3)+
  coord_cartesian(ylim = c(0,4.6))+
  ggtitle(label = "LUSC")

p2

```




# S3D diff abundance smoking cellohood

```{r, fig.height=15, fig.width=25}
condition <- c("Non-smoker","Non-smoker","Ex-smoker")
names(condition) <- c("never_vs_current_smoker","never_vs_Ex_smoker","current_vs_Ex_smoker")

p.list <- list()

# merge with clinical data
anno <- clin_dat %>%
  filter(!is.na(smoking_status),
         simple_histology == "Adenocarcinoma")
  
for (i in c("cello1","cello2")) {

  abundances <- cello_dat %>%
    select(contains(i))
  abundances <- t(abundances)
  colnames(abundances) <- cello_dat$immucan_id
  
  abundances <- abundances[,colnames(abundances) %in% anno$immucan_id]
  
  meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
  Size <- factor(meta[,"smoking_status"])
  meta$Size <- Size
  
  # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
  y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
  design <- model.matrix(~0+group,data = y.ab$samples)
  colnames(design) <- levels(y.ab$samples$group)
  # we do not fit a trend 
  y.ab <- estimateDisp(y.ab, design, trend.method = "none")
  y.ab$common.dispersion
  #plotBCV(y.ab,pch = 5)
  
  # we only compare the diff abundance for Adeno vs Sqam.
  fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
  fit_1_2 <- glmLRT(glmfit = fit.ab,contrast = c(1,-1,0))
  fit_1_3 <- glmLRT(glmfit = fit.ab,contrast = c(1,0,-1))
  fit_2_3 <- glmLRT(glmfit = fit.ab,contrast = c(0,1,-1))
  res_list <- list(fit_1_2,fit_1_3,fit_2_3)
  
  cur_list <- lapply(res_list,function(x){
    sig_dat <- x$table
    sig_dat$celltypes <- rownames(sig_dat)
    sig_dat$comparison <- gsub("Size","",x$comparison)
    return(sig_dat)
  })
  
  p.list[[i]] <- do.call(rbind,cur_list)
}   

cur_dat_LUAD <- do.call(rbind,p.list)
cur_dat_LUAD <- cur_dat_LUAD %>%
  mutate(histology = "LUAD")
```

## smoking status LUSC
of note, there are no non-smokers for LUSC histology in our cohort 
```{r}
condition <- c("Non-smoker","Non-smoker","Ex-smoker")
names(condition) <- c("never_vs_current_smoker","never_vs_Ex_smoker","current_vs_Ex_smoker")
p.list <- list()

# merge with clinical data
anno <- clin_dat %>%
  filter(!is.na(smoking_status),
         simple_histology == "Squamous cell carcinoma")
  
for (i in c("cello1","cello2")) {

  abundances <- cello_dat %>%
    select(contains(i))
  abundances <- t(abundances)
  colnames(abundances) <- cello_dat$immucan_id
  
  abundances <- abundances[,colnames(abundances) %in% anno$immucan_id]
  
  meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
  Size <- factor(meta[,"smoking_status"])
  meta$Size <- Size
  
  # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
  y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
  design <- model.matrix(~0+group,data = y.ab$samples)
  colnames(design) <- levels(y.ab$samples$group)
  # we do not fit a trend 
  y.ab <- estimateDisp(y.ab, design, trend.method = "none")
  y.ab$common.dispersion
  #plotBCV(y.ab,pch = 5)
  
  # we only compare the diff abundance for Adeno vs Sqam.
  fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
  fit_1_2 <- glmLRT(glmfit = fit.ab,contrast = c(1,-1))
  res_list <- list(fit_1_2)
  
  cur_list <- lapply(res_list,function(x){
    sig_dat <- x$table
    sig_dat$celltypes <- rownames(sig_dat)
    sig_dat$comparison <- gsub("Size","",x$comparison)
    return(sig_dat)
  })
  
  p.list[[i]] <- do.call(rbind,cur_list)
}   

cur_dat_LUSC <- do.call(rbind,p.list)

cur_dat_LUSC <- cur_dat_LUSC %>%
  mutate(histology = "LUSC")
```


## smoking signature based

```{r}
# Merge with mutational signatures
Lung_NSCLC2_filtered <-read.delim(paste0(mount_path,"data/WES/mutation_signatures/Lung_NSCLC2_filtered.tsv"), sep="\t")
names(Lung_NSCLC2_filtered)[names(Lung_NSCLC2_filtered)=="patient"] <- "immucan_id"
anno_WES <- merge(clin_dat, Lung_NSCLC2_filtered, by="immucan_id", sort=F)

# Combine smoking signatures
anno_WES[,"SBS4_SBS92"] <- apply(anno_WES[,c("SBS4","SBS92")],1,sum)

# Dichotomise smoking signature
cutoff <- median(anno_WES$SBS4_SBS92)
anno_WES <- anno_WES %>%
      filter(!is.na(SBS4_SBS92))  %>%
      mutate(SBS4_SBS92_dicho = ifelse(SBS4_SBS92 > cutoff, "high","low"))


# diff abundance
anno <- anno_WES
p.list <- list()

for (i in c("cello1","cello2")) {

  abundances <- cello_dat %>%
    select(contains(i))
  abundances <- t(abundances)
  colnames(abundances) <- cello_dat$immucan_id
  abundances <- as.data.frame(abundances[,colnames(abundances) %in% anno$immucan_id])
   
  meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
  stopifnot(all(colnames(abundances)==meta$immucan_id))
    
  Size <- factor(meta[,"SBS4_SBS92_dicho"], levels = c("low", "high"))

  # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
  y.ab <- DGEList(counts = abundances, group = Size)
  design <- model.matrix(~0+Size)
  colnames(design) <- levels(Size)
    
  # we do not fit a trend 
  y.ab <- estimateDisp(y.ab, design, trend.method = "none")
  y.ab$common.dispersion

  fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
  fit_1_2 <- glmLRT(glmfit = fit.ab,contrast = c(-1,1))

  res_list <- list(fit_1_2)
    
  cur_list <- lapply(res_list,function(x){
      sig_dat <- x$table
      sig_dat$celltypes <- rownames(sig_dat)
      sig_dat$comparison <- gsub("Size","",x$comparison)
      return(sig_dat)
    })
  p.list[[i]] <- do.call(rbind,cur_list)
}

cur_dat.SBS <- do.call(rbind,p.list)
cur_dat.SBS <-cur_dat.SBS %>%
  mutate(histology = "all")
```


```{r, fig.height=10, fig.width=20}
cur_dat <- rbind(cur_dat_LUAD,cur_dat_LUSC,cur_dat.SBS)

cur_dat <- cur_dat %>%
  mutate(tech = gsub("_density","",str_split_i(celltypes,"__",2)),
         cond = ifelse(comparison == "1*Current smoker -1*Ex-smoker","CurrentSmoker_over_ExSmoker","ExSmoker_over_NonSmoker"),
         cond = ifelse(comparison == "1*Current smoker -1*Non-smoker","CurrentSmoker_over_NonSmoker",cond),
         cond = paste0(histology,"_",cond),
         cond = ifelse(comparison == "-1*low 1*high","SBS4-SBS92_High_over_Low",cond)) %>%
  group_by(tech) %>%
  mutate(FDR = p.adjust(PValue,method = "BH"),
         sig = ifelse(PValue < 0.05 & logFC > 0, "UP_trend",NA),
         sig = ifelse(FDR < 0.1 & logFC > 0, "UP",sig),
         sig = ifelse(PValue < 0.05 & logFC < 0, "DOWN_trend",sig),
         sig = ifelse(FDR < 0.1 & logFC < 0, "DOWN",sig),
         sig = ifelse(is.na(sig),"NOTSIG", sig),
         ct_name = gsub("_density","",str_split_i(celltypes,"__",1)))
  
to_show <- cur_dat %>%
  filter(sig %in% c("UP","DOWN")) %>%
  pull(celltypes)

cur_dat <- cur_dat %>%
  filter(celltypes %in% to_show) %>%
  mutate(celltypes = gsub("_abundance","",celltypes),
         celltypes = gsub("cello", "IMC",celltypes))

# get max and min fold change values
max_up <- max(cur_dat$logFC)+1
min_down <- min(cur_dat$logFC)-1

# plot the data
p <- cur_dat %>%
  ggplot(aes(x=celltypes, y = factor(cond, levels = c("SBS4-SBS92_High_over_Low",
                                                      "LUAD_CurrentSmoker_over_ExSmoker",
                                                      "LUAD_ExSmoker_over_NonSmoker",
                                                      "LUAD_CurrentSmoker_over_NonSmoker",
                                                      "LUSC_CurrentSmoker_over_ExSmoker")),col = logFC, size= -log10(PValue)))+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(
    colours = c("blue", "white", "red"), 
    values = rescale(c(round(min_down), 0, round(max_up))),
    limits = c(round(min_down), round(max_up)),
    guide = guide_colorbar(
      title = "logFC",                   
      title.theme = element_text(size = 12),
      label.theme = element_text(size = 12)
    )
  )+
  scale_size(range = c(1,4))+
  new_scale_colour() +
  geom_point(data = cur_dat,shape = 1,aes(size = -log10(PValue), col = sig), stroke =1.5)+
  scale_color_manual(values = c("UP" = "#9f1214","UP_trend" = "#e73032","DOWN" = "#25567d","DOWN_trend" = "#428bc7","NOTSIG" = "snow2"), name = "stats")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 16),
        axis.text.y = element_text(size = 16),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 12),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        legend.location = "plot",
        legend.position = "bottom") +
  guides(color = guide_legend(nrow = 2),  
         shape = guide_legend(nrow = 1))
p

ggsave(filename = paste0(out_folder,"S3D_diff_abundance_smoking.pdf"),
       plot = p,
       device = "pdf",
       width = 10,
       height = 6)
```



# S3E mutation signatures

```{r}
# Merge with mutational signatures
Lung_NSCLC2_filtered <-read.delim(paste0(mount_path,"data/WES/mutation_signatures/Lung_NSCLC2_filtered.tsv"), sep="\t")
names(Lung_NSCLC2_filtered)[names(Lung_NSCLC2_filtered)=="patient"] <- "immucan_id"
anno_WES <- merge(traits, Lung_NSCLC2_filtered, by="immucan_id", sort=F)

# Combine smoking signatures
anno_WES[,"SBS4_SBS92"] <- apply(anno_WES[,c("SBS4","SBS92")],1,sum)

# Dichotomise smoking signature
cutoff <- median(anno_WES$SBS4_SBS92)
anno_WES <- anno_WES %>%
      filter(!is.na(SBS4_SBS92))  %>%
      mutate(SBS4_SBS92_dicho = ifelse(SBS4_SBS92 > cutoff, "high","low"))

hist="ALL"
feature="SBS4_SBS92"
  
data_to_plot <- anno_WES[,c("smoking_status", feature)]
names(data_to_plot) <- c("smoking_status", "feature")
  
data_to_plot <- data_to_plot %>%
  filter(!is.na(smoking_status))

sample_sizes <- data_to_plot %>%
    group_by(smoking_status) %>%
    summarise(n = n()) %>%
    mutate(label = paste0(smoking_status, "\n(n=", n, ")"))

p <- ggplot(data_to_plot, aes(x = smoking_status, y = feature, fill=smoking_status)) +
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(color = smoking_status),position = position_jitterdodge(jitter.width = 0.2), size= 1.5)+
  scale_fill_manual(values = alpha(c("blue", "lightblue", "cyan3"),alpha = 0.5))+
  scale_color_manual(values = c("blue", "lightblue", "cyan3"))+
  scale_x_discrete(labels = sample_sizes$label) + 
  labs(y = "# of mutations",
       title=paste0(feature, " (", hist, ")"),
       subtitle = "Wilcoxon Rank Sum test") +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 16, color = "black"),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 14, color = "black"),
        axis.text.y = element_text(size = 14, color = "black"),
        plot.title = element_text(size = 16, color = "black"),
        plot.subtitle = element_text(size = 14, color = "black"),
        axis.line = element_line(color = "black"),
        axis.ticks = element_line(color = "black"),
        legend.text = element_text(size = 14),
        legend.position = "bottom",
        legend.location = "plot") +
  stat_compare_means(comparisons = list(c("Non-smoker", "Ex-smoker"),
                                        c("Non-smoker", "Current smoker"),
                                        c("Ex-smoker", "Current smoker")),
                     method = "wilcox.test", label = "p.format",size = 5,label.y = c(1200, 1400, 1600))+
  coord_cartesian(ylim = c(0,1800))

p
ggsave(filename = paste0(out_folder,"S3E_mutation_signatures_for_smoking.pdf"),
       plot = p,
       device = "pdf",
       width = 6,
       height = 6)
```

# S3F inflammation signature for smoking

```{r}
cur_dat <- read.table(paste0(mount_path,"data/RNASeq/CURIE/inflammation_quantification_NSCLC2/outputs/pathwayActivity/PathwayActivity_NSCLC2_mlm"))
my_comp <-  list( c("Current smoker", "Non-smoker"), c("Ex-smoker", "Non-smoker"), c("Current smoker", "Non-smoker") )

cur_dat <- cur_dat %>%
  as.data.frame() %>%
  mutate(immucan_id = paste0("LUNG-",rownames(cur_dat))) %>%
  left_join(traits,by = "immucan_id")

cur_dat <- cur_dat %>%
  pivot_longer(cols = c(JAK.STAT,NFkB,TNFa))
p <- cur_dat %>%
  filter(!is.na(smoking_status)) %>%
  ggplot(aes(x = smoking_status,y = value, fill = smoking_status))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(color = smoking_status),position = position_jitterdodge(jitter.width = 0.2), size= 1)+
  facet_wrap(.~name)+
  scale_fill_manual(values = alpha(c("blue", "lightblue", "cyan3"),alpha = 0.5))+
  scale_color_manual(values = c("blue", "lightblue", "cyan3"))+
  stat_compare_means(size = 5,label = "p.signif", comparisons = my_comp)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5,hjust = 1, size = 16),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 14),
        legend.text = element_text(size = 14),
        strip.text.x = element_text(size = 14),
        legend.position = "bottom",
        legend.location = "plot")+
  ylab("Pathway score")

p

ggsave(filename = paste0(out_folder,"S3F_inflammation_score_for_smoking_status.pdf"),
       plot = p,
       device = "pdf",
       width = 7,
       height = 7)
```
