---
title: "Figure 3"
author: "Daniel Schulz"
date: "2024-11-15"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(tidyverse)
library(ggpubr)
library(SingleCellExperiment)
library(survivalAnalysis)
library(ggsurvfit)
library(survminer)
library(survival)
library(edgeR)
library(cowplot)
library(ggrepel)
library(igraph)
library(ggraph)
library(ggrepel)
library(scales)
library(ggnewscale)
library(BiocParallel)
library(imcRtools)
```

```{r}
mount_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/"
out_folder <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/Figure_3/"

if (!dir.exists(out_folder)) {
  dir.create(out_folder)
}
```

# load data
we will load the data we generated at the beginning of figure 2
```{r}
dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/abundances_dat.rds"))
dens_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/densities_dat.rds"))
cello_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/cello_dat.rds"))
clin_dat <- read.csv("/mnt/immucan_volume/clinical_data/Lung_NSCLC2/20230825/NSCLC2_for_R_update_2024.csv")
traits <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/all_traits.csv"))

# variables
data_type <- unique(str_split_i(string = colnames(dat[,-1]),pattern = "__",i = 2))

# colors
stage_col = setNames(c("palevioletred3","darkseagreen3","dodgerblue"),c("I","II","III"))
stage_col_num = setNames(c("palevioletred3","darkseagreen3","dodgerblue"),c("1","2","3"))
hist_col_2 = setNames(c("blue","orange"),c("LUAD","LUSC"))
hist_col_3 = setNames(c("blue","orange", "gray40"),c("LUAD","LUSC","Other"))

```

# 3B enrichment in clinical traits
```{r, fig.height=10, fig.width=6}
cur_traits <- traits %>% select("TPS_score","age_dicho", "gender","stage_1_2","stage_1_3","stage_2_3","LUAD_vs_LUSC","never_vs_current_smoker",
            "never_vs_Ex_smoker","current_vs_Ex_smoker", "N0_vs_Nx","M0_vs_Mx", "long_survivors","T1_vs_Tother")
condition <- c("positive","low", "Female","I","I","II","Adenocarcinoma","Non-smoker","Non-smoker","Ex-smoker", "node_neg", "met_neg", "alive","T1")

names(condition) <- colnames(cur_traits)

p.list <- list()
l.list <- list()
for (i in names(condition)) {
  x <- factor(cur_traits$TPS_score)
  x <- relevel(x, ref = "negative")
  
  y <- factor(unlist(cur_traits[,i]))
  y <- relevel(y, ref = condition[i])

  p.list[[i]] <- fisher.test(table(x, y))
  l.list[[i]] <- c(levels(y)[1])
}

out <- data.frame(feature = names(p.list),
                  p.value = unlist(lapply(p.list, function(x){
                    x$p.value})),
                  odds_ratio = unlist(lapply(p.list, function(x){
                    x$estimate})),
                  reference = unlist(l.list))

out <- out[-1,]

out <- out %>%
  mutate(feature = ifelse(feature == "age_dicho","Old_over_Young", feature),
         feature = ifelse(feature == "gender","Male_over_Female", feature),
         feature = ifelse(feature == "stage_1_2","Stage_II_over_I", feature),
         feature = ifelse(feature == "stage_1_3","Stage_III_over_I", feature),
         feature = ifelse(feature == "stage_2_3","Stage_III_over_II", feature),
         feature = ifelse(feature == "LUAD_vs_LUSC","LUSC_over_LUAD", feature),
         feature = ifelse(feature == "never_vs_current_smoker","Smoker_over_Never-smoker", feature),
         feature = ifelse(feature == "never_vs_Ex_smoker","Ex-smoker_over_Never-smoker", feature),
         feature = ifelse(feature == "current_vs_Ex_smoker","Smoker_over_Ex-smoker", feature),
         feature = ifelse(feature == "N0_vs_Nx","Node-pos_over_Node-neg", feature),
         feature = ifelse(feature == "M0_vs_Mx","Met-pos_over_Met-neg", feature),
         feature = ifelse(feature == "long_survivors","Dead_over_Alive", feature),
         feature = ifelse(feature == "T1_vs_Tother","T1_over_T-other", feature))


out <- out %>%
  mutate(TPS_score = "high",
         FDR = p.adjust(p.value,method = "BH"),
         sig = ifelse(p.value < 0.05 & odds_ratio > 1, "UP_trend",NA),
         sig = ifelse(FDR < 0.1 & odds_ratio > 1, "UP",sig),
         sig = ifelse(p.value < 0.05 & odds_ratio < 1, "DOWN_trend",sig),
         sig = ifelse(FDR < 0.1 & odds_ratio < 1, "DOWN",sig),
         sig = ifelse(is.na(sig),"NOTSIG", sig))

p <- out %>%
  ggplot(aes(x=feature, y = TPS_score,col = odds_ratio, size= -log10(p.value)))+
  #geom_tile(aes(fill = sig))+
  #theme_bw()+
  geom_point()+
  theme_bw()+
  #scale_colour_viridis_c()+
  #theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  #      panel.background = element_blank(), axis.line = element_line(colour = "black"))+
  #theme_bw()+
  scale_color_gradient2(low = "blue",mid = "white",   high = "red",midpoint = 1)+
  scale_size(range = c(1,4))+
  new_scale_colour() +
  geom_point(data = out,shape = 1,aes(size = -log10(p.value), col = sig), stroke =1.5)+
  scale_color_manual(values = c("UP" = "#9f1214","UP_trend" = "#e73032","DOWN" = "#25567d","DOWN_trend" = "#428bc7","NOTSIG" = "snow2"), name = "stats")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 12),
        axis.text.y = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 12),
        axis.title.y = element_blank())+
  coord_flip()
p

ggsave(filename = paste0(out_folder,"3B_TPS_enrichment_in_clinical.pdf"),
       plot = p,
       device = "pdf",
       width = 4.5,
       height = 6)
```



# 3C diff abundance TPS

we will estimate the dispersion from the whole data
```{r, fig.height=20, fig.width=10}
condition <- c("negative")
names(condition) <- c("TPS_score")

# loop over conditions and do two way abundance difference tests
p.list <- list()
for (i in c(data_type,"cello1","cello2")) {
    
    if(grepl("cello",i)){
      abundances <- cello_dat %>%
        select(contains(i))
      abundances <- t(abundances)
      colnames(abundances) <- cello_dat$immucan_id
    } else {
      abundances <- dat %>%
        select(contains(i)) %>%
        filter(complete.cases(.))
      abundances <- t(abundances)
      colnames(abundances) <-dat %>%
        select(contains(i),immucan_id) %>%
        filter(complete.cases(.)) %>%
        pull(immucan_id)
    }
    anno <- traits %>%
      filter(!is.na(.data[["TPS_score"]]))
    
    #new from here
    meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
    meta$histology_TPS <- paste0(meta$simple_histology,"_",meta$TPS_score)
    Size <- factor(meta[,"histology_TPS"])
    meta$Size <- Size
    
    # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
    y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
    design <- model.matrix(~0+group,data = y.ab$samples)
    colnames(design) <- levels(y.ab$samples$group)
    # we do not fit a trend 
    y.ab <- estimateDisp(y.ab, design, trend.method = "none")
    y.ab$common.dispersion
    #plotBCV(y.ab,pch = 5)
    
    # we only compare the diff abundance for Adeno vs Sqam.
    fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
    fit_LUAD <- glmLRT(glmfit = fit.ab,contrast = c(-1,1,0,0,0,0,0,0,0,0,0,0,0))
    fit_LUSC <- glmLRT(glmfit = fit.ab,contrast = c(0,0,0,0,0,0,0,0,0,-1,1,0,0))
    #plotQLDisp(fit.ab, pch = 5)

    res_list <- list(fit_LUAD,fit_LUSC)
    
    cur_list <- lapply(res_list,function(x){
      sig_dat <- x$table
      sig_dat$celltypes <- rownames(sig_dat)
      sig_dat$comparison <- grep("Adeno|Squamous",x$comparison,value = TRUE)
      return(sig_dat)
    })

    p.list[[i]] <- do.call(rbind,cur_list)
    
    
}

cur_dat <- do.call(rbind,p.list)
cur_dat <- cur_dat %>%
  mutate(tech = gsub("_abundance","",str_split_i(celltypes,"__",2)),
         cond = ifelse(grepl("Adeno",comparison),"TPS-high_LUAD","TPS-high_LUSC")) %>%
  group_by(tech) %>%
  mutate(FDR = p.adjust(PValue,method = "BH"),
         sig = ifelse(PValue < 0.05 & logFC > 0, "UP_trend",NA),
         sig = ifelse(FDR < 0.1 & logFC > 0, "UP",sig),
         sig = ifelse(PValue < 0.05 & logFC < 0, "DOWN_trend",sig),
         sig = ifelse(FDR < 0.1 & logFC < 0, "DOWN",sig),
         sig = ifelse(is.na(sig),"NOTSIG", sig),
         ct_name = gsub("_density","",str_split_i(celltypes,"__",1)))
  
to_show <- cur_dat %>%
  filter(sig %in% c("UP","DOWN")) %>%
  pull(celltypes)

cur_dat <- cur_dat %>%
  filter(celltypes %in% to_show) %>%
  mutate(celltypes = gsub("_abundance","",celltypes))

## spatial cluster renaming
cur_dat <- cur_dat %>%
  mutate(celltypes = gsub(pattern = "_cello",replacement = "IMC",x = celltypes))

# get max and min fold change values
max_up <- max(cur_dat$logFC)+1
min_down <- min(cur_dat$logFC)-1

# plot the data
p <- cur_dat %>%
  ggplot(aes(x=celltypes, y = cond,col = logFC, size= -log10(PValue)))+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(round(min_down),0,round(max_up))),
                         guide = "colorbar", limits=c(round(min_down),round(max_up)))+
  scale_size(range = c(1,10))+
  new_scale_colour() +
  geom_point(data = cur_dat,shape = 1,aes(size = -log10(PValue), col = sig), stroke =1)+
  scale_color_manual(values = c("UP" = "#9f1214","UP_trend" = "#e73032","DOWN" = "#25567d","DOWN_trend" = "#428bc7","NOTSIG" = "snow2"), name = "stats")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 16),
        axis.text.y = element_text(size = 16),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 20),
        legend.location = "plot",
        axis.title.x = element_blank(),
        axis.title.y = element_blank())+
  coord_flip()
p

ggsave(filename = paste0(out_folder,"3C_diff_abundance_TPS_LUSC_LUAD_bubble_plot.pdf"),
       plot = p,
       device = "pdf",
       width = 7.5,
       height = 19)
```

# 3G - Cox model for infiltration

```{r}
inf_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/infiltration_harmonized.rds"))

# compare densities of infiltrating, bystanding and ignored from IMC1, IMC2 and mIF
inf_dat <- inf_dat %>%
  group_by(immucan_id,tech) %>%
  mutate(total_CD8 = sum(bystanding,ignored,infiltrating,na.rm = TRUE)/im_size,
         bystanding = round(bystanding/im_size),
         ignored = round(ignored/im_size),
         infiltrating = round(infiltrating/im_size)) %>%
  ungroup()


p1 <- inf_dat %>%
  ggplot()+
  geom_density(aes(x = bystanding, col = tech), lwd = 1)+
  ggtitle("bystanding")
p2 <- inf_dat %>%
  ggplot()+
  geom_density(aes(x = ignored, col = tech), lwd = 1)+
  ggtitle("ignored")
p3 <- inf_dat %>%
  ggplot()+
  geom_density(aes(x = infiltrating, col = tech), lwd = 1)+
  ggtitle("infiltrating")

plot_grid(p1, p2,p3)

# define interactor over bystander ratio
inf_dat <- inf_dat %>%
  group_by(immucan_id, tech) %>%
  mutate(int_over_bystander = (infiltrating+bystanding)/ignored)
  
inf_dat %>%
  ggplot()+
  geom_density(aes(x = int_over_bystander, col = tech), lwd = 1)+
  ggtitle("infiltrating")

inf_dat[(is.infinite(inf_dat$int_over_bystander)),]$int_over_bystander <- NA

# we will average the ratios across technologies
inf_dat <- inf_dat %>%
  group_by(immucan_id) %>%
  mutate(mean_inf_ratio = mean(int_over_bystander,na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(mean_inf_ratio_dicho = ntile(mean_inf_ratio,2),
         mean_inf_ratio_dicho = ifelse(mean_inf_ratio_dicho == 2, "high", "low"))

res <- left_join(inf_dat %>% select(immucan_id,mean_inf_ratio_dicho) %>% distinct(), traits,"immucan_id")

cm_LUSC <- coxph( Surv(time_to_event, censoring) ~  mean_inf_ratio_dicho ,data = as.data.frame(res) %>% filter(simple_histology == "Squamous cell carcinoma"))
cm_LUAD <- coxph( Surv(time_to_event, censoring) ~  mean_inf_ratio_dicho ,data = as.data.frame(res) %>% filter(simple_histology == "Adenocarcinoma"))

dfLUAD <- data.frame(histology = "LUAD",
                     HR = summary(cm_LUAD)$coefficients[,"exp(coef)"],
                     p.value = summary(cm_LUAD)$coefficients[, "Pr(>|z|)"],
                     lower_conf = summary(cm_LUAD)$conf.int[, "lower .95"],
                     upper_conf = summary(cm_LUAD)$conf.int[, "upper .95"]) 
dfLUSC <- data.frame(histology = "LUSC",
                     HR = summary(cm_LUSC)$coefficients[,"exp(coef)"],
                     p.value = summary(cm_LUSC)$coefficients[, "Pr(>|z|)"],
                     lower_conf = summary(cm_LUSC)$conf.int[, "lower .95"],
                     upper_conf = summary(cm_LUSC)$conf.int[, "upper .95"]) 
df1 <- rbind(dfLUAD,dfLUSC)

p1 <- df1 %>%
  ggplot(aes(x = HR,y = histology, col = histology))+
  geom_point(size = 2)+
  geom_errorbar(aes(xmin=lower_conf, xmax = upper_conf),lwd =1,width = 0.3)+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust= 1))+
  #scale_y_discrete(labels = function(x) str_split_i(x,"_",2))+
  scale_x_log10()+
  theme_bw()+
  geom_vline(xintercept = 1,linetype = "dashed")+
  geom_text(aes(label = paste0("p : ",round(p.value,digits = 3))),
            col = "black",
            vjust = -1,
            inherit.aes = TRUE)+
  scale_color_manual(values = c("LUAD" = "blue", "LUSC" = "orange"))+
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        legend.position = "none",
        axis.title.y = element_blank(),
        plot.title = element_text(hjust = 0.5))+
  ggtitle("HR for low infiltration ratio")
  p1

ggsave(filename = paste0(out_folder,"3G_cox_model_infiltration.pdf"),
       plot = p1,
       device = "pdf",
       width = 3,
       height = 2)
```

check if infiltration ratio also works for each individual modality
```{r}
res <- inf_dat %>%
  filter(tech  == "IMC1") %>%
  group_by(immucan_id) %>%
  ungroup() %>%
  mutate(mean_inf_ratio_dicho = ntile(int_over_bystander,2),
         mean_inf_ratio_dicho = ifelse(mean_inf_ratio_dicho == 2, "high", "low"))

res <- left_join(res, traits,"immucan_id")
p1 <- ggforest(coxph( Surv(time_to_event, censoring) ~  mean_inf_ratio_dicho ,data = as.data.frame(res) %>% filter(simple_histology == "Squamous cell carcinoma")))
p2 <- ggforest(coxph( Surv(time_to_event, censoring) ~ mean_inf_ratio_dicho,data = as.data.frame(res) %>% filter(simple_histology == "Adenocarcinoma")))
p <- plot_grid(p1, p2,ncol = 1)
p

# check individual techs IMC2
res <- inf_dat %>%
  filter(tech  == "IMC2") %>%
  group_by(immucan_id) %>%
  ungroup() %>%
  mutate(mean_inf_ratio_dicho = ntile(int_over_bystander,2),
         mean_inf_ratio_dicho = ifelse(mean_inf_ratio_dicho == 2, "high", "low"))

res <- left_join(res, traits,"immucan_id")
p1 <- ggforest(coxph( Surv(time_to_event, censoring) ~  mean_inf_ratio_dicho ,data = as.data.frame(res) %>% filter(simple_histology == "Squamous cell carcinoma")))
p2 <- ggforest(coxph( Surv(time_to_event, censoring) ~ mean_inf_ratio_dicho,data = as.data.frame(res) %>% filter(simple_histology == "Adenocarcinoma")))
p <- plot_grid(p1, p2,ncol = 1)
p

# check individual techs mIF
res <- inf_dat %>%
  filter(tech  == "mIF") %>%
  ungroup() %>%
  mutate(mean_inf_ratio_dicho = ntile(int_over_bystander,2),
         mean_inf_ratio_dicho = ifelse(mean_inf_ratio_dicho == 2, "high", "low"))

res <- left_join(res, traits,"immucan_id")
p1 <- ggforest(coxph( Surv(time_to_event, censoring) ~  mean_inf_ratio_dicho ,data = as.data.frame(res) %>% filter(simple_histology == "Squamous cell carcinoma")))
p2 <- ggforest(coxph( Surv(time_to_event, censoring) ~ mean_inf_ratio_dicho,data = as.data.frame(res) %>% filter(simple_histology == "Adenocarcinoma")))
p <- plot_grid(p1, p2,ncol = 1)
p

# correlation mIF CD8 and TPS
res <- left_join(inf_dat,traits,"immucan_id")
res %>%
  filter(tech == "mIF") %>%
  ggplot(aes(x = TPS, y = total_CD8))+
  geom_point()
```

# 3D correlation of T cell density and TPS

```{r}
inf_dat <- inf_dat %>%
  group_by(immucan_id) %>%
  mutate(mean_total_CD8_density = mean(total_CD8,na.rm =TRUE),
         mean_bystanding = mean(bystanding,na.rm =TRUE),
         mean_infiltrating = mean(infiltrating,na.rm =TRUE),
         mean_ignored = mean(ignored,na.rm =TRUE)) %>%
  select(immucan_id,mean_total_CD8_density, mean_bystanding, mean_ignored, mean_infiltrating,mean_inf_ratio,mean_inf_ratio_dicho) %>%
  distinct()
cur_traits <- left_join(traits, inf_dat, "immucan_id")

# LUAD
cur_dat <- cur_traits %>%
    filter(simple_histology == "Adenocarcinoma")

labs <- cor.test(cur_dat[["TPS"]],cur_dat[["mean_total_CD8_density"]],use = "complete.obs",method = "spearman")
p1 <- cur_traits %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  ggplot(aes(x = log2(TPS),y = log2(.data[["mean_total_CD8_density"]])))+
  geom_point()+
  ggtitle(paste0("cor: ",round(labs$estimate,digits = 3)," p.value = ",formatC(labs$p.value,format = "e",digits = 1)))+
  theme(plot.title = element_text(size=6))+
  theme_bw()

# LUSC
cur_dat <- cur_traits %>%
    filter(simple_histology == "Squamous cell carcinoma")

labs <- cor.test(cur_dat[["TPS"]],cur_dat[["mean_total_CD8_density"]],use = "complete.obs",method = "spearman")
p2 <- cur_traits %>%
  filter(simple_histology == "Squamous cell carcinoma") %>%
  ggplot(aes(x = log2(TPS),y = log2(.data[["mean_total_CD8_density"]])))+
  geom_point()+
  ggtitle(paste0("cor: ",round(labs$estimate,digits = 3)," p.value = ",round(labs$p.value,digits = 2)))+
  theme(plot.title = element_text(size=6))+
  theme_bw()


p <- plot_grid(p1,p2, nrow = 2)

p


# combined plot with colors
p1 <- cur_traits %>%
  mutate(LUAD_vs_LUSC = ifelse(is.na(LUAD_vs_LUSC), "other", LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma","LUAD", LUAD_vs_LUSC),
          LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Squamous cell carcinoma","LUSC", LUAD_vs_LUSC)) %>%
  ggplot(aes(x = log2(TPS),y = log2(mean_total_CD8_density), col = (LUAD_vs_LUSC)))+
  geom_point()+
  geom_smooth(method = "lm",alpha = 0.2, show.legend = FALSE)+
  ggpmisc::stat_poly_eq(
    aes(label = after_stat(paste(rr.label, p.value.label, sep = "*\", \"*"))),
    formula = y ~ x, 
    parse = TRUE,
    size = 6,
    vstep = 0.07)+
  theme_bw()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.x =  element_text(size = 14),
        axis.title.y = element_text(size = 14),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 15))+
  scale_colour_manual(values = hist_col_3)+
  labs(y = expression("Log2 CD8 density [mm"^2*"]"),color = "Histology")+
  coord_cartesian(ylim = c(4,13))+
  guides(color = guide_legend(override.aes = list(size = 5)))

p1 
ggsave(filename = paste0(out_folder,"3D_correlation_T_cell_density_and_TPS_colored.pdf"),
       plot = p1,
       device = "pdf",
       width = 5,
       height = 4)
```

# 3F correlation interaction ratio and TPS

```{r}
# LUAD
cur_dat <- cur_traits %>%
    filter(simple_histology == "Adenocarcinoma")

labs <- cor.test(cur_dat[["TPS"]],cur_dat[["mean_inf_ratio"]],use = "complete.obs",method = "spearman")
p1 <- cur_traits %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  ggplot(aes(x = log2(TPS),y = log2(.data[["mean_inf_ratio"]])))+
  geom_point()+
  ggtitle(paste0("cor: ",round(labs$estimate,digits = 3)," p.value = ",round(labs$p.value,digits = 2)))+
  theme(plot.title = element_text(size=6))+
  theme_bw()

# LUSC
cur_dat <- cur_traits %>%
    filter(simple_histology == "Squamous cell carcinoma")

labs <- cor.test(cur_dat[["TPS"]],cur_dat[["mean_inf_ratio"]],use = "complete.obs",method = "spearman")
p2 <- cur_traits %>%
  filter(simple_histology == "Squamous cell carcinoma") %>%
  ggplot(aes(x = log2(TPS),y = log2(.data[["mean_inf_ratio"]])))+
  geom_point()+
  ggtitle(paste0("cor: ",round(labs$estimate,digits = 3)," p.value = ",round(labs$p.value,digits = 2)))+
  theme(plot.title = element_text(size=6))+
  theme_bw()



p <- plot_grid(p1, p2, nrow = 2)

p

p1 <- cur_traits %>%
  mutate(LUAD_vs_LUSC = ifelse(is.na(LUAD_vs_LUSC), "other", LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma","LUAD", LUAD_vs_LUSC),
          LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Squamous cell carcinoma","LUSC", LUAD_vs_LUSC)) %>%
  ggplot(aes(x = log2(TPS),y = log2(mean_inf_ratio), col = (LUAD_vs_LUSC)))+
  geom_point()+
  geom_smooth(method = "lm",alpha = 0.2, show.legend = FALSE)+
  ggpmisc::stat_poly_eq(
    aes(label = after_stat(paste(rr.label, p.value.label, sep = "*\", \"*"))),
    formula = y ~ x, 
    parse = TRUE,
    size = 6,
    vstep = 0.07)+
  theme_bw()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.x =  element_text(size = 14),
        axis.title.y = element_text(size = 14),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 15))+
  scale_colour_manual(values = hist_col_3)+
  labs(y = "Log2 Infiltration ratio",color = "Histology")+
  guides(color = guide_legend(override.aes = list(size = 5)))

p1 
ggsave(filename = paste0(out_folder,"3F_correlation_interaction_ratio_and_TPS_colored.pdf"),
       plot = p1,
       device = "pdf",
       width = 4,
       height = 4)

```


# 3H
RNA based pathway activity as calculated by Jane Merlevede [here](../../RNA/CURIE/inflammation_quantification_NSCLC2/inflammation_quantification_analysis.Rmd). We will extract the relevant information to prepare the figure here

```{r}
cur_dat <- read.table(paste0(mount_path,"data/RNAseq/CURIE/inflammation_quantification_NSCLC2/outputs/pathwayActivity/PathwayActivity_NSCLC2_mlm"))
p <- cur_dat %>%
  as.data.frame() %>%
  mutate(immucan_id = paste0("LUNG-",rownames(cur_dat))) %>%
  left_join(traits,by = "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(is.na(LUAD_vs_LUSC),"Other",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma","LUAD",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Squamous cell carcinoma","LUSC",LUAD_vs_LUSC)) %>%
  filter(!is.na(TPS_score)) %>%
  mutate(TPS_score = ifelse(TPS_score == "positive", "high","low")) %>%
  ggplot(aes(x = LUAD_vs_LUSC,y = JAK.STAT, fill = TPS_score))+
  geom_boxplot()+
  geom_jitter(aes(color = TPS_score),position = position_jitterdodge(jitter.width = 0.2), size= 1.75, show.legend = FALSE)+
  scale_fill_manual(values = alpha(c("orange", "darkgreen"),alpha = 0.6))+
  scale_color_manual(values = c("orange", "darkgreen"))+
  stat_compare_means(size = 8,label = "p.signif",method = "t.test")+
  theme_bw()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90, vjust = 0.5,hjust = 1, size = 16),
        axis.text.y= element_text(size = 14),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 14),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 16))+
    guides(fill = guide_legend(override.aes = list(size = 5)))+
  coord_cartesian(ylim = c(-2,4.1))
  

p

ggsave(filename = paste0(out_folder,"3H_inflammation_score_across_TPS.pdf"),
       plot = p,
       device = "pdf",
       width = 3,
       height = 6)
```

# 3I enrichment in clinical traits for TMB
```{r, fig.height=5, fig.width=5}
cur_traits <- traits %>% select("TMB_score","age_dicho", "gender","stage_1_2","stage_1_3","stage_2_3","LUAD_vs_LUSC","never_vs_current_smoker",
            "never_vs_Ex_smoker","current_vs_Ex_smoker", "N0_vs_Nx","M0_vs_Mx", "long_survivors","T1_vs_Tother", "TPS_score")
condition <- c("high","low", "Female","I","I","II","Adenocarcinoma","Non-smoker","Non-smoker","Ex-smoker", "node_neg", "met_neg", "alive","T1", "negative")

names(condition) <- colnames(cur_traits)

p.list <- list()
l.list <- list()
for (i in names(condition)) {
  x <- factor(cur_traits$TMB_score)
  x <- relevel(x, ref = "low")
  
  y <- factor(unlist(cur_traits[,i]))
  y <- relevel(y, ref = condition[i])

  p.list[[i]] <- fisher.test(table(x, y))
  l.list[[i]] <- c(levels(y)[1])
}

out <- data.frame(feature = names(p.list),
                  p.value = unlist(lapply(p.list, function(x){
                    x$p.value})),
                  odds_ratio = unlist(lapply(p.list, function(x){
                    x$estimate})),
                  reference = unlist(l.list))

out <- out[-1,]

out <- out %>%
  mutate(feature = ifelse(feature == "TPS_score","TPS-High_over_Low", feature),
         feature = ifelse(feature == "age_dicho","Old_over_Young", feature),
         feature = ifelse(feature == "gender","Male_over_Female", feature),
         feature = ifelse(feature == "stage_1_2","Stage_II_over_I", feature),
         feature = ifelse(feature == "stage_1_3","Stage_III_over_I", feature),
         feature = ifelse(feature == "stage_2_3","Stage_III_over_II", feature),
         feature = ifelse(feature == "LUAD_vs_LUSC","LUSC_over_LUAD", feature),
         feature = ifelse(feature == "never_vs_current_smoker","Smoker_over_Never-smoker", feature),
         feature = ifelse(feature == "never_vs_Ex_smoker","Ex-smoker_over_Never-smoker", feature),
         feature = ifelse(feature == "current_vs_Ex_smoker","Smoker_over_Ex-smoker", feature),
         feature = ifelse(feature == "N0_vs_Nx","Node-pos_over_Node-neg", feature),
         feature = ifelse(feature == "M0_vs_Mx","Met-pos_over_Met-neg", feature),
         feature = ifelse(feature == "long_survivors","Dead_over_Alive", feature),
         feature = ifelse(feature == "T1_vs_Tother","T1_over_T-other", feature))


out <- out %>%
  mutate(TMB_score = "high",
         FDR = p.adjust(p.value,method = "BH"),
         sig = ifelse(p.value < 0.05 & odds_ratio > 1, "UP_trend",NA),
         sig = ifelse(FDR < 0.1 & odds_ratio > 1, "UP",sig),
         sig = ifelse(p.value < 0.05 & odds_ratio < 1, "DOWN_trend",sig),
         sig = ifelse(FDR < 0.1 & odds_ratio < 1, "DOWN",sig),
         sig = ifelse(is.na(sig),"NOTSIG", sig))

p <- out %>%
  ggplot(aes(x=feature, y = TMB_score,col = odds_ratio, size= -log10(p.value)))+
  #geom_tile(aes(fill = sig))+
  #theme_bw()+
  geom_point()+
  theme_bw()+
  #scale_colour_viridis_c()+
  #theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  #      panel.background = element_blank(), axis.line = element_line(colour = "black"))+
  #theme_bw()+
  scale_color_gradient2(low = "blue",mid = "white",   high = "red",midpoint = 1)+
  scale_size(range = c(1,4))+
  new_scale_colour() +
  geom_point(data = out,shape = 1,aes(size = -log10(p.value), col = sig), stroke =1.5)+
  scale_color_manual(values = c("UP" = "#9f1214","UP_trend" = "#e73032","DOWN" = "#25567d","DOWN_trend" = "#428bc7","NOTSIG" = "snow2"), name = "stats")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 10),
        axis.text.y = element_text(size = 10),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10),
        axis.title.y = element_blank())+
  coord_flip()
p

ggsave(filename = paste0(out_folder,"3I_TMB_enrichment_in_clinical.pdf"),
       plot = p,
       device = "pdf",
       width = 4,
       height = 4.9)
```





# 3J bubble plot for TMB with histology

```{r, fig.height=15, fig.width=10}
condition <- c("low")
names(condition) <- c("TMB_score")

# loop over conditions and do two way abundance difference tests
p.list <- list()
for (i in c(data_type,"cello1","cello2")) {
    
    if(grepl("cello",i)){
      abundances <- cello_dat %>%
        select(contains(i))
      abundances <- t(abundances)
      colnames(abundances) <- cello_dat$immucan_id
    } else {
      abundances <- dat %>%
        select(contains(i)) %>%
        filter(complete.cases(.))
      abundances <- t(abundances)
      colnames(abundances) <- dat %>%
        select(contains(i),immucan_id) %>%
        filter(complete.cases(.)) %>%
        pull(immucan_id)
    }
    anno <- traits %>%
      filter(!is.na(.data[["TMB_score"]]))
    
    # remove patients without TMB info from the abundance data
    abundances <- abundances[,colnames(abundances) %in% anno$immucan_id]
    
    # new from here
    meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
    meta$histology_TMB <- paste0(meta$simple_histology,"_",meta$TMB_score)
    Size <- factor(meta[,"histology_TMB"])
    meta$Size <- Size
    
    # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
    y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
    design <- model.matrix(~0+group,data = y.ab$samples)
    colnames(design) <- levels(y.ab$samples$group)
    # we do not fit a trend 
    y.ab <- estimateDisp(y.ab, design, trend.method = "none")
    y.ab$common.dispersion
    #plotBCV(y.ab,pch = 5)
    
    # we only compare the diff abundance for Adeno vs Sqam.
    fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
    fit_LUAD <- glmLRT(glmfit = fit.ab,contrast = c(1,-1,0,0,0,0,0,0,0,0,0))
    fit_LUSC <- glmLRT(glmfit = fit.ab,contrast = c(0,0,0,0,0,0,0,0,1,-1,0))
    #plotQLDisp(fit.ab, pch = 5)

    res_list <- list(fit_LUAD,fit_LUSC)
    
    cur_list <- lapply(res_list,function(x){
      sig_dat <- x$table
      sig_dat$celltypes <- rownames(sig_dat)
      sig_dat$comparison <- grep("Adeno|Squamous",x$comparison,value = TRUE)
      return(sig_dat)
    })

    p.list[[i]] <- do.call(rbind,cur_list)
    
    
}

cur_dat <- do.call(rbind,p.list)
cur_dat <- cur_dat %>%
  mutate(tech = gsub("_abundance","",str_split_i(celltypes,"__",2)),
         cond = ifelse(grepl("Adeno",comparison),"TMB-high_LUAD","TMB-high_LUSC")) %>%
  group_by(tech) %>%
  mutate(FDR = p.adjust(PValue,method = "BH"),
         sig = ifelse(PValue < 0.05 & logFC > 0, "UP_trend",NA),
         sig = ifelse(FDR < 0.1 & logFC > 0, "UP",sig),
         sig = ifelse(PValue < 0.05 & logFC < 0, "DOWN_trend",sig),
         sig = ifelse(FDR < 0.1 & logFC < 0, "DOWN",sig),
         sig = ifelse(is.na(sig),"NOTSIG", sig),
         ct_name = gsub("_density","",str_split_i(celltypes,"__",1)))
  
to_show <- cur_dat %>%
  filter(sig %in% c("UP","DOWN")) %>%
  pull(celltypes)

cur_dat <- cur_dat %>%
  filter(celltypes %in% to_show) %>%
  mutate(celltypes = gsub("_abundance","",celltypes))

## spatial cluster renaming
cur_dat <- cur_dat %>%
  mutate(celltypes = gsub(pattern = "_cello",replacement = "IMC",x = celltypes))

# get max and min fold change values
max_up <- max(cur_dat$logFC)+1
min_down <- min(cur_dat$logFC)-1

# plot the data
p <- cur_dat %>%
  ggplot(aes(x=celltypes, y = cond,col = logFC, size= -log10(PValue)))+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(round(min_down),0,round(max_up))),
                         guide = "colorbar", limits=c(round(min_down),round(max_up)))+
  scale_size(range = c(1,10))+
  new_scale_colour() +
  geom_point(data = cur_dat,shape = 1,aes(size = -log10(PValue), col = sig), stroke =1)+
  scale_color_manual(values = c("UP" = "#9f1214","UP_trend" = "#e73032","DOWN" = "#25567d","DOWN_trend" = "#428bc7","NOTSIG" = "snow2"), name = "stats")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 16),
        axis.text.y = element_text(size = 16),
        legend.text = element_text(size = 16),
        legend.title = element_text(size = 16),
        legend.location = "plot",
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        title = element_text(size = 16))
  

p
ggsave(filename = paste0(out_folder,"3J_diff_abundance_TMB_LUSC_LUAD_bubble_plot.pdf"),
       plot = p,
       device = "pdf",
       width = 15,
       height = 6)
```


sanity checks

```{r}
cello_dat %>%
  select(immucan_id,CN_1__cello2) %>%
  left_join(traits,"immucan_id") %>%
  filter(simple_histology == "Squamous cell carcinoma")%>%
  ggplot(aes(x = TMB_score,y = log10(CN_1__cello2)))+
  geom_boxplot()+
  stat_compare_means()

cello_dat %>%
  select(immucan_id,CN_9__cello2) %>%
  left_join(traits,"immucan_id") %>%
  filter(simple_histology == "Squamous cell carcinoma")%>%
  ggplot(aes(x = TMB_score,y = log10(CN_9__cello2)))+
  geom_boxplot()+
  stat_compare_means()
```

# 3K TMB vs TPS

```{r, fig.height=6, fig.width=15}
table(traits$TMB_score,traits$TPS_score)
traits <- traits %>%
  mutate(TMB_TPS = ifelse(TMB_score == "low" & TPS_score == "positive", "TMBlow_TPShigh",ifelse(TMB_score == "high" & TPS_score == "negative", "TMBhigh_TPSlow", NA))) 
condition <- c("TMBhigh_TPSlow")
names(condition) <- c("TMB_TPS")

# loop over conditions and do two way abundance difference tests
p.list <- list()
for (i in c(data_type,"cello1","cello2")) {
    
    if(grepl("cello",i)){
      abundances <- cello_dat %>%
        select(contains(i))
      abundances <- t(abundances)
      colnames(abundances) <- cello_dat$immucan_id
    } else {
      abundances <- dat %>%
        select(contains(i)) %>%
        filter(complete.cases(.))
      abundances <- t(abundances)
      colnames(abundances) <- dat %>%
        select(contains(i),immucan_id) %>%
        filter(complete.cases(.)) %>%
        pull(immucan_id)
    }
    anno <- traits %>%
      filter(!is.na(.data[["TMB_TPS"]]))
    
    # remove patients without TMB info from the abundance data
    abundances <- abundances[,colnames(abundances) %in% anno$immucan_id]
    
    meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
    meta$histology_TMB_TPS <- paste0(meta$simple_histology,"_",meta$TMB_TPS)
    Size <- factor(meta[,"histology_TMB_TPS"])
    meta$Size <- Size
    
    # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
    y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
    design <- model.matrix(~0+group,data = y.ab$samples)
    colnames(design) <- levels(y.ab$samples$group)
    # we do not fit a trend 
    y.ab <- estimateDisp(y.ab, design, trend.method = "none")
    y.ab$common.dispersion
    #plotBCV(y.ab,pch = 5)
    
    # we only compare the diff abundance for Adeno vs Sqam.
    fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
    fit_LUAD <- glmLRT(glmfit = fit.ab,contrast = c(-1,1,0,0,0,0,0,0))
    fit_LUSC <- glmLRT(glmfit = fit.ab,contrast = c(0,0,0,0,0,-1,1,0))
    #plotQLDisp(fit.ab, pch = 5)

    res_list <- list(fit_LUAD,fit_LUSC)
    
    cur_list <- lapply(res_list,function(x){
      sig_dat <- x$table
      sig_dat$celltypes <- rownames(sig_dat)
      sig_dat$comparison <- grep("Adeno|Squamous",x$comparison,value = TRUE)
      return(sig_dat)
    })

    p.list[[i]] <- do.call(rbind,cur_list)
    
    
}

cur_dat <- do.call(rbind,p.list)
cur_dat <- cur_dat %>%
  mutate(tech = gsub("_abundance","",str_split_i(celltypes,"__",2)),
         cond = ifelse(grepl("Adeno",comparison),"LUAD","LUSC")) %>%
  group_by(tech) %>%
  mutate(FDR = p.adjust(PValue,method = "BH"),
         sig = ifelse(PValue < 0.05 & logFC > 0, "UP_trend",NA),
         sig = ifelse(FDR < 0.1 & logFC > 0, "UP",sig),
         sig = ifelse(PValue < 0.05 & logFC < 0, "DOWN_trend",sig),
         sig = ifelse(FDR < 0.1 & logFC < 0, "DOWN",sig),
         sig = ifelse(is.na(sig),"NOTSIG", sig),
         ct_name = gsub("_density","",str_split_i(celltypes,"__",1)))
  
to_show <- cur_dat %>%
  filter(sig %in% c("UP","DOWN")) %>%
  pull(celltypes)

cur_dat <- cur_dat %>%
  filter(celltypes %in% to_show) %>%
  mutate(celltypes = gsub("_abundance","",celltypes))

## spatial cluster renaming
cur_dat <- cur_dat %>%
  mutate(celltypes = gsub(pattern = "_cello",replacement = "IMC",x = celltypes))

# get max and min fold change values
max_up <- max(cur_dat$logFC)+1
min_down <- min(cur_dat$logFC)-1

# plot the data
p <- cur_dat %>%
  ggplot(aes(x=celltypes, y = cond,col = logFC, size= -log10(PValue)))+
  #geom_tile(aes(fill = sig))+
  #theme_bw()+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(round(min_down),0,round(max_up))),
                         guide = "colorbar", limits=c(round(min_down),round(max_up)))+
  #theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  #      panel.background = element_blank(), axis.line = element_line(colour = "black"))+
  #theme_bw()+
  #scale_color_gradient(low = "white",   high = "red")+
  scale_size(range = c(1,10))+
  new_scale_colour() +
  geom_point(data = cur_dat,shape = 1,aes(size = -log10(PValue), col = sig), stroke =1)+
  scale_color_manual(values = c("UP" = "#9f1214","UP_trend" = "#e73032","DOWN" = "#25567d","DOWN_trend" = "#428bc7","NOTSIG" = "snow2"), name = "stats")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 16),
        axis.text.y = element_text(size = 16),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 14),
        legend.location = "plot",
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        title = element_text(size = 14))+
  ggtitle("Differential abundance for TPS-high/TMB-low over TPS-low/TMB-high")
  

p

ggsave(filename = paste0(out_folder,"3K_diff_abundance_TMB_vs_TPS.pdf"),
       plot = p,
       device = "pdf",
       width = 20,
       height = 5.5)
```


# 3L inflammation signature for TMB

```{r}
cur_dat <- read.table(paste0(mount_path,"data/RNASeq/CURIE/inflammation_quantification_NSCLC2/outputs/pathwayActivity/PathwayActivity_NSCLC2_mlm"))
my_comp <-  list( c("low", "high"))

cur_dat <- cur_dat %>%
  as.data.frame() %>%
  mutate(immucan_id = paste0("LUNG-",rownames(cur_dat))) %>%
  left_join(traits,by = "immucan_id")

cur_dat <- cur_dat %>%
  pivot_longer(cols = c(JAK.STAT,NFkB,TNFa))

p <- cur_dat %>%
  select(-TMB) %>%
  dplyr::rename(TMB = TMB_score) %>%
  filter(!is.na(TMB)) %>%
  ggplot(aes(x = name,y = value, fill = TMB))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(color = TMB),position = position_jitterdodge(jitter.width = 0.2), size= 1)+
  #facet_wrap(.~name)+
  scale_fill_manual(values = alpha(c("blue", "red"),alpha = 0.5))+
  scale_color_manual(values = c("blue", "red"))+
  stat_compare_means(size = 8,label = "p.signif",method = "t.test")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5,hjust = 1, size = 18),
        axis.text.y = element_text(size = 18),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 18),
        legend.text = element_text(size = 18),
        legend.title = element_text(size = 18),
        strip.text.x = element_text(size = 18))+
  ylab("Pathway score")

p
ggsave(filename = paste0(out_folder,"3L_inflammation_score_for_TMB.pdf"),
       plot = p,
       device = "pdf",
       width = 4.5,
       height = 7)
```

not in figures.
testing for TMB and TPS subgroups

```{r, fig.height=6, fig.width=15}
table(traits$TMB_score,traits$TPS_score)
traits <- traits %>%
  mutate(TMB_TPS = ifelse(TMB_score == "high" & TPS_score == "positive", "TMBhigh_TPShigh",ifelse(TMB_score == "low" & TPS_score == "positive", "TMBlow_TPShigh", NA)))
condition <- c("TMBlow_TPShigh")
names(condition) <- c("TMB_TPS")

# loop over conditions and do two way abundance difference tests
p.list <- list()
for (i in c(data_type,"cello1","cello2")) {
    
    if(grepl("cello",i)){
      abundances <- cello_dat %>%
        select(contains(i))
      abundances <- t(abundances)
      colnames(abundances) <- cello_dat$immucan_id
    } else {
      abundances <- dat %>%
        select(contains(i)) %>%
        filter(complete.cases(.))
      abundances <- t(abundances)
      colnames(abundances) <- dat %>%
        select(contains(i),immucan_id) %>%
        filter(complete.cases(.)) %>%
        pull(immucan_id)
    }
    anno <- traits %>%
      filter(!is.na(.data[["TMB_TPS"]]))
    
    # remove patients without TMB info from the abundance data
    abundances <- abundances[,colnames(abundances) %in% anno$immucan_id]
    
    # new from here
    meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
    Size <- factor(meta[,"TMB_TPS"])
    meta$Size <- Size
    
    # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
    y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
    design <- model.matrix(~0+group,data = y.ab$samples)
    colnames(design) <- levels(y.ab$samples$group)
    # we do not fit a trend 
    y.ab <- estimateDisp(y.ab, design, trend.method = "none")
    y.ab$common.dispersion
    #plotBCV(y.ab,pch = 5)
    
    # we only compare the diff abundance for Adeno vs Sqam.
    fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
    fit <- glmLRT(glmfit = fit.ab,contrast = c(1,-1))
    #plotQLDisp(fit.ab, pch = 5)

    sig_dat <- fit$table
    sig_dat$celltypes <- rownames(sig_dat)

    p.list[[i]] <- sig_dat
    
    
}

# define logFC cutoff
cutoff <- 0.7
cur_dat <- do.call(rbind,p.list)
cur_dat <- cur_dat %>%
  mutate(tech = gsub("_abundance","",str_split_i(celltypes,"__",2)),
         cond = "TMBhigh_TPShigh-over-TMBlow_TPShigh") %>%
  group_by(tech) %>%
  mutate(FDR = p.adjust(PValue,method = "BH"),
         sig_overall = ifelse(FDR < 0.1 ,TRUE,FALSE),
         sig = ifelse(FDR < 0.1 & abs(logFC) > cutoff,TRUE,FALSE),
         direction = ifelse(logFC > 0 ,"up", "down"),
         ct_name = gsub("_abundance","",str_split_i(celltypes,"__",1)))
  
to_show <- cur_dat %>%
  filter(sig == TRUE) %>%
  pull(celltypes)

cur_dat <- cur_dat %>%
  filter(celltypes %in% to_show) %>%
  mutate(celltypes = gsub("_abundance","",celltypes))

## spatial cluster renaming
cur_dat <- cur_dat %>%
  mutate(celltypes = gsub(pattern = "_cello",replacement = "IMC",x = celltypes))

# get max and min fold change values
max_up <- max(cur_dat$logFC)+1
min_down <- min(cur_dat$logFC)-1

# plot the data
plot_dat <- cur_dat
p <- plot_dat %>%
  ggplot(aes(x=celltypes, y = cond,col = logFC, size= -log10(PValue)))+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(
    colours = c("blue", "white", "red"), 
    values = rescale(c(round(min_down), 0, round(max_up))),
    limits = c(round(min_down), round(max_up)),
    guide = guide_colorbar(
      title = "logFC",                   
      barwidth = 12,                     
      barheight = 1.5,                   
      title.theme = element_text(size = 16),
      label.theme = element_text(size = 14)
    )
  )+
  scale_size(range = c(1,8))+
  new_scale_colour() +
  geom_point(data = plot_dat %>% filter(sig == TRUE),shape = 1,aes(size = -log10(PValue), col = sig), stroke =2.5)+
  scale_color_manual(values = c("TRUE" = "black","FALSE" = "white"), name = "FDR < 0.1")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 14),
        axis.text.y = element_text(size = 14),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 14),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.position = "bottom",
        plot.title = element_text(hjust = 0.5, size = 14))+
  ggtitle("Differential abundance for TMB-high")
  

p
```


