---
title: "Figure 3"
author: "Daniel Schulz"
date: "2024-11-15"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, message=FALSE}
library(tidyverse)
library(ggpubr)
library(SingleCellExperiment)
library(survivalAnalysis)
library(ggsurvfit)
library(survminer)
library(survival)
library(edgeR)
library(cowplot)
library(ggrepel)
library(igraph)
library(ggraph)
library(ggrepel)
library(scales)
library(ggnewscale)
library(BiocParallel)
library(imcRtools)
```

```{r}
mount_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/"
out_folder <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/Figure_2/"

if (!dir.exists(out_folder)) {
  dir.create(out_folder)
}
```

# load data
we will load the data we generated at the beginning of figure 2
```{r}
dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/abundances_dat.rds"))
dens_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/densities_dat.rds"))
cello_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/cello_dat.rds"))
clin_dat <- read.csv("/mnt/immucan_volume/clinical_data/Lung_NSCLC2/20230825/NSCLC2_for_R_update_2024.csv")
traits <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/all_traits.csv"))

# variables
data_type <- unique(str_split_i(string = colnames(dat[,-1]),pattern = "__",i = 2))
```

# 3B enrichment in clinical traits
```{r, fig.height=10, fig.width=6}
cur_traits <- traits %>% select("TPS_score","age_dicho", "gender","stage_1_2","stage_1_3","stage_2_3","LUAD_vs_LUSC","never_vs_current_smoker",
            "never_vs_Ex_smoker","current_vs_Ex_smoker", "N0_vs_Nx","M0_vs_Mx", "long_survivors","T1_vs_Tother")
condition <- c("positive","low", "Female","I","I","II","Adenocarcinoma","Non-smoker","Non-smoker","Ex-smoker", "node_neg", "met_neg", "alive","T1")

names(condition) <- colnames(cur_traits)

p.list <- list()
l.list <- list()
for (i in names(condition)) {
  x <- factor(cur_traits$TPS_score)
  x <- relevel(x, ref = "negative")
  
  y <- factor(unlist(cur_traits[,i]))
  y <- relevel(y, ref = condition[i])

  p.list[[i]] <- fisher.test(table(x, y))
  l.list[[i]] <- c(levels(y)[1])
}

out <- data.frame(feature = names(p.list),
                  p.value = unlist(lapply(p.list, function(x){
                    x$p.value})),
                  odds_ratio = unlist(lapply(p.list, function(x){
                    x$estimate})),
                  reference = unlist(l.list))

out <- out[-1,]

p <- out %>%
  mutate(TPS_score = "negative",
         FDR = p.adjust(p.value,method = "BH"),
         sig = ifelse(FDR < 0.1,TRUE, FALSE)) %>%
  ggplot(aes(x=feature, y = TPS_score,col = odds_ratio, size= -log10(p.value)))+
  #geom_tile(aes(fill = sig))+
  #theme_bw()+
  geom_point()+
  theme_bw()+
  #scale_colour_viridis_c()+
  #theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  #      panel.background = element_blank(), axis.line = element_line(colour = "black"))+
  #theme_bw()+
  scale_color_gradient2(low = "blue",mid = "white",   high = "red",midpoint = 1)+
  scale_size(range = c(1,4))+
  new_scale_colour() +
  geom_point(shape = 1,aes(size = -log10(p.value), col = sig), stroke =1.5)+
  scale_color_manual(values = c("TRUE" = "black","FALSE" = "white"), name = "FDR < 0.1")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 10),
        axis.text.y = element_text(size = 10),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10),
        axis.title.y = element_blank())+
  coord_flip()
p

ggsave(filename = paste0(out_folder,"3B_TPS_enrichment_in_clinical.pdf"),
       plot = p,
       device = "pdf",
       width = 3.5,
       height = 6)
```



# 3C diff abundance TPS

we will estimate the dispersion from the whole data
```{r, fig.height=20, fig.width=10}
condition <- c("negative")
names(condition) <- c("TPS_score")

# loop over conditions and do two way abundance difference tests
p.list <- list()
for (i in c(data_type,"cello1","cello2")) {
    
    if(grepl("cello",i)){
      abundances <- cello_dat %>%
        select(contains(i))
      abundances <- t(abundances)
      colnames(abundances) <- cello_dat$immucan_id
    } else {
      abundances <- dat %>%
        select(contains(i)) %>%
        filter(complete.cases(.))
      abundances <- t(abundances)
      colnames(abundances) <-dat %>%
        select(contains(i),immucan_id) %>%
        filter(complete.cases(.)) %>%
        pull(immucan_id)
    }
    anno <- traits %>%
      filter(!is.na(.data[["TPS_score"]]))
    
    #new from here
    meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
    meta$histology_TPS <- paste0(meta$simple_histology,"_",meta$TPS_score)
    Size <- factor(meta[,"histology_TPS"])
    meta$Size <- Size
    
    # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
    y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
    design <- model.matrix(~0+group,data = y.ab$samples)
    colnames(design) <- levels(y.ab$samples$group)
    # we do not fit a trend 
    y.ab <- estimateDisp(y.ab, design, trend.method = "none")
    y.ab$common.dispersion
    #plotBCV(y.ab,pch = 5)
    
    # we only compare the diff abundance for Adeno vs Sqam.
    fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
    fit_LUAD <- glmLRT(glmfit = fit.ab,contrast = c(-1,1,0,0,0,0,0,0,0,0,0,0,0))
    fit_LUSC <- glmLRT(glmfit = fit.ab,contrast = c(0,0,0,0,0,0,0,0,0,-1,1,0,0))
    #plotQLDisp(fit.ab, pch = 5)

    res_list <- list(fit_LUAD,fit_LUSC)
    
    cur_list <- lapply(res_list,function(x){
      sig_dat <- x$table
      sig_dat$celltypes <- rownames(sig_dat)
      sig_dat$comparison <- grep("Adeno|Squamous",x$comparison,value = TRUE)
      return(sig_dat)
    })

    p.list[[i]] <- do.call(rbind,cur_list)
    
    
}

# define fold change cutoff for plotting
cutoff <- 0.7

cur_dat <- do.call(rbind,p.list)
cur_dat <- cur_dat %>%
  mutate(tech = gsub("_abundance","",str_split_i(celltypes,"__",2)),
         cond = ifelse(grepl("Adeno",comparison),"TPShigh_LUAD","TPShigh_LUSC")) %>%
  group_by(tech) %>%
  mutate(FDR = p.adjust(PValue,method = "BH"),
         sig_overall = ifelse(FDR < 0.1 ,TRUE,FALSE),
         sig = ifelse(FDR < 0.1 & abs(logFC) > cutoff,TRUE,FALSE),
         direction = ifelse(logFC > 0 ,"up", "down"),
         ct_name = gsub("_abundance","",str_split_i(celltypes,"__",1)))
  
to_show <- cur_dat %>%
  filter(sig == TRUE) %>%
  pull(celltypes)

cur_dat <- cur_dat %>%
  filter(celltypes %in% to_show) %>%
  mutate(celltypes = gsub("_abundance","",celltypes))

# get max and min fold change values
max_up <- max(cur_dat$logFC)+1
min_down <- min(cur_dat$logFC)-1

# plot the data
p <- cur_dat %>%
  ggplot(aes(x=celltypes, y = cond,col = logFC, size= -log10(PValue)))+
  #geom_tile(aes(fill = sig))+
  #theme_bw()+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(round(min_down),0,round(max_up))),
                         guide = "colorbar", limits=c(round(min_down),round(max_up)))+
  #theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  #      panel.background = element_blank(), axis.line = element_line(colour = "black"))+
  #theme_bw()+
  #scale_color_gradient(low = "white",   high = "red")+
  scale_size(range = c(1,8))+
  new_scale_colour() +
  geom_point(data = cur_dat %>% filter(sig == TRUE),shape = 1,aes(size = -log10(PValue), col = sig), stroke =1)+
  scale_color_manual(values = c("TRUE" = "black","FALSE" = "white"), name = "FDR < 0.1")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 20),
        axis.text.y = element_text(size = 20),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 20),
        legend.location = "plot",
        axis.title.x = element_blank(),
        axis.title.y = element_blank())+
  coord_flip()
  #theme(plot.margin = unit(c(5, 2.5, 2.5, 2.5), "cm"))
p

ggsave(filename = paste0(out_folder,"3C_diff_abundance_TPS_LUSC_LUAD_bubble_plot.pdf"),
       plot = p,
       device = "pdf",
       width = 7.5,
       height = 19)

```

# plots not in the paper

```{r, fig.height=10, fig.width=13}
condition <- c("negative","negative")
names(condition) <- c("TPS_score","CPS_score")

# loop over conditions and do two way abundance difference tests
l.list <- list()
for (i in c(data_type,"cello1","cello2")) {
  p.list <- list()
  for (j in names(condition)) {
    
    if(grepl("cello",i)){
      abundances <- cello_dat %>%
        select(contains(i))
      abundances <- t(abundances)
      colnames(abundances) <- cello_dat$immucan_id
    } else {
      abundances <- dat %>%
        select(contains(i))
      abundances <- t(abundances)
      colnames(abundances) <- dat$immucan_id
    }
    
    anno <- traits %>%
      filter(!is.na(.data[[j]]))

    abundances <- abundances[,colnames(abundances) %in% anno$immucan_id]
    
    meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
    Size <- factor(meta[,j])
    meta$Size <- relevel(Size,ref =condition[j])
    
    y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
    design <- model.matrix(~Size,data = y.ab$samples)
    
    # we do not fit a trend 
    y.ab <- estimateDisp(y.ab, design, trend.method = "none")
    y.ab$common.dispersion
    
    #plotBCV(y.ab,pch = 5)
    
    
    fit.ab <- glmQLFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
    #plotQLDisp(fit.ab, pch = 5)
    
    res <- glmQLFTest(fit.ab)
    
    sig_dat <- res$table
    sig_dat$celltypes <- rownames(sig_dat)
    sig_dat$cond <- j
    sig_dat$comparison <- gsub("Size","",res$comparison)
    p.list[[j]] <- sig_dat
    
    
  }
  l.list[[i]] <- do.call(rbind,p.list)
}

cur_dat <- do.call(rbind,l.list)
cutoff <- 1

cur_dat <- cur_dat %>%
  # we only look at TPS
  filter(!grepl("CPS",cond))%>%
  mutate(tech = gsub("_abundance","",str_split_i(celltypes,"__",2))) %>%
  group_by(tech) %>%
  mutate(cond = paste0(cond,"_comparison: ",comparison),
         FDR = p.adjust(PValue,method = "BH"),
         sig_overall = ifelse(FDR < 0.1 ,TRUE,FALSE),
         sig = ifelse(FDR < 0.1  & abs(logFC) > cutoff,TRUE,FALSE),
         direction = ifelse(logFC > 0 ,"up", "down"),
         ct_name = gsub("_abundance","",str_split_i(celltypes,"__",1)))
  
p <- cur_dat %>%
  ggplot(aes(x = logFC,y = -log10(PValue), color = tech))+
  geom_point()+
  geom_label_repel(data = cur_dat %>% filter(sig == TRUE),
                  aes(x = logFC, y = -log10(PValue),label = ct_name),
                  max.overlaps = 40,
                  min.segment.length = 0,
                  nudge_x = 0.5,nudge_y = 1,
                  size= 4)+
  scale_color_manual(values = c("IF1" = "blue", "IF2" = "deepskyblue","IF3" = "steelblue","IMC1" ="red", "IMC2" ="darkorange","cello1" = "green4", "cello2" = "magenta3"))+
    theme_bw()+
  coord_cartesian(xlim = c(-1.5,7.5),ylim = c(0,25))+
  geom_vline(xintercept = c(-cutoff, cutoff), lwd = 0.7, col = "gray", linetype = "dashed")
  
p
# ggsave(filename = paste0(out_folder,"3C_diff_abundance_TPS.pdf"),
#        plot = p,
#        device = "pdf",
#        width = 10,
#        height = 8)
```




## TPS differential interactions for TPS

Note: the code below was also run for LUSC and we found no differences!
```{r}
TPSpos_pats <- traits %>%
  filter(TPS_score == "positive") %>%
  pull(immucan_id)
TPSneg_pats <- traits %>%
  filter(TPS_score == "negative") %>%
  pull(immucan_id)
```

### test differences

```{r, fig.height=20, fig.width=30}
out_dat <- lapply(colnames(int_dat)[-1], function(x){
  pred <- int_dat %>% select(all_of(c("immucan_id",x)))
  pred <- pred %>% filter(.data[[x]] != 0,
                          !is.na(.data[[x]]))
  
  TPSpos <- pred %>% filter(immucan_id %in% TPSpos_pats)
  TPSneg <- pred %>% filter(immucan_id %in% TPSneg_pats)
  if(length(TPSpos$immucan_id) > 5 & length(TPSneg$immucan_id) > 5) {
    res <- wilcox.test(TPSpos[,2],TPSneg[,2])
    out <- data.frame(feature = x,
                      p_value = res$p.value,
                      TPSpos_mean = mean(TPSpos[,2]),
                      TPSneg_mean = mean(TPSneg[,2]),
                      TPSpos_median = median(TPSpos[,2]),
                      TPSneg_median = median(TPSneg[,2]),
                      TPSpos_cases = length(TPSpos$immucan_id),
                      TPSneg_cases = length(TPSneg$immucan_id))
    
    return(out)
    
  }
  
})

out <- do.call(rbind,out_dat)
out %>%
  dplyr::arrange(p_value)
out <- out %>% 
  mutate(tech = str_split_i(feature,"__",2)) %>%
  group_by(tech) %>%
  mutate(FDR = p.adjust(p_value,method = "BH"),
         logFC = log2(TPSpos_mean/TPSneg_mean),
         show = ifelse(FDR < 0.1 & abs(logFC) > 0.5,TRUE, FALSE)) %>%
  ungroup()

out <- out %>%
  filter(!tech %in% c("cello1","cello2"))

out %>%
  filter(show == TRUE)
out %>% 
  arrange(p_value)

p2 <- out %>%
  filter(!grepl("__IF2_int", feature)) %>%
  ggplot(aes(x = logFC,y = -log10(FDR), color = tech))+
  geom_point()+
  geom_label_repel(data = out %>% filter(show == TRUE, !grepl("__IF2_int", feature)),
                  aes(x = logFC, y = -log10(FDR),label = feature),
                  max.overlaps = 50,
                  min.segment.length = 0,
                  nudge_x = 0.5,nudge_y = 0.05)+
    theme_bw()+
  coord_cartesian(xlim = c(-3,4),ylim = c(0,5))+
  scale_color_manual(values = c("IF1_int" = "blue", "IF2_int" = "deepskyblue","IF3_int" = "steelblue","IMC1_int" ="red", "IMC2_int" ="darkorange"))

p2
```

### differential graph
```{r, fig.height=30, fig.width=30}
diff_graph <- out %>%
  filter(show == TRUE,
         !grepl("cello",feature)) %>%
  select(feature,logFC,p_value) %>%
  mutate(tech = str_split_i(feature, "__",2),
         A = str_split_i(feature,"\\.",1),
         B = gsub("__.*","",str_split_i(feature,"\\.",2)),
         A = paste0(A,"__", tech),
         B = paste0(B,"__", tech)) %>%
  select(A,B,logFC) %>%
  mutate(direction = ifelse(logFC < 0, "down-regulated","up-regulated"),
         logFC = abs(logFC))

node_dat <- data.frame(nodes = c(diff_graph$A, diff_graph$B))
node_dat <- node_dat %>%
  mutate(tech = str_split_i(nodes,"__",2)) %>%
  distinct()


cur_graph <- graph_from_data_frame(diff_graph,directed = FALSE,vertices = node_dat)
E(cur_graph)$weight <- diff_graph$logFC
E(cur_graph)$weight_category <- as.factor(diff_graph$direction)

set.seed(1254)
ggraph(cur_graph, layout = 'fr') +  # 'fr' is the Fruchterman-Reingold layout
  geom_edge_link(aes(width = weight, color = weight_category)) +  # Edge width is based on weight
  geom_node_point(aes(fill = tech), size = 8, shape = 21) +              # Customize node appearance
  geom_node_text(aes(label = name), vjust = "inward", hjust = "inward") +
  scale_fill_manual(values = c("IF1_int" = "blue", "IF2_int" = "deepskyblue","IF3_int" = "steelblue","IMC1_int" ="red", "IMC2_int" ="darkorange")) +# Add labels for nodes
  theme_void() +    
  #scale_edge_width(range = c(0.1, 5)) +
  ggtitle("mutated over wild type")
```



## Infiltration in TPS
this is a score derived some year ago by Ruben Casanova and used now by a number of people in the bodenmiller lab. CD8 T cells are classified into three categroies based on their number of interactions with Tumor cells.
1. no tumor neighbors: 'ignored' 
2. 1-2 tumor neighbors: 'bystanding'
3. >2 tumor neighbors: 'infiltrating'

## old - to be removed
```{r, fig.height=20, fig.width=20}
#sce <- readRDS("/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/IMC/Rout/data/sce.rds")
sce <- sce[,sce$ROI != 4]
sce <- aggregateNeighbors(sce, 
                          colPairName = "neighborhood", # select appropriate spatial graph
                          aggregate_by = "metadata", 
                          count_by = "celltypes", # cell type level
                          proportions = FALSE, 
                          name = "count_neighbors")

cur_dat <- sce$count_neighbors %>% as.data.frame
rownames(cur_dat) <- sce$cell_id
cur_dat$celltypes <- sce$celltypes
cur_dat$immucan_id <- sce$immucan_id

# Define CD8+ T cells as infiltrating, bystanding, ignored
cur_dat <- cur_dat %>% filter(celltypes == "CD8") %>% 
  mutate(CD8Tcell_type = ifelse(Tumor >= 3, "infiltrating", 
                                ifelse(Tumor < 3 & Tumor > 0, "bystanding", "ignored")))

sce$CD8Tcell_type <- cur_dat$CD8Tcell_type[match(sce$cell_id, rownames(cur_dat))]

cur_sce <- sce[,sce$sample_id %in% unique(sce$sample_id)[c(10,50,100,130,180, 220, 300,400)]]

plotSpatial(
  object = cur_sce[,cur_sce$celltypes %in% c("Tumor", "CD8")], 
  img_id = "sample_id", 
  node_color_by = "CD8Tcell_type", 
  node_shape_by = "celltypes", 
  colPairName = "neighborhood", 
  coords = c("Pos_X", "Pos_Y"),
  draw_edges = TRUE, 
  nodes_first = FALSE, 
  directed = FALSE)+
  scale_color_manual(values = c("navy", "steelblue1", "firebrick"))

```

## CD8 spatial phenotype per patient
We will now derive a per patient score based on the CD8 T cell phenotypes

```{r, fig.height=10, fig.width=8}
# calculate image/tissue area
im_size <- data.frame(immucan_id = sce$immucan_id,
                      sample_id = sce$sample_id,
                      im_size = sce$width_px*sce$height_px/10^6) %>%
  distinct() %>%
  group_by(immucan_id) %>%
  summarise(im_size = sum(im_size))

# generate spatial score and normalize by image/tissue area
out <- cur_dat %>% group_by(immucan_id, CD8Tcell_type) %>% dplyr::count() %>% 
  pivot_wider(id_cols = immucan_id, values_from = n, names_from = CD8Tcell_type, values_fill = 0) %>% 
  left_join(im_size,by = "immucan_id") %>%
  mutate(bystanding = round(bystanding/im_size),
         ignored = round(ignored/im_size),
         infiltrating = round(infiltrating/im_size),
         ratio_inf_ex = bystanding / infiltrating, number_inf_ex = bystanding + infiltrating,
         ratio_inf_ex = ifelse(is.infinite(ratio_inf_ex),max(out$ratio_inf_ex[!is.infinite(out$ratio_inf_ex)]),ratio_inf_ex))

# generate score. NOTE: we will not used the score. it is
out <- out %>%
  mutate(spatial_CD8_phenotype = ifelse(ratio_inf_ex >= 4 | infiltrating <= 20, "excluded", "inflamed")) %>% 
  mutate(spatial_CD8_phenotype = ifelse(number_inf_ex <= 60 | is.na(number_inf_ex), "cold", spatial_CD8_phenotype))
table(out$spatial_CD8_phenotype)
res <- left_join(out, traits,"immucan_id")
ggforest(coxph( Surv(time_to_event, censoring) ~  simple_stage + spatial_CD8_phenotype,data = as.data.frame(res)))


# generate general interaction score based solely on CD8 T cells with or without tumor neighbors
out <- out %>%
  ungroup() %>%
  mutate(int_over_bystander = (bystanding+infiltrating)/ignored,
         int_over_bystander_dicho = as.factor(ntile(int_over_bystander,2)),
         int_over_bystander_mean_split = int_over_bystander>mean(out$int_over_bystander),
         int_over_bystander_mean_split = ifelse(int_over_bystander_mean_split == TRUE, "high", "low"),
         int_over_bystander_dicho = ifelse(int_over_bystander_dicho == 2, "high", "low"))

out <- out %>%
  mutate(int_over_bystander_mean_split = ifelse(number_inf_ex < 50, "cold", int_over_bystander_mean_split))
res <- left_join(out, traits,"immucan_id")

# Survival in LUAD and LUSC 
p1 <- ggforest(coxph( Surv(time_to_event, censoring) ~  int_over_bystander_dicho ,data = as.data.frame(res) %>% filter(simple_histology == "Squamous cell carcinoma")))
p2 <- ggforest(coxph( Surv(time_to_event, censoring) ~ int_over_bystander_dicho,data = as.data.frame(res) %>% filter(simple_histology == "Adenocarcinoma")))
p <- plot_grid(p1, p2,ncol = 1)
p

ggsave(filename = paste0(out_folder,"survival_by_infiltration.pdf"),
       plot = p,
       device = "pdf",
       width = 8,
       height = 5)

# surv_fit <- survfit(Surv(time_to_event, censoring) ~ int_over_bystander_mean_split, data = as.data.frame(res))
# 
# ggsurv <- ggsurvplot(surv_fit, data = as.data.frame(res), risk.table = TRUE, pval = TRUE)
# ggsurv$plot <- ggsurv$plot + 
# theme(legend.position = "none")
# ggsurv

# enrichment of TPS and infiltration
table(res[res$simple_histology=="Adenocarcinoma",]$int_over_bystander_dicho,
      res[res$simple_histology=="Adenocarcinoma",]$TPS_score)
table(res[res$simple_histology=="Squamous cell carcinoma",]$int_over_bystander_dicho,
      res[res$simple_histology=="Squamous cell carcinoma",]$TPS_score)
ftest <- fisher.test(res[res$simple_histology=="Squamous cell carcinoma",]$int_over_bystander_dicho,res[res$simple_histology=="Squamous cell carcinoma",]$TPS_score)
p <- res %>%
  filter(simple_histology == "Squamous cell carcinoma") %>%
  ggplot()+
  geom_bar(aes(x = int_over_bystander_dicho, fill=TPS_score))+
  theme_bw()+
  scale_fill_manual(values = c("negative" = "blue", "positive" = "red"))+
  ggtitle(paste0("Fisher's exact test: ",round(ftest$p.value,3)))
p

ftest <- fisher.test(res[res$simple_histology=="Adenocarcinoma",]$int_over_bystander_dicho,res[res$simple_histology=="Adenocarcinoma",]$TPS_score)
p <- res %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  ggplot()+
  geom_bar(aes(x = int_over_bystander_dicho, fill=TPS_score))+
  theme_bw()+
  scale_fill_manual(values = c("negative" = "blue", "positive" = "red"))+
  ggtitle(paste0("Fisher's exact test: ",round(ftest$p.value,3)))
p

# 
cur_sce <- sce[,!is.na(sce$CD8Tcell_type)]
hist_vec <- clin_dat$LUAD_vs_LUSC
names(hist_vec) <- clin_dat$immucan_id
cur_sce$LUAD_vs_LUSC <- hist_vec[cur_sce$immucan_id]
cur_sce <- cur_sce[,!is.na(cur_sce$LUAD_vs_LUSC)]
mean_sce <- scater::aggregateAcrossCells(cur_sce,ids = DataFrame(CD8Tcell_type = cur_sce$CD8Tcell_type,
                                                                 histology = cur_sce$LUAD_vs_LUSC),
                                         statistics = "mean")
assay(mean_sce,"exprs") <- asinh(counts(mean_sce))
colnames(mean_sce) <- paste0(mean_sce$CD8Tcell_type,"_", mean_sce$histology)
dittoSeq::dittoHeatmap(mean_sce,genes = rownames(mean_sce)[grepl("CD7|CD8|PD1|LAG3|PDL1|Ido1|CD45RO|CD45RA|ICOS|GrzB|VISTA|Ki67|Ecad", rownames(mean_sce))],annot.by = c("CD8Tcell_type","histology"), main= "CD8Tcell_type", heatmap.colors = colorRampPalette(c("dark blue", "white", "dark red"))(100),breaks = seq(-2, 2, length.out = 101), cluster_rows =TRUE)

```


# 3G nfiltration ratio and cox model
T
```{r}
inf_dat <- readRDS("/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/Figures/scores_for_sharing/infiltration_harmonized.rds")

# compare densities of infiltrating, bystanding and ignored from IMC1, IMC2 and mIF
inf_dat <- inf_dat %>%
  group_by(immucan_id,tech) %>%
  mutate(total_CD8 = sum(bystanding,ignored,infiltrating,na.rm = TRUE)/im_size,
         bystanding = round(bystanding/im_size),
         ignored = round(ignored/im_size),
         infiltrating = round(infiltrating/im_size)) %>%
  ungroup()


p1 <- inf_dat %>%
  ggplot()+
  geom_density(aes(x = bystanding, col = tech), lwd = 1)+
  ggtitle("bystanding")
p2 <- inf_dat %>%
  ggplot()+
  geom_density(aes(x = ignored, col = tech), lwd = 1)+
  ggtitle("ignored")
p3 <- inf_dat %>%
  ggplot()+
  geom_density(aes(x = infiltrating, col = tech), lwd = 1)+
  ggtitle("infiltrating")

plot_grid(p1, p2,p3)

# define interactor over bystander ratio
inf_dat <- inf_dat %>%
  group_by(immucan_id, tech) %>%
  mutate(int_over_bystander = (infiltrating+bystanding)/ignored)
  
inf_dat %>%
  ggplot()+
  geom_density(aes(x = int_over_bystander, col = tech), lwd = 1)+
  ggtitle("infiltrating")

inf_dat[(is.infinite(inf_dat$int_over_bystander)),]$int_over_bystander <- NA

# we will average the ratios across technologies
inf_dat <- inf_dat %>%
  group_by(immucan_id) %>%
  mutate(mean_inf_ratio = mean(int_over_bystander,na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(mean_inf_ratio_dicho = ntile(mean_inf_ratio,2),
         mean_inf_ratio_dicho = ifelse(mean_inf_ratio_dicho == 2, "high", "low"))

res <- left_join(inf_dat %>% select(immucan_id,mean_inf_ratio_dicho) %>% distinct(), traits,"immucan_id")
p1 <- ggforest(coxph( Surv(time_to_event, censoring) ~  mean_inf_ratio_dicho ,data = as.data.frame(res) %>% filter(simple_histology == "Squamous cell carcinoma")),refLabel = "")
p2 <- ggforest(coxph( Surv(time_to_event, censoring) ~ mean_inf_ratio_dicho,data = as.data.frame(res) %>% filter(simple_histology == "Adenocarcinoma")),refLabel = "")
p <- plot_grid(p1, p2,ncol = 1)
p


ggsave(filename = paste0(out_folder,"3G_cox_model_infiltration.pdf"),
       plot = p,
       device = "pdf",
       width = 7,
       height = 3.5)
```

check if infiltration ratio also works for each individual modality
```{r}
res <- inf_dat %>%
  filter(tech  == "IMC1") %>%
  group_by(immucan_id) %>%
  ungroup() %>%
  mutate(mean_inf_ratio_dicho = ntile(int_over_bystander,2),
         mean_inf_ratio_dicho = ifelse(mean_inf_ratio_dicho == 2, "high", "low"))

res <- left_join(res, traits,"immucan_id")
p1 <- ggforest(coxph( Surv(time_to_event, censoring) ~  mean_inf_ratio_dicho ,data = as.data.frame(res) %>% filter(simple_histology == "Squamous cell carcinoma")))
p2 <- ggforest(coxph( Surv(time_to_event, censoring) ~ mean_inf_ratio_dicho,data = as.data.frame(res) %>% filter(simple_histology == "Adenocarcinoma")))
p <- plot_grid(p1, p2,ncol = 1)
p

# check individual techs IMC2
res <- inf_dat %>%
  filter(tech  == "IMC2") %>%
  group_by(immucan_id) %>%
  ungroup() %>%
  mutate(mean_inf_ratio_dicho = ntile(int_over_bystander,2),
         mean_inf_ratio_dicho = ifelse(mean_inf_ratio_dicho == 2, "high", "low"))

res <- left_join(res, traits,"immucan_id")
p1 <- ggforest(coxph( Surv(time_to_event, censoring) ~  mean_inf_ratio_dicho ,data = as.data.frame(res) %>% filter(simple_histology == "Squamous cell carcinoma")))
p2 <- ggforest(coxph( Surv(time_to_event, censoring) ~ mean_inf_ratio_dicho,data = as.data.frame(res) %>% filter(simple_histology == "Adenocarcinoma")))
p <- plot_grid(p1, p2,ncol = 1)
p

# check individual techs mIF
res <- inf_dat %>%
  filter(tech  == "mIF") %>%
  ungroup() %>%
  mutate(mean_inf_ratio_dicho = ntile(int_over_bystander,2),
         mean_inf_ratio_dicho = ifelse(mean_inf_ratio_dicho == 2, "high", "low"))

res <- left_join(res, traits,"immucan_id")
p1 <- ggforest(coxph( Surv(time_to_event, censoring) ~  mean_inf_ratio_dicho ,data = as.data.frame(res) %>% filter(simple_histology == "Squamous cell carcinoma")))
p2 <- ggforest(coxph( Surv(time_to_event, censoring) ~ mean_inf_ratio_dicho,data = as.data.frame(res) %>% filter(simple_histology == "Adenocarcinoma")))
p <- plot_grid(p1, p2,ncol = 1)
p

# correlation mIF CD8 and TPS
res <- left_join(inf_dat,traits,"immucan_id")
res %>%
  filter(tech == "mIF") %>%
  ggplot(aes(x = TPS, y = total_CD8))+
  geom_point()
```

# 3D correlation of T cell density and TPS

```{r}
inf_dat <- inf_dat %>%
  group_by(immucan_id) %>%
  mutate(mean_total_CD8_density = mean(total_CD8,na.rm =TRUE),
         mean_bystanding = mean(bystanding,na.rm =TRUE),
         mean_infiltrating = mean(infiltrating,na.rm =TRUE),
         mean_ignored = mean(ignored,na.rm =TRUE)) %>%
  select(immucan_id,mean_total_CD8_density, mean_bystanding, mean_ignored, mean_infiltrating,mean_inf_ratio,mean_inf_ratio_dicho) %>%
  distinct()
cur_traits <- left_join(traits, inf_dat, "immucan_id")

# LUAD
cur_dat <- cur_traits %>%
    filter(simple_histology == "Adenocarcinoma")

labs <- cor.test(cur_dat[["TPS"]],cur_dat[["mean_total_CD8_density"]],use = "complete.obs",method = "spearman")
p1 <- cur_traits %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  ggplot(aes(x = log2(TPS),y = log2(.data[["mean_total_CD8_density"]])))+
  geom_point()+
  ggtitle(paste0("cor: ",round(labs$estimate,digits = 3)," p.value = ",formatC(labs$p.value,format = "e",digits = 1)))+
  theme(plot.title = element_text(size=6))+
  theme_bw()

# LUSC
cur_dat <- cur_traits %>%
    filter(simple_histology == "Squamous cell carcinoma")

labs <- cor.test(cur_dat[["TPS"]],cur_dat[["mean_total_CD8_density"]],use = "complete.obs",method = "spearman")
p2 <- cur_traits %>%
  filter(simple_histology == "Squamous cell carcinoma") %>%
  ggplot(aes(x = log2(TPS),y = log2(.data[["mean_total_CD8_density"]])))+
  geom_point()+
  ggtitle(paste0("cor: ",round(labs$estimate,digits = 3)," p.value = ",round(labs$p.value,digits = 2)))+
  theme(plot.title = element_text(size=6))+
  theme_bw()


p <- plot_grid(p1,p2, nrow = 2)

p


# combined plot with colors
p1 <- cur_traits %>%
  mutate(LUAD_vs_LUSC = ifelse(is.na(LUAD_vs_LUSC), "other", LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma","LUAD", LUAD_vs_LUSC),
          LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Squamous cell carcinoma","LUSC", LUAD_vs_LUSC)) %>%
  ggplot(aes(x = log2(TPS),y = log2(mean_total_CD8_density), col = (LUAD_vs_LUSC)))+
  geom_point()+
  geom_smooth(method = "lm",alpha = 0.2)+
  ggpmisc::stat_poly_eq(
    aes(label = after_stat(paste(rr.label, p.value.label, sep = "*\", \"*"))),
    formula = y ~ x, 
    parse = TRUE,
    size = 6,
    vstep = 0.07)+
  theme_bw()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
         axis.title.x =  element_text(size = 14),
        axis.title.y = element_text(size = 14))+
  scale_colour_manual(values = c("LUAD" = "blue", "LUSC" = "forestgreen","other" = "black"))+
  labs(y = expression("Log2 CD8 density [mm"^2*"]"),color = "Histology")+
  coord_cartesian(ylim = c(4,13))

p1 
ggsave(filename = paste0(out_folder,"3D_correlation_T_cell_density_and_TPS_colored.pdf"),
       plot = p1,
       device = "pdf",
       width = 4,
       height = 4)
```

# 3F correlation interaction ratio and TPS

```{r}
# LUAD
cur_dat <- cur_traits %>%
    filter(simple_histology == "Adenocarcinoma")

labs <- cor.test(cur_dat[["TPS"]],cur_dat[["mean_inf_ratio"]],use = "complete.obs",method = "spearman")
p1 <- cur_traits %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  ggplot(aes(x = log2(TPS),y = log2(.data[["mean_inf_ratio"]])))+
  geom_point()+
  ggtitle(paste0("cor: ",round(labs$estimate,digits = 3)," p.value = ",round(labs$p.value,digits = 2)))+
  theme(plot.title = element_text(size=6))+
  theme_bw()

# LUSC
cur_dat <- cur_traits %>%
    filter(simple_histology == "Squamous cell carcinoma")

labs <- cor.test(cur_dat[["TPS"]],cur_dat[["mean_inf_ratio"]],use = "complete.obs",method = "spearman")
p2 <- cur_traits %>%
  filter(simple_histology == "Squamous cell carcinoma") %>%
  ggplot(aes(x = log2(TPS),y = log2(.data[["mean_inf_ratio"]])))+
  geom_point()+
  ggtitle(paste0("cor: ",round(labs$estimate,digits = 3)," p.value = ",round(labs$p.value,digits = 2)))+
  theme(plot.title = element_text(size=6))+
  theme_bw()



p <- plot_grid(p1, p2, nrow = 2)

p

p1 <- cur_traits %>%
  mutate(LUAD_vs_LUSC = ifelse(is.na(LUAD_vs_LUSC), "other", LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma","LUAD", LUAD_vs_LUSC),
          LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Squamous cell carcinoma","LUSC", LUAD_vs_LUSC)) %>%
  ggplot(aes(x = log2(TPS),y = log2(mean_inf_ratio), col = (LUAD_vs_LUSC)))+
  geom_point()+
  geom_smooth(method = "lm",alpha = 0.2)+
  ggpmisc::stat_poly_eq(
    aes(label = after_stat(paste(rr.label, p.value.label, sep = "*\", \"*"))),
    formula = y ~ x, 
    parse = TRUE,
    size = 6,
    vstep = 0.07)+
  theme_bw()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(size = 14),
        axis.text.y = element_text(size = 14),
         axis.title.x =  element_text(size = 14),
        axis.title.y = element_text(size = 14))+
  scale_colour_manual(values = c("LUAD" = "blue", "LUSC" = "forestgreen","other" = "black"))+
  labs(y = "Log2 Infiltration ratio",color = "Histology")

p1 
ggsave(filename = paste0(out_folder,"3F_correlation_interaction_ratio_and_TPS_colored.pdf"),
       plot = p1,
       device = "pdf",
       width = 4,
       height = 4)

```

# plot not in paper
```{r, fig.height=8, fig.width=10}
mean_exp_1 <- read_csv("/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/Figures/scores_for_sharing/mean_marker_expr_per_celltype_p1.csv")
mean_exp_2 <- read_csv("/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/Figures/scores_for_sharing/mean_marker_expr_per_celltype_p2.csv")

mean_exp_1 %>%
  filter(celltypes == "CD8",
         grepl("PD1|LAG|GrzB|Ki67|CD38|TCF7|CD27|ICOS",name)) %>%
  left_join(traits, by = "immucan_id") %>%
  ggplot(aes(x = name, y = mean_exp,fill= TPS_score))+
  geom_boxplot()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
```



# 3H
RNA based pathway activity as calculated by Jane Merlevede [here](../../RNA/CURIE/inflammation_quantification_NSCLC2/inflammation_quantification_analysis.Rmd). We will extract the relevant information to prepare the figure here

```{r}
cur_dat <- read.table("../../RNA/CURIE/inflammation_quantification_NSCLC2/outputs/pathwayActivity/PathwayActivity_NSCLC2_mlm")
p <- cur_dat %>%
  as.data.frame() %>%
  mutate(immucan_id = paste0("LUNG-",rownames(cur_dat))) %>%
  left_join(traits,by = "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(is.na(LUAD_vs_LUSC),"Other",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma","LUAD",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Squamous cell carcinoma","LUSC",LUAD_vs_LUSC)) %>%
  filter(!is.na(TPS_score)) %>%
  ggplot(aes(x = LUAD_vs_LUSC,y = JAK.STAT, fill = TPS_score))+
  geom_boxplot()+
  geom_jitter(aes(color = TPS_score),position = position_jitterdodge(jitter.width = 0.2), size= 0.3)+
  scale_fill_manual(values = alpha(c("orange", "darkgreen"),alpha = 0.6))+
  scale_color_manual(values = c("orange", "darkgreen"))+
  stat_compare_means(size = 8,label = "p.signif")+
  theme_bw()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 90, vjust = 0.5,hjust = 1, size = 14),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 14),
        legend.text = element_text(size = 16))

p

ggsave(filename = paste0(out_folder,"3H_inflammation_score_across_TPS.pdf"),
       plot = p,
       device = "pdf",
       width = 4,
       height = 8)
```

# 3I enrichment in clinical traits for TMB
```{r, fig.height=5, fig.width=5}
cur_traits <- traits %>% select("TMB_score","age_dicho", "gender","stage_1_2","stage_1_3","stage_2_3","LUAD_vs_LUSC","never_vs_current_smoker",
            "never_vs_Ex_smoker","current_vs_Ex_smoker", "N0_vs_Nx","M0_vs_Mx", "long_survivors","T1_vs_T2b", "TPS_score")
condition <- c("high","low", "Female","I","I","II","Adenocarcinoma","Non-smoker","Non-smoker","Ex-smoker", "node_neg", "met_neg", "alive","T1a", "negative")

names(condition) <- colnames(cur_traits)

p.list <- list()
l.list <- list()
for (i in names(condition)) {
  x <- factor(cur_traits$TMB_score)
  x <- relevel(x, ref = "low")
  
  y <- factor(unlist(cur_traits[,i]))
  y <- relevel(y, ref = condition[i])

  p.list[[i]] <- fisher.test(table(x, y))
  l.list[[i]] <- c(levels(y)[1])
}

out <- data.frame(feature = names(p.list),
                  p.value = unlist(lapply(p.list, function(x){
                    x$p.value})),
                  odds_ratio = unlist(lapply(p.list, function(x){
                    x$estimate})),
                  reference = unlist(l.list))

out <- out[-1,]

p <- out %>%
  mutate(TMB_score = "low",
         FDR = p.adjust(p.value,method = "BH"),
         sig = ifelse(FDR < 0.1,TRUE, FALSE)) %>%
  ggplot(aes(x=feature, y = TMB_score,col = odds_ratio, size= -log10(p.value)))+
  #geom_tile(aes(fill = sig))+
  #theme_bw()+
  geom_point()+
  theme_bw()+
  #scale_colour_viridis_c()+
  #theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  #      panel.background = element_blank(), axis.line = element_line(colour = "black"))+
  #theme_bw()+
  scale_color_gradient2(low = "blue",mid = "white",   high = "red",midpoint = 1)+
  scale_size(range = c(1,4))+
  new_scale_colour() +
  geom_point(shape = 1,aes(size = -log10(p.value), col = sig), stroke =1.5)+
  scale_color_manual(values = c("TRUE" = "black","FALSE" = "white"), name = "FDR < 0.1")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 10),
        axis.text.y = element_text(size = 10),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10),
        axis.title.y = element_blank())+
  coord_flip()
p

ggsave(filename = paste0(out_folder,"3I_TMB_enrichment_in_clinical.pdf"),
       plot = p,
       device = "pdf",
       width = 3.5,
       height = 4)
```





# 3J bubble plot for TMB with histology

```{r, fig.height=15, fig.width=10}
condition <- c("low")
names(condition) <- c("TMB_score")

# loop over conditions and do two way abundance difference tests
p.list <- list()
for (i in c(data_type,"cello1","cello2")) {
    
    if(grepl("cello",i)){
      abundances <- cello_dat %>%
        select(contains(i))
      abundances <- t(abundances)
      colnames(abundances) <- cello_dat$immucan_id
    } else {
      abundances <- dat %>%
        select(contains(i)) %>%
        filter(complete.cases(.))
      abundances <- t(abundances)
      colnames(abundances) <- dat %>%
        select(contains(i),immucan_id) %>%
        filter(complete.cases(.)) %>%
        pull(immucan_id)
    }
    anno <- traits %>%
      filter(!is.na(.data[["TMB_score"]]))
    
    # remove patients without TMB info from the abundance data
    abundances <- abundances[,colnames(abundances) %in% anno$immucan_id]
    
    # new from here
    meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
    meta$histology_TMB <- paste0(meta$simple_histology,"_",meta$TMB_score)
    Size <- factor(meta[,"histology_TMB"])
    meta$Size <- Size
    
    # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
    y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
    design <- model.matrix(~0+group,data = y.ab$samples)
    colnames(design) <- levels(y.ab$samples$group)
    # we do not fit a trend 
    y.ab <- estimateDisp(y.ab, design, trend.method = "none")
    y.ab$common.dispersion
    #plotBCV(y.ab,pch = 5)
    
    # we only compare the diff abundance for Adeno vs Sqam.
    fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
    fit_LUAD <- glmLRT(glmfit = fit.ab,contrast = c(1,-1,0,0,0,0,0,0,0,0,0))
    fit_LUSC <- glmLRT(glmfit = fit.ab,contrast = c(0,0,0,0,0,0,0,0,1,-1,0))
    #plotQLDisp(fit.ab, pch = 5)

    res_list <- list(fit_LUAD,fit_LUSC)
    
    cur_list <- lapply(res_list,function(x){
      sig_dat <- x$table
      sig_dat$celltypes <- rownames(sig_dat)
      sig_dat$comparison <- grep("Adeno|Squamous",x$comparison,value = TRUE)
      return(sig_dat)
    })

    p.list[[i]] <- do.call(rbind,cur_list)
    
    
}

# define logFC cutoff
cutoff <- 0.7
cur_dat <- do.call(rbind,p.list)
cur_dat <- cur_dat %>%
  mutate(tech = gsub("_abundance","",str_split_i(celltypes,"__",2)),
         cond = ifelse(grepl("Adeno",comparison),"TMBhigh_LUAD","TMBhigh_LUSC")) %>%
  group_by(tech) %>%
  mutate(FDR = p.adjust(PValue,method = "BH"),
         sig_overall = ifelse(FDR < 0.1 ,TRUE,FALSE),
         sig = ifelse(FDR < 0.1 & abs(logFC) > cutoff,TRUE,FALSE),
         direction = ifelse(logFC > 0 ,"up", "down"),
         ct_name = gsub("_abundance","",str_split_i(celltypes,"__",1)))
  
to_show <- cur_dat %>%
  filter(sig == TRUE) %>%
  pull(celltypes)

cur_dat <- cur_dat %>%
  filter(celltypes %in% to_show) %>%
  mutate(celltypes = gsub("_abundance","",celltypes))

# get max and min fold change values
max_up <- max(cur_dat$logFC)+1
min_down <- min(cur_dat$logFC)-1

# plot the data
p <- cur_dat %>%
  ggplot(aes(x=celltypes, y = cond,col = logFC, size= -log10(PValue)))+
  #geom_tile(aes(fill = sig))+
  #theme_bw()+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(round(min_down),0,round(max_up))),
                         guide = "colorbar", limits=c(round(min_down),round(max_up)))+
  #theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  #      panel.background = element_blank(), axis.line = element_line(colour = "black"))+
  #theme_bw()+
  #scale_color_gradient(low = "white",   high = "red")+
  scale_size(range = c(1,8))+
  new_scale_colour() +
  geom_point(data = cur_dat %>% filter(sig == TRUE),shape = 1,aes(size = -log10(PValue), col = sig), stroke =2.5)+
  scale_color_manual(values = c("TRUE" = "black","FALSE" = "white"), name = "FDR < 0.1")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 14),
        axis.text.y = element_text(size = 10),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10),
        legend.location = "plot",
        axis.title.x = element_blank(),
        axis.title.y = element_blank())

p
ggsave(filename = paste0(out_folder,"3J_diff_abundance_TMB_LUSC_LUAD_bubble_plot.pdf"),
       plot = p,
       device = "pdf",
       width = 15,
       height = 5)
```


