---
title: "Figure 3"
author: "Daniel Schulz"
date: "2024-11-15"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggpubr)
library(SingleCellExperiment)
library(survivalAnalysis)
library(ggsurvfit)
library(survminer)
library(survival)
library(edgeR)
library(cowplot)
library(ggrepel)
library(igraph)
library(ggraph)
library(ggrepel)
library(scales)
library(ggnewscale)
```

```{r}
mount_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/"
out_folder <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/Figure_3_supp/"
old_mount <- "/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/"
if (!dir.exists(out_folder)) {
  dir.create(out_folder)
}
```

# load data
we will load the data we generated at the beginning of figure 2
```{r}
dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/abundances_dat.rds"))
dens_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/densities_dat.rds"))
cello_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/cello_dat.rds"))
clin_dat <- read.csv("/mnt/immucan_volume/clinical_data/Lung_NSCLC2/20230825/NSCLC2_for_R_update_2024.csv")
traits <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/all_traits.csv"))

# variables
data_type <- unique(str_split_i(string = colnames(dat[,-1]),pattern = "__",i = 2))
```






# TMB & TPS groups - not in paper

```{r}
cur_traits <- traits %>%
  filter(! is.na(TPS_score),
         ! is.na(TMB_score)) %>%
  mutate(TMB_TPS = paste0(TMB_score,"-",TPS_score)) %>%
  filter(simple_histology != "unknown") %>% 
  as.data.frame()


ggforest(coxph( Surv(time_to_event, censoring) ~ simple_stage + simple_histology + TPS_score,
                data = cur_traits))
ggforest(coxph( Surv(time_to_event, censoring) ~ simple_stage + simple_histology + TMB_score,
                data = cur_traits))

ggforest(coxph( Surv(time_to_event, censoring) ~ simple_stage  + TPS_score,
                data = cur_traits %>% filter(simple_histology == "Adenocarcinoma")))
ggforest(coxph( Surv(time_to_event, censoring) ~ simple_stage  + TMB_score,
                data = cur_traits %>% filter(simple_histology == "Adenocarcinoma")))

ggforest(coxph( Surv(time_to_event, censoring) ~ simple_stage  + TMB_TPS,
                data = cur_traits))


```



# B cell patches

## caculate B cell patch summaries
note: the code is not evaluated automatically as the output is saved. we will keep these files also at their original location to avoid duplicated data
```{r, fig.height=20, fig.width=20, eval=FALSE}
library(BiocParallel)
cur_files <- list.files(paste0(old_mount"IF/IF1/tsv"), 
                        pattern = ".rds$", full.names = TRUE)

bplapply(cur_files,function(x){
  cur_sce <- readRDS(x)
  
  if (is.numeric(cur_sce$TLS.ID)) {
    dat <- as.data.frame(colData(cur_sce))
    sample_id <- unique(cur_sce$sample_id)

    dat <- dat %>%
      dplyr::select(sample_id,celltype, TLS.ID,cell.ID, in.ROI.tumor_tissue, nucleus.x,nucleus.y, dist_to_tumor)

    # here we will collect the information of the size of TLS and their
    # location. We will identify the location based on subset of cells in
    # the tumor_tissue
    dat <- dat  %>%
      filter(!is.na(TLS.ID)) %>%
      group_by(TLS.ID, in.ROI.tumor_tissue) %>%
      summarise(size = n(),
                TLS_dist_to_tumor = mean(dist_to_tumor)) %>%
      mutate(tissue_location = "in_tumor",
             tissue_location = ifelse(TLS_dist_to_tumor > -500 & TLS_dist_to_tumor < 500, "boundary",tissue_location),
             tissue_location = ifelse(TLS_dist_to_tumor > 500, "outside_tumor",tissue_location))


    cell_props <- as.data.frame(colData(cur_sce))
    cell_props <- cell_props %>%
      filter(!is.na(TLS.ID)) %>%
      dplyr::select(sample_id,celltype, TLS.ID,cell.ID, in.ROI.tumor_tissue, nucleus.x,nucleus.y) %>%
      group_by(TLS.ID) %>%
      dplyr::count(celltype) %>%
      mutate(total_cellcount = sum(n),
             frac = n/total_cellcount) %>%
      pivot_wider(id_cols = c(TLS.ID),names_from = celltype,values_from = frac, values_fill = 0)


    dat <- left_join(dat,cell_props,"TLS.ID")
    dat$location <- NULL
    dat$sample_id <- sample_id

  }
  # No TLS in sample
  else {
    dat <- data.frame(TLS.ID = NA,
                      size = NA,
                      TLS_dist_to_tumor =NA,
                      tissue_location = NA,
                      B = NA,
                      BnT = NA,
                      DC = NA,
                      `T`= NA,
                      other = NA,
                      Tumor = NA,
                      MacCD163 = NA,
                      Tumor_CD15 = NA,
                      sample_id = unique(cur_sce$sample_id))
                      
  }
  
  write_csv(dat,paste0(old_mount,"IF/Rout/IF1_features/TLS_summary/",unique(dat$sample_id),".csv"))
}, BPPARAM = MulticoreParam(workers = 12))
gc()
```

## calculate compartment based cell counts for tumor and outside only

Here we calculate the number of cells in the normal and tumor tissue

```{r, fig.height=20, fig.width=20, eval=FALSE}
cur_files <- list.files(paste0(old_mount,"IF/IF1/tsv"),
                        pattern = ".rds$", full.names = TRUE)

library(BiocParallel)
if (!file.exists(paste0(mount_path,"Figures/scores_for_sharing/dist_dat_all_cells_inTumor_outside.rds"))) {
    cur_list <- bplapply(cur_files,function(x){
    cur_sce <- readRDS(x)
    
    
    cur_dat <- as.data.frame(colData(cur_sce))
    sample_id <- unique(cur_sce$sample_id)
    rownames(cur_dat) <- NULL
    
    cur_dat <- cur_dat %>%
      dplyr::select(sample_id,celltype, dist_to_tumor) %>%
      mutate(location = ifelse(dist_to_tumor < 500, "in_tumor", "outside")) %>%
      group_by(celltype,location) %>%
      summarise(n = n(), .groups = "drop") %>%
      mutate(sample_id = sample_id)
    
    return(cur_dat)
    
  }, BPPARAM = MulticoreParam(workers = 12))
  gc()
  dist_dat <- do.call(rbind, cur_list)
  saveRDS(dist_dat,paste0(mount_path,"Figures/scores_for_sharing/dist_dat_all_cells_inTumor_outside.rds"))
}

```


# S3A B cell patch image
```{r, fig.height=40, fig.width=40}
cur_files <- list.files(paste0(old_mount,"IF/Rout/IF1_features/TLS_summary"),full.names = TRUE)
cur_list <- list()
for (i in cur_files) {
  cur_file <- read.csv(i, check.names = FALSE)
  cur_list[[i]] <- cur_file %>%
    pivot_longer(cols = -c(colnames(cur_file)[1:4],colnames(cur_file)[ncol(cur_file)]))

}

TLS_summary <- do.call(rbind,cur_list)

TLS_summary %>%
  mutate(immucan_id = str_split_i(sample_id,"-FIXT",1)) %>%
  select(immucan_id,size, TLS.ID) %>%
  distinct() %>%
  group_by(immucan_id) %>%
  filter(!is.na(TLS.ID)) %>%
  dplyr::count(immucan_id) %>%
  arrange(desc(n))

cur_sce_2 <- readRDS(paste0(old_mount,"IF/IF1/tsv/LUNG-NSCLC2-0565-FIXT-01-IF1-01_#_cells_properties_#_12cf4d92f8de6204055505affdf85e90.rds"))
# we will assign all cells within 500 µm of the tumor annotation to tumor
cur_sce_2[,cur_sce_2$dist_to_tumor < 500]$in.ROI.tumor_tissue <- TRUE

cur_sce_2$TLS <- !is.na(cur_sce_2$TLS.ID)
df <- as.data.frame(colData(cur_sce_2))
df <- df %>%
  mutate(TLS = ifelse(celltype %in% c("Tumor","Tumor_CD15"),"Tumor cells","Stroma cells"),
         TLS = ifelse(in.ROI.tumor_tissue == FALSE,"Healthy",TLS),
         TLS = ifelse(!is.na(TLS.ID),"TLS",TLS))
colData(cur_sce_2) <- DataFrame(df)

p <- imcRtools::plotSpatial(cur_sce_2,img_id = "sample_id",coords = c("nucleus.x","nucleus.y"),node_color_by = "TLS",node_size_fix = 0.2)+
  scale_colour_manual(values = c("Healthy" = "gray", "Stroma cells" = "darkred", "Tumor cells" = "magenta", "TLS" = "darkorange1"))
#p2 <- imcRtools::plotSpatial(cur_sce_2,img_id = "sample_id",coords = c("nucleus.x","nucleus.y"),node_color_by = "celltype",node_size_fix = 0.2)+
#  scale_colour_manual(values = col_vec)

p
#p2
ggsave(filename = paste0(out_folder,"S3A_TLS_example.png"),
       plot = p,
       device = "png",
       width = 4000,
       height = 4000,
       units = "px")
```

## patients with TLS 
percentage of patients at least 10 patches of 100 B cells
```{r}
TLS_pats <- TLS_summary %>%
  as.data.frame() %>%
  mutate(immucan_id = str_split_i(sample_id,"-FIXT",1),
         B_count = size*value) %>%
  filter(name == "B") %>%
  select(immucan_id,B_count, TLS.ID) %>%
  distinct() %>%
  filter(B_count > 100) %>%
  dplyr::count(immucan_id) %>%
  filter(n >10) %>%
  ungroup() %>%
  left_join(clin_dat[,c("immucan_id","simple_histology")], by = "immucan_id")

dim(TLS_pats)[1]/length(unique(clin_dat$immucan_id))
df <- data.frame(hist_count = unclass(table(clin_dat$simple_histology)))

# per histology
Adeno_pats <- TLS_pats %>%
  filter(simple_histology == "Adenocarcinoma")
dim(Adeno_pats)[1]/df["Adenocarcinoma",]

LUSC_pats <- TLS_pats %>%
  filter(simple_histology == "Squamous cell carcinoma")
dim(LUSC_pats)[1]/df["Squamous cell carcinoma",]
```



## S3B TLS size and number

```{r}
plot_dat <- TLS_summary %>%
  as.data.frame() %>%
  mutate(immucan_id = str_split_i(sample_id,"-FIXT",1),
         B_count = size*value) %>%
  filter(name == "B") %>%
  select(immucan_id,B_count, TLS.ID) %>%
  distinct() %>%
  group_by(immucan_id) %>%
  summarise(n = n(),
            mean_size = as.integer(median(B_count)))

p <- plot_dat %>%
  filter(!is.na(mean_size)) %>%
  ggplot(aes(x = mean_size,y = n))+
  geom_point(size =0.6)+
  theme_bw()+
  ylab("B cell patch count per patient")+
  xlab("average size of B cell patches")+
  scale_y_log10()+
  scale_x_log10(limits = c(10,350))

p
ggsave(filename = paste0(out_folder,"S3B_TLS_count_and_size.pdf"),
       plot = p,
       device = "pdf",
       width = 2.5,
       height = 2)
```

## S3C TLS abundance in compartments

```{r}
dist_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/dist_dat_all_cells_inTumor_outside.rds"))
dist_dat <- dist_dat %>%
  mutate(immucan_id = str_split_i(sample_id,"-FIXT",1)) %>%
  select(-sample_id)

# check the fraction of B cells across compartments
dist_dat %>%
  group_by(immucan_id, location) %>%
  mutate(total_cells = sum(n),
         frac = n/total_cells) %>%
  ungroup() %>%
  filter(celltype == "B") %>%
  ggplot(aes(x = location,y = frac))+
  geom_boxplot()+
  theme(legend.position = "none")+
  stat_compare_means()

# cound the number of TLS in each compartment
TLS_dat <- TLS_summary %>%
  as.data.frame() %>%
  mutate(immucan_id = str_split_i(sample_id,"-FIXT",1)) %>%
  select(immucan_id,size, TLS.ID,TLS_dist_to_tumor) %>%
  distinct() %>%
 mutate(location = ifelse(TLS_dist_to_tumor > 500, "outside","in_tumor")) %>%
  group_by(immucan_id) %>%
  dplyr::count(location,name = "TLS_count") %>%
  filter(!is.na(location))

# calculate the number of total cells per compartment
cell_count <- dist_dat %>%
  group_by(immucan_id, location) %>%
  mutate(total_cells = sum(n)) %>%
  select(immucan_id,total_cells, location) %>%
  distinct()

# merge data
cur_dat <- left_join(TLS_dat,cell_count,c("immucan_id", "location"))

# plot
p <- cur_dat %>%
  mutate(frac = TLS_count/total_cells) %>%
  ggplot(aes(x = location,y = (frac)))+
  geom_boxplot(outliers = FALSE)+
  theme(legend.position = "none")+
  geom_jitter(width = 0.2, col = "gray40")+
  stat_compare_means()+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = .5, hjust = 1, size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_text(size = 15),
        axis.title.y = element_text(size = 15))+
  scale_y_log10()+
  labs(x = "Location", y = "Frequency of patches among all cells")


p

ggsave(filename = paste0(out_folder,"S3C_TLS_location.pdf"),
       plot = p,
       device = "pdf",
       width = 2.75,
       height = 8)
```

# S3D size of patches by location
check if patches are larger in border
```{r}
p <- TLS_summary %>%
  as.data.frame() %>%
  mutate(immucan_id = str_split_i(sample_id,"-FIXT",1)) %>%
  select(immucan_id,size, TLS.ID,TLS_dist_to_tumor) %>%
  distinct() %>%
  mutate(location = ifelse(TLS_dist_to_tumor > 500, "outside","in_tumor")) %>%
  #group_by(immucan_id, location) %>%
  #summarise(mean_size = mean(size)) %>%
  filter(!is.na(location)) %>%
  ggplot(aes(y = (size),x = location))+
  geom_boxplot()+
  scale_y_log10()+
  stat_compare_means()+
  stat_summary(fun = median, geom = "text", 
               aes(label = round(..y.., 2)), 
               vjust = -0.5, color = "red")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = .5, hjust = 1, size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_text(size = 15),
        axis.title.y = element_text(size = 15))+
  scale_y_log10()+
  labs(x = "Location",y = "Patch size")

p

ggsave(filename = paste0(out_folder,"S3D_patch_size.pdf"),
       plot = p,
       device = "pdf",
       width = 3,
       height = 8)
```

# S3E T and B cell proportions in Patches by location
composition of outside TLS
```{r, fig.height=12, fig.width=10}
p <- TLS_summary %>%
  as.data.frame() %>%
  filter(name %in% c("B","T")) %>%
  mutate(immucan_id = str_split_i(sample_id,"-FIXT",1)) %>%
  mutate(location = ifelse(TLS_dist_to_tumor > 500, "outside","in_tumor")) %>%
  filter(!is.na(location)) %>%
  ggplot(aes(x = location,y = value, fill=location))+
  geom_boxplot(outliers = TRUE)+
  stat_compare_means()+
  facet_wrap(.~name, scales = "free",ncol = 1)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = .5, hjust = 1, size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_text(size = 15),
        axis.title.y = element_text(size = 15))+
  scale_fill_manual(values = c("in_tumor" = "magenta","outside" = "gray40"))

p

ggsave(filename = paste0(out_folder,"S3E_B_and_T_cell_by_location.pdf"),
       plot = p,
       device = "pdf",
       width = 4,
       height = 8)
```

# S3F density of patches by histology and stage

only check inside and outside tumor. with 500 µm expansion of tumor

```{r}
# get the tumor size
IF1 <- read_csv(paste0(mount_path,"data/IF/Rout/IF1_features/IF1_cell_type_densities.csv"))
IF1 <- IF1 %>%
  mutate(immucan_id = str_split_i(sample_id,"-FIXT",1)) %>%
  select(immucan_id,tumor_size)

# count the number of TLS in each compartment
TLS_dat <- TLS_summary %>%
  as.data.frame() %>%
  mutate(immucan_id = str_split_i(sample_id,"-FIXT",1)) %>%
  select(immucan_id,size, TLS.ID,TLS_dist_to_tumor) %>%
  distinct() %>%
 mutate(location = ifelse(TLS_dist_to_tumor > 500, "outside","in_tumor")) %>%
  group_by(immucan_id) %>%
  dplyr::count(location,name = "TLS_count") %>%
  filter(!is.na(location))

# merge with tumor size data
cur_dat <- left_join(TLS_dat,IF1,c("immucan_id"))

# only keep in_tumor information
cur_dat <- cur_dat %>%
  filter(location == "in_tumor")

# merge with clincial data
cur_dat <- left_join(cur_dat, traits,"immucan_id")

my_comp <-  list( c("LUAD", "LUSC"), c("LUAD", "Other"), c("LUSC", "Other") )

# calculate patch density per histology and plot
p <- cur_dat %>%
  mutate(patch_density = TLS_count/tumor_size,
         LUAD_vs_LUSC = ifelse(is.na(LUAD_vs_LUSC),"Other",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma","LUAD",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Squamous cell carcinoma","LUSC",LUAD_vs_LUSC)) %>%
  ggplot(aes(x = LUAD_vs_LUSC,y = (patch_density), fill = LUAD_vs_LUSC))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(color = LUAD_vs_LUSC),position = position_jitterdodge(jitter.width = 0.2), size= 0.8)+
  scale_fill_manual(values = alpha(c("LUAD" ="orange","LUSC"= "darkgreen","Other"= "gray40"),alpha = 0.6))+
  scale_color_manual(values = c("orange", "darkgreen", "gray40"))+
  stat_compare_means(comparisons = my_comp,label = "p.signif")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = .5, hjust = 1, size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_text(size = 15),
        axis.title.y = element_text(size = 15))+
  scale_y_log10()+
  labs(x = "Histology", y = expression("Density of patches in tumor [mm"^2*"]"))


p


ggsave(filename = paste0(out_folder,"S3F_Patch_density_for_histology.pdf"),
       plot = p,
       device = "pdf",
       width = 3.5,
       height = 5)
```



# S3G heatmap of correlations
```{r, fig.height=4, fig.width=8}
# cound the number of TLS in each compartment
TLS_dat <- TLS_summary %>%
  as.data.frame() %>%
  mutate(immucan_id = str_split_i(sample_id,"-FIXT",1)) %>%
  select(immucan_id,size, TLS.ID,TLS_dist_to_tumor) %>%
  distinct() %>%
  mutate(location = ifelse(TLS_dist_to_tumor > 500, "outside","in_tumor")) %>%
  group_by(immucan_id) %>%
  dplyr::count(location,name = "TLS_count") %>%
  filter(!is.na(location))

# merge data
cur_dat <- left_join(TLS_dat,IF1,c("immucan_id"))%>%
  filter(location == "in_tumor") %>%
  #left_join(traits,by= "immucan_id") %>%
  ungroup() %>%
  mutate(Patch_density= TLS_count / tumor_size,
         Patch_density_dicho = as.factor(ntile(Patch_density, 2)),
         Patch_density_group_mean = ifelse(Patch_density > mean(Patch_density),TRUE, FALSE))

y <- cur_dat %>%
  filter(location == "in_tumor") %>%
  left_join(dens_dat %>% pivot_longer(cols = c(-immucan_id)),"immucan_id") %>%
  group_by(name) %>%
  filter(!is.na(name)) %>%
  summarise(correlation = cor.test(Patch_density, value,method= "spearman")$estimate,
            p.value = cor.test(Patch_density, value,method= "spearman")$p.value) %>%
  arrange(desc(correlation))

x <- dens_dat %>%
  pivot_longer(cols = c(-immucan_id, -B_density__IF1)) %>%
  group_by(name) %>%
  summarise(correlation_B_cells = cor.test(value, B_density__IF1,method= "spearman")$estimate,
            p.value_ct = cor.test(value, B_density__IF1,method= "spearman")$p.value) %>%
  arrange(desc(correlation_B_cells))

mat <- left_join(x,y,"name")

# Load necessary libraries
library(ComplexHeatmap)
library(circlize)



# Convert to matrix
mat <- as.data.frame(mat) %>%
  mutate(FDR = p.adjust(p.value,method = "BH"))
  #filter(! name %in% c("CD20_all_density__IF1","BnT_density__IF1"),
  #FDR < 0.1)
rownames(mat) <- mat$name
plot_mat_matrix <- mat %>%
  mutate(diff = correlation- correlation_B_cells) %>%
  select(-p.value_ct,-p.value, -name) %>%
  arrange(desc(correlation)) %>%
  slice_head(n=25) %>%
  arrange(desc(diff)) %>%
  as.matrix()

cor_matrix <- as.matrix(plot_mat_matrix[, c("correlation", "correlation_B_cells")])
diff_matrix <- as.matrix(plot_mat_matrix[, "diff", drop = FALSE])

# Define color scales
col_fun_corr <- colorRamp2(c(-0.75, 0, 0.75), c("blue", "white", "red"))
col_fun_corr2 <- colorRamp2(c(-0.2, 0, 0.15), c("purple", "white", "darkgreen"))

# Create two heatmaps and combine them
ht1 <- Heatmap(cor_matrix, name = "Correlation", col = col_fun_corr,
               cluster_rows = FALSE, cluster_columns = FALSE,
               show_column_dend = FALSE, show_row_dend = TRUE)

ht2 <- Heatmap(diff_matrix, name = "Diff", col =col_fun_corr2,
               cluster_rows = FALSE, cluster_columns = FALSE,
               show_column_dend = FALSE, show_row_dend = FALSE)  # No row dendrogram for second heatmap

#plot
ht1 + ht2

pdf(file = paste0(out_folder,"S3G_heatmap_of_correlations.pdf"),
    width = 5,height = 7)
ht1 + ht2
dev.off()
```

# S3H overlap with PDL1, TMB and stage

```{r}
p1 <- cur_dat %>%
  left_join(traits,"immucan_id") %>%
  mutate(patch_density = TLS_count/tumor_size,
         LUAD_vs_LUSC = ifelse(is.na(LUAD_vs_LUSC),"Other",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma","LUAD",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Squamous cell carcinoma","LUSC",LUAD_vs_LUSC)) %>%
  ggplot(aes(x = log2(TPS),y = log2(Patch_density), col = (LUAD_vs_LUSC)))+
  geom_point()+
  geom_smooth(method = "lm",alpha = 0.2)+
   ggpmisc::stat_poly_eq(
    aes(label = after_stat(p.value.label)),
    formula = y ~ x, 
    parse = TRUE,
    size = 6,
    vstep = 0.07)+
  theme_bw()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
         axis.title.x =  element_text(size = 10),
        axis.title.y = element_text(size = 10))+
  scale_colour_manual(values = c("LUAD" = "blue", "LUSC" = "forestgreen","Other" = "gray40"))+
  labs(y = expression("Log2 B cell patch density [mm"^2*"]"),color = "Histology")+
  coord_cartesian(ylim = c(-7.5,2.5))

p2 <- cur_dat %>%
  left_join(traits,"immucan_id") %>%
  mutate(patch_density = TLS_count/tumor_size,
         LUAD_vs_LUSC = ifelse(is.na(LUAD_vs_LUSC),"Other",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma","LUAD",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Squamous cell carcinoma","LUSC",LUAD_vs_LUSC)) %>%
  ggplot(aes(x = log2(TMB),y = log2(Patch_density), col = (LUAD_vs_LUSC)))+
  geom_point()+
  geom_smooth(method = "lm",alpha = 0.2)+
   ggpmisc::stat_poly_eq(
    aes(label = after_stat(p.value.label)),
    formula = y ~ x, 
    parse = TRUE,
    size = 6,
    vstep = 0.07)+
  theme_bw()+
  theme(legend.position = "bottom",
        axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
         axis.title.x =  element_text(size = 10),
        axis.title.y = element_text(size = 10))+
  scale_colour_manual(values = c("LUAD" = "blue", "LUSC" = "forestgreen","Other" = "gray40"))+
  labs(y = expression("Log2 B cell patch density [mm"^2*"]"),color = "Histology", y = "Log2 TMB")+
  coord_cartesian(ylim = c(-7.5,2.5))
p2

p <- plot_grid(p1,p2, ncol = 2)
p


ggsave(filename = paste0(out_folder,"S3H_Overlap_with_clinical_and_PDL1.pdf"),
       plot = p,
       device = "pdf",
       width = 8,
       height = 4)
```

# S3I stage and patch density

```{r}
my_comp <-  list( c("I", "II"), c("I", "III"), c("II", "III") )
p <- cur_dat %>%
  left_join(traits,"immucan_id") %>%
  mutate(patch_density = TLS_count/tumor_size) %>%
  filter(simple_stage != "IV") %>%
  ggplot(aes(y = patch_density,x = simple_stage, fill = simple_stage))+
  geom_boxplot()+
  scale_fill_manual(values = c("I" = "palevioletred3", "II" = "darkseagreen3", "III" = "darkorange2"))+
  stat_compare_means(comparisons =  my_comp,label = "p.signif")+
  theme_bw()

p
ggsave(filename = paste0(out_folder,"S3I_Patches_stage.pdf"),
       plot = p,
       device = "pdf",
       width = 2.5,
       height = 2.5)
```


# S3J B cell and Patch density survival


```{r}
cur_dat <- left_join(TLS_dat,IF1,c("immucan_id"))%>%
  filter(location == "in_tumor") %>%
  left_join(traits,by= "immucan_id") %>%
  ungroup() %>%
  mutate(Patch_density= TLS_count / tumor_size,
         Patch_density_dicho = as.factor(ntile(Patch_density, 2)),
         Patch_density_group_median = ifelse(Patch_density > median(Patch_density),TRUE, FALSE),
         LUAD_vs_LUSC = ifelse(is.na(LUAD_vs_LUSC),"Other",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma","LUAD",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Squamous cell carcinoma","LUSC",LUAD_vs_LUSC))



p.list <- list()
for(j in c("LUAD", "LUSC", "Other", "All")) {
  if (j == "All") {
    df <- cur_dat %>%
      filter(!is.na(cur_dat[,"Patch_density_dicho"]))
  } else{
        df <- cur_dat %>%
      filter(!is.na(cur_dat[,"Patch_density_dicho"]),
             LUAD_vs_LUSC == j)
  }
 
  surv_ob <- Surv(df$time_to_event,df$censoring)
  p.list[[paste0(j, "_Patch_density_dicho")]] <- coxph( as.formula(paste0("surv_ob ~ + Patch_density_dicho")),
                                                        data = df)
}

df <- data.frame(HR =      unlist(lapply(p.list, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p.list, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "upper .95"]}))) 

df1 <- df %>% 
  mutate(hist = str_split_i(rownames(df),"_",1))
  
df1

p <- df1 %>%
  ggplot(aes(x = HR,y = hist, col = hist))+
  geom_point(size = 2)+
  geom_errorbar(aes(xmin=lower_conf, xmax = upper_conf),lwd =1,width = 0.3)+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust= 1))+
  #scale_y_discrete(labels = function(x) str_split_i(x,"_",2))+
  scale_x_log10()+
  theme_bw()+
  geom_vline(xintercept = 1,linetype = "dashed")+
  geom_text(aes(label = paste0("p : ",round(p.value,digits = 3))),
            col = "black",
            vjust = -1,
            inherit.aes = TRUE)+
  scale_color_manual(values = c("LUAD" = "blue", "LUSC" = "forestgreen", "Other" = "gray40", "All" = "black"),name = "Histology")+
  theme(axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10))+
  labs(x = "HR",y = "Histology")

p

ggsave(filename = paste0(out_folder,"S3j_survival.pdf"),
       plot = p,
       device = "pdf",
       width = 3.5,
       height = 3)
```

