---
title: "Figure_4"
author: "Daniel Schulz"
date: "2024-10-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(tidyverse)
library(ggpubr)
library(SingleCellExperiment)
library(scales)
library(ggnewscale)
library(edgeR)
library(cowplot)
library(igraph)
library(ggraph)
library(ggrepel)
```

```{r}
mount_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/"
out_folder <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/Figure_4/"

if (!dir.exists(out_folder)) {
  dir.create(out_folder)
}
```



# load data
we will load the data we generated at the beginning of figure 2
```{r}
dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/abundances_dat.rds"))
dens_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/densities_dat.rds"))
cello_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/cello_dat.rds"))
clin_dat <- read.csv("/mnt/immucan_volume/clinical_data/Lung_NSCLC2/20230825/NSCLC2_for_R_update_2024.csv")
int_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/interaction_dat.rds"))
traits <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/all_traits.csv"))

# variables
data_type <- unique(str_split_i(string = colnames(dat[,-1]),pattern = "__",i = 2))
```

# 4A Waterfall plot

The code for this plot is here: [here](../../WES/analysis/immucan_nsclc2_wes_tableReport.Rmd)


# 4B Driver mutations adenocarcinoma

```{r, fig.height=15, fig.width=20}
condition <- c(FALSE,FALSE,FALSE, FALSE)
names(condition) <- c("KRAS","EGFR","STK11", "TP53")

# loop over conditions and do two way abundance difference tests
p.list <- list()
for (i in c(data_type,"cello1","cello2")) {
  for (j in names(condition)) {
    
    if(grepl("cello",i)){
      abundances <- cello_dat %>%
        select(contains(i))
      abundances <- t(abundances)
      colnames(abundances) <- cello_dat$immucan_id
    } else {
      abundances <- dat %>%
        select(contains(i)) %>%
        filter(complete.cases(.))
      abundances <- t(abundances)
      colnames(abundances) <-dat %>%
        select(contains(i),immucan_id) %>%
        filter(complete.cases(.)) %>%
        pull(immucan_id)
    }
    anno <- traits
    
    meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
    meta$histology_cond <- paste0(meta$simple_histology,"_",meta[,j])
    Size <- factor(meta[,"histology_cond"])
    meta$Size <- Size
    
    # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
    y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
    design <- model.matrix(~group,data = y.ab$samples)
    colnames(design) <- levels(y.ab$samples$group)
    # we do not fit a trend 
    y.ab <- estimateDisp(y.ab, design, trend.method = "none")
    y.ab$common.dispersion
    #plotBCV(y.ab,pch = 5)
    
    # subset design matrix for Adenocarcinoma
    design <- design[,c("Adenocarcinoma_FALSE","Adenocarcinoma_TRUE")]
    
    fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
    fit_LUAD <- glmLRT(glmfit = fit.ab,contrast = c(-1,1))

    res_list <-list(fit_LUAD)
    cur_list <- lapply(as.list(res_list),function(x){
      sig_dat <- x$table
      sig_dat$celltypes <- rownames(sig_dat)
      sig_dat$cond <- j
      sig_dat$comparison <- grep("Adeno|Squamous",x$comparison,value = TRUE)
      return(sig_dat)
    })

    p.list[[paste0(i,"_",j)]] <- do.call(rbind,cur_list)
    
  }
}

# define fold change cutoff for plotting
cutoff <- 0

cur_dat <- do.call(rbind,p.list)
cur_dat <- cur_dat %>%
  mutate(tech = gsub("_abundance","",str_split_i(celltypes,"__",2)),
         cond = paste0(cond, "_mut")) %>%
  group_by(tech, cond) %>%
  mutate(FDR = p.adjust(PValue,method = "BH"),
         sig_overall = ifelse(FDR < 0.1 ,TRUE,FALSE),
         sig = ifelse(FDR < 0.1 & abs(logFC) > cutoff,TRUE,FALSE),
         direction = ifelse(logFC > 0 ,"up", "down"),
         ct_name = gsub("_abundance","",str_split_i(celltypes,"__",1)))
  
cur_dat_LUAD <- cur_dat 
to_show_LUAD <- cur_dat %>%
  filter(sig == TRUE) %>%
  pull(celltypes)

cur_dat <- cur_dat %>%
  filter(celltypes %in% to_show_LUAD) %>%
  mutate(celltypes = gsub("_abundance","",celltypes))

# get max and min fold change values
max_up <- max(cur_dat$logFC)+1
min_down <- min(cur_dat$logFC)-1

# plot the data
p <- cur_dat %>%
  ggplot(aes(x=celltypes, y = cond,col = logFC, size= -log10(PValue)))+
  #geom_tile(aes(fill = sig))+
  #theme_bw()+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(round(min_down),0,round(max_up))),
                         guide = "colorbar", limits=c(round(min_down),round(max_up)))+
  #theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  #      panel.background = element_blank(), axis.line = element_line(colour = "black"))+
  #theme_bw()+
  #scale_color_gradient(low = "white",   high = "red")+
  scale_size(range = c(1,8))+
  new_scale_colour() +
  geom_point(data = cur_dat %>% filter(sig == TRUE),shape = 1,aes(size = -log10(PValue), col = sig), stroke =2.5)+
  scale_color_manual(values = c("TRUE" = "black","FALSE" = "white"), name = "FDR < 0.1")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 20),
        xis.text.y = element_text(size = 20),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 20),
        legend.location = "plot",
        axis.title.x = element_blank(),
        axis.title.y = element_blank())+
  coord_flip()
  #theme(plot.margin = unit(c(5, 2.5, 2.5, 2.5), "cm"))
p
```

# Wilcoxon test for EGFR - not in paper
here we see that actually the differential abundance test does not capture the difference in HLADR+ tumor cells while there is a strong difference based on a Wilcoxon or t-test.
```{r}
dat %>%
  left_join(traits,"immucan_id") %>%
  filter(LUAD_vs_LUSC == "Adenocarcinoma") %>%
  ggplot(aes(x = EGFR,y = log2(Tumor_HLADR_abundance__IMC1)))+
  geom_boxplot()+
  stat_compare_means(method = "t.test")
dens_dat %>%
  left_join(traits,"immucan_id") %>%
  filter(LUAD_vs_LUSC == "Adenocarcinoma") %>%
  ggplot(aes(x = EGFR,y = log2(Tumor_HLADR_density__IMC1)))+
  geom_boxplot()+
  stat_compare_means(method = "t.test")

# we run systematic wilcoxon tests for all markers for EGFR
dat %>%
  select(contains("immucan_id"), contains("IMC1")) %>%
  pivot_longer(cols = -immucan_id) %>%
  left_join(traits,"immucan_id") %>%
  filter(LUAD_vs_LUSC == "Adenocarcinoma") %>%
  group_by(name) %>%
  nest() %>%
  mutate(test = map(data, ~ wilcox.test(value ~ EGFR, data = .x)),
         summary = map(test, broom::tidy))%>%
  select(name, summary) %>%
  unnest(summary)

w_test <- dat %>%
  select(contains("immucan_id"), contains("IMC1")) %>%
  pivot_longer(cols = -immucan_id) %>%
  left_join(traits,"immucan_id") %>%
  filter(LUAD_vs_LUSC == "Adenocarcinoma") %>%
  group_by(name) %>%
  nest() %>%
  mutate(
    # Wilcoxon test
    test = map(data, ~ wilcox.test(value ~ EGFR, data = .x)),
    summary = map(test, broom::tidy),
    
    # Compute medians for fold change
    medians = map(data, ~ .x %>%
                    group_by(EGFR) %>%
                    summarise(median = median(value, na.rm = TRUE), .groups = "drop") %>%
                    pivot_wider(names_from = EGFR, values_from = median, names_prefix = "EGFR_")
                 ),
    
    # Compute fold change: TRUE / FALSE
    fold_change = map_dbl(medians, ~ .x$EGFR_TRUE / .x$EGFR_FALSE)
  ) %>%
  select(name, summary, fold_change) %>%
  unnest(summary)

w_test <- w_test %>%
  mutate(test = "EGFR",
         sig = p.value < 0.05) 
  
w_test %>%
  ggplot(aes(x=name, y = test,col = log2(fold_change), size= -log10(p.value)))+
  #geom_tile(aes(fill = sig))+
  #theme_bw()+
  geom_point()+
  theme_bw()+
  scale_colour_gradient2(low = "blue", mid = "white",high ="red")+
  new_scale_colour() +
  geom_point(data = w_test %>% filter(sig == TRUE),shape = 1,aes(size = -log10(p.value), col = sig), stroke =2.5)+
  scale_color_manual(values = c("TRUE" = "black","FALSE" = "white"))+
coord_flip()
```

to be deleted. here we test if the results are the same if we estimate the dispersion only for LUAD samples
```{r, fig.height=10}
condition <- c(FALSE,FALSE,FALSE, FALSE)
names(condition) <- c("KRAS","EGFR","STK11", "TP53")

# loop over conditions and do two way abundance difference tests
p.list <- list()
i <- "IMC1"
j <- "EGFR"

if(grepl("cello",i)){
  abundances <- cello_dat %>%
    select(contains(i))
  abundances <- t(abundances)
  colnames(abundances) <- cello_dat$immucan_id
} else {
  abundances <- dat %>%
    select(contains(i)) %>%
    filter(complete.cases(.))
  abundances <- t(abundances)
  colnames(abundances) <-dat %>%
    select(contains(i),immucan_id) %>%
    filter(complete.cases(.)) %>%
    pull(immucan_id)
}
anno <- traits

meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
meta$histology_cond <- paste0(meta$simple_histology,"_",meta[,j])
Size <- factor(meta[,"histology_cond"])
meta$Size <- Size

####### test if the results are different if only use the adenocarcinoma and EGFR samples
meta <- meta[meta$histology_cond %in% c("Adenocarcinoma_FALSE","Adenocarcinoma_TRUE"),]
abundances <- abundances[,colnames(abundances) %in% meta$immucan_id]

y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
design <- model.matrix(~0+group,data = y.ab$samples)
colnames(design) <- levels(y.ab$samples$group)
# we do not fit a trend 
y.ab <- estimateDisp(y.ab, design, trend.method = "none")
y.ab$common.dispersion
#plotBCV(y.ab,pch = 5)

cur_mat <- cpm(abundances,prior.count = 1,log = TRUE)

library(ComplexHeatmap)
Heatmap(cur_mat, 
        col = viridis::viridis(100),
        column_split = meta$histology_cond)%v%
  columnAnnotation(histology_cond = anno_simple(meta$histology_cond))

# subset design matrix for Adenocarcinoma
design <- design[,c("Adenocarcinoma_FALSE","Adenocarcinoma_TRUE")]

fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
fit_LUAD <- glmLRT(glmfit = fit.ab,contrast = c(-1,1))

res_list <-list(fit_LUAD)
cur_list <- lapply(as.list(res_list),function(x){
  sig_dat <- x$table
  sig_dat$celltypes <- rownames(sig_dat)
  sig_dat$cond <- j
  sig_dat$comparison <- grep("Adeno|Squamous",x$comparison,value = TRUE)
  return(sig_dat)
})


cur_list
cur_dat <- cur_list[[1]] %>%
  mutate(tech = gsub("_abundance","",str_split_i(celltypes,"__",2)),
         cond = paste0(cond, "_mut")) %>%
  group_by(tech, cond) %>%
  mutate(FDR = p.adjust(PValue,method = "BH"),
         sig_overall = ifelse(FDR < 0.1 ,TRUE,FALSE),
         sig = ifelse(FDR < 0.1 & abs(logFC) > cutoff,TRUE,FALSE),
         direction = ifelse(logFC > 0 ,"up", "down"),
         ct_name = gsub("_abundance","",str_split_i(celltypes,"__",1)))
  
cur_dat_LUAD <- cur_dat 
to_show_LUAD <- cur_dat %>%
  filter(sig == TRUE) %>%
  pull(celltypes)

cur_dat <- cur_dat %>%
  filter(celltypes %in% to_show_LUAD) %>%
  mutate(celltypes = gsub("_abundance","",celltypes))

# get max and min fold change values
max_up <- max(cur_dat$logFC)+1
min_down <- min(cur_dat$logFC)-1

# plot the data
p <- cur_dat %>%
  ggplot(aes(x=celltypes, y = cond,col = logFC, size= -log10(PValue)))+
  #geom_tile(aes(fill = sig))+
  #theme_bw()+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(round(min_down),0,round(max_up))),
                         guide = "colorbar", limits=c(round(min_down),round(max_up)))+
  #theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  #      panel.background = element_blank(), axis.line = element_line(colour = "black"))+
  #theme_bw()+
  #scale_color_gradient(low = "white",   high = "red")+
  scale_size(range = c(1,8))+
  new_scale_colour() +
  geom_point(data = cur_dat %>% filter(sig == TRUE),shape = 1,aes(size = -log10(PValue), col = sig), stroke =2.5)+
  scale_color_manual(values = c("TRUE" = "black","FALSE" = "white"), name = "FDR < 0.1")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 20),
        xis.text.y = element_text(size = 20),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 20),
        legend.location = "plot",
        axis.title.x = element_blank(),
        axis.title.y = element_blank())+
  coord_flip()
  #theme(plot.margin = unit(c(5, 2.5, 2.5, 2.5), "cm"))
p
```


# Driver mutations LUSC

```{r, fig.height=10, fig.width=20}
condition <- c(FALSE,FALSE)
names(condition) <- c("PIK3CA", "TP53")

# loop over conditions and do two way abundance difference tests
p.list <- list()
for (i in c(data_type,"cello1","cello2")) {
  for (j in names(condition)) {
    
    if(grepl("cello",i)){
      abundances <- cello_dat %>%
        select(contains(i))
      abundances <- t(abundances)
      colnames(abundances) <- cello_dat$immucan_id
    } else {
      abundances <- dat %>%
        select(contains(i)) %>%
        filter(complete.cases(.))
      abundances <- t(abundances)
      colnames(abundances) <-dat %>%
        select(contains(i),immucan_id) %>%
        filter(complete.cases(.)) %>%
        pull(immucan_id)
    }
    anno <- traits
    
    meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
    meta$histology_cond <- paste0(meta$simple_histology,"_",meta[,j])
    Size <- factor(meta[,"histology_cond"])
    meta$Size <- Size
    
    # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
    y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
    design <- model.matrix(~0+group,data = y.ab$samples)
    colnames(design) <- levels(y.ab$samples$group)
    # we do not fit a trend 
    y.ab <- estimateDisp(y.ab, design, trend.method = "none")
    y.ab$common.dispersion
    #plotBCV(y.ab,pch = 5)
    
    # subset design matrix  for Adenocarcinoma
    design <- design[,c("Squamous cell carcinoma_FALSE","Squamous cell carcinoma_TRUE")]
    # we only compare the diff abundance for Adeno vs Sqam.
    fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
    fit_LUSC <- glmLRT(glmfit = fit.ab,contrast = c(-1,1))

    res_list <- list( fit_LUSC)
    
    cur_list <- lapply(res_list,function(x){
      sig_dat <- x$table
      sig_dat$celltypes <- rownames(sig_dat)
      sig_dat$cond <- j
      sig_dat$comparison <- grep("Adeno|Squamous",x$comparison,value = TRUE)
      return(sig_dat)
    })

    p.list[[paste0(i,"_",j)]] <- do.call(rbind,cur_list)
    
  }
}

# define fold change cutoff for plotting
cutoff <- 0

cur_dat <- do.call(rbind,p.list)
cur_dat <- cur_dat %>%
  mutate(tech = gsub("_abundance","",str_split_i(celltypes,"__",2)),
         cond = paste0(cond, "_mut")) %>%
  group_by(tech, cond) %>%
  mutate(FDR = p.adjust(PValue,method = "BH"),
         sig_overall = ifelse(FDR < 0.1 ,TRUE,FALSE),
         sig = ifelse(FDR < 0.1 & abs(logFC) > cutoff,TRUE,FALSE),
         direction = ifelse(logFC > 0 ,"up", "down"),
         ct_name = gsub("_abundance","",str_split_i(celltypes,"__",1)))
  
cur_dat_LUSC <- cur_dat

to_show_LUSC <- cur_dat %>%
  filter(sig == TRUE) %>%
  pull(celltypes)

cur_dat <- cur_dat %>%
  filter(celltypes %in% to_show_LUSC) %>%
  mutate(celltypes = gsub("_abundance","",celltypes))

# get max and min fold change values
max_up <- max(cur_dat$logFC)+1
min_down <- min(cur_dat$logFC)-1

# plot the data
p <- cur_dat %>%
  ggplot(aes(x=celltypes, y = cond,col = logFC, size= -log10(PValue)))+
  #geom_tile(aes(fill = sig))+
  #theme_bw()+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(round(min_down),0,round(max_up))),
                         guide = "colorbar", limits=c(round(min_down),round(max_up)))+
  #theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  #      panel.background = element_blank(), axis.line = element_line(colour = "black"))+
  #theme_bw()+
  #scale_color_gradient(low = "white",   high = "red")+
  scale_size(range = c(1,8))+
  new_scale_colour() +
  geom_point(data = cur_dat %>% filter(sig == TRUE),shape = 1,aes(size = -log10(PValue), col = sig), stroke =2.5)+
  scale_color_manual(values = c("TRUE" = "black","FALSE" = "white"), name = "FDR < 0.1")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 20),
        xis.text.y = element_text(size = 20),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 20),
        legend.location = "plot",
        axis.title.x = element_blank(),
        axis.title.y = element_blank())+
  coord_flip()
  #theme(plot.margin = unit(c(5, 2.5, 2.5, 2.5), "cm"))
p

```

## plot  LUSC and LUAD together
```{r, fig.height=15, fig.width=10}
# unify names
cur_dat_LUAD <- cur_dat_LUAD %>% 
  mutate(cond = paste0("LUAD_",cond))
cur_dat_LUSC <- cur_dat_LUSC %>% 
  mutate(cond = paste0("LUSC_",cond))

cur_dat <- rbind(cur_dat_LUAD,cur_dat_LUSC)

cur_dat <- cur_dat %>%
  filter(celltypes %in% c(to_show_LUAD,to_show_LUSC)) %>%
  mutate(celltypes = gsub("_abundance","",celltypes))

max_up <- max(cur_dat$logFC)+1
min_down <- min(cur_dat$logFC)-1

# plot the data
p <- cur_dat %>%
  ggplot(aes(x=(celltypes), y = cond,col = logFC, size= -log10(PValue)))+
  #geom_tile(aes(fill = sig))+
  #theme_bw()+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(round(min_down),0,round(max_up))),
                         guide = "colorbar", limits=c(round(min_down),round(max_up)))+
  #theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  #      panel.background = element_blank(), axis.line = element_line(colour = "black"))+
  #theme_bw()+
  #scale_color_gradient(low = "white",   high = "red")+
  scale_size(range = c(1,3.5))+
  new_scale_colour() +
  geom_point(data = cur_dat %>% filter(sig == TRUE),shape = 1,aes(size = -log10(PValue), col = sig), stroke =1)+
  scale_color_manual(values = c("TRUE" = "black","FALSE" = "white"), name = "FDR < 0.1")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 7),
        axis.text.y = element_text(size = 7),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 7),
        legend.location = "plot",
        axis.title.x = element_blank(),
        axis.title.y = element_blank())+
  coord_flip()

p

ggsave(filename = paste0(out_folder,"4B_diff_abundance_driver_mutations.pdf"),
       plot = p,
       device = "pdf",
       width = 4,
       height = 8)
```



# expression differences across cell types
```{r, fig.height=7, fig.width=15}
mean_exp1 <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/mean_marker_expr_per_celltype_p1.csv"))
mean_exp2 <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/mean_marker_expr_per_celltype_p2.csv"))

# long format
mean_exp1 <- mean_exp1 %>%
  pivot_longer(-immucan_id)
mean_exp2 <- mean_exp2 %>%
  pivot_longer(-immucan_id)

mean_exp1$tech <- "IMC1"
mean_exp2$tech <- "IMC2"
mean_exp <- rbind(mean_exp1,mean_exp2)

mean_exp <- mean_exp %>%
  mutate(celltypes = str_split_i(name,"-",1),
         name = str_split_i(name,"-",2))
# load sce object panel 1
sce1 <- readRDS(paste0(mount_path,"data/IMC/Panel_1/Rout/sce.rds"))
sce1 <- sce1[,sce1$ROI != 4]
# Panel 2
sce2 <- readRDS(paste0(mount_path,"data/IMC/Panel_2/Rout/sce.rds"))
```



```{r}

mean_exp %>%
  filter(celltypes == "CD8",
         grepl("PD1|LAG|GrzB|Ki67|CD38|TCF7|CD27|ICOS|CXCL13|GITR|Tbet|GATA3|NKG2A|Tim3|Ki67|EOMES",name)) %>%
  left_join(traits, by = "immucan_id") %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  ggplot(aes(x = EGFR, y = value,fill= EGFR))+
  geom_boxplot()+
  facet_wrap(.~name, scales= "free")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  stat_compare_means()+
  coord_cartesian(ylim =c(0,3))
```



# 4C HLADR on tumor cells in EGFR mutated

## Tumor cells HLADR in EGFR
```{r, fig.height=5, fig.width=5}
p_Tumor_HLADR_EGFR <- mean_exp %>%
  filter(celltypes == "Tumor",
         name %in% c("HLADR_p1")) %>%
  left_join(traits, by = "immucan_id") %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  ggplot(aes(x = EGFR, y = value,fill= EGFR))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(aes(color = EGFR),position = position_jitterdodge(jitter.width = 0.2), size= 0.3)+
  scale_fill_manual(values = alpha(c("red", "blue"),alpha = 0.6))+
  scale_color_manual(values = c("red", "blue"))+
  facet_wrap(.~name, scales= "free")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  stat_compare_means(size = 1.5,label.y = 4)+
    theme_bw()+
  theme(axis.text.y = element_text(size = 6),
        axis.text.x = element_text(size = 6,angle = 90, vjust = 1, hjust = 0.5),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 6),
        strip.text = element_text(size = 6),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7))+
  theme(legend.position = "none")
p_Tumor_HLADR_EGFR

ggsave(filename = paste0(out_folder,"4C_HLADR_on_tumor_cells_EGFR.pdf"),
       plot = p_Tumor_HLADR_EGFR,
       device = "pdf",
       width = 1,
       height = 2)
```

## generate IMC images HLADR

```{r, fig.height=20, fig.width=20}
# get the expression matrix and caculate stats
counts <- as.data.frame(t(assay(sce1,"exprs")))
head(counts)
dim(counts)

counts$celltypes <- sce1$celltypes
counts$immucan_id <- sce1$immucan_id
counts$sample_id <- sce1$sample_id
counts <- left_join(counts, traits,"immucan_id")

# select samples from the frequency of positive cells
counts %>%
  mutate(HLADRpos = ifelse(HLADR > 3, TRUE,FALSE)) %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  group_by(immucan_id,EGFR,.drop = FALSE) %>%
  dplyr::count(HLADRpos) %>%
  mutate(total_cells = sum(n),
         frac = n/total_cells) %>%
  filter(HLADRpos == TRUE) %>%
  arrange(frac)

EGFRpos <- c("LUNG-NSCLC2-0624","LUNG-NSCLC2-0689","LUNG-NSCLC2-0659")

# load images
library(cytomapper)
masks <- readRDS(paste0(mount_path,"data/IMC/Panel_1/Rout/masks.rds"))
images <- readRDS(paste0(mount_path,"data/IMC/Panel_1/Rout/images.rds"))

cur_df <- as.data.frame(colData(sce1))
rownames(cur_df) <- NULL
cur_df <- cur_df %>%
  select(immucan_id,image, sample_id) %>%
  distinct() %>%
  mutate(image = gsub(".tiff","",image))
rownames(cur_df) <- cur_df$image


# select images
ims <- cur_df %>% 
  filter(immucan_id %in% EGFRpos,
         grepl("002$", sample_id)) %>%
  pull(sample_id)

# unify metadata of masks and images
mcols(images)$immucan_id <- cur_df[names(images),"immucan_id"]
mcols(masks)$immucan_id <- cur_df[names(masks),"immucan_id"]
mcols(images)$sample_id <- cur_df[names(images),"sample_id"]
mcols(masks)$sample_id <- cur_df[names(masks),"sample_id"]


cur_sce <- sce1[,sce1$sample_id %in% ims]
channelNames(images) <- rownames(cur_sce)

cur_ims <- getImages(images,ims)
cur_masks <- getImages(masks, ims)

# read images into memory and normalize
cur_images <- CytoImageList(cur_ims, on_disk = FALSE)
cur_masks <- CytoImageList(cur_masks, on_disk = FALSE)
cur_images <- scaleImages(cur_images,value = 2^16-1)
cur_images <- normalize(cur_images, separateImages = TRUE)
cur_images <- normalize(cur_images, separateImages = TRUE, 
                        inputRange = c(0, 0.2))

# define all tumor and non-tumor cells
cur_sce$tumor <- "non-tumor"
cur_sce[,cur_sce$celltypes == "Tumor"]$tumor <- "tumor"
plot_sce <- cur_sce[,cur_sce$celltypes == "Tumor"]

# generate images
plotPixels(image = cur_images,
           object = plot_sce,
           cell_id = "ObjectNumber",
           img_id = "sample_id",
           colour_by = c("Ecad", "HLADR", "DNA1"),
           colour = list(Ecad = c("black", "green"),HLADR = c("black","red"),DNA1 = c("black","blue")),
           bcg = list(Ecad  = c(0,2,1),
                      HLADR = c(0,1.5,1),
                      DNA1  = c(0,1,1)),
            image_title = NULL,save_plot = list(filename = paste0(out_folder,"HLADR_tumor_2.png")))
```



# 4D CD8 T cells CXC13 in EGFR
```{r, fig.height=5, fig.width=5}
p_CD8_CXCL13_EGFR <- mean_exp %>%
  filter(celltypes == "CD8",
         name %in% c("CXCL13_p2", "PD1_p2","LAG3_p1", "Tim3_p2")) %>%
  left_join(traits, by = "immucan_id") %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  ggplot(aes(x = EGFR, y = value,fill= EGFR))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(color = EGFR),position = position_jitterdodge(jitter.width = 0.2), size= 0.5)+
  scale_fill_manual(values = alpha(c("red", "blue"),alpha = 0.6))+
  scale_color_manual(values = c("red", "blue"))+
  facet_wrap(.~name, scales= "free",nrow = 1)+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  stat_compare_means(size = 1.5)+
    theme_bw()+
  theme(axis.text.y = element_text(size = 6),
        axis.text.x = element_text(size = 6, angle = 90, vjust = 1,hjust = 0.5),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 6),
        strip.text = element_text(size = 6))+
  theme(legend.position = "none")
p_CD8_CXCL13_EGFR

ggsave(filename = paste0(out_folder,"4D_exhaustion_on_CD8_Tcells.pdf"),
       plot = p_CD8_CXCL13_EGFR,
       device = "pdf",
       width = 3.5,
       height = 2.5)
```


# 4E Tumor cells EGFR HLADR in STK11
```{r, fig.height=5, fig.width=5}
p_Tumor_HLADR_STK11 <- mean_exp %>%
  filter(celltypes == "Tumor",
         name %in% c("EGFR_p2", "HLADR_p1")) %>%
  left_join(traits, by = "immucan_id") %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  ggplot(aes(x = STK11, y = value,fill= STK11))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(color = STK11),position = position_jitterdodge(jitter.width = 0.2), size= 0.5)+
  scale_fill_manual(values = alpha(c("red", "blue"),alpha = 0.6))+
  scale_color_manual(values = c("red", "blue"))+
  facet_wrap(.~name, scales= "free")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  stat_compare_means(size = 1.5)+
    theme_bw()+
  theme(axis.text.y = element_text(size = 6),
        axis.text.x = element_text(size = 6, angle = 90, vjust = 1, hjust = 0.5),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 6),
        strip.text = element_text(size = 6),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7))+
  theme(legend.position = "none")
p_Tumor_HLADR_STK11

ggsave(filename = paste0(out_folder,"4E_HLADR_and_EGFR_on_tumor_cells_STK11.pdf"),
       plot = p_Tumor_HLADR_STK11,
       device = "pdf",
       width = 1.8,
       height = 2)
```

# 4F Ki67 and p53 on tumor cells in TP53 mutated in LUAD

```{r, fig.height=5, fig.width=5}
p_Tumor_Ki67_P53_TP53 <- mean_exp %>%
  filter(celltypes == "Tumor",
         name %in% c("p53_p2","Ki67_p1")) %>%
  left_join(traits, by = "immucan_id") %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  ggplot(aes(x = TP53, y = value,fill= TP53))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(aes(color = TP53),position = position_jitterdodge(jitter.width = 0.2), size= 0.75)+
  scale_fill_manual(values = alpha(c("red", "blue"),alpha = 0.6))+
  scale_color_manual(values = c("red", "blue"))+
  facet_wrap(.~name, scales= "free")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  stat_compare_means(size = 1)+
    theme_bw()+
  theme(axis.text.y = element_text(size = 6),
        axis.text.x = element_text(size = 6, angle = 90, vjust = 1, hjust = 0.5),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 6),
        strip.text = element_text(size = 6),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7))+
  theme(legend.position = "none")
p_Tumor_Ki67_P53_TP53

ggsave(filename = paste0(out_folder,"4F_Ki67_p53_on_tumor_cells_TP53.pdf"),
       plot = p_Tumor_Ki67_P53_TP53,
       device = "pdf",
       width = 1.8,
       height = 2)
```


## P53 images
```{r, fig.height=20, fig.width=20}
counts2 <- as.data.frame(t(assay(sce2,"exprs")))
head(counts2)
dim(counts2)

counts2$celltypes <- sce2$celltypes
counts2$immucan_id <- sce2$immucan_id
counts2$sample_id <- sce2$sample_id
counts2 %>%
  filter(celltypes == "Tumor") %>%
  ggplot(aes(x = Ecad,y = p53))+
  geom_point()

counts2 <- left_join(counts2, traits,"immucan_id")

counts2 %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  mutate(p53pos = ifelse(p53 > 1, TRUE,FALSE)) %>%
  group_by(sample_id,.drop = FALSE) %>%
  dplyr::count(p53pos) %>%
  mutate(total_cells = sum(n),
         frac = n/total_cells) %>%
  filter(p53pos == TRUE) %>%
  arrange(frac)

p53pos <- c("10040769","10040838")
p53pos <- traits$immucan_id[grepl(paste(p53pos,collapse = "|"),traits$ibbl_kit_ids)]

# load images
library(cytomapper)
masks <- readRDS(paste0(mount_path,"data/IMC/Panel_2/Rout/masks.rds"))
images <- readRDS(paste0(mount_path,"data/IMC/Panel_2/Rout/images.rds"))

cur_df <- as.data.frame(colData(sce2))
rownames(cur_df) <- NULL
cur_df <- cur_df %>%
  select(immucan_id,image, sample_id) %>%
  distinct() %>%
  mutate(image = gsub(".tiff","",image))
rownames(cur_df) <- cur_df$image


# select images
ims <- cur_df %>% 
  filter(immucan_id %in% p53pos,
         grepl("002$", sample_id)) %>%
  pull(sample_id)

# unify metadata of masks and images
mcols(images)$immucan_id <- cur_df[names(images),"immucan_id"]
mcols(masks)$immucan_id <- cur_df[names(masks),"immucan_id"]
mcols(images)$sample_id <- cur_df[names(images),"sample_id"]
mcols(masks)$sample_id <- cur_df[names(masks),"sample_id"]


cur_sce <- sce2[,sce2$sample_id %in% ims]


cur_ims <- getImages(images,ims)
cur_masks <- getImages(masks, ims)

# read images into memory and normalize
cur_images <- CytoImageList(cur_ims, on_disk = FALSE)
cur_masks <- CytoImageList(cur_masks, on_disk = FALSE)
cur_images <- scaleImages(cur_images,value = 2^16-1)
cur_images <- normalize(cur_images, separateImages = TRUE)
cur_images <- normalize(cur_images, separateImages = TRUE, 
                        inputRange = c(0, 0.2))

channelNames(cur_images) <- rownames(cur_sce)

# define all tumor and non-tumor cells
cur_sce$tumor <- "non-tumor"
cur_sce[,cur_sce$celltypes == "Tumor"]$tumor <- "tumor"
plot_sce <- cur_sce[,cur_sce$celltypes == "Tumor"]

# generate images
plotPixels(image = cur_images,
           object = plot_sce,
           cell_id = "ObjectNumber",
           img_id = "sample_id",
           colour_by = c("Ecad", "p53", "DNA1"),
           colour = list(Ecad = c("black", "green"),p53 = c("black","red"),DNA1 = c("black","blue")),
           bcg = list(Ecad  = c(0,2,1),
                      p53 = c(0,1.5,1),
                      DNA1  = c(0,1,1)),
            image_title = NULL,save_plot = list(filename = paste0(out_folder,"p53_tumor.png")))
```

```{r}
counts2 %>%
  filter(celltypes == "Tumor") %>%
  ggplot(aes(x = Ecad,y = p53))+
  geom_point(size = 0.1)+
  facet_wrap(LUAD_vs_LUSC~TP53)+
  theme_bw()

hist(counts2$p53)

counts2 %>%
  mutate(p53pos = ifelse(p53 > 2, TRUE,FALSE)) %>%
  group_by(immucan_id,TP53,.drop = FALSE) %>%
  dplyr::count(p53pos) %>%
  left_join(traits[,c("immucan_id","LUAD_vs_LUSC")],"immucan_id") %>%
  mutate(total_cells = sum(n),
         frac = n/total_cells) %>%
  filter(p53pos == TRUE) %>%
  ggplot(aes(x = TP53,y = log2(frac)))+
  geom_boxplot()+
  facet_wrap(.~LUAD_vs_LUSC, scales = "free")+
  theme_bw()+
  stat_compare_means(method = "wilcox.test")
```




# 4G

## Treg cells GITR PIK3CA
```{r, fig.height=5, fig.width=5}
p_Treg_GITR_PIK3CA <- mean_exp %>%
  filter(celltypes == "Treg",
         name %in% c("GITR_p2","CD137_p2", "GATA3_p2")) %>%
  left_join(traits, by = "immucan_id") %>%
  filter(simple_histology == "Squamous cell carcinoma") %>%
  ggplot(aes(x = PIK3CA, y = value,fill= PIK3CA))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(color = PIK3CA),position = position_jitterdodge(jitter.width = 0.2), size= 0.5)+
  scale_fill_manual(values = alpha(c("red", "blue"),alpha = 0.6))+
  scale_color_manual(values = c("red", "blue"))+
  facet_wrap(.~name, scales= "free")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  stat_compare_means(size = 1)+
    theme_bw()+
  theme(axis.text.y = element_text(size = 6),
        axis.text.x = element_text(size = 6, angle = 90, vjust = 1, hjust = 0.5),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 6),
        strip.text = element_text(size = 6),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7))+
  theme(legend.position = "none")
p_Treg_GITR_PIK3CA

ggsave(filename = paste0(out_folder,"4G_activation_markers_on_Tregs.pdf"),
       plot = p_Treg_GITR_PIK3CA,
       device = "pdf",
       width = 2.5,
       height = 2)
```




