---
title: "Figure S5"
author: "Daniel Schulz"
date: "2025-03-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggpubr)
library(SingleCellExperiment)
library(scales)
library(ggnewscale)
library(edgeR)
library(cowplot)
library(igraph)
library(ggraph)
library(ggrepel)
library(survminer)
library(survivalAnalysis)
library(survival)
```


```{r}
mount_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/"
out_folder <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/Figure_5_supp/"

if (!dir.exists(out_folder)) {
  dir.create(out_folder)
}
```


# load data
we will load the data we generated at the beginning of figure 2
```{r}
dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/abundances_dat.rds"))
dens_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/densities_dat.rds"))
cello_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/cello_dat.rds"))
clin_dat <- read.csv("/mnt/immucan_volume/clinical_data/Lung_NSCLC2/20230825/NSCLC2_for_R_update_2024.csv")
int_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/interaction_dat.rds"))
traits <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/all_traits.csv"))

# variables
data_type <- unique(str_split_i(string = colnames(dat[,-1]),pattern = "__",i = 2))

# colors
hist_col_2 = setNames(c("blue","orange"),c("LUAD","LUSC"))
hist_col_3 = setNames(c("blue","orange", "gray40"),c("LUAD","LUSC","Other"))
```

# hazards for oncogenes
not in paper
```{r}
p.list <- list()
for (i in c("TP53","KRAS", "EGFR","STK11", "PIK3CA")) {
  for(j in c("Adenocarcinoma", "Squamous cell carcinoma")) {
    cur_dat <- as.data.frame(traits)
    #cur_dat[,i] <- as.factor(cur_dat[,i])
    cur_dat <- cur_dat %>%
      filter(!is.na(cur_dat[,i]),
             simple_histology == j)
    surv_ob <- Surv(cur_dat$time_to_event,cur_dat$censoring)
    p.list[[paste0(j, "_",i)]] <- coxph( as.formula(paste0("surv_ob ~ simple_stage + ",i)),
                          data = cur_dat)
  }
}

df <- data.frame(HR =      unlist(lapply(p.list, function(x){summary(x)$coefficients[,"exp(coef)"][4]})),
                 p.value = unlist(lapply(p.list, function(x){summary(x)$coefficients[, "Pr(>|z|)"][4]})),
                 lower_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "lower .95"][4]})),
                 upper_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "upper .95"][4]}))) 

df1 <- df %>% 
  mutate(hist = str_split_i(rownames(df),"_",1),
         cond = str_split_i(rownames(df),"\\.",2),
         cond = gsub("TRUE","mut", cond)) %>%
  filter(!is.na(HR),
         !(hist=="Squamous cell carcinoma" & cond == "EGFRmut"),
         !(hist=="Adenocarcinoma" & cond == "PIK3CAmut"),
         !(hist=="Adenocarcinoma" & cond == "EGFRmut"))

p1 <- df1 %>%
  mutate(cond = paste0(hist,"_",cond)) %>%
  ggplot(aes(x = HR,y = cond, col = hist))+
  geom_point(size = 2)+
  geom_errorbar(aes(xmin=lower_conf, xmax = upper_conf),lwd =1,width = 0.3)+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust= 1))+
  scale_y_discrete(labels = function(x) str_split_i(x,"_",2))+
  scale_x_log10()+
  theme_bw()+
  geom_vline(xintercept = 1,linetype = "dashed")+
  geom_text(aes(label = paste0("p : ",round(p.value,digits = 3))),
            col = "black",
            vjust = -1,
            inherit.aes = TRUE)+
  scale_color_manual(values = c("Adenocarcinoma" = "blue", "Squamous cell carcinoma" = "orange"))

# Kaplan Meier plot for EGFR since no patients die
surv_fit <- survfit(Surv(time_to_event, censoring) ~ EGFR, data = traits)

p2 <- ggsurvplot(surv_fit, data = traits %>% filter(simple_histology == "Adenocarcinoma"), risk.table = TRUE, pval = TRUE,palette = c("red","blue"))+
  ylab("OS probability")+
  xlab("Time (days)")

p2 <- plot_grid(p2$plot,ncol = 1, rel_heights = c(1,0.5))
p2

p <- plot_grid(p1,p2,ncol = 1,rel_heights = c(1,1))
p

```


# S5A driver mutation enrichment in histology

```{r, fig.height=8}
cur_traits <- traits %>% select("LUAD_vs_LUSC")
condition <- c("Adenocarcinoma")

names(condition) <- colnames(cur_traits)
plots_list <- list()
for(j in c("TP53","KRAS","EGFR","STK11","PIK3CA")){
  p.list <- list()
  l.list <- list()
  for (i in names(condition)) {
    x <- factor(traits[[j]])
    x <- relevel(x, ref = "FALSE")
    
    y <- factor(unlist(traits[,i]))
    y <- relevel(y, ref = condition[i])
    
    p.list[[i]] <- fisher.test(table(x, y))
    l.list[[i]] <- c(levels(y)[1])
  }
  
  out <- data.frame(feature = names(p.list),
                    p.value = unlist(lapply(p.list, function(x){
                      x$p.value})),
                    odds_ratio = unlist(lapply(p.list, function(x){
                      x$estimate})),
                    reference = unlist(l.list))
  
  
  # generate annotation vector:
  anno_vec <- c("LUSC_vs_LUAD")
  out$annotation = anno_vec
  out[,"mutation"] <- paste0(j,"mut")
  plots_list[[j]]<- out 
}

cur_dat <- do.call(rbind,plots_list)

cur_dat <- cur_dat %>%
    mutate(FDR = p.adjust(p.value,method = "BH"),
           sig = ifelse(p.value < 0.05 & odds_ratio > 1, "UP_trend",NA),
           sig = ifelse(FDR < 0.1 & odds_ratio > 1, "UP",sig),
           sig = ifelse(p.value < 0.05 & odds_ratio < 1, "DOWN_trend",sig),
           sig = ifelse(FDR < 0.1 & odds_ratio < 1, "DOWN",sig),
           sig = ifelse(is.na(sig),"NOTSIG", sig),
           p.value = ifelse(is.infinite(odds_ratio),NA,p.value),
           odds_ratio = ifelse(odds_ratio == 0,1/max(odds_ratio),odds_ratio),
           log_odds = log2(odds_ratio))

min_down <- min(cur_dat$log_odds-1)
max_up <- max(cur_dat$log_odds)

p1 <- cur_dat %>%
  ggplot(aes(x=mutation, y = annotation,col = log_odds, size= -log10(p.value)))+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(round(min_down),0,round(max_up+1))),
                         guide = "colorbar", limits=c(round(min_down),round(max_up+1)),name = "log2 odds ratio")+
  scale_size(range = c(1,6))+
  new_scale_colour() +
  geom_point(data =cur_dat, shape = 1,aes(x = mutation,size = -log10(p.value), col = sig), stroke =1.75)+
  scale_color_manual(values = c("UP" = "#9f1214","UP_trend" = "#e73032","DOWN" = "#25567d","UP_trend" = "#e73032","NOTSIG" = "snow2"), name = "stats")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 10),
        axis.text.y = element_text(size = 10),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10),
        plot.margin = unit(c(0.5, 0,0,0), "cm"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.position = "none")
p1

ggsave(filename = paste0(out_folder,"S5A_Driver_enrichment_in_histology.pdf"),
       plot = p1,
       device = "pdf",
       width = 3,
       height = 1.5)
```
# S5A enrichment of clinical in oncogenes

## LUAD
```{r, fig.height=8}
cur_traits <- traits %>% select("TPS_score", "TMB_score","age_dicho", "gender","stage_1_2","stage_1_3","stage_2_3","never_vs_current_smoker",
            "never_vs_Ex_smoker","current_vs_Ex_smoker", "N0_vs_Nx","M0_vs_Mx")
condition <- c("negative", "low","low", "Female","I","I","II","Non-smoker","Non-smoker","Ex-smoker", "node_neg", "met_neg")

names(condition) <- colnames(cur_traits)
plots_list <- list()
cur_traits <- traits %>% filter(simple_histology == "Adenocarcinoma")
for(j in c("TP53","KRAS","EGFR","STK11")){
  p.list <- list()
  l.list <- list()
  for (i in names(condition)) {
    x <- factor(cur_traits[[j]])
    x <- relevel(x, ref = "FALSE")
    
    y <- factor(unlist(cur_traits[,i]))
    y <- relevel(y, ref = condition[i])
    
    p.list[[i]] <- fisher.test(table(x, y))
    l.list[[i]] <- c(levels(y)[1])
  }
  
  out <- data.frame(feature = names(p.list),
                    p.value = unlist(lapply(p.list, function(x){
                      x$p.value})),
                    odds_ratio = unlist(lapply(p.list, function(x){
                      x$estimate})),
                    reference = unlist(l.list))
  
  
  # generate annotation vector:
  anno_vec <- c("TPSpos_vs_TPSneg","TMBhigh_vs_TMBlow","HighAge_vs_lowAge","Male_vs_Female","StageII_vs_I","StageIII_vs_I","StageIII_vs_II","Smoker_vs_neverSmoker","formerSmoker_vs_neverSmoker","Smoker_vs_formerSmoker","NodePos_vs_NodeNeg","MetPos_Vs_MetNeg")
  out$annotation = anno_vec
  out[,"mutation"] <- paste0(j,"mut")
  plots_list[[j]]<- out 
}

cur_dat <- do.call(rbind,plots_list)

cur_dat_LUAD <- cur_dat %>%
    mutate(FDR = p.adjust(p.value,method = "BH"),
           sig = ifelse(p.value < 0.05 & odds_ratio > 1, "UP_trend",NA),
           sig = ifelse(FDR < 0.1 & odds_ratio > 1, "UP",sig),
           sig = ifelse(p.value < 0.05 & odds_ratio < 1, "DOWN_trend",sig),
           sig = ifelse(FDR < 0.1 & odds_ratio < 1, "DOWN",sig),
           sig = ifelse(is.na(sig),"NOTSIG", sig),
           p.value = ifelse(is.infinite(odds_ratio),NA,p.value),
           odds_ratio = ifelse(odds_ratio == 0,1/max(odds_ratio),odds_ratio),
           log_odds = log2(odds_ratio))


p <- cur_dat_LUAD %>%
  ggplot(aes(x=mutation, y = annotation,col = log_odds, size= -log10(p.value)))+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(round(min_down),0,round(max_up+1))),
                         guide = "colorbar", limits=c(round(min_down),round(max_up+1)))+
  scale_size(range = c(1,6))+
  new_scale_colour() +
  geom_point(shape = 1,aes(size = -log10(p.value), col = sig), stroke =1.75)+
  scale_color_manual(values = c("UP" = "#9f1214","UP_trend" = "#e73032","DOWN" = "#25567d","UP_trend" = "#e73032","NOTSIG" = "snow2"), name = "stats")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 10),
        axis.text.y = element_text(size = 10),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10),
        plot.margin = unit(c(0.5, 0,0,0), "cm"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank())
p

```

## LUSC
```{r, fig.height=8}
cur_traits <- traits %>% select("TPS_score", "TMB_score","age_dicho", "gender","stage_1_2","stage_1_3","stage_2_3","current_vs_Ex_smoker", "N0_vs_Nx","M0_vs_Mx")
condition <- c("negative", "low","low", "Female","I","I","II","Ex-smoker", "node_neg", "met_neg")

names(condition) <- colnames(cur_traits)
plots_list <- list()
cur_traits <- traits %>% filter(simple_histology == "Squamous cell carcinoma")
for(j in c("TP53","PIK3CA")){
  p.list <- list()
  l.list <- list()
  for (i in names(condition)) {
    x <- factor(cur_traits[[j]])
    x <- relevel(x, ref = "FALSE")
    
    y <- factor(unlist(cur_traits[,i]))
    y <- relevel(y, ref = condition[i])
    
    p.list[[i]] <- fisher.test(table(x, y))
    l.list[[i]] <- c(levels(y)[1])
  }
  
  out <- data.frame(feature = names(p.list),
                    p.value = unlist(lapply(p.list, function(x){
                      x$p.value})),
                    odds_ratio = unlist(lapply(p.list, function(x){
                      x$estimate})),
                    reference = unlist(l.list))
  
  
  # generate annotation vector:
  anno_vec <- c("TPSpos_vs_TPSneg","TMBhigh_vs_TMBlow","HighAge_vs_lowAge","Male_vs_Female","StageII_vs_I","StageIII_vs_I","StageIII_vs_II","Smoker_vs_formerSmoker","NodePos_vs_NodeNeg","MetPos_Vs_MetNeg")
  out$annotation = anno_vec
  out[,"mutation"] <- paste0(j,"mut")
  plots_list[[j]]<- out
}

cur_dat <- do.call(rbind,plots_list)


cur_dat_LUSC <-  cur_dat %>%
    mutate(FDR = p.adjust(p.value,method = "BH"),
           sig = ifelse(p.value < 0.05 & odds_ratio > 1, "UP_trend",NA),
           sig = ifelse(FDR < 0.1 & odds_ratio > 1, "UP",sig),
           sig = ifelse(p.value < 0.05 & odds_ratio < 1, "DOWN_trend",sig),
           sig = ifelse(FDR < 0.1 & odds_ratio < 1, "DOWN",sig),
           sig = ifelse(is.na(sig),"NOTSIG", sig),
           p.value = ifelse(is.infinite(odds_ratio),NA,p.value),
           odds_ratio = ifelse(odds_ratio == 0,1/max(odds_ratio),odds_ratio),
           odds_ratio = ifelse(is.infinite(odds_ratio),NA,odds_ratio),
           log_odds = log2(odds_ratio))

p <- cur_dat_LUSC %>%
  ggplot(aes(x=mutation, y = annotation,col = log_odds, size= -log10(p.value)))+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(round(min_down),0,round(max_up+1))),
                         guide = "colorbar", limits=c(round(min_down),round(max_up+1)))+
  scale_size(range = c(1,6))+
  new_scale_colour() +
  geom_point(shape = 1,aes(size = -log10(p.value), col = sig), stroke =1.75)+
  scale_color_manual(values = c("UP" = "#9f1214","UP_trend" = "#e73032","DOWN" = "#25567d","UP_trend" = "#e73032","NOTSIG" = "snow2"), name = "stats")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 10),
        axis.text.y = element_text(size = 10),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10),
        plot.margin = unit(c(0.5, 0,0,0), "cm"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank())
p

```


## plot LUSC and LUAD together

```{r}
cur_dat_LUAD$histology <- "LUAD"
cur_dat_LUSC$histology <- "LUSC"

cur_dat <- rbind(cur_dat_LUAD,cur_dat_LUSC)

cur_dat <- cur_dat %>%
  mutate(FDR = p.adjust(p.value,method = "BH"),
         sig = ifelse(p.value < 0.05 & odds_ratio > 1, "UP_trend",NA),
         sig = ifelse(FDR < 0.1 & odds_ratio > 1, "UP",sig),
         sig = ifelse(p.value < 0.05 & odds_ratio < 1, "DOWN_trend",sig),
         sig = ifelse(FDR < 0.1 & odds_ratio < 1, "DOWN",sig),
         sig = ifelse(is.na(sig),"NOTSIG", sig),
         p.value = ifelse(is.infinite(odds_ratio),NA,p.value),
         odds_ratio = ifelse(odds_ratio == 0,1/max(odds_ratio),odds_ratio),
         log_odds = log2(odds_ratio),
         mutation = paste0(histology,"_",mutation))
p <- cur_dat %>%
  ggplot(aes(x=mutation, y = annotation,col = log_odds, size= -log10(p.value)))+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(round(min_down),0,round(max_up+1))),
                         guide = "colorbar", limits=c(round(min_down),round(max_up+1)))+
  scale_size(range = c(1,6))+
  new_scale_colour() +
  geom_point(data = cur_dat ,shape = 1,aes(size = -log10(p.value), col = sig), stroke =1.75)+
  scale_color_manual(values = c("UP" = "#9f1214","UP_trend" = "#e73032","DOWN" = "#25567d","UP_trend" = "#e73032","NOTSIG" = "snow2"), name = "stats")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 10),
        axis.text.y = element_text(size = 10),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10),
        plot.margin = unit(c(0.5, 0,0,0), "cm"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.location = "plot")

p
p2 <- plot_grid(p1,p,ncol = 1, rel_heights = c(1,4.5))
ggsave(filename = paste0(out_folder,"S5A_driver_mutation_enrichment_in_clinical.pdf"),
       plot = p2,
       device = "pdf",
       width = 6,
       height = 7)
```

# S5B mean expression of markers across cell types and driver mutations
```{r}
mean_exp1 <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/mean_marker_expr_per_celltype_p1.csv"))
mean_exp2 <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/mean_marker_expr_per_celltype_p2.csv"))

# long format
mean_exp1 <- mean_exp1 %>%
  pivot_longer(-immucan_id)
mean_exp2 <- mean_exp2 %>%
  pivot_longer(-immucan_id)

mean_exp1$tech <- "IMC1"
mean_exp2$tech <- "IMC2"
mean_exp <- rbind(mean_exp1,mean_exp2)

mean_exp <- mean_exp %>%
  mutate(celltypes = str_split_i(name,"-",1),
         name = str_split_i(name,"-",2))
```

## panel 1 markers of interest - to be removeed
```{r}
markers_1 <- list()
# we will not subcluters "uncertain" cells as the merging will be very difficult. we may inspect this at a later time point again

mean_exp1 <- mean_exp %>%
  filter(tech == "IMC1")
#functional markers_1 for T cells
markers_1[["CD4"]] <- mean_exp1 %>%
  filter(grepl("CD27|LAG3|PD1|CD7|GrzB|ICOS|Ki67|TCF7",name)) %>%
  pull(name) %>%
  unique()

markers_1[["CD8"]] <- mean_exp1 %>%
  filter(grepl(pattern = "CD27|LAG3|PD1|CD7|GrzB|ICOS|Ki67|TCF7",name)) %>%
  pull(name) %>%
  unique()

markers_1[["Treg"]] <- mean_exp1 %>%
  filter(grepl(pattern = "CD27|LAG3|ICOS|Ki67|TCF7",name)) %>%
  pull(name) %>% 
  unique()

# B and BnT cell markers_1
markers_1[["BnT"]] <- mean_exp1 %>%
  filter(grepl(pattern = "CD27|CD45RA|CD45RO|LAG3|PD1|CD7|GrzB|FOXP3|ICOS|Ki67|TCF7|VISTA|CD40|HLADR|CD38",name))%>% 
  pull(name) %>%  
  unique()

markers_1[["B"]] <- mean_exp1 %>%  
  filter(grepl(pattern = "CD27|CD40|Ki67",name)) %>% 
  pull(name) %>% 
  unique()

# myeloid markers_1
markers_1[["MacCD163"]] <- mean_exp1 %>%  
  filter(grepl(pattern = "CD68|CD11c|PDL1|VISTA|CD40|CD14|CD206|IDO1|CD16",name)) %>% 
  pull(name) %>% 
  unique()

# we will also use CD20 for DCs to detect those DCs that sit in TLS
markers_1[["DC"]] <- mean_exp1 %>%  
  filter(grepl(pattern = "PDL1|VISTA|CD40|IDO1|CD20",name)) %>% 
  pull(name) %>%  
  unique()

# we will also use E-cad for HLADR cells as there may still be tumor cells in these cells
markers_1[["HLADR"]] <- mean_exp1 %>%  
  filter(grepl(pattern = "CD11b|CD163|CD68|CD11c|PDL1|ICOS|VISTA|TIM3|CD40|CD14|CD206|CD163|HLADR|Ecad",name)) %>%  
  pull(name) %>%   
  unique()

# NK markers_1
markers_1[["NK"]] <- mean_exp1 %>%  
  filter(grepl(pattern = "CD7|GrzB|CD16|CD45RA|CD45RO",name)) %>%
  pull(name) %>% 
  unique()

# Neutrophil markers_1
markers_1[["Neutrophil"]] <- mean_exp1 %>%  
  filter(grepl(pattern = "MPO|PDL1|VISTA",name)) %>% 
  pull(name) %>% 
  unique()

# pDC markers_1
markers_1[["pDC"]] <- mean_exp1 %>%  
  filter(grepl(pattern = "CD303|CD4|GrzB|CD45RA|CD45RO|CD27",name)) %>%  
  pull(name) %>%  
  unique()

# Plasma cell markers_1
markers_1[["plasma"]] <- mean_exp1 %>%  
  filter(grepl(pattern = "CD38|TCF7|CD27|CD45RA|CD45RO",name)) %>% 
  pull(name) %>% 
  unique()

# stroma markers_1
markers_1[["Mural"]] <- mean_exp1 %>%   
  filter(grepl(pattern = "PDGFR|SMA",name)) %>% 
  pull(name) %>% 
  unique()

# tumor markers_1
markers_1[["Tumor"]] <- mean_exp1 %>%   
  filter(grepl(pattern = "HLADR|CarbonicAnhydrase|cleavedPARP|PDL1|IDO1|CD15|Ki67|Ido|B2M",name)) %>%  
  pull(name) %>%  
  unique()
```

## panel 2 markers of interest - to be removed

```{r}
mean_exp2 <- mean_exp %>%
  filter(tech == "IMC2")
markers_2 <- list()
# we will not subcluters "uncertain" cells as the merging will be very difficult. we may inspect this at a later time point again

markers_2[["CD8"]] <- mean_exp2 %>% 
  filter(grepl(pattern = "CD7$|PD1|Bcl2|GATA3|PD1|NKG2A|CXCL13|EOMES|CD137|CD134|Tbet|GITR|Tim3|Ki67",name)) %>%
  pull(name) %>%  
  unique()
  
markers_2[["CD4"]] <- mean_exp2 %>% 
  filter(grepl(pattern = "CD7$|PD1|Bcl2|GATA3|PD1|NKG2A|CXCL13|EOMES|CD137|CD134|Tbet|GITR|Tim3|Ki67",name)) %>%
  pull(name) %>%  
  unique()
markers_2[["Treg"]] <- mean_exp2 %>%
  filter(grepl(pattern = "CD7$|Bcl2|GATA3|PD1|NKG2A|CXCL13|EOMES|CD137|CD134|Tbet|GITR|Tim3|Ki67",name)) %>%
  pull(name) %>%  
  unique()


# NK markers_2
markers_2[["NK"]] <- mean_exp2 %>% 
  filter(grepl(pattern = "Tbet|Bcl2|GATA3|NKG2A|EOMES|GITR",name)) %>%
  pull(name) %>%  
  unique()


# B and BnT cell markers_2
markers_2[["B"]] <- mean_exp2 %>%
  filter(grepl(pattern = "Bcl2|Ki67",name)) %>%
  pull(name) %>%  
  unique()
markers_2[["BnT"]] <- mean_exp2 %>% 
  filter(grepl(pattern = "CD7$|PD1|FOXP3|Bcl2|GATA3|CD8a|LAG3|PD1|NKG2A|CXCL13|EOMES|CD137|CD134|Tbet|GITR|Tim3|Ki67",name)) %>%
  pull(name) %>%  
  unique()



# myeloid markers_2
markers_2[["MacCD204"]] <- mean_exp2 %>% 
  filter(grepl(pattern = "CD11c|Arg1|MMP9|CD73|CD10|Tim3",name)) %>%
  pull(name) %>%  
  unique()
markers_2[["MacCD209"]] <- mean_exp2 %>% 
  filter(grepl(pattern = "CD11c|Arg1|MMP9|CD73|CD10|Tim3",name)) %>%
  pull(name) %>%  
  unique()
markers_2[["MacCD204_CD209"]] <- mean_exp2 %>%
  filter(grepl(pattern = "CD11c|Arg1|MMP9|CD73|CD10|Tim3",name)) %>%
  pull(name) %>%  
  unique()
# DCs
markers_2[["DC"]] <- mean_exp2 %>% 
  filter(grepl(pattern = "CD137|Tim3|CD20$",name)) %>%
  pull(name) %>%  
  unique()
markers_2[["mregDC"]] <- mean_exp2 %>% 
  filter(grepl(pattern = "CD11c|CD137|Tim3",name)) %>%
  pull(name) %>%  
  unique()


# MDSCs
markers_2[["MDSC"]] <- mean_exp2 %>%
  filter(grepl(pattern = "CD11b|CD10|CXCL8|Arg1|MMP9",name)) %>%
  pull(name) %>%  
  unique()
markers_2[["PMN_MDSC"]] <- mean_exp2 %>%
  filter(grepl(pattern = "CD11b|CD15$|CD66b|CD10|CXCL8|Arg1|MMP9",name)) %>%
  pull(name) %>%  
  unique()
markers_2[["Neutrophil"]] <- mean_exp2 %>%
  filter(grepl(pattern = "CD11b|CD15$|CD66b|CD10|CXCL8|Arg1|MMP9",name)) %>%
  pull(name) %>%  
  unique()


# stroma markers_2
markers_2[["Vasculature"]] <- mean_exp2 %>% filter(grepl(pattern = "CD31|CD146",name)) %>%
  pull(name) %>%  
  unique()

markers_2[["vCAF"]] <- mean_exp2 %>%
  filter(grepl(pattern = "SMA|CD146",name)) %>%
  pull(name) %>%  
  unique()

markers_2[["Fibro"]] <- mean_exp2 %>% 
  filter(grepl(pattern = "Podoplanin|CD10|MMP9|SMA",name)) %>%
  pull(name) %>%  
  unique()


# tumor markers_2
markers_2[["Tumor"]] <- mean_exp2 %>%
  filter(grepl(pattern = "panCK|Ecad|Arg1|EGFR|MMP9|p53|Lamp3|CD73|Bcl2|GATA3|CD155|CD10|CD56|CXCL8|Ki67|Podoplanin|IgG",name)) %>%
  pull(name) %>%  
  unique()

```

```{r, fig.height=15, fig.width=15}
cur_files_1 <- lapply(names(markers_1), function(x){
  cur_dat <- mean_exp %>%
    filter(celltypes == x,
           name %in% markers_1[[x]])
})
cur_files_2 <- lapply(names(markers_2), function(x){
  mean_exp %>%
    filter(celltypes == x,
           name %in% markers_2[[x]])
})

mean_exp1 <- do.call(rbind,cur_files_1)
mean_exp2 <- do.call(rbind,cur_files_2)

mean_data <- distinct(rbind(mean_exp1,mean_exp2))

mean_exp %>%
  left_join(traits,"immucan_id") %>%
  group_by(EGFR, celltypes,name) %>%
  summarise(mean(value))

library(dplyr)
library(rstatix)

p.list <- list()

# LUAD EGFR
p.list[["LUAD_EGFR"]] <- mean_data %>%
  left_join(traits,"immucan_id") %>%
  filter(!is.na(EGFR),
         simple_histology == "Adenocarcinoma") %>%
  group_by(celltypes, name) %>%
  summarise(
    wilcox_test = list(wilcox_test(value ~ EGFR, data = cur_data(), detailed = TRUE)),
    logFC = log2(mean(value[EGFR == TRUE], na.rm = TRUE) / mean(value[EGFR == FALSE], na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  unnest(wilcox_test) %>%
  mutate(condition = "LUAD_EGFR")

# LUAD KRAS
p.list[["LUAD_KRAS"]] <- mean_data %>%
  left_join(traits,"immucan_id") %>%
  filter(!is.na(KRAS),
         simple_histology == "Adenocarcinoma") %>%
  group_by(celltypes, name) %>%
  summarise(
    wilcox_test = list(wilcox_test(value ~ KRAS, data = cur_data(), detailed = TRUE)),
    logFC = log2(mean(value[KRAS == TRUE], na.rm = TRUE) / mean(value[KRAS == FALSE], na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  unnest(wilcox_test) %>%
  mutate(condition = "LUAD_KRAS")

#LUAD STK11
p.list[["LUAD_STK11"]] <- mean_data %>%
  left_join(traits,"immucan_id") %>%
  filter(!is.na(STK11),
         simple_histology == "Adenocarcinoma") %>%
  group_by(celltypes, name) %>%
  summarise(
    wilcox_test = list(wilcox_test(value ~ STK11, data = cur_data(), detailed = TRUE)),
    logFC = log2(mean(value[STK11 == TRUE], na.rm = TRUE) / mean(value[STK11 == FALSE], na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  unnest(wilcox_test) %>%
  mutate(condition = "LUAD_STK11")

# LUAD TP53
p.list[["LUAD_p53"]] <- mean_data %>%
  left_join(traits,"immucan_id") %>%
  filter(!is.na(TP53),
         simple_histology == "Adenocarcinoma") %>%
  group_by(celltypes, name) %>%
  summarise(
    wilcox_test = list(wilcox_test(value ~ TP53, data = cur_data(), detailed = TRUE)),
    logFC = log2(mean(value[TP53 == TRUE], na.rm = TRUE) / mean(value[TP53 == FALSE], na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  unnest(wilcox_test) %>%
  mutate(condition = "LUAD_TP53")

# LUSC PIK3CA
p.list[["LUSC_PIK3CA"]] <- mean_data %>%
  left_join(traits,"immucan_id") %>%
  filter(!is.na(PIK3CA),
         simple_histology == "Squamous cell carcinoma") %>%
  group_by(celltypes, name) %>%
  summarise(
    wilcox_test = list(wilcox_test(value ~ PIK3CA, data = cur_data(), detailed = TRUE)),
    logFC = log2(mean(value[PIK3CA == TRUE], na.rm = TRUE) / mean(value[PIK3CA == FALSE], na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  unnest(wilcox_test) %>%
  mutate(condition = "LUSC_PIK3CA")

# LUSC TP53
p.list[["LUSC_TP53"]] <- mean_data %>%
  left_join(traits,"immucan_id") %>%
  filter(!is.na(TP53),
         simple_histology == "Squamous cell carcinoma") %>%
  group_by(celltypes, name) %>%
  summarise(
    wilcox_test = list(wilcox_test(value ~ TP53, data = cur_data(), detailed = TRUE)),
    logFC = log2(mean(value[TP53 == TRUE], na.rm = TRUE) / mean(value[TP53 == FALSE], na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  unnest(wilcox_test) %>%
  mutate(condition = "LUSC_TP53")

dat <- do.call(rbind,p.list)

dat <- dat %>%
  group_by(celltypes, condition) %>%
  mutate(p_adjust =p.adjust(p = p,method = "BH"),
         FCcutoff = ifelse(abs(logFC)>0.33,TRUE,FALSE),
         show =  p_adjust < 0.1 & FCcutoff == TRUE)
dat %>%
  filter(celltypes == "Treg",
         condition == "LUSC_PIK3CA")

library(ggrepel)
sce1 <- readRDS(paste0(mount_path,"data/IMC/Panel_1/Rout/sce.rds"))
sce2 <- readRDS(paste0(mount_path,"data/IMC/Panel_2/Rout/sce.rds"))

col_vec <- c(metadata(sce1)$colour_vectors$celltypes,metadata(sce2)$colour_vectors$cell_types)
col_vec <- col_vec[!duplicated(names(col_vec))]



p <- dat %>%
  ggplot(aes(x = logFC,y = -log10(p), col= celltypes, shape = factor(condition)))+
  geom_point(size = 4)+
  geom_text_repel(data = dat %>% filter(show == TRUE),aes(x = logFC,y = -log10(p),label= paste(name, celltypes,sep = "-")), col = "black",size = 5,box.padding = 0.5,max.overlaps = 20)+
  scale_color_manual(values = col_vec)+
  theme_bw()+
  theme(legend.position = "bottom",
        legend.text = element_text(size = 12),
        axis.text.x = element_text(size =12),
        axis.text.y = element_text(size =12),
        axis.title.x = element_text(size =12),
        axis.title.y = element_text(size =12))

p

ggsave(filename = paste0(out_folder,"S5B_marker_exp_fold_changes_all_celltypes.pdf"),
       plot = p,
       device = "pdf",
       width = 15,
       height = 10)

```

# S5C number HLADR positive tumor cells in EGFR

```{r, fig.width=4, fig.height=8}
sce1 <- sce1[,sce1$ROI != 4]

counts <- as.data.frame(t(assay(sce1,"exprs")))
head(counts)
dim(counts)

counts$celltypes <- sce1$celltypes
counts$immucan_id <- sce1$immucan_id
counts$sample_id <- sce1$sample_id
counts <- left_join(counts, traits,"immucan_id")

p <- counts %>%
  filter(celltypes == "Tumor", 
         simple_histology == "Adenocarcinoma") %>%
  ggplot(aes(x = Ecad,y = HLADR))+
  geom_point(size = 0.1)+
  facet_wrap(.~EGFR)+
  theme_bw()

counts_EGFR <- counts %>%
  filter(celltypes == "Tumor", 
         simple_histology == "Adenocarcinoma",
         EGFR == TRUE)

counts_EGFRneg <- counts %>%
  filter(celltypes == "Tumor", 
         simple_histology == "Adenocarcinoma",
         EGFR == FALSE)

library(mclust)

fit2 <- mixtools::normalmixEM(counts_EGFRneg$HLADR,k = 2)
# get cut point
post <- fit2$posterior[, 1]
# Approximate cut point: where posteriors are equal (0.5)
idx <- which.min(abs(post - 0.5))
cut_point_HLADR <- counts_EGFRneg$HLADR[idx]
cut_point_HLADR

png(filename = paste0(out_folder,"S5C_HLADR_tumor_cell_EGFRmutated.png"),width = 1000, height = 500,units = "px")
par(mfrow = c(1,2))
LSD::heatscatter(x = counts_EGFR$Ecad, y = counts_EGFR$HLADR,main = "EGFR mutated")
abline(h = cut_point_HLADR, lty = 2, lwd = 5)
LSD::heatscatter(x = counts_EGFRneg$Ecad, y = counts_EGFRneg$HLADR, main = "EGFR wild type")
abline(h = cut_point_HLADR, lty = 2, lwd = 5)
dev.off()
```

# S5D Ki67 and p53 on tumor cells in LUSC

```{r}
p_Tumor_Ki67_P53_TP53_LUSC <- mean_exp %>%
  filter(celltypes == "Tumor",
         name %in% c("p53_p2","Ki67_p1","Ki67_p2")) %>%
  left_join(traits, by = "immucan_id") %>%
  filter(simple_histology == "Squamous cell carcinoma", 
         !is.na(TP53)) %>%
  ggplot(aes(x = TP53, y = value,fill= TP53))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(color = TP53),position = position_jitterdodge(jitter.width = 0.2), size= 1)+
  scale_fill_manual(values = alpha(c("red", "blue"),alpha = 0.6))+
  scale_color_manual(values = c("red", "blue"))+
  facet_wrap(.~name, scales= "free")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  stat_compare_means(size = 2.5)+
    theme_bw()+
  theme(axis.text.y = element_text(size = 6),
        axis.text.x = element_text(size = 6),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 6),
        strip.text = element_text(size = 7))+
  theme(legend.position = "none")
p_Tumor_Ki67_P53_TP53_LUSC

ggsave(filename = paste0(out_folder,"S5D_Ki67_p53_tumor_cells_LUSC.pdf"),
       plot = p_Tumor_Ki67_P53_TP53_LUSC,
       device = "pdf",
       width = 5,
       height = 3)
```

# S5E GITR on Tregs

```{r}
counts2 <- as.data.frame(t(assay(sce2,"exprs")))
head(counts2)
dim(counts2)

counts2$celltypes <- sce2$celltypes
counts2$immucan_id <- sce2$immucan_id
counts2$sample_id <- sce2$sample_id
counts2 %>%
  filter(celltypes == "Treg") %>%
  ggplot(aes(x = FOXP3,y = GITR))+
  geom_point()

counts2 <- left_join(counts2, traits,"immucan_id")

counts2 %>%
  filter(simple_histology == "Squamous cell carcinoma",
         celltypes == "Treg") %>%
  mutate(PIK3CApos = ifelse(GITR > 1, TRUE,FALSE)) %>%
  group_by(sample_id,.drop = FALSE) %>%
  dplyr::count(PIK3CApos) %>%
  mutate(total_cells = sum(n),
         frac = n/total_cells) %>%
  filter(PIK3CApos == TRUE) %>%
  arrange(frac)

# selected images
ims <- c("IMMUcan_Batch20210921_LUNG_10041582-LUNG-VAR-TIS-01-IMC-02_003","IMMUcan_Batch20210701_LUNG_10040761-LUNG-VAR-TIS-01-IMC-02_001")


# load images
library(cytomapper)
masks <- readRDS(paste0(mount_path,"data/IMC/Panel_2/Rout/masks.rds"))
images <- readRDS(paste0(mount_path,"data/IMC/Panel_2/Rout/images.rds"))

cur_df <- as.data.frame(colData(sce2))
rownames(cur_df) <- NULL
cur_df <- cur_df %>%
  select(immucan_id,image, sample_id) %>%
  distinct() %>%
  mutate(image = gsub(".tiff","",image))
rownames(cur_df) <- cur_df$image


# unify metadata of masks and images
mcols(images)$immucan_id <- cur_df[names(images),"immucan_id"]
mcols(masks)$immucan_id <- cur_df[names(masks),"immucan_id"]
mcols(images)$sample_id <- cur_df[names(images),"sample_id"]
mcols(masks)$sample_id <- cur_df[names(masks),"sample_id"]


cur_sce <- sce2[,sce2$sample_id %in% ims]


cur_ims <- getImages(images,ims)
cur_masks <- getImages(masks, ims)

# read images into memory and normalize
cur_images <- CytoImageList(cur_ims, on_disk = FALSE)
cur_masks <- CytoImageList(cur_masks, on_disk = FALSE)
cur_images <- scaleImages(cur_images,value = 2^16-1)
cur_masks <- scaleImages(cur_masks,value = 2^16-1)
cur_images <- normalize(cur_images, separateImages = TRUE)
cur_images <- normalize(cur_images, separateImages = TRUE, 
                        inputRange = c(0, 0.2))

channelNames(cur_images) <- rownames(cur_sce)

# define all tumor and non-tumor cells
cur_sce$Treg <- "nonTreg"
cur_sce[,cur_sce$celltypes == "Treg"]$Treg <- "Treg"
plot_sce <- cur_sce[,cur_sce$celltypes == "Treg"]

test_images <- endoapply(cur_images, gblur, sigma = 1)

# generate images
plotPixels(image = test_images,
           mask = cur_masks,
           object = plot_sce,
           cell_id = "ObjectNumber",
           img_id = "sample_id",
           colour_by = c("CD137", "GITR", "FOXP3"),
           outline_by = "Treg",
           colour = list(CD137 = c("black", "green"),GITR = c("black","red"),FOXP3 = c("black","blue"),
                        Treg = c(Treg = "white")),
           bcg = list(CD137  = c(0,1.5,1),
                      GITR = c(0,1.5,1),
                      FOXP3  = c(0,3,1)),
            image_title = NULL,save_plot = list(filename = paste0(out_folder,"S5E_GITR_Tregs_blurred.png")))

plotPixels(image = cur_images,
           mask = cur_masks,
           object = plot_sce,
           cell_id = "ObjectNumber",
           img_id = "sample_id",
           colour_by = c("CD137", "GITR", "FOXP3"),
           outline_by = "Treg",
           colour = list(CD137 = c("black", "green"),GITR = c("black","red"),FOXP3 = c("black","blue"),
                        Treg = c(Treg = "white")),
           bcg = list(CD137  = c(0,1.5,1),
                      GITR = c(0,1.5,1),
                      FOXP3  = c(0,3,1)),
            image_title = NULL,save_plot = list(filename = paste0(out_folder,"S5E_GITR_Tregs.png")))

```

