---
title: "Supplementary Figure 4"
author: "Daniel Schulz"
date: "2025-03-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggpubr)
library(SingleCellExperiment)
library(scales)
library(ggnewscale)
library(edgeR)
library(cowplot)
library(igraph)
library(ggraph)
library(ggrepel)
library(survminer)
library(survivalAnalysis)
library(survival)
```


```{r}
out_folder <- "/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/Figures/Figure_4_supp/"
```


# load data
we will load the data we generated at the beginning of figure 2
```{r}
dat <- readRDS("/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/Figures/scores_for_sharing/abundances_dat.rds")
dens_dat <- readRDS("/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/Figures/scores_for_sharing/densities_dat.rds")
cello_dat <- readRDS("/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/Figures/scores_for_sharing/cello_dat.rds")
clin_dat <- read.csv("/mnt/immucan_volume/clinical_data/Lung_NSCLC2/20230825/NSCLC2_for_R_update_2024.csv")
int_dat <- readRDS("/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/Figures/scores_for_sharing/interaction_dat.rds")
traits <- read_csv("/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/IMC/Rout/data/scores_for_sharing/all_traits.csv")

# variables
data_type <- unique(str_split_i(string = colnames(dat[,-1]),pattern = "__",i = 2))
```

# S4A
```{r}
p.list <- list()
for (i in c("TP53","KRAS", "EGFR","STK11", "PIK3CA")) {
  for(j in c("Adenocarcinoma", "Squamous cell carcinoma")) {
    cur_dat <- as.data.frame(traits)
    #cur_dat[,i] <- as.factor(cur_dat[,i])
    cur_dat <- cur_dat %>%
      filter(!is.na(cur_dat[,i]),
             simple_histology == j)
    surv_ob <- Surv(cur_dat$time_to_event,cur_dat$censoring)
    p.list[[paste0(j, "_",i)]] <- coxph( as.formula(paste0("surv_ob ~ simple_stage + ",i)),
                          data = cur_dat)
  }
}

df <- data.frame(HR =      unlist(lapply(p.list, function(x){summary(x)$coefficients[,"exp(coef)"][4]})),
                 p.value = unlist(lapply(p.list, function(x){summary(x)$coefficients[, "Pr(>|z|)"][4]})),
                 lower_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "lower .95"][4]})),
                 upper_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "upper .95"][4]}))) 

df1 <- df %>% 
  mutate(hist = str_split_i(rownames(df),"_",1),
         cond = str_split_i(rownames(df),"\\.",2),
         cond = gsub("TRUE","mut", cond)) %>%
  filter(!is.na(HR),
         !(hist=="Squamous cell carcinoma" & cond == "EGFRmut"),
         !(hist=="Adenocarcinoma" & cond == "PIK3CAmut"),
         !(hist=="Adenocarcinoma" & cond == "EGFRmut"))

p1 <- df1 %>%
  mutate(cond = paste0(hist,"_",cond)) %>%
  ggplot(aes(x = HR,y = cond, col = hist))+
  geom_point(size = 2)+
  geom_errorbar(aes(xmin=lower_conf, xmax = upper_conf),lwd =1,width = 0.3)+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust= 1))+
  scale_y_discrete(labels = function(x) str_split_i(x,"_",2))+
  scale_x_log10()+
  theme_bw()+
  geom_vline(xintercept = 1,linetype = "dashed")+
  geom_text(aes(label = paste0("p : ",round(p.value,digits = 3))),
            col = "black",
            vjust = -1,
            inherit.aes = TRUE)+
  scale_color_manual(values = c("Adenocarcinoma" = "blue", "Squamous cell carcinoma" = "orange"))

df2 <- df %>% 
  mutate(hist = str_split_i(rownames(df),"_",1),
         cond = str_split_i(rownames(df),"\\.",2),
         cond = gsub("TRUE","mut", cond)) %>%
  filter((hist=="Adenocarcinoma" & cond == "EGFRmut"))

p2 <- df2 %>%
  mutate(cond = paste0(hist,"_",cond)) %>%
  ggplot(aes(x = HR,y = cond, col = hist))+
  geom_point(size = 2)+
  geom_errorbar(aes(xmin=lower_conf, xmax = upper_conf),lwd =1,width = 0.3)+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust= 1))+
  scale_y_discrete(labels = function(x) str_split_i(x,"_",2))+
  scale_x_log10()+
  theme_bw()+
  geom_vline(xintercept = 1,linetype = "dashed")+
  geom_text(aes(label = paste0("p : ",round(p.value,digits = 3))),
            col = "black",
            vjust = -1,
            inherit.aes = TRUE)+
  scale_color_manual(values = c("Adenocarcinoma" = "blue", "Squamous cell carcinoma" = "orange"))
p2
p <- plot_grid(p1,p2,ncol = 1,rel_heights = c(2,1))
ggsave(filename = paste0(out_folder,"S4A_hazard_for_driver_mutations.pdf"),
       plot = p,
       device = "pdf",
       width = 6,
       height = 6)
```


# S4B enrichment of TPS, TMB in drivers
```{r, fig.height=8}
cur_traits <- traits %>% select("TPS_score", "TMB_score","age_dicho", "gender","stage_1_2","stage_1_3","stage_2_3","LUAD_vs_LUSC","never_vs_current_smoker",
            "never_vs_Ex_smoker","current_vs_Ex_smoker", "N0_vs_Nx","M0_vs_Mx")
condition <- c("negative", "low","low", "Female","I","I","II","Adenocarcinoma","Non-smoker","Non-smoker","Ex-smoker", "node_neg", "met_neg")

names(condition) <- colnames(cur_traits)
plots_list <- list()
for(j in c("TP53","KRAS","EGFR","STK11","PIK3CA")){
  p.list <- list()
  l.list <- list()
  for (i in names(condition)) {
    x <- factor(traits[[j]])
    x <- relevel(x, ref = "FALSE")
    
    y <- factor(unlist(traits[,i]))
    y <- relevel(y, ref = condition[i])
    
    p.list[[i]] <- fisher.test(table(x, y))
    l.list[[i]] <- c(levels(y)[1])
  }
  
  out <- data.frame(feature = names(p.list),
                    p.value = unlist(lapply(p.list, function(x){
                      x$p.value})),
                    odds_ratio = unlist(lapply(p.list, function(x){
                      x$estimate})),
                    reference = unlist(l.list))
  
  
  # generate annotation vector:
  anno_vec <- c("TPSpos_vs_TPSneg","TMBhigh_vs_TMBlow","HighAge_vs_lowAge","Male_vs_Female","StageII_vs_I","StageIII_vs_I","StageIII_vs_II","LUSC_vs_LUAD","Smoker_vs_neverSmoker","formerSmoker_vs_neverSmoker","Smoker_vs_formerSmoker","NodePos_vs_NodeNeg","MetPos_Vs_MetNeg")
  out$annotation = anno_vec
  out[,"mutation"] <- paste0(j,"mut")
  plots_list[[j]]<- out %>%
    mutate(FDR = p.adjust(p.value,method = "BH"),
           sig = ifelse(FDR < 0.1,TRUE, FALSE),
           p.value = ifelse(is.infinite(odds_ratio),NA,p.value))
}

plots_list

cur_dat <- do.call(rbind,plots_list)

cur_dat <- cur_dat %>%
  mutate(odds_ratio = ifelse(odds_ratio == 0,0.01,odds_ratio),
         odds_ratio = ifelse(is.infinite(odds_ratio),10,odds_ratio),
         p.value = ifelse(is.na(p.value),min(cur_dat$p.value,na.rm = TRUE),p.value),
         log_odds = log(odds_ratio))
min_down <- min(cur_dat$log_odds)
max_up <- max(cur_dat$log_odds)

p <- cur_dat %>%
  ggplot(aes(x=mutation, y = annotation,col = log_odds, size= -log10(p.value)))+
  #geom_tile(aes(fill = sig))+
  #theme_bw()+
  geom_point()+
  theme_bw()+
  #scale_colour_viridis_c()+
  #theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  #      panel.background = element_blank(), axis.line = element_line(colour = "black"))+
  #theme_bw()+
  scale_colour_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(round(min_down),0,round(max_up+1))),
                         guide = "colorbar", limits=c(round(min_down),round(max_up+1)))+
  scale_size(range = c(1,6))+
  new_scale_colour() +
  geom_point(shape = 1,aes(size = -log10(p.value), col = sig), stroke =1.75)+
  scale_color_manual(values = c("TRUE" = "black","FALSE" = "white"), name = "FDR < 0.1")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 10),
        axis.text.y = element_text(size = 10),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10),
        plot.margin = unit(c(0.5, 0,0,0), "cm"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank())
p
ggsave(filename = paste0(out_folder,"S4B_driver_mutation_enrichment_in_clinical.pdf"),
       plot = p,
       device = "pdf",
       width = 4.5,
       height = 5)
```



# S4C mean expression of markers across cell types and driver mutations
```{r}
mean_exp1 <- read_csv("/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/Figures/scores_for_sharing/mean_marker_expr_per_celltype_p1.csv")
mean_exp2 <- read_csv("/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/Figures/scores_for_sharing/mean_marker_expr_per_celltype_p2.csv")

mean_exp1$tech <- "IMC1"
mean_exp2$tech <- "IMC2"
mean_exp <- rbind(mean_exp1,mean_exp2)
```

## panel 1 markers of interest - to be removeed
```{r}
markers_1 <- list()
# we will not subcluters "uncertain" cells as the merging will be very difficult. we may inspect this at a later time point again

#functional markers_1 for T cells
markers_1[["CD4"]] <- mean_exp %>%
  filter(grepl("CD27|LAG3|PD1|CD7|GrzB|ICOS|Ki67|TCF7",name)) %>%
  pull(name) %>%
  unique()

markers_1[["CD8"]] <- mean_exp %>%
  filter(grepl(pattern = "CD27|LAG3|PD1|CD7|GrzB|ICOS|Ki67|TCF7",name)) %>%
  pull(name) %>%
  unique()

markers_1[["Treg"]] <- mean_exp %>%
  filter(grepl(pattern = "CD27|LAG3|ICOS|Ki67|TCF7",name)) %>%
  pull(name) %>% 
  unique()

# B and BnT cell markers_1
markers_1[["BnT"]] <- mean_exp %>%
  filter(grepl(pattern = "CD27|CD45RA|CD45RO|LAG3|PD1|CD7|GrzB|FOXP3|ICOS|Ki67|TCF7|VISTA|CD40|HLADR|CD38",name))%>% 
  pull(name) %>%  
  unique()

markers_1[["B"]] <- mean_exp %>%  
  filter(grepl(pattern = "CD27|CD40|Ki67",name)) %>% 
  pull(name) %>% 
  unique()

# myeloid markers_1
markers_1[["MacCD163"]] <- mean_exp %>%  
  filter(grepl(pattern = "CD68|CD11c|PDL1|VISTA|CD40|CD14|CD206|IDO1|CD16",name)) %>% 
  pull(name) %>% 
  unique()

# we will also use CD20 for DCs to detect those DCs that sit in TLS
markers_1[["DC"]] <- mean_exp %>%  
  filter(grepl(pattern = "PDL1|VISTA|CD40|IDO1|CD20",name)) %>% 
  pull(name) %>%  
  unique()

# we will also use E-cad for HLADR cells as there may still be tumor cells in these cells
markers_1[["HLADR"]] <- mean_exp %>%  
  filter(grepl(pattern = "CD11b|CD163|CD68|CD11c|PDL1|ICOS|VISTA|TIM3|CD40|CD14|CD206|CD163|HLADR|Ecad",name)) %>%  
  pull(name) %>%   
  unique()

# NK markers_1
markers_1[["NK"]] <- mean_exp %>%  
  filter(grepl(pattern = "CD7|GrzB|CD16|CD45RA|CD45RO",name)) %>%
  pull(name) %>% 
  unique()

# Neutrophil markers_1
markers_1[["Neutrophil"]] <- mean_exp %>%  
  filter(grepl(pattern = "MPO|PDL1|VISTA",name)) %>% 
  pull(name) %>% 
  unique()

# pDC markers_1
markers_1[["pDC"]] <- mean_exp %>%  
  filter(grepl(pattern = "CD303|CD4|GrzB|CD45RA|CD45RO|CD27",name)) %>%  
  pull(name) %>%  
  unique()

# Plasma cell markers_1
markers_1[["plasma"]] <- mean_exp %>%  
  filter(grepl(pattern = "CD38|TCF7|CD27|CD45RA|CD45RO",name)) %>% 
  pull(name) %>% 
  unique()

# stroma markers_1
markers_1[["Mural"]] <- mean_exp %>%   
  filter(grepl(pattern = "PDGFR|SMA",name)) %>% 
  pull(name) %>% 
  unique()

# tumor markers_1
markers_1[["Tumor"]] <- mean_exp %>%   
  filter(grepl(pattern = "HLADR|CarbonicAnhydrase|cleavedPARP|PDL1|IDO1|CD15|Ki67|Ido|B2M",name)) %>%  
  pull(name) %>%  
  unique()
```

## panel 2 markers of interest - to be removed

```{r}
markers_2 <- list()
# we will not subcluters "uncertain" cells as the merging will be very difficult. we may inspect this at a later time point again

markers_2[["CD8"]] <- mean_exp2 %>% 
  filter(grepl(pattern = "CD7$|PD1|Bcl2|GATA3|PD1|NKG2A|CXCL13|EOMES|CD137|CD134|Tbet|GITR|Tim3|Ki67",name)) %>%
  pull(name) %>%  
  unique()
  
markers_2[["CD4"]] <- mean_exp2 %>% 
  filter(grepl(pattern = "CD7$|PD1|Bcl2|GATA3|PD1|NKG2A|CXCL13|EOMES|CD137|CD134|Tbet|GITR|Tim3|Ki67",name)) %>%
  pull(name) %>%  
  unique()
markers_2[["Treg"]] <- mean_exp2 %>%
  filter(grepl(pattern = "CD7$|Bcl2|GATA3|PD1|NKG2A|CXCL13|EOMES|CD137|CD134|Tbet|GITR|Tim3|Ki67",name)) %>%
  pull(name) %>%  
  unique()


# NK markers_2
markers_2[["NK"]] <- mean_exp2 %>% 
  filter(grepl(pattern = "Tbet|Bcl2|GATA3|NKG2A|EOMES|GITR",name)) %>%
  pull(name) %>%  
  unique()


# B and BnT cell markers_2
markers_2[["B"]] <- mean_exp2 %>%
  filter(grepl(pattern = "Bcl2|Ki67",name)) %>%
  pull(name) %>%  
  unique()
markers_2[["BnT"]] <- mean_exp2 %>% 
  filter(grepl(pattern = "CD7$|PD1|FOXP3|Bcl2|GATA3|CD8a|LAG3|PD1|NKG2A|CXCL13|EOMES|CD137|CD134|Tbet|GITR|Tim3|Ki67",name)) %>%
  pull(name) %>%  
  unique()



# myeloid markers_2
markers_2[["MacCD204"]] <- mean_exp2 %>% 
  filter(grepl(pattern = "CD11c|Arg1|MMP9|CD73|CD10|Tim3",name)) %>%
  pull(name) %>%  
  unique()
markers_2[["MacCD209"]] <- mean_exp2 %>% 
  filter(grepl(pattern = "CD11c|Arg1|MMP9|CD73|CD10|Tim3",name)) %>%
  pull(name) %>%  
  unique()
markers_2[["MacCD204_CD209"]] <- mean_exp2 %>%
  filter(grepl(pattern = "CD11c|Arg1|MMP9|CD73|CD10|Tim3",name)) %>%
  pull(name) %>%  
  unique()
# DCs
markers_2[["DC"]] <- mean_exp2 %>% 
  filter(grepl(pattern = "CD137|Tim3|CD20$",name)) %>%
  pull(name) %>%  
  unique()
markers_2[["mregDC"]] <- mean_exp2 %>% 
  filter(grepl(pattern = "CD11c|CD137|Tim3",name)) %>%
  pull(name) %>%  
  unique()


# MDSCs
markers_2[["MDSC"]] <- mean_exp2 %>%
  filter(grepl(pattern = "CD11b|CD10|CXCL8|Arg1|MMP9",name)) %>%
  pull(name) %>%  
  unique()
markers_2[["PMN_MDSC"]] <- mean_exp2 %>%
  filter(grepl(pattern = "CD11b|CD15$|CD66b|CD10|CXCL8|Arg1|MMP9",name)) %>%
  pull(name) %>%  
  unique()
markers_2[["Neutrophil"]] <- mean_exp2 %>%
  filter(grepl(pattern = "CD11b|CD15$|CD66b|CD10|CXCL8|Arg1|MMP9",name)) %>%
  pull(name) %>%  
  unique()


# stroma markers_2
markers_2[["Vasculature"]] <- mean_exp2 %>% filter(grepl(pattern = "CD31|CD146",name)) %>%
  pull(name) %>%  
  unique()

markers_2[["vCAF"]] <- mean_exp2 %>%
  filter(grepl(pattern = "SMA|CD146",name)) %>%
  pull(name) %>%  
  unique()

markers_2[["Fibro"]] <- mean_exp2 %>% 
  filter(grepl(pattern = "Podoplanin|CD10|MMP9|SMA",name)) %>%
  pull(name) %>%  
  unique()


# tumor markers_2
markers_2[["Tumor"]] <- mean_exp2 %>%
  filter(grepl(pattern = "panCK|Ecad|Arg1|EGFR|MMP9|p53|Lamp3|CD73|Bcl2|GATA3|CD155|CD10|CD56|CXCL8|Ki67|Podoplanin|IgG",name)) %>%
  pull(name) %>%  
  unique()
```

```{r, fig.height=15, fig.width=15}
cur_files_1 <- lapply(names(markers_1), function(x){
  cur_dat <- mean_exp %>%
    filter(celltypes == x,
           name %in% markers_1[[x]])
})
cur_files_2 <- lapply(names(markers_2), function(x){
  mean_exp %>%
    filter(celltypes == x,
           name %in% markers_2[[x]])
})

mean_exp1 <- do.call(rbind,cur_files_1)
mean_exp2 <- do.call(rbind,cur_files_2)

mean_data <- distinct(rbind(mean_exp1,mean_exp2))

mean_exp %>%
  left_join(traits,"immucan_id") %>%
  group_by(EGFR, celltypes,name) %>%
  summarise(mean(mean_exp))

library(dplyr)
library(rstatix)

p.list <- list()

# LUAD EGFR
p.list[["LUAD_EGFR"]] <- mean_data %>%
  left_join(traits,"immucan_id") %>%
  filter(!is.na(EGFR),
         simple_histology == "Adenocarcinoma") %>%
  group_by(celltypes, name) %>%
  summarise(
    wilcox_test = list(wilcox_test(mean_exp ~ EGFR, data = cur_data(), detailed = TRUE)),
    logFC = log2(mean(mean_exp[EGFR == TRUE], na.rm = TRUE) / mean(mean_exp[EGFR == FALSE], na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  unnest(wilcox_test) %>%
  mutate(condition = "LUAD_EGFR")

# LUAD KRAS
p.list[["LUAD_KRAS"]] <- mean_data %>%
  left_join(traits,"immucan_id") %>%
  filter(!is.na(KRAS),
         simple_histology == "Adenocarcinoma") %>%
  group_by(celltypes, name) %>%
  summarise(
    wilcox_test = list(wilcox_test(mean_exp ~ KRAS, data = cur_data(), detailed = TRUE)),
    logFC = log2(mean(mean_exp[KRAS == TRUE], na.rm = TRUE) / mean(mean_exp[KRAS == FALSE], na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  unnest(wilcox_test) %>%
  mutate(condition = "LUAD_KRAS")

#LUAD STK11
p.list[["LUAD_STK11"]] <- mean_data %>%
  left_join(traits,"immucan_id") %>%
  filter(!is.na(STK11),
         simple_histology == "Adenocarcinoma") %>%
  group_by(celltypes, name) %>%
  summarise(
    wilcox_test = list(wilcox_test(mean_exp ~ STK11, data = cur_data(), detailed = TRUE)),
    logFC = log2(mean(mean_exp[STK11 == TRUE], na.rm = TRUE) / mean(mean_exp[STK11 == FALSE], na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  unnest(wilcox_test) %>%
  mutate(condition = "LUAD_STK11")

# LUAD TP53
p.list[["LUAD_p53"]] <- mean_data %>%
  left_join(traits,"immucan_id") %>%
  filter(!is.na(TP53),
         simple_histology == "Adenocarcinoma") %>%
  group_by(celltypes, name) %>%
  summarise(
    wilcox_test = list(wilcox_test(mean_exp ~ TP53, data = cur_data(), detailed = TRUE)),
    logFC = log2(mean(mean_exp[TP53 == TRUE], na.rm = TRUE) / mean(mean_exp[TP53 == FALSE], na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  unnest(wilcox_test) %>%
  mutate(condition = "LUAD_TP53")

# LUSC PIK3CA
p.list[["LUSC_PIK3CA"]] <- mean_data %>%
  left_join(traits,"immucan_id") %>%
  filter(!is.na(PIK3CA),
         simple_histology == "Squamous cell carcinoma") %>%
  group_by(celltypes, name) %>%
  summarise(
    wilcox_test = list(wilcox_test(mean_exp ~ PIK3CA, data = cur_data(), detailed = TRUE)),
    logFC = log2(mean(mean_exp[PIK3CA == TRUE], na.rm = TRUE) / mean(mean_exp[PIK3CA == FALSE], na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  unnest(wilcox_test) %>%
  mutate(condition = "LUSC_PIK3CA")

# LUSC TP53
p.list[["LUSC_PIK3CA"]] <- mean_data %>%
  left_join(traits,"immucan_id") %>%
  filter(!is.na(TP53),
         simple_histology == "Squamous cell carcinoma") %>%
  group_by(celltypes, name) %>%
  summarise(
    wilcox_test = list(wilcox_test(mean_exp ~ TP53, data = cur_data(), detailed = TRUE)),
    logFC = log2(mean(mean_exp[TP53 == TRUE], na.rm = TRUE) / mean(mean_exp[TP53 == FALSE], na.rm = TRUE)),
    .groups = "drop"
  ) %>%
  unnest(wilcox_test) %>%
  mutate(condition = "LUSC_TP53")

dat <- do.call(rbind,p.list)

dat <- dat %>%
  group_by(celltypes, condition) %>%
  mutate(p_adjust =p.adjust(p = p,method = "BH"),
         FCcutoff = ifelse(abs(logFC)>0.4,TRUE,FALSE),
         show =  p_adjust < 0.1 & FCcutoff == TRUE)
dat %>%
  filter(celltypes == "Treg",
         condition == "LUSC_PIK3CA")

library(ggrepel)
col_vec1 <- readRDS("../data/col_vec1.rds")
col_vec2 <- readRDS("../data/col_vec2.rds")
col_vec <- c(col_vec1,col_vec2)
col_vec <- col_vec[!duplicated(col_vec)]


p <- dat %>%
  ggplot(aes(x = logFC,y = -log10(p), col= celltypes, shape = factor(condition)))+
  geom_point(size = 4)+
  geom_text_repel(data = dat %>% filter(show == TRUE),aes(x = logFC,y = -log10(p),label= paste(name, celltypes,sep = "-")), col = "black",size = 6.5,box.padding = 0.5)+
  scale_color_manual(values = col_vec)+
  theme_bw()+
  theme(legend.position = "bottom",
        legend.text = element_text(size = 10))

p

ggsave(filename = paste0(out_folder,"S4C_marker_exp_fold_changes_all_celltypes.pdf"),
       plot = p,
       device = "pdf",
       width = 15,
       height = 10)
```

# S4D number HLADR positive tumor cells in EGFR

```{r, fig.width=4, fig.height=8}
sce1 <- readRDS("/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/IMC/Rout/data/sce.rds")
sce1 <- sce1[,sce1$ROI != 4]

counts <- as.data.frame(t(assay(sce1,"exprs")))
head(counts)
dim(counts)

counts$celltypes <- sce1$celltypes
counts$immucan_id <- sce1$immucan_id
counts$sample_id <- sce1$sample_id
counts <- left_join(counts, traits,"immucan_id")

p <- counts %>%
  filter(celltypes == "Tumor", 
         simple_histology == "Adenocarcinoma") %>%
  ggplot(aes(x = Ecad,y = HLADR))+
  geom_point(size = 0.1)+
  facet_wrap(.~EGFR)+
  theme_bw()

counts_EGFR <- counts %>%
  filter(celltypes == "Tumor", 
         simple_histology == "Adenocarcinoma",
         EGFR == TRUE)

counts_EGFRneg <- counts %>%
  filter(celltypes == "Tumor", 
         simple_histology == "Adenocarcinoma",
         EGFR == FALSE)

png(filename = paste0(out_folder,"S4D_HLADR_tumor_cell_EGFRmutated.png"),width = 1000, height = 500,units = "px")
par(mfrow = c(1,2))
LSD::heatscatter(x = counts_EGFR$Ecad, y = counts_EGFR$HLADR,main = "EGFR mutated")
abline(h = 2, lty = 2, lwd = 5)
LSD::heatscatter(x = counts_EGFRneg$Ecad, y = counts_EGFRneg$HLADR, main = "EGFR wild type")
abline(h = 2, lty = 2, lwd = 5)
dev.off()
```

# Ki67 and p53 on tumor cells in LUSC

```{r}
mean_exp1 <- read_csv("/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/Figures/scores_for_sharing/mean_marker_expr_per_celltype_p1.csv")
mean_exp2 <- read_csv("/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/Figures/scores_for_sharing/mean_marker_expr_per_celltype_p2.csv")

mean_exp1$tech <- "IMC1"
mean_exp2$tech <- "IMC2"
mean_exp <- rbind(mean_exp1,mean_exp2)

p_Tumor_Ki67_P53_TP53_LUSC <- mean_exp %>%
  filter(celltypes == "Tumor",
         name %in% c("p53_p2","Ki67_p1","Ki67_p2")) %>%
  left_join(traits, by = "immucan_id") %>%
  filter(simple_histology == "Squamous cell carcinoma", 
         !is.na(TP53)) %>%
  ggplot(aes(x = TP53, y = mean_exp,fill= TP53))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(color = TP53),position = position_jitterdodge(jitter.width = 0.2), size= 1)+
  scale_fill_manual(values = alpha(c("red", "blue"),alpha = 0.6))+
  scale_color_manual(values = c("red", "blue"))+
  facet_wrap(.~name, scales= "free")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  stat_compare_means(size = 2.5)+
    theme_bw()+
  theme(axis.text.y = element_text(size = 6),
        axis.text.x = element_text(size = 6),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 6),
        strip.text = element_text(size = 7))+
  theme(legend.position = "none")
p_Tumor_Ki67_P53_TP53_LUSC

ggsave(filename = paste0(out_folder,"S4E_Ki67_p53_tumor_cells_LUSC.pdf"),
       plot = p_Tumor_Ki67_P53_TP53_LUSC,
       device = "pdf",
       width = 5,
       height = 3)
```


# check IF data for PDL1 expression on Tumor in EGFR

```{r}
files <- list.files("/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/IF/IF2/tsv",pattern = "*.rds",full.names = TRUE)
library(BiocParallel)
cur_files <- bplapply(files,function(x){
  cur_file <- readRDS(x)
  cur_counts <- as.data.frame(t(assay(cur_file,"normalized"))) 
  cur_counts$celltype <- cur_file$celltype

  cur_counts <- cur_counts %>%
    group_by(celltype) %>%
    summarise(across(where(is.numeric),mean))
  cur_counts$immucan_id <- str_split_i(unique(cur_file$sample_id),"-FIXT",1)
  return(cur_counts)
}, BPPARAM = MulticoreParam(workers = 10))

cur_dat <- do.call(rbind,cur_files)
cur_dat <- left_join(cur_dat, traits,"immucan_id")
cur_dat %>%
  filter(simple_histology == "Adenocarcinoma",
         celltype == "Tumor") %>%
  ggplot(aes(x = EGFR,y = PDL1))+
  geom_boxplot()+
  geom_jitter(position = position_dodge2(width = 0.2))+
  theme_bw()
  
```

# check IF data for PD1 expression on CD8 in EGFR
```{r}
cur_dat %>%
  filter(simple_histology == "Adenocarcinoma",
         celltype == "CD8") %>%
  ggplot(aes(x = EGFR,y = PD1))+
  geom_boxplot()+
  geom_jitter(position = position_dodge2(width = 0.2))+
  theme_bw()
```


# TP53 further analyis
Here we investigate the differences between patients that die with TP53 mutation and those that don't

```{r}
traits <- traits %>%
  mutate(TP53_death = ifelse(TP53 == TRUE & censoring == 1,"TP53_poor",NA),
         TP53_death = ifelse(TP53 == TRUE & censoring == 0,"TP53_good",TP53_death))

table(traits$TP53_death, traits$LUAD_vs_LUSC)

table(traits$TP53,traits$metastatic_treatment)
table(traits$censoring,traits$metastatic_treatment)
table(traits$TP53_death,traits$KRAS)
```

```{r}
condition <- c(0)
names(condition) <- c("TP53_death")

pats <- traits %>%
      filter(!is.na(TP53_death)) %>%
  pull(immucan_id)
# loop over conditions and do two way abundance difference tests
p.list <- list()
for (i in c(data_type,"cello1","cello2")) {
  for (j in names(condition)) {
    
    if(grepl("cello",i)){
      abundances <- cello_dat %>%
        select(contains(i))
      abundances <- t(abundances)
      colnames(abundances) <- cello_dat$immucan_id
    } else {
      abundances <- dat %>%
        select(contains(i)) %>%
        filter(complete.cases(.))
      abundances <- t(abundances)
      colnames(abundances) <-dat %>%
        select(contains(i),immucan_id) %>%
        filter(complete.cases(.)) %>%
        pull(immucan_id)
    }
    anno <- traits %>%
      filter(!is.na(TP53_death),
             LUAD_vs_LUSC == "Squamous cell carcinoma")
    
    # subset abundances
    abundances <- abundances[,which(colnames(abundances) %in% anno$immucan_id)]
    
    
    meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
    Size <- factor(meta[,j])
    meta$Size <- Size
    dim(meta)
    
    
    # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
    y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
    design <- model.matrix(~0+group,data = y.ab$samples)
    colnames(design) <- levels(y.ab$samples$group)
    # we do not fit a trend 
    y.ab <- estimateDisp(y.ab, design, trend.method = "none")
    y.ab$common.dispersion
    #plotBCV(y.ab,pch = 5)

    # we only compare the diff abundance for Adeno vs Sqam.
    fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
    fit_LUAD <- glmLRT(glmfit = fit.ab,contrast = c(1,-1))


      sig_dat <- fit_LUAD$table
      sig_dat$celltypes <- rownames(sig_dat)
      sig_dat$cond <- j
      sig_dat$comparison <- grep("TP53",fit_LUAD$comparison,value = TRUE)


  }
}

# define fold change cutoff for plotting
cutoff <- 0

cur_dat <- sig_dat %>%
  mutate(tech = gsub("_abundance","",str_split_i(celltypes,"__",2)),
         cond = "TP53_good_over_poor") %>%
  group_by(tech, cond) %>%
  mutate(FDR = p.adjust(PValue,method = "BH"),
         sig_overall = ifelse(FDR < 0.1 ,TRUE,FALSE),
         sig = ifelse(FDR < 0.1 & abs(logFC) > cutoff,TRUE,FALSE),
         direction = ifelse(logFC > 0 ,"up", "down"),
         ct_name = gsub("_abundance","",str_split_i(celltypes,"__",1)))
  
cur_dat_LUAD <- cur_dat 
to_show_LUAD <- cur_dat %>%
  filter(sig == TRUE) %>%
  pull(celltypes)

cur_dat <- cur_dat %>%
  filter(celltypes %in% to_show_LUAD) %>%
  mutate(celltypes = gsub("_abundance","",celltypes))

# get max and min fold change values
max_up <- max(cur_dat$logFC)+1
min_down <- min(cur_dat$logFC)-1

# plot the data
p <- cur_dat %>%
  ggplot(aes(x=celltypes, y = cond,col = logFC, size= -log10(PValue)))+
  #geom_tile(aes(fill = sig))+
  #theme_bw()+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(round(min_down),0,round(max_up))),
                         guide = "colorbar", limits=c(round(min_down),round(max_up)))+
  #theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  #      panel.background = element_blank(), axis.line = element_line(colour = "black"))+
  #theme_bw()+
  #scale_color_gradient(low = "white",   high = "red")+
  scale_size(range = c(1,8))+
  new_scale_colour() +
  geom_point(data = cur_dat %>% filter(sig == TRUE),shape = 1,aes(size = -log10(PValue), col = sig), stroke =2.5)+
  scale_color_manual(values = c("TRUE" = "black","FALSE" = "white"), name = "FDR < 0.1")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 20),
        xis.text.y = element_text(size = 20),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 20),
        legend.location = "plot",
        axis.title.x = element_blank(),
        axis.title.y = element_blank())+
  coord_flip()
  #theme(plot.margin = unit(c(5, 2.5, 2.5, 2.5), "cm"))
p

sig_dat %>%
  filter(celltypes == "Tumor_IDO1_abundance__IMC1")
```

```{r}
dat %>%
  left_join(traits, "immucan_id") %>%
  ggplot(aes(x = TP53_death,y = log2(Tumor_IDO1_abundance__IMC1)))+
  geom_boxplot()
test
```

# PIK3CA differential expression of  PIK3CA mutated tumors - not included in manuscript

note: we only have RNAseq data for 4 patients that are PIK3CA mutated.
```{r, fig.height=8, fig.width=8}
library(DESeq2)
library(org.Hs.eg.db)
library(clusterProfiler)
counts <- read.table("/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/RNAseq/Rout/tables/RNAseq/rawcounts_proteins.txt", sep = "\t",check.names = FALSE)
# we will fix the version names of the genes
clean_ids <- gsub("\\..*", "", rownames(counts))
rownames(counts) <- clean_ids

# metadata
metadata <- traits[,c("immucan_id","PIK3CA")]

qc_matrix <- read.delim("/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/RNAseq/STAR_counts/qc_matrix.tsv", sep="\t", header=T)
rownames(qc_matrix) <- qc_matrix$ID

seq_matrix <- read.delim("/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/RNAseq/STAR_counts/sequencing_qc.csv", sep=",", header = T)
seq_matrix$lib.kit <- factor(ifelse(grepl("Takara",seq_matrix$assay_library), "takara", "kapa"))
seq_matrix <- unique(seq_matrix[,c("sample_id","date_of_extraction", "number_of_reads", "percent_unique","lib.kit")])
rownames(seq_matrix) <- seq_matrix$sample_id

metadata <- merge(seq_matrix, qc_matrix, by=0, sort=F)
metadata$Row.names <- NULL
metadata$immucan_id <- str_split_i(metadata$sample_id,"-FIXT",1)


# add condition to metadata
metadata <- left_join(metadata,as.data.frame(traits),"immucan_id")
rownames(metadata) <- metadata$sample_id

metadata <- metadata %>%
  filter(sample_id %in% colnames(counts))

stopifnot(all(colnames(counts) == rownames(metadata)))

# factorize treatment
metadata <- metadata %>%
  mutate(PIK3CA = as.factor(ifelse(PIK3CA == TRUE,"mutated", "wildtype")))

# we will only keep LUSC samples
ids <- metadata %>%
  filter(LUAD_vs_LUSC == "Squamous cell carcinoma") %>%
  pull(sample_id)
counts <- counts[,ids]
metadata <- metadata %>%
  filter(sample_id %in% ids)
# perform differential expression analysis
dds <- DESeqDataSetFromMatrix(countData = counts,
                              colData = metadata,
                              design = ~ lib.kit + PIK3CA)
# Filter low-count genes
dds <- dds[rowSums(counts(dds)) > 10,]

dds <- DESeq(dds)

res <- results(dds, contrast = c("PIK3CA", "mutated", "wildtype"))

# Order and remove NAs
res <- res[order(res$pvalue), ]
res <- na.omit(res)

# Add gene symbols as a column if not rownames already
res$symbol <- rownames(res)

# Map SYMBOL to ENTREZID
res$entrez <- mapIds(org.Hs.eg.db,
                     keys = res$symbol,
                     column = "SYMBOL",
                     keytype = "ENSEMBL",
                     multiVals = "first")

# Drop NAs
res <- res[!is.na(res$entrez), ]

# Create ranked list (can use log2FoldChange or stat)
gene_list <- res$log2FoldChange
names(gene_list) <- res$entrez
gene_list <- sort(gene_list, decreasing = TRUE)


gsea_res <- gseGO(geneList = gene_list,
                  OrgDb = org.Hs.eg.db,
                  ont = "BP",
                  keyType = "SYMBOL",
                  minGSSize = 10,
                  maxGSSize = 500,
                  pvalueCutoff = 0.05,
                  verbose = FALSE)

# View top results
head(gsea_res)

dotplot(gsea_res, showCategory = 20, title = "GSEA - GO: Biological Process")
```


```{r}
library(DESeq2)
expr <- read.table(file="/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/RNAseq/Rout/tables/RNAseq/vstcounts_adj_proteins.txt", header=TRUE, stringsAsFactors=FALSE, row.names=1, check.names = F)

MetadataFile <- "/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/RNAseq/Rout/tables/RNAseq/metadata.txt"
```

