---
title: "general stats for NSCLC2"
author: "Daniel Schulz"
date: "2024-08-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggplot2)
library(survivalAnalysis)
library(ggsurvfit)
library(survminer)
library(survival)
```

```{r}
mount_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/"
out_folder <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/Figure_1_supp/"

if (!dir.exists(out_folder)) {
  dir.create(out_folder)
}
```


```{r}
traits <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/all_traits.csv"))
```

# number of cells in mIF
```{r}
dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/abundances_dat.rds"))
dat <- dat[,grepl("IF",colnames(dat))]
x <- sum(rowSums(dat,na.rm = TRUE),na.rm = TRUE)/10^6
x
```

in mIF we have `r round(x)` million cells

# number of cells in IMC
```{r}
dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/abundances_dat.rds"))
dat <- dat[,grepl("IMC1",colnames(dat))]
(x <- sum(rowSums(dat),na.rm = TRUE)/10^6)

dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/abundances_dat.rds"))
dat <- dat[,grepl("IMC2",colnames(dat))]
(y <- sum(rowSums(dat),na.rm = TRUE)/10^6)
```

in IMC panel 1 we have `r round(x)` and in panel 2 `r round(y)` million cells


```{r, eval=FALSE}
sce1 <- readRDS(paste0(mount_path,"/data/IMC/Panel_1/Rout/sce.rds"))
cur_sce <- sce1[,sce1$ROI != 4]
dat <- as.data.frame(sort(table(cur_sce$celltypes)/ncol(cur_sce)))

dat %>%
  filter(Var1 %in% c("Treg","CD4","CD8")) %>%
  summarise(sum(Freq))

dat %>%
  filter(Var1 %in% c("DC","MacCD163", "HLADR","Neutrophil")) %>%
  summarise(sum(Freq))
```

# survival for neo-adjuvant and adjuvant treatment
```{r}
ggforest(coxph( Surv(time_to_event, censoring) ~  neo_adjuvant_treatment,data = traits))
x<- coxph( Surv(time_to_event, censoring) ~  neo_adjuvant_treatment,data = traits)

(p <- summary(x)$coefficients[, "Pr(>|z|)"])
(HR <- summary(x)$coefficients[, "exp(coef)"])
```

Neo-adjuvant treatment has a HR of `r round(HR, digits = 2)` and a p value of `r round(p, digits = 2)`

# survival for adjuvant and adjuvant treatment
```{r}
ggforest(coxph( Surv(time_to_event, censoring) ~  adjuvant_treatment,data = traits))
x<- coxph( Surv(time_to_event, censoring) ~  adjuvant_treatment,data = traits)

(p <- summary(x)$coefficients[, "Pr(>|z|)"])
(HR <- summary(x)$coefficients[, "exp(coef)"])
```

Adjuvant treatment has a HR of `r round(HR, digits = 2)` and a p value of `r round(p, digits = 2)`

# survival for TPS and CPS
```{r}
ggforest(coxph( Surv(time_to_event, censoring) ~  TPS_score,data = traits))
x<- coxph( Surv(time_to_event, censoring) ~  TPS_score,data = traits)
summary(x)

table(traits$TPS_score)/nrow(traits)

(p_TPS <- summary(x)$coefficients[, "Pr(>|z|)"])
(HR_TPS <- summary(x)$coefficients[, "exp(coef)"])

ggforest(coxph( Surv(time_to_event, censoring) ~  CPS_score,data = traits))
x<- coxph( Surv(time_to_event, censoring) ~  CPS_score,data = traits)
summary(x)

(p_CPS <- summary(x)$coefficients[, "Pr(>|z|)"])
(HR_CPS <- summary(x)$coefficients[, "exp(coef)"])
```

TPS > 1% has a HR of `r round(HR_TPS, digits = 2)` and a p value of `r round(p_TPS, digits = 2)` and CPS has a HR of `r round(HR_CPS, digits = 2)` and a p value of `r round(p_CPS, digits = 2)`

# survival for TMB
```{r}
ggforest(coxph( Surv(time_to_event, censoring) ~  TMB_score ,data = traits))
x<- coxph( Surv(time_to_event, censoring) ~  TMB_score* LUAD_vs_LUSC,data = traits)
summary(x)

(p <- summary(x)$coefficients[, "Pr(>|z|)"])
(HR <- summary(x)$coefficients[, "exp(coef)"])
```

TMB > 10 mut/MB has a HR of `r round(HR, digits = 2)` and a p value of `r round(p, digits = 2)` 

# survival for smoking
```{r}
cur_traits <- traits
cur_traits$smoking_status <- factor(cur_traits$smoking_status,levels = c("Non-smoker" ,"Ex-smoker", "Current smoker"))
ggforest(coxph( Surv(time_to_event, censoring) ~  smoking_status,data = cur_traits))
x<- coxph( Surv(time_to_event, censoring) ~  smoking_status,data = cur_traits)
summary(x)

(p <- summary(x)$coefficients[, "Pr(>|z|)"])
(HR <- summary(x)$coefficients[, "exp(coef)"])
```

# association of smoking with TP53 mutation
```{r}
table(traits$smoking_status,traits$TP53)
cur_traits <- traits
cur_traits <- cur_traits %>%
  mutate(smoking_status = ifelse(smoking_status == "Ex-smoker", "smoker",smoking_status),
         smoking_status = ifelse(smoking_status == "Current smoker", "smoker",smoking_status))
table(cur_traits$smoking_status, cur_traits$TP53)
fisher.test(cur_traits$smoking_status, cur_traits$TP53)

cur_traits <- traits[which(traits$smoking_status %in% c("Current smoker", "Non-smoker")),]
fisher.test(cur_traits$smoking_status,cur_traits$TP53)

cur_traits <- traits[which(traits$smoking_status %in% c("Ex-smoker", "Non-smoker")),]
fisher.test(cur_traits$smoking_status,cur_traits$TP53)
```


# TMB and TPS
```{r}
res <- fisher.test(traits$TMB_score,traits$TPS_score)
res$p.value
```

TMB and TPS are not asociated in fisher test `r round(res$p.value, digits = 2)`

# p53 mutaion
```{r}
ggforest(coxph( Surv(time_to_event, censoring) ~  TP53,data = traits %>% filter(simple_histology == "Adenocarcinoma")))
ggforest(coxph( Surv(time_to_event, censoring) ~  TP53,data = traits))
ggforest(coxph( Surv(time_to_event, censoring) ~  TP53,data = traits %>% filter(simple_histology == "Squamous cell carcinoma")))
x<- coxph( Surv(time_to_event, censoring) ~  TP53,data =  traits %>% filter(simple_histology == "Adenocarcinoma"))

(p_LUAD <- summary(x)$coefficients[, "Pr(>|z|)"])
(HR_LUAD <- summary(x)$coefficients[, "exp(coef)"])

y <- coxph( Surv(time_to_event, censoring) ~  TP53,data =  traits %>% filter(simple_histology == "Squamous cell carcinoma"))
(p_LUSC <- summary(y)$coefficients[, "Pr(>|z|)"])
(HR_LUSC <- summary(y)$coefficients[, "exp(coef)"])
```



TP53 has a HR of `r round(HR_LUAD, digits = 2)` and a p value of `r round(p_LUAD, digits = 2)` and CPS has a HR of `r round(HR_LUSC, digits = 2)` and a p value of `r round(p_LUSC, digits = 2)`

# percent of TPS > 1 % patients
```{r}
table(traits$TPS_score)/nrow(traits)
```

```{r}
cur_traits <- traits %>%
  filter(simple_histology == "Adenocarcinoma")
table(cur_traits$smoking_status, cur_traits$TPS_score)
```


# check stuff - to be removed eventually

```{r, fig.height=20}
dat <- read.csv("/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/Figures/scores_for_sharing/cell-pair_interactions_celltypes_minor_directional_P1.csv")
dat <- read.csv("/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/Figures/scores_for_sharing/cell-pair_interactions_celltypes_directional_P1.csv")
dat <- read.csv("/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/Figures/scores_for_sharing/cell-cell-pair_interactions_directional_IF1.csv")
dat <- dat %>%
  select(immucan_id, contains("MacCD163"))

p.list <- list()
for (i in colnames(dat[,-1]) ) {
  cur_dat <- dat %>%
    select(immucan_id,all_of(i)) %>%
    mutate(test = as.factor(ntile(.data[[i]],2))) %>%
    left_join(traits, "immucan_id")
  surv_ob <- Surv(cur_dat$time_to_event,cur_dat$censoring)
 p.list[[i]] <- coxph( as.formula(paste0("surv_ob ~ + test")),data = cur_dat)
}

df <- data.frame(HR =      unlist(lapply(p.list, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p.list, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "upper .95"]}))) 
df %>%
  mutate(name = rownames(df)) %>%
  filter(grepl("Tumor", name)) %>%
  arrange(HR)

dat %>%
  select(contains("MacCD163.Tumor"))


cur_dat <- dat %>%
    select(immucan_id,all_of("MacCD163_PDL1.Tumor_CD15_Ki67")) %>%
    mutate(test = as.factor(ntile(MacCD163_PDL1.Tumor_CD15_Ki67,2))) %>%
    left_join(traits, "immucan_id")
ggforest(coxph( Surv(time_to_event, censoring) ~  test,data = cur_dat ))

surv_fit <- survfit(Surv(time_to_event, censoring) ~ test, data = cur_dat)

ggsurv <- ggsurvplot(surv_fit, data = cur_dat, risk.table = TRUE, pval = TRUE)
p <- ggsurv$plot
p
table(cur_dat$censoring,cur_dat$test)


# PDGFRb analysis

# mean exp
exp <- read.csv("/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/Figures/scores_for_sharing/mean_marker_expr_per_celltype_p2.csv")
cur_dat <- exp %>%
  filter(celltypes == "Vasculature",name == "CD31_p2") %>%
  mutate(test = as.factor(ntile(mean_exp,2))) %>%
  left_join(traits,"immucan_id") %>%
  filter(simple_histology == "Adenocarcinoma")
surv_fit <- survfit(Surv(time_to_event, censoring) ~ test, data = cur_dat)

ggsurv <- ggsurvplot(surv_fit, data = cur_dat, risk.table = TRUE, pval = TRUE)
ggsurv$plot

cur_dat <- exp %>%
  filter(name == "CD31_p2") %>%
  group_by(immucan_id) %>%
  summarise(mean_exp = mean(mean_exp)) %>%
  ungroup() %>%
  mutate(test = as.factor(ntile(mean_exp,2))) %>%
  left_join(traits,"immucan_id") %>%
  filter(simple_histology == "Squamous cell carcinoma")
surv_fit <- survfit(Surv(time_to_event, censoring) ~ test, data = cur_dat)

ggsurv <- ggsurvplot(surv_fit, data = cur_dat, risk.table = TRUE, pval = TRUE)
ggsurv$plot

cur_dat <- exp %>%
  filter(name == "PDGFRb_p1")  %>%
  pivot_wider(id_cols = c(immucan_id),values_from =mean_exp, names_from = celltypes) %>%
  select(-immucan_id)
pheatmap::pheatmap(cur_dat)

cur_dat <- exp %>%
  filter(celltypes == "Mural",name == "PDGFRb_p1") %>%
  left_join(dens_dat, "immucan_id") %>%
  select(-immucan_id,-celltypes, -name)
  
plot_dat <- data.frame(cors = apply(cur_dat[,-1], 2, function(x) cor(x,cur_dat$mean_exp,use = "complete")))
rownames(plot_dat) <- names(apply(cur_dat[,-1], 2, function(x) cor(x,cur_dat$mean_exp,use = "complete")))
pheatmap::pheatmap(plot_dat, cluster_cols = FALSE)

# correlation with Treg_density! I checked and the Treg_density is not prognostic.
cur_dat <- dens_dat %>%
  select(Treg_density__IMC1, immucan_id) %>%
  left_join(traits, "immucan_id") %>%
  mutate(test = as.factor(ntile(Treg_density__IMC1, 2)))
surv_fit <- survfit(Surv(time_to_event, censoring) ~ test, data = cur_dat)

ggsurv <- ggsurvplot(surv_fit, data = cur_dat, risk.table = TRUE, pval = TRUE)
ggsurv$plot

```


