---
title: "Data processing"
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: inline
---


NOTE: The IF data is large and is stored in a different location than the IMC, RNAseq and WES data for NSCLC2 to avoid duplication of data.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r}
library(tidyverse)
library(stringr)
library(R.utils)
library(SingleCellExperiment)
library(readr)
library(BiocParallel)
```

```{r}
mount_path <- "/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/IF/"
```


# sample removal
We will generally remove the following samples due to low cell count, low quality or disrupted tissues. this list has been manually curated after a first round of data processing and visualization of the `plotSpatial` images and looking at overall cell numbers. many samples from NSCLC cohort had below 10000 cells per panel and often very disrupted non-connected tissues.


```{r}
sample_removal <- c()
```

LUNG-NSCLC2-0669-FIXT-01-IF2-01 has mechanical damage on the tissue

In this script, we will read in the mIF data, do some initial QC and save
data as `SingleCellExperiment` objects.

# IF Panel 1

## check if and what files are there
```{r}
data_path <- paste0(mount_path, "IF1/tsv")
cur_files <- list.files(data_path, pattern = ".*properties_#_.*.gz$")
cur_files_2 <- list.files(data_path, pattern = ".*properties2_#_.*.gz$")
IF1 <- str_split(cur_files,pattern = "_#_",simplify = TRUE)[,1]
IF1_2 <- str_split(cur_files_2,pattern = "_#_",simplify = TRUE)[,1]

IF1_pat <- str_split(IF1,"-FIXT",simplify = T)[,1]

# check for unique samples per patient (not sample id)
paste0("no duplicated patient samples: ", length(IF1_pat) == length(unique(IF1_pat)))

#a check if all properties files are available
paste0("files for which no properties2 file ist available: ",length(IF1[!IF1 %in% IF1_2]))
# there was a properties2 file missing for one sample.

paste0("Number of sample IDs: ",length(IF1))
paste0("Unique number of sample IDs: ",length(unique(IF1)))
paste0("Number of NSCLC2 samples: ",sum(grepl("NSCLC2-",IF1)))
paste0("Number of NSCLC samples: ",sum(grepl("NSCLC-",IF1)))

IF1_x <- str_split(cur_files,pattern = "-FIXT",simplify = TRUE)[,1]
paste0("patients with more than one sample: ",paste0(IF1_x[duplicated(IF1_x)], collapse = "; "))
```


## generate single cell files for each patient

we have gone through the data and identified all the possible columns that the individual tsv files have. to generate `SingleCellExperiment`objects we will load each file and generate the set of `colData` entries in  the right order to merge them later on.
the metadata files ("properties2") contains the same column entries across all files. we will also read this file and add the information to the single cell object

```{r, message=FALSE}
# read phenotype key
key <- read_csv("data/phenotype_key_IF1.csv")
qc_chuv <- read.csv(paste0(mount_path,"IF_QC_CHUV/qc_matrix_IF1.txt"), sep = "\t")

data_path <- paste0(mount_path, "IF1/tsv")
cur_files <- list.files(data_path, full.names = TRUE, pattern = ".*properties_#_.*.gz$")

bplapply(cur_files,function(x){
  cur_name <- str_split(basename(x), "_", simplify = TRUE)[,1]
  cur_rds <- sub(".tsv.gz$", ".rds", x)
  
  cur_data <- read_tsv(x)
  cur_meta <- cur_data[,!grepl("score", colnames(cur_data))]

  cur_meta$sample_id <- cur_name
  cur_exprs <- cur_data[,grepl("score.normalized", colnames(cur_data))]

   if(!"in.ROI.necrosis" %in% colnames(cur_meta)){
    cur_meta$in.ROI.necrosis <- NA
  }
  if(!"in.ROI.next_to_tumor_tissue" %in% colnames(cur_meta)){
    cur_meta$in.ROI.next_to_tumor_tissue <- NA
  }
  if(!"in.ROI.tumor_tissue" %in% colnames(cur_meta)){
    cur_meta$in.ROI.tumor_tissue <- NA
  }
  if(!"sample_id" %in% colnames(cur_meta)){
    cur_meta$sample_id <- NA
  }
  if(!"in.ROI.adipose_tissue" %in% colnames(cur_meta)){
    cur_meta$in.ROI.adipose_tissue <- NA
  }
  if(!"in.ROI.tls" %in% colnames(cur_meta)){
    cur_meta$in.ROI.tls <- NA
  }
#  if(!"in.ROI.1" %in% colnames(cur_meta)){
#    cur_meta$in.ROI.1 <- NA
#  }
#  if(!"in.ROI.2" %in% colnames(cur_meta)){
#    cur_meta$in.ROI.2 <- NA
#  }
#  if(!"in.ROI.3" %in% colnames(cur_meta)){
#    cur_meta$in.ROI.3 <- NA
#  }

  # now we will read the metadata file
  j <- list.files(data_path, full.names = TRUE)[grepl(paste0(cur_name,".*properties2.*.gz$"),list.files(data_path))]
  cur_meta_name <- str_split(basename(j), "_", simplify = TRUE)[1,1]

  cur_metadata <- read_tsv(j)
  if(dim(cur_metadata)[1] == dim(cur_meta)[1]){
      cur_meta <- left_join(cur_meta,cur_metadata[,c("cell.ID","cell.area","TLS.ID")],"cell.ID")

      cur_meta <- cur_meta[,c("sample_id","cell.ID","nucleus.x","nucleus.y","cell.area","tissue.type","phenotype",
                              "in.ROI.necrosis","in.ROI.next_to_tumor_tissue","in.ROI.tumor_tissue",
                              "in.ROI.adipose_tissue","in.ROI.tls","TLS.ID")]
  } else if (dim(cur_metadata)[1] != dim(cur_meta)[1]) {
    cur_meta$cell.area <- NA
    cur_meta$TLS.ID <- NA

    cur_meta <- cur_meta[,c("sample_id","cell.ID","nucleus.x","nucleus.y","cell.area","tissue.type","phenotype",
                            "in.ROI.necrosis","in.ROI.next_to_tumor_tissue","in.ROI.tumor_tissue",
                            "in.ROI.adipose_tissue","in.ROI.tls","TLS.ID")]
  }

  # check that the phenotype column is ordered the same
  cur_pheno <- str_extract_all(cur_meta$phenotype, "[A-Za-z]{1,3}[0-9]{0,3}\\W|[A-Za-z]{1,3}[0-9]{0,3}c{1}\\W", simplify = TRUE)

  colnames(cur_pheno) <- as.character(sapply(cur_pheno[1,],function(x) gsub("\\+|-","",x)))

  cur_pheno <- cur_pheno[,c("CD15", "CK", "CD3", "CD11c", "CD20", "CD163")]

  cur_pheno <- apply(cur_pheno, 1, function(x) paste0(x, collapse = ""))
  cur_meta$phenotype <- cur_pheno

  # Collect positivity
  cur_meta$CD3 <- str_detect(cur_meta$phenotype, "CD3\\+")
  cur_meta$CD15 <- str_detect(cur_meta$phenotype, "CD15\\+")
  cur_meta$CK <- str_detect(cur_meta$phenotype, "CK\\+")
  cur_meta$CD163 <- str_detect(cur_meta$phenotype, "CD163\\+")
  cur_meta$CD11c <- str_detect(cur_meta$phenotype, "CD11c\\+")
  cur_meta$CD20 <- str_detect(cur_meta$phenotype, "CD20\\+")

  cur_sce <- SingleCellExperiment(assays = list("normalized" = t(cur_exprs)))
  colData(cur_sce) <- as(cur_meta, "DataFrame")
  rownames(cur_sce) <- str_split(rownames(cur_sce), "\\.", simplify = TRUE)[,1]

  # add phenotype information
  colData(cur_sce) <- DataFrame(left_join(as.data.frame(colData(cur_sce)), key, by = "phenotype"))
  colnames(cur_sce) <- paste0(cur_sce$sample_id,"_",cur_sce$cell.ID)

  
  # add qc_chuv to metadata
  id <- str_extract_all(cur_name, "\\b\\d{4}\\b")[[1]]
  markers_qc <- paste0("qc_",c("CD3", "CD15", "CK", "CD163", "CD11c", "CD20"))
  qc <- qc_chuv %>%
    filter(grepl(id, sample_id)) %>%
    dplyr::select(all_of(c("sample_id","qc_reception_status", "qc_staining_status", "qc_scanning_status", "qc_scanning_comment", "qc_analysis_comment", "qc_analysis_status", markers_qc))) %>%
    distinct()
  
  metadata(cur_sce)$qc_chuv <- qc
  
  # other qc
  cur_sce$flag_no_cells <- rep(ifelse(ncol(cur_sce) < 20000,TRUE,FALSE),ncol(cur_sce))
  
  saveRDS(cur_sce, cur_rds)
  
},BPPARAM = MulticoreParam(workers = 10))

```


# IF Panel 2


## check if and what files are there
```{r}
data_path <- paste0(mount_path, "IF2/tsv")
cur_files <- list.files(data_path, pattern = ".*properties_#_.*.gz$")
cur_files_2 <- list.files(data_path, pattern = ".*properties2_#_.*.gz$")
IF2 <- str_split(cur_files,pattern = "_#_",simplify = TRUE)[,1]
IF2_2 <- str_split(cur_files_2,pattern = "_#_",simplify = TRUE)[,1]

IF2_pat <- str_split(IF2,"-FIXT",simplify = T)[,1]

# check for unique samples per patient (not sample id)
paste0("no duplicated patient samples: ", length(IF2_pat) == length(unique(IF2_pat)))
IF2_pat[duplicated(IF2_pat)]

#a check if all properties files are available
paste0("files for which no properties2 file ist available: ",length(IF2[!IF2 %in% IF2_2]))


paste0("Number of sample IDs: ",length(IF2))
paste0("Unique number of sample IDs: ",length(unique(IF2)))
paste0("Number of NSCLC2 samples: ",sum(grepl("NSCLC2-",IF2)))
paste0("Number of NSCLC samples: ",sum(grepl("NSCLC-",IF2)))

IF2_x <- str_split(cur_files,pattern = "-FIXT",simplify = TRUE)[,1]
paste0("patients with more than one sample: ",paste0(IF2_x[duplicated(IF2_x)], collapse = "; "))
```


## generate single cell files for each patient

```{r, message=FALSE}
# read phenotype key
key <- read_csv("data/phenotype_key_IF2.csv")
qc_chuv <- read.csv(paste0(mount_path,"IF_QC_CHUV/qc_matrix_IF2.txt"), sep = "\t")

data_path <- paste0(mount_path, "IF2/tsv")
cur_files <- list.files(data_path, full.names = TRUE, pattern = ".*properties_#_.*.gz$")


bplapply(cur_files,function(x){
  cur_name <- str_split(basename(x), "_", simplify = TRUE)[,1]
  cur_rds <- sub(".tsv.gz$", ".rds", x)

  cur_data <- read_tsv(x)
  cur_meta <- cur_data[,!grepl("score", colnames(cur_data))]

  cur_meta$sample_id <- cur_name
  cur_exprs <- cur_data[,grepl("score.normalized", colnames(cur_data))]

   if(!"in.ROI.necrosis" %in% colnames(cur_meta)){
    cur_meta$in.ROI.necrosis <- NA
  }
  if(!"in.ROI.next_to_tumor_tissue" %in% colnames(cur_meta)){
    cur_meta$in.ROI.next_to_tumor_tissue <- NA
  }
  if(!"in.ROI.tumor_tissue" %in% colnames(cur_meta)){
    cur_meta$in.ROI.tumor_tissue <- NA
  }
  if(!"in.ROI.adipose_tissue" %in% colnames(cur_meta)){
    cur_meta$in.ROI.adipose_tissue <- NA
  }

    # now we will read the metadata file
  j <- list.files(data_path, full.names = TRUE)[grepl(paste0(cur_name,".*properties2.*.gz$"),list.files(data_path))]
  cur_meta_name <- str_split(basename(j), "_", simplify = TRUE)[,1]

  cur_metadata <- read_tsv(j)

  if(dim(cur_metadata)[1] == dim(cur_meta)[1]){
    cur_meta <- left_join(cur_meta,cur_metadata[,c("cell.ID","cell.area")],"cell.ID")

    cur_meta <- cur_meta[,c("sample_id","cell.ID","nucleus.x","nucleus.y","cell.area","tissue.type","phenotype",
                            "in.ROI.necrosis","in.ROI.next_to_tumor_tissue","in.ROI.tumor_tissue",
                            "in.ROI.adipose_tissue")]
  } else if (dim(cur_metadata)[1] != dim(cur_meta)[1]) {
    cur_meta$cell.area <- NA

    cur_meta <- cur_meta[,c("sample_id","cell.ID","nucleus.x","nucleus.y","cell.area","tissue.type","phenotype",
                            "in.ROI.necrosis","in.ROI.next_to_tumor_tissue","in.ROI.tumor_tissue",
                            "in.ROI.adipose_tissue")]
  }

  # check that the phenotype column is ordered the same
  cur_pheno <- str_extract_all(cur_meta$phenotype, "[A-Za-z]{1,3}[0-9]{0,3}\\W", simplify = TRUE)
  colnames(cur_pheno) <- as.character(sapply(cur_pheno[1,],function(x) gsub("\\+|-","",x)))

  cur_pheno <- cur_pheno[,c("PD1", "CK","Ki67", "GB", "CD8", "PDL1")]

  cur_pheno <- apply(cur_pheno, 1, function(x) paste0(x, collapse = ""))
  cur_meta$phenotype <- cur_pheno

  # Collect positivity
  cur_meta$PD1 <- str_detect(cur_meta$phenotype, "PD1\\+")
  cur_meta$CK <- str_detect(cur_meta$phenotype, "CK\\+")
  cur_meta$Ki67 <- str_detect(cur_meta$phenotype, "Ki67\\+")
  cur_meta$GB <- str_detect(cur_meta$phenotype, "GB\\+")
  cur_meta$CD8 <- str_detect(cur_meta$phenotype, "CD8\\+")
  cur_meta$PDL1 <- str_detect(cur_meta$phenotype, "PDL1\\+")

  cur_sce <- SingleCellExperiment(assays = list("normalized" = t(cur_exprs)))
  colData(cur_sce) <- as(cur_meta, "DataFrame")
  rownames(cur_sce) <- str_split(rownames(cur_sce), "\\.", simplify = TRUE)[,1]

  # add phenotype information
  colData(cur_sce) <- DataFrame(left_join(as.data.frame(colData(cur_sce)), key, by = "phenotype"))
  colnames(cur_sce) <- paste0(cur_sce$sample_id,"_",cur_sce$cell.ID)
    
  # add qc_chuv to metadata
  id <- str_extract_all(cur_name, "\\b\\d{4}\\b")[[1]]
  markers_qc <- paste0("qc_",c("PD1", "CK", "Ki67", "GB", "CD8", "PDL1"))
  qc <- qc_chuv %>%
    filter(grepl(id, sample_id)) %>%
    dplyr::select(all_of(c("sample_id","qc_reception_status", "qc_staining_status", "qc_scanning_status", "qc_scanning_comment", "qc_analysis_comment", "qc_analysis_status", markers_qc))) %>%
    distinct()
  
  metadata(cur_sce)$qc_chuv <- qc
  
  # other qc
  cur_sce$flag_no_cells <- rep(ifelse(ncol(cur_sce) < 20000,TRUE,FALSE),ncol(cur_sce))
  
  saveRDS(cur_sce, cur_rds)
  
},BPPARAM = MulticoreParam(workers = 10))

gc()
```

#  IF panel 3

## check if and what files are there
```{r}
data_path <- paste0(mount_path, "IF3/tsv")
cur_files <- list.files(data_path, pattern = ".*properties_#_.*.gz$")
cur_files_2 <- list.files(data_path, pattern = ".*properties2_#_.*.gz$")
IF3 <- str_split(cur_files,pattern = "_#_",simplify = TRUE)[,1]
IF3_2 <- str_split(cur_files_2,pattern = "_#_",simplify = TRUE)[,1]

IF3_pat <- str_split(IF3,"-FIXT",simplify = T)[,1]

# check for unique samples per patient (not sample id)
paste0("no duplicated patient samples: ", length(IF3_pat) == length(unique(IF3_pat)))
IF3_pat[duplicated(IF3_pat)]

#a check if all properties files are available
paste0("files for which no properties2 file ist available: ",length(IF3[!IF3 %in% IF3_2]))
IF3[!IF3 %in% IF3_2]

paste0("Number of sample IDs: ",length(IF3))
paste0("Unique number of sample IDs: ",length(unique(IF3)))
paste0("Number of NSCLC2 samples: ",sum(grepl("NSCLC2-",IF3)))
paste0("Number of NSCLC samples: ",sum(grepl("NSCLC-",IF3)))

IF3_x <- str_split(cur_files,pattern = "-FIXT",simplify = TRUE)[,1]
paste0("patients with more than one sample: ",paste0(IF3_x[duplicated(IF3_x)], collapse = "; "))
```


## generate single cell files for each patient

```{r, message=FALSE}
key <- read_csv("data/phenotype_key_IF3.csv")
qc_chuv <- read.csv(paste0(mount_path,"IF_QC_CHUV/qc_matrix_IF3.txt"), sep = "\t")

data_path <- paste0(mount_path, "IF3/tsv")
cur_files <- list.files(data_path, full.names = TRUE, pattern = ".*properties_#_.*.gz$")


bplapply(cur_files,function(x){
  cur_name <- str_split(basename(x), "_", simplify = TRUE)[,1]
  cur_rds <- sub(".tsv.gz$", ".rds", x)

  cur_data <- read_tsv(x)
  cur_meta <- cur_data[,!grepl("score", colnames(cur_data))]

  cur_meta$sample_id <- cur_name
  cur_exprs <- cur_data[,grepl("score.normalized", colnames(cur_data))]

   if(!"in.ROI.necrosis" %in% colnames(cur_meta)){
    cur_meta$in.ROI.necrosis <- NA
  }
  if(!"in.ROI.next_to_tumor_tissue" %in% colnames(cur_meta)){
    cur_meta$in.ROI.next_to_tumor_tissue <- NA
  }
  if(!"in.ROI.tumor_tissue" %in% colnames(cur_meta)){
    cur_meta$in.ROI.tumor_tissue <- NA
  }
  if(!"in.ROI.adipose_tissue" %in% colnames(cur_meta)){
    cur_meta$in.ROI.adipose_tissue <- NA
  }

  # now we will read the metadata file
  if (identical(list.files(data_path, full.names = TRUE)[grepl(paste0(cur_name,".*properties2.*.gz$"),list.files(data_path))], character(0))) {
    cur_meta$cell.area <- NA
    cur_meta <- cur_meta[,c("sample_id","cell.ID","nucleus.x","nucleus.y","cell.area","tissue.type","phenotype",
                            "in.ROI.necrosis","in.ROI.next_to_tumor_tissue","in.ROI.tumor_tissue",
                            "in.ROI.adipose_tissue")]
  } else {
    j <- list.files(data_path, full.names = TRUE)[grepl(paste0(cur_name,".*properties2.*.gz$"),list.files(data_path))]
    cur_meta_name <- str_split(basename(j), "_", simplify = TRUE)[,1]

    cur_metadata <- read_tsv(j)

    if(dim(cur_metadata)[1] == dim(cur_meta)[1]){
      cur_meta <- left_join(cur_meta,cur_metadata[,c("cell.ID","cell.area")],"cell.ID")

      cur_meta <- cur_meta[,c("sample_id","cell.ID","nucleus.x","nucleus.y","cell.area","tissue.type","phenotype",
                              "in.ROI.necrosis","in.ROI.next_to_tumor_tissue","in.ROI.tumor_tissue",
                              "in.ROI.adipose_tissue")]
    } else if (dim(cur_metadata)[1] != dim(cur_meta)[1]) {
      cur_meta$cell.area <- NA

      cur_meta <- cur_meta[,c("sample_id","cell.ID","nucleus.x","nucleus.y","cell.area","tissue.type","phenotype",
                              "in.ROI.necrosis","in.ROI.next_to_tumor_tissue","in.ROI.tumor_tissue",
                              "in.ROI.adipose_tissue")]
    }
  }


  # check that the phenotype column is ordered the same
  cur_pheno <- str_extract_all(cur_meta$phenotype, "[A-Za-z]{1,4}[0-9]{0,3}\\W", simplify = TRUE)
  colnames(cur_pheno) <- as.character(sapply(cur_pheno[1,],function(x) gsub("\\+|-","",x)))

  cur_pheno <- cur_pheno[,c("CD4", "CK","CD3", "CD56", "CD8", "FOXP3")]

  cur_pheno <- apply(cur_pheno, 1, function(x) paste0(x, collapse = ""))
  cur_meta$phenotype <- cur_pheno

  # Collect positivity
  cur_meta$CD4 <- str_detect(cur_meta$phenotype, "CD4\\+")
  cur_meta$CK <- str_detect(cur_meta$phenotype, "CK\\+")
  cur_meta$CD3 <- str_detect(cur_meta$phenotype, "CD3\\+")
  cur_meta$CD56 <- str_detect(cur_meta$phenotype, "CD56\\+")
  cur_meta$CD8 <- str_detect(cur_meta$phenotype, "CD8\\+")
  cur_meta$FOXP3 <- str_detect(cur_meta$phenotype, "FOXP3\\+")

  cur_sce <- SingleCellExperiment(assays = list("normalized" = t(cur_exprs)))
  colData(cur_sce) <- as(cur_meta, "DataFrame")
  rownames(cur_sce) <- str_split(rownames(cur_sce), "\\.", simplify = TRUE)[,1]

  # add phenotype information
  colData(cur_sce) <- DataFrame(left_join(as.data.frame(colData(cur_sce)), key, by = "phenotype"))
  colnames(cur_sce) <- paste0(cur_sce$sample_id,"_",cur_sce$cell.ID)

  # add qc_chuv to metadata
  id <- str_extract_all(cur_name, "\\b\\d{4}\\b")[[1]]
  markers_qc <- paste0("qc_",c("CD4", "CK", "CD3", "CD56", "CD8", "FOXP3"))
  qc <- qc_chuv %>%
    filter(grepl(id, sample_id)) %>%
    dplyr::select(all_of(c("sample_id","qc_reception_status", "qc_staining_status", "qc_scanning_status", "qc_scanning_comment", "qc_analysis_comment", "qc_analysis_status", markers_qc))) %>%
    distinct()
  
  metadata(cur_sce)$qc_chuv <- qc
  # other qc
  cur_sce$flag_no_cells <- rep(ifelse(ncol(cur_sce) < 20000,TRUE,FALSE),ncol(cur_sce))
  
  saveRDS(cur_sce, cur_rds)
  
},BPPARAM = MulticoreParam(workers = 10))
```
