---
title: "imi_nsclc2_wes_overview"
author: "Timo Trefzer <timo.trefzer@bayer.com>"
date: '`r format(Sys.time(), "%d %B, %Y")`'
output:
  html_notebook:
    code_folding: hide
    highlight: haddock
    theme: yeti
    toc: true
    toc_depth: 4
    number_sections: true
    toc_float: true
    fig_height: 5
    fig_width: 8
    #css: styles.css
  html_document:
    highlight: haddock
    theme: journal
    toc: true
    toc_depth: 4
    number_sections: keep
    fig_height: 5
    fig_width: 8
    css: styles.css
  rmdformats::readthedown:
    code_folding: hide
    self_contained: true
    thumbnails: true
    lightbox: true
    gallery: true
    highlight: tango
    toc_depth: 4
    use_bookdown: true
    css: readthedown.css
  pdf_document:
    latex_engine: xelatex
    toc: true
    toc_depth: 4
    number_sections: keep
    fig_height: 5
    fig_width: 8
csl: vancouver.csl
---

```{r setup, include = FALSE}
library(pander)
library(ggplot2)
theme_set(theme_minimal())
```
  
-------------------------------------------------------------------------------

********************************************************************************
  


# Introduction

Whole exome sequencing data was processed by the "vegan-pipelin" for calling and annotation of genomic variants.  
  
In this notebook variants are filtered by folowing criteria:  
    - Only variants that passed "vegan" quality control are kept ("PASS" filter)
    - Variants with a population allele frequency above 1%, as by "vegan" from the gnomAD data base are removed from further analysis.  
    - Only variants included in the COSMIC data base as annotated by the "vegan-pipeline" are kept for further analysis
    - Variants are further filtered to only include functionally relevant alterations for downstream analysis

   
  
********************************************************************************

\newpage

# Libraries
```{r loading_libraries, message=FALSE}
library(data.table)
library(dplyr)
library(ComplexHeatmap)
library(tidyr)
library(ggtext)
```

# Variables and directories
```{r}
# The folowing firectories need to change according to the setup
mount_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/WES/data/"
path_output <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/WES/data/Rout/"


#following directories are set automatically
path_data_clin <- file.path( "/mnt/immucan_volume/clinical_data/Lung_NSCLC2/20230825/NSCLC2_for_R_update_2024.csv")
path_data_tmb <- file.path("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/WES/data/tmb")
path_vegan <-file.path("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/WES/data/tableReports") # results of main vegan run


# Define driver Genes
Genes_driver <- c("KRAS", "EGFR", "BRAF", "MET", "ALK", "ROS", "RET", "NTRK", "NRG1", "FGFR1", "FGFR2", "HRAS", "NRAS", "MAP2K1", "ERBB2", "ROS1", "PIK3CA", "NTRK1", "STK11")
Genes_sup <- c("TP53", "KEAP1","STK11","RB1","CDKN2A", "CDKN2B","NKX2", "NF1")

genes_sel <- c(Genes_driver, Genes_sup)
```


# Main Program with Plots and Results

## Load data

### Clinical

```{r clinical_data}
data_clin <- fread(path_data_clin)
data_clin
```

### VeganID to IMMUcanID
```{r}
vegan_lookup <- list.files(path_vegan, pattern = "design.csv", full.names = T)
vegan_lookup <- lapply(vegan_lookup, fread)
vegan_lookup <- do.call("rbind", vegan_lookup)
vegan_lookup[["immucan_id"]] <- gsub("-FIXT.*", "", vegan_lookup[["PAIR_ID"]])
vegan_lookup
```


### TMB

```{r}
data_tmb <- list.files(path_data_tmb, pattern = "tmb.txt", recursive = T, full.names = T)
data_tmb <- lapply(data_tmb, function(x){
  name <- gsub("_Mutect2.*", "", 
               gsub(".*tmb/", "", x)
               )
  data_tmp <- read.delim(x) %>%
    separate(col = 1, sep = "=", into = c("a", "b")) %>%
    filter(a == "TMB")
  
  data_tmp <- data_tmp[,"b",drop=T]
  data_tmp <- as.numeric(gsub(" ", "", data_tmp))

  
  names(data_tmp) <- name
  
  return(data_tmp)
  
  
})

data_tmb <-unlist(data_tmb)

data_tmb <- data.frame(TUMOR_ID = names(data_tmb),
           TMB = data_tmb)

data_tmb <- merge(data_tmb, vegan_lookup, all.x = T, by = "TUMOR_ID")

data_tmb
```

### Combine meta-data per sample
```{r}
data_meta <- merge(data_clin, data_tmb, by="immucan_id") # TMB data
data_meta
```

```{r}
data_mut <- list.files(path = path_vegan, 
                       pattern = "Mutect2_filtered_pass_norm_COSMIC_CancerHotspots_GnomAD_dbNSFP.tsv",
                       recursive = T,
                       full.names = T)


data_mut <- lapply(data_mut, function(path){
  
  name <- gsub("_.*", "", 
               gsub(".*tableReports/", "", path)
               )
  data_tmp <- fread(path)
  colnames(data_tmp) <- c(colnames(data_tmp)[1:14],"AD","DP","AF")
  data_tmp$TUMOR_ID <- name
  
  return(data_tmp)
  
})

data_mut <- do.call("rbind", data_mut)

data_mut <- merge(data_mut, data_meta, all.x = T, by = "TUMOR_ID")
data_mut <-  as.data.frame(data_mut)

# Format some column names
#colnames(data_mut) <- gsub("ANN\\[0\\]\\.", "", colnames(data_mut))
data_mut <- data_mut %>%
  transform(AF = as.numeric(AF),
            DP = as.numeric(DP))

data_mut
```

### Filter mutations {.tabsets}

#### mutation frequency

coverage of selected gene variants

```{r}
data_mut %>%
  filter(Gene %in% genes_sel,
         FILTER_AF == "keep",
         FILTER_COSMIC == "keep",
         FILTER_Effect == "keep",
         ) %>%
  ggplot(aes(x = Gene,y = DP))+
  geom_boxplot()+
  theme(axis.text.x = element_text(angle = 90,vjust = 0.5,hjust = 1))+
  geom_hline(yintercept = 75)
data_mut %>%
  filter(Gene %in% genes_sel,
         FILTER_AF == "keep",
         FILTER_COSMIC == "keep",
         FILTER_Effect == "keep",
         ) %>%
  summarise(med_var = median(DP),
            min_var = min(DP))
```

Remove all variants that do not occur in at least 5% of the reads
```{r}

hist(data_mut$DP)
data_mut[["FILTER_DP"]] <- ifelse(data_mut$DP >= 75, "keep", "remove")
data_mut[["FILTER_AF"]] <- ifelse(data_mut$AF >= 0.1, "keep", "remove")
data_mut

# check if our selected Genes are affected by the AF filter
data_mut %>%
  filter(Gene %in% genes_sel) %>%
  select(Effect,Gene,HGVS_C,AF,AD, FILTER_AF) %>%
  filter(FILTER_AF == "remove")

```

#### COSMIC
Remove all variants that are not described in COSMIC
```{r}
data_mut$FILTER_COSMIC <- ifelse(data_mut$COSMIC == "TRUE",
                                   "keep", "remove"
                                   )
```

#### Functional filter

```{r fig.height=15}
data_plot <- data_mut %>% 
  group_by(Effect) %>% 
  dplyr::count() %>% 
  as.data.frame()

ggplot(data=data_plot, aes(x=reorder(Effect,-n), y=n)) +
  geom_bar(stat="identity", fill="grey") +
  theme(axis.text.x = element_markdown(angle = 90, vjust = 0.5, hjust=1), axis.title.x = element_blank()) +
  scale_y_log10()
```

Keep following functional annotations

```{r fig.height=13}
keep_func = c("missense_variant", 
              "stop_gained", 
            #  "synonymous_variant", 
            #  "downstream_Gene_variant", 
              "disruptive_inframe_insertion",
            #  "upstream_Gene_variant",
              "frameshift_variant",
            #  "intron_variant",
            #  "non_coding_transcript_exon_variant",
            #  "splice_donor_variant&intron_variant",
              "disruptive_inframe_deletion",
            #  "5_prime_UTR_variant",
              "missense_variant&splice_region_variant",
            #  "intergenic_region",
              "stop_gained&splice_region_variant",
              "start_lost",
              "5_prime_UTR_premature_start_codon_gain_variant",
              "conservative_inframe_deletion",
              "frameshift_variant&stop_gained",
            #  "3_prime_UTR_variant",
            #  "splice_region_variant&intron_variant",
            #  "splice_region_variant&synonymous_variant",
              "stop_lost",
            #  "splice_region_variant",
              "conservative_inframe_insertion",
            #  "splice_acceptor_variant&splice_region_variant&intron_variant",
              "frameshift_variant&splice_region_variant",
            #  "synonymous_variant",
            #  "downstream_Gene_variant",
            #  "upstream_Gene_variant",
            #  "intron_variant",
            #  "non_coding_transcript_exon_variant",
            #  "splice_donor_variant&intron_variant",
            #  "5_prime_UTR_variant",
              "missense_variant&splice_region_variant",
            #  "intergenic_region",
              "stop_gained&splice_region_variant", #!
            #  "3_prime_UTR_variant",
            #  "splice_region_variant&intron_variant",
            #  "splice_region_variant&synonymous_variant",
              "splice_region_variant",
            #  "splice_acceptor_variant&splice_region_variant&intron_variant",
            #  "frameshift_variant&splice_region_variant",
            #  "splice_donor_variant&missense_variant&splice_region_variant&intron_variant",
            #  "splice_region_variant&non_coding_transcript_exon_variant",
            #  "frameshift_variant&splice_donor_variant&splice_region_variant&intron_variant",
            #  "stop_retained_variant",
            #  "splice_acceptor_variant&intron_variant",
              "stop_gained&disruptive_inframe_deletion",
            #  "splice_acceptor_variant&splice_donor_variant&intron_variant",
              "initiator_codon_variant",
              "stop_lost&splice_region_variant", #!
              "frameshift_variant&splice_acceptor_variant&splice_region_variant&intron_variant", #!
              "splice_acceptor_variant&missense_variant&splice_region_variant&intron_variant", #!
              "splice_acceptor_variant&disruptive_inframe_deletion&splice_region_variant&intron_variant", #!
              "splice_acceptor_variant&conservative_inframe_deletion&splice_region_variant&intron_variant", #!
              "splice_donor_variant&conservative_inframe_deletion&splice_region_variant&intron_variant", #!
              "disruptive_inframe_deletion&splice_region_variant",
            #  "splice_donor_variant&splice_region_variant&intron_variant",
              "stop_lost&conservative_inframe_deletion",
              "frameshift_variant&stop_lost",
              "conservative_inframe_deletion&splice_region_variant",
              "frameshift_variant&start_lost",
              "splice_donor_variant&disruptive_inframe_deletion&splice_region_variant&intron_variant", #!
              "start_lost&splice_region_variant",
              "stop_gained&disruptive_inframe_insertion",
              "splice_acceptor_variant&splice_donor_variant&disruptive_inframe_deletion&splice_region_variant&intron_variant", #!
              "exon_loss_variant&splice_acceptor_variant&splice_donor_variant&splice_region_variant&intron_variant", #!
              "disruptive_inframe_insertion&splice_region_variant",
              "stop_gained&splice_donor_variant&splice_region_variant&intron_variant", #!
              "bidirectional_Gene_fusion",
              "start_lost&conservative_inframe_deletion",
            #  "splice_donor_variant&splice_region_variant&synonymous_variant&intron_variant",
              "stop_lost&disruptive_inframe_deletion",
              "stop_gained&splice_donor_variant&disruptive_inframe_deletion&splice_region_variant&intron_variant", #!
              "stop_gained&conservative_inframe_insertion",
              "frameshift_variant&stop_lost&splice_region_variant",
            #  "splice_donor_variant&5_prime_UTR_variant&intron_variant",
              "stop_lost&conservative_inframe_deletion&splice_region_variant",
              "start_lost&conservative_inframe_insertion",
            #  "splice_acceptor_variant&splice_region_variant&intron_variant&non_coding_transcript_exon_variant",
              "frameshift_variant&start_lost&splice_region_variant",
              "start_lost&disruptive_inframe_deletion",
            #  "start_retained_variant",
              "stop_gained&disruptive_inframe_insertion&splice_region_variant",
              "stop_gained&disruptive_inframe_deletion&splice_region_variant",
              "stop_lost&3_prime_UTR_truncation&exon_loss_variant&conservative_inframe_deletion&splice_region_variant",
            #  "splice_region_variant&stop_retained_variant",
              "stop_gained&splice_acceptor_variant&splice_region_variant&intron_variant", #!
            #  "splice_acceptor_variant&splice_region_variant&5_prime_UTR_variant&intron_variant",
            #  "splice_donor_variant&splice_region_variant&intron_variant&non_coding_transcript_exon_variant",
              "frameshift_variant&stop_gained&splice_region_variant",
              "stop_gained&conservative_inframe_insertion&splice_region_variant",
              "start_lost&conservative_inframe_deletion&splice_region_variant"
            )

data_plot$keep <- ifelse(data_plot$Effect %in% keep_func,
                         "keep",
                         "remove"
                         )

data_plot$col_label <- ifelse(data_plot$keep == "keep", "darkgreen", "black")
data_plot$xlabel <- glue::glue("<span style='color:{data_plot$col_label}'>{data_plot$Effect}</span>")

ggplot(data=data_plot, aes(x=reorder(xlabel,-n), y=n, fill=keep)) +
  geom_bar(stat="identity") +
  theme(axis.text.x = element_markdown(angle = 90, vjust = 0.5, hjust=1), axis.title.x = element_blank()) +
  scale_y_log10() +
  scale_fill_manual(values = c("keep" = "darkgreen", "remove"="grey"), guide = FALSE ) +
  ggtitle("Functions in green are kept for analysis")
```
 
```{r}
data_mut$FILTER_Effect <- ifelse(data_mut$Effect %in% keep_func,
                         "keep",
                         "remove"
                         )
```

Translate multiple annotations to one 
```{r}
new_Effect <- c(
  "missense_variant&splice_region_variant" = "missense_variant",
  "stop_gained&splice_region_variant" = "stop_gained",
  "frameshift_variant&stop_gained" = "stop_gained",
  "frameshift_variant&splice_region_variant" = "frameshift_variant",
  "stop_gained&disruptive_inframe_deletion" = "stop_gained",
  "disruptive_inframe_deletion&splice_region_variant" = "disruptive_inframe_deletion",
  "stop_lost&conservative_inframe_deletion" = "stop_lost",
  "frameshift_variant&stop_lost" = "stop_lost",
  "conservative_inframe_deletion&splice_region_variant" = "conservative_inframe_deletion",
  "frameshift_variant&start_lost" = "start_lost",
  "start_lost&splice_region_variant" = "start_lost",
  "stop_gained&disruptive_inframe_insertion" = "stop_gained",
  "disruptive_inframe_insertion&splice_region_variant" = "disruptive_inframe_insertion",
  "start_lost&conservative_inframe_deletion" = "start_lost",
  "stop_lost&disruptive_inframe_deletion" = "stop_lost",
  "stop_gained&conservative_inframe_insertion" = "stop_gained",
  "frameshift_variant&stop_lost&splice_region_variant" = "stop_lost",
  "stop_lost&conservative_inframe_deletion&splice_region_variant" = "stop_lost",
  "start_lost&conservative_inframe_insertion" = "start_lost",
  "frameshift_variant&start_lost&splice_region_variant" = "start_lost",
  "start_lost&disruptive_inframe_deletion" = "start_lost",
  "stop_gained&disruptive_inframe_insertion&splice_region_variant" = "stop_gained",
  "stop_gained&disruptive_inframe_deletion&splice_region_variant" = "stop_gained",
  "stop_lost&3_prime_UTR_truncation&exon_loss_variant&conservative_inframe_deletion&splice_region_variant" = "stop_lost",
  "frameshift_variant&stop_gained&splice_region_variant" = "stop_gained",
  "stop_gained&conservative_inframe_insertion&splice_region_variant" = "stop_gained",
  "start_lost&conservative_inframe_deletion&splice_region_variant" = "start_lost",
  "5_prime_UTR_premature_start_codon_gain_variant" = "start_gained"
  )

data_mut$Effect_NEW <- plyr::mapvalues(data_mut$Effect, names(new_Effect), new_Effect)
```

```{r}
data_mut
```

### Oncoprint old

```{r}
# Prepare plot data
data_plot <- data_mut %>% 
  filter(if_all(starts_with("FILTER_"), ~ . == "keep")) %>% 
  select(immucan_id, Gene, Effect_NEW) %>%
  pivot_wider(names_from = immucan_id, values_from = Effect_NEW, values_fn = function(x) paste(x, collapse="|")) %>%
  as.data.frame()

data_plot <- data_plot %>% 
  filter(Gene %in% Genes_driver)


rownames(data_plot) <- data_plot[["Gene"]]
data_plot[["Gene"]] <- NULL
```

```{r}
# Prepare meta data
data_meta <- as.data.frame(data_meta)
rownames(data_meta) <- data_meta[["immucan_id"]]
data_meta
```


```{r fig.width=20, fig.height=8}
# Define Variables

separator <- "|"
#get_type_fun <- function(x) strsplit(x, separator)[[1]]
col_meta <- c("gender", "simple_histology", "simple_stage", "environmental_exposure", "smoking_status", "long_term_ex_smoker")

col_fun <- function(df_meta, cl_name, brew_pallete = "Set1"){
  names_tmp <- unique(df_meta[[cl_name]])
  names_tmp[is.na(names_tmp)] <- "unkown"
  pal_tmp <- RColorBrewer::brewer.pal(length(names_tmp), brew_pallete)
  pal_tmp <- setNames(pal_tmp, names_tmp)
  
  return(pal_tmp)
}

col_list <- list(gender = c("Male" = "#6f50a1", "Female" = "#0d6d64"),
               simple_histology = col_fun(data_meta, cl_name = "simple_histology"),
               simple_stage = col_fun(data_meta, cl_name = "simple_stage"),
               environmental_exposure = col_fun(data_meta, cl_name = "environmental_exposure"),
               smoking_status = col_fun(data_meta, cl_name = "smoking_status"),
               long_term_ex_smoker = col_fun(data_meta, cl_name = "long_term_ex_smoker")
               )

## Define alteration functions
alt_predef <- unique(unlist(sapply(data_plot, unique)))
alt_predef <- alt_predef[!is.na(alt_predef)]
# Add 'mutated' for simple annotation
alt_predef <- c(alt_predef, "mutated")
alt_predef <- setNames(alt_predef, rainbow(length(alt_predef)))
alt_predef<- setNames(alt_predef, pals::alphabet(n = length(alt_predef)))
alt_fun_predef = list(
  "mutated" = function(x, y, w, h){
    grid.rect(x, y, w, h,
              gp = gpar(fill =  "#bd1a26", col = "white"))
  },
  "missense_variant" = function(x, y, w, h){
    grid.rect(x, y, w, h,
              gp = gpar(fill =  names(alt_predef[1]), col = "white"))
  },
  "stop_gained" = function(x, y, w, h){
    grid.rect(x-0.5*x, y, w*0.4, h,
              gp = gpar(fill =  names(alt_predef[2]), col = "white"))
  },
  "frameshift_variant" = function(x, y, w, h){
    grid.rect(x-0.3*x, y, w*0.4, h,
              gp = gpar(fill =  names(alt_predef[3]), col = "white"))
  },
  "frameshift_variant|missense_variant" = function(x, y, w, h){
    grid.rect(x-0.1*x, y, w*0.4, h,
              gp = gpar(fill =  names(alt_predef[4]), col = "white"))
  },
  "conservative_inframe_deletion" = function(x, y, w, h){
    grid.rect(x+0.1*x, y, w*0.4, h,
              gp = gpar(fill =  names(alt_predef[5]), col = "white"))
  },
  "disruptive_inframe_deletion" = function(x, y, w, h){
    grid.rect(x+0.3*x, y, w*0.4, h,
              gp = gpar(fill =  names(alt_predef[6]), col = "white"))
  },
  "missense_variant|missense_variant" = function(x, y, w, h){
    grid.rect(x, y, w*0.4, h,
              gp = gpar(fill =  names(alt_predef[7]), col = "white"))
  },
   "stop_gained|missense_variant" = function(x, y, w, h){
    grid.rect(x, y, w*0.4, h,
              gp = gpar(fill =  names(alt_predef[8]), col = "white"))
  }
)


background_fun <- list(
      background = function(...) NULL
    )

alt_fun <- append(background_fun, alt_fun_predef)

# Annotation
df_top <- data_meta[, col_meta]
df_top <- df_top[colnames(data_plot), ]
top_annotation = HeatmapAnnotation(
      cbar = anno_oncoprint_barplot(),
      df = df_top,
      col = col_list,
      annotation_name_side = "left",
      annotation_legend_param = list(title_gp = gpar(fontsize = 12),
                                     labels_gp = gpar(fontsize = 12))
      , annotation_name_gp= gpar(fontsize = 12)
    )

data_plot[is.na(data_plot)] <- ""
```

#### Driver
```{r fig.width=15, fig.height=10}
top_annotation = HeatmapAnnotation(
      cbar = anno_oncoprint_barplot(),
      df = df_top,
      col = col_list,
      annotation_name_side = "left",
      annotation_legend_param = list(title_gp = gpar(fontsize = 12),
                                     labels_gp = gpar(fontsize = 12))
      , annotation_name_gp= gpar(fontsize = 12)
    )
ht <- ComplexHeatmap::oncoPrint(data_plot,
                    #, get_type = get_type_fun
                    , alter_fun = alt_fun
                    , col = setNames(names(alt_predef), alt_predef)
                    , remove_empty_columns = FALSE
                    , remove_empty_rows = FALSE
                    , top_annotation = top_annotation
                    , row_names_gp = gpar(fontsize = 12)
                    #, heatmap_legend_param = list(legend_gp = gpar(fontsize = 12))
  )
ht
#pdf(file = "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/Figure_4/4A_oncoprint.pdf",width = 10,height = 9)
#draw(ht,
#     heatmap_legend_side = "bottom"
#     ,annotation_legend_side = "bottom"
#     ,padding = unit(c(2, 5, 2, 2), "mm"))
#dev.off()
```


### Oncoprint

```{r}
# Prepare plot data
data_plot <- data_mut %>% 
  filter(Gene %in% genes_sel,
         FILTER_Effect == "keep",
         FILTER_AF == "keep",
         FILTER_COSMIC == "keep",
         FILTER_DP == "keep") %>%
  select(immucan_id, Gene, Effect_NEW) %>%
  group_by(Gene, immucan_id) %>%
  dplyr::count(Effect_NEW) %>%
  mutate(Effect_NEW = ifelse(n >= 2, paste0(Effect_NEW,"s"), Effect_NEW)) %>%
  ungroup() %>%
  select(-n) %>%
  distinct() %>%
  pivot_wider(names_from = immucan_id,values_from = Effect_NEW,values_fn = function(x) paste(x, collapse = " & ")) %>%
  as.data.frame()

data_plot %>%
  filter(Gene == "EGFR") %>%
  pivot_longer(-Gene) %>%
  dplyr::count(value)

rownames(data_plot) <- data_plot[["Gene"]]
data_plot[["Gene"]] <- NULL

# add patients with no mutation back
missing_pats <- unique(data_mut$immucan_id)[!unique(data_mut$immucan_id) %in% colnames(data_plot)]
no_mut <- as.data.frame(matrix(nrow = nrow(data_plot),ncol = length(missing_pats)))
colnames(no_mut) <- missing_pats
rownames(no_mut) <- rownames(data_plot)

data_plot <- cbind(data_plot,no_mut)
```

```{r}
# Prepare meta data
data_meta <- as.data.frame(data_meta)
rownames(data_meta) <- data_meta[["immucan_id"]]
data_meta
```


```{r fig.width=20, fig.height=8}
# Define Variables
separator <- "&"
#get_type_fun <- function(x) strsplit(x, separator)[[1]]
col_meta <- c("gender", "simple_histology", "simple_stage", "environmental_exposure", "smoking_status", "long_term_ex_smoker")

col_fun <- function(df_meta, cl_name, brew_pallete = "Set1"){
  names_tmp <- unique(df_meta[[cl_name]])
  names_tmp[is.na(names_tmp)] <- "unkown"
  pal_tmp <- RColorBrewer::brewer.pal(length(names_tmp), brew_pallete)
  pal_tmp <- setNames(pal_tmp, names_tmp)
  
  return(pal_tmp)
}

col_list <- list(Gender = c("Male" = "#6f50a1", "Female" = "#0d6d64"),
               Histology = col_fun(data_meta, cl_name = "simple_histology"),
               Stage = col_fun(data_meta, cl_name = "simple_stage"),
               Environmental_exposure = col_fun(data_meta, cl_name = "environmental_exposure"),
               Smoking_status = col_fun(data_meta, cl_name = "smoking_status"),
               Long_term_ex_smoker = col_fun(data_meta, cl_name = "long_term_ex_smoker")
               )

## Define alteration functions
## Define alteration functions
alt_predef <- unique(unlist(sapply(data_plot, unique)))
alt_predef <- alt_predef[!is.na(alt_predef)]
alt_predef <- setNames( pals::alphabet(n = length(alt_predef)),alt_predef)
alt_predef

col_fun1 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[1]), col = "white"))}
col_fun2 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[2]), col = "white"))}
col_fun3<- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[3]), col = "white"))}
col_fun4 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[4]), col = "white"))}
col_fun5 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[5]), col = "white"))}
col_fun6 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[6]), col = "white"))}
col_fun7 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[7]), col = "white"))}
col_fun8 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[8]), col = "white"))}
col_fun9 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[9]), col = "white"))}
col_fun10 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[10]), col = "white"))}
col_fun11 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[11]), col = "white"))}
col_fun12 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[12]), col = "white"))}
col_fun13 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[13]), col = "white"))}
col_fun14 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[14]), col = "white"))}
col_fun15 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[15]), col = "white"))}


alt_fun_predef <- list(col_fun1,col_fun2,col_fun3,col_fun4,col_fun5,col_fun6,col_fun7,col_fun8
                       ,col_fun9
                       ,col_fun10
                       ,col_fun11
                       ,col_fun12
                       ,col_fun13
                       ,col_fun14
                       ,col_fun15
                       )
names(alt_fun_predef) <- names(alt_predef)

background_fun <- list(
      background = function(...) NULL
    )

alt_fun <- append(background_fun, alt_fun_predef)

# Annotation
df_top <- data_meta[, col_meta]
df_top <- df_top[colnames(data_plot), ]

# clean up colnames
colnames(df_top) <- c("Gender","Histology","Stage","Environmental_exposure","Smoking_status","Long_term_ex_smoker")

data_plot[is.na(data_plot)] <- ""
```

#### Driver {-}
```{r fig.width=15, fig.height=10}
top_annotation = HeatmapAnnotation(
      cbar = anno_oncoprint_barplot(),
      df = df_top,
      col = col_list,
      annotation_name_side = "left",
      annotation_legend_param = list(title_gp = gpar(fontsize = 12),
                                     labels_gp = gpar(fontsize = 12))
      , annotation_name_gp= gpar(fontsize = 12)
    )
ht <- ComplexHeatmap::oncoPrint(data_plot,
                    #, get_type = get_type_fun
                    , alter_fun = alt_fun
                    , col = alt_predef,
                    , remove_empty_columns = FALSE
                    , remove_empty_rows = FALSE
                    , top_annotation = top_annotation
                    , row_names_gp = gpar(fontsize = 12)
                    #, heatmap_legend_param = list(legend_gp = gpar(fontsize = 12))
  )
ht
pdf(file = "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/Figure_4/20251010_4A_oncoprint.pdf",width = 10,height = 10)
draw(ht,
    heatmap_legend_side = "bottom"
    ,annotation_legend_side = "bottom"
    ,padding = unit(c(2, 5, 2, 2), "mm"))
dev.off()
```





#### Driver - simple
```{r fig.width=15, fig.height=5}
data_plot[!is.na(data_plot)] <- "mutated"
ComplexHeatmap::oncoPrint(data_plot
                    #, get_type = get_type_fun
                    , alter_fun = alt_fun
                    , col = setNames("#bd1a26", "mutated")
                    , remove_empty_columns = FALSE
                    , remove_empty_rows = FALSE
                    , top_annotation = top_annotation
  )


```

### Save filtered data
```{r}
# Whole table 
data_save <- data_mut %>%
  filter(if_all(starts_with("FILTER_"), ~ . == "keep"))

data_save
write.csv(data_save, file = file.path(path_output, "NSCLC2_mutation_table_full.csv"), quote = F)
```

```{r}
# Matrix
data_save <- data_mut %>%
  filter(if_all(starts_with("FILTER_"), ~ . == "keep"))

data_save <- data_save %>%
  group_by(immucan_id, Gene) %>%
  dplyr::count() %>%
  pivot_wider(names_from = immucan_id, values_from = n, values_fill = 0)

data_save
write.csv(data_save, file = file.path(path_output, "20251010_NSCLC2_mutation_matrix.csv"), quote = F)
```

```{r}
# Driver only
## Whole table 
data_save <- data_mut %>%
  filter(if_all(starts_with("FILTER_"), ~ . == "keep")) %>%
  filter(Gene %in% genes_sel)

data_save
write.csv(data_save, file = file.path(path_output, "NSCLC2_mutation_table_driver.csv"), quote = F)
```

```{r}
# Driver only
## Matrix
data_save <- data_mut %>%
  filter(if_all(starts_with("FILTER_"), ~ . == "keep")) %>%
  filter(Gene %in% genes_sel)

data_save <- data_save %>%
  group_by(immucan_id, Gene) %>%
  dplyr::count() %>%
  pivot_wider(names_from = immucan_id, values_from = n, values_fill = 0)

data_save
write.csv(data_save, file = file.path(path_output, "NSCLC2_mutation_matrix_driver.csv"), quote = F)
```

```{r}
# TMB
write.csv(data_tmb, file = file.path(path_output, "NSCLC2_TMB.csv"), quote = F)
```

********************************************************************************
  
# Session info
  
```{r session_info, results='asis'}
sessionInfo() %>% pander()
Sys.time() %>% pander()
```

```{r system_info, results='asis'}
pander::pandoc.table(as.data.frame(Sys.info()), justify = 'left')
```


********************************************************************************

