---
title: "mutation calling from vcf"
author: "Daniel Schulz"
date: "2026-02-10"
output: html_document
---

in this script we will read in the vcf files, clean them and merge them. The original code to read the vcfs was written for a snakeMake pipeline by Sergi Beneyto and adapted here for a pure R script.

the data was downloaded from the ftp using:
`mget *_vs_IMMU-NSCLC*/design.csv`
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggplot2)
library(ggtext)
```

# selected genes for NSCLC 

```{r}
genes_sel <- c("KRAS", "EGFR", "BRAF", "MET", "ALK", "ROS", "RET", "NTRK", "NRG1", "FGFR1", "FGFR2", "HRAS", "NRAS", "MAP2K1", "ERBB2", "ROS1", "PIK3CA", "NTRK1", "STK11")
genes_sup <- c("TP53", "KEAP1","STK11","RB1","CDKN2A", "CDKN2B","NKX2", "NF1")

genes_sel <- c(genes_sel, genes_sup)
```

# availability of data vcf
```{r}
vcf_files <- list.files("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/WES/data/vcf_mutect2/", pattern = "COSMIC_CancerHotspots_GnomAD_dbNSFP_#_.*.vcf.gz$")

# exclude PON files
vcf_files <- vcf_files[!grepl("_PON_", vcf_files)]

# get unique patient ids
vcf_samps <- str_split_i(vcf_files, "-DNA",1)

# samples
cat("unique number of WES samples with vcf files",length(vcf_samps), "\n")


# duplicated samples
cat("duplicated samples: \n")
vcf_samps[duplicated(vcf_samps)]

# patient samples
vcf_pats <- str_split_i(vcf_samps,"-FIXT",1)
cat("duplicated patients: \n")
vcf_pats[duplicated(vcf_pats)]
```

# design files availabe
```{r}
cur_files <- list.files("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/WES/data/design",pattern = "*.design.csv")
des_files <- str_split_i(cur_files,"-FIXT",1)

cat("unique number of patients for which design files are available",length(unique(des_files)),"\n")

# overlap with vcf files - 
cat("are all patient found in design and vcf:",all(unique(des_files) %in% unique(vcf_pats)),"\n")
# there is one extra design file for which the vcf is not available
cat("patient for which no vcf is available:",unique(des_files)[!unique(des_files) %in% unique(vcf_pats)],"\n")
cat("all vcf files do have a design file:",all(unique(vcf_pats) %in% unique(des_files)),"\n")
# all TRUE
```


# Design matrix
downloaded the design files from the vegan directory using: `mget *_vs_IMMU-NSCLC*/design.csv`
```{r, warning=FALSE}
cur_files <- list.files("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/WES/data/design",pattern = "*.design.csv",full.names = TRUE)

files <- lapply(cur_files, read.csv)
lapply(files, colnames)
cur_dat <- do.call(rbind, files)

tumor_ids <- cur_dat$TUMOR_ID
ref <- cur_dat$GERMLINE_ID
```

# read vcf files
```{r, message=FALSE,results='hide'}
library(vcfR)
source("~/Git/NSCLC2_retrospective/WES/analysis/read_vcf.R")
vcf_files <- list.files("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/WES/data/vcf_mutect2/", pattern = "COSMIC_CancerHotspots_GnomAD_dbNSFP_#_.*.vcf.gz$", full.names = TRUE)


vcf_list <- lapply(vcf_files, read_vcf)

data_mut <- do.call(rbind, vcf_list)

cat("did we find mutations for all patients:",all(des_files %in% unique(data_mut$immucan_id)))

# separate allelic depth column
data_mut <- data_mut %>%
  separate(col = gt_AD,into = c("ref_counts","var_counts"),sep = ",")
```

## Functional filter

```{r fig.height=15}
data_plot <- data_mut %>% 
  group_by(type) %>% 
  dplyr::count() %>% 
  as.data.frame()

ggplot(data=data_plot, aes(x=reorder(type,-n), y=n)) +
  geom_bar(stat="identity", fill="grey") +
  theme(axis.text.x = element_markdown(angle = 90, vjust = 0.5, hjust=1), axis.title.x = element_blank()) +
  scale_y_log10()
```

Keep following functional annotations

```{r}
data_mut <- data_mut %>%
  mutate(type = ifelse(grepl("stop_gained",type),"stop_gained",type),
         type = ifelse(grepl("stop_lost",type),"stop_lost",type),
         type = ifelse(grepl("start_lost",type),"start_lost",type),
         type = ifelse(grepl("frameshift_variant",type),"frameshift_variant",type),
         type = ifelse(grepl("disruptive_inframe_deletion",type),"disruptive_inframe_deletion",type),
         type = ifelse(grepl("splice_acceptor_variant",type),"splice_acceptor_variant",type),
         type = ifelse(grepl("splice_donor_variant",type),"splice_donor_variant",type),
         type = ifelse(grepl("missense_variant",type),"missense_variant",type),
         type = gsub("intron_variant&|intron_variant|&intron_variant|&intron_variant&","", type),
         type = gsub("conservative_inframe_deletion&|conservative_inframe_insertion&","", type),
         type = gsub("synonymous_variant|&synonymous_variant","", type))
unique(data_mut$type)

# remove non effective mutations
data_mut <- data_mut %>%
  filter(type != "")

data_plot <- data_mut %>% 
  group_by(type) %>% 
  dplyr::count() %>% 
  as.data.frame()

ggplot(data=data_plot, aes(x=reorder(type,-n), y=n)) +
  geom_bar(stat="identity") +
  theme(axis.text.x = element_markdown(angle = 90, vjust = 0.5, hjust=1), axis.title.x = element_blank()) +
  scale_y_log10() +
  ggtitle("Functions in green are kept for analysis")
```


```{r fig.height=13}
unique(data_mut$type)
keep_func = c("missense_variant", 
              "stop_gained", 
            #  "synonymous_variant", 
            #  "downstream_gene_variant", 
              "disruptive_inframe_insertion",
            #  "upstream_gene_variant",
              "frameshift_variant",
            #  "intron_variant",
            #  "non_coding_transcript_exon_variant",
            #  "splice_donor_variant&intron_variant",
              "disruptive_inframe_deletion",
            #  "5_prime_UTR_variant",
              "missense_variant&splice_region_variant",
            #  "intergenic_region",
              "stop_gained&splice_region_variant",
              "start_lost",
              "5_prime_UTR_premature_start_codon_gain_variant",
              "conservative_inframe_deletion",
              "frameshift_variant&stop_gained",
            #  "3_prime_UTR_variant",
            #  "splice_region_variant&intron_variant",
            #  "splice_region_variant&synonymous_variant",
              "stop_lost",
            #  "splice_region_variant",
              "conservative_inframe_insertion",
            #  "splice_acceptor_variant&splice_region_variant&intron_variant",
              "frameshift_variant&splice_region_variant",
            #  "synonymous_variant",
            #  "downstream_gene_variant",
            #  "upstream_gene_variant",
            #  "intron_variant",
            #  "non_coding_transcript_exon_variant",
            #  "splice_donor_variant&intron_variant",
            #  "5_prime_UTR_variant",
              "missense_variant&splice_region_variant",
            #  "intergenic_region",
              "stop_gained&splice_region_variant", #!
            #  "3_prime_UTR_variant",
            #  "splice_region_variant&intron_variant",
            #  "splice_region_variant&synonymous_variant",
              "splice_region_variant",
            #  "splice_acceptor_variant&splice_region_variant&intron_variant",
            #  "frameshift_variant&splice_region_variant",
            #  "splice_donor_variant&missense_variant&splice_region_variant&intron_variant",
            #  "splice_region_variant&non_coding_transcript_exon_variant",
            #  "frameshift_variant&splice_donor_variant&splice_region_variant&intron_variant",
            #  "stop_retained_variant",
            #  "splice_acceptor_variant&intron_variant",
              "stop_gained&disruptive_inframe_deletion",
            #  "splice_acceptor_variant&splice_donor_variant&intron_variant",
              "initiator_codon_variant",
              "stop_lost&splice_region_variant", #!
              "frameshift_variant&splice_acceptor_variant&splice_region_variant&intron_variant", #!
              "splice_acceptor_variant&missense_variant&splice_region_variant&intron_variant", #!
              "splice_acceptor_variant&disruptive_inframe_deletion&splice_region_variant&intron_variant", #!
              "splice_acceptor_variant&conservative_inframe_deletion&splice_region_variant&intron_variant", #!
              "splice_donor_variant&conservative_inframe_deletion&splice_region_variant&intron_variant", #!
              "disruptive_inframe_deletion&splice_region_variant",
            #  "splice_donor_variant&splice_region_variant&intron_variant",
              "stop_lost&conservative_inframe_deletion",
              "frameshift_variant&stop_lost",
              "conservative_inframe_deletion&splice_region_variant",
              "frameshift_variant&start_lost",
              "splice_donor_variant&disruptive_inframe_deletion&splice_region_variant&intron_variant", #!
              "start_lost&splice_region_variant",
              "stop_gained&disruptive_inframe_insertion",
              "splice_acceptor_variant&splice_donor_variant&disruptive_inframe_deletion&splice_region_variant&intron_variant", #!
              "exon_loss_variant&splice_acceptor_variant&splice_donor_variant&splice_region_variant&intron_variant", #!
              "disruptive_inframe_insertion&splice_region_variant",
              "stop_gained&splice_donor_variant&splice_region_variant&intron_variant", #!
              "bidirectional_gene_fusion",
              "start_lost&conservative_inframe_deletion",
            #  "splice_donor_variant&splice_region_variant&synonymous_variant&intron_variant",
              "stop_lost&disruptive_inframe_deletion",
              "stop_gained&splice_donor_variant&disruptive_inframe_deletion&splice_region_variant&intron_variant", #!
              "stop_gained&conservative_inframe_insertion",
              "frameshift_variant&stop_lost&splice_region_variant",
            #  "splice_donor_variant&5_prime_UTR_variant&intron_variant",
              "stop_lost&conservative_inframe_deletion&splice_region_variant",
              "start_lost&conservative_inframe_insertion",
            #  "splice_acceptor_variant&splice_region_variant&intron_variant&non_coding_transcript_exon_variant",
              "frameshift_variant&start_lost&splice_region_variant",
              "start_lost&disruptive_inframe_deletion",
            #  "start_retained_variant",
              "stop_gained&disruptive_inframe_insertion&splice_region_variant",
              "stop_gained&disruptive_inframe_deletion&splice_region_variant",
              "stop_lost&3_prime_UTR_truncation&exon_loss_variant&conservative_inframe_deletion&splice_region_variant",
            #  "splice_region_variant&stop_retained_variant",
              "stop_gained&splice_acceptor_variant&splice_region_variant&intron_variant", #!
            #  "splice_acceptor_variant&splice_region_variant&5_prime_UTR_variant&intron_variant",
            #  "splice_donor_variant&splice_region_variant&intron_variant&non_coding_transcript_exon_variant",
              "frameshift_variant&stop_gained&splice_region_variant",
              "stop_gained&conservative_inframe_insertion&splice_region_variant",
              "start_lost&conservative_inframe_deletion&splice_region_variant"
            )

data_plot$keep <- ifelse(data_plot$type %in% keep_func,
                         "keep",
                         "remove"
                         )

data_plot$col_label <- ifelse(data_plot$keep == "keep", "darkgreen", "black")
data_plot$xlabel <- glue::glue("<span style='color:{data_plot$col_label}'>{data_plot$type}</span>")

ggplot(data=data_plot, aes(x=reorder(xlabel,-n), y=n, fill=keep)) +
  geom_bar(stat="identity") +
  theme(axis.text.x = element_markdown(angle = 90, vjust = 0.5, hjust=1), axis.title.x = element_blank()) +
  scale_y_log10() +
  scale_fill_manual(values = c("keep" = "darkgreen", "remove"="grey"), guide = FALSE ) +
  ggtitle("Functions in green are kept for analysis")
```
 
```{r}
data_mut$FILTER_type <- ifelse(data_mut$type %in% keep_func,
                         "keep",
                         "remove")
```


# QC

## VAF 
```{r}

# cut off for ceoverage of 100 reads
hist(data_mut$gt_DP, breaks = 100)
abline(v = 100,col = "red")

data_mut %>%
  filter(gene %in% genes_sel,
         FILTER_type == "keep",
         ) %>%
  ggplot(aes(x = gene,y = as.numeric(gt_AF)))+
  geom_boxplot()+
  geom_hline(yintercept = 0.05)+
  theme(axis.text.x = element_text(angle = 90,vjust = 0.5,hjust = 1))

```

# save

```{r}
write.csv(x = data_mut,file = "/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/WES/data/Rout/20260210_mutations_all.csv",quote = FALSE,row.names = FALSE)
```
