---
title: "oncoplot"
author: "Daniel Schulz"
date: "2026-02-10"
output: html_document
---

in this script we will read in the vcf files, clean them and merge them. The original code to read the vcfs was written for a snakeMake pipeline by Sergi Beneyto and adapted here for a pure R script.
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# read file
```{r}
data_mut <- read_csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/WES/data/Rout/20260210_mutations_all.csv")
```

# filtering

min VAF: 5%
ref VAF: < 1%
min coverage: 100 reads

```{r}
VAF <- 0.05
ref_VAF <- 0.01
min_cov <- 100
disruptive_variants <- TRUE
```

# selected genes for NSCLC 

```{r}
genes_sel <- c("KRAS", "EGFR", "BRAF", "MET", "ALK", "ROS", "RET", "NTRK", "NRG1", "FGFR1", "FGFR2", "HRAS", "NRAS", "MAP2K1", "ERBB2", "ROS1", "PIK3CA", "NTRK1", "STK11")
genes_sup <- c("TP53", "KEAP1","STK11","RB1","CDKN2A", "CDKN2B","NKX2", "NF1")

genes_sel <- c(genes_sel, genes_sup)
```


# Filtering

```{r}
cur_dat <- data_mut %>%
  filter(gt_AF > VAF,
         gt_DP > min_cov,
         if(disruptive_variants == TRUE) FILTER_type == "keep" else FILTER_type %in% c("keep", "remove"))
```

# Metadata
```{r}
# Prepare meta data
clin_dat <- read_csv("/mnt/immucan_volume/clinical_data/Lung_NSCLC2/20230825/NSCLC2_for_R_update_2024.csv")

data_meta <- as.data.frame(clin_dat)
rownames(data_meta) <- data_meta[["immucan_id"]]
data_meta

```

# Oncoprint selected genes
no cosmic filter
```{r}
# Prepare plot data
data_plot <- cur_dat %>% 
  filter(gene %in% genes_sel) %>%
  select(immucan_id, gene, type) %>%
  group_by(gene, immucan_id) %>%
  dplyr::count(type) %>%
  mutate(type = ifelse(n >= 2, paste0(type,"s"), type),
         type = ifelse(type == "stop_gaineds", "stops_gained", type)) %>%
  ungroup() %>%
  select(-n) %>%
  distinct() %>%
  pivot_wider(names_from = immucan_id,values_from = type,values_fn = function(x) paste(x, collapse = " & ")) %>%
  as.data.frame()

# add patients with no mutation back
missing_pats <- unique(data_mut$immucan_id)[!unique(data_mut$immucan_id) %in% colnames(data_plot)]
no_mut <- as.data.frame(matrix(nrow = nrow(data_plot),ncol = length(missing_pats)))
colnames(no_mut) <- missing_pats
rownames(no_mut) <- rownames(data_plot)

data_plot <- cbind(data_plot,no_mut)

unique(unique(unlist(sapply(data_plot, unique))))

rownames(data_plot) <- data_plot[["gene"]]
data_plot[["gene"]] <- NULL

data_mut %>% filter(grepl("stop_gaineds", type))
```



```{r fig.width=20, fig.height=8}
col_meta <- c("gender", "Histology", "simple_stage", "smoking_status", "age_dicho")

# Define colors
Stage_col = setNames(c("palevioletred3","darkseagreen3","dodgerblue", "darkblue"),c("I","II","III","IV"))
Stage_col_num = setNames(c("palevioletred3","darkseagreen3","dodgerblue", "darkblue"),c("1","2","3", "4"))
hist_col_2= setNames(c("blue","orange"),c("LUAD","LUSC"))
hist_col_3 = setNames(c("blue","orange", "gray40"),c("LUAD","LUSC","Other"))
# smoking
smoking_status_col_3 <- setNames(c("blue", "lightblue", "cyan3"),sort(c("Current","Former", "Never")))
smoking_status_col_4 <- setNames(c("blue", "lightblue", "cyan3", "gray40"),sort(c("Current","Former", "Never", "Unknown")))

data_meta <- data_meta %>%
  mutate(Histology = LUAD_vs_LUSC,
         Histology = ifelse(is.na(LUAD_vs_LUSC),"Other", ifelse(Histology == "Adenocarcinoma", "LUAD", "LUSC")),
         smoking_status = ifelse(is.na(smoking_status),"Unknown",smoking_status),
         smoking_status = ifelse(smoking_status == "Ex-smoker", "Former", smoking_status),
         smoking_status = ifelse(smoking_status == "Current smoker", "Current", smoking_status),
         smoking_status = ifelse(smoking_status == "Non-smoker", "Never", smoking_status))


col_list <- list(gender = c("Male" = "#6f50a1", "Female" = "#0d6d64"),
               Histology =hist_col_3,
               simple_stage = Stage_col,
               smoking_status = smoking_status_col_4,
               age_dicho = c("low" = "blue3", "high" = "red4")
               )

# drivers
# mutations: KRAS
KRAS_col <- setNames(c("red", "blue"),sort(c(TRUE,FALSE)))
EGFR_col <- setNames(c("red", "blue"),sort(c(TRUE,FALSE)))
PIK3CA_col <- setNames(c("red", "blue"),sort(c(TRUE,FALSE)))
STK11_col <- setNames(c("red", "blue"),sort(c(TRUE,FALSE)))
KEAP1_col <- setNames(c("red", "blue"),sort(c(TRUE,FALSE)))
BRAF_col <- setNames(c("red", "blue"),sort(c(TRUE,FALSE)))
RET_col <- setNames(c("red", "blue"),sort(c(TRUE,FALSE)))
MET_col <- setNames(c("red", "blue"),sort(c(TRUE,FALSE)))
ROS1_col <- setNames(c("red", "blue"),sort(c(TRUE,FALSE)))
TP53_col <- setNames(c("red", "blue"),sort(c(TRUE,FALSE)))



## Define alteration functions
alt_predef <- unique(unlist(sapply(data_plot, unique)))
#alt_predef <- alt_predef[c(1,3:length(alt_predef))]
alt_predef <- alt_predef[!is.na(alt_predef)]
alt_predef <- setNames( pals::alphabet(n = length(alt_predef)),alt_predef)
alt_predef
```


```{r}

col_fun1 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[1]), col = "white"))}
col_fun2 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[2]), col = "white"))}
col_fun3<- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[3]), col = "white"))}
col_fun4 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[4]), col = "white"))}
col_fun5 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[5]), col = "white"))}
col_fun6 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[6]), col = "white"))}
col_fun7 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[7]), col = "white"))}
col_fun8 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[8]), col = "white"))}
col_fun9 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[9]), col = "white"))}
col_fun10 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[10]), col = "white"))}
col_fun11 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[11]), col = "white"))}
col_fun12 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[12]), col = "white"))}
col_fun13 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[13]), col = "white"))}
col_fun14 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[14]), col = "white"))}
col_fun15 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[15]), col = "white"))}
col_fun16 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[16]), col = "white"))}
col_fun17 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[17]), col = "white"))}
col_fun18 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[18]), col = "white"))}
col_fun19 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[19]), col = "white"))}

alt_fun_predef <- list(col_fun1,col_fun2,col_fun3,col_fun4,col_fun5,col_fun6,col_fun7,col_fun8
                       ,col_fun9
                       ,col_fun10
                       ,col_fun11
                       ,col_fun12
                       ,col_fun13
                       ,col_fun14
                       ,col_fun15
                       ,col_fun16
                       ,col_fun17
                       ,col_fun18
                       ,col_fun19
                       )

names(alt_fun_predef) <- names(alt_predef)

background_fun <- list(
      background = function(...) NULL
    )

alt_fun <- append(background_fun, alt_fun_predef)

# note: currently we have not all clinical data yet. we will subset the data to plot to only contain those we have clinical data for and fix this later
cur_pats <- intersect(colnames(data_plot),clin_dat$immucan_id)
data_plot <- data_plot %>%
  select(all_of(cur_pats))

# Annotation
df_top <- as.data.frame(data_meta[, col_meta])
rownames(df_top) <- data_meta$immucan_id
df_top <- df_top[colnames(data_plot), ]


data_plot[is.na(data_plot)] <- ""
```


```{r fig.width=20, fig.height=10}
library(ComplexHeatmap)

top_annotation = HeatmapAnnotation(
      cbar = anno_oncoprint_barplot(),
      df = df_top,
      col = col_list,
      annotation_name_side = "left",
      annotation_legend_param = list(title_gp = gpar(fontsize = 12),
                                     labels_gp = gpar(fontsize = 12))
      , annotation_name_gp= gpar(fontsize = 12)
    )
ht <- ComplexHeatmap::oncoPrint(data_plot,
                    #, get_type = get_type_fun
                    , alter_fun = alt_fun
                    #, alter_fun_is_vectorized = FALSE
                    #, col = setNames(names(alt_predef), alt_predef)
                    , col = alt_predef
                    , remove_empty_columns = FALSE
                    , remove_empty_rows = FALSE
                    , top_annotation = top_annotation
                    , row_names_gp = gpar(fontsize = 12)
                    #, heatmap_legend_param = list(legend_gp = gpar(fontsize = 12))
  )
ht
pdf(file = "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/Figure_4/oncoprint_selected_genes_20260210.pdf",width = 10,height = 9)
draw(ht,
    heatmap_legend_side = "bottom"
    ,annotation_legend_side = "bottom"
    ,padding = unit(c(2, 5, 2, 2), "mm"))
dev.off()
```


# Oncoprint COSMIC genes

```{r}
# Prepare plot data
data_plot <- cur_dat %>% 
  filter(gene %in% genes_sel,
         COSMIC == TRUE) %>%
  select(immucan_id, gene, type) %>%
  group_by(gene, immucan_id) %>%
  dplyr::count(type) %>%
  mutate(type = ifelse(n >= 2, paste0(type,"s"), type),
         type = ifelse(type == "stop_gaineds", "stops_gained", type)) %>%
  ungroup() %>%
  select(-n) %>%
  distinct() %>%
  pivot_wider(names_from = immucan_id,values_from = type,values_fn = function(x) paste(x, collapse = " & ")) %>%
  as.data.frame()

# add patients with no mutation back
missing_pats <- unique(data_mut$immucan_id)[!unique(data_mut$immucan_id) %in% colnames(data_plot)]
no_mut <- as.data.frame(matrix(nrow = nrow(data_plot),ncol = length(missing_pats)))
colnames(no_mut) <- missing_pats
rownames(no_mut) <- rownames(data_plot)

data_plot <- cbind(data_plot,no_mut)

unique(unique(unlist(sapply(data_plot, unique))))

rownames(data_plot) <- data_plot[["gene"]]
data_plot[["gene"]] <- NULL

data_mut %>% filter(grepl("stop_gaineds", type))
```



```{r fig.width=20, fig.height=8}
## Define alteration functions
alt_predef <- unique(unlist(sapply(data_plot, unique)))
#alt_predef <- alt_predef[c(1,3:length(alt_predef))]
alt_predef <- alt_predef[!is.na(alt_predef)]
alt_predef <- setNames( pals::alphabet(n = length(alt_predef)),alt_predef)
alt_predef
```


```{r}

col_fun1 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[1]), col = "white"))}
col_fun2 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[2]), col = "white"))}
col_fun3<- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[3]), col = "white"))}
col_fun4 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[4]), col = "white"))}
col_fun5 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[5]), col = "white"))}
col_fun6 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[6]), col = "white"))}
col_fun7 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[7]), col = "white"))}
col_fun8 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[8]), col = "white"))}
col_fun9 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[9]), col = "white"))}
col_fun10 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[10]), col = "white"))}
col_fun11 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[11]), col = "white"))}
col_fun12 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[12]), col = "white"))}
col_fun13 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[13]), col = "white"))}
col_fun14 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[14]), col = "white"))}
col_fun15 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[15]), col = "white"))}
col_fun16 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[16]), col = "white"))}
col_fun17 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[17]), col = "white"))}
col_fun18 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[18]), col = "white"))}
col_fun19 <- function(x,y,w,h) {
    grid.rect(x, y, w, h,gp = gpar(fill = unname(alt_predef[19]), col = "white"))}

alt_fun_predef <- list(col_fun1,col_fun2,col_fun3,col_fun4,col_fun5,col_fun6,col_fun7,col_fun8
                       ,col_fun9
                       ,col_fun10
                       ,col_fun11
                       ,col_fun12
                       ,col_fun13
                       ,col_fun14
                       ,col_fun15
                       ,col_fun16
                       ,col_fun17
                       ,col_fun18
                       ,col_fun19
                       )

names(alt_fun_predef) <- names(alt_predef)

background_fun <- list(
      background = function(...) NULL
    )

alt_fun <- append(background_fun, alt_fun_predef)

# note: currently we have not all clinical data yet. we will subset the data to plot to only contain those we have clinical data for and fix this later
cur_pats <- intersect(colnames(data_plot),clin_dat$immucan_id)
data_plot <- data_plot %>%
  select(all_of(cur_pats))

# Annotation
df_top <- as.data.frame(data_meta[, col_meta])
rownames(df_top) <- data_meta$immucan_id
df_top <- df_top[colnames(data_plot), ]


data_plot[is.na(data_plot)] <- ""
```


```{r fig.width=20, fig.height=10}
top_annotation = HeatmapAnnotation(
      cbar = anno_oncoprint_barplot(),
      df = df_top,
      col = col_list,
      annotation_name_side = "left",
      annotation_legend_param = list(title_gp = gpar(fontsize = 12),
                                     labels_gp = gpar(fontsize = 12))
      , annotation_name_gp= gpar(fontsize = 12)
    )
ht <- ComplexHeatmap::oncoPrint(data_plot,
                    #, get_type = get_type_fun
                    , alter_fun = alt_fun
                    #, alter_fun_is_vectorized = FALSE
                    #, col = setNames(names(alt_predef), alt_predef)
                    , col = alt_predef
                    , remove_empty_columns = FALSE
                    , remove_empty_rows = FALSE
                    , top_annotation = top_annotation
                    , row_names_gp = gpar(fontsize = 12)
                    #, heatmap_legend_param = list(legend_gp = gpar(fontsize = 12))
  )
ht
pdf(file = "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/Figure_4/4A_oncoprint_COSMIC_20260210.pdf",width = 10,height = 9)
draw(ht,
    heatmap_legend_side = "bottom"
    ,annotation_legend_side = "bottom"
    ,padding = unit(c(2, 5, 2, 2), "mm"))
dev.off()
```
