---
title: "data_prep"
author: "Daniel Schulz"
date: "2026-09-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## 02_data_prep.R
## Merge all data modalities by immucan_id, build a feature dictionary
## (name -> category -> spatial flag -> modality) and a tidy long-format
## table (immucan_id, name, value, category, spatial). Save both to results/.
##
## Base R only (survival/boot + base). Must run on R 4.2+.

```{r}

suppressWarnings(suppressMessages(library(survival)))

DATA_DIR    <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/"
RESULTS_DIR <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/survival/Rout/"

log_msg <- function(...) cat(format(Sys.time(), "%H:%M:%S"), "-", ..., "\n")

```

# read data
```{r}
traits <- read.csv(file.path(DATA_DIR, "all_traits.csv"), stringsAsFactors = FALSE)

dens <- readRDS(file.path(DATA_DIR, "densities_dat.rds"))

inter <- readRDS(file.path(DATA_DIR, "interaction_dat.rds"))

cello <- readRDS(file.path(DATA_DIR, "cello_dat.rds"))

img_mat <- readRDS(file.path(DATA_DIR, "patient_imaging_meta_cluster_scores_tpm.rds"))

rna_gene <- read.csv(file.path(DATA_DIR, "MOFA/RNA_named.csv"), stringsAsFactors = FALSE,
                      check.names = FALSE)

rna_mod <- readRDS(file.path(DATA_DIR, "rna_scores_tpm.rds"))
```



# 2. Sanity checks / structure verification (printed, not assumed)
```{r}

log_msg("Raw dims: traits=", paste(dim(traits), collapse = "x"),
        " dens=", paste(dim(dens), collapse = "x"),
        " inter=", paste(dim(inter), collapse = "x"),
        " cello=", paste(dim(cello), collapse = "x"),
        " img_mat=", paste(dim(img_mat), collapse = "x"),
        " rna_gene=", paste(dim(rna_gene), collapse = "x"),
        " rna_mod=", paste(dim(rna_mod), collapse = "x"))

stopifnot("immucan_id" %in% colnames(traits))
stopifnot("immucan_id" %in% colnames(dens))
stopifnot("immucan_id" %in% colnames(inter))
stopifnot("immucan_id" %in% colnames(cello))
stopifnot("Immucan_id" %in% colnames(rna_gene))  # confirmed capitalisation differs

```

## death vs censoring cross-check: confirm which value of `censoring` = event.

```{r}
xtab <- table(traits$death, traits$censoring, useNA = "always")
log_msg("death x censoring crosstab:")
print(xtab)
event_value <- 1
if (!(all(traits$death[traits$censoring == 1] == "Yes", na.rm = TRUE) &&
      all(traits$death[traits$censoring == 0] == "No", na.rm = TRUE))) {
  stop("censoring==1 does not cleanly correspond to death=='Yes' -- re-check coding before proceeding.")
}
log_msg("Confirmed: censoring == 1 corresponds to death == 'Yes' (event). Using event_value = 1.")
```



## histology levels present (for info; filtering to LUAD/LUSC happens in downstream scripts)
```{r}
print(table(traits$simple_histology, useNA = "always"))
```

# 3. Fix RNA names
```{r}
rna_gene$immucan_id <- gsub("_", "-", rna_gene$Immucan_id)
rna_gene$Immucan_id <- NULL
stopifnot(all(rna_gene$immucan_id %in% traits$immucan_id))
```


# 4. image module matrix prep
```{r}
stopifnot(all(rownames(img_mat) %in% traits$immucan_id))
img_df <- as.data.frame(img_mat, stringsAsFactors = FALSE)
img_df$immucan_id <- rownames(img_mat)
rownames(img_df) <- NULL

names(img_df) <- as.character(names(img_df))
log_msg("image module colnames: ", paste(setdiff(colnames(img_df), "immucan_id"), collapse = ", "))

```


# 5. RNA module score matrix 
```{r}
stopifnot(all(rownames(rna_mod) %in% traits$immucan_id))
rna_mod_df <- as.data.frame(rna_mod, stringsAsFactors = FALSE)
rna_mod_df$immucan_id <- rownames(rna_mod)
rownames(rna_mod_df) <- NULL
```

# 6. Biomarker + mutation columns pulled out of traits
```{r}
biomarker_cols <- c("TPS", "CPS", "TMB")
mutation_cols  <- c("KRAS","EGFR","BRAF","MET","ALK","RET","NRG1","FGFR1","FGFR2",
                     "ERBB2","ROS1","PIK3CA","NTRK1","STK11","TP53","KEAP1","RB1",
                     "CDKN2A","NF1")
stopifnot(all(biomarker_cols %in% colnames(traits)))
stopifnot(all(mutation_cols %in% colnames(traits)))

biomarker_df <- traits[, c("immucan_id", biomarker_cols)]
mutation_df  <- traits[, c("immucan_id", mutation_cols)]

# mutation columns are logical TRUE/FALSE/NA -> convert to numeric 0/1 for modelling
for (cc in mutation_cols) mutation_df[[cc]] <- as.numeric(mutation_df[[cc]])
```


# 7. Name-collisions 
handling: driver-gene mutation symbols (e.g. ROS1, RET,
    CDKN2A, ERBB2) also exist as RNA_gene expression column names (same
    gene symbol, completely different measurement: binary mutation call vs
    continuous bulk expression). Disambiguate RNA_gene columns with a
    "__RNA_gene" suffix before merging (documented in decisions log).

```{r}
rna_gene_feat_cols <- setdiff(colnames(rna_gene), "immucan_id")
overlap_before <- intersect(rna_gene_feat_cols, mutation_cols)
log_msg("Gene-symbol collisions between RNA_gene expression and mutation calls: ",
        paste(overlap_before, collapse = ", "))
new_names <- paste0(rna_gene_feat_cols, "__RNA_gene")
names(rna_gene)[match(rna_gene_feat_cols, colnames(rna_gene))] <- new_names
rna_gene_feat_cols <- new_names

```


# 8. Merge all modalities 

```{r}
merge_all <- function(a, b) merge(a, b, by = "immucan_id", all = TRUE)

feature_tables <- list(dens, inter, cello, img_df, rna_gene, rna_mod_df,
                        biomarker_df, mutation_df)
full_wide <- Reduce(merge_all, feature_tables)
log_msg("Merged feature-only wide table dim: ", paste(dim(full_wide), collapse = "x"))

```


## clinical/core columns needed by downstream scripts
```{r}
clinical_core <- traits[, c("immucan_id", "simple_histology", "time_to_event",
                             "censoring", "death")]
clinical_core$status <- as.integer(clinical_core$censoring == event_value)
```



## merge with clinical
```{r}
full_wide <- merge(clinical_core, full_wide, by = "immucan_id", all.x = TRUE)
log_msg("Final full_wide dim (patients x [clinical + features]): ",
        paste(dim(full_wide), collapse = "x"))
stopifnot(!any(duplicated(full_wide$immucan_id)))
```


# 9. Feature dictionary
```{r}
extract_modality <- function(nm, strip_suffix = NULL) {
  m <- sub("^.*__", "", nm)
  if (!is.null(strip_suffix)) m <- sub(paste0(strip_suffix, "$"), "", m)
  m
}

dict_density <- data.frame(
  name = setdiff(colnames(dens), "immucan_id"), category = "density",
  spatial = FALSE, stringsAsFactors = FALSE)
dict_density$modality <- extract_modality(dict_density$name)

dict_interaction <- data.frame(
  name = setdiff(colnames(inter), "immucan_id"), category = "interaction",
  spatial = TRUE, stringsAsFactors = FALSE)
dict_interaction$modality <- extract_modality(dict_interaction$name, strip_suffix = "_int")

dict_cello <- data.frame(
  name = setdiff(colnames(cello), "immucan_id"), category = "cellohood",
  spatial = TRUE, stringsAsFactors = FALSE)
dict_cello$modality <- extract_modality(dict_cello$name)

dict_img <- data.frame(
  name = setdiff(colnames(img_df), "immucan_id"), category = "image_module",
  spatial = NA, modality = "imaging_meta_cluster", stringsAsFactors = FALSE)

dict_rna_gene <- data.frame(
  name = rna_gene_feat_cols, category = "RNA_gene",
  spatial = FALSE, modality = "RNA_gene", stringsAsFactors = FALSE)

dict_rna_mod <- data.frame(
  name = setdiff(colnames(rna_mod_df), "immucan_id"), category = "RNA_module",
  spatial = FALSE, modality = "RNA_module", stringsAsFactors = FALSE)

dict_biomarker <- data.frame(
  name = biomarker_cols, category = "biomarker",
  spatial = FALSE, modality = "clinical_biomarker", stringsAsFactors = FALSE)

dict_mutation <- data.frame(
  name = mutation_cols, category = "mutation",
  spatial = FALSE, modality = "WES_driver_mutation", stringsAsFactors = FALSE)

feature_dict <- rbind(dict_density, dict_interaction, dict_cello, dict_img,
                       dict_rna_gene, dict_rna_mod, dict_biomarker, dict_mutation)
stopifnot(!any(duplicated(feature_dict$name)))
stopifnot(all(feature_dict$name %in% colnames(full_wide)))
log_msg("Feature dictionary: ", nrow(feature_dict), " features across categories:")
print(table(feature_dict$category))

```


# 10. Tidy long-format table
immucan_id, name, value, category, spatial
(values coerced to numeric; logicals from mutation_df already numeric)

```{r}
log_msg("Building long-format table ...")
feat_names <- feature_dict$name
n_pat <- nrow(full_wide)
val_list <- vector("list", length(feat_names))
for (k in seq_along(feat_names)) {
  val_list[[k]] <- suppressWarnings(as.numeric(full_wide[[feat_names[k]]]))
}
long_dat <- data.frame(
  immucan_id = rep(full_wide$immucan_id, times = length(feat_names)),
  name       = rep(feat_names, each = n_pat),
  value      = unlist(val_list, use.names = FALSE),
  stringsAsFactors = FALSE
)
long_dat <- merge(long_dat, feature_dict[, c("name", "category", "spatial")],
                   by = "name", all.x = TRUE)
long_dat <- long_dat[, c("immucan_id", "name", "value", "category", "spatial")]
log_msg("Long table dim: ", paste(dim(long_dat), collapse = "x"),
        " (expected ", n_pat, " x ", length(feat_names), " = ", n_pat * length(feat_names), " rows)")
stopifnot(nrow(long_dat) == n_pat * length(feat_names))
```



# 11. Save
```{r}
saveRDS(feature_dict,  file.path(RESULTS_DIR, "feature_dictionary.rds"))
saveRDS(long_dat,      file.path(RESULTS_DIR, "features_long.rds"))
saveRDS(full_wide,     file.path(RESULTS_DIR, "full_wide.rds"))
saveRDS(clinical_core, file.path(RESULTS_DIR, "clinical_core.rds"))

log_msg("Saved: results/feature_dictionary.rds, results/features_long.rds, ",
        "results/full_wide.rds, results/clinical_core.rds")
log_msg("DONE 02_data_prep.R")
```


