---
title: "QC"
author: "Daniel Schulz"
date: "2026-09-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
## 03_qc_plots.R
## "Difference in detected features for LUAD vs LUSC"
## For each feature category, compute how many features pass basic QC
## (non-degenerate: >=2 unique observed values; <50% missing) within
## LUAD-only rows and LUSC-only rows separately.
##
## Base R only. Must run on R 4.2+.

RESULTS_DIR <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/survival/Rout/"
PLOTS_DIR   <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/survival/plots/"
if (!dir.exists(PLOTS_DIR)) dir.create(PLOTS_DIR, recursive = TRUE)

log_msg <- function(...) cat(format(Sys.time(), "%H:%M:%S"), "-", ..., "\n")

long_dat      <- readRDS(file.path(RESULTS_DIR, "features_long.rds"))
feature_dict  <- readRDS(file.path(RESULTS_DIR, "feature_dictionary.rds"))
clinical_core <- readRDS(file.path(RESULTS_DIR, "clinical_core.rds"))

## Restrict the LUAD-vs-LUSC comparison to the two histologies of interest;
## Adenosquamous / LCNEC / Other / unknown are excluded per the project spec.
histo_map <- clinical_core[clinical_core$simple_histology %in%
                              c("Adenocarcinoma", "Squamous cell carcinoma"),
                            c("immucan_id", "simple_histology", "status")]
histo_map$histology <- ifelse(histo_map$simple_histology == "Adenocarcinoma", "LUAD", "LUSC")
log_msg("Patients in LUAD/LUSC comparison: ", nrow(histo_map),
        " (LUAD=", sum(histo_map$histology == "LUAD"),
        ", LUSC=", sum(histo_map$histology == "LUSC"), ")")

long_hs <- merge(long_dat, histo_map[, c("immucan_id", "histology")], by = "immucan_id")

```

# Per-feature QC
within each histology: non-degenerate (>=2 unique observed values) and missingness < 50%.

```{r}
qc_one_group <- function(df, denom_n) {
  ## df: rows for ONE feature, ONE histology (immucan_id, value, ...)
  n_obs    <- sum(!is.na(df$value))
  frac_na  <- 1 - n_obs / denom_n
  n_unique <- length(unique(df$value[!is.na(df$value)]))
  data.frame(n_obs = n_obs, frac_na = frac_na, n_unique = n_unique)
}

histologies <- c("LUAD", "LUSC")
n_per_histo <- table(histo_map$histology)

qc_list <- list()
idx <- 1
for (h in histologies) {
  denom_n <- as.integer(n_per_histo[h])
  sub_h <- long_hs[long_hs$histology == h, ]
  spl <- split(sub_h, sub_h$name)
  qc_rows <- lapply(spl, qc_one_group, denom_n = denom_n)
  qc_df <- do.call(rbind, qc_rows)
  qc_df$name <- rownames(qc_df)
  rownames(qc_df) <- NULL
  qc_df$histology <- h
  qc_list[[idx]] <- qc_df
  idx <- idx + 1
}
qc_all <- do.call(rbind, qc_list)
qc_all <- merge(qc_all, feature_dict[, c("name", "category", "spatial")], by = "name")
qc_all$pass_qc <- qc_all$n_unique >= 2 & qc_all$frac_na < 0.5

log_msg("QC table dim: ", paste(dim(qc_all), collapse = "x"))
saveRDS(qc_all, file.path(RESULTS_DIR, "qc_by_category_histology.rds"))
```

# QC 2
Aggregate: features passing QC, by category x histology
qc_features_by_category_histology
```{r}
categories <- sort(unique(feature_dict$category))

pass_tab <- table(qc_all$category[qc_all$pass_qc], qc_all$histology[qc_all$pass_qc])
pass_mat <- matrix(0, nrow = length(categories), ncol = length(histologies),
                    dimnames = list(categories, histologies))
for (r in rownames(pass_tab)) for (cn in colnames(pass_tab)) pass_mat[r, cn] <- pass_tab[r, cn]

## missingness rate (mean frac_na across features) by category x histology
miss_mat <- matrix(NA_real_, nrow = length(categories), ncol = length(histologies),
                    dimnames = list(categories, histologies))
for (r in categories) for (h in histologies) {
  vals <- qc_all$frac_na[qc_all$category == r & qc_all$histology == h]
  miss_mat[r, h] <- if (length(vals)) mean(vals) else NA_real_
}

## total feature count per category (for context / denominator on the bar plot)
n_feat_cat <- table(feature_dict$category)[categories]

log_msg("Features passing QC by category x histology:")
print(pass_mat)
log_msg("Mean missingness by category x histology:")
print(round(miss_mat, 3))

#pdf(file.path(PLOTS_DIR, "qc_features_by_category_histology.pdf"), width = 10, height = 6)
op <- par(mar = c(8, 5, 4, 2))
bp <- barplot(t(pass_mat), beside = TRUE, las = 2,
              col = c("blue", "orange"),
              ylim = c(0, max(pass_mat, na.rm = TRUE) * 1.25),
              ylab = "Number of features passing QC\n(>=2 unique values, <50% missing)",
              main = paste0("Features passing basic QC by category: LUAD (n=",
                             sum(histo_map$histology == "LUAD"), ", ",
                             sum(histo_map$status[histo_map$histology == "LUAD"]), " events) vs LUSC (n=",
                             sum(histo_map$histology == "LUSC"), ", ",
                             sum(histo_map$status[histo_map$histology == "LUSC"]), " events, EXPLORATORY)"),
              cex.names = 0.9)
text(x = bp, y = t(pass_mat), labels = t(pass_mat), pos = 3, cex = 0.65, xpd = NA)
## annotate total available features per category above the bars
for (i in seq_along(categories)) {
  mid <- mean(bp[, i])
  text(mid, max(pass_mat[i, ], na.rm = TRUE) * 1.15,
       labels = paste0("of ", n_feat_cat[categories[i]]), cex = 0.6, col = "grey30")
}
legend("topright", legend = c("LUAD", "LUSC (exploratory)"), fill = c("blue", "orange"), bty = "n")
mtext("image_module is a spatially-derived per-patient summary index (own category, not scored as spatial/non-spatial)",
      side = 1, line = 6.5, cex = 0.65, col = "grey40")
par(op)
#dev.off()

```

# QC 3
qc_missingness_by_category_histology
```{r}
#pdf(file.path(PLOTS_DIR, "qc_missingness_by_category_histology.pdf"), width = 10, height = 6)
op <- par(mar = c(8, 5, 4, 2))
bp2 <- barplot(t(miss_mat) * 100, beside = TRUE, las = 2,
               col = c("blue", "orange"),
               ylim = c(0, 100),
               ylab = "Mean missingness rate (%) across features in category",
               main = "Missingness by category: LUAD vs LUSC (exploratory)",
               cex.names = 0.9)
text(x = bp2, y = t(miss_mat) * 100, labels = round(t(miss_mat) * 100, 1),
     pos = 3, cex = 0.6, xpd = NA)
abline(h = 50, lty = 2, col = "grey50")
legend("topright", legend = c("LUAD", "LUSC (exploratory)"), fill = c("blue", "orange"), bty = "n")
par(op)
#dev.off()
```

# QC 4
qc_sample_event_summary

n patients and n events for LUAD vs LUSC, per modality/category
(since densities/interactions/cellohood/RNA all have different n due to
per-patient modality availability)

```{r}
# Determine, for each category, how many LUAD/LUSC patients have >=1
# non-missing value in that category (a proxy for "modality available").
avail_mat <- matrix(0, nrow = length(categories), ncol = length(histologies),
                     dimnames = list(categories, histologies))
event_mat <- matrix(0, nrow = length(categories), ncol = length(histologies),
                     dimnames = list(categories, histologies))
for (r in categories) {
  feats_r <- feature_dict$name[feature_dict$category == r]
  sub_r <- long_hs[long_hs$name %in% feats_r, ]
  ## a patient counts as "has this modality" if ANY feature in the category is non-NA
  has_any <- tapply(sub_r$value, sub_r$immucan_id, function(v) any(!is.na(v)))
  ids_with <- names(has_any)[has_any]
  for (h in histologies) {
    ids_h <- histo_map$immucan_id[histo_map$histology == h]
    ids_with_h <- intersect(ids_with, ids_h)
    avail_mat[r, h] <- length(ids_with_h)
    event_mat[r, h] <- sum(histo_map$status[histo_map$immucan_id %in% ids_with_h])
  }
}

log_msg("N patients with >=1 non-missing feature, by category x histology:")
print(avail_mat)
log_msg("N events among those patients, by category x histology:")
print(event_mat)

#pdf(file.path(PLOTS_DIR, "qc_sample_event_summary.pdf"), width = 11, height = 7)
op <- par(mfrow = c(1, 2), mar = c(8, 5, 4, 1))
bp3 <- barplot(t(avail_mat), beside = TRUE, las = 2,
               col = c("blue", "orange"),
               ylab = "N patients with modality available",
               main = "N patients by category x histology",
               cex.names = 0.85)
text(x = bp3, y = t(avail_mat), labels = t(avail_mat), pos = 3, cex = 0.6, xpd = NA)
legend("topright", legend = c("LUAD", "LUSC (exploratory)"), fill = c("blue", "orange"), bty = "n", cex = 0.8)

bp4 <- barplot(t(event_mat), beside = TRUE, las = 2,
               col = c("blue", "orange"),
               ylab = "N events (deaths) among available patients",
               main = "N EVENTS by category x histology\n(note the small LUSC counts)",
               cex.names = 0.85)
text(x = bp4, y = t(event_mat), labels = t(event_mat), pos = 3, cex = 0.6, xpd = NA)
legend("topright", legend = c("LUAD", "LUSC (exploratory)"), fill = c("blue", "orange"), bty = "n", cex = 0.8)
mtext(paste0("Overall: LUAD n=", sum(histo_map$histology == "LUAD"),
             " events=", sum(histo_map$status[histo_map$histology == "LUAD"]),
             "  |  LUSC n=", sum(histo_map$histology == "LUSC"),
             " events=", sum(histo_map$status[histo_map$histology == "LUSC"]),
             " -- LUSC results are EXPLORATORY throughout due to low event count"),
      outer = TRUE, line = -1.2, cex = 0.75, col = "grey20")
par(op)
#dev.off()

log_msg("Saved plots: qc_features_by_category_histology.pdf, qc_missingness_by_category_histology.pdf, qc_sample_event_summary.pdf")
log_msg("DONE 03_qc_plots.R")

```

