---
title: "survival_modeling"
author: "Daniel Schulz"
date: "2026-09-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


Univariate Cox screen, LUAD and LUSC separately.
Re-implements the rank-based inverse-normal transform + MIN_EVENTS/MIN_PREV
style filtering from code/01_feature_selection.R (reference only), in base
R. LUSC is systematically underpowered (12 events cohort-wide, as low as 7
once RNA missingness is applied) -- thresholds are relaxed for LUSC and
every LUSC output is labelled EXPLORATORY.

QC gate (previously computed in 03_qc_plots.R but never consumed
downstream -- pass_qc there only requires >=2 unique values and <50%
missing within a histology, and was purely descriptive): candidate
features are now first restricted, per histology, to those with
pass_qc==TRUE in results/qc_by_category_histology.rds, BEFORE the
MIN_EVENTS/MIN_PREV screening below is applied. In practice this drops
only a handful of rare driver-mutation columns per histology (features
that clear the QC bar but never had enough mutation-positive patients to
clear MIN_PREV anyway), so screened results are numerically almost
unchanged -- the point of this change is that the QC step is now an
explicit, traceable upstream gate instead of a coincidentally-similar
parallel computation.


# setup
```{r}
suppressWarnings(suppressMessages(library(survival)))

RESULTS_DIR <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/survival/Rout/"
#PLOTS_DIR   <- "plots"


log_msg <- function(...) cat(format(Sys.time(), "%H:%M:%S"), "-", ..., "\n")

full_wide     <- readRDS(file.path(RESULTS_DIR, "full_wide.rds"))
feature_dict  <- readRDS(file.path(RESULTS_DIR, "feature_dictionary.rds"))
qc_all        <- readRDS(file.path(RESULTS_DIR, "qc_by_category_histology.rds"))

```

# helper functions
 Thresholds. LUAD keeps the reference defaults; LUSC's thresholds are
 deliberately relaxed because with only 12 events cohort-wide (and as few
 as 7 once RNA missingness is applied), MIN_EVENTS=10/MIN_PREV=10 would
 leave almost nothing to test. This is documented explicitly in the
 decisions log; LUSC results are exploratory/hypothesis-generating only.

```{r}

THRESH <- list(
  LUAD = list(MIN_EVENTS = 10, MIN_PREV = 10),
  LUSC = list(MIN_EVENTS = 5,  MIN_PREV = 5)
)

inv_norm <- function(x) {
  n <- sum(!is.na(x)); if (n < 3) return(rep(NA_real_, length(x)))
  qnorm((rank(x, na.last = "keep", ties.method = "average") - 0.5) / n)
}
tie_fraction <- function(x) {
  x <- x[!is.na(x)]; if (!length(x)) NA_real_ else max(table(x)) / length(x)
}

## single-predictor Cox model; returns one-row data.frame (or NA-row on failure)
fit_one <- function(d, xcol, lab) {
  na_row <- data.frame(scale = lab, hr = NA_real_, lo = NA_real_, hi = NA_real_,
                        coef = NA_real_, se = NA_real_, p_wald = NA_real_,
                        lr_chisq = NA_real_, p_lr = NA_real_,
                        cindex = NA_real_, p_ph = NA_real_, stringsAsFactors = FALSE)
  if (length(unique(stats::na.omit(d[[xcol]]))) < 2) return(na_row)

  f <- stats::as.formula(paste("Surv(time_to_event, status) ~", xcol))
  fit <- try(coxph(f, data = d, ties = "efron"), silent = TRUE)
  if (inherits(fit, "try-error") || !all(is.finite(stats::coef(fit)))) return(na_row)

  s  <- summary(fit)
  cf <- s$coefficients[xcol, ]
  ci <- s$conf.int[xcol, ]
  lr <- 2 * (fit$loglik[2] - fit$loglik[1])

  zph  <- try(cox.zph(fit), silent = TRUE)
  p_ph <- if (inherits(zph, "try-error") || !xcol %in% rownames(zph$table)) NA_real_
          else unname(zph$table[xcol, "p"])

  data.frame(scale = lab,
             hr = unname(ci["exp(coef)"]), lo = unname(ci["lower .95"]),
             hi = unname(ci["upper .95"]), coef = unname(cf["coef"]),
             se = unname(cf["se(coef)"]), p_wald = unname(cf["Pr(>|z|)"]),
             lr_chisq = lr,
             p_lr = if (is.na(lr)) NA_real_ else stats::pchisq(lr, 1, lower.tail = FALSE),
             cindex = unname(concordance(fit)$concordance),
             p_ph = p_ph, stringsAsFactors = FALSE)
}

fit_feature <- function(nm, category, d_base, min_events, min_prev) {
  x <- d_base[[nm]]
  ok <- !is.na(x)
  n_ev <- sum(d_base$status[ok])
  if (n_ev < min_events) return(NULL)

  is_binary <- identical(category, "mutation")
  d <- d_base[ok, , drop = FALSE]
  xv <- x[ok]

  if (is_binary) {
    tb <- table(xv)
    if (length(tb) < 2 || min(tb) < min_prev) return(NULL)
    d$x_rank <- xv; d$x_sd <- xv
    prev <- mean(xv == 1); tf <- NA_real_; unit <- "present vs absent"
  } else {
    if (length(unique(xv)) < 3) return(NULL)
    d$x_rank <- inv_norm(xv)
    sdv <- stats::sd(xv)
    d$x_sd <- if (sdv > 0) (xv - mean(xv)) / sdv else NA_real_
    prev <- NA_real_; tf <- tie_fraction(xv); unit <- "per 1 SD normal scores"
  }

  n_ev2 <- sum(d$status)
  if (n_ev2 < min_events) return(NULL)

  out <- rbind(fit_one(d, "x_rank", "rank_INT"), fit_one(d, "x_sd", "per_SD"))
  out$name <- nm; out$category <- category
  out$n <- nrow(d); out$n_events <- n_ev2
  out$prevalence <- prev; out$tie_frac <- tf; out$scale_unit <- unit
  out
}

run_screen <- function(histology_label, full_wide, feature_dict, qc_all, min_events, min_prev) {
  d_base <- full_wide[full_wide$simple_histology ==
                         ifelse(histology_label == "LUAD", "Adenocarcinoma", "Squamous cell carcinoma"), ]
  d_base <- d_base[!is.na(d_base$time_to_event) & d_base$time_to_event > 0 & !is.na(d_base$status), ]
  log_msg(histology_label, ": n=", nrow(d_base), " events=", sum(d_base$status))

  ## QC gate: only screen features that passed the per-histology QC check
  ## from 03_qc_plots.R (>=2 unique values, <50% missing within this
  ## histology). See header comment for why this used to be a no-op.
  qc_pass_names <- qc_all$name[qc_all$histology == histology_label & qc_all$pass_qc]
  feature_dict_h <- feature_dict[feature_dict$name %in% qc_pass_names, ]
  log_msg(histology_label, ": ", nrow(feature_dict) - nrow(feature_dict_h), " / ", nrow(feature_dict),
          " candidate features dropped by the QC gate (failed >=2 unique values or <50% missing",
          " within ", histology_label, ")")

  res_list <- vector("list", nrow(feature_dict_h))
  for (k in seq_len(nrow(feature_dict_h))) {
    nm  <- feature_dict_h$name[k]
    cat_k <- feature_dict_h$category[k]
    res_list[[k]] <- fit_feature(nm, cat_k, d_base, min_events, min_prev)
  }
  res <- do.call(rbind, res_list)
  res$histology <- histology_label
  res
}

```

# run survival association
```{r}
if(!file.exists(file.path(RESULTS_DIR, "cox_luad.rds")) & !file.exists(file.path(RESULTS_DIR, "cox_lusc.rds"))){
  log_msg("Running LUAD screen (MIN_EVENTS=", THRESH$LUAD$MIN_EVENTS,
          ", MIN_PREV=", THRESH$LUAD$MIN_PREV, ") ...")
  t0 <- Sys.time()
  cox_luad <- run_screen("LUAD", full_wide, feature_dict, qc_all,
                         THRESH$LUAD$MIN_EVENTS, THRESH$LUAD$MIN_PREV)
  log_msg("LUAD screen done in ", round(as.numeric(difftime(Sys.time(), t0, units = "secs")), 1),
          "s -- ", nrow(cox_luad) / 2, " features tested")
  
  log_msg("Running LUSC screen (EXPLORATORY; MIN_EVENTS=", THRESH$LUSC$MIN_EVENTS,
          ", MIN_PREV=", THRESH$LUSC$MIN_PREV, ") ...")
  t0 <- Sys.time()
  cox_lusc <- run_screen("LUSC", full_wide, feature_dict, qc_all,
                         THRESH$LUSC$MIN_EVENTS, THRESH$LUSC$MIN_PREV)
  log_msg("LUSC screen done in ", round(as.numeric(difftime(Sys.time(), t0, units = "secs")), 1),
          "s -- ", nrow(cox_lusc) / 2, " features tested")
  # FDR


add_fdr <- function(res) {
  res$fdr_within_category <- NA_real_
  for (sc in unique(res$scale)) {
    for (cc in unique(res$category)) {
      idx <- res$scale == sc & res$category == cc
      res$fdr_within_category[idx] <- stats::p.adjust(res$p_lr[idx], method = "BH")
    }
  }
  res$fdr_global <- NA_real_
  for (sc in unique(res$scale)) {
    idx <- res$scale == sc
    res$fdr_global[idx] <- stats::p.adjust(res$p_lr[idx], method = "BH")
  }
  res
}
cox_luad <- add_fdr(cox_luad)
cox_lusc <- add_fdr(cox_lusc)

cox_all <- rbind(cox_luad, cox_lusc)

saveRDS(cox_luad, file.path(RESULTS_DIR, "cox_luad.rds"))
saveRDS(cox_lusc, file.path(RESULTS_DIR, "cox_lusc.rds"))
saveRDS(cox_all,  file.path(RESULTS_DIR, "cox_all.rds"))
log_msg("Saved results/cox_luad.rds, results/cox_lusc.rds, results/cox_all.rds")

} else {
  cox_luad <- readRDS(file.path(RESULTS_DIR, "cox_luad.rds"))
  cox_lusc <- readRDS(file.path(RESULTS_DIR, "cox_lusc.rds"))
}

```

# Feature count
Final per-category, per-histology feature counts (for the Methods text):
how many candidate features existed, how many the QC gate dropped, and how
many were ultimately fit and tested (survived QC + MIN_EVENTS/MIN_PREV) --
i.e. the actual denominator behind cox_luad.rds/cox_lusc.rds and the
screen's implicit FDR correction, not just 03_qc_plots.R's descriptive
pass/fail count.

```{r}
categories_all <- sort(unique(feature_dict$category))

count_tested <- function(res, histology_label) {
  sub <- res[res$scale == "rank_INT" & !is.na(res$p_lr), ]
  tab <- table(factor(sub$category, levels = categories_all))
  data.frame(category = names(tab), n_tested = as.integer(tab),
             histology = histology_label, stringsAsFactors = FALSE)
}
count_dropped_qc <- function(histology_label) {
  qc_pass_names <- qc_all$name[qc_all$histology == histology_label & qc_all$pass_qc]
  dropped <- feature_dict[!feature_dict$name %in% qc_pass_names, ]
  tab <- table(factor(dropped$category, levels = categories_all))
  data.frame(category = names(tab), n_dropped_qc = as.integer(tab),
             histology = histology_label, stringsAsFactors = FALSE)
}

n_candidate <- data.frame(category = categories_all,
                           n_candidate = as.integer(table(factor(feature_dict$category, levels = categories_all))),
                           stringsAsFactors = FALSE)

tested_tab <- merge(count_tested(cox_luad, "LUAD"),  count_dropped_qc("LUAD"),  by = c("category", "histology"))
tested_tab <- rbind(tested_tab,
                     merge(count_tested(cox_lusc, "LUSC"), count_dropped_qc("LUSC"), by = c("category", "histology")))
tested_tab <- merge(n_candidate, tested_tab, by = "category")
tested_tab$n_dropped_threshold <- with(tested_tab, n_candidate - n_dropped_qc - n_tested)
tested_tab <- tested_tab[order(tested_tab$category, tested_tab$histology),
                          c("category", "histology", "n_candidate", "n_dropped_qc",
                            "n_dropped_threshold", "n_tested")]

log_msg("Final feature counts by category x histology (candidate -> QC-dropped -> ",
        "MIN_EVENTS/MIN_PREV-dropped -> tested):")
print(tested_tab)

saveRDS(tested_tab, file.path(RESULTS_DIR, "n_features_tested_by_category.rds"))
write.csv(tested_tab, file.path(RESULTS_DIR, "n_features_tested_by_category.csv"), row.names = FALSE)
log_msg("Saved results/n_features_tested_by_category.rds, results/n_features_tested_by_category.csv")

## summary table, printed for the log
summarize_screen <- function(res, label) {
  sub <- res[res$scale == "rank_INT", ]
  agg <- aggregate(cbind(tested = !is.na(p_lr),
                          raw05 = p_lr < 0.05,
                          fdr10 = fdr_within_category < 0.1) ~ category,
                    data = sub, FUN = function(v) sum(v, na.rm = TRUE))
  agg$histology <- label
  agg
}
summ_luad <- summarize_screen(cox_luad, "LUAD")
summ_lusc <- summarize_screen(cox_lusc, "LUSC")
log_msg("LUAD summary (rank_INT scale):"); print(summ_luad)
log_msg("LUSC summary (rank_INT scale, EXPLORATORY):"); print(summ_lusc)
```




# PLOT 1: plots/qq_pvalues_luad_lusc.pdf -- observed vs expected -log10(p)
```{r}
#pdf(file.path(PLOTS_DIR, "qq_pvalues_luad_lusc.pdf"), width = 10, height = 5.5)
op <- par(mfrow = c(1, 2), mar = c(5, 5, 4, 2))
qq_panel <- function(res, title_lab, col_pt) {
  p <- res$p_lr[res$scale == "rank_INT"]
  p <- p[!is.na(p)]
  p <- sort(p)
  n <- length(p)
  expected <- -log10(((seq_len(n)) - 0.5) / n)
  observed <- -log10(p)
  lim <- c(0, max(c(expected, observed), na.rm = TRUE) * 1.05)
  plot(expected, observed, pch = 16, cex = 0.5, col = col_pt,
       xlim = lim, ylim = lim,
       xlab = expression(Expected~-log[10](p)), ylab = expression(Observed~-log[10](p)),
       main = title_lab)
  abline(0, 1, col = "grey40", lty = 2)
}
qq_panel(cox_luad, paste0("LUAD (n=", sum(cox_luad$n[cox_luad$scale=="rank_INT"] > 0, na.rm=TRUE),
                           " features tested, ", length(unique(full_wide$immucan_id[full_wide$simple_histology=="Adenocarcinoma"])),
                           " patients, 22 events)"), "blue")
qq_panel(cox_lusc, "LUSC -- EXPLORATORY (only 12 events cohort-wide)", "#d95f02")
par(op)
#dev.off()
```



# PLOT 2: plots/volcano_luad_lusc.pdf -- log(HR) vs -log10(p), by category

```{r}
cats <- sort(unique(feature_dict$category))
cat_cols <- setNames(c("blue","#d95f02","#7570b3","#e7298a","#66a61e","#e6ab02","#a6761d","#666666")[seq_along(cats)], cats)

#pdf(file.path(PLOTS_DIR, "volcano_luad_lusc.pdf"), width = 12, height = 6)
op <- par(mfrow = c(1, 2), mar = c(5, 5, 4, 8), xpd = FALSE)
volcano_panel <- function(res, title_lab) {
  sub <- res[res$scale == "rank_INT" & !is.na(res$p_lr), ]
  plot(sub$coef, -log10(sub$p_lr), pch = 16, cex = 0.6,
       col = cat_cols[sub$category],
       xlab = "log(HR)  [rank-based inverse-normal scale]",
       ylab = expression(-log[10](p[LR])),
       main = title_lab)
  abline(h = -log10(0.05), lty = 2, col = "grey50")
  abline(v = 0, lty = 3, col = "grey70")
}
volcano_panel(cox_luad, "LUAD")
volcano_panel(cox_lusc, "LUSC -- EXPLORATORY (12 events cohort-wide)")
par(xpd = NA)
legend(x = par("usr")[2] * 1.05, y = par("usr")[4],
       legend = names(cat_cols), col = cat_cols, pch = 16, bty = "n", cex = 0.8)
par(op)
#dev.off()
```



## PLOT 3: plots/hitcounts_by_category_histology.pdf
```{r}
hit_tab <- function(res) {
  sub <- res[res$scale == "rank_INT", ]
  data.frame(
    category = cats,
    raw05 = sapply(cats, function(cc) sum(sub$p_lr[sub$category == cc] < 0.05, na.rm = TRUE)),
    fdr10 = sapply(cats, function(cc) sum(sub$fdr_within_category[sub$category == cc] < 0.1, na.rm = TRUE))
  )
}
h_luad <- hit_tab(cox_luad); h_lusc <- hit_tab(cox_lusc)

hit_mat <- rbind(LUAD_raw05 = h_luad$raw05, LUAD_fdr10 = h_luad$fdr10,
                  LUSC_raw05 = h_lusc$raw05, LUSC_fdr10 = h_lusc$fdr10)
colnames(hit_mat) <- cats

#pdf(file.path(PLOTS_DIR, "hitcounts_by_category_histology.pdf"), width = 11, height = 6.5)
op <- par(mar = c(8, 5, 4, 2))
cols4 <- c("blue", "lightblue", "orange", "yellow2")
bp <- barplot(hit_mat, beside = TRUE, las = 2, col = cols4,
              ylab = "Number of features",
              main = "Nominal p<0.05 and FDR<0.1 hit counts by category\n(rank_INT scale; LUSC is EXPLORATORY, 12 events cohort-wide)",
              cex.names = 0.9)
legend("topright", legend = c("LUAD nominal p<0.05", "LUAD FDR<0.1",
                                "LUSC nominal p<0.05 (exploratory)", "LUSC FDR<0.1 (exploratory)"),
       fill = cols4, bty = "n", cex = 0.75)
par(op)
#dev.off()
```




# PLOT 4: plots/overlap_luad_lusc_hits.pdf -- overlap of nominal hits
```{r}
hits_luad <- unique(cox_luad$name[cox_luad$scale == "rank_INT" & !is.na(cox_luad$p_lr) & cox_luad$p_lr < 0.05])
hits_lusc <- unique(cox_lusc$name[cox_lusc$scale == "rank_INT" & !is.na(cox_lusc$p_lr) & cox_lusc$p_lr < 0.05])
n_luad_only <- length(setdiff(hits_luad, hits_lusc))
n_shared    <- length(intersect(hits_luad, hits_lusc))
n_lusc_only <- length(setdiff(hits_lusc, hits_luad))
log_msg("Nominal p<0.05 hits: LUAD-only=", n_luad_only, " shared=", n_shared, " LUSC-only=", n_lusc_only,
        " (LUAD total=", length(hits_luad), ", LUSC total=", length(hits_lusc), ")")

#pdf(file.path(PLOTS_DIR, "overlap_luad_lusc_hits.pdf"), width = 7, height = 6)
op <- par(mar = c(5, 5, 4, 2))
counts <- c(`LUAD only` = n_luad_only, `Shared` = n_shared, `LUSC only (exploratory)` = n_lusc_only)
bp <- barplot(counts, col = c("blue", "red", "orange"),
              ylab = "Number of nominally significant features (p<0.05, rank_INT)",
              main = paste0("Overlap of nominal hits, LUAD vs LUSC\n",
                             "(LUAD ", length(hits_luad), " hits / LUSC ", length(hits_lusc),
                             " hits -- different feature sets tested at different thresholds;\n",
                             "LUSC hits are EXPLORATORY given only 12 events)"),
              cex.main = 0.85, ylim = c(0, max(counts) * 1.2))
text(bp, counts, labels = counts, pos = 3)
par(op)
#dev.off()

log_msg("Saved plots: qq_pvalues_luad_lusc.pdf, volcano_luad_lusc.pdf, hitcounts_by_category_histology.pdf, overlap_luad_lusc_hits.pdf")
log_msg("DONE 04_luad_lusc_survival.R")
```


