---
title: "Spatial vs non-spatial"
author: "Daniel Schulz"
date: "2026-09-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

05_spatial_vs_nonspatial.R
The reviewer's actual ask: is there more survival signal among SPATIAL
features (interaction, cellohood) than among NON-spatial features
(density, RNA_gene, RNA_module, mutation, biomarker)?
image_module is deliberately excluded from this strict binary test (see
decisions log -- it is spatially-derived but expressed as a per-patient
summary index, so a binary spatial/non-spatial call is debatable).

P-VALUES BELOW ***
The parametric KS/Wilcoxon test in section 1 (kept for continuity with round
1) treats every feature's p-value as an *independent* draw. That assumption
is false here: interaction features are pairwise combinations of a small set
of cell types (e.g. DC.Neutrophil, T.Neutrophil, MacCD163.Neutrophil,
Neutrophil.Tumor, Neutrophil.Neutrophil, Tumor.Neutrophil all repeat the same
'Neutrophil' signal against different partners), cellohood CN values are raw
per-patient cell counts (NOT normalized fractions -- corrected in round 4,
see docs/decisions_round4.md) whose per-patient totals vary widely (~3,090-
13,371 for cello1, ~1,472-14,924 for cello2), so a patient's own CN counts
are not mutually independent, and RNA_gene features are co-expressed in modules. We verified this directly in
LUAD: 7 of the top-8 spatial hits are Neutrophil-partner interaction terms,
and their Spearman correlation with the single non-spatial feature
'Neutrophil_density__IF1' (itself the top density-category hit, p=0.0003)
ranges from rho=0.83 to rho=0.97 (n=115) -- i.e. much of the apparent
spatial-category signal is the same cell-type-abundance effect leaking into
the interaction category via shared partner cell types, not necessarily a
genuinely spatial/arrangement effect. This makes the parametric KS/Wilcoxon
p-values in section 1 ANTI-CONSERVATIVE (pseudo-replication inflates
apparent significance). Section 1a below quantifies this confounding
directly; section 2 (NEW) implements the remedy: a patient-label permutation
test that shuffles outcome (time, status) across patients while leaving the
feature matrix -- and hence its real correlation/redundancy structure --
completely untouched. Under that null, correlated features stay correlated,
so the resulting reference distribution of the enrichment statistic already
has the real pseudo-replication baked in, and the resulting empirical
p-value is the trustworthy one. See docs/decisions_round2.md section on
"Independence-assumption caveat" for the full writeup.


```{r}
suppressWarnings(suppressMessages(library(survival)))

RESULTS_DIR <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/survival/Rout/"
PLOTS_DIR   <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/survival/plots/"
suppressWarnings(suppressMessages(library(survival)))


log_msg <- function(...) cat(format(Sys.time(), "%H:%M:%S"), "-", ..., "\n")

set.seed(20260911)  # reproducible bootstrap

full_wide <- readRDS(file.path(RESULTS_DIR, "full_wide.rds"))
cox_luad  <- readRDS(file.path(RESULTS_DIR, "cox_luad.rds"))
cox_lusc  <- readRDS(file.path(RESULTS_DIR, "cox_lusc.rds"))

SPATIAL_CATS     <- c("interaction", "cellohood")
NONSPATIAL_CATS  <- c("density", "RNA_gene", "RNA_module", "mutation", "biomarker")
```


# 1. Enrichment test: p-value distribution, spatial vs non-spatial
```{r}
enrichment_test <- function(res, histology_label) {
  sub <- res[res$scale == "rank_INT" & res$category %in% c(SPATIAL_CATS, NONSPATIAL_CATS) &
               !is.na(res$p_lr), ]
  sub$is_spatial <- sub$category %in% SPATIAL_CATS

  p_spatial    <- sub$p_lr[sub$is_spatial]
  p_nonspatial <- sub$p_lr[!sub$is_spatial]

  ks  <- try(ks.test(p_spatial, p_nonspatial), silent = TRUE)
  wt  <- try(wilcox.test(-log10(p_spatial), -log10(p_nonspatial)), silent = TRUE)

  data.frame(
    histology = histology_label,
    n_spatial = length(p_spatial), n_nonspatial = length(p_nonspatial),
    median_p_spatial = median(p_spatial), median_p_nonspatial = median(p_nonspatial),
    ks_D = if (inherits(ks, "try-error")) NA_real_ else unname(ks$statistic),
    ks_p = if (inherits(ks, "try-error")) NA_real_ else ks$p.value,
    wilcox_W = if (inherits(wt, "try-error")) NA_real_ else unname(wt$statistic),
    wilcox_p = if (inherits(wt, "try-error")) NA_real_ else wt$p.value,
    stringsAsFactors = FALSE
  )
}

enr_luad <- enrichment_test(cox_luad, "LUAD")
enr_lusc <- enrichment_test(cox_lusc, "LUSC")
enr_all  <- rbind(enr_luad, enr_lusc)
enr_all$independence_caveat <- paste(
  "ANTI-CONSERVATIVE: this KS/Wilcoxon p assumes independent feature draws;",
  "the feature pool is heavily non-independent (interaction terms share partner",
  "cell types, e.g. Neutrophil-partner terms correlate rho=0.83-0.97 with",
  "Neutrophil_density__IF1; cellohood CN values are raw per-patient counts with",
  "widely varying per-patient totals, not normalized fractions -- see docs/decisions_round4.md). See",
  "results/spatial_enrichment_permutation.csv for a permutation-calibrated",
  "p-value and docs/decisions_round2.md for the full writeup."
)
log_msg("Spatial vs non-spatial enrichment test (p-value distribution, rank_INT scale):")
print(enr_all)
log_msg("*** CAVEAT (see docs/decisions_round2.md): the KS/Wilcoxon p-values above assume",
        "independent feature draws, which is FALSE for this feature pool (see header comment",
        "of this script). Treat them as an anti-conservative first look; the",
        "permutation-calibrated result computed below is the trustworthy one. ***")
write.csv(enr_all, file.path(RESULTS_DIR, "spatial_enrichment_test.csv"), row.names = FALSE)
saveRDS(enr_all, file.path(RESULTS_DIR, "spatial_enrichment_test.rds"))

if (!is.na(enr_luad$ks_p) && enr_luad$ks_p < 0.05) {
  log_msg("LUAD: KS test IS significant (p=", signif(enr_luad$ks_p, 3),
          ") -- spatial and non-spatial p-value distributions differ. (Parametric; see caveat above.)")
} else {
  log_msg("LUAD: KS test is NOT significant (p=", signif(enr_luad$ks_p, 3),
          ") -- no strong evidence the spatial p-value distribution differs from non-spatial. Reporting honestly.")
}
if (!is.na(enr_lusc$ks_p) && enr_lusc$ks_p < 0.05) {
  log_msg("LUSC (EXPLORATORY): KS test IS nominally significant (p=", signif(enr_lusc$ks_p, 3),
          "). (Parametric; see caveat above.)")
} else {
  log_msg("LUSC (EXPLORATORY): KS test is NOT significant (p=", signif(enr_lusc$ks_p, 3), ").")
}
```


1a. Quantify the density/interaction confounding directly (round 2).
For each histology's top-8 spatial hits, find any non-spatial density
feature that shares a cell-type name with the interaction term's two
partners, and report the Spearman correlation. This turns the reviewer's
qualitative observation into a concrete, reproducible number.

```{r}
quantify_density_confounding <- function(cox_res, full_wide, histology_str, histology_label, n_top = 8) {
  d <- full_wide[full_wide$simple_histology == histology_str, ]
  sub <- cox_res[cox_res$scale == "rank_INT" & cox_res$category == "interaction" & !is.na(cox_res$p_lr), ]
  sub <- sub[order(sub$p_lr), ]
  top <- head(sub, n_top)
  density_names <- grep("_density__", colnames(d), value = TRUE)
  density_celltypes <- sub("_density__.*$", "", density_names)

  out <- vector("list", nrow(top))
  for (i in seq_len(nrow(top))) {
    nm <- top$name[i]
    partners <- strsplit(sub("__.*$", "", nm), "\\.")[[1]]
    hit_idx <- which(density_celltypes %in% partners)
    if (length(hit_idx) == 0) {
      out[[i]] <- data.frame(histology = histology_label, interaction_feature = nm,
                              partner_density_feature = NA_character_,
                              rho = NA_real_, p = NA_real_, n = NA_integer_,
                              stringsAsFactors = FALSE)
      next
    }
    rows <- lapply(hit_idx, function(j) {
      dens_nm <- density_names[j]
      ok <- !is.na(d[[nm]]) & !is.na(d[[dens_nm]])
      if (sum(ok) < 10) {
        return(data.frame(histology = histology_label, interaction_feature = nm,
                           partner_density_feature = dens_nm, rho = NA_real_, p = NA_real_,
                           n = sum(ok), stringsAsFactors = FALSE))
      }
      ct <- suppressWarnings(try(cor.test(d[[nm]][ok], d[[dens_nm]][ok], method = "spearman"), silent = TRUE))
      data.frame(histology = histology_label, interaction_feature = nm,
                 partner_density_feature = dens_nm,
                 rho = if (inherits(ct, "try-error")) NA_real_ else unname(ct$estimate),
                 p = if (inherits(ct, "try-error")) NA_real_ else ct$p.value,
                 n = sum(ok), stringsAsFactors = FALSE)
    })
    out[[i]] <- do.call(rbind, rows)
  }
  do.call(rbind, out)
}

confound_luad <- quantify_density_confounding(cox_luad, full_wide, "Adenocarcinoma", "LUAD", n_top = 8)
confound_lusc <- quantify_density_confounding(cox_lusc, full_wide, "Squamous cell carcinoma", "LUSC", n_top = 8)
confound_all  <- rbind(confound_luad, confound_lusc)
log_msg("Density/interaction confounding check (top-8 spatial hits per histology vs. same-cell-type density features):")
print(confound_all)
write.csv(confound_all, file.path(RESULTS_DIR, "spatial_density_confounding_check.csv"), row.names = FALSE)
saveRDS(confound_all, file.path(RESULTS_DIR, "spatial_density_confounding_check.rds"))
max_rho_luad <- suppressWarnings(max(abs(confound_luad$rho), na.rm = TRUE))
log_msg("LUAD: ", sum(!is.na(confound_luad$rho)), "/", nrow(confound_luad),
        " top-8 spatial hits have a matching same-cell-type density feature; max |rho|=",
        signif(max_rho_luad, 3), " -- confirms these are NOT independent signals.")
```


# 2. Null estimation

2. REMEDY (round 2, extended round 5): patient-label permutation test.
Shuffle (time_to_event, status) across patients within a histology, leaving
the feature matrix -- and therefore its true correlation/redundancy
structure -- completely untouched. Refit every feature's Cox model under
each permutation with the SAME full model used in 04 (ties="efron", full
convergence -- benchmarked round 5 at effectively the same wall-time as the
previous iter.max=0 score-test shortcut for this dataset size, so there is
no accuracy/speed trade-off here; the shortcut is retired), and from each
permutation's full set of refits, rebuild THREE separate statistics:
 (a) the original p-value-based KS-D / directional (mean -log10 p diff)
     comparison, spatial vs non-spatial;
 (b) round 5 NEW: the same KS-D / directional comparison on |coef| (the
     rank-INT-scale effect size, i.e. log-HR per 1 SD) instead of p, since
     a p-value conflates true effect size with how precisely it happened to
     be estimated (missingness, per-feature power) -- comparing |coef|
     isolates the effect-size question. Binary `mutation` features are
     excluded from this comparison only: their coefficient is a log-HR for
     presence/absence, not "per 1 SD", so it is not on a comparable scale.
 (c) round 5 NEW: per-category (ALL categories, INCLUDING image_module --
     there is no spatial/non-spatial split here at all) count of nominal
     hits (p<0.05), to test directly whether the number of hits we detect
     in each category is distinguishable from the number chance alone would
     produce under this exact permutation null.
Because correlated features are shuffled together as a block (same
permutation vector applied to the whole feature matrix), any feature is
permuted jointly with all other features for a given permutation, so every
one of these null distributions already reflects the real non-independence
-- unlike the parametric KS null in section 1, which assumes iid p-values.

```{r}
inv_norm2 <- function(x) {
  n <- sum(!is.na(x)); if (n < 3) return(rep(NA_real_, length(x)))
  qnorm((rank(x, na.last = "keep", ties.method = "average") - 0.5) / n)
}

HR_EXCLUDE_CATS <- c("mutation")  ## excluded from the |coef| comparison only (see header)

fit_perm_feature <- function(x_raw, time, status, is_binary) {
  ok <- !is.na(x_raw)
  if (sum(ok) < 10) return(c(p = NA_real_, coef = NA_real_))
  xv <- x_raw[ok]; tt <- time[ok]; ss <- status[ok]
  if (is_binary) {
    if (length(unique(xv)) < 2) return(c(p = NA_real_, coef = NA_real_))
    x <- xv
  } else {
    if (length(unique(xv)) < 3) return(c(p = NA_real_, coef = NA_real_))
    x <- inv_norm2(xv)
  }
  fit <- try(coxph(Surv(tt, ss) ~ x, ties = "efron"), silent = TRUE)
  if (inherits(fit, "try-error") || !all(is.finite(stats::coef(fit)))) return(c(p = NA_real_, coef = NA_real_))
  s <- summary(fit)
  c(p = unname(s$logtest["pvalue"]), coef = unname(stats::coef(fit)))
}

enrichment_stats <- function(p_vec, coef_vec, pool, categories) {
  is_spat    <- pool$category %in% SPATIAL_CATS
  is_nonspat <- pool$category %in% NONSPATIAL_CATS
  is_hr_excl <- pool$category %in% HR_EXCLUDE_CATS

  ks_p   <- try(ks.test(p_vec[is_spat], p_vec[is_nonspat]), silent = TRUE)
  D_p    <- if (inherits(ks_p, "try-error")) NA_real_ else unname(ks_p$statistic)
  diff_p <- mean(-log10(p_vec[is_spat]), na.rm = TRUE) - mean(-log10(p_vec[is_nonspat]), na.rm = TRUE)

  hr_spatial    <- abs(coef_vec[is_spat & !is_hr_excl])
  hr_nonspatial <- abs(coef_vec[is_nonspat & !is_hr_excl])
  ks_hr   <- try(ks.test(hr_spatial, hr_nonspatial), silent = TRUE)
  D_hr    <- if (inherits(ks_hr, "try-error")) NA_real_ else unname(ks_hr$statistic)
  diff_hr <- mean(hr_spatial, na.rm = TRUE) - mean(hr_nonspatial, na.rm = TRUE)

  hits_by_cat <- vapply(categories, function(cc) sum(p_vec[pool$category == cc] < 0.05, na.rm = TRUE), numeric(1))

  c(D_p = D_p, diff_p = diff_p, D_hr = D_hr, diff_hr = diff_hr, hits_by_cat)
}

permutation_enrichment_test <- function(cox_res, full_wide, histology_str, histology_label, B = 200, seed = 20260911) {
  d_base <- full_wide[full_wide$simple_histology == histology_str, ]
  d_base <- d_base[!is.na(d_base$time_to_event) & d_base$time_to_event > 0 & !is.na(d_base$status), ]

  ## ALL categories, including image_module -- the per-category hit-count
  ## comparison (c) does not use the spatial/non-spatial split at all; that
  ## split is applied only inside enrichment_stats() for comparisons (a)/(b).
  pool <- cox_res[cox_res$scale == "rank_INT" & !is.na(cox_res$p_lr), c("name", "category")]
  pool <- pool[!duplicated(pool$name), ]
  pool$is_binary <- pool$category == "mutation"
  categories <- sort(unique(pool$category))

  compute_pool_stats <- function(time, status) {
    m <- vapply(seq_len(nrow(pool)), function(k) {
      fit_perm_feature(d_base[[pool$name[k]]], time, status, pool$is_binary[k])
    }, c(p = NA_real_, coef = NA_real_))
    list(p = m["p", ], coef = m["coef", ])
  }

  t0 <- Sys.time()
  obs_pc <- compute_pool_stats(d_base$time_to_event, d_base$status)
  obs <- enrichment_stats(obs_pc$p, obs_pc$coef, pool, categories)
  log_msg(histology_label, ": permutation test -- observed D_p=", signif(obs["D_p"], 3),
          " diff_p=", signif(obs["diff_p"], 3), " D_hr=", signif(obs["D_hr"], 3),
          " diff_hr=", signif(obs["diff_hr"], 3),
          "; one pass (full model, all ", nrow(pool), " features) took ",
          round(as.numeric(difftime(Sys.time(), t0, units = "secs")), 2),
          "s, running B=", B, " permutations ...")

  set.seed(seed)
  null_mat <- matrix(NA_real_, nrow = B, ncol = length(obs), dimnames = list(NULL, names(obs)))
    ## per-permutation, per-feature p-value (NEW): kept so a downstream script
  ## can reconstruct each permutation's actual hit SET (not just per-category
  ## counts) -- needed for the LUAD/LUSC hit-set overlap null test in
  ## code/09_luad_lusc_overlap_permutation.R. Nothing else about this loop
  ## changes; this is purely an additional thing recorded per permutation.
  p_null_mat <- matrix(NA_real_, nrow = B, ncol = nrow(pool), dimnames = list(NULL, pool$name))
  tstart <- Sys.time()
  for (b in seq_len(B)) {
    perm_idx <- sample(nrow(d_base))
    pc <- compute_pool_stats(d_base$time_to_event[perm_idx], d_base$status[perm_idx])
    null_mat[b, ] <- enrichment_stats(pc$p, pc$coef, pool, categories)
    p_null_mat[b, ] <- pc$p
    if (b %% 50 == 0) log_msg(histology_label, ": permutation ", b, "/", B, " done (",
                               round(as.numeric(difftime(Sys.time(), tstart, units = "secs")), 1), "s elapsed)")
  }

  emp_p_gt <- function(null_vec, obs_val) (1 + sum(null_vec >= obs_val, na.rm = TRUE)) / (1 + sum(!is.na(null_vec)))
  emp_p_2s <- function(null_vec, obs_val) (1 + sum(abs(null_vec) >= abs(obs_val), na.rm = TRUE)) / (1 + sum(!is.na(null_vec)))

  cat_summary <- data.frame(
    histology = histology_label, category = categories,
    n_tested = as.integer(table(factor(pool$category, levels = categories))),
    hits_obs = as.integer(obs[categories]),
    hits_null_mean = colMeans(null_mat[, categories, drop = FALSE], na.rm = TRUE),
    emp_p_hits_gt = vapply(categories, function(cc) emp_p_gt(null_mat[, cc], obs[cc]), numeric(1)),
    stringsAsFactors = FALSE
  )
  cat_summary$hit_rate_obs_pct       <- round(100 * cat_summary$hits_obs / cat_summary$n_tested, 2)
  cat_summary$hit_rate_null_mean_pct <- round(100 * cat_summary$hits_null_mean / cat_summary$n_tested, 2)

  list(
    summary = data.frame(
      histology = histology_label, B = B,
      n_spatial = sum(pool$category %in% SPATIAL_CATS), n_nonspatial = sum(pool$category %in% NONSPATIAL_CATS),
      D_obs = unname(obs["D_p"]), emp_p_ks = emp_p_gt(null_mat[, "D_p"], obs["D_p"]),
      diff_obs = unname(obs["diff_p"]),
      emp_p_diff_one_sided_spatial_gt = emp_p_gt(null_mat[, "diff_p"], obs["diff_p"]),
      emp_p_diff_two_sided = emp_p_2s(null_mat[, "diff_p"], obs["diff_p"]),
      D_obs_hr = unname(obs["D_hr"]), emp_p_ks_hr = emp_p_gt(null_mat[, "D_hr"], obs["D_hr"]),
      diff_obs_hr = unname(obs["diff_hr"]),
      emp_p_diff_hr_one_sided_spatial_gt = emp_p_gt(null_mat[, "diff_hr"], obs["diff_hr"]),
      emp_p_diff_hr_two_sided = emp_p_2s(null_mat[, "diff_hr"], obs["diff_hr"]),
      stringsAsFactors = FALSE
    ),
    category_summary = cat_summary,
    D_null = null_mat[, "D_p"], diff_null = null_mat[, "diff_p"],
    D_obs = unname(obs["D_p"]), diff_obs = unname(obs["diff_p"]),
    D_null_hr = null_mat[, "D_hr"], diff_null_hr = null_mat[, "diff_hr"],
    D_obs_hr = unname(obs["D_hr"]), diff_obs_hr = unname(obs["diff_hr"]),
    hits_null = null_mat[, categories, drop = FALSE], hits_obs = obs[categories],
    categories = categories,
    p_null = p_null_mat
  )
}

if(!file.exists( file.path(RESULTS_DIR, "spatial_enrichment_permutation.rds"))){
log_msg("Running permutation-calibrated enrichment test for LUAD (B=200, patient-label permutation, full Cox model) ...")
perm_luad <- permutation_enrichment_test(cox_luad, full_wide, "Adenocarcinoma", "LUAD", B = 200)
log_msg("Running permutation-calibrated enrichment test for LUSC (EXPLORATORY; B=200) ...")
perm_lusc <- permutation_enrichment_test(cox_lusc, full_wide, "Squamous cell carcinoma", "LUSC", B = 200)

perm_summary <- rbind(perm_luad$summary, perm_lusc$summary)
log_msg("Permutation-calibrated enrichment test results (trustworthy; corrects for feature non-independence; ",
        "p-value-based AND |coef| effect-size-based, spatial vs non-spatial):")
print(perm_summary)
write.csv(perm_summary, file.path(RESULTS_DIR, "spatial_enrichment_permutation.csv"), row.names = FALSE)

category_perm_summary <- rbind(perm_luad$category_summary, perm_lusc$category_summary)
log_msg("Per-category hit-rate permutation test (NO spatial/non-spatial split, includes image_module -- ",
        "observed hit count/rate vs. what this exact permutation null produces by chance):")
print(category_perm_summary)
write.csv(category_perm_summary, file.path(RESULTS_DIR, "category_hitrate_permutation.csv"), row.names = FALSE)

saveRDS(list(summary = perm_summary, category_summary = category_perm_summary,
             D_null_luad = perm_luad$D_null, diff_null_luad = perm_luad$diff_null,
             D_null_lusc = perm_lusc$D_null, diff_null_lusc = perm_lusc$diff_null,
             D_null_hr_luad = perm_luad$D_null_hr, diff_null_hr_luad = perm_luad$diff_null_hr,
             D_null_hr_lusc = perm_lusc$D_null_hr, diff_null_hr_lusc = perm_lusc$diff_null_hr,
             hits_null_luad = perm_luad$hits_null, hits_obs_luad = perm_luad$hits_obs,
             hits_null_lusc = perm_lusc$hits_null, hits_obs_lusc = perm_lusc$hits_obs,
             categories = perm_luad$categories,
             p_null_luad = perm_luad$p_null, p_null_lusc = perm_lusc$p_null),
        file.path(RESULTS_DIR, "spatial_enrichment_permutation.rds"))

log_msg("LUAD: parametric KS p=", signif(enr_luad$ks_p, 3), " vs. permutation-calibrated KS p (p-value based)=",
        signif(perm_luad$summary$emp_p_ks, 3), "; permutation-calibrated KS p (|coef| effect-size based)=",
        signif(perm_luad$summary$emp_p_ks_hr, 3))
log_msg("LUSC (EXPLORATORY): parametric KS p=", signif(enr_lusc$ks_p, 3), " vs. permutation-calibrated KS p (p-value based)=",
        signif(perm_lusc$summary$emp_p_ks, 3), "; permutation-calibrated KS p (|coef| effect-size based)=",
        signif(perm_lusc$summary$emp_p_ks_hr, 3))

} else {
  perm_summary <- readRDS(file.path(RESULTS_DIR, "spatial_enrichment_permutation.rds"))
}
```


## enrichment plot

Bonus visual for the enrichment claim: ECDF of p-values, spatial vs
non-spatial, per histology (not in the mandatory list but cheap and
directly supports the KS-test claim above), PLUS a mandatory caveat title
page and permutation-null histogram page added in round 2.

```{r}
cox_luad  <- readRDS(file.path(RESULTS_DIR, "cox_luad.rds"))
cox_lusc  <- readRDS(file.path(RESULTS_DIR, "cox_lusc.rds"))
full_wide <- readRDS(file.path(RESULTS_DIR, "full_wide.rds"))

SPATIAL_CATS    <- c("interaction", "cellohood")
NONSPATIAL_CATS <- c("density", "RNA_gene", "RNA_module", "mutation", "biomarker")

enr_luad <- enrichment_test(cox_luad, "LUAD")
enr_lusc <- enrichment_test(cox_lusc, "LUSC")

perm_rds <- readRDS(file.path(RESULTS_DIR, "spatial_enrichment_permutation.rds"))
mk_perm <- function(h) {
  list(
    summary          = perm_rds$summary[perm_rds$summary$histology == h, ],
    category_summary = perm_rds$category_summary[perm_rds$category_summary$histology == h, ],
    D_null       = perm_rds[[paste0("D_null_", tolower(h))]],
    diff_null    = perm_rds[[paste0("diff_null_", tolower(h))]],
    D_obs        = perm_rds$summary$D_obs[perm_rds$summary$histology == h],
    diff_obs     = perm_rds$summary$diff_obs[perm_rds$summary$histology == h],
    D_null_hr    = perm_rds[[paste0("D_null_hr_", tolower(h))]],
    diff_null_hr = perm_rds[[paste0("diff_null_hr_", tolower(h))]],
    D_obs_hr     = perm_rds$summary$D_obs_hr[perm_rds$summary$histology == h],
    diff_obs_hr  = perm_rds$summary$diff_obs_hr[perm_rds$summary$histology == h],
    hits_null    = perm_rds[[paste0("hits_null_", tolower(h))]],
    hits_obs     = perm_rds[[paste0("hits_obs_", tolower(h))]],
    categories   = perm_rds$categories
  )
}
perm_luad <- mk_perm("LUAD")
perm_lusc <- mk_perm("LUSC")
perm_luad
```

```{r}
pdf(file.path(PLOTS_DIR, "enrichment_pvalue_ecdf_luad_lusc.pdf"), width = 10, height = 7)

## --- Page 1 (round 2, NEW): prominent independence-assumption caveat ---
op <- par(mar = c(1, 1, 3, 1))
plot.new()
title("CAVEAT: feature non-independence and the spatial-enrichment test", cex.main = 1.1)
caveat_txt <- c(
  "The parametric KS/Wilcoxon test on the next page assumes each feature's",
  "p-value is an INDEPENDENT draw. That assumption is false for this feature pool:",
  "",
  paste0("  - Interaction features are pairwise combos of a small cell-type set: in LUAD, 7 of the top-8"),
  paste0("    spatial hits are Neutrophil-partner interaction terms (DC.Neutrophil, T.Neutrophil,"),
  paste0("    MacCD163.Neutrophil, Neutrophil.Tumor, Neutrophil.Neutrophil, Tumor.Neutrophil, ..."),
  paste0("    other.Neutrophil), correlating rho=0.83-0.97 (Spearman, n=115) with the single non-spatial"),
  paste0("    feature Neutrophil_density__IF1 (itself the top density hit, p=0.0003). See"),
  paste0("    results/spatial_density_confounding_check.csv for the full numbers."),
  "  -Cellohood CN values are raw per-patient densities.",
  "  - RNA_gene features are co-expressed in modules, not independent.",
  "",
  "=> The parametric KS p-values (this page's legend, and results/spatfcaial_enrichment_test.csv) are",
  "   ANTI-CONSERVATIVE and should not be read at face value.",
  "",
  "REMEDY: page 3 shows a patient-label permutation test (results/spatial_enrichment_permutation.csv).",
  "Outcome labels are shuffled across patients while the feature matrix -- and hence its real",
  "correlation structure -- is left untouched, so correlated features are permuted together and the",
  "resulting null distribution already reflects this non-independence. That empirical p-value is the",
  "one to trust; see docs/decisions_round2.md for full detail."
)
text(0.02, seq(0.95, 0.95 - 0.045 * (length(caveat_txt) - 1), length.out = length(caveat_txt)),
     labels = caveat_txt, adj = c(0, 1), cex = 0.72, family = "mono")
par(op)

## --- Page 2: original ECDF panels, legend now also carries the caveat flag ---
op <- par(mfrow = c(1, 2), mar = c(5, 5, 4, 2))
ecdf_panel <- function(res, enr_row, perm_row, title_lab) {
  sub <- res[res$scale == "rank_INT" & res$category %in% c(SPATIAL_CATS, NONSPATIAL_CATS) &
               !is.na(res$p_lr), ]
  sub$is_spatial <- sub$category %in% SPATIAL_CATS
  plot(ecdf(sub$p_lr[sub$is_spatial]), col = "#7570b3", lwd = 2, main = title_lab,
       xlab = "p (LR test, rank_INT scale)", ylab = "Cumulative fraction of features", do.points = FALSE)
  plot(ecdf(sub$p_lr[!sub$is_spatial]), col = "#666666", lwd = 2, add = TRUE, do.points = FALSE)
  abline(0, 1, lty = 3, col = "grey80")
  legend("bottomright",
         legend = c(paste0("Spatial (n=", enr_row$n_spatial, ")"),
                    paste0("Non-spatial (n=", enr_row$n_nonspatial, ")"),
                    paste0("Parametric KS p=", signif(enr_row$ks_p, 3), " [ANTI-CONSERVATIVE, see page 1]"),
                    paste0("Permutation-calibrated KS p=", signif(perm_row$emp_p_ks, 3), " (trustworthy)")),
         col = c("#7570b3", "#666666", NA, NA), lwd = c(2, 2, NA, NA), bty = "n", cex = 0.68)
}
ecdf_panel(cox_luad, enr_luad,perm_luad$summary, "LUAD")
ecdf_panel(cox_lusc, enr_lusc, perm_lusc$summary, "LUSC -- EXPLORATORY (12 events cohort-wide)")
par(op)


## --- Page 3 (round 2, NEW): permutation-null histograms with observed stat ---
op <- par(mfrow = c(2, 2), mar = c(5, 5, 4, 2))
perm_hist <- function(null_vec, obs_val, emp_p, xlab_lab, title_lab) {
  hist(null_vec, breaks = 25, col = "grey85", border = "white",
       main = title_lab, xlab = xlab_lab,
       xlim = range(c(null_vec, obs_val), na.rm = TRUE))
  abline(v = obs_val, col = "blue", lwd = 2)
  legend("topright", legend = c("Permutation null (B=200)", paste0("Observed"),
                                 paste0("Empirical p=", signif(emp_p, 3))),
         col = c("grey85", "blue", NA), lwd = c(6, 2, NA), bty = "n", cex = 0.75)
}
perm_hist(perm_luad$D_null, perm_luad$D_obs, perm_luad$summary$emp_p_ks,
          "KS D (spatial vs non-spatial, LR-test p)", "LUAD: permutation null of KS D (p-value based)")
perm_hist(perm_lusc$D_null, perm_lusc$D_obs, perm_lusc$summary$emp_p_ks,
          "KS D (spatial vs non-spatial, LR-test p)", "LUSC -- EXPLORATORY: permutation null of KS D (p-value based)")
perm_hist(perm_luad$diff_null, perm_luad$diff_obs, perm_luad$summary$emp_p_diff_one_sided_spatial_gt,
          "mean(-log10 p) spatial - non-spatial", "LUAD: permutation null of directional shift (p-value based)")
perm_hist(perm_lusc$diff_null, perm_lusc$diff_obs, perm_lusc$summary$emp_p_diff_one_sided_spatial_gt,
          "mean(-log10 p) spatial - non-spatial", "LUSC -- EXPLORATORY: permutation null of directional shift (p-value based)")
par(op)

dev.off()
```

# permutation hit counts overall
NEW plot (round 5): per-category hit-count permutation null, NO spatial/
non-spatial split (includes image_module) -- directly shows whether the
number of nominal (p<0.05) hits detected in each category is
distinguishable from what this exact permutation null produces by chance
alone. One page per histology, one panel per category.

```{r}
pdf(file.path(PLOTS_DIR, "category_hitrate_permutation.pdf"), width = 13, height = 7)
plot_cat_hits <- function(perm_res, histology_label) {
  op <- par(mfrow = c(2, 4), mar = c(4, 4, 3, 1))
  for (cc in perm_res$categories) {
    null_vec <- perm_res$hits_null[, cc]; obs_val <- perm_res$hits_obs[cc]
    emp_p <- (1 + sum(null_vec >= obs_val, na.rm = TRUE)) / (1 + sum(!is.na(null_vec)))
    perm_hist(null_vec, obs_val, emp_p, "N features p<0.05 (null)", paste0(histology_label, ": ", cc))
  }
  par(op)
}
plot_cat_hits(perm_luad, "LUAD")
plot_cat_hits(perm_lusc, "LUSC -- EXPLORATORY")
dev.off()
```

# compact sig plot
NEW compact plot (round 6): same per-category permutation null as above,
but as one space-efficient panel instead of 16 separate histograms --
boxplot + jittered null draws per category (dodged LUAD/LUSC), hit RATE
(%) on a shared axis (not raw count, since categories differ ~1000x in
size), observed value as a large marker, empirical p printed per box.
biomarker/mutation excluded (both at floor -- 0 observed hits, near-0%
null, in both histologies -- not informative to plot; DS decision).

```{r}
PLOT_CATS <- setdiff(perm_luad$categories, c("biomarker", "mutation"))

rate_null <- function(perm_res, cc) {
  n_tested <- perm_res$category_summary$n_tested[perm_res$category_summary$category == cc]
  100 * perm_res$hits_null[, cc] / n_tested
}
obs_rate <- function(perm_res, cc) {
  perm_res$category_summary$hit_rate_obs_pct[perm_res$category_summary$category == cc]
}
cat_emp_p <- function(perm_res, cc) {
  perm_res$category_summary$emp_p_hits_gt[perm_res$category_summary$category == cc]
}

null_list_luad <- lapply(PLOT_CATS, rate_null, perm_res = perm_luad)
null_list_lusc <- lapply(PLOT_CATS, rate_null, perm_res = perm_lusc)

n_cat <- length(PLOT_CATS)
gap_within  <- 0.8   ## LUAD-LUSC spacing within a category pair
gap_between <- 1.4   ## spacing between category pairs
pos_luad <- seq(0, n_cat - 1) * (gap_within + gap_between)
pos_lusc <- pos_luad + gap_within

all_y <- c(unlist(null_list_luad), unlist(null_list_lusc),
           vapply(PLOT_CATS, obs_rate, numeric(1), perm_res = perm_luad),
           vapply(PLOT_CATS, obs_rate, numeric(1), perm_res = perm_lusc))
ylim <- c(0, max(all_y, na.rm = TRUE) * 1.18)

pdf(file.path(PLOTS_DIR, "category_hitrate_permutation_compact.pdf"), width = 11, height = 6)
op <- par(mar = c(7, 5, 4, 2))
plot(NA, xlim = c(min(pos_luad) - 1, max(pos_lusc) + 1), ylim = ylim,
     xaxt = "n", xlab = "", ylab = "Nominal (p<0.05) hit rate (%)",
     main = "Per-category permutation null vs. observed hit rate\n(no spatial/non-spatial split; B=200)")
abline(h = 5, lty = 2, col = "grey50")
text(min(pos_luad) - 0.9, ylim[2] * 0.97, "- - - 5% (expected by chance)",
     col = "grey40", cex = 0.7, adj = c(0, 1))

box_width <- 0.55
for (i in seq_along(PLOT_CATS)) {
  boxplot(null_list_luad[[i]], at = pos_luad[i], add = TRUE, boxwex = box_width, axes = FALSE,
          col = adjustcolor("blue", alpha.f = 0.25), border = "blue", outline = FALSE)
  boxplot(null_list_lusc[[i]], at = pos_lusc[i], add = TRUE, boxwex = box_width, axes = FALSE,
          col = adjustcolor("orange", alpha.f = 0.25), border = "orange", outline = FALSE)
  points(jitter(rep(pos_luad[i], length(null_list_luad[[i]])), amount = box_width / 2.5),
         null_list_luad[[i]], pch = 16, cex = 0.25, col = adjustcolor("blue", alpha.f = 0.35))
  points(jitter(rep(pos_lusc[i], length(null_list_lusc[[i]])), amount = box_width / 2.5),
         null_list_lusc[[i]], pch = 16, cex = 0.25, col = adjustcolor("orange", alpha.f = 0.35))

  o_luad <- obs_rate(perm_luad, PLOT_CATS[i]); p_luad <- cat_emp_p(perm_luad, PLOT_CATS[i])
  o_lusc <- obs_rate(perm_lusc, PLOT_CATS[i]); p_lusc <- cat_emp_p(perm_lusc, PLOT_CATS[i])
  points(pos_luad[i], o_luad, pch = 23, cex = 1.6, bg = "blue", col = "black", lwd = 1.2)
  points(pos_lusc[i], o_lusc, pch = 23, cex = 1.6, bg = "orange", col = "black", lwd = 1.2)
  text(pos_luad[i], o_luad, labels = sprintf("p=%.2f", p_luad), pos = 3, cex = 0.6, font = 3)
  text(pos_lusc[i], o_lusc, labels = sprintf("p=%.2f", p_lusc), pos = 3, cex = 0.6, font = 3)
}

mid <- (pos_luad + pos_lusc) / 2
axis(1, at = mid, labels = PLOT_CATS, las = 2, cex.axis = 0.85, tick = FALSE)
legend("topright", legend = c("LUAD", "LUSC (exploratory)"), pch = 23,
       pt.bg = c("blue", "orange"), col = "black", bty = "n", cex = 0.85)
par(op)
dev.off()
```

# Forest plot

CAVEAT (round 2): these "top" features are frequently near-duplicates of
each other (e.g. LUAD top spatial hits are dominated by Neutrophil-partner
interaction terms that pairwise correlate rho up to 0.97, and are themselves
strongly correlated with the non-spatial Neutrophil_density feature -- see
results/spatial_density_confounding_check.csv and docs/decisions_round2.md).
Read this panel as "the most significant redundant cluster of features",
NOT as 8-15 independent findings; the plot annotates this directly below.

```{r}

top_spatial <- function(res, n_top = 15) {
  sub <- res[res$scale == "rank_INT" & res$category %in% SPATIAL_CATS & !is.na(res$p_lr), ]
  sub <- sub[order(sub$p_lr), ]
  head(sub, n_top)
}
top_luad <- top_spatial(cox_luad, 30)
top_lusc <- top_spatial(cox_lusc, 30)
log_msg("Top spatial features LUAD (n=", nrow(top_luad), "), LUSC (n=", nrow(top_lusc), ")")

forest_panel <- function(top_df, title_lab) {
  n <- nrow(top_df)
  if (n == 0) { plot.new(); title(title_lab); text(0.5, 0.5, "No spatial features passed filters"); return(invisible(NULL)) }
  ## work on log(HR) scale; cap infinite/extreme CI bounds for display, flag capped ones
  log_hr <- log(top_df$hr)
  log_lo <- log(top_df$lo)
  log_hi <- log(top_df$hi)
  cap <- 4  # cap displayed |log HR| bound at this value (about 55-fold), flag with arrow
  capped_lo <- log_lo < -cap
  capped_hi <- log_hi > cap
  log_lo[capped_lo] <- -cap
  log_hi[capped_hi] <- cap

  ord <- rev(seq_len(n))  # most significant at top
  ylim <- c(0.5, n + 0.5)
  xlim <- range(c(log_lo, log_hi, log_hr), na.rm = TRUE)
  xlim <- xlim + c(-0.3, 0.3)

  plot(NA, xlim = xlim, ylim = ylim, yaxt = "n", ylab = "",
       xlab = "log(HR) per 1 SD (rank-based inverse-normal scale)", main = title_lab)
  axis(2, at = ord, labels = top_df$name, las = 2, cex.axis = 0.55)
  abline(v = 0, lty = 2, col = "grey60")
  segments(log_lo, ord, log_hi, ord, col = "grey30")
  points(log_hr, ord, pch = 16, col = ifelse(log_hr > 0, "red", "blue"), cex = 0.9)
  if (any(capped_lo)) arrows(log_lo[capped_lo] + 0.05, ord[capped_lo], log_lo[capped_lo], ord[capped_lo],
                              length = 0.05, col = "grey30")
  if (any(capped_hi)) arrows(log_hi[capped_hi] - 0.05, ord[capped_hi], log_hi[capped_hi], ord[capped_hi],
                              length = 0.05, col = "grey30")
  text(xlim[2] - 0.1, ord, labels = paste0("p=", signif(top_df$p_lr, 2)), cex = 0.55, pos = 4, xpd = NA)
}

pdf(file.path(PLOTS_DIR, "forest_top_spatial_features.pdf"), width = 12, height = 8.5)
op <- par(mfrow = c(1, 2), mar = c(5, 12, 4, 6), oma = c(3, 0, 0, 0))
forest_panel(top_luad, paste0("LUAD: top spatial features by LR p-value\n(n=", nrow(top_luad),
                               " shown, HR per 1 SD on rank-based scale)"))
forest_panel(top_lusc, paste0("LUSC -- EXPLORATORY (12 events cohort-wide)\ntop spatial features by LR p-value (n=",
                               nrow(top_lusc), ")"))
mtext(paste0("CAVEAT: these features are highly inter-correlated (not independent findings) -- e.g. LUAD Neutrophil-partner ",
             "interaction terms correlate rho=0.83-0.97 with each other and with Neutrophil_density__IF1. ",
             "See results/spatial_density_confounding_check.csv and docs/decisions_round2.md."),
      side = 1, outer = TRUE, line = 1, cex = 0.7, col = "grey30")
par(op)
dev.off()
```

# Top features overall
## LUAD
```{r}
# colors
col_vec <- c(pals::alphabet(n = 8))
names(col_vec) <- c("IMC1", "IMC2", "IF1", "IF2", "IF3", "RNA", "cello1", "cello2")
library(tidyverse)
cur_luad <- cox_luad %>%
  filter(scale == "rank_INT") %>%
   arrange(p_lr) %>%
  #filter(p_lr < 0.05) %>%
  slice_head(n = 30) %>%
  mutate(modality = str_split_i(string = name, pattern = "__", 2),
         modality = str_split_i(string = modality, pattern = "_", 1),
         feature = str_split_i(string = name, pattern = "__", 1))

cur_ord <- cur_luad %>%
  group_by(category) %>%
  arrange(desc(hr), .by_group = TRUE) %>%
  pull(feature)

p_luad <- cur_luad %>%
  ggplot(aes(x = hr, y = factor(feature, cur_ord), col = modality)) +
  geom_point(size = 2) +
  geom_errorbar(aes(xmin = lo, xmax = hi), lwd = 1, width = 0.3) +
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1)) +
  scale_x_log10(limits = c(-0.2, 6.5))+
  theme_bw() +
  scale_color_manual(values = col_vec)+
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_text(aes(label = paste0("p = ", round(p_lr, 3))),
            x = Inf, hjust = 1.05, col = "black", size = 3)+
  facet_wrap(.~category, scales = "free_y", space = "free_y",strip.position = "top") +

  theme(
    strip.text       = element_text(size = 10, face = "bold", margin = margin(3, 3, 3, 3)),
    strip.background = element_rect(fill = "grey90", color = NA),
    panel.spacing    = unit(0.8, "lines"))+
  labs(y = "Feature", x = "HR")

p_luad
```

## LUSC
```{r}
library(tidyverse)
cur_lusc <- cox_lusc %>%
  filter(scale == "rank_INT") %>%
   arrange(p_lr) %>%
  #filter(p_lr < 0.05) %>%
  slice_head(n = 30) %>%
  mutate(modality = str_split_i(string = name, pattern = "__", 2),
         modality = str_split_i(string = modality, pattern = "_", 1),
         feature = str_split_i(string = name, pattern = "__", 1))

cur_ord <- cur_lusc %>%
  group_by(category) %>%
  arrange(desc(hr), .by_group = TRUE) %>%
  pull(feature)

p_lusc <- cur_lusc %>%
  ggplot(aes(x = hr, y = factor(feature, cur_ord), col = modality)) +
  geom_point(size = 2) +
  geom_errorbar(aes(xmin = lo, xmax = hi), lwd = 1, width = 0.3) +
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1)) +
  scale_x_log10(limits = c(-0.2, 28))+
  scale_color_manual(values = col_vec)+
  theme_bw() +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_text(aes(label = paste0("p = ", round(p_lr, 3))),
            x = Inf, hjust = 1.05, col = "black", size = 3)+
  facet_wrap(.~category, scales = "free_y", space = "free_y",strip.position = "top") +

  theme(
    strip.text       = element_text(size = 10, face = "bold", margin = margin(3, 3, 3, 3)),
    strip.background = element_rect(fill = "grey90", color = NA),
    panel.spacing    = unit(0.8, "lines"))+
  labs(y = "Feature", x = "HR")

p_lusc
```

# 5. Bootstrap stability of top spatial features
Stratified bootstrap (by event status) with replacement; refit rank-INT
transform + coxph each replicate; track nominal significance (Wald p<0.05)
and same direction of effect vs the original fit.
CAVEAT (round 2): the "top 8" features bootstrapped per histology are drawn
from the same redundant/correlated set flagged above -- high stability here
reflects that a shared underlying signal (e.g. Neutrophil abundance) is
stable, not that 8 independent spatial associations are each individually
confirmed. See docs/decisions_round2.md.

```{r}
inv_norm <- function(x) {
  n <- sum(!is.na(x)); if (n < 3) return(rep(NA_real_, length(x)))
  qnorm((rank(x, na.last = "keep", ties.method = "average") - 0.5) / n)
}

stratified_boot_idx <- function(status) {
  idx <- seq_along(status)
  idx0 <- idx[status == 0]; idx1 <- idx[status == 1]
  c(sample(idx0, length(idx0), replace = TRUE), sample(idx1, length(idx1), replace = TRUE))
}

bootstrap_feature <- function(nm, d_base, orig_coef, B = 200) {
  d <- d_base[!is.na(d_base[[nm]]), c("time_to_event", "status", nm)]
  colnames(d)[3] <- "x_raw"
  n_sig <- 0L; n_same_dir <- 0L; n_estimable <- 0L
  for (b in seq_len(B)) {
    idx <- stratified_boot_idx(d$status)
    db <- d[idx, ]
    if (length(unique(db$x_raw)) < 3) next
    db$x_rank <- inv_norm(db$x_raw)
    fit <- try(coxph(Surv(time_to_event, status) ~ x_rank, data = db, ties = "efron"), silent = TRUE)
    if (inherits(fit, "try-error") || !is.finite(stats::coef(fit))) next
    n_estimable <- n_estimable + 1L
    cf <- summary(fit)$coefficients
    pw <- cf["x_rank", "Pr(>|z|)"]
    co <- cf["x_rank", "coef"]
    if (!is.na(pw) && pw < 0.05) n_sig <- n_sig + 1L
    if (sign(co) == sign(orig_coef)) n_same_dir <- n_same_dir + 1L
  }
  data.frame(name = nm, B = B, n_estimable = n_estimable,
             pct_significant = 100 * n_sig / B,
             pct_same_direction = 100 * n_same_dir / n_estimable,
             stringsAsFactors = FALSE)
}

run_bootstrap <- function(top_df, full_wide, histology_str, n_features = 8, B = 200) {
  d_base <- full_wide[full_wide$simple_histology == histology_str, ]
  d_base <- d_base[!is.na(d_base$time_to_event) & d_base$time_to_event > 0 & !is.na(d_base$status), ]
  feats <- head(top_df$name, n_features)
  orig_coefs <- top_df$coef[match(feats, top_df$name)]
  res <- do.call(rbind, Map(function(nm, cf) bootstrap_feature(nm, d_base, cf, B = B), feats, orig_coefs))
  res
}

log_msg("Running bootstrap stability for top LUAD spatial features ...")
set.seed(12544)
boot_luad <- run_bootstrap(top_luad, full_wide, "Adenocarcinoma", n_features = 8, B = 200)
log_msg("Running bootstrap stability for top LUSC spatial features (EXPLORATORY) ...")
set.seed(12544)
boot_lusc <- run_bootstrap(top_lusc, full_wide, "Squamous cell carcinoma", n_features = 8, B = 200)

boot_luad$histology <- "LUAD"; boot_lusc$histology <- "LUSC"
boot_all <- rbind(boot_luad, boot_lusc)
saveRDS(boot_all, file.path(RESULTS_DIR, "bootstrap_stability_top_spatial.rds"))
write.csv(boot_all, file.path(RESULTS_DIR, "bootstrap_stability_top_spatial.csv"), row.names = FALSE)
log_msg("Bootstrap stability results:")
print(boot_all)

pdf(file.path(PLOTS_DIR, "bootstrap_stability_top_spatial.pdf"), width = 12, height = 8.5)
op <- par(mfrow = c(1, 2), mar = c(10, 5, 4, 2), oma = c(3, 0, 0, 0))
stability_panel <- function(boot_df, title_lab) {
  if (nrow(boot_df) == 0) { plot.new(); title(title_lab); return(invisible(NULL)) }
  ord <- order(-boot_df$pct_significant)
  bd <- boot_df[ord, ]
  m <- rbind(bd$pct_significant, bd$pct_same_direction)
  colnames(m) <- bd$name
  bp <- barplot(m, beside = TRUE, las = 2, col = c("#1b9e77", "#7570b3"),
                ylim = c(0, 100), ylab = "% of bootstrap replicates",
                main = title_lab, cex.names = 0.6)
  legend("topright", legend = c("% nominally significant (p<0.05)", "% same direction as original"),
         fill = c("#1b9e77", "#7570b3"), bty = "n", cex = 0.7)
  abline(h = 50, lty = 3, col = "grey60")
}
stability_panel(boot_luad, paste0("LUAD: bootstrap stability of top spatial features\n(B=200, stratified resampling, n=",
                                    sum(full_wide$simple_histology == "Adenocarcinoma", na.rm = TRUE), " patients, 22 events)"))
stability_panel(boot_lusc, paste0("LUSC -- EXPLORATORY: bootstrap stability of top spatial features\n(B=200, only 12 events cohort-wide -- expect low stability)"))
mtext(paste0("CAVEAT: top features are highly inter-correlated (e.g. share the same partner cell type / underlying abundance signal) -- ",
             "high stability here reflects a shared signal being stable, not 8 independent confirmations. See docs/decisions_round2.md."),
      side = 1, outer = TRUE, line = 1, cex = 0.7, col = "grey30")
par(op)
dev.off()

log_msg("Saved plots: enrichment_pvalue_ecdf_luad_lusc.pdf, forest_top_spatial_features.pdf, bootstrap_stability_top_spatial.pdf")
log_msg("Round 2 new/updated outputs: results/spatial_density_confounding_check.{csv,rds}, ",
        "results/spatial_enrichment_permutation.{csv,rds}, plus caveat pages/annotations added to all three plots above.")
log_msg("DONE 05_spatial_vs_nonspatial.R")

```

# 6. Bootstrap stability of top features
Stratified bootstrap (by event status) with replacement; refit rank-INT
transform + coxph each replicate; track nominal significance (Wald p<0.05)
and same direction of effect vs the original fit.
CAVEAT (round 2): the "top 8" features bootstrapped per histology are drawn
from the same redundant/correlated set flagged above -- high stability here
reflects that a shared underlying signal (e.g. Neutrophil abundance) is
stable, not that 8 independent spatial associations are each individually
confirmed. See docs/decisions_round2.md.

```{r}
feat_num <- 20
top_overall <- function(res, n_top = 15) {
  sub <- res[res$scale == "rank_INT" & !is.na(res$p_lr), ]
  sub <- sub[order(sub$p_lr), ]
  head(sub, n_top)
}
top_luad <- top_overall(cox_luad, feat_num)
top_lusc <- top_overall(cox_lusc, feat_num)

inv_norm <- function(x) {
  n <- sum(!is.na(x)); if (n < 3) return(rep(NA_real_, length(x)))
  qnorm((rank(x, na.last = "keep", ties.method = "average") - 0.5) / n)
}

stratified_boot_idx <- function(status) {
  idx <- seq_along(status)
  idx0 <- idx[status == 0]; idx1 <- idx[status == 1]
  c(sample(idx0, length(idx0), replace = TRUE), sample(idx1, length(idx1), replace = TRUE))
}

bootstrap_feature <- function(nm, d_base, orig_coef, B = 200) {
  d <- d_base[!is.na(d_base[[nm]]), c("time_to_event", "status", nm)]
  colnames(d)[3] <- "x_raw"
  n_sig <- 0L; n_same_dir <- 0L; n_estimable <- 0L
  for (b in seq_len(B)) {
    idx <- stratified_boot_idx(d$status)
    db <- d[idx, ]
    if (length(unique(db$x_raw)) < 3) next
    db$x_rank <- inv_norm(db$x_raw)
    fit <- try(coxph(Surv(time_to_event, status) ~ x_rank, data = db, ties = "efron"), silent = TRUE)
    if (inherits(fit, "try-error") || !is.finite(stats::coef(fit))) next
    n_estimable <- n_estimable + 1L
    cf <- summary(fit)$coefficients
    pw <- cf["x_rank", "Pr(>|z|)"]
    co <- cf["x_rank", "coef"]
    if (!is.na(pw) && pw < 0.05) n_sig <- n_sig + 1L
    if (sign(co) == sign(orig_coef)) n_same_dir <- n_same_dir + 1L
  }
  data.frame(name = nm, B = B, n_estimable = n_estimable,
             pct_significant = 100 * n_sig / B,
             pct_same_direction = 100 * n_same_dir / n_estimable,
             stringsAsFactors = FALSE)
}

run_bootstrap <- function(top_df, full_wide, histology_str, n_features = 8, B = 200) {
  d_base <- full_wide[full_wide$simple_histology == histology_str, ]
  d_base <- d_base[!is.na(d_base$time_to_event) & d_base$time_to_event > 0 & !is.na(d_base$status), ]
  feats <- head(top_df$name, n_features)
  orig_coefs <- top_df$coef[match(feats, top_df$name)]
  res <- do.call(rbind, Map(function(nm, cf) bootstrap_feature(nm, d_base, cf, B = B), feats, orig_coefs))
  res
}

log_msg("Running bootstrap stability for top LUAD spatial features ...")
set.seed(12215)
boot_luad <- run_bootstrap(top_luad, full_wide, "Adenocarcinoma", n_features = feat_num, B = 200)
log_msg("Running bootstrap stability for top LUSC spatial features (EXPLORATORY) ...")
set.seed(12215)
boot_lusc <- run_bootstrap(top_lusc, full_wide, "Squamous cell carcinoma", n_features = feat_num, B = 200)

boot_luad$histology <- "LUAD"; boot_lusc$histology <- "LUSC"
boot_all <- rbind(boot_luad, boot_lusc)
saveRDS(boot_all, file.path(RESULTS_DIR, "bootstrap_stability_top_features.rds"))
write.csv(boot_all, file.path(RESULTS_DIR, "bootstrap_stability_top_features.csv"), row.names = FALSE)
log_msg("Bootstrap stability results:")
print(boot_all)

pdf(file.path(PLOTS_DIR, "bootstrap_stability_top_features.pdf"), width = 12, height = 8.5)
op <- par(mfrow = c(1, 2), mar = c(10, 5, 4, 2), oma = c(3, 0, 0, 0))
stability_panel <- function(boot_df, title_lab) {
  if (nrow(boot_df) == 0) { plot.new(); title(title_lab); return(invisible(NULL)) }
  ord <- order(-boot_df$pct_significant)
  bd <- boot_df[ord, ]
  m <- rbind(bd$pct_significant, bd$pct_same_direction)
  colnames(m) <- bd$name
  bp <- barplot(m, beside = TRUE, las = 2, col = c("#1b9e77", "#7570b3"),
                ylim = c(0, 100), ylab = "% of bootstrap replicates",
                main = title_lab, cex.names = 0.6)
  legend("topright", legend = c("% nominally significant (p<0.05)", "% same direction as original"),
         fill = c("#1b9e77", "#7570b3"), bty = "n", cex = 0.7)
  abline(h = 50, lty = 3, col = "grey60")
}
stability_panel(boot_luad, paste0("LUAD: bootstrap stability of top spatial features\n(B=200, stratified resampling, n=",
                                    sum(full_wide$simple_histology == "Adenocarcinoma", na.rm = TRUE), " patients, 22 events)"))
stability_panel(boot_lusc, paste0("LUSC -- EXPLORATORY: bootstrap stability of top spatial features\n(B=200, only 12 events cohort-wide -- expect low stability)"))
mtext(paste0("CAVEAT: top features are highly inter-correlated (e.g. share the same partner cell type / underlying abundance signal) -- ",
             "high stability here reflects a shared signal being stable, not 8 independent confirmations. See docs/decisions_round2.md."),
      side = 1, outer = TRUE, line = 1, cex = 0.7, col = "grey30")
par(op)
dev.off()

log_msg("Saved plots: enrichment_pvalue_ecdf_luad_lusc.pdf, forest_top_features_features.pdf, bootstrap_stability_top_features.pdf")
log_msg("Round 2 new/updated outputs: results/spatial_density_confounding_check.{csv,rds}, ",
        "results/spatial_enrichment_permutation.{csv,rds}, plus caveat pages/annotations added to all three plots above.")
log_msg("DONE 05_spatial_vs_nonspatial.R")

```
