---
title: "Venn LUAD LUSC"
author: "Daniel Schulz"
date: "2026-09-15"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

06_venn_luad_lusc_overlap.R
Per-category (modality) overlap of nominally significant univariate Cox
hits between LUAD and LUSC. Hit definition matches the pooled overlap
already computed in 04_luad_lusc_survival.R (Plot 4): rank_INT scale,
likelihood-ratio p_lr < 0.05 -- here broken out per feature category
instead of pooled across all categories.

For shared hits (nominally significant in BOTH histologies), we also
report whether the direction of association agrees: sign(coef) on the
rank_INT scale is positive when a higher feature value is associated with
WORSE survival (higher hazard) and negative when associated with better
survival, so "same direction" (concordant) vs "opposite direction"
(discordant) is directly comparable across features and histologies.
IMPORTANT CAVEATS carried over from 04_luad_lusc_survival.R (also printed
on each panel):
- Nothing here survives FDR correction (see cox_luad.rds/cox_lusc.rds
  fdr_within_category/fdr_global); this is a nominal-p, hypothesis-
  generating comparison, consistent with how the rest of the project
  frames these screens.
- LUAD and LUSC are screened at different MIN_EVENTS/MIN_PREV thresholds,
  so the *tested* feature set per category can already differ between
  histologies before any hit-calling happens -- n_tested is reported per
  panel so this isn't silently folded into the overlap counts.
- LUSC remains EXPLORATORY throughout (12 events cohort-wide).
Base R only -- no extra plotting packages; the Venn circles are drawn by
hand with polygon(). NOT area-proportional: circle size is fixed and
identical in every panel, only the printed counts carry the quantitative
information. Must run on R 4.2+.

```{r}


RESULTS_DIR <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/survival/Rout/"
PLOTS_DIR   <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/survival/plots/"
if (!dir.exists(PLOTS_DIR)) dir.create(PLOTS_DIR, recursive = TRUE)

log_msg <- function(...) cat(format(Sys.time(), "%H:%M:%S"), "-", ..., "\n")

cox_luad <- readRDS(file.path(RESULTS_DIR, "cox_luad.rds"))
cox_lusc <- readRDS(file.path(RESULTS_DIR, "cox_lusc.rds"))

P_THRESH <- 0.05  ## nominal rank_INT / LR-test p; nothing survives FDR (see docs) --
                  ## deliberately framed as nominal/exploratory throughout.

```


# Per-category tested sets, hit sets, and (for shared hits) direction
## concordance, LUAD vs LUSC

```{r}
summarize_category <- function(cc) {
  luad <- cox_luad[cox_luad$category == cc & cox_luad$scale == "rank_INT", ]
  lusc <- cox_lusc[cox_lusc$category == cc & cox_lusc$scale == "rank_INT", ]

  tested_luad <- luad$name[!is.na(luad$p_lr)]
  tested_lusc <- lusc$name[!is.na(lusc$p_lr)]

  hits_luad <- luad$name[!is.na(luad$p_lr) & luad$p_lr < P_THRESH]
  hits_lusc <- lusc$name[!is.na(lusc$p_lr) & lusc$p_lr < P_THRESH]

  shared    <- intersect(hits_luad, hits_lusc)
  luad_only <- setdiff(hits_luad, hits_lusc)
  lusc_only <- setdiff(hits_lusc, hits_luad)

  dir_luad <- setNames(sign(luad$coef), luad$name)
  dir_lusc <- setNames(sign(lusc$coef), lusc$name)
  concordant <- shared[dir_luad[shared] == dir_lusc[shared]]
  discordant <- shared[dir_luad[shared] != dir_lusc[shared]]

  list(category = cc,
       n_tested_luad = length(tested_luad), n_tested_lusc = length(tested_lusc),
       n_hits_luad = length(hits_luad), n_hits_lusc = length(hits_lusc),
       luad_only = luad_only, lusc_only = lusc_only,
       shared = shared, concordant = concordant, discordant = discordant)
}

cats <- sort(unique(c(cox_luad$category, cox_lusc$category)))
overlap_list <- lapply(cats, summarize_category)
names(overlap_list) <- cats

## flat table for saving / manuscript text (one row per category)
overlap_tab <- do.call(rbind, lapply(overlap_list, function(o) {
  data.frame(category = o$category,
             n_tested_luad = o$n_tested_luad, n_tested_lusc = o$n_tested_lusc,
             n_hits_luad = o$n_hits_luad, n_hits_lusc = o$n_hits_lusc,
             n_luad_only = length(o$luad_only), n_lusc_only = length(o$lusc_only),
             n_shared = length(o$shared), n_concordant = length(o$concordant),
             n_discordant = length(o$discordant), stringsAsFactors = FALSE)
}))
rownames(overlap_tab) <- NULL
log_msg("Per-category LUAD/LUSC nominal-hit overlap (rank_INT, p_lr<", P_THRESH, "):")
print(overlap_tab)

saveRDS(overlap_list, file.path(RESULTS_DIR, "venn_overlap_by_category.rds"))
write.csv(overlap_tab, file.path(RESULTS_DIR, "venn_overlap_by_category.csv"), row.names = FALSE)
log_msg("Saved results/venn_overlap_by_category.rds, results/venn_overlap_by_category.csv")
```

## Hand-drawn 2-circle schematic Venn, one panel per category.
LUAD/LUSC colors match qq_pvalues_luad_lusc.pdf and
hitcounts_by_category_histology.pdf ("blue" / "orange").
```{r}
draw_circle <- function(cx, cy, r, col, alpha_fill = 0.15, n = 200) {
  th <- seq(0, 2 * pi, length.out = n)
  polygon(cx + r * cos(th), cy + r * sin(th),
          border = col, col = adjustcolor(col, alpha.f = alpha_fill), lwd = 2)
}

venn_panel <- function(o, col_luad = "blue", col_lusc = "orange") {
  r <- 1; offset <- 0.75
  plot(NA, xlim = c(-2.2, 2.2), ylim = c(-1.6, 1.9), asp = 1, axes = FALSE,
       xlab = "", ylab = "", main = o$category, cex.main = 1.1)
  draw_circle(-offset, 0, r, col_luad)
  draw_circle( offset, 0, r, col_lusc)

  n_luad_only <- length(o$luad_only); n_lusc_only <- length(o$lusc_only)
  n_shared    <- length(o$shared)
  dir_lab <- if (n_shared > 0)
    paste0(n_shared, "\n(", length(o$concordant), " same dir,\n",
           length(o$discordant), " opp. dir)")
  else "0"

  text(-offset * 1.55, 0, n_luad_only, cex = 1.2, font = 2)
  text( offset * 1.55, 0, n_lusc_only, cex = 1.2, font = 2)
  text(0, 0, dir_lab, cex = 0.8)

  text(-offset, 1.3, "LUAD", col = col_luad, font = 2, cex = 1.0)
  text( offset, 1.3, "LUSC", col = col_lusc, font = 2, cex = 1.0)
  mtext(paste0("tested: LUAD n=", o$n_tested_luad, ", LUSC n=", o$n_tested_lusc,
               " | hits (p<", P_THRESH, "): LUAD=", o$n_hits_luad, ", LUSC=", o$n_hits_lusc),
        side = 1, line = 0.8, cex = 0.6)
}

n_col <- ceiling(length(cats) / 2)
pdf(file.path(PLOTS_DIR, "venn_luad_lusc_by_category.pdf"), width = 3.2 * n_col, height = 7.5)
op <- par(mfrow = c(2, n_col), mar = c(3, 1, 3, 1), oma = c(0, 0, 4, 0))
for (cc in cats) venn_panel(overlap_list[[cc]])
mtext(paste0("Overlap of nominally significant (p_lr<", P_THRESH, ", rank_INT) univariate Cox hits, ",
             "LUAD vs LUSC, by feature category\n",
             "Schematic Venn (NOT area-proportional) -- LUSC is EXPLORATORY (12 events cohort-wide); ",
             "nothing here survives FDR correction; tested feature sets differ by histology ",
             "(see MIN_EVENTS/MIN_PREV in 04_luad_lusc_survival.R)"),
      side = 3, outer = TRUE, line = 0.5, cex = 0.75)
par(op)
dev.off()

log_msg("Saved plots/venn_luad_lusc_by_category.pdf")
log_msg("DONE 06_venn_luad_lusc_overlap.R")

```

