---
title: "correlated features survival"
author: "Daniel Schulz"
date: "2026-09-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



Post-hoc redundancy pruning of the nominal univariate Cox hits (rank_INT
scale, p_lr < 0.05) in cox_luad.rds / cox_lusc.rds, FOR REPORTING PURPOSES
ONLY -- this does not change any p-value, FDR result, or model; it just
avoids listing multiple near-duplicate rows (e.g. an interaction term and
the density feature of the same cell type -- the exact confound quantified
in 05_spatial_vs_nonspatial.R/docs/decisions_round2.md) in a hit table or
figure when they carry the same signal.

 - Hits are pooled ACROSS ALL feature categories together per histology
   (not clustered within each category separately), specifically because
   the known confound is cross-category (interaction <-> density).
 - Pairwise Spearman correlation among that histology's hit features,
   computed on full_wide (pairwise-complete observations, >=10 required or
   the pair is treated as uncorrelated) -- same convention already used in
   05_spatial_vs_nonspatial.R's density-confounding check. Spearman is
   invariant to the rank-based inverse-normal transform used for the Cox
   fits, so correlating the raw full_wide columns gives identical rho to
   correlating the rank_INT-transformed values.
 - Features are grouped by connected components of the graph "|rho| > 0.8"
   (transitive: if A-B and B-C both exceed 0.8 but A-C does not, all three
   still collapse into one group -- standard for this kind of redundancy
   pruning, but worth stating explicitly since it is not the only possible
   convention).
 - Within each group, the representative is the feature with the smallest
   p_lr (matches the project's existing "rank hits by ascending LR p"
   convention from docs/decisions_round2.md's top-8 lists).
 - Both the full per-feature mapping (which group / representative each
   hit was assigned to) and the pruned representatives-only list are
   saved, so nothing is silently dropped from the record.

# set up
```{r}
RESULTS_DIR <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/survival/Rout/"
PLOTS_DIR <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/survival/plots/"
log_msg <- function(...) cat(format(Sys.time(), "%H:%M:%S"), "-", ..., "\n")

cox_luad  <- readRDS(file.path(RESULTS_DIR, "cox_luad.rds"))
cox_lusc  <- readRDS(file.path(RESULTS_DIR, "cox_lusc.rds"))
full_wide <- readRDS(file.path(RESULTS_DIR, "full_wide.rds"))

P_THRESH   <- 0.05  ## nominal rank_INT / LR-test p, matching the hit definition used elsewhere
R_THRESH   <- 0.8   ## |Spearman rho| above which two hit features are considered redundant
MIN_PAIR_N <- 10    ## minimum pairwise-complete N to trust a correlation estimate at all

```

# connected components
Connected components of an adjacency matrix (simple BFS, base R only --
no igraph dependency for what is a small graph, at most a few hundred
nodes per histology).

```{r}

connected_components <- function(adj) {
  n <- nrow(adj)
  comp <- rep(NA_integer_, n)
  current <- 0L
  for (i in seq_len(n)) {
    if (!is.na(comp[i])) next
    current <- current + 1L
    queue <- i
    comp[i] <- current
    while (length(queue)) {
      j <- queue[1]; queue <- queue[-1]
      nb <- which(adj[j, ] & is.na(comp))
      if (length(nb)) { comp[nb] <- current; queue <- c(queue, nb) }
    }
  }
  comp
}


# Per-histology dedup: pull nominal hits, correlate, cluster, pick
# representatives.

dedup_histology <- function(cox_res, full_wide, histology_str, histology_label) {
  hits <- cox_res[cox_res$scale == "rank_INT" & !is.na(cox_res$p_lr) & cox_res$p_lr < P_THRESH,
                   c("name", "category", "p_lr", "coef", "hr")]
  hits <- hits[order(hits$p_lr), ]
  log_msg(histology_label, ": ", nrow(hits), " nominal hits (p_lr<", P_THRESH, ") pooled across categories")

  if (nrow(hits) <= 1) {
    hits$cluster_id <- seq_len(nrow(hits))
    hits$cluster_size <- 1L
    hits$is_representative <- TRUE
    hits$representative_name <- hits$name
    hits$histology <- histology_label
    return(hits)
  }

  d <- full_wide[full_wide$simple_histology == histology_str, hits$name, drop = FALSE]
  mat <- as.matrix(sapply(d, as.numeric))

  cor_mat <- suppressWarnings(cor(mat, method = "spearman", use = "pairwise.complete.obs"))
  not_na  <- !is.na(mat)
  n_mat   <- crossprod(not_na, not_na)

  adj <- (abs(cor_mat) > R_THRESH) & (n_mat >= MIN_PAIR_N)
  adj[is.na(adj)] <- FALSE
  diag(adj) <- FALSE

  comp <- connected_components(adj)
  hits$cluster_id <- comp[match(hits$name, colnames(mat))]

  cluster_size <- table(hits$cluster_id)
  hits$cluster_size <- as.integer(cluster_size[as.character(hits$cluster_id)])

  ## representative = smallest p_lr within the cluster (ties -> alphabetical, deterministic)
  hits <- hits[order(hits$cluster_id, hits$p_lr, hits$name), ]
  rep_name <- tapply(hits$name, hits$cluster_id, `[`, 1)
  hits$representative_name <- rep_name[as.character(hits$cluster_id)]
  hits$is_representative <- hits$name == hits$representative_name
  hits$histology <- histology_label
  hits
}

dedup_luad <- dedup_histology(cox_luad, full_wide, "Adenocarcinoma", "LUAD")
dedup_lusc <- dedup_histology(cox_lusc, full_wide, "Squamous cell carcinoma", "LUSC")
dedup_all  <- rbind(dedup_luad, dedup_lusc)

n_before <- c(LUAD = nrow(dedup_luad), LUSC = nrow(dedup_lusc))
n_after  <- c(LUAD = sum(dedup_luad$is_representative), LUSC = sum(dedup_lusc$is_representative))
log_msg("Nominal hits before -> after correlation dedup (|rho|>", R_THRESH, "): LUAD ",
        n_before["LUAD"], " -> ", n_after["LUAD"], " | LUSC ", n_before["LUSC"], " -> ", n_after["LUSC"])

## per-category breakdown of the collapse, for the methods text
cat_summary <- aggregate(cbind(n_hits = rep(1, nrow(dedup_all)),
                                n_kept = as.integer(dedup_all$is_representative)) ~
                            histology + category, data = dedup_all, FUN = sum)
cat_summary <- cat_summary[order(cat_summary$histology, cat_summary$category), ]
log_msg("Per-category hit counts, before -> after dedup:")
print(cat_summary)

saveRDS(dedup_all, file.path(RESULTS_DIR, "hit_dedup_full.rds"))
write.csv(dedup_all, file.path(RESULTS_DIR, "hit_dedup_full.csv"), row.names = FALSE)

hit_dedup_representatives <- dedup_all[dedup_all$is_representative,
                                        c("histology", "category", "name", "p_lr", "coef", "hr",
                                          "cluster_id", "cluster_size")]
hit_dedup_representatives <- hit_dedup_representatives[order(hit_dedup_representatives$histology,
                                                               hit_dedup_representatives$p_lr), ]
saveRDS(hit_dedup_representatives, file.path(RESULTS_DIR, "hit_dedup_representatives.rds"))
write.csv(hit_dedup_representatives, file.path(RESULTS_DIR, "hit_dedup_representatives.csv"), row.names = FALSE)
```

# Bootstrap stability of top pruned features (not used in the paper)

Same feature selection as Figure S8B: within each correlation cluster
(per histology), density features are preferred, otherwise the
representative (lowest p). Top 20 per histology, stratified bootstrap
(B = 200). % significant and % same direction are relative to the number
of successfully fitted replicates.

```{r}
library(dplyr)
suppressWarnings(suppressMessages(library(survival)))

N_TOP    <- 20
HIST_STR <- c(LUAD = "Adenocarcinoma", LUSC = "Squamous cell carcinoma")

## same selection as in the figure script (S8B)
select_top <- function(dedup, histology_label, n_top) {
  dedup %>%
    group_by(histology, cluster_id) %>%
    mutate(final_selection = if (any(category == "density")) category == "density" else is_representative) %>%
    ungroup() %>%
    filter(final_selection, histology == histology_label) %>%
    arrange(p_lr) %>%
    slice_head(n = n_top)
}

inv_norm <- function(x) {
  n <- sum(!is.na(x)); if (n < 3) return(rep(NA_real_, length(x)))
  qnorm((rank(x, na.last = "keep", ties.method = "average") - 0.5) / n)
}

stratified_boot_idx <- function(status) {
  idx <- seq_along(status)
  idx0 <- idx[status == 0]; idx1 <- idx[status == 1]
  c(sample(idx0, length(idx0), replace = TRUE), sample(idx1, length(idx1), replace = TRUE))
}

bootstrap_feature <- function(nm, d_base, orig_coef, B = 200) {
  d <- d_base[!is.na(d_base[[nm]]), c("time_to_event", "status", nm)]
  colnames(d)[3] <- "x_raw"
  n_sig <- 0L; n_same_dir <- 0L; n_estimable <- 0L
  for (b in seq_len(B)) {
    db <- d[stratified_boot_idx(d$status), ]
    if (length(unique(db$x_raw)) < 3) next
    db$x_rank <- inv_norm(db$x_raw)
    fit <- try(coxph(Surv(time_to_event, status) ~ x_rank, data = db, ties = "efron"), silent = TRUE)
    if (inherits(fit, "try-error") || !is.finite(stats::coef(fit))) next
    n_estimable <- n_estimable + 1L
    cf <- summary(fit)$coefficients
    if (!is.na(cf["x_rank", "Pr(>|z|)"]) && cf["x_rank", "Pr(>|z|)"] < 0.05) n_sig <- n_sig + 1L
    if (sign(cf["x_rank", "coef"]) == sign(orig_coef)) n_same_dir <- n_same_dir + 1L
  }
  data.frame(name = nm, B = B, n_estimable = n_estimable,
             pct_significant    = 100 * n_sig / n_estimable,
             pct_same_direction = 100 * n_same_dir / n_estimable,
             stringsAsFactors = FALSE)
}

boot_all <- do.call(rbind, lapply(names(HIST_STR), function(h) {
  set.seed(12215)
  top <- select_top(dedup_all, h, N_TOP)
  d_base <- full_wide[full_wide$simple_histology == HIST_STR[[h]], ]
  d_base <- d_base[!is.na(d_base$time_to_event) & d_base$time_to_event > 0 & !is.na(d_base$status), ]
  res <- do.call(rbind, Map(function(nm, cf) bootstrap_feature(nm, d_base, cf), top$name, top$coef))
  res$histology <- h
  res
}))

saveRDS(boot_all, file.path(RESULTS_DIR, "bootstrap_stability_top_features_dedup.rds"))
write.csv(boot_all, file.path(RESULTS_DIR, "bootstrap_stability_top_features_dedup.csv"), row.names = FALSE)
boot_all
```




# 6. Bootstrap stability of top features
Stratified bootstrap (by event status) with replacement; refit rank-INT
transform + coxph each replicate; track nominal significance (Wald p<0.05)
and same direction of effect vs the original fit.
CAVEAT (round 2): the "top 8" features bootstrapped per histology are drawn
from the same redundant/correlated set flagged above -- high stability here
reflects that a shared underlying signal (e.g. Neutrophil abundance) is
stable, not that 8 independent spatial associations are each individually
confirmed. See docs/decisions_round2.md.

```{r}
feat_num <- 10
dat1 <- readRDS(file.path(RESULTS_DIR,"hit_dedup_full.rds"))

cur_names_luad <- dat1 %>%
  group_by(cluster_id) %>%
  mutate(density_filter   = category == "density",
         final_selection  = if (any(density_filter)) density_filter else is_representative) %>%
  ungroup() %>%
  filter(final_selection == TRUE,
         histology == "LUAD") %>%
  arrange(p_lr) %>%
  #filter(p_lr < 0.05) %>%
  slice_head(n = feat_number) %>%
  pull(name)

cur_names_lusc <- dat1 %>%
  group_by(cluster_id) %>%
  mutate(density_filter   = category == "density",
         final_selection  = if (any(density_filter)) density_filter else is_representative) %>%
  ungroup() %>%
  filter(final_selection == TRUE,
         histology == "LUSC") %>%
  arrange(p_lr) %>%
  #filter(p_lr < 0.05) %>%
  slice_head(n = feat_number) %>%
  pull(name)


top_overall <- function(res, n_top = 15, name_sel) {
  sub <- res[res$scale == "rank_INT" & res$name %in% name_sel, ]
  sub <- sub[order(sub$p_lr), ]
  head(sub, n_top)
}


top_luad <- top_overall(cox_luad, feat_num, cur_names_luad)
top_lusc <- top_overall(cox_lusc, feat_num, cur_names_lusc)

inv_norm <- function(x) {
  n <- sum(!is.na(x)); if (n < 3) return(rep(NA_real_, length(x)))
  qnorm((rank(x, na.last = "keep", ties.method = "average") - 0.5) / n)
}

stratified_boot_idx <- function(status) {
  idx <- seq_along(status)
  idx0 <- idx[status == 0]; idx1 <- idx[status == 1]
  c(sample(idx0, length(idx0), replace = TRUE), sample(idx1, length(idx1), replace = TRUE))
}

bootstrap_feature <- function(nm, d_base, orig_coef, B = 200) {
  d <- d_base[!is.na(d_base[[nm]]), c("time_to_event", "status", nm)]
  colnames(d)[3] <- "x_raw"
  n_sig <- 0L; n_same_dir <- 0L; n_estimable <- 0L
  for (b in seq_len(B)) {
    idx <- stratified_boot_idx(d$status)
    db <- d[idx, ]
    if (length(unique(db$x_raw)) < 3) next
    db$x_rank <- inv_norm(db$x_raw)
    fit <- try(coxph(Surv(time_to_event, status) ~ x_rank, data = db, ties = "efron"), silent = TRUE)
    if (inherits(fit, "try-error") || !is.finite(stats::coef(fit))) next
    n_estimable <- n_estimable + 1L
    cf <- summary(fit)$coefficients
    pw <- cf["x_rank", "Pr(>|z|)"]
    co <- cf["x_rank", "coef"]
    if (!is.na(pw) && pw < 0.05) n_sig <- n_sig + 1L
    if (sign(co) == sign(orig_coef)) n_same_dir <- n_same_dir + 1L
  }
  data.frame(name = nm, B = B, n_estimable = n_estimable,
             pct_significant = 100 * n_sig / B,
             pct_same_direction = 100 * n_same_dir / n_estimable,
             stringsAsFactors = FALSE)
}

run_bootstrap <- function(top_df, full_wide, histology_str, n_features = 8, B = 200) {
  d_base <- full_wide[full_wide$simple_histology == histology_str, ]
  d_base <- d_base[!is.na(d_base$time_to_event) & d_base$time_to_event > 0 & !is.na(d_base$status), ]
  feats <- head(top_df$name, n_features)
  orig_coefs <- top_df$coef[match(feats, top_df$name)]
  res <- do.call(rbind, Map(function(nm, cf) bootstrap_feature(nm, d_base, cf, B = B), feats, orig_coefs))
  res
}

log_msg("Running bootstrap stability for top LUAD spatial features ...")
set.seed(12215)
boot_luad <- run_bootstrap(top_luad, full_wide, "Adenocarcinoma", n_features = feat_num, B = 200)
log_msg("Running bootstrap stability for top LUSC spatial features (EXPLORATORY) ...")
set.seed(12215)
boot_lusc <- run_bootstrap(top_lusc, full_wide, "Squamous cell carcinoma", n_features = feat_num, B = 200)

boot_luad$histology <- "LUAD"; boot_lusc$histology <- "LUSC"
boot_all <- rbind(boot_luad, boot_lusc)
saveRDS(boot_all, file.path(RESULTS_DIR, "bootstrap_stability_top_features.rds"))
write.csv(boot_all, file.path(RESULTS_DIR, "bootstrap_stability_top_features.csv"), row.names = FALSE)
log_msg("Bootstrap stability results:")
print(boot_all)

pdf(file.path(PLOTS_DIR, "bootstrap_stability_top_features.pdf"), width = 12, height = 8.5)
op <- par(mfrow = c(1, 2), mar = c(10, 5, 4, 2), oma = c(3, 0, 0, 0))
stability_panel <- function(boot_df, title_lab) {
  if (nrow(boot_df) == 0) { plot.new(); title(title_lab); return(invisible(NULL)) }
  ord <- order(-boot_df$pct_significant)
  bd <- boot_df[ord, ]
  m <- rbind(bd$pct_significant, bd$pct_same_direction)
  colnames(m) <- bd$name
  bp <- barplot(m, beside = TRUE, las = 2, col = c("#1b9e77", "#7570b3"),
                ylim = c(0, 100), ylab = "% of bootstrap replicates",
                main = title_lab, cex.names = 0.6)
  legend("topright", legend = c("% nominally significant (p<0.05)", "% same direction as original"),
         fill = c("#1b9e77", "#7570b3"), bty = "n", cex = 0.7)
  abline(h = 50, lty = 3, col = "grey60")
}
stability_panel(boot_luad, paste0("LUAD: bootstrap stability of top spatial features\n(B=200, stratified resampling, n=",
                                    sum(full_wide$simple_histology == "Adenocarcinoma", na.rm = TRUE), " patients, 22 events)"))
stability_panel(boot_lusc, paste0("LUSC -- EXPLORATORY: bootstrap stability of top spatial features\n(B=200, only 12 events cohort-wide -- expect low stability)"))
mtext(paste0("CAVEAT: top features are highly inter-correlated (e.g. share the same partner cell type / underlying abundance signal) -- ",
             "high stability here reflects a shared signal being stable, not 8 independent confirmations. See docs/decisions_round2.md."),
      side = 1, outer = TRUE, line = 1, cex = 0.7, col = "grey30")
par(op)
dev.off()

log_msg("Saved plots: enrichment_pvalue_ecdf_luad_lusc.pdf, forest_top_features_features.pdf, bootstrap_stability_top_features.pdf")
log_msg("Round 2 new/updated outputs: results/spatial_density_confounding_check.{csv,rds}, ",
        "results/spatial_enrichment_permutation.{csv,rds}, plus caveat pages/annotations added to all three plots above.")
log_msg("DONE 05_spatial_vs_nonspatial.R")

```

# fixed subsamplin 70%
```{r, fig.width=12, fig.height=5}
subsample_idx <- function(status, frac = 0.7) {
  idx <- seq_along(status)
  idx0 <- idx[status == 0]; idx1 <- idx[status == 1]
  c(sample(idx0, round(frac * length(idx0))), sample(idx1, round(frac * length(idx1))))
}

subsample_feature <- function(nm, d_base, orig_coef, frac = 0.7, B = 200) {
  d <- d_base[!is.na(d_base[[nm]]), c("time_to_event", "status", nm)]
  colnames(d)[3] <- "x_raw"
  n_sig <- 0L; n_same_dir <- 0L; n_estimable <- 0L
  for (b in seq_len(B)) {
    idx <- subsample_idx(d$status, frac = frac)
    db <- d[idx, ]
    if (length(unique(db$x_raw)) < 3) next
    db$x_rank <- inv_norm(db$x_raw)
    fit <- try(coxph(Surv(time_to_event, status) ~ x_rank, data = db, ties = "efron"), silent = TRUE)
    if (inherits(fit, "try-error") || !is.finite(stats::coef(fit))) next
    n_estimable <- n_estimable + 1L
    cf <- summary(fit)$coefficients
    pw <- cf["x_rank", "Pr(>|z|)"]
    co <- cf["x_rank", "coef"]
    if (!is.na(pw) && pw < 0.05) n_sig <- n_sig + 1L
    if (sign(co) == sign(orig_coef)) n_same_dir <- n_same_dir + 1L
  }
  data.frame(name = nm, B = B, frac = frac, n_estimable = n_estimable,
             pct_significant = 100 * n_sig / B,
             pct_same_direction = 100 * n_same_dir / n_estimable,
             stringsAsFactors = FALSE)
}

run_subsample <- function(top_df, full_wide, histology_str, n_features = 8, frac = 0.7, B = 200) {
  d_base <- full_wide[full_wide$simple_histology == histology_str, ]
  d_base <- d_base[!is.na(d_base$time_to_event) & d_base$time_to_event > 0 & !is.na(d_base$status), ]
  feats <- head(top_df$name, n_features)
  orig_coefs <- top_df$coef[match(feats, top_df$name)]
  do.call(rbind, Map(function(nm, cf) subsample_feature(nm, d_base, cf, frac = frac, B = B), feats, orig_coefs))
}

log_msg("Running 70% subsampling (no replacement) sanity check for top LUAD spatial features ...")
set.seed(1235)
subsample_luad <- run_subsample(top_luad, full_wide, "Adenocarcinoma", n_features = feat_num, frac = 0.7, B = 200)
log_msg("Running 70% subsampling (no replacement) sanity check for top LUSC spatial features (EXPLORATORY) ...")
set.seed(1235)
subsample_lusc <- run_subsample(top_lusc, full_wide, "Squamous cell carcinoma", n_features = feat_num, frac = 0.7, B = 200)

subsample_luad$histology <- "LUAD"; subsample_lusc$histology <- "LUSC"
subsample_all <- rbind(subsample_luad, subsample_lusc)
saveRDS(subsample_all, file.path(RESULTS_DIR, "subsample70_stability_top_spatial.rds"))
write.csv(subsample_all, file.path(RESULTS_DIR, "subsample70_stability_top_spatial.csv"), row.names = FALSE)
log_msg("70% subsampling (no replacement) stability results:")
print(subsample_all)

pdf(file.path(PLOTS_DIR, "subsample70_stability_top_spatial.pdf"), width = 12, height = 8.5)
op <- par(mfrow = c(1, 2), mar = c(10, 5, 4, 2), oma = c(3, 0, 0, 0))
stability_panel(subsample_luad, paste0("LUAD: 70% subsampling stability of top spatial features\n(B=200, no replacement, n=",
                                        round(0.7 * sum(full_wide$simple_histology == "Adenocarcinoma", na.rm = TRUE)), " patients per replicate)"))
stability_panel(subsample_lusc, "LUSC -- EXPLORATORY: 70% subsampling stability of top spatial features\n(B=200, no replacement -- expect low stability)")
mtext(paste0("Sanity check alongside the with-replacement bootstrap above: does the finding survive training on a fixed, ",
             "duplicate-free 70% of patients? Same inter-correlation caveat applies -- see docs/decisions_round2.md."),
      side = 1, outer = TRUE, line = 1, cex = 0.7, col = "grey30")
par(op)
dev.off()
p1 <- subsample_luad %>%
  ggplot(aes(y = factor(name, levels = subsample_luad %>% arrange(pct_significant) %>% pull(name)), x = histology, fill = pct_significant))+
  geom_tile()+
  geom_text(aes(label= pct_significant) )+
  scale_fill_viridis_c(limits = c(0,100))+
  ylab("Top 20 features")+
  theme_bw()

p2 <- subsample_lusc %>%
  ggplot(aes(y = factor(name, levels = subsample_lusc %>% arrange(pct_significant) %>% pull(name)), x = histology, fill = pct_significant))+
  geom_tile()+  
  geom_text(aes(label= pct_significant) )+
  scale_fill_viridis_c(limits = c(0,100))+
  ylab("Top 20 features")+
  theme_bw()

p <- cowplot::plot_grid(p1, p2, ncol = 2)
p

```

