---
title: "NSCLC"
author: "Elena"
date: "2026-06-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggtext)
library(glue)
library(singscore)
```


```{r}
rna_mat <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/rna_mat.rds")
traits <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/all_traits.csv"))
signGenes <- read.csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/RNASeq/annotations/tSn_sign.csv")
```



```{r}
luad_ids <- traits %>% filter(LUAD_vs_LUSC == "Adenocarcinoma") %>% pull(immucan_id)

luad_ids_present <- luad_ids[luad_ids %in% rownames(rna_mat)]

rna_luad <- rna_mat[luad_ids_present, , drop = FALSE] # the object to proceed to work with

dim(rna_luad)
length(luad_ids)          # 125 all samples
length(luad_ids_present)  # 85 samples in the RNA countmat
length(setdiff(luad_ids, rownames(rna_mat)))  # 40
```



```{r}
expr_luad <- t(rna_luad)

program_cols <- c(
  tS1 = "tS1_sign",
  tS2 = "tS2_sign",
  tS3 = "tS3_sign"
)

# only available genes
program_genes <- lapply(program_cols, function(col) {
  genes <- signGenes[[col]]
  genes <- genes[!is.na(genes) & genes != ""]
  intersect(genes, rownames(expr_luad))
})

sapply(program_genes, length)
```

## sincScore

It ranks all genes within one sample/cell, then asks whether the genes in your signature are mostly near the top of that ranked list.

It scores each sample independently, as a signature activity/enrichment score

```{r}
rankData <- rankGenes(expr_luad)

program_scores <- data.frame(sample = colnames(expr_luad))

for (program in names(program_genes)) {
  scored <- simpleScore(rankData, upSet = program_genes[[program]])
  program_scores[[paste0(program, "_singscore")]] <- scored[
    program_scores$sample,
    "TotalScore"
  ]
}

head(program_scores)
```

# raw score distributions

```{r}
score_cols <- c("tS1_singscore", "tS2_singscore", "tS3_singscore")

score_long <- program_scores %>%
  select(sample, all_of(score_cols)) %>%
  pivot_longer(
    cols = all_of(score_cols),
    names_to = "program",
    values_to = "singscore"
  ) %>%
  mutate(
    program = recode(
      program,
      tS1_singscore = "tS1",
      tS2_singscore = "tS2",
      tS3_singscore = "tS3"
    )
  )

ggplot(score_long, aes(x = singscore, fill = program, color = program)) +
  geom_density(alpha = 0.25, linewidth = 1.1) +
  scale_color_manual(values = c(tS1 = "firebrick3", tS2 = "steelblue", tS3 = "darkgreen")) +
  scale_fill_manual(values = c(tS1 = "firebrick3", tS2 = "steelblue", tS3 = "darkgreen")) +
  labs(
    x = "singScore",
    y = "Density",
    fill = NULL,
    color = NULL
  ) +
  theme_classic() +
 theme(
    axis.title.y = element_text(size = 13),
    axis.ticks.y = element_line(color = "gray"),
    axis.text.y = element_text(size = 13),
    axis.ticks.x = element_line(color = "gray"),
    axis.text.x = element_text(size = 13),
    axis.title.x = element_text(size = 15),
    legend.text = element_text(size = 15),
    legend.position = "top"
  )
```
# raw singscores are clearly on different scales

tS1 mean = 0.349, sd = 0.084
tS2 mean = 0.281, sd = 0.038
tS3 mean = 0.006, sd = 0.035

```{r}
summary(program_scores[, score_cols])
```


```{r}
sapply(program_scores[, score_cols], sd)
```

## z-score each program score across samples

```{r}
z_scores <- scale(program_scores[, score_cols])

colnames(z_scores) <- paste0(names(program_genes), "_z")

program_scores <- cbind(
  program_scores,
  as.data.frame(z_scores)
)

z_cols <- colnames(z_scores)
```


## assign each sample to the program with max z-score

```{r}
z_mat <- as.matrix(program_scores[, z_cols])

program_scores$Kim_subtype <- names(program_genes)[
  max.col(z_mat, ties.method = "first")
]

table(program_scores$Kim_subtype)
```



```{r}
head(program_scores)
```

```{r}
pheatmap::pheatmap(program_scores[,c(5:7)], color = colorRampPalette(c("dark blue", "white", "dark red"))(100))
```

# survival
```{r}
test <- program_scores %>%
  rename(sample = "immucan_id") %>%
  left_join(traits, "immucan_id")

survminer::ggforest(coxph( Surv(time_to_event, censoring) ~ Kim_subtype,
                data =test))

ggforest(coxph(Surv(time_to_event, censoring) ~ tS3_singscore, data= test))
```


## check: does each program show a cluster split?

```{r}
cluster_results <- list()

for (program in names(program_genes)) {
  
  genes <- program_genes[[program]]
  score_col <- paste0(program, "_singscore")
  
  mat <- rna_luad[, genes, drop = FALSE]
  
  # z-score each gene across samples
  mat_z <- scale(mat)
  
  # cluster samples using only this program's genes
  sample_hclust <- hclust(
    dist(mat_z),
    method = "complete"
  )
  
  clusters <- cutree(sample_hclust, k = 2) # 2 clusters: high & low
  
  score_vec <- program_scores[[score_col]]
  names(score_vec) <- program_scores$sample
  
  cluster_df <- data.frame(
    sample = names(clusters),
    program = program,
    cluster = paste0("cluster_", clusters),
    singscore = score_vec[names(clusters)]
  )
  
  cluster_medians <- aggregate(singscore ~ cluster, cluster_df, median)
  
  high_cluster <- cluster_medians$cluster[
    which.max(cluster_medians$singscore)
  ]
  
  cluster_df$group <- ifelse(
    cluster_df$cluster == high_cluster,
    paste0(program, "_high"),
    paste0(program, "_low")
  )
  
  wilcox_res <- wilcox.test(singscore ~ group, data = cluster_df)
  
  cluster_results[[program]] <- list(
    cluster_df = cluster_df,
    counts = table(cluster_df$group),
    medians = cluster_medians,
    p_value = wilcox_res$p.value,
    hclust = sample_hclust
  )
}
```



```{r}
plot_df <- do.call(
  rbind,
  lapply(cluster_results, function(x) x$cluster_df)
)

plot_df <- plot_df |> 
  mutate(
    group_short = if_else(grepl("_high$", group), "high", "low"),
    plot_group = paste(program, group_short),
    plot_group = factor(
      plot_group,
      levels = c(
        "tS1 low", "tS1 high",
        "tS2 low", "tS2 high",
        "tS3 low", "tS3 high"
      )
    )
  )

annot_df <- plot_df |> 
  group_by(program) |> 
  summarise(
    x_start = min(as.numeric(plot_group)) + 0.15,
    x_end = max(as.numeric(plot_group)) - 0.15,
    x_mid = mean(as.numeric(plot_group)),
    y_line = max(singscore, na.rm = TRUE) +
      0.08 * diff(range(singscore, na.rm = TRUE)),
    y_star = max(singscore, na.rm = TRUE) +
      0.14 * diff(range(singscore, na.rm = TRUE)),
    .groups = "drop"
  )
```


```{r, fig.width=10, fig.height=6}
clusters <- ggplot(plot_df, aes(x = plot_group, y = singscore)) +
  geom_boxplot(
    fill = "gray",
    color = "darkgray",
    staplewidth = 0.3
  ) +
  stat_summary(
    geom = "crossbar",
    fun = median,
    fun.min = median,
    fun.max = median,
    width = 0.75,
    linewidth = 0.15
  ) +
  annotate(
    geom = "segment",
    x = 1.1, xend = 1.9,
    y = max(plot_df$singscore[plot_df$program == "tS1"]) + 0.02,
    yend = max(plot_df$singscore[plot_df$program == "tS1"]) + 0.02
  ) +
  annotate(
    geom = "text",
    x = 1.5,
    y = max(plot_df$singscore[plot_df$program == "tS1"]) + 0.035,
    label = "*",
    size = 5
  ) +
  annotate(
    geom = "segment",
    x = 3.1, xend = 3.9,
    y = max(plot_df$singscore[plot_df$program == "tS2"]) + 0.02,
    yend = max(plot_df$singscore[plot_df$program == "tS2"]) + 0.02
  ) +
  annotate(
    geom = "text",
    x = 3.5,
    y = max(plot_df$singscore[plot_df$program == "tS2"]) + 0.035,
    label = "*",
    size = 5
  ) +
  annotate(
    geom = "segment",
    x = 5.1, xend = 5.9,
    y = max(plot_df$singscore[plot_df$program == "tS3"]) + 0.02,
    yend = max(plot_df$singscore[plot_df$program == "tS3"]) + 0.02
  ) +
  annotate(
    geom = "text",
    x = 5.5,
    y = max(plot_df$singscore[plot_df$program == "tS3"]) + 0.035,
    label = "*",
    size = 5
  ) +
  labs(
    x = "Kim LUAD program",
    y = "singScore"
  ) +
  coord_cartesian(clip = "off") +
  theme_classic(base_family = "Helvetica") +
  theme(
    axis.title.y = element_text(size = 13),
    axis.ticks.y = element_line(color = "gray"),
    axis.text.y = element_text(size = 13),
    axis.ticks.x = element_blank(),
    axis.text.x = element_text(size = 15),
    axis.title.x = element_text(size = 15)
  )

clusters
```

```{r}
ggsave("clusters.png", width = 8, height = 4)
```


