---
title: "Figure_2"
author: "Daniel Schulz"
date: "2023-10-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r}
library(tidyverse)
library(ggpubr)
library(SingleCellExperiment)
library(survivalAnalysis)
library(ggsurvfit)
library(survminer)
library(survival)
library(edgeR)
library(cowplot)
library(ggrepel)
library(igraph)
library(ggraph)
library(ggrepel)
library(scales)
library(ggnewscale)
library(imcRtools)
library(purrr)
library(broom)
```

```{r}
mount_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/"
out_folder <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/revised_figures/Figure_2/"

if (!dir.exists(out_folder)) {
  dir.create(out_folder)
}

source("code/lollipop_function.R")
options(bitmapType = "cairo")
```

# load data
we will load the data we generated at the beginning of figure 2
```{r}
dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/abundances_dat.rds"))
dens_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/densities_dat.rds"))
cello_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/cello_dat.rds"))
clin_dat <- read.csv("/mnt/immucan_volume/clinical_data/Lung_NSCLC2/2026_final/clinical.csv")
traits <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/all_traits.csv"))
modules <- as.data.frame(readRDS( "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/patient_imaging_meta_cluster_scores_tpm.rds"))

rna <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/rna_scores_tpm.rds")
rna$immucan_id  <- rownames(rna)
rownames(rna) <- NULL

modules <- modules  %>%
  mutate(immucan_id = rownames(modules))
rownames(modules) <- NULL
module_names <- colnames(modules)[1:7]

# variables
data_type <- unique(str_split_i(string = colnames(dat[,-1]),pattern = "__",i = 2))
```

# 2A Imaging modules

code is here: [imaging module plot](../../../imaging_modules/imaging_module_generation.Rmd)

# Histology

# 2B (alternativ) images

```{r, fig.height=20, fig.width=20}

sce1 <- readRDS(paste0(mount_path,"data/IMC/Panel_1/Rout/sce.rds"))
sce1 <- sce1[,sce1$ROI != 4]


pats <- modules %>%
  pivot_longer(cols = -immucan_id) %>%
  filter(!immucan_id %in% c("LUNG-NSCLC2-0688", "LUNG-NSCLC2-0587","LUNG-NSCLC2-0679","LUNG-NSCLC2-0557","LUNG-NSCLC2-0638", "LUNG-NSCLC2-0576","LUNG-NSCLC2-0677","LUNG-NSCLC2-0721","LUNG-NSCLC2-0669")) %>%
  group_by(name) %>%
  slice_max(value,n = 4)

pats %>% arrange(immucan_id)
unique(pats$immucan_id)


cur_sce <- sce1[,sce1$immucan_id %in% pats$immucan_id]
cur_sce1 <- cur_sce[,cur_sce$ROI == 1]
cur_sce1$im_type <- pats$name[match(cur_sce1$immucan_id,pats$immucan_id)]
cur_sce1$title <- paste0(cur_sce1$im_type, "_", cur_sce1$immucan_id,"_", cur_sce1$ROI)
cur_sce2 <- cur_sce[,cur_sce$ROI == 2]
cur_sce2$im_type <- pats$name[match(cur_sce2$immucan_id,pats$immucan_id)]
cur_sce2$title <- paste0(cur_sce2$im_type, "_", cur_sce2$immucan_id,"_", cur_sce2$ROI)
cur_sce3 <- cur_sce[,cur_sce$ROI == 3]
cur_sce3$im_type <- pats$name[match(cur_sce3$immucan_id,pats$immucan_id)]
cur_sce3$title <- paste0(cur_sce3$im_type, "_", cur_sce3$immucan_id,"_", cur_sce3$ROI)


# 1. Order columns so each class's 4 images are grouped consecutively,
#    then lock that order into the "title" factor levels
cur_sce1 <- cur_sce1[, order(cur_sce1$im_type, cur_sce1$immucan_id)]
cur_sce1$title <- factor(cur_sce1$title, levels = unique(cur_sce1$title))

p1 <- plotSpatial(cur_sce1, img_id = "title", node_color_by = "celltypes",
                  ncols = 7, node_size_fix = 0.1) +
  scale_color_manual(values = metadata(sce1)$colour_vectors$celltypes) +
  facet_wrap(~title, ncol = 7, dir = "v")   # fill down each column before moving right

cur_sce2 <- cur_sce2[, order(cur_sce2$im_type, cur_sce2$immucan_id)]
cur_sce2$title <- factor(cur_sce2$title, levels = unique(cur_sce2$title))

p2 <- plotSpatial(cur_sce2, img_id = "title", node_color_by = "celltypes",
                  ncols = 7, node_size_fix = 0.1) +
  scale_color_manual(values = metadata(sce1)$colour_vectors$celltypes) +
  facet_wrap(~title, ncol = 7, dir = "v")   # fill down each column before moving right

cur_sce3 <- cur_sce3[, order(cur_sce3$im_type, cur_sce3$immucan_id)]
cur_sce3$title <- factor(cur_sce3$title, levels = unique(cur_sce3$title))

p3 <- plotSpatial(cur_sce3, img_id = "title", node_color_by = "celltypes",
                  ncols = 7, node_size_fix = 0.1) +
  scale_color_manual(values = metadata(sce1)$colour_vectors$celltypes) +
  facet_wrap(~title, ncol = 7, dir = "v")   # fill down each column before moving right

p1
p2
p3
ggsave(plot = p2+
  theme(strip.text.x = element_blank(), legend.position = "bottom"), filename = file.path(out_folder,"2B_module_exmaple_images.pdf"),device = cairo_pdf,width = 10, height = 6)
ggsave(plot = p2+
  theme(strip.text.x = element_blank(), legend.position = "bottom", legend.text = element_text(size = 12)), filename = file.path(out_folder,"2B_module_exmaple_images.png"),device = png,width = 3000, height = 1800, units = "px")

```


# 2C imaging modules for histology
```{r, fig.height=10, fig.width=8}
pat_nums <-  left_join(modules, traits, "immucan_id") %>%
  filter(!is.na(LUAD_vs_LUSC)) %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"))
table(pat_nums[["LUAD_vs_LUSC"]])

dat_z <- left_join(modules, traits, "immucan_id") %>%
  filter(!is.na(LUAD_vs_LUSC)) %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC")) %>%
  pivot_longer(cols = all_of(module_names))

result <- lollipop_plot(dat_z, "LUAD_vs_LUSC", "LUAD", "LUSC", col_A = "blue", col_B = "orange")
p <- result$plot +
  coord_cartesian(xlim = c(-.5, .5))
p
ggsave(plot = p, path = out_folder,filename = "2C_Imaging_modules_for_histology.pdf",width = 6, height = 3.5,device = cairo_pdf)
```

# 2D diff abundance histology 
here we will estimate the dispersion across the whole data set and then do the linear model of count comparison only for the groups of interest.
```{r, fig.height=20, fig.width=10}
condition <- c("Adenocarcinoma")
names(condition) <- c("LUAD_vs_LUSC")
j <- names(condition)
p.list <- list()

for (i in c(data_type,"cello1","cello2")) {
    
    if(grepl("cello",i)){
      abundances <- cello_dat %>%
        select(contains(i))
      abundances <- t(abundances)
      colnames(abundances) <- cello_dat$immucan_id
    } else {
      abundances <- dat %>%
        select(contains(i)) %>%
        filter(complete.cases(.))
      abundances <- t(abundances)
      colnames(abundances) <-dat %>%
        select(contains(i),immucan_id) %>%
        filter(complete.cases(.)) %>%
        pull(immucan_id)
    }

  
  # merge with clinical data
  anno <- clin_dat %>%
    filter(!is.na(.data[["simple_histology"]]))
  
  abundances <- abundances[,colnames(abundances) %in% anno$immucan_id]
  
  meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
  Size <- factor(meta[,"simple_histology"])
  meta$Size <- Size
  
  # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
  y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
  design <- model.matrix(~0+group,data = y.ab$samples)
  colnames(design) <- levels(y.ab$samples$group)
  # we do not fit a trend 
  y.ab <- estimateDisp(y.ab, design, trend.method = "none")
  y.ab$common.dispersion
  #plotBCV(y.ab,pch = 5)
  
  # we only compare the diff abundance for Adeno vs Sqam.
  fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
  fit <- glmLRT(glmfit = fit.ab,contrast = c(-1,0,0,0,1,0))
  #plotQLDisp(fit.ab, pch = 5)
  
  sig_dat <- fit$table
  sig_dat$celltypes <- rownames(sig_dat)
  sig_dat$cond <- j
  sig_dat$comparison <- gsub("Size","",fit$comparison)
  p.list[[i]] <- sig_dat
}   


cur_dat <- do.call(rbind,p.list)

cur_dat <- cur_dat %>%
  mutate(tech = gsub("_abundance","",str_split_i(celltypes,"__",2))) %>%
  group_by(tech) %>%
  mutate(FDR = p.adjust(PValue,method = "BH"),
         sig = ifelse(PValue < 0.05 & logFC > 0, "UP_trend",NA),
         sig = ifelse(FDR < 0.1 & logFC > 0, "UP",sig),
         sig = ifelse(PValue < 0.05 & logFC < 0, "DOWN_trend",sig),
         sig = ifelse(FDR < 0.1 & logFC < 0, "DOWN",sig),
         sig = ifelse(is.na(sig),"NOTSIG", sig),
         ct_name = gsub("_density","",str_split_i(celltypes,"__",1)))
  
to_show_dens <- cur_dat %>%
  filter(tech %in% c("IMC1", "IMC2", "IF1", "IF2","IF3")) %>%
  arrange((abs(FDR))) %>%
  ungroup() %>%
  slice_head(n = 15) %>%
  pull(celltypes)
to_show_spat <- cur_dat %>%
  filter(tech %in% c("cello1", "cello2")) %>%
  arrange((abs(FDR))) %>%
  ungroup() %>%
  slice_head(n = 5) %>%
  pull(celltypes)

cur_dat <- cur_dat %>%
  filter(celltypes %in% c(to_show_dens, to_show_spat)) %>%
  mutate(celltypes = gsub("_abundance","",celltypes),
         cond = "LUSC_over_LUAD")

# get max and min fold change values
max_up <- max(cur_dat$logFC)+1
min_down <- min(cur_dat$logFC)-1

p <- cur_dat %>%
  ggplot(aes(x=celltypes, y = cond,col = logFC, size= -log10(PValue)))+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(
    colours = c("blue", "white", "red"), 
    values = rescale(c(round(min_down), 0, round(max_up))),
    limits = c(round(min_down), round(max_up)),
    guide = guide_colorbar(
      title = "logFC",                   
      #barwidth = 2,                     
      #barheight = 1,                   
      title.theme = element_text(size = 10),
      label.theme = element_text(size = 10)
    )
  )+
  scale_size(range = c(1,4))+
  new_scale_colour() +
  geom_point(data = cur_dat,shape = 1,aes(size = -log10(PValue), col = sig), stroke =1.5)+
  scale_color_manual(values = c("UP" = "#9f1214","UP_trend" = "#e73032","DOWN" = "#25567d","DOWN_trend" = "#428bc7","NOTSIG" = "snow2"), name = "stats")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 10),
        axis.text.y = element_text(size = 10),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        legend.location = "plot",
        legend.position = "right",legend)+
  coord_flip()
p

ggsave(filename = paste0(out_folder,"2D_diff_abundance_histology.pdf"),
       plot = p,
       device = "pdf",
       width = 4,
       height = 7)

```


# 2E RNA scores

```{r, fig.height=15, fig.width=15}
pat_nums <-  left_join(rna, traits, "immucan_id") %>%
  filter(!is.na(LUAD_vs_LUSC)) %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"))
table(pat_nums[["LUAD_vs_LUSC"]])

dat_z <- left_join(rna, traits, "immucan_id") %>%
  filter(!is.na(LUAD_vs_LUSC)) %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC")) %>%
  pivot_longer(cols = 1:29)

result <- lollipop_plot(dat_z, condition = "LUAD_vs_LUSC", var_A = "LUAD", var_B = "LUSC", col_A = "blue", col_B = "orange",zscore_data = TRUE)
sigs <- result$data %>% filter(sig == "FDR < 0.1") %>% pull(name)
result_sel <- lollipop_plot(dat_z %>% filter(name %in% sigs), condition = "LUAD_vs_LUSC", var_A = "LUAD", var_B = "LUSC", col_A = "blue", col_B = "orange",zscore_data = TRUE)
p <- result_sel$plot +
  ggtitle(paste0("RNA modules for Histology"))+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))+
  theme(axis.text.y = element_text(size = 10),
        legend.text = element_text(size = 10),
        axis.title.x = element_text(size = 10),
        legend.position = "bottom", legend.box = "vertical")+
  guides(color = guide_legend(nrow = 1, byrow = TRUE))+
  xlab("Δ z-score")
p

ggsave(plot = p, path = out_folder,filename = "2E_RNA_difference_for_histology_selected.pdf",width = 5.5, height = 5,device = cairo_pdf)
```

# inflammation in histology

```{r}
library(org.Hs.eg.db)
library(AnnotationDbi)
library(msigdbr)
library(singscore)

rna_mat <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/rna_mat_tpm.rds")


# ── gene signatures ────────────────────────────────────────────────────
hallmark_sets <- msigdbr(species = "Homo sapiens", collection = "H") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  split(.$gs_name) %>%
  lapply(function(x) x$gene_symbol)

# Targeted Hallmark gene sets of interest
target_sets <- c(
  "HALLMARK_INFLAMMATORY_RESPONSE",
  "HALLMARK_INTERFERON_GAMMA_RESPONSE",
  "HALLMARK_INTERFERON_ALPHA_RESPONSE",
  "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
  "HALLMARK_IL6_JAK_STAT3_SIGNALING",
  "HALLMARK_IL2_STAT5_SIGNALING"
)
gs_list_hallmark <- hallmark_sets[intersect(target_sets, names(hallmark_sets))]

# ── 2. Custom gene sets ────────────────────────────────────────────────────────
# CXCL9/10/11 chemokine axis (T cell recruitment)
gs_list_custom <- list(
  CXCL9_10_11_axis = c("CXCL9", "CXCL10", "CXCL11", "CXCR3",
                       "CXCL13", "CXCR5"),
  T_cell_activation = c("CD8A", "CD8B", "GZMB", "PRF1", "IFNG",
                        "TNF", "IL2", "PDCD1", "HAVCR2", "LAG3"),
  type_I_IFN = c("IFNB1", "IRF7", "MX1", "OAS1", "OAS2", "OAS3",
                 "ISG15", "IFIT1", "IFIT2", "IFIT3", "RSAD2", "HERC5"),
  NF_kB_target  = c("NFKB1", "RELA", "RELB", "NFKB2", "REL",
                    "BCL2L1", "IL1B", "IL6", "CXCL8", "TNF", "ICAM1"),
  TCR_signaling = c("CD3D", "CD3E", "CD3G", "ZAP70", "LCK", "LAT",
                    "ITK", "PLCG1", "NFATC1", "NFATC2"),
  cytotoxic = c("GZMB", "GZMA", "GZMK", "PRF1", "NKG7", "IFNG", "KLRG1"),
  exhaustion = c("PDCD1","LAG3","HAVCR2","TX","CTLA4", "IDO1")
  )

all_gs <- c(gs_list_hallmark, gs_list_custom)

# compute score
rna_for_score <- t(rna_mat)

# Filter gene sets to genes present in the RNA-seq matrix
all_gs <- lapply(all_gs, function(g) intersect(g, rownames(rna_for_score)))
all_gs <- all_gs[sapply(all_gs, length) >= 4]



overlap <- sapply(all_gs, function(g) sum(g %in% rownames(rna_for_score)))
total   <- sapply(all_gs, length)
print(data.frame(found = overlap, total = total, pct = round(100 * overlap / total)))

# ── Score with singscore ──────────────────────────────────────────────────────
ranked <- rankGenes(rna_for_score)

scores_list <- lapply(names(all_gs), function(nm) {
  sc <- simpleScore(ranked, upSet = all_gs[[nm]], centerScore = TRUE)
  setNames(sc$TotalScore, rownames(sc))
})


rna_scores <- as.data.frame(do.call(cbind, scores_list))
colnames(rna_scores) <- names(all_gs)
rownames(rna_scores) <- colnames(rna_for_score)
rna_scores$immucan_id <- rownames(rna_scores)
rownames(rna_scores) <- NULL

dat_z <- left_join(rna_scores, traits, "immucan_id") %>%
  filter(!is.na(LUAD_vs_LUSC)) %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC")) %>%
  pivot_longer(cols = 1:length(all_gs))

result <- lollipop_plot(dat_z, condition =  "LUAD_vs_LUSC",var_A = "LUAD", var_B = "LUSC", col_A = "blue", col_B = "orange", zscore_data = TRUE)
p <- result$plot+
    ylab("RNA modules")+
  ggtitle("LUAD vs LUSC")+
  theme(axis.title.y = element_blank(),
        axis.title.x = element_text(size= 18),
        legend.position = "bottom", 
        legend.text = element_text(size = 18),
        legend.title = element_text(size = 18))+
  guides(color = guide_legend(nrow = 2, byrow = TRUE),
         size = guide_legend(nrow = 2, byrow = TRUE))

p
```

```{r, fig.height=10, fig.width=10}
p <- result$plot+
    ylab("RNA modules")+
  ggtitle("LUAD vs LUSC")+
  theme(axis.title.y = element_blank(),
        axis.title.x = element_text(size= 18),
        #legend.position = "bottom", 
        legend.text = element_text(size = 18),
        legend.title = element_text(size = 18),
        title = element_text(size = 18))
p
ggsave(plot = p, path = out_folder,filename = "alternative_inflammation.pdf",width = 10, height = 6 ,device = cairo_pdf)

```


# 2F tumor types imaging modules
we will not subtype LUSC samples because we have too few samples
```{r}
library(ComplexHeatmap)
ttypes <- read.csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/tumor_subtypes.csv")

dat_z <- ttypes %>% left_join(modules, "immucan_id")

# numbers
table(dat_z$subtype)

dat_z$immucan_id <- NULL
luad_dat <- dat_z %>%
  filter(subtype %in% c("TRU","PI","PP")) %>%
  mutate(across(.cols = -subtype, scale)) %>%
  group_by(subtype) %>%
  summarise(across(.fns = mean))

luad_mat <- as.matrix(luad_dat[,-1])
rownames(luad_mat) <- luad_dat$subtype

compute_subtype_sig <- function(data, subtypes) {
  df <- data %>% filter(subtype %in% subtypes)
  modules <- setdiff(colnames(df), "subtype")

  purrr::map_dfr(subtypes, function(s) {
    purrr::map_dfr(modules, function(m) {
      x <- df[[m]][df$subtype == s]
      y <- df[[m]][df$subtype != s]
      pval <- tryCatch(t.test(x, y)$p.value, error = function(e) NA_real_)
      tibble::tibble(subtype = s, module = m, pvalue = pval)
    })
  }) %>%
    mutate(padj = p.adjust(pvalue, method = "BH"))
}

# one helper for either statistic, so p and FDR matrices stay perfectly aligned
to_stat_matrix <- function(sig_df, ref_mat, value_col) {
  sig_df %>%
    dplyr::select(subtype, module, dplyr::all_of(value_col)) %>%
    tidyr::pivot_wider(names_from = module, values_from = dplyr::all_of(value_col)) %>%
    tibble::column_to_rownames("subtype") %>%
    as.matrix() %>%
    .[rownames(ref_mat), colnames(ref_mat)]
}

luad_sig <- compute_subtype_sig(dat_z, c("TRU", "PI", "PP"))

# transpose to match t(luad_mat): modules x subtypes
luad_p_mat   <- t(to_stat_matrix(luad_sig, luad_mat, "pvalue"))
luad_fdr_mat <- t(to_stat_matrix(luad_sig, luad_mat, "padj"))

# star tier from raw p, but only shown when BH-FDR < 0.1
sig_star <- function(p, fdr) {
  if (is.na(p) || is.na(fdr) || fdr >= 0.1) return("")
  if (p < 0.001) return("***")
  if (p < 0.01)  return("**")
  if (p < 0.05)  return("*")
  ""
}

ann_df <- data.frame(Subtype = c(unique(luad_dat$subtype)),
                     subtype_col = c("blue","deepskyblue","dodgerblue2"))
ha <- HeatmapAnnotation(subtype = ann_df$Subtype,
                        col = list(subtype = structure(ann_df$subtype_col, names = ann_df$Subtype)))

z_col <- circlize::colorRamp2(c(-1, 0, 1), c("#2166AC", "white", "#B2182B"))

h <- ComplexHeatmap::Heatmap(
  t(luad_mat), name = "mean z_score",
  cell_fun = function(j, i, x, y, width, height, fill) {
    lab <- sig_star(luad_p_mat[i, j], luad_fdr_mat[i, j])
    if (nzchar(lab)) {
      grid::grid.text(lab, x, y - height * 0.15,
                      gp = grid::gpar(fontsize = 12))
    }
  },
  top_annotation = ha,
  col = z_col
)
h

pdf(file = paste0(out_folder,"2F_imaging_modules_for_tumor_subtypes_LUAD.pdf"), width = 4.5, height = 4)
draw(h, merge_legend = TRUE)
dev.off()

```



# 2G tumor types for stages

```{r}
scores <- read.csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/tumor_subtype_scores.csv")

scores <- left_join(scores, traits,"immucan_id")


scores <- scores %>%
  rename(magnoid = "PP",
         squamoid = "PI",
         bronchioid = "TRU")

dat_z <-  scores %>%
  pivot_longer(cols = 1:3) %>% 
  filter(simple_stage != "IV",
         !is.na(simple_stage), 
         LUAD_vs_LUSC == "Adenocarcinoma")

table(dat_z$simple_stage)/3
sigs <- dat_z %>%
  select(immucan_id,name, value,simple_stage) %>%
  mutate(stage = ifelse(simple_stage == "I",1,NA),
         stage = ifelse(simple_stage == "II",2,stage),
         stage = ifelse(simple_stage == "III",3,stage),
         stage = ifelse(simple_stage == "IV",4,stage),
         stage = as.numeric(as.factor(stage))) %>%
  filter(stage != 4) %>%
  group_by(name) %>%
  nest() %>%
  mutate(model = map(data, ~ lm(value ~ stage, data = .x)),
         summary = map(model, broom::tidy)) %>%
  select(name, summary) %>%
  unnest(c(summary)) %>%
  filter(term == "stage") %>%
  mutate(tech = str_split_i(name,"__",2)) %>%
  group_by(tech) %>%
  mutate(FDR = p.adjust(p = p.value,method = "BH"),
         sig_p = ifelse(p.value < 0.05,TRUE, FALSE),
         sig_fdr = ifelse(FDR < 0.1,"T", "F"))


summary_dat <- sigs %>%
  #filter(sig_p == TRUE) %>%
  #filter(grepl("CD8_GATA3_BCL2|^CD8_density|CD8_Ki67_density|CD8_TCF7_density|Tumor_IDO1",name)) %>%
  mutate(y_position = 1000,
         name = gsub(pattern = "density_",replacement = "",x = name)) %>%
  ungroup() %>%
  as.data.frame()

p <- dat_z %>%
  select(immucan_id,name, value,simple_stage) %>%
  mutate(stage = ifelse(simple_stage == "I",1,NA),
         stage = ifelse(simple_stage == "II",2,stage),
         stage = ifelse(simple_stage == "III",3,stage),
         stage = ifelse(simple_stage == "IV",4,stage),
         stage = as.numeric(as.factor(stage)),
         name = gsub(pattern = "density_",replacement = "",x = name)) %>%
  filter(name %in% summary_dat$name,
         stage != 4) %>%
  ggplot(aes(x = name,y = value,fill=as.factor(stage))) +
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(col = as.factor(stage)),position = position_jitterdodge(jitter.width = 0.2),size= 0.5)+
  scale_fill_manual(values = alpha(c("1" = "palevioletred3", "2" = "darkseagreen3", "3" = "dodgerblue"),alpha = 0.6))+
  scale_color_manual(values = c("1" = "palevioletred3", "2" = "darkseagreen3", "3" = "dodgerblue"))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90,vjust= 0.5, hjust = 1),
        axis.title.x = element_blank(),
        title = element_text(size = 10))+
  guides(fill=guide_legend(title="Stage"),
         col=guide_legend(title="Stage"))+
  geom_text(data = summary_dat, aes(x = name,y=1,label = paste0("fdr: ",sig_fdr)), inherit.aes = FALSE, size =3)+
  geom_text(data = summary_dat, aes(x = name,y=1.12,label = paste0("p : ",round(p.value,digits = 3))), inherit.aes = FALSE, size = 3)+
  #coord_cartesian(ylim = c(-0.35, 0.5))+
  ggtitle(label = "Tumor subtypes across stages in LUAD")+
  labs(y = "Tumor subtype score")
p  
ggsave(plot = p, path = out_folder,filename = "2G_tumor_types_for_stages_LUAD.pdf",width = 3.5, height = 3.5,device = cairo_pdf)
```


# Stage
## 2H imaging modules -remove/not show?
```{r}
dat_z <- left_join(modules, traits, "immucan_id") %>%
  filter(!is.na(simple_stage),
         simple_histology == "Adenocarcinoma") %>%
  pivot_longer(cols = all_of(module_names))

sigs <- dat_z %>%
  dplyr::select(immucan_id,name, value,simple_stage) %>%
  mutate(stage = ifelse(simple_stage == "I",1,NA),
         stage = ifelse(simple_stage == "II",2,stage),
         stage = ifelse(simple_stage == "III",3,stage),
         stage = ifelse(simple_stage == "IV",4,stage),
         stage = as.numeric(as.factor(stage))) %>%
  filter(stage != 4) %>%
  group_by(name) %>%
  nest() %>%
  mutate(model = map(data, ~ lm(value ~ stage, data = .x)),
         summary = map(model, broom::tidy)) %>%
  dplyr::select(name, summary) %>%
  unnest(c(summary)) %>%
  filter(term == "stage") %>%
  mutate(tech = str_split_i(name,"__",2)) %>%
  group_by(tech) %>%
  mutate(FDR = p.adjust(p = p.value,method = "BH"),
         sig_p = ifelse(p.value < 0.05,TRUE, FALSE),
         sig_fdr = ifelse(FDR < 0.1,"T", "F"))


summary_dat <- sigs %>%
  #filter(sig_p == TRUE) %>%
  #filter(grepl("CD8_GATA3_BCL2|^CD8_density|CD8_Ki67_density|CD8_TCF7_density|Tumor_IDO1",name)) %>%
  mutate(y_position = 1000,
         name = gsub(pattern = "density_",replacement = "",x = name)) %>%
  ungroup() %>%
  as.data.frame()

p <- dat_z %>%
  dplyr::select(immucan_id,name, value,simple_stage) %>%
  mutate(stage = ifelse(simple_stage == "I",1,NA),
         stage = ifelse(simple_stage == "II",2,stage),
         stage = ifelse(simple_stage == "III",3,stage),
         stage = ifelse(simple_stage == "IV",4,stage),
         stage = as.numeric(as.factor(stage)),
         name = gsub(pattern = "density_",replacement = "",x = name)) %>%
  filter(name %in% summary_dat$name,
         stage != 4) %>%
  ggplot(aes(x = name,y = log10(value+1),fill=as.factor(stage))) +
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(col = as.factor(stage)),position = position_jitterdodge(jitter.width = 0.2),size= 0.5)+
  scale_fill_manual(values = alpha(c("1" = "palevioletred3", "2" = "darkseagreen3", "3" = "dodgerblue"),alpha = 0.6))+
  scale_color_manual(values = c("1" = "palevioletred3", "2" = "darkseagreen3", "3" = "dodgerblue"))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90,vjust= 0.5, hjust = 1, size = 12),
        axis.title = element_blank())+
  guides(fill=guide_legend(title="Stage"),
         col=guide_legend(title="Stage"))+
  geom_text(data = summary_dat, aes(x = name,y=0.5,label = paste0("fdr: ",sig_fdr)), inherit.aes = FALSE, size = 2.5)+
  geom_text(data = summary_dat, aes(x = name,y=0.6,label = paste0("p : ",round(p.value,digits = 3))), inherit.aes = FALSE, size = 2.5)+
  coord_cartesian(ylim = c(-.55,0.6))+
  ggtitle(label = "Imaging module differences across stages in LUAD")
  
p

#ggsave(plot = p, path = out_folder,filename = "2H_Imaging_modules_for_stage_LUAD.pdf",width = 6, height = 3.5,device = cairo_pdf)


```


# Smoking

## 2H imaging

### imaging modules never vs current
```{r}
pat_nums <- left_join(modules, traits, "immucan_id") %>%
  filter(LUAD_vs_LUSC == "Adenocarcinoma")
table(pat_nums[["never_vs_current_smoker"]])

dat_z <- left_join(modules, traits, "immucan_id") %>%
  pivot_longer(cols = all_of(module_names)) %>%
  filter(LUAD_vs_LUSC == "Adenocarcinoma")

result <- lollipop_plot(dat_z, "never_vs_current_smoker", "Current smoker", "Non-smoker", col_A = "blue", col_B = "cyan3")
p1 <- result$plot+
  ggtitle("Current smokers vs Non-smokers in LUAD")+
  ylab("Imaging modules")+
  theme(axis.title.y = element_text(size= 14))+
  coord_cartesian(xlim = c(-0.4,0.4))
p1

cur_order <- result$data$name

ggsave(plot = p1, path = out_folder,filename = "2H_Imaging_modules_for_smoking_LUAD.pdf",width = 6, height = 3.5,device = cairo_pdf)
```

### imaging modules never vs ex
```{r}
result <- lollipop_plot(dat_z, "never_vs_Ex_smoker", "Ex-smoker", "Non-smoker", order = FALSE, order_vec = cur_order, col_A = "lightblue", col_B = "cyan3")
p2 <- result$plot+
  theme(axis.text.y = element_blank())+
  ggtitle("Ex-smokers vs Non-smokers")
```

### imaging modules current vs ex
```{r}
result <- lollipop_plot(dat_z, "current_vs_Ex_smoker", "Current smoker", "Ex-smoker", order = FALSE, order_vec =cur_order, col_A = "blue", col_B = "lightblue")
p3 <- result$plot +
  theme(axis.text.y = element_blank())+
  ggtitle("Current smokers vs Ex-smokers")
```

### plot imaging
```{r, fig.height=4, fig.width= 20}
p <- cowplot::plot_grid(p1, p2, p3, ncol = 3, rel_widths = c(1.05,0.7,0.7))
p

#ggsave(plot = p, path = out_folder,filename = "2J_Imaging_modules_for_smoking_LUAD.pdf",width = 20, height = 3.75,device = cairo_pdf)
```







# 2I diff abundance smoking bubble plot

```{r, fig.height=15, fig.width=25}
condition <- c("Non-smoker","Non-smoker","Ex-smoker")
names(condition) <- c("never_vs_current_smoker","never_vs_Ex_smoker","current_vs_Ex_smoker")

p.list <- list()

# clinical data prep
anno <- clin_dat %>%
  filter(!is.na(smoking_status),
         simple_histology == "Adenocarcinoma")
# numbers
table(anno$current_vs_Ex_smoker)
table(anno$never_vs_current_smoker)

for (i in c(data_type,"cello1","cello2")) {
    
    if(grepl("cello",i)){
      abundances <- cello_dat %>%
        select(contains(i))
      abundances <- t(abundances)
      colnames(abundances) <- cello_dat$immucan_id
    } else {
      abundances <- dat %>%
        select(contains(i)) %>%
        filter(complete.cases(.))
      abundances <- t(abundances)
      colnames(abundances) <-dat %>%
        select(contains(i),immucan_id) %>%
        filter(complete.cases(.)) %>%
        pull(immucan_id)
    }
  
  abundances <- abundances[,colnames(abundances) %in% anno$immucan_id]
  
  meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
  Size <- factor(meta[,"smoking_status"])
  meta$Size <- Size
  
  # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
  y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
  design <- model.matrix(~0+group,data = y.ab$samples)
  colnames(design) <- levels(y.ab$samples$group)
  # we do not fit a trend 
  y.ab <- estimateDisp(y.ab, design, trend.method = "none")
  y.ab$common.dispersion
  #plotBCV(y.ab,pch = 5)
  
  # we only compare the diff abundance for smoker 
  fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
  fit_1_2 <- glmLRT(glmfit = fit.ab,contrast = c(1,-1,0))
  fit_1_3 <- glmLRT(glmfit = fit.ab,contrast = c(1,0,-1))
  #fit_2_3 <- glmLRT(glmfit = fit.ab,contrast = c(0,1,-1))
  res_list <- list(fit_1_2,fit_1_3)
  
  cur_list <- lapply(res_list,function(x){
    sig_dat <- x$table
    sig_dat$celltypes <- rownames(sig_dat)
    sig_dat$comparison <- gsub("Size","",x$comparison)
    return(sig_dat)
  })
  
  p.list[[i]] <- do.call(rbind,cur_list)
}   

cur_dat_LUAD <- do.call(rbind,p.list)
cur_dat_LUAD <- cur_dat_LUAD %>%
  mutate(histology = "LUAD")

cur_dat <- cur_dat_LUAD %>%
  mutate(tech = gsub("_density","",str_split_i(celltypes,"__",2)),
         cond = ifelse(comparison == "1*Current smoker -1*Ex-smoker","CurrentSmoker_over_ExSmoker","ExSmoker_over_NonSmoker"),
         cond = ifelse(comparison == "1*Current smoker -1*Non-smoker","CurrentSmoker_over_NonSmoker",cond),
         cond = paste0(histology,"_",cond),
         cond = ifelse(comparison == "-1*low 1*high","SBS4-SBS92_High_over_Low",cond)) %>%
  group_by(tech) %>%
  mutate(FDR = p.adjust(PValue,method = "BH"),
         sig = ifelse(PValue < 0.05 & logFC > 0, "UP_trend",NA),
         sig = ifelse(FDR < 0.1 & logFC > 0, "UP",sig),
         sig = ifelse(PValue < 0.05 & logFC < 0, "DOWN_trend",sig),
         sig = ifelse(FDR < 0.1 & logFC < 0, "DOWN",sig),
         sig = ifelse(is.na(sig),"NOTSIG", sig),
         ct_name = gsub("_density","",str_split_i(celltypes,"__",1)))
  
to_show <- cur_dat %>%
  filter(sig %in% c("UP","DOWN")) %>%
  pull(ct_name)

cur_dat <- cur_dat %>%
  filter(ct_name %in% to_show) %>%
  mutate(celltypes = gsub("_abundance","",celltypes))

# get max and min fold change values
max_up <- max(cur_dat$logFC)+1
min_down <- min(cur_dat$logFC)-1

# plot the data
p <- cur_dat %>%
  ggplot(aes(x=celltypes, y = factor(cond, levels = c("SBS4-SBS92_High_over_Low",
                                                      "LUAD_CurrentSmoker_over_ExSmoker",
                                                      "LUAD_ExSmoker_over_NonSmoker",
                                                      "LUAD_CurrentSmoker_over_NonSmoker",
                                                      "LUSC_CurrentSmoker_over_ExSmoker")),col = logFC, size= -log10(PValue)))+
  #geom_tile(aes(fill = sig))+
  #theme_bw()+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(
    colours = c("blue", "white", "red"), 
    values = rescale(c(round(min_down), 0, round(max_up))),
    limits = c(round(min_down), round(max_up)),
    guide = guide_colorbar(
      title = "logFC",                   
      #barwidth = 2,                     
      #barheight = 1,                   
      title.theme = element_text(size = 10),
      label.theme = element_text(size = 10)
    )
  )+
  scale_size(range = c(1,4))+
  new_scale_colour() +
  geom_point(data = cur_dat,shape = 1,aes(size = -log10(PValue), col = sig), stroke =1.5)+
  scale_color_manual(values = c("UP" = "#9f1214","UP_trend" = "#e73032","DOWN" = "#25567d","DOWN_trend" = "#428bc7","NOTSIG" = "snow2"), name = "stats")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 10),
        axis.text.y = element_text(size = 10),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        legend.location = "plot",
        legend.position = "right",legend)
p

ggsave(filename = paste0(out_folder,"2I_diff_abundance_smoking.pdf"),
      plot = p,
      device = "pdf",
      width = 12,
      height = 3.5)
```

### smoking status LUSC
of note, there are no non-smokers for LUSC histology in our cohort 
```{r}
condition <- c("Non-smoker","Non-smoker","Ex-smoker")
names(condition) <- c("never_vs_current_smoker","never_vs_Ex_smoker","current_vs_Ex_smoker")
p.list <- list()

# merge with clinical data
anno <- clin_dat %>%
  filter(!is.na(smoking_status),
         simple_histology == "Squamous cell carcinoma")
  
for (i in data_type) {
  
  # get cell abundances and remove NAs
  abundances <- dat %>%
    dplyr::select(contains(i)) %>%
    filter(complete.cases(.))
  abundances <- t(abundances)
  colnames(abundances) <-dat %>%
    dplyr::select(contains(i),immucan_id) %>%
    filter(complete.cases(.)) %>%
    pull(immucan_id)
  
  abundances <- abundances[,colnames(abundances) %in% anno$immucan_id]
  
  meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
  Size <- factor(meta[,"smoking_status"])
  meta$Size <- Size
  
  # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
  y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
  design <- model.matrix(~0+group,data = y.ab$samples)
  colnames(design) <- levels(y.ab$samples$group)
  # we do not fit a trend 
  y.ab <- estimateDisp(y.ab, design, trend.method = "none")
  y.ab$common.dispersion
  #plotBCV(y.ab,pch = 5)
  
  # we only compare the diff abundance for Adeno vs Sqam.
  fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
  fit_1_2 <- glmLRT(glmfit = fit.ab,contrast = c(1,-1))
  res_list <- list(fit_1_2)
  
  cur_list <- lapply(res_list,function(x){
    sig_dat <- x$table
    sig_dat$celltypes <- rownames(sig_dat)
    sig_dat$comparison <- gsub("Size","",x$comparison)
    return(sig_dat)
  })
  
  p.list[[i]] <- do.call(rbind,cur_list)
}   

cur_dat_LUSC <- do.call(rbind,p.list)

cur_dat_LUSC <- cur_dat_LUSC %>%
  mutate(histology = "LUSC")
```


### smoking signature based

```{r}
# Merge with mutational signatures
Lung_NSCLC2_filtered <-read.delim(paste0(mount_path,"data/WES/mutation_signatures/Lung_NSCLC2_filtered.tsv"), sep="\t")
names(Lung_NSCLC2_filtered)[names(Lung_NSCLC2_filtered)=="patient"] <- "immucan_id"
anno_WES <- merge(clin_dat, Lung_NSCLC2_filtered, by="immucan_id", sort=F)

# Combine smoking signatures
anno_WES[,"SBS4_SBS92"] <- apply(anno_WES[,c("SBS4","SBS92")],1,sum)

# Dichotomise smoking signature
cutoff <- median(anno_WES$SBS4_SBS92)
anno_WES <- anno_WES %>%
      filter(!is.na(SBS4_SBS92))  %>%
      mutate(SBS4_SBS92_dicho = ifelse(SBS4_SBS92 > cutoff, "high","low"))


# diff abundance
anno <- anno_WES
p.list <- list()

for (i in data_type) {

  # get cell abundances and remove NAs
  abundances <- dat %>%
    dplyr::select(contains(i)) %>%
    filter(complete.cases(.))
  abundances <- t(abundances)
  colnames(abundances) <-dat %>%
    dplyr::select(contains(i),immucan_id) %>%
    filter(complete.cases(.)) %>%
    pull(immucan_id)
  
  abundances <- as.data.frame(abundances[,colnames(abundances) %in% anno$immucan_id])
   
  meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
  stopifnot(all(colnames(abundances)==meta$immucan_id))
    
  Size <- factor(meta[,"SBS4_SBS92_dicho"], levels = c("low", "high"))

  # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
  y.ab <- DGEList(counts = abundances, group = Size)
  design <- model.matrix(~0+Size)
  colnames(design) <- levels(Size)
    
  # we do not fit a trend 
  y.ab <- estimateDisp(y.ab, design, trend.method = "none")
  y.ab$common.dispersion

  fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
  fit_1_2 <- glmLRT(glmfit = fit.ab,contrast = c(-1,1))

  res_list <- list(fit_1_2)
    
  cur_list <- lapply(res_list,function(x){
      sig_dat <- x$table
      sig_dat$celltypes <- rownames(sig_dat)
      sig_dat$comparison <- gsub("Size","",x$comparison)
      return(sig_dat)
    })
  p.list[[i]] <- do.call(rbind,cur_list)
}

cur_dat.SBS <- do.call(rbind,p.list)
cur_dat.SBS <-cur_dat.SBS %>%
  mutate(histology = "all")
```

### plot together
```{r, fig.height=10, fig.width=20}
cur_dat <- rbind(cur_dat_LUAD,cur_dat_LUSC,cur_dat.SBS)

cur_dat <- cur_dat %>%
  mutate(tech = gsub("_density","",str_split_i(celltypes,"__",2)),
         cond = ifelse(comparison == "1*Current smoker -1*Ex-smoker","CurrentSmoker_over_ExSmoker","ExSmoker_over_NonSmoker"),
         cond = ifelse(comparison == "1*Current smoker -1*Non-smoker","CurrentSmoker_over_NonSmoker",cond),
         cond = paste0(histology,"_",cond),
         cond = ifelse(comparison == "-1*low 1*high","SBS4-SBS92_High_over_Low",cond)) %>%
  group_by(tech) %>%
  mutate(FDR = p.adjust(PValue,method = "BH"),
         sig = ifelse(PValue < 0.05 & logFC > 0, "UP_trend",NA),
         sig = ifelse(FDR < 0.1 & logFC > 0, "UP",sig),
         sig = ifelse(PValue < 0.05 & logFC < 0, "DOWN_trend",sig),
         sig = ifelse(FDR < 0.1 & logFC < 0, "DOWN",sig),
         sig = ifelse(is.na(sig),"NOTSIG", sig),
         ct_name = gsub("_density","",str_split_i(celltypes,"__",1)))
  
to_show <- cur_dat %>%
  filter(sig %in% c("UP","DOWN")) %>%
  pull(ct_name)

cur_dat <- cur_dat %>%
  filter(ct_name %in% to_show) %>%
  mutate(celltypes = gsub("_abundance","",celltypes))

# get max and min fold change values
max_up <- max(cur_dat$logFC)+1
min_down <- min(cur_dat$logFC)-1

# plot the data
p <- cur_dat %>%
  ggplot(aes(x=celltypes, y = factor(cond, levels = c("SBS4-SBS92_High_over_Low",
                                                      "LUAD_CurrentSmoker_over_ExSmoker",
                                                      "LUAD_ExSmoker_over_NonSmoker",
                                                      "LUAD_CurrentSmoker_over_NonSmoker",
                                                      "LUSC_CurrentSmoker_over_ExSmoker")),col = logFC, size= -log10(PValue)))+
  #geom_tile(aes(fill = sig))+
  #theme_bw()+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(
    colours = c("blue", "white", "red"), 
    values = rescale(c(round(min_down), 0, round(max_up))),
    limits = c(round(min_down), round(max_up)),
    guide = guide_colorbar(
      title = "logFC",                   
      #barwidth = 2,                     
      #barheight = 1,                   
      title.theme = element_text(size = 12),
      label.theme = element_text(size = 12)
    )
  )+
  scale_size(range = c(1,4))+
  new_scale_colour() +
  geom_point(data = cur_dat,shape = 1,aes(size = -log10(PValue), col = sig), stroke =1.5)+
  scale_color_manual(values = c("UP" = "#9f1214","UP_trend" = "#e73032","DOWN" = "#25567d","DOWN_trend" = "#428bc7","NOTSIG" = "snow2"), name = "stats")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 12),
        axis.text.y = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 12),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        legend.location = "plot",
        legend.position = "right")
p

#ggsave(filename = paste0(out_folder,"2F_diff_abundance_smoking.pdf"),
#       plot = p,
#       device = "pdf",
#       width = 16,
#       height = 5)
```

