---
title: "Figure_2"
author: "Daniel Schulz"
date: "2023-10-18"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

```{r}
library(tidyverse)
library(ggpubr)
library(SingleCellExperiment)
library(survivalAnalysis)
library(ggsurvfit)
library(survminer)
library(survival)
library(edgeR)
library(cowplot)
library(ggrepel)
library(igraph)
library(ggraph)
library(ggrepel)
library(scales)
library(ggnewscale)
library(imcRtools)
```

```{r}
mount_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/"
out_folder <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/revised_figures/Figure_2/"

if (!dir.exists(out_folder)) {
  dir.create(out_folder)
}

source("code/lollipop_function.R")
```

# load data
we will load the data we generated at the beginning of figure 2
```{r}
dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/abundances_dat.rds"))
dens_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/densities_dat.rds"))
cello_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/cello_dat.rds"))
clin_dat <- read.csv("/mnt/immucan_volume/clinical_data/Lung_NSCLC2/20230825/NSCLC2_for_R_update_2024.csv")
traits <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/all_traits.csv"))
modules <- as.data.frame(readRDS( "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/patient_imaging_meta_cluster_scores.rds"))

rna <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/rna_scores.rds")
rna$immucan_id  <- rownames(rna)
rownames(rna) <- NULL

modules <- modules  %>%
  mutate(immucan_id = rownames(modules))
rownames(modules) <- NULL
# variables
data_type <- unique(str_split_i(string = colnames(dat[,-1]),pattern = "__",i = 2))
```

# 2A Imaging modules

code is here: [imaging module plot](../../../imaging_modules/imaging_module_generation.Rmd)

# Histology

## 2B RNA scores

```{r, fig.height=15, fig.width=15}
left_join(rna, traits, "immucan_id") %>%
  filter(!is.na(LUAD_vs_LUSC)) %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC")) %>%
  pivot_longer(cols = 1:29) %>%
  ggplot(aes(x = LUAD_vs_LUSC,y = value))+
  geom_boxplot()+
  facet_wrap(.~name, scales = "free")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  ggpubr::stat_compare_means()

dat_z <- left_join(rna, traits, "immucan_id") %>%
  filter(!is.na(LUAD_vs_LUSC)) %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC")) %>%
  pivot_longer(cols = 1:29)

library(dplyr)
library(tidyr)
library(purrr)
library(broom)

volcano_df <- dat_z %>%
  group_by(name) %>%
  mutate(value_z = as.numeric(scale(value))) %>%  # z-score within each signature
  summarise(
    delta_z = mean(value_z[LUAD_vs_LUSC == "LUAD"], na.rm = TRUE) -
              mean(value_z[LUAD_vs_LUSC == "LUSC"], na.rm = TRUE),
    pvalue  = t.test(value_z[LUAD_vs_LUSC == "LUAD"],
                      value_z[LUAD_vs_LUSC == "LUSC"])$p.value,
    .groups = "drop"
  ) %>%
  mutate(
    padj        = p.adjust(pvalue, method = "BH"),
    neg_log10_p = -log10(pvalue),
    direction   = ifelse(delta_z > 0, "Up in LUAD", "Up in LUSC"),
    sig = ifelse(padj < 0.1, TRUE, FALSE)
  ) %>%
  arrange(delta_z)


volcano_df <- volcano_df %>%
  mutate(name = factor(name, levels = name[order(delta_z)]))  # order for plotting

p <- ggplot(volcano_df, aes(x = delta_z, y = name)) +
  geom_segment(aes(x = 0, xend = delta_z, y = name, yend = name), color = "grey70") +
  geom_point(aes(color = direction, size = -log10(padj))) +
  geom_point(
    data = subset(volcano_df, sig == TRUE),
    aes(x = delta_z, y = name, size = -log10(padj)),
    shape = 21, color = "black", fill = NA, stroke = 1
  )+
  scale_color_manual(values = c("Up in LUAD" = "blue", "Up in LUSC" = "orange")) +
  geom_vline(xintercept = 0, linetype = "dashed", color = "grey40") +
  labs(x = "Δ z-score (LUAD - LUSC)", y = NULL,
       color = "Direction", size = "-log10(FDR)") +
  guides(
    color = guide_legend(nrow = 1, title.position = "top"),
    size  = guide_legend(nrow = 1, title.position = "top")
  ) +
  theme_minimal() +
  theme(
    axis.title.x = element_text(size = 10),
    axis.text.y     = element_text(size = 9),
    legend.position = "bottom",
    legend.box      = "vertical",     # stacks the color legend and size legend on top of each other
    legend.margin   = margin(t = 0),
    legend.spacing  = unit(0.2, "cm")
  )+
  ggtitle("RNA signatures")
p
ggsave(plot = p, path = out_folder,filename = "RNA_difference_for_histology.pdf",width = 4.5, height = 7,device = cairo_pdf)
```


## 2C imaging modules for histology
```{r, fig.height=10, fig.width=8}
left_join(modules, traits, "immucan_id") %>%
  filter(!is.na(LUAD_vs_LUSC)) %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC")) %>%
  pivot_longer(cols = 1:8) %>%
  ggplot(aes(x = LUAD_vs_LUSC,y = value))+
  geom_boxplot()+
  facet_wrap(.~name, scales = "free")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  ggpubr::stat_compare_means(label.y = )

dat_z <- left_join(modules, traits, "immucan_id") %>%
  filter(!is.na(LUAD_vs_LUSC)) %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC")) %>%
  pivot_longer(cols = 1:8)

result <- lollipop_plot(dat_z, "LUAD_vs_LUSC", "LUAD", "LUSC")
p <- result$plot
p

ggsave(plot = p, path = out_folder,filename = "Imaging_modules_for_histology.pdf",width = 6, height = 3.5,device = cairo_pdf)
```

## to  be removed
```{r}
left_join(modules, traits, "immucan_id") %>%
  filter(!is.na(TPS_score)) %>%
  pivot_longer(cols = 1:8) %>%
  ggplot(aes(x = TPS_score,y = value))+
  geom_boxplot()+
  facet_wrap(.~name)+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  ggpubr::stat_compare_means(method = "t.test")

left_join(modules, traits, "immucan_id") %>%
  filter(!is.na(TMB_score)) %>%
  pivot_longer(cols = 1:8) %>%
  ggplot(aes(x = TMB_score,y = value))+
  geom_boxplot()+
  facet_wrap(.~name)+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  ggpubr::stat_compare_means(method = "t.test")

left_join(modules, traits, "immucan_id") %>%
  filter(!is.na(EGFR),
         LUAD_vs_LUSC == "Adenocarcinoma") %>%
  pivot_longer(cols = 1:8) %>%
  ggplot(aes(x = EGFR,y = value))+
  geom_boxplot()+
  facet_wrap(.~name)+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  ggpubr::stat_compare_means(method = "t.test")

left_join(modules, traits, "immucan_id") %>%
  filter(!is.na(KRAS),
         LUAD_vs_LUSC == "Adenocarcinoma") %>%
  pivot_longer(cols = 1:8) %>%
  ggplot(aes(x = KRAS,y = value))+
  geom_boxplot()+
  facet_wrap(.~name)+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  ggpubr::stat_compare_means(method = "t.test")

left_join(modules, traits, "immucan_id") %>%
  filter(!is.na(simple_stage),
         simple_stage != "IV") %>%
  pivot_longer(cols = 1:8) %>%
  ggplot(aes(x = simple_stage,y = value))+
  geom_boxplot()+
  facet_wrap(.~name)+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  ggpubr::stat_compare_means()

```

```{r}
left_join(modules, traits, "immucan_id") %>%
  left_join(inf_dat,"immucan_id") %>%
  mutate(inf_score = as.factor(ntile(x = mean_inf_ratio, n = 2))) %>%
  filter(!is.na(inf_score)) %>%
  pivot_longer(cols = 1:8) %>%
  select(immucan_id, name, value, inf_score) %>%
  ggplot(aes(x = inf_score,y = value))+
  geom_boxplot()+
    facet_wrap(.~name)+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  ggpubr::stat_compare_means()
```




# 2A diff abundance histology - move to supplements
here we will estimate the dispersion across the whole data set and then do the linear model of count comparison only for the groups of interest.
```{r, fig.height=10, fig.width=10}
condition <- c("Adenocarcinoma")
names(condition) <- c("LUAD_vs_LUSC")
j <- names(condition)
p.list <- list()
for (i in data_type) {

  # get cell abundances and remove NAs
  abundances <- dat %>%
    select(contains(i)) %>%
    filter(complete.cases(.))
  abundances <- t(abundances)
  colnames(abundances) <-dat %>%
    select(contains(i),immucan_id) %>%
    filter(complete.cases(.)) %>%
    pull(immucan_id)
  
  # merge with clinical data
  anno <- clin_dat %>%
    filter(!is.na(.data[["simple_histology"]]))
  
  abundances <- abundances[,colnames(abundances) %in% anno$immucan_id]
  
  meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
  Size <- factor(meta[,"simple_histology"])
  meta$Size <- Size
  
  # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
  y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
  design <- model.matrix(~0+group,data = y.ab$samples)
  colnames(design) <- levels(y.ab$samples$group)
  # we do not fit a trend 
  y.ab <- estimateDisp(y.ab, design, trend.method = "none")
  y.ab$common.dispersion
  #plotBCV(y.ab,pch = 5)
  
  # we only compare the diff abundance for Adeno vs Sqam.
  fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
  fit <- glmLRT(glmfit = fit.ab,contrast = c(-1,0,0,0,1,0))
  #plotQLDisp(fit.ab, pch = 5)
  
  sig_dat <- fit$table
  sig_dat$celltypes <- rownames(sig_dat)
  sig_dat$cond <- j
  sig_dat$comparison <- gsub("Size","",fit$comparison)
  p.list[[i]] <- sig_dat
}   

# fold change cutoff
cutoff <- 0.7

cur_dat <- do.call(rbind,p.list)
cur_dat <- cur_dat %>%
  mutate(tech = gsub("_abundance","",str_split_i(celltypes,"__",2))) %>%
  group_by(tech) %>%
  mutate(cond = paste0(cond,"_comparison: ",comparison),
         FDR = p.adjust(PValue,method = "BH"),
         sig_overall = ifelse(FDR < 0.1 ,TRUE,FALSE),
         sig = ifelse(FDR < 0.1 & abs(logFC) > cutoff ,TRUE,FALSE),
         direction = ifelse(logFC > 0 ,"up", "down"),
         ct_name = gsub("_abundance","",str_split_i(celltypes,"__",1)))
  

p <- cur_dat %>%
  ggplot(aes(x = logFC,y = -log10(PValue), color = tech))+
  geom_point()+
  geom_label_repel(data = cur_dat %>% filter(sig == TRUE),
                  aes(x = logFC, y = -log10(PValue),label = ct_name),
                  max.overlaps = 20,
                  min.segment.length = 0,
                  nudge_x = 0.5,nudge_y = 0.05)+
    scale_color_manual(values = c("IF1" = "blue", "IF2" = "deepskyblue","IF3" = "steelblue","IMC1" ="red", "IMC2" ="darkorange"))+
    theme_bw()+
  coord_cartesian(xlim = c(-2,4),ylim = c(0,7))+
  geom_vline(xintercept = c(-cutoff, cutoff), lwd = 0.7, col = "gray", linetype = "dashed")+
  ggtitle("Cell type abundance enrichment for histology")+
  theme(axis.text.x = element_text(size = 15),
        axis.text.y = element_text(size = 15),
        axis.title.x = element_text(size = 20),
        axis.title.y = element_text(size = 20),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 15),
        plot.title = element_text(hjust = 0.5, size = 16, face = "bold"))+
  guides(size = guide_legend(override.aes = list(size = 10)))

p
ggsave(filename = paste0(out_folder,"2A_diff_abundance_histology.pdf"),
       plot = p,
       device = "pdf",
       width = 10,
       height = 8)
```

```{r}
dens_dat %>%
  left_join()
colSums(cello_dat[,grepl("cello1",colnames(cello_dat))])/sum(cello_dat[,grepl("cello1",colnames(cello_dat))])
```


# 2B diff abundance cellohood for histology - move to supplements

```{r}
condition <- c("Adenocarcinoma")
names(condition) <- c("LUAD_vs_LUSC")
j <- names(condition)
p.list <- list()
for (i in c("cello1","cello2")) {

    abundances <- cello_dat %>%
      select(contains(i))
    abundances <- t(abundances)
    colnames(abundances) <- cello_dat$immucan_id
    anno <- clin_dat %>%
      filter(!is.na(.data[["simple_histology"]]))

    abundances <- abundances[,colnames(abundances) %in% anno$immucan_id]

    meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
    Size <- factor(meta[,"simple_histology"])
    meta$Size <- Size
    
    # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
    y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
    design <- model.matrix(~0+group,data = y.ab$samples)
    colnames(design) <- levels(y.ab$samples$group)
    # we do not fit a trend 
    y.ab <- estimateDisp(y.ab, design, trend.method = "none")
    y.ab$common.dispersion
    #plotBCV(y.ab,pch = 5)
    
    # we only compare the diff abundance for Adeno vs Sqam.
    fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
    fit <- glmLRT(glmfit = fit.ab,contrast = c(-1,0,0,0,1,0))
    #plotQLDisp(fit.ab, pch = 5)
    
    sig_dat <- fit$table
    sig_dat$celltypes <- rownames(sig_dat)
    sig_dat$cond <- j
    sig_dat$comparison <- gsub("Size","",fit$comparison)
    p.list[[i]] <- sig_dat
}   
# fold change cutoff
cutoff <- 0.7 

cur_dat <- do.call(rbind,p.list)
cur_dat <- cur_dat %>%
  mutate(tech = gsub("_abundance","",str_split_i(celltypes,"__",2))) %>%
  group_by(tech) %>%
  mutate(cond = paste0(cond,"_comparison: ",comparison),
         FDR = p.adjust(PValue,method = "BH"),
         sig_overall = ifelse(FDR < 0.1 ,TRUE,FALSE),
         sig = ifelse(FDR < 0.1 & abs(logFC) > cutoff ,TRUE,FALSE),
         direction = ifelse(logFC > 0 ,"up", "down"),
         ct_name = gsub("_abundance","",str_split_i(celltypes,"__",1)))
  

cur_dat <- cur_dat %>%
  mutate(tech = gsub(pattern = "cello",replacement = "IMC",x = tech))

p <- cur_dat %>%
  ggplot(aes(x = logFC,y = -log10(PValue), color = tech))+
  geom_point()+
  geom_label_repel(data = cur_dat %>% filter(sig == TRUE),
                  aes(x = logFC, y = -log10(PValue),label = ct_name),
                  max.overlaps = 20,
                  min.segment.length = 0,
                  nudge_x = 0.5,nudge_y = 0.05)+
    scale_color_manual(values = c("IF1" = "blue", "IF2" = "deepskyblue","IF3" = "steelblue","IMC1" ="red", "IMC2" ="darkorange"))+
    theme_bw()+
  coord_cartesian(xlim = c(-2.5,5),ylim = c(0,12.5))+
  scale_color_manual(values = c("IMC1" = "green4", "IMC2" = "magenta3"))+
  geom_vline(xintercept = c(-cutoff, cutoff), lwd = 0.7, col = "gray", linetype = "dashed")+
  ggtitle("Cellohood cluster enrichment for histology")+
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 14),
        axis.title.y = element_text(size = 14),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 12),
        plot.title = element_text(hjust = 0.5, size = 11, face = "bold"))+
  guides(size = guide_legend(override.aes = list(size = 10)))

p
ggsave(filename = paste0(out_folder,"2B_diff_abundance_celloClusters_histology.pdf"),
       plot = p,
       device = "pdf",
       width = 9,
       height = 6)
```

# 2D images P1 CN_2
```{r, fig.height=12, fig.width=16}
sce1 <- readRDS(paste0(mount_path,"data/IMC/Panel_1/Rout/sce.rds"))
sce1 <- sce1[,sce1$ROI != 4]
counts <- as.data.frame(t(assay(sce1,"exprs")))

# identify images with high amuounts of differentially abundant cello clusters
#cello_dat %>% 
#  left_join(clin_dat,"immucan_id") %>%
#  select(immucan_id,CN_13__cello1, CN_2__cello1, CN_10__cello1, simple_histology)

#LUNG-NSCLC2-0640|0688 # LUSC high for CN_13 -> rare cluster 
#LUNG-NSCLC2-0567|0617 # high for CN_2
#LUNG-NSCLC2-0689|0582 # high for CN_10
#LUNG-NSCLC2-0699|0588 # high for CN_28

cur_sce <- sce1[,grepl("0640|0688|0567|0617|0689|0582|0699|0588",sce1$immucan_id)]


plotSpatial(cur_sce[,grepl("0567|0617", cur_sce$immucan_id)],img_id = "sample_id",node_color_by = "celltypes_minor")
plotSpatial(cur_sce[,grepl("0567|0617", cur_sce$immucan_id)],img_id = "sample_id",node_color_by = "cellohood")

library(cytomapper)
masks <- readRDS(paste0(mount_path,"data/IMC/Panel_1/Rout/masks.rds"))
images <- readRDS(paste0(mount_path,"data/IMC/Panel_1/Rout/images.rds"))

cur_df <- as.data.frame(colData(sce1))
rownames(cur_df) <- NULL
cur_df <- cur_df %>%
  select(immucan_id,image, sample_id) %>%
  distinct() %>%
  mutate(image = gsub(".tiff","",image))
rownames(cur_df) <- cur_df$image

mcols(images)$immucan_id <- cur_df[names(images),"immucan_id"]
mcols(masks)$immucan_id <- cur_df[names(masks),"immucan_id"]
mcols(images)$sample_id <- cur_df[names(images),"sample_id"]
mcols(masks)$sample_id <- cur_df[names(masks),"sample_id"]

plot_sce <- cur_sce[,cur_sce$immucan_id == "LUNG-NSCLC2-0617" & cur_sce$ROI == 1]
channelNames(images) <- rownames(plot_sce)
im_sel <- unique(plot_sce$sample_id)

cur_ims <- getImages(images, im_sel)
cur_masks <- getImages(masks, im_sel)

cur_images <- CytoImageList(cur_ims, on_disk = FALSE)
cur_masks <- CytoImageList(cur_masks, on_disk = FALSE)
cur_images <- scaleImages(cur_images,value = 2^16-1)
cur_masks <- scaleImages(cur_masks,value = 2^16-1)

cur_images[[1]]
cur_images <- normalize(cur_images, separateImages = TRUE)
cur_images <- normalize(cur_images, separateImages = TRUE, 
                        inputRange = c(0, 0.2))



plot_sce$CN_2 <- "other"
plot_sce[,plot_sce$cellohood == "CN_2"]$CN_2 <- "CN_2"

plotCells(mask = cur_masks,
          object = plot_sce,
          cell_id = "ObjectNumber",
          img_id = "sample_id",
          image_title = NULL,
          colour_by = "celltypes",colour = list(celltypes = metadata(plot_sce)$colour_vectors$celltypes),
          legend = list(colour_by.labels.cex = 0.1),
                    scale_bar = list(length = 100,
                            label = expression("100 " ~ mu * "m"),
                            cex = 1.2,
                            lwidth = 5),
           save_plot = list(filename = paste0(out_folder,"LUNG-NSCLC2-0617_ROI_01_celltypes.png")))
plotCells(mask = cur_masks,
          object = plot_sce,
          cell_id = "ObjectNumber",
          img_id = "sample_id",
          colour_by = "CN_2",
          colour = list(CN_2 = c(CN_2 = "red",
                                 other = "gray")),
          image_title = NULL,
          legend = list(colour_by.labels.cex = 0.7),
          scale_bar = list(length = 100,
                            label = expression("100 " ~ mu * "m"),
                            cex = 1.2,
                            lwidth = 5),
           save_plot = list(filename = paste0(out_folder,"LUNG-NSCLC2-0617_ROI_01_CN_2.png")))

plotPixels(image = cur_images,
           
           object = plot_sce,
           cell_id = "ObjectNumber",
           img_id = "sample_id",
           colour_by = c("PDL1", "DNA1", "Ecad"),
           image_title = NULL,
           colour = list(PDL1 = c("black","red"),
                         DNA1 = c("black","blue"),
                         Ecad = c("black","green")),
           bcg = list(PDL1 = c(0,4,1)),
           scale_bar = list(length = 100,
                            label = expression("100 " ~ mu * "m"),
                            cex = 1.2,
                            lwidth = 5),
           save_plot = list(filename = paste0(out_folder,"LUNG-NSCLC2-0617_ROI_01_PDL1_expr.png")))
```


# 2E images P1 CN_2
```{r, fig.height=12, fig.width=16}
sce2 <- readRDS(paste0(mount_path,"data/IMC/Panel_2/Rout/sce.rds"))

# identify images with high amuounts of differentiall abundant cello clusters
#cello_dat %>% 
#  left_join(clin_dat,"immucan_id") %>%
#  select(immucan_id,CN_9__cello2, simple_histology) %>%
#  arrange(desc(CN_9__cello2))

#LUNG-NSCLC2-0720|0729 # LUSC high for CN_9 -> rare cluster 

cur_sce <- sce2[,grepl("0729",sce2$immucan_id)]

plotSpatial(cur_sce,img_id = "sample_id",node_color_by = "celltypes_minor")
plotSpatial(cur_sce,img_id = "sample_id",node_color_by = "cellohood")

masks <- readRDS(paste0(mount_path,"data/IMC/Panel_2/Rout/masks.rds"))
images <- readRDS(paste0(mount_path,"data/IMC/Panel_2/Rout/images.rds"))

cur_df <- as.data.frame(colData(sce2))
rownames(cur_df) <- NULL
cur_df <- cur_df %>%
  select(immucan_id,image, sample_id) %>%
  distinct() %>%
  mutate(image = gsub(".tiff","",image))
rownames(cur_df) <- cur_df$image

mcols(images)$immucan_id <- cur_df[names(images),"immucan_id"]
mcols(masks)$immucan_id <- cur_df[names(masks),"immucan_id"]
mcols(images)$sample_id <- cur_df[names(images),"sample_id"]
mcols(masks)$sample_id <- cur_df[names(masks),"sample_id"]

plot_sce <- cur_sce[,cur_sce$ROI == 3]
im_sel <- unique(plot_sce$sample_id)

cur_ims <- getImages(images, im_sel)
cur_masks <- getImages(masks, im_sel)

channelNames(cur_ims) <- rownames(plot_sce)

cur_images <- CytoImageList(cur_ims, on_disk = FALSE)
cur_masks <- CytoImageList(cur_masks, on_disk = FALSE)
cur_images <- scaleImages(cur_images,value = 2^16-1)
cur_masks <- scaleImages(cur_masks,value = 2^16-1)

cur_images <- normalize(cur_images, separateImages = TRUE)
cur_images <- normalize(cur_images, separateImages = TRUE, 
                        inputRange = c(0, 0.2))


plot_sce$CN_9 <- "other"
plot_sce[,plot_sce$cellohood == "CN_9"]$CN_9 <- "CN_9"

plotCells(mask = cur_masks,
          object = plot_sce,
          cell_id = "ObjectNumber",
          img_id = "sample_id",
          image_title = NULL,
          colour_by = "celltypes",
          colour = list(celltypes = metadata(plot_sce)$colour_vectors$cell_types),
          legend = list(colour_by.labels.cex = 0.1),
          scale_bar = list(length = 100,
                           label = expression("100 " ~ mu * "m"),
                           cex = 1.2,
                           lwidth = 5),
          save_plot = list(filename = paste0(out_folder,"LUNG-NSCLC2-0729_ROI_01_celltypes.png")))
plotCells(mask = cur_masks,
          object = plot_sce,
          cell_id = "ObjectNumber",
          img_id = "sample_id",
          colour_by = "CN_9",
          colour = list(CN_9 = c(CN_9 = "red",
                                 other = "gray")),
          image_title = NULL,
          legend = list(colour_by.labels.cex = 0.7),
          scale_bar = list(length = 100,
                            label = expression("100 " ~ mu * "m"),
                            cex = 1.2,
                            lwidth = 5),
           save_plot = list(filename =  paste0(out_folder,"LUNG-NSCLC2-0729_ROI_01_CN_9.png")))

plotPixels(image = cur_images,
           object = plot_sce,
           cell_id = "ObjectNumber",
           img_id = "sample_id",
           colour_by = c("CD73", "DNA1", "Ecad"),
           image_title = NULL,
           colour = list(CD73 = c("black","red"),
                         DNA1 = c("black","blue"),
                         Ecad = c("black","green")),
           bcg = list(CD73 = c(0,4,1)),
           scale_bar = list(length = 100,
                            label = expression("100 " ~ mu * "m"),
                            cex = 1.2,
                            lwidth = 5),
           save_plot = list(filename = paste0(out_folder,"LUNG-NSCLC2-0729_ROI_01_CD73_expr.png")))
```




# linear model densities for stages both histologies - to be removed
not in paper
```{r, fig.height=10, fig.width=20}
library(purrr)
library(broom)
sigs <- dens_dat %>%
  pivot_longer(-immucan_id) %>%
    left_join(clin_dat,"immucan_id") %>%
  select(immucan_id,name, value,simple_stage) %>%
  mutate(stage = ifelse(simple_stage == "I",1,NA),
         stage = ifelse(simple_stage == "II",2,stage),
         stage = ifelse(simple_stage == "III",3,stage),
         stage = ifelse(simple_stage == "IV",4,stage),
         stage = as.numeric(as.factor(stage))) %>%
  filter(stage != 4) %>%
  group_by(name) %>%
  nest() %>%
  mutate(model = map(data, ~ lm(value ~ stage, data = .x)),
         summary = map(model, broom::tidy)) %>%
  select(name, summary) %>%
  unnest(c(summary)) %>%
  filter(term == "stage") %>%
  mutate(tech = str_split_i(name,"__",2)) %>%
  group_by(tech) %>%
  mutate(FDR = p.adjust(p = p.value,method = "BH"),
         sig = ifelse(FDR < 0.1,TRUE, FALSE))

FDR_sig <- sigs %>%
  filter(sig == TRUE) %>%
  pull(name)
p_sig <- sigs %>%
  filter(p.value < 0.05) %>%
  pull(name)

p_sig_dat <- sigs %>%
  filter(sig ==TRUE) %>%
  mutate(name = gsub("_density_","",name))
  

p_sig <- c(p_sig,"NKT_density__IF3")
p <- dens_dat %>%
  pivot_longer(-immucan_id) %>%
  mutate(tech = str_split_i(name,"__",2),
         name = gsub("_density_","",name)) %>%
  left_join(clin_dat,"immucan_id") %>%
  select(immucan_id,name, value,simple_stage, tech) %>%
  mutate(stage = ifelse(simple_stage == "I",1,NA),
         stage = ifelse(simple_stage == "II",2,stage),
         stage = ifelse(simple_stage == "III",3,stage),
         stage = ifelse(simple_stage == "IV",4,stage),
         stage = as.numeric(as.factor(stage))) %>%
  filter(name %in% gsub("_density_","",FDR_sig),
         stage != 4) %>%
  ggplot(aes(x = name,y = log10(value+1),fill=as.factor(stage))) +
  geom_boxplot(outliers = FALSE)+
  geom_jitter(position = position_jitterdodge(jitter.width = 0.2),size= 0.5, show.legend = FALSE)+
  scale_fill_manual(values = c("palevioletred3", "darkseagreen3", "dodgerblue"))+
  theme_bw()+
  geom_text(data = p_sig_dat,
            aes(x = name,
                y=3.5,
                label = paste0("p : ",round(p.value,digits = 3))),
            inherit.aes = FALSE,
            size = 6)+
  coord_cartesian(ylim = c(0,3.5))+
  theme(axis.text.x = element_text(angle = 90,vjust= 0.5, hjust = 1, size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 16),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 15))+
  guides(fill = guide_legend(override.aes = list(size = 8,
                                                 shape = 22)))+
  labs(y = expression("Log10(density)"),
       fill = "Stage")

p 

```


# 2F tumor types imaging modules
we will not subtype LUSC samples because we have too few samples
```{r}
ttypes <- read.csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/tumor_subtypes.csv")

dat_z <- ttypes %>% left_join(modules, "immucan_id")

dat_z$immucan_id <- NULL
luad_dat <- dat_z %>%
  filter(subtype %in% c("bronchioid","squamoid","magnoid")) %>%
  mutate(across(.cols = -subtype,scale)) %>%
  group_by(subtype) %>%
  summarise(across(.fns = mean))
  
lusc_dat <- dat_z %>%
  filter(subtype %in% c("classical","primitive","basal", "secretory")) %>%
  mutate(across(.cols = -subtype,scale)) %>%
  group_by(subtype) %>%
  summarise(across(.fns = mean))

luad_mat <- as.matrix(luad_dat[,-1])
rownames(luad_mat) <- luad_dat$subtype

lusc_mat <- as.matrix(lusc_dat[,-1])
rownames(lusc_mat) <- lusc_dat$subtype


h1 <- ComplexHeatmap::Heatmap(t(lusc_mat))
h2 <- ComplexHeatmap::Heatmap(t(luad_mat))
h1+h2
# annotation
ann_df <- data.frame(Subtype = c(unique(luad_dat$subtype)),
                     subtype_col = c( "blue","deepskyblue","dodgerblue2"))
ha <- HeatmapAnnotation(subtype = ann_df$Subtype,
                        col = list(subtype = structure(ann_df$subtype_col,names = ann_df$Subtype)))

h <- ComplexHeatmap::Heatmap(t( as.matrix(luad_dat[,-1])),
                             name = "mean z_score",
                             top_annotation = ha)


pdf(file = paste0(out_folder,"imaging_modules_for_tumor_subtypes_LUAD.pdf"), width = 4, height = 4)
draw(h, merge_legend = TRUE)
dev.off()


## this code generate the two heatmaps for LUAD and LUSC together. but we will exclude LUSC because we have too few samples
# plot_dat <- dat_z %>%
#   mutate(across(.cols = -subtype,scale)) %>%
#   group_by(subtype) %>%
#   summarise(across(.fns = mean))
# 
# ann_df <- data.frame(Subtype = c(unique(plot_dat$subtype)),
#                      Histology = c("LUSC","LUAD","LUSC","LUAD","LUSC","LUSC","LUAD"),
#                      subtype_col = c("darkgoldenrod1", "lightskyblue","yellow2","deepskyblue","orange3", "burlywood2","dodgerblue2"),
#                      hist_col = c("orange", "blue", "orange", "blue","orange","orange","blue"))
# mat <- as.matrix(plot_dat[,-1])
# rownames(mat) <- plot_dat$subtype
# ha <- HeatmapAnnotation(subtype = ann_df$Subtype,
#                         Histology = ann_df$Histology,
#                         col = list(subtype = structure(ann_df$subtype_col,names = ann_df$Subtype),
#                                    Histology = structure(ann_df$hist_col,names = ann_df$Histology)))
# h <- ComplexHeatmap::Heatmap(t(mat),
#                              name = "mean z_score",
#                              top_annotation = ha,column_split = ann_df$Histology)
# 
# pdf(file = paste0(out_folder,"imaging_modules_for_tumor_subtypes_complex.pdf"), width = 5, height = 4)
# draw(h, merge_legend = TRUE,
#                       heatmap_legend_side = "right",
#                       annotation_legend_side = "right")
# dev.off()
```



# 2G tumor types for stages

```{r}
scores <- read.csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/tumor_subtype_scores.csv")

scores <- left_join(scores, traits,"immucan_id")

p <- scores %>%
  pivot_longer(cols = 1:3) %>% 
  filter(simple_stage != "IV",
         !is.na(simple_stage)) %>%
  ggplot(aes(x = simple_stage, y = value, fill = simple_stage))+
  geom_boxplot()+
  geom_jitter(aes(col = simple_stage),position = position_jitterdodge(jitter.width = 0.2),size= 0.5)+
  scale_fill_manual(values = alpha(c("I" = "palevioletred3", "II" = "darkseagreen3", "III" = "dodgerblue"),alpha = 0.6))+
  scale_color_manual(values = c("I" = "palevioletred3", "II" = "darkseagreen3", "III" = "dodgerblue"))+
  facet_wrap(~name)+
  stat_compare_means(aes(group = simple_stage),
                     comparisons = list(c("I", "II"), c("I","III"), c("II", "III")),
                     method = "t.test")+
  theme_bw()+
  coord_cartesian(ylim = c(-0.7, 1.5))+
  xlab("Stage")+
  ylab("Tumor type score")

ggsave(plot = p, path = out_folder,filename = "tumor_types_for_stages_LUAD.pdf",width = 4, height = 3.5,device = cairo_pdf)

```


# Stage
## 2H imaging modules
```{r}
dat_z <- left_join(modules, traits, "immucan_id") %>%
  filter(!is.na(simple_stage),
         simple_histology == "Adenocarcinoma") %>%
  pivot_longer(cols = 1:8)

sigs <- dat_z %>%
  select(immucan_id,name, value,simple_stage) %>%
  mutate(stage = ifelse(simple_stage == "I",1,NA),
         stage = ifelse(simple_stage == "II",2,stage),
         stage = ifelse(simple_stage == "III",3,stage),
         stage = ifelse(simple_stage == "IV",4,stage),
         stage = as.numeric(as.factor(stage))) %>%
  filter(stage != 4) %>%
  group_by(name) %>%
  nest() %>%
  mutate(model = map(data, ~ lm(value ~ stage, data = .x)),
         summary = map(model, broom::tidy)) %>%
  select(name, summary) %>%
  unnest(c(summary)) %>%
  filter(term == "stage") %>%
  mutate(tech = str_split_i(name,"__",2)) %>%
  group_by(tech) %>%
  mutate(FDR = p.adjust(p = p.value,method = "BH"),
         sig_p = ifelse(p.value < 0.05,TRUE, FALSE),
         sig_fdr = ifelse(FDR < 0.1,"T", "F"))


summary_dat <- sigs %>%
  #filter(sig_p == TRUE) %>%
  #filter(grepl("CD8_GATA3_BCL2|^CD8_density|CD8_Ki67_density|CD8_TCF7_density|Tumor_IDO1",name)) %>%
  mutate(y_position = 1000,
         name = gsub(pattern = "density_",replacement = "",x = name)) %>%
  ungroup() %>%
  as.data.frame()

p <- dat_z %>%
  select(immucan_id,name, value,simple_stage) %>%
  mutate(stage = ifelse(simple_stage == "I",1,NA),
         stage = ifelse(simple_stage == "II",2,stage),
         stage = ifelse(simple_stage == "III",3,stage),
         stage = ifelse(simple_stage == "IV",4,stage),
         stage = as.numeric(as.factor(stage)),
         name = gsub(pattern = "density_",replacement = "",x = name)) %>%
  filter(name %in% summary_dat$name,
         stage != 4) %>%
  ggplot(aes(x = name,y = log10(value+1),fill=as.factor(stage))) +
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(col = as.factor(stage)),position = position_jitterdodge(jitter.width = 0.2),size= 0.5)+
  scale_fill_manual(values = alpha(c("1" = "palevioletred3", "2" = "darkseagreen3", "3" = "dodgerblue"),alpha = 0.6))+
  scale_color_manual(values = c("1" = "palevioletred3", "2" = "darkseagreen3", "3" = "dodgerblue"))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90,vjust= 0.5, hjust = 1, size = 12),
        axis.title = element_blank())+
  guides(fill=guide_legend(title="Stage"),
         col=guide_legend(title="Stage"))+
  geom_text(data = summary_dat, aes(x = name,y=0.5,label = paste0("fdr: ",sig_fdr)), inherit.aes = FALSE, size = 2.5)+
  geom_text(data = summary_dat, aes(x = name,y=0.6,label = paste0("p : ",round(p.value,digits = 3))), inherit.aes = FALSE, size = 2.5)+
  coord_cartesian(ylim = c(-.55,0.6))+
  ggtitle(label = "Imaging module differences across stages in LUAD")
  
p

ggsave(plot = p, path = out_folder,filename = "Imaging_modules_for_stage_LUAD.pdf",width = 6, height = 3.5,device = cairo_pdf)


```


## linear model densities for stages - LUAD only - to be removed

```{r, fig.height=10, fig.width=30}
library(purrr)
library(broom)

sigs <- dens_dat %>%
  pivot_longer(-immucan_id) %>%
    left_join(clin_dat,"immucan_id") %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  select(immucan_id,name, value,simple_stage) %>%
  mutate(stage = ifelse(simple_stage == "I",1,NA),
         stage = ifelse(simple_stage == "II",2,stage),
         stage = ifelse(simple_stage == "III",3,stage),
         stage = ifelse(simple_stage == "IV",4,stage),
         stage = as.numeric(as.factor(stage))) %>%
  filter(stage != 4) %>%
  group_by(name) %>%
  nest() %>%
  mutate(model = map(data, ~ lm(value ~ stage, data = .x)),
         summary = map(model, broom::tidy)) %>%
  select(name, summary) %>%
  unnest(c(summary)) %>%
  filter(term == "stage") %>%
  mutate(tech = str_split_i(name,"__",2)) %>%
  group_by(tech) %>%
  mutate(FDR = p.adjust(p = p.value,method = "BH"),
         sig_p = ifelse(p.value < 0.05,TRUE, FALSE),
         sig_fdr = ifelse(FDR < 0.1,TRUE, FALSE))

summary_dat <- sigs %>%
  filter(sig_p == TRUE) %>%
  #filter(grepl("CD8_GATA3_BCL2|^CD8_density|CD8_Ki67_density|CD8_TCF7_density|Tumor_IDO1",name)) %>%
  mutate(y_position = 1000,
         name = gsub(pattern = "density_",replacement = "",x = name)) %>%
  ungroup() %>%
  as.data.frame()

p2 <- dens_dat %>%
  pivot_longer(-immucan_id) %>%
  mutate(tech = str_split_i(name,"__",2)) %>%
  left_join(clin_dat,"immucan_id") %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  select(immucan_id,name, value,simple_stage, tech) %>%
  mutate(stage = ifelse(simple_stage == "I",1,NA),
         stage = ifelse(simple_stage == "II",2,stage),
         stage = ifelse(simple_stage == "III",3,stage),
         stage = ifelse(simple_stage == "IV",4,stage),
         stage = as.numeric(as.factor(stage)),
         name = gsub(pattern = "density_",replacement = "",x = name)) %>%
  filter(name %in% summary_dat$name,
         stage != 4) %>%
  ggplot(aes(x = name,y = log10(value+1),fill=as.factor(stage))) +
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(col = as.factor(stage)),position = position_jitterdodge(jitter.width = 0.2),size= 0.5)+
  scale_fill_manual(values = alpha(c("1" = "palevioletred3", "2" = "darkseagreen3", "3" = "dodgerblue"),alpha = 0.6))+
  scale_color_manual(values = c("1" = "palevioletred3", "2" = "darkseagreen3", "3" = "dodgerblue"))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90,vjust= 0.5, hjust = 1),
        axis.title = element_blank())+
  guides(fill=guide_legend(title="Stage"),
         col=guide_legend(title="Stage"))+
  geom_text(data = summary_dat, aes(x = name,y=4.2,label = paste0("fdr: ",sig_fdr)), inherit.aes = FALSE)+
  geom_text(data = summary_dat, aes(x = name,y=4.6,label = paste0("p : ",round(p.value,digits = 3))), inherit.aes = FALSE)+
  coord_cartesian(ylim = c(0,4.6))+
  ggtitle(label = "LUAD")
  
p2

ggsave(filename = paste0(out_folder,"2E_linear_model_densities_for_stages_LUAD.pdf"),
      plot = p2,
      device = "pdf",
      width = 14,
      height = 4)
```


## fractions in tumor and stroma across stages - to be removed
this code is here for sanity checks. We could not identify any of the either tumor or stroma based cell type fractions associated with stage in a linear model
```{r}
stroma_IF1 <- read.csv(paste0(mount_path,"Figures/scores_for_sharing/celltype_fractions_stroma_compartment_IF1.csv"))
stroma_IF2 <- read.csv(paste0(mount_path,"Figures/scores_for_sharing/celltype_fractions_stroma_compartment_IF2.csv"))
stroma_IF3 <- read.csv(paste0(mount_path,"Figures/scores_for_sharing/celltype_fractions_stroma_compartment_IF3.csv"))
tumor_IF1 <- read.csv(paste0(mount_path,"Figures/scores_for_sharing/celltype_fractions_tumor_compartment_IF1.csv"))
tumor_IF2 <- read.csv(paste0(mount_path,"Figures/scores_for_sharing/celltype_fractions_tumor_compartment_IF2.csv"))
tumor_IF3 <- read.csv(paste0(mount_path,"Figures/scores_for_sharing/celltype_fractions_tumor_compartment_IF3.csv"))


colnames(stroma_IF1) <- c("immucan_id",paste0(colnames(stroma_IF1)[-1],"__stroma_IF1"))
colnames(stroma_IF2) <- c("immucan_id",paste0(colnames(stroma_IF2)[-1],"__stroma_IF2"))
colnames(stroma_IF3) <- c("immucan_id",paste0(colnames(stroma_IF3)[-1],"__stroma_IF3"))
colnames(tumor_IF1) <- c("immucan_id",paste0(colnames(tumor_IF1)[-1],"__tumor_IF1"))
colnames(tumor_IF2) <- c("immucan_id",paste0(colnames(tumor_IF2)[-1],"__tumor_IF2"))
colnames(tumor_IF3) <- c("immucan_id",paste0(colnames(tumor_IF3)[-1],"__tumor_IF3"))

tumor_IF2 <- tumor_IF2 %>%
  filter(immucan_id != "LUNG-NSCLC2-0564")
cur_dat <- plyr::join_all(list(stroma_IF1, stroma_IF2, stroma_IF3, tumor_IF1, tumor_IF2,tumor_IF3), by='immucan_id', type='left')

# take only patients for which all data is available
pats <- Reduce(intersect, list(stroma_IF1$immucan_id, stroma_IF2$immucan_id, stroma_IF3$immucan_id,tumor_IF1$immucan_id,tumor_IF2$immucan_id, tumor_IF3$immucan_id))
cur_dat <- cur_dat %>% 
  filter(immucan_id %in% pats)

dim(cur_dat)
tumor_IF2  %>% summarise(across(-immucan_id, ~ sum(. != 0)))
# filtering: we will only keep data where we have at least 50 patients with non-zero counts
columns_to_keep <- cur_dat %>%
  summarise(across(-immucan_id, ~ sum(. != 0))) %>%
  pivot_longer(everything(),names_to = "column", values_to = "zero_count") %>%
  filter(zero_count > 50) %>%
  pull(column)
cur_dat <- cur_dat %>%
  select(immucan_id, all_of(columns_to_keep))

colnames(cur_dat) <- gsub("_GB_","_GrzB_",colnames(cur_dat))


sigs <- cur_dat %>%
  pivot_longer(-immucan_id) %>%
  filter(grepl("tumor",name)) %>%
  left_join(clin_dat,"immucan_id") %>%
  select(immucan_id,name, value,simple_stage) %>%
  mutate(stage = ifelse(simple_stage == "I",1,NA),
         stage = ifelse(simple_stage == "II",2,stage),
         stage = ifelse(simple_stage == "III",3,stage),
         stage = ifelse(simple_stage == "IV",4,stage),
         stage = as.numeric(as.factor(stage))) %>%
  filter(stage != 4) %>%
  group_by(name) %>%
  nest() %>%
  mutate(model = map(data, ~ lm(value ~ stage, data = .x)),
         summary = map(model, broom::tidy)) %>%
  select(name, summary) %>%
  unnest(c(summary)) %>%
  filter(term == "stage") %>%
  mutate(tech = str_split_i(name,"__",2)) %>%
  ungroup() %>%

  mutate(FDR = p.adjust(p = p.value,method = "BH"),
         sig = ifelse(FDR < 0.1,TRUE, FALSE))

sigs %>%
  filter(sig > 0.1)
# no significant hits!
```






# Smoking

## 2I imaging

### imaging modules never vs current
```{r}
dat_z <- left_join(modules, traits, "immucan_id") %>%
  pivot_longer(cols = 1:8)

result <- lollipop_plot(dat_z, "never_vs_current_smoker", "Current smoker", "Non-smoker")
p1 <- result$plot+
  ggtitle("Current smokers vs Non-smokers")+
  ylab("Imaging modules")+
  theme(axis.title.y = element_text(size= 14))+
  coord_cartesian(xlim = c(-0.8,0.9))
```

### imaging modules never vs ex
```{r}
result <- lollipop_plot(dat_z, "never_vs_Ex_smoker", "Ex-smoker", "Non-smoker", order = FALSE, order_vec = c("Thelper_enriched","cold","medium_inflamed","proliferating","highly_inflamed","Neutrophil_enriched","mixed","inflamed_proliferating"))
p2 <- result$plot+
  theme(axis.text.y = element_blank())+
  ggtitle("Ex-smokers vs Non-smokers")
```

### imaging modules current vs ex
```{r}
result <- lollipop_plot(dat_z, "current_vs_Ex_smoker", "Current smoker", "Ex-smoker", order = FALSE, order_vec = c("Thelper_enriched","cold","medium_inflamed","proliferating","highly_inflamed","Neutrophil_enriched","mixed","inflamed_proliferating"))
p3 <- result$plot +
  theme(axis.text.y = element_blank())+
  ggtitle("Current smokers vs Ex-smokers")
```

### plot imaging
```{r, fig.height=4, fig.width= 20}
p <- cowplot::plot_grid(p1, p2, p3, ncol = 3, rel_widths = c(1.05,0.7,0.7))
p

ggsave(plot = p, path = out_folder,filename = "Imaging_modules_for_smoking_LUAD.pdf",width = 15, height = 3.75,device = cairo_pdf)
```



## 2J Inflammation signature in smoking 

```{r}
cur_dat <- read.table(paste0(mount_path,"data/RNASeq/CURIE/inflammation_quantification_NSCLC2/outputs/pathwayActivity/PathwayActivity_NSCLC2_mlm"))
my_comp <-  list( c("Current smoker", "Non-smoker"), c("Ex-smoker", "Non-smoker"), c("Current smoker", "Non-smoker") )

cur_dat <- cur_dat %>%
  as.data.frame() %>%
  mutate(immucan_id = paste0("LUNG-",rownames(cur_dat))) %>%
  left_join(traits,by = "immucan_id")

cur_dat <- cur_dat %>%
  pivot_longer(cols = c(JAK.STAT,NFkB,TNFa))
p <- cur_dat %>%
  filter(!is.na(smoking_status)) %>%
  ggplot(aes(x = smoking_status,y = value, fill = smoking_status))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(color = smoking_status),position = position_jitterdodge(jitter.width = 0.2), size= 1)+
  facet_wrap(.~name)+
  scale_fill_manual(values = alpha(c("blue", "lightblue", "cyan3"),alpha = 0.5))+
  scale_color_manual(values = c("blue", "lightblue", "cyan3"))+
  stat_compare_means(size = 5,label = "p.signif", comparisons = my_comp,method = "t.test")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5,hjust = 1, size = 16),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 14),
        legend.text = element_text(size = 14),
        strip.text.x = element_text(size = 14),
        legend.position = "bottom",
        legend.location = "plot")+
  ylab("Pathway score")+ 
  guides(fill = guide_legend(nrow = 2, byrow = TRUE))+
  coord_cartesian(ylim=c(-3, 6))

p 

ggsave(filename = paste0(out_folder,"inflammation_score_for_smoking_status.pdf"),
       plot = p,
       device = "pdf",
       width = 4.5,
       height = 5)
```

# diff abundance smoking bubble plot - to be removed

### smoking status LUAD

```{r, fig.height=15, fig.width=25}
condition <- c("Non-smoker","Non-smoker","Ex-smoker")
names(condition) <- c("never_vs_current_smoker","never_vs_Ex_smoker","current_vs_Ex_smoker")

p.list <- list()

# merge with clinical data
anno <- clin_dat %>%
  filter(!is.na(smoking_status),
         simple_histology == "Adenocarcinoma")
  
for (i in data_type) {
  
  # get cell abundances and remove NAs
  abundances <- dat %>%
    select(contains(i)) %>%
    filter(complete.cases(.))
  abundances <- t(abundances)
  colnames(abundances) <-dat %>%
    select(contains(i),immucan_id) %>%
    filter(complete.cases(.)) %>%
    pull(immucan_id)
  
  abundances <- abundances[,colnames(abundances) %in% anno$immucan_id]
  
  meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
  Size <- factor(meta[,"smoking_status"])
  meta$Size <- Size
  
  # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
  y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
  design <- model.matrix(~0+group,data = y.ab$samples)
  colnames(design) <- levels(y.ab$samples$group)
  # we do not fit a trend 
  y.ab <- estimateDisp(y.ab, design, trend.method = "none")
  y.ab$common.dispersion
  #plotBCV(y.ab,pch = 5)
  
  # we only compare the diff abundance for Adeno vs Sqam.
  fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
  fit_1_2 <- glmLRT(glmfit = fit.ab,contrast = c(1,-1,0))
  fit_1_3 <- glmLRT(glmfit = fit.ab,contrast = c(1,0,-1))
  fit_2_3 <- glmLRT(glmfit = fit.ab,contrast = c(0,1,-1))
  res_list <- list(fit_1_2,fit_1_3,fit_2_3)
  
  cur_list <- lapply(res_list,function(x){
    sig_dat <- x$table
    sig_dat$celltypes <- rownames(sig_dat)
    sig_dat$comparison <- gsub("Size","",x$comparison)
    return(sig_dat)
  })
  
  p.list[[i]] <- do.call(rbind,cur_list)
}   

cur_dat_LUAD <- do.call(rbind,p.list)
cur_dat_LUAD <- cur_dat_LUAD %>%
  mutate(histology = "LUAD")
```

### smoking status LUSC
of note, there are no non-smokers for LUSC histology in our cohort 
```{r}
condition <- c("Non-smoker","Non-smoker","Ex-smoker")
names(condition) <- c("never_vs_current_smoker","never_vs_Ex_smoker","current_vs_Ex_smoker")
p.list <- list()

# merge with clinical data
anno <- clin_dat %>%
  filter(!is.na(smoking_status),
         simple_histology == "Squamous cell carcinoma")
  
for (i in data_type) {
  
  # get cell abundances and remove NAs
  abundances <- dat %>%
    select(contains(i)) %>%
    filter(complete.cases(.))
  abundances <- t(abundances)
  colnames(abundances) <-dat %>%
    select(contains(i),immucan_id) %>%
    filter(complete.cases(.)) %>%
    pull(immucan_id)
  
  abundances <- abundances[,colnames(abundances) %in% anno$immucan_id]
  
  meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
  Size <- factor(meta[,"smoking_status"])
  meta$Size <- Size
  
  # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
  y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
  design <- model.matrix(~0+group,data = y.ab$samples)
  colnames(design) <- levels(y.ab$samples$group)
  # we do not fit a trend 
  y.ab <- estimateDisp(y.ab, design, trend.method = "none")
  y.ab$common.dispersion
  #plotBCV(y.ab,pch = 5)
  
  # we only compare the diff abundance for Adeno vs Sqam.
  fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
  fit_1_2 <- glmLRT(glmfit = fit.ab,contrast = c(1,-1))
  res_list <- list(fit_1_2)
  
  cur_list <- lapply(res_list,function(x){
    sig_dat <- x$table
    sig_dat$celltypes <- rownames(sig_dat)
    sig_dat$comparison <- gsub("Size","",x$comparison)
    return(sig_dat)
  })
  
  p.list[[i]] <- do.call(rbind,cur_list)
}   

cur_dat_LUSC <- do.call(rbind,p.list)

cur_dat_LUSC <- cur_dat_LUSC %>%
  mutate(histology = "LUSC")
```


### smoking signature based

```{r}
# Merge with mutational signatures
Lung_NSCLC2_filtered <-read.delim(paste0(mount_path,"data/WES/mutation_signatures/Lung_NSCLC2_filtered.tsv"), sep="\t")
names(Lung_NSCLC2_filtered)[names(Lung_NSCLC2_filtered)=="patient"] <- "immucan_id"
anno_WES <- merge(clin_dat, Lung_NSCLC2_filtered, by="immucan_id", sort=F)

# Combine smoking signatures
anno_WES[,"SBS4_SBS92"] <- apply(anno_WES[,c("SBS4","SBS92")],1,sum)

# Dichotomise smoking signature
cutoff <- median(anno_WES$SBS4_SBS92)
anno_WES <- anno_WES %>%
      filter(!is.na(SBS4_SBS92))  %>%
      mutate(SBS4_SBS92_dicho = ifelse(SBS4_SBS92 > cutoff, "high","low"))


# diff abundance
anno <- anno_WES
p.list <- list()

for (i in data_type) {

  # get cell abundances and remove NAs
  abundances <- dat %>%
    select(contains(i)) %>%
    filter(complete.cases(.))
  abundances <- t(abundances)
  colnames(abundances) <-dat %>%
    select(contains(i),immucan_id) %>%
    filter(complete.cases(.)) %>%
    pull(immucan_id)
  
  abundances <- as.data.frame(abundances[,colnames(abundances) %in% anno$immucan_id])
   
  meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
  stopifnot(all(colnames(abundances)==meta$immucan_id))
    
  Size <- factor(meta[,"SBS4_SBS92_dicho"], levels = c("low", "high"))

  # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
  y.ab <- DGEList(counts = abundances, group = Size)
  design <- model.matrix(~0+Size)
  colnames(design) <- levels(Size)
    
  # we do not fit a trend 
  y.ab <- estimateDisp(y.ab, design, trend.method = "none")
  y.ab$common.dispersion

  fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
  fit_1_2 <- glmLRT(glmfit = fit.ab,contrast = c(-1,1))

  res_list <- list(fit_1_2)
    
  cur_list <- lapply(res_list,function(x){
      sig_dat <- x$table
      sig_dat$celltypes <- rownames(sig_dat)
      sig_dat$comparison <- gsub("Size","",x$comparison)
      return(sig_dat)
    })
  p.list[[i]] <- do.call(rbind,cur_list)
}

cur_dat.SBS <- do.call(rbind,p.list)
cur_dat.SBS <-cur_dat.SBS %>%
  mutate(histology = "all")
```

### plot together
```{r, fig.height=10, fig.width=20}
cur_dat <- rbind(cur_dat_LUAD,cur_dat_LUSC,cur_dat.SBS)

cur_dat <- cur_dat %>%
  mutate(tech = gsub("_density","",str_split_i(celltypes,"__",2)),
         cond = ifelse(comparison == "1*Current smoker -1*Ex-smoker","CurrentSmoker_over_ExSmoker","ExSmoker_over_NonSmoker"),
         cond = ifelse(comparison == "1*Current smoker -1*Non-smoker","CurrentSmoker_over_NonSmoker",cond),
         cond = paste0(histology,"_",cond),
         cond = ifelse(comparison == "-1*low 1*high","SBS4-SBS92_High_over_Low",cond)) %>%
  group_by(tech) %>%
  mutate(FDR = p.adjust(PValue,method = "BH"),
         sig = ifelse(PValue < 0.05 & logFC > 0, "UP_trend",NA),
         sig = ifelse(FDR < 0.1 & logFC > 0, "UP",sig),
         sig = ifelse(PValue < 0.05 & logFC < 0, "DOWN_trend",sig),
         sig = ifelse(FDR < 0.1 & logFC < 0, "DOWN",sig),
         sig = ifelse(is.na(sig),"NOTSIG", sig),
         ct_name = gsub("_density","",str_split_i(celltypes,"__",1)))
  
to_show <- cur_dat %>%
  filter(sig %in% c("UP","DOWN")) %>%
  pull(ct_name)

cur_dat <- cur_dat %>%
  filter(ct_name %in% to_show) %>%
  mutate(celltypes = gsub("_abundance","",celltypes))

# get max and min fold change values
max_up <- max(cur_dat$logFC)+1
min_down <- min(cur_dat$logFC)-1

# plot the data
p <- cur_dat %>%
  ggplot(aes(x=celltypes, y = factor(cond, levels = c("SBS4-SBS92_High_over_Low",
                                                      "LUAD_CurrentSmoker_over_ExSmoker",
                                                      "LUAD_ExSmoker_over_NonSmoker",
                                                      "LUAD_CurrentSmoker_over_NonSmoker",
                                                      "LUSC_CurrentSmoker_over_ExSmoker")),col = logFC, size= -log10(PValue)))+
  #geom_tile(aes(fill = sig))+
  #theme_bw()+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(
    colours = c("blue", "white", "red"), 
    values = rescale(c(round(min_down), 0, round(max_up))),
    limits = c(round(min_down), round(max_up)),
    guide = guide_colorbar(
      title = "logFC",                   
      #barwidth = 2,                     
      #barheight = 1,                   
      title.theme = element_text(size = 12),
      label.theme = element_text(size = 12)
    )
  )+
  scale_size(range = c(1,4))+
  new_scale_colour() +
  geom_point(data = cur_dat,shape = 1,aes(size = -log10(PValue), col = sig), stroke =1.5)+
  scale_color_manual(values = c("UP" = "#9f1214","UP_trend" = "#e73032","DOWN" = "#25567d","DOWN_trend" = "#428bc7","NOTSIG" = "snow2"), name = "stats")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 12),
        axis.text.y = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 12),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        legend.location = "plot",
        legend.position = "right")
p

ggsave(filename = paste0(out_folder,"2F_diff_abundance_smoking.pdf"),
       plot = p,
       device = "pdf",
       width = 16,
       height = 5)
```



## sanity check smoking effects only in LUSC
```{r, fig.width=12, fig.height=6}
condition <- c("Ex-smoker")
names(condition) <- c("current_vs_Ex_smoker")

p.list <- list()

# merge with clinical data
anno <- clin_dat %>%
  filter(!is.na(smoking_status),
         simple_histology == "Squamous cell carcinoma")
  
for (i in data_type) {
  
  # get cell abundances and remove NAs
  abundances <- dat %>%
    select(contains(i)) %>%
    filter(complete.cases(.))
  abundances <- t(abundances)
  colnames(abundances) <-dat %>%
    select(contains(i),immucan_id) %>%
    filter(complete.cases(.)) %>%
    pull(immucan_id)
  
  abundances <- abundances[,colnames(abundances) %in% anno$immucan_id]
  
  meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
  Size <- factor(meta[,"smoking_status"])
  meta$Size <- Size
  
  # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
  y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
  design <- model.matrix(~0+group,data = y.ab$samples)
  colnames(design) <- levels(y.ab$samples$group)
  # we do not fit a trend 
  y.ab <- estimateDisp(y.ab, design, trend.method = "none")
  y.ab$common.dispersion
  #plotBCV(y.ab,pch = 5)
  
  # we only compare the diff abundance for Adeno vs Sqam.
  fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
  fit_1_2 <- glmLRT(glmfit = fit.ab,contrast = c(1,-1))
  res_list <- list(fit_1_2,fit_1_3,fit_2_3)
  
  cur_list <- lapply(res_list,function(x){
    sig_dat <- x$table
    sig_dat$celltypes <- rownames(sig_dat)
    sig_dat$comparison <- gsub("Size","",x$comparison)
    return(sig_dat)
  })
  
  p.list[[i]] <- do.call(rbind,cur_list)
}   

cur_dat <- do.call(rbind,p.list)

cur_dat <- cur_dat %>%
  mutate(tech = gsub("_density","",str_split_i(celltypes,"__",2)),
         cond = ifelse(comparison == "1*Current smoker -1*Ex-smoker","CurrentSmoker_over_ExSmoker","ExSmoker_over_NonSmoker"),
         cond = ifelse(comparison == "1*Current smoker -1*Non-smoker","CurrentSmoker_over_NonSmoker",cond),
         cond = ifelse(comparison == "-1*low 1*high","SBS4-SBS92_High_over_Low",cond)) %>%
  group_by(tech) %>%
  mutate(FDR = p.adjust(PValue,method = "BH"),
         sig_overall = ifelse(FDR < 0.1 ,TRUE,FALSE),
         sig = ifelse(FDR < 0.1 ,TRUE,FALSE),
         direction = ifelse(logFC > 0 ,"up", "down"),
         ct_name = gsub("_density","",str_split_i(celltypes,"__",1)))
  
to_show <- cur_dat %>%
  filter(sig_overall == TRUE) %>%
  pull(ct_name)

cur_dat <- cur_dat %>%
  filter(ct_name %in% to_show) %>%
  mutate(celltypes = gsub("_abundance","",celltypes))

# get max and min fold change values
max_up <- max(cur_dat$logFC)+1
min_down <- min(cur_dat$logFC)-1

# plot the data
p <- cur_dat %>%
  ggplot(aes(x=celltypes, y = factor(cond, levels = c("SBS4-SBS92_High_over_Low",
                                                      "CurrentSmoker_over_ExSmoker",
                                                      "ExSmoker_over_NonSmoker",
                                                      "CurrentSmoker_over_NonSmoker")),col = logFC, size= -log10(PValue)))+
  #geom_tile(aes(fill = sig))+
  #theme_bw()+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(
    colours = c("blue", "white", "red"), 
    values = rescale(c(round(min_down), 0, round(max_up))),
    limits = c(round(min_down), round(max_up)),
    guide = guide_colorbar(
      title = "logFC",                   
      #barwidth = 2,                     
      #barheight = 1,                   
      title.theme = element_text(size = 12),
      label.theme = element_text(size = 12)
    )
  )+
  scale_size(range = c(1,4))+
  new_scale_colour() +
  geom_point(data = cur_dat %>% filter(sig == TRUE),shape = 1,aes(size = -log10(PValue), col = sig), stroke =1.5)+
  scale_color_manual(values = c("TRUE" = "black","FALSE" = "white"), name = "FDR < 0.1")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 12),
        axis.text.y = element_text(size = 12),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 12),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        legend.location = "plot",
        legend.position = "right")
p
```

# images for smoking - to be removed
not in paper
```{r, fig.height=12, fig.width=16}
# identify images with high amuounts of differentiall abundant cello clusters 18 from panel 1
cello_dat %>% 
  left_join(clin_dat,"immucan_id") %>%
  select(immucan_id,CN_13__cello1,CN_17__cello1, smoking_status) %>%
  arrange(desc(CN_13__cello1))

#LUNG-NSCLC2-0644|0558 # 0644 is current and 0558 ex-smoker 

masks <- readRDS(paste0(mount_path,"data/IMC/Panel_1/Rout/masks.rds"))
images <- readRDS(paste0(mount_path,"data/IMC/Panel_1/Rout/images.rds"))

cur_df <- as.data.frame(colData(sce1))
rownames(cur_df) <- NULL
cur_df <- cur_df %>%
  select(immucan_id,image, sample_id) %>%
  distinct() %>%
  mutate(image = gsub(".tiff","",image))
rownames(cur_df) <- cur_df$image

mcols(images)$immucan_id <- cur_df[names(images),"immucan_id"]
mcols(masks)$immucan_id <- cur_df[names(masks),"immucan_id"]
mcols(images)$sample_id <- cur_df[names(images),"sample_id"]
mcols(masks)$sample_id <- cur_df[names(masks),"sample_id"]

cur_sce <- sce1[,grepl("0688|0665|0671",sce1$immucan_id)]
# 0688 is ex-smoker and 0680|0713 is current smoker and 0671 is a non-smoker
for(i in c("0688","0665","0671")){
  ROI <- 1
  plot_sce <- cur_sce[,cur_sce$ROI == ROI & grepl(i,cur_sce$immucan_id)]
  im_sel <- unique(plot_sce$sample_id)
  
  cur_ims <- getImages(images, im_sel)
  cur_masks <- getImages(masks, im_sel)
  
  channelNames(cur_ims) <- rownames(plot_sce)
  
  cur_images <- CytoImageList(cur_ims, on_disk = FALSE)
  cur_masks <- CytoImageList(cur_masks, on_disk = FALSE)
  cur_images <- scaleImages(cur_images,value = 2^16-1)
  cur_masks <- scaleImages(cur_masks,value = 2^16-1)
  
  cur_images <- normalize(cur_images, separateImages = TRUE)
  cur_images <- normalize(cur_images, separateImages = TRUE, 
                          inputRange = c(0, 0.2))
  
  
  plot_sce$CNs <- "other"
  plot_sce[,plot_sce$cellohood %in% c("CN_13","CN_17")]$CNs <- "CN13_and_CN17"
  plot_sce[,plot_sce$cellohood %in% c("CN_13","CN_17")]$CNs <- plot_sce[,plot_sce$cellohood %in% c("CN_13","CN_17")]$cellohood
  
  plotCells(mask = cur_masks,
            object = plot_sce,
            cell_id = "ObjectNumber",
            img_id = "sample_id",
            image_title = NULL,
            colour_by = "celltypes",
            colour = list(celltypes = metadata(plot_sce)$colour_vectors$celltypes),
            legend = list(colour_by.labels.cex = 0.1),
            scale_bar = list(length = 100,
                             label = expression("100 " ~ mu * "m"),
                             cex = 1.2,
                             lwidth = 5)
            ,save_plot = list(filename = paste0(out_folder,"LUNG-NSCLC2-",i,"_ROI_",ROI,"_celltypes.png"))
            )
  plotCells(mask = cur_masks,
            object = plot_sce,
            cell_id = "ObjectNumber",
            img_id = "sample_id",
            colour_by = "CNs",
            colour = list(CNs = c(CN_13 = "red",
                                  CN_17 = "blue",
                                  other = "gray")),
            image_title = NULL,
            legend = list(colour_by.labels.cex = 0.7),
            scale_bar = list(length = 100,
                             label = expression("100 " ~ mu * "m"),
                             cex = 1.2,
                             lwidth = 5)
            ,save_plot = list(filename = paste0(out_folder,"LUNG-NSCLC2-",i,"_ROI_",ROI,"_CN_13_17.png"))
            )
  
  if (i == "0688") {
    plotPixels(image = cur_images,
             object = plot_sce[,plot_sce$CNs != "other"],
             #mask = cur_masks,
             cell_id = "ObjectNumber",
             img_id = "sample_id",
             colour_by = c("CD38", "CD8a", "Ecad"),
             #outline_by = "CNs",
             image_title = NULL,
             colour = list(CD38 = c("black","green"),
                           CD8a = c("black","red"),
                           Ecad = c("black","blue"),
                           CNs = c(CN_13 = "white",
                                   CN_17 = "yellow")),
             bcg = list(CD38 = c(0,2,1),
                        Ecad = c(0,4,1),
                        CD8a = c(0,2,1)),
             scale_bar = list(length = 100,
                              label = expression("100 " ~ mu * "m"),
                              cex = 1.2,
                              lwidth = 5)
             ,save_plot = list(filename =paste0(out_folder,"LUNG-NSCLC2-",i,"_ROI_",ROI,"_CD38_expr.png"))
             )
  } else {
    plotPixels(image = cur_images,
             object = plot_sce[,plot_sce$CNs != "other"],
             #mask = cur_masks,
             cell_id = "ObjectNumber",
             img_id = "sample_id",
             colour_by = c("CD38", "CD8a", "Ecad"),
             #outline_by = "CNs",
             image_title = NULL,
             colour = list(CD38 = c("black","green"),
                           CD8a = c("black","red"),
                           Ecad = c("black","blue"),
                           CNs = c(CN_13 = "white",
                                   CN_17 = "yellow")),
             bcg = list(CD38 = c(0,2,1),
                        CD8a = c(0,2,1),
                        Ecad = c(0,1,1)),
             scale_bar = list(length = 100,
                              label = expression("100 " ~ mu * "m"),
                              cex = 1.2,
                              lwidth = 5)
             ,save_plot = list(filename =paste0(out_folder,"LUNG-NSCLC2-",i,"_ROI_",ROI,"_CD38_expr.png"))
             )
  }
  
}
```

