---
title: "Figure S4"
author: "Daniel Schulz"
date: "2024-11-13"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggpubr)
library(SingleCellExperiment)
library(survivalAnalysis)
library(ggsurvfit)
library(survminer)
library(survival)
library(edgeR)
library(cowplot)
library(ggrepel)
library(igraph)
library(ggraph)
library(ggrepel)
library(scales)
library(ggnewscale)
```

```{r}
mount_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/"
out_folder <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/revised_figures/Figure_4_supp/"

if (!dir.exists(out_folder)) {
  dir.create(out_folder)
}

source("../code/lollipop_function.R")
```

# load data
we will load the data we generated at the beginning of figure 2
```{r}
dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/abundances_dat.rds"))
dens_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/densities_dat.rds"))
cello_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/cello_dat.rds"))
clin_dat <- read.csv("/mnt/immucan_volume/clinical_data/Lung_NSCLC2/2026_final/clinical.csv")
traits <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/all_traits.csv"))

rna <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/rna_scores_tpm.rds")
rna$immucan_id  <- rownames(rna)
rownames(rna) <- NULL

modules <- as.data.frame(readRDS( "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/patient_imaging_meta_cluster_scores_tpm.rds"))
modules$immucan_id <- rownames(modules)
rownames(modules) <- NULL

# variables
data_type <- unique(str_split_i(string = colnames(dat[,-1]),pattern = "__",i = 2))
```

# S4A Kaplan Meier stage

```{r, fig.height=6}
cur_clin <- clin_dat
cur_clin$Stage <- cur_clin$simple_stage
cur_clin <- cur_clin %>%
  mutate(Stage = ifelse(Stage == "I","1",Stage),
         Stage = ifelse(Stage == "II","2",Stage),
         Stage = ifelse(Stage == "III","3",Stage),
         Stage = ifelse(Stage == "IV","4",Stage))
surv_fit <- survfit(Surv(time_to_event, censoring) ~ Stage, data = cur_clin)

p <- ggsurvplot(surv_fit, data = cur_clin, risk.table = TRUE, pval = TRUE,palette = c("palevioletred3","darkseagreen3","dodgerblue", "darkorange"))

p <- plot_grid(p$plot,p$table,ncol = 1, rel_heights = c(1,0.5))
p
ggsave(p,filename = paste0(out_folder,"S4A_Kaplan_Meier_CPS_score.pdf"),device = "pdf",width = 6, height = 5)
```


# S4B Imaging modules for stage 
```{r}
dat_z <- left_join(modules, traits, "immucan_id") %>%
  filter(!is.na(simple_stage),
         simple_histology == "Adenocarcinoma") %>%
  pivot_longer(cols = 1:7) %>%
  ungroup()

sigs <- dat_z %>%
  select(immucan_id,name, value,simple_stage) %>%
  mutate(stage = ifelse(simple_stage == "I",1,NA),
         stage = ifelse(simple_stage == "II",2,stage),
         stage = ifelse(simple_stage == "III",3,stage),
         stage = ifelse(simple_stage == "IV",4,stage),
         stage = as.numeric(as.factor(stage))) %>%
  filter(stage != 4) %>%
  group_by(name) %>%
  nest() %>%
  mutate(model = map(data, ~ lm(value ~ stage, data = .x)),
         summary = map(model, broom::tidy)) %>%
  select(name, summary) %>%
  unnest(c(summary)) %>%
  filter(term == "stage") %>%
  mutate(tech = str_split_i(name,"__",2)) %>%
  group_by(tech) %>%
  mutate(FDR = p.adjust(p = p.value,method = "BH"),
         sig_p = ifelse(p.value < 0.05,TRUE, FALSE),
         sig_fdr = ifelse(FDR < 0.1,"T", "F"))


summary_dat <- sigs %>%
  #filter(sig_p == TRUE) %>%
  #filter(grepl("CD8_GATA3_BCL2|^CD8_density|CD8_Ki67_density|CD8_TCF7_density|Tumor_IDO1",name)) %>%
  mutate(y_position = 1000,
         name = gsub(pattern = "density_",replacement = "",x = name)) %>%
  ungroup() %>%
  as.data.frame()

p_im <- dat_z %>%
  select(immucan_id,name, value,simple_stage) %>%
  mutate(stage = ifelse(simple_stage == "I",1,NA),
         stage = ifelse(simple_stage == "II",2,stage),
         stage = ifelse(simple_stage == "III",3,stage),
         stage = ifelse(simple_stage == "IV",4,stage),
         stage = as.numeric(as.factor(stage)),
         name = gsub(pattern = "density_",replacement = "",x = name)) %>%
  filter(name %in% summary_dat$name,
         stage != 4) %>%
  ggplot(aes(x = name,y = value,fill=as.factor(stage))) +
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(col = as.factor(stage)),position = position_jitterdodge(jitter.width = 0.2),size= 0.5)+
  scale_fill_manual(values = alpha(c("1" = "palevioletred3", "2" = "darkseagreen3", "3" = "dodgerblue"),alpha = 0.6))+
  scale_color_manual(values = c("1" = "palevioletred3", "2" = "darkseagreen3", "3" = "dodgerblue"))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90,vjust= 0.5, hjust = 1),
        axis.title.x = element_blank())+
  guides(fill=guide_legend(title="Stage"),
         col=guide_legend(title="Stage"))+
  geom_text(data = summary_dat, aes(x = name,y=2.3,label = paste0("fdr: ",sig_fdr)), inherit.aes = FALSE, size =3)+
  geom_text(data = summary_dat, aes(x = name,y=2.8,label = paste0("p : ",round(p.value,digits = 3))), inherit.aes = FALSE, size = 3)+
  #coord_cartesian(ylim = c(-0.35, 0.5))+
  ggtitle(label = "RNA module differences in LUAD")+
  labs(y = "RNA module score")
  
p_im
ggsave(plot = p_im, path = out_folder,filename = "S4B_imaging_modules_for_stage.pdf",width = 6, height = 3,device = cairo_pdf)
# plotting in chunk below
```


# S4C RNA modules for stage 
```{r}
dat_z <- left_join(rna, traits, "immucan_id") %>%
  filter(!is.na(simple_stage),
         simple_histology == "Adenocarcinoma") %>%
  pivot_longer(cols = 1:29) %>%
  group_by(name) %>%
  mutate(value = as.numeric(scale(value))) %>%  # z-score within each signature
  ungroup()

sigs <- dat_z %>%
  select(immucan_id,name, value,simple_stage) %>%
  mutate(stage = ifelse(simple_stage == "I",1,NA),
         stage = ifelse(simple_stage == "II",2,stage),
         stage = ifelse(simple_stage == "III",3,stage),
         stage = ifelse(simple_stage == "IV",4,stage),
         stage = as.numeric(as.factor(stage))) %>%
  filter(stage != 4) %>%
  group_by(name) %>%
  nest() %>%
  mutate(model = map(data, ~ lm(value ~ stage, data = .x)),
         summary = map(model, broom::tidy)) %>%
  select(name, summary) %>%
  unnest(c(summary)) %>%
  filter(term == "stage") %>%
  mutate(tech = str_split_i(name,"__",2)) %>%
  group_by(tech) %>%
  mutate(FDR = p.adjust(p = p.value,method = "BH"),
         sig_p = ifelse(p.value < 0.05,TRUE, FALSE),
         sig_fdr = ifelse(FDR < 0.1,"T", "F"))


summary_dat <- sigs %>%
  filter(sig_p == TRUE) %>%
  #filter(grepl("CD8_GATA3_BCL2|^CD8_density|CD8_Ki67_density|CD8_TCF7_density|Tumor_IDO1",name)) %>%
  mutate(y_position = 1000,
         name = gsub(pattern = "density_",replacement = "",x = name)) %>%
  ungroup() %>%
  as.data.frame()

p_rna <- dat_z %>%
  select(immucan_id,name, value,simple_stage) %>%
  mutate(stage = ifelse(simple_stage == "I",1,NA),
         stage = ifelse(simple_stage == "II",2,stage),
         stage = ifelse(simple_stage == "III",3,stage),
         stage = ifelse(simple_stage == "IV",4,stage),
         stage = as.numeric(as.factor(stage)),
         name = gsub(pattern = "density_",replacement = "",x = name)) %>%
  filter(name %in% summary_dat$name,
         stage != 4) %>%
  ggplot(aes(x = name,y = value,fill=as.factor(stage))) +
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(col = as.factor(stage)),position = position_jitterdodge(jitter.width = 0.2),size= 0.5)+
  scale_fill_manual(values = alpha(c("1" = "palevioletred3", "2" = "darkseagreen3", "3" = "dodgerblue"),alpha = 0.6))+
  scale_color_manual(values = c("1" = "palevioletred3", "2" = "darkseagreen3", "3" = "dodgerblue"))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90,vjust= 0.5, hjust = 1),
        axis.title.x = element_blank())+
  guides(fill=guide_legend(title="Stage"),
         col=guide_legend(title="Stage"))+
  geom_text(data = summary_dat, aes(x = name,y=3,label = paste0("fdr: ",sig_fdr)), inherit.aes = FALSE, size =3)+
  geom_text(data = summary_dat, aes(x = name,y=3.4,label = paste0("p : ",round(p.value,digits = 3))), inherit.aes = FALSE, size = 3)+
  #coord_cartesian(ylim = c(-0.35, 0.5))+
  ggtitle(label = "RNA module differences in LUAD")+
  labs(y = "RNA module score")
  
p_rna

# plotting in chunk below
```

# S4C_D TCGA RNA modules
```{r, fig.width=10, fig.height=6}
library(singscore)
tcga <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/MOFA/tcga_LUAD_LUSC.rds")
sce <- as(tcga, "SingleCellExperiment")
# subset to LUAD
sce <- sce[,sce$histology == "LUAD"]

# use length normalized data from TCGA
tpm_mat <- assay(tcga, "tpm_unstrand")
tpm_mat <- as.matrix(tpm_mat)   # in case it's a DelayedArray/sparse matrix

# if your gene set is symbols rather than Ensembl IDs, map via rowData
if ("gene_name" %in% colnames(rowData(tcga))) {
  symbols <- rowData(tcga)$gene_name
  keep <- !is.na(symbols) & !duplicated(symbols)   # drop NA + duplicate symbols
  tpm_mat <- tpm_mat[keep, ]
  rownames(tpm_mat) <- symbols[keep]
}

# 3. rank
ranked_data <- rankGenes(tpm_mat)

#ranked_data <- rankGenes(sce)

# load Bageav genes
df <- readxl::read_xlsx(
  "/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/RNASeq/annotations/Bagaev_FGES.xlsx",
  sheet = 1, skip = 1)

df <- df %>%
  dplyr::select(-ID) %>%
  pivot_longer(cols = -`Process/Signature`) %>%
  dplyr::select(-name) %>%
  filter(!is.na(value))

colnames(df) <- c("name", "gene")

signatures <- split(df$gene, df$name)

# can we find all genes in tcga
overlap <- sapply(signatures, function(g) sum(g %in% rownames(ranked_data)))
total   <- sapply(signatures, length)
print(data.frame(found = overlap, total = total, pct = round(100 * overlap / total)))

# score with signscore
scores_list <- lapply(names(signatures), function(nm) {
  sc <- simpleScore(ranked_data, upSet = signatures[[nm]], centerScore = TRUE)
  setNames(sc$TotalScore, rownames(sc))
})
rna_scores <- as.data.frame(do.call(cbind, scores_list))
colnames(rna_scores) <- names(signatures)

# z score the signscore data
rn <- rownames(rna_scores)
rna_scores <- as.data.frame(apply(rna_scores, 2, scale))
rownames(rna_scores) <- rn

# investigate stage
table(sce$tobacco_smoking_status)
table(sce$ajcc_pathologic_stage)
colnames(colData(sce))
clin_tcga <- as.data.frame(colData(sce)) %>%
  select(patient, ajcc_pathologic_stage) %>%
  distinct()

cur_dat <- cbind(clin_tcga, rna_scores[rownames(clin_tcga),])

# define stages

dat_z <- cur_dat %>%
  pivot_longer(cols = -c(patient,ajcc_pathologic_stage))

sigs <- dat_z %>%
  select(patient,name, value,ajcc_pathologic_stage) %>%
  mutate(stage = ifelse(grepl("I",ajcc_pathologic_stage),1,NA),
         stage = ifelse(grepl("II",ajcc_pathologic_stage),2,stage),
         stage = ifelse(grepl("III",ajcc_pathologic_stage),3,stage),
         stage = ifelse(grepl("IV",ajcc_pathologic_stage),4,stage),
         stage = as.numeric(as.factor(stage))) %>%
  filter(stage != 4) %>%
  group_by(name) %>%
  nest() %>%
  mutate(model = map(data, ~ lm(value ~ stage, data = .x)),
         summary = map(model, broom::tidy)) %>%
  select(name, summary) %>%
  unnest(c(summary)) %>%
  filter(term == "stage") %>%
  ungroup() %>%
  mutate(FDR = p.adjust(p = p.value,method = "BH"),
         sig_p = ifelse(p.value < 0.05,TRUE, FALSE),
         sig_fdr = ifelse(FDR < 0.1,"T", "F"))


summary_dat_val <- sigs %>%
  filter(name %in% c("Matrix","EMT signature","Cancer-associated fibroblasts","Granulocyte traffic","Myeloid cells traffic","Protumor cytokines"))%>%
  ungroup() %>%
  as.data.frame()

summary_dat_sig <- sigs %>%
  filter(sig_p == TRUE) %>%
  ungroup() %>%
  as.data.frame()

p1 <- dat_z %>%
  select(patient,name, value,ajcc_pathologic_stage) %>%
  mutate(stage = ifelse(grepl("I",ajcc_pathologic_stage),1,NA),
         stage = ifelse(grepl("II",ajcc_pathologic_stage),2,stage),
         stage = ifelse(grepl("III",ajcc_pathologic_stage),3,stage),
         stage = ifelse(grepl("IV",ajcc_pathologic_stage),4,stage),
         stage = as.numeric(as.factor(stage)),
         name = gsub(pattern = "density_",replacement = "",x = name)) %>%
  filter(name %in% summary_dat_sig$name ,
         stage != 4) %>%
  ggplot(aes(x = name,y = value,fill=as.factor(stage))) +
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(col = as.factor(stage)),position = position_jitterdodge(jitter.width = 0.2),size= 0.5)+
  scale_fill_manual(values = alpha(c("1" = "palevioletred3", "2" = "darkseagreen3", "3" = "dodgerblue"),alpha = 0.6))+
  scale_color_manual(values = c("1" = "palevioletred3", "2" = "darkseagreen3", "3" = "dodgerblue"))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90,vjust= 0.5, hjust = 1),
        axis.title = element_blank())+
  guides(fill=guide_legend(title="Stage"),
         col=guide_legend(title="Stage"))+
  geom_text(data = summary_dat_sig, aes(x = name,y=2.8,label = paste0("fdr: ",sig_fdr)), inherit.aes = FALSE, size = 3)+
  geom_text(data = summary_dat_sig, aes(x = name,y=3.5,label = paste0("p : ",round(p.value,digits = 3))), inherit.aes = FALSE, size = 3)+
  coord_cartesian(ylim = c(-4, 3.5))+
  ggtitle(label = "RNA modules across stages in LUAD - TCGA exploratory")+
  theme(title = element_text(size = 10))

p_val <- dat_z %>%
  select(patient,name, value,ajcc_pathologic_stage) %>%
  mutate(stage = ifelse(grepl("I",ajcc_pathologic_stage),1,NA),
         stage = ifelse(grepl("II",ajcc_pathologic_stage),2,stage),
         stage = ifelse(grepl("III",ajcc_pathologic_stage),3,stage),
         stage = ifelse(grepl("IV",ajcc_pathologic_stage),4,stage),
         stage = as.numeric(as.factor(stage)),
         name = gsub(pattern = "density_",replacement = "",x = name)) %>%
  filter(name %in% summary_dat_val$name ,
         stage != 4) %>%
  ggplot(aes(x = name,y = value,fill=as.factor(stage))) +
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(col = as.factor(stage)),position = position_jitterdodge(jitter.width = 0.2),size= 0.5)+
  scale_fill_manual(values = alpha(c("1" = "palevioletred3", "2" = "darkseagreen3", "3" = "dodgerblue"),alpha = 0.6))+
  scale_color_manual(values = c("1" = "palevioletred3", "2" = "darkseagreen3", "3" = "dodgerblue"))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90,vjust= 0.5, hjust = 1),
        axis.title.x = element_blank())+
  guides(fill=guide_legend(title="Stage"),
         col=guide_legend(title="Stage"))+
  geom_text(data = summary_dat_val, aes(x = name,y=2.5,label = paste0("fdr: ",sig_fdr)), inherit.aes = FALSE, size = 3)+
  geom_text(data = summary_dat_val, aes(x = name,y=3.2,label = paste0("p : ",round(p.value,digits = 3))), inherit.aes = FALSE, size = 3)+
  coord_cartesian(ylim = c(-4, 3.2))+
  ggtitle(label = "RNA modules across stages in LUAD - TCGA validation")+
  theme(title = element_text(size = 10))+
  labs(y = "RNA module score")
p1
p_val

p <- plot_grid(p2 + theme(axis.text.x = element_blank()), p_val, ncol = 1, rel_heights = c(1.2,2))
p
ggsave(plot = p, path = out_folder,filename = "S4C_RNA_modules_for_stage_and_TCGA_validation.pdf",width = 6, height = 5,device = cairo_pdf)


#p1
#ggsave(plot = p1, path = out_folder,filename = "S4D_RNA_modules_for_stage_and_TCGA_exploratory.pdf",width = 6, height = 3,device = cairo_pdf)

dens_dat %>%
  select(immucan_id, B_density__IMC2) %>%
  left_join(traits, "immucan_id") %>%
  filter(simple_stage != "IV") %>%
  ggplot(aes(x = simple_stage,y = B_density__IMC2))+
  geom_boxplot()+
  stat_compare_means()
```

# S4D
code here: to be included in this script: [tumor subtyping](../../../RNA/tumor_subtypes/tumor_subtypes_oncotypes.Rmd)


```{r, fig.height=7, fig.width=20}
summary_dat <- sigs %>%
  #filter(sig_p == TRUE) %>%
  ungroup() %>%
  as.data.frame()

p2 <- dat_z %>%
  select(patient,name, value,ajcc_pathologic_stage) %>%
  mutate(stage = ifelse(grepl("I",ajcc_pathologic_stage),1,NA),
         stage = ifelse(grepl("II",ajcc_pathologic_stage),2,stage),
         stage = ifelse(grepl("III",ajcc_pathologic_stage),3,stage),
         stage = ifelse(grepl("IV",ajcc_pathologic_stage),4,stage),
         stage = as.numeric(as.factor(stage)),
         name = gsub(pattern = "density_",replacement = "",x = name)) %>%
  filter(name %in% summary_dat$name,
         stage != 4) %>%
  ggplot(aes(x = name,y = log10(value+1),fill=as.factor(stage))) +
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(col = as.factor(stage)),position = position_jitterdodge(jitter.width = 0.2),size= 0.5)+
  scale_fill_manual(values = alpha(c("1" = "palevioletred3", "2" = "darkseagreen3", "3" = "dodgerblue"),alpha = 0.6))+
  scale_color_manual(values = c("1" = "palevioletred3", "2" = "darkseagreen3", "3" = "dodgerblue"))+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90,vjust= 0.5, hjust = 1),
        axis.title = element_blank())+
  guides(fill=guide_legend(title="Stage"),
         col=guide_legend(title="Stage"))+
  geom_text(data = summary_dat, aes(x = name,y=0.85,label = paste0("fdr: ",sig_fdr)), inherit.aes = FALSE)+
  geom_text(data = summary_dat, aes(x = name,y=1,label = paste0("p : ",round(p.value,digits = 3))), inherit.aes = FALSE)+
  coord_cartesian(ylim = c(-1, 1))+
  ggtitle(label = "LUAD")
  
p2
sigs
```


# S4E Smoking
## RNA

### RNA modules never vs current
```{r}
dat_z <- left_join(rna, traits, "immucan_id") %>%
  filter(LUAD_vs_LUSC == "Adenocarcinoma") %>%
  pivot_longer(cols = 1:29)

result <- lollipop_plot(dat_z, "never_vs_current_smoker", "Current smoker", "Non-smoker",col_A = "blue", col_B = "cyan", zscore_data = TRUE)
order_vec <- result$data %>%
  arrange(delta_z) %>%
  pull(name)
p4 <- result$plot+
  
    ylab("RNA modules")+
  ggtitle("Current vs Non-smoker")+
  theme(axis.title.y = element_text(size= 14))
p4
```



### RNA modules never vs ex
```{r}
result <- lollipop_plot(dat_z, "never_vs_Ex_smoker", "Ex-smoker", "Non-smoker",col_A = "lightblue",col_B ="cyan", order = FALSE, order_vec = order_vec,zscore_data = TRUE)
p5 <- result$plot+
  theme(axis.text.y = element_blank())+
  ggtitle("Ex-smokers vs Non-smokers")
p5
```

### RNA modules current vs ex
```{r}
result <- lollipop_plot(dat_z, "current_vs_Ex_smoker", "Current smoker", "Ex-smoker", col_A = "blue",col_B ="lightblue", order = FALSE, order_vec = order_vec,zscore_data = TRUE)
p6 <- result$plot+
  theme(axis.text.y = element_blank())+
  ggtitle("Current smokers vs Ex-smokers")
p6
```

### plot RNA

```{r, fig.height=5.5, fig.width= 22}
p <- cowplot::plot_grid(p4, p5, p6, ncol = 3, rel_widths = c(1.5, 0.7, 0.7))
p
ggsave(plot = p4+
         theme(legend.position = "bottom", legend.box = "vertical")+
           guides(color = guide_legend(nrow = 2, byrow = TRUE),
         size = guide_legend(nrow = 2, byrow = TRUE)), path = out_folder,filename = "S4E_RNA_modules_for_smoking.pdf",width = 8, height = 8,device = cairo_pdf)
```

# S4F inflammation in smoking

```{r}
library(org.Hs.eg.db)
library(AnnotationDbi)
library(msigdbr)
library(singscore)

rna_mat <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/rna_mat_tpm.rds")


# ── gene signatures ────────────────────────────────────────────────────
hallmark_sets <- msigdbr(species = "Homo sapiens", collection = "H") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  split(.$gs_name) %>%
  lapply(function(x) x$gene_symbol)

# Targeted Hallmark gene sets of interest
target_sets <- c(
  "HALLMARK_INFLAMMATORY_RESPONSE",
  "HALLMARK_INTERFERON_GAMMA_RESPONSE",
  "HALLMARK_INTERFERON_ALPHA_RESPONSE",
  "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
  "HALLMARK_IL6_JAK_STAT3_SIGNALING",
  "HALLMARK_IL2_STAT5_SIGNALING"
)
gs_list_hallmark <- hallmark_sets[intersect(target_sets, names(hallmark_sets))]

# ── 2. Custom gene sets ────────────────────────────────────────────────────────
# CXCL9/10/11 chemokine axis (T cell recruitment)
gs_list_custom <- list(
  CXCL9_10_11_axis = c("CXCL9", "CXCL10", "CXCL11", "CXCR3",
                       "CXCL13", "CXCR5"),
  T_cell_activation = c("CD8A", "CD8B", "GZMB", "PRF1", "IFNG",
                        "TNF", "IL2", "PDCD1", "HAVCR2", "LAG3"),
  type_I_IFN = c("IFNB1", "IRF7", "MX1", "OAS1", "OAS2", "OAS3",
                 "ISG15", "IFIT1", "IFIT2", "IFIT3", "RSAD2", "HERC5"),
  NF_kB_target  = c("NFKB1", "RELA", "RELB", "NFKB2", "REL",
                    "BCL2L1", "IL1B", "IL6", "CXCL8", "TNF", "ICAM1"),
  TCR_signaling = c("CD3D", "CD3E", "CD3G", "ZAP70", "LCK", "LAT",
                    "ITK", "PLCG1", "NFATC1", "NFATC2"),
  cytotoxic = c("GZMB", "GZMA", "GZMK", "PRF1", "NKG7", "IFNG", "KLRG1"),
  exhaustion = c("PDCD1","LAG3","HAVCR2","TX","CTLA4", "IDO1")
  )

all_gs <- c(gs_list_hallmark, gs_list_custom)

# compute score
rna_for_score <- t(rna_mat)

# Filter gene sets to genes present in the RNA-seq matrix
all_gs <- lapply(all_gs, function(g) intersect(g, rownames(rna_for_score)))
all_gs <- all_gs[sapply(all_gs, length) >= 4]



overlap <- sapply(all_gs, function(g) sum(g %in% rownames(rna_for_score)))
total   <- sapply(all_gs, length)
print(data.frame(found = overlap, total = total, pct = round(100 * overlap / total)))

# ── Score with singscore ──────────────────────────────────────────────────────
ranked <- rankGenes(rna_for_score)

scores_list <- lapply(names(all_gs), function(nm) {
  sc <- simpleScore(ranked, upSet = all_gs[[nm]], centerScore = TRUE)
  setNames(sc$TotalScore, rownames(sc))
})


rna_scores <- as.data.frame(do.call(cbind, scores_list))
colnames(rna_scores) <- names(all_gs)
rownames(rna_scores) <- colnames(rna_for_score)
rna_scores$immucan_id <- rownames(rna_scores)
rownames(rna_scores) <- NULL

dat_z <- left_join(rna_scores, traits, "immucan_id") %>%
  filter(LUAD_vs_LUSC == "Adenocarcinoma") %>%
  pivot_longer(cols = 1:length(all_gs))

result <- lollipop_plot(dat_z, "never_vs_current_smoker", "Current smoker", "Non-smoker", col_A = "blue", col_B = "cyan", zscore_data = TRUE)
p <- result$plot+
    ylab("RNA modules")+
  ggtitle("Current vs Non-smoker")+
  theme(axis.title.y = element_blank(),
        axis.title.x = element_text(size= 18),
        legend.position = "bottom", 
        legend.text = element_text(size = 18),
        legend.title = element_text(size = 18))+
  guides(color = guide_legend(nrow = 2, byrow = TRUE),
         size = guide_legend(nrow = 2, byrow = TRUE))

p
```

```{r, fig.height=10, fig.width=10}
p <- result$plot+
    ylab("RNA modules")+
  ggtitle("Current vs Non-smoker")+
  theme(axis.title.y = element_blank(),
        axis.title.x = element_text(size= 18),
        #legend.position = "bottom", 
        legend.text = element_text(size = 18),
        legend.title = element_text(size = 18),
        title = element_text(size = 18))
p
ggsave(plot = p, path = out_folder,filename = "S4F_alternative_inflammation.pdf",width = 10, height = 6 ,device = cairo_pdf)

```

# S4G tumor types for smoking

```{r}
scores <- read.csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/tumor_subtype_scores.csv")

scores <- left_join(scores, traits,"immucan_id")

scores <- scores %>%
  rename(magnoid = "PP",
         squamoid = "PI",
         bronchioid = "TRU")

plot_data <- scores %>%
  pivot_longer(cols = 1:3) %>%
  filter(!is.na(smoking_status))

library(rstatix)
stat_test <- plot_data %>%
  group_by(name) %>%
  t_test(value ~ smoking_status,
         comparisons = list(c("Current smoker", "Ex-smoker"),
                             c("Current smoker", "Non-smoker"),
                             c("Ex-smoker", "Non-smoker"))) %>%
  mutate(p_round = round(p, 3)) %>%
  add_xy_position(x = "smoking_status")

p <- ggplot(plot_data, aes(x = smoking_status, y = value, fill = smoking_status)) +
  geom_boxplot() +
  geom_jitter(aes(col = smoking_status), position = position_jitterdodge(jitter.width = 0.2), size = 0.5) +
  scale_fill_manual(values = alpha(c("blue", "lightblue", "cyan3"), alpha = 0.5)) +
  scale_color_manual(values = c("blue", "lightblue", "cyan3")) +
  facet_wrap(~name) +
  stat_pvalue_manual(stat_test, label = "p_round", tip.length = 0.01,y.position = c(1, 1.2,1.4)) +
  theme_bw() +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  coord_cartesian(ylim = c(-0.7, 1.5)) +
  xlab("Smoking status") +
  ylab("Tumor type score")

p
ggsave(plot = p, path = out_folder,filename = "S4G_tumor_types_for_smoking.pdf",width = 5, height = 3.5,device = cairo_pdf)

```


# diff abundance smoking cellohood

```{r, fig.height=15, fig.width=25}
condition <- c("Non-smoker","Non-smoker","Ex-smoker")
names(condition) <- c("never_vs_current_smoker","never_vs_Ex_smoker","current_vs_Ex_smoker")

p.list <- list()

# merge with clinical data
anno <- clin_dat %>%
  filter(!is.na(smoking_status),
         simple_histology == "Adenocarcinoma")
  
for (i in c("cello1","cello2")) {

  abundances <- cello_dat %>%
    select(contains(i))
  abundances <- t(abundances)
  colnames(abundances) <- cello_dat$immucan_id
  
  abundances <- abundances[,colnames(abundances) %in% anno$immucan_id]
  
  meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
  Size <- factor(meta[,"smoking_status"])
  meta$Size <- Size
  
  # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
  y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
  design <- model.matrix(~0+group,data = y.ab$samples)
  colnames(design) <- levels(y.ab$samples$group)
  # we do not fit a trend 
  y.ab <- estimateDisp(y.ab, design, trend.method = "none")
  y.ab$common.dispersion
  #plotBCV(y.ab,pch = 5)
  
  # we only compare the diff abundance for Adeno vs Sqam.
  fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
  fit_1_2 <- glmLRT(glmfit = fit.ab,contrast = c(1,-1,0))
  fit_1_3 <- glmLRT(glmfit = fit.ab,contrast = c(1,0,-1))
  fit_2_3 <- glmLRT(glmfit = fit.ab,contrast = c(0,1,-1))
  res_list <- list(fit_1_2,fit_1_3,fit_2_3)
  
  cur_list <- lapply(res_list,function(x){
    sig_dat <- x$table
    sig_dat$celltypes <- rownames(sig_dat)
    sig_dat$comparison <- gsub("Size","",x$comparison)
    return(sig_dat)
  })
  
  p.list[[i]] <- do.call(rbind,cur_list)
}   

cur_dat_LUAD <- do.call(rbind,p.list)
cur_dat_LUAD <- cur_dat_LUAD %>%
  mutate(histology = "LUAD")
```

## smoking status LUSC
of note, there are no non-smokers for LUSC histology in our cohort 
```{r}
condition <- c("Non-smoker","Non-smoker","Ex-smoker")
names(condition) <- c("never_vs_current_smoker","never_vs_Ex_smoker","current_vs_Ex_smoker")
p.list <- list()

# merge with clinical data
anno <- clin_dat %>%
  filter(!is.na(smoking_status),
         simple_histology == "Squamous cell carcinoma")
  
for (i in c("cello1","cello2")) {

  abundances <- cello_dat %>%
    select(contains(i))
  abundances <- t(abundances)
  colnames(abundances) <- cello_dat$immucan_id
  
  abundances <- abundances[,colnames(abundances) %in% anno$immucan_id]
  
  meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
  Size <- factor(meta[,"smoking_status"])
  meta$Size <- Size
  
  # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
  y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
  design <- model.matrix(~0+group,data = y.ab$samples)
  colnames(design) <- levels(y.ab$samples$group)
  # we do not fit a trend 
  y.ab <- estimateDisp(y.ab, design, trend.method = "none")
  y.ab$common.dispersion
  #plotBCV(y.ab,pch = 5)
  
  # we only compare the diff abundance for Adeno vs Sqam.
  fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
  fit_1_2 <- glmLRT(glmfit = fit.ab,contrast = c(1,-1))
  res_list <- list(fit_1_2)
  
  cur_list <- lapply(res_list,function(x){
    sig_dat <- x$table
    sig_dat$celltypes <- rownames(sig_dat)
    sig_dat$comparison <- gsub("Size","",x$comparison)
    return(sig_dat)
  })
  
  p.list[[i]] <- do.call(rbind,cur_list)
}   

cur_dat_LUSC <- do.call(rbind,p.list)

cur_dat_LUSC <- cur_dat_LUSC %>%
  mutate(histology = "LUSC")
```


## smoking signature based

```{r}
# Merge with mutational signatures
Lung_NSCLC2_filtered <-read.delim(paste0(mount_path,"data/WES/mutation_signatures/Lung_NSCLC2_filtered.tsv"), sep="\t")
names(Lung_NSCLC2_filtered)[names(Lung_NSCLC2_filtered)=="patient"] <- "immucan_id"
anno_WES <- merge(clin_dat, Lung_NSCLC2_filtered, by="immucan_id", sort=F)

# Combine smoking signatures
anno_WES[,"SBS4_SBS92"] <- apply(anno_WES[,c("SBS4","SBS92")],1,sum)

# Dichotomise smoking signature
cutoff <- median(anno_WES$SBS4_SBS92)
anno_WES <- anno_WES %>%
      filter(!is.na(SBS4_SBS92))  %>%
      mutate(SBS4_SBS92_dicho = ifelse(SBS4_SBS92 > cutoff, "high","low"))


# diff abundance
anno <- anno_WES
p.list <- list()

for (i in c("cello1","cello2")) {

  abundances <- cello_dat %>%
    select(contains(i))
  abundances <- t(abundances)
  colnames(abundances) <- cello_dat$immucan_id
  abundances <- as.data.frame(abundances[,colnames(abundances) %in% anno$immucan_id])
   
  meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
  stopifnot(all(colnames(abundances)==meta$immucan_id))
    
  Size <- factor(meta[,"SBS4_SBS92_dicho"], levels = c("low", "high"))

  # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
  y.ab <- DGEList(counts = abundances, group = Size)
  design <- model.matrix(~0+Size)
  colnames(design) <- levels(Size)
    
  # we do not fit a trend 
  y.ab <- estimateDisp(y.ab, design, trend.method = "none")
  y.ab$common.dispersion

  fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
  fit_1_2 <- glmLRT(glmfit = fit.ab,contrast = c(-1,1))

  res_list <- list(fit_1_2)
    
  cur_list <- lapply(res_list,function(x){
      sig_dat <- x$table
      sig_dat$celltypes <- rownames(sig_dat)
      sig_dat$comparison <- gsub("Size","",x$comparison)
      return(sig_dat)
    })
  p.list[[i]] <- do.call(rbind,cur_list)
}

cur_dat.SBS <- do.call(rbind,p.list)
cur_dat.SBS <-cur_dat.SBS %>%
  mutate(histology = "all")
```


```{r, fig.height=10, fig.width=20}
cur_dat <- rbind(cur_dat_LUAD,cur_dat_LUSC,cur_dat.SBS)

cur_dat <- cur_dat %>%
  mutate(tech = gsub("_density","",str_split_i(celltypes,"__",2)),
         cond = ifelse(comparison == "1*Current smoker -1*Ex-smoker","CurrentSmoker_over_ExSmoker","ExSmoker_over_NonSmoker"),
         cond = ifelse(comparison == "1*Current smoker -1*Non-smoker","CurrentSmoker_over_NonSmoker",cond),
         cond = paste0(histology,"_",cond),
         cond = ifelse(comparison == "-1*low 1*high","SBS4-SBS92_High_over_Low",cond)) %>%
  group_by(tech) %>%
  mutate(FDR = p.adjust(PValue,method = "BH"),
         sig = ifelse(PValue < 0.05 & logFC > 0, "UP_trend",NA),
         sig = ifelse(FDR < 0.1 & logFC > 0, "UP",sig),
         sig = ifelse(PValue < 0.05 & logFC < 0, "DOWN_trend",sig),
         sig = ifelse(FDR < 0.1 & logFC < 0, "DOWN",sig),
         sig = ifelse(is.na(sig),"NOTSIG", sig),
         ct_name = gsub("_density","",str_split_i(celltypes,"__",1)))
  
to_show <- cur_dat %>%
  filter(sig %in% c("UP","DOWN")) %>%
  pull(celltypes)

cur_dat <- cur_dat %>%
  filter(celltypes %in% to_show) %>%
  mutate(celltypes = gsub("_abundance","",celltypes),
         celltypes = gsub("cello", "IMC",celltypes))

# get max and min fold change values
max_up <- max(cur_dat$logFC)+1
min_down <- min(cur_dat$logFC)-1

# plot the data
p <- cur_dat %>%
  ggplot(aes(x=celltypes, y = factor(cond, levels = c("SBS4-SBS92_High_over_Low",
                                                      "LUAD_CurrentSmoker_over_ExSmoker",
                                                      "LUAD_ExSmoker_over_NonSmoker",
                                                      "LUAD_CurrentSmoker_over_NonSmoker",
                                                      "LUSC_CurrentSmoker_over_ExSmoker")),col = logFC, size= -log10(PValue)))+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(
    colours = c("blue", "white", "red"), 
    values = rescale(c(round(min_down), 0, round(max_up))),
    limits = c(round(min_down), round(max_up)),
    guide = guide_colorbar(
      title = "logFC",                   
      title.theme = element_text(size = 12),
      label.theme = element_text(size = 12)
    )
  )+
  scale_size(range = c(1,4))+
  new_scale_colour() +
  geom_point(data = cur_dat,shape = 1,aes(size = -log10(PValue), col = sig), stroke =1.5)+
  scale_color_manual(values = c("UP" = "#9f1214","UP_trend" = "#e73032","DOWN" = "#25567d","DOWN_trend" = "#428bc7","NOTSIG" = "snow2"), name = "stats")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 16),
        axis.text.y = element_text(size = 16),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 12),
        axis.title.y = element_blank(),
        axis.title.x = element_blank(),
        legend.location = "plot",
        legend.position = "bottom") +
  guides(color = guide_legend(nrow = 2),  
         shape = guide_legend(nrow = 1))
p
```



# S4H mutation signatures

```{r}
# Merge with mutational signatures
Lung_NSCLC2_filtered <-read.delim(paste0(mount_path,"data/WES/mutation_signatures/Lung_NSCLC2_filtered.tsv"), sep="\t")
names(Lung_NSCLC2_filtered)[names(Lung_NSCLC2_filtered)=="patient"] <- "immucan_id"
anno_WES <- merge(traits, Lung_NSCLC2_filtered, by="immucan_id", sort=F)

apply(Lung_NSCLC2_filtered[, 4:12], 2, hist)
# Combine smoking signatures
anno_WES[,"SBS4_SBS92"] <- apply(anno_WES[,c("SBS4","SBS92")],1,sum)

# Dichotomise smoking signature
cutoff <- median(anno_WES$SBS4_SBS92)
anno_WES <- anno_WES %>%
      filter(!is.na(SBS4_SBS92))  %>%
      mutate(SBS4_SBS92_dicho = ifelse(SBS4_SBS92 > cutoff, "high","low"))

hist="ALL"
feature="SBS4_SBS92"
  
data_to_plot <- anno_WES[,c("smoking_status", feature)]
names(data_to_plot) <- c("smoking_status", "feature")
  
data_to_plot <- data_to_plot %>%
  filter(!is.na(smoking_status),
         !feature == 0
         )

sample_sizes <- data_to_plot %>%
    group_by(smoking_status) %>%
    summarise(n = n()) %>%
    mutate(label = paste0(smoking_status, "\n(n=", n, ")"))

p <- ggplot(data_to_plot, aes(x = smoking_status, y = feature, fill=smoking_status)) +
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(color = smoking_status),position = position_jitterdodge(jitter.width = 0.2), size= 1.5)+
  scale_fill_manual(values = alpha(c("blue", "lightblue", "cyan3"),alpha = 0.5))+
  scale_color_manual(values = c("blue", "lightblue", "cyan3"))+
  scale_x_discrete(labels = sample_sizes$label) + 
  labs(y = "# of mutations",
       title=paste0(feature, " (", hist, ")"),
       subtitle = "Wilcoxon Rank Sum test") +
  theme_minimal() +
  theme(panel.grid = element_blank(),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 16, color = "black"),
        axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 14, color = "black"),
        axis.text.y = element_text(size = 14, color = "black"),
        plot.title = element_text(size = 16, color = "black"),
        plot.subtitle = element_text(size = 14, color = "black"),
        axis.line = element_line(color = "black"),
        axis.ticks = element_line(color = "black"),
        legend.text = element_text(size = 14),
        legend.position = "bottom",
        legend.location = "plot") +
  stat_compare_means(comparisons = list(c("Non-smoker", "Ex-smoker"),
                                        c("Non-smoker", "Current smoker"),
                                        c("Ex-smoker", "Current smoker")),
                     method = "wilcox.test", label = "p.format",size = 5,label.y = c(1200, 1400, 1600))+
  coord_cartesian(ylim = c(0,1800))

p
ggsave(filename = paste0(out_folder,"S4H_mutation_signatures_for_smoking.pdf"),
       plot = p,
       device = "pdf",
       width = 6,
       height = 6)

```

# S4I imaging modules for mutation signatures

```{r}
cur_plot <- anno_WES %>%
  select(immucan_id, SBS4_SBS92) %>%
  filter(!SBS4_SBS92 == 0) %>%
  left_join(modules, "immucan_id")


mat <- rstatix::cor_mat(cur_plot[,-1],)[-1,1:2]
mat_p <- rstatix::cor_pmat(cur_plot[,-1],)[-1,1:2]


mat <- left_join(mat, mat_p, "rowname")
mat <- mat %>%
  mutate(SBS4_SBS92 = "SBS4_SBS92",
         label = case_when(
      SBS4_SBS92.y < 0.001 ~ "***",
      SBS4_SBS92.y < 0.01  ~ "**",
      SBS4_SBS92.y < 0.05  ~ "*")
    )
p <- mat %>%
 ggplot(aes(x = SBS4_SBS92, y = rowname, fill = SBS4_SBS92.x)) +
    geom_tile() +
    geom_text(aes(label = label), color = "black") +
    scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0) +
    #facet_wrap(~method) +
    theme_minimal() +
    labs(
      x = "Mutation", y = "Imaging module", fill = paste0("Pearson r")
    )+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1, size = 16),
        axis.text.y = element_text(size = 16),
        axis.title.y = element_text(size = 16),
        axis.title.x = element_text(size = 16),
        legend.title = element_text(size = 16))
pdf(file = paste0(out_folder,"S4I_imaging_modules_correlation_SBS4_SBS92.pdf"), width = 4, height = 5)
p
dev.off()
plot(cur_plot$SBS4_SBS92, cur_plot$Medium_inflamed)
```


