---
title: "Figure 3"
author: "Daniel Schulz"
date: "2024-11-15"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
options(bitmapType = "cairo")
```

```{r, message=FALSE}
library(tidyverse)
library(ggpubr)
library(SingleCellExperiment)
library(survivalAnalysis)
library(ggsurvfit)
library(survminer)
library(survival)
library(edgeR)
library(cowplot)
library(ggrepel)
library(igraph)
library(ggraph)
library(ggrepel)
library(scales)
library(ggnewscale)
library(BiocParallel)
library(imcRtools)
```

```{r}
mount_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/"
out_folder <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/revised_figures/Figure_3/"

if (!dir.exists(out_folder)) {
  dir.create(out_folder)
}

source("code/lollipop_function.R")
```

# load data
we will load the data we generated at the beginning of figure 2
```{r}
dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/abundances_dat.rds"))
dens_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/densities_dat.rds"))
cello_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/cello_dat.rds"))
clin_dat <- read.csv("/mnt/immucan_volume/clinical_data/Lung_NSCLC2/2026_final/clinical.csv")
traits <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/all_traits.csv"))

modules <- as.data.frame(readRDS( "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/patient_imaging_meta_cluster_scores_tpm.rds"))

rna <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/rna_scores_tpm.rds")
rna$immucan_id  <- rownames(rna)
rownames(rna) <- NULL

modules <- modules  %>%
  mutate(immucan_id = rownames(modules))
rownames(modules) <- NULL

# variables
data_type <- unique(str_split_i(string = colnames(dat[,-1]),pattern = "__",i = 2))

# colors
stage_col = setNames(c("palevioletred3","darkseagreen3","dodgerblue"),c("I","II","III"))
stage_col_num = setNames(c("palevioletred3","darkseagreen3","dodgerblue"),c("1","2","3"))
hist_col_2 = setNames(c("blue","orange"),c("LUAD","LUSC"))
hist_col_3 = setNames(c("blue","orange", "gray40"),c("LUAD","LUSC","Other"))

```

# TPS stats

```{r}
cur_var <- "TPS_score"
var_A <- "positive"
var_B <-  "negative"
col_A <- "orange"
col_B <- "darkgreen"
title_name <- "TPS"
```

```{r}
fisher.test(table(traits$TPS_score,traits$LUAD_vs_LUSC))
table(traits$TPS_score,traits$LUAD_vs_LUSC)
table(traits$simple_histology)
table(traits$TPS_score)
57/184
```

# 3B RNA modules for TPS

## LUAD
```{r}
dat_z <- left_join(rna, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC")) %>%
  filter(LUAD_vs_LUSC == "LUAD") %>%
  pivot_longer(cols = 1:29)

library(scales)
result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B, zscore_data = TRUE)
result$data
p_LUAD <- result$plot +
  ggtitle(paste0("RNA modules for ", title_name," in LUAD"))+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))
p_LUAD
cur_order <- result$data$name

```

## LUSC
```{r, fig.height=5, fig.width=15}
dat_z <- left_join(rna, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC")) %>%
  filter(LUAD_vs_LUSC == "LUSC") %>%
  pivot_longer(cols = 1:29)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B,order = FALSE, order_vec = cur_order, zscore_data = TRUE)
p_LUSC <- result$plot+ 
  ggtitle(paste0("RNA modules for ", title_name," in LUSC"))+
  theme(axis.text.y = element_blank())+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))
p_LUSC

p <- plot_grid(p_LUAD, p_LUSC, nrow = 1, rel_widths = c(1.1, 0.6))
p
ggsave(plot = p, path = out_folder,filename = "3B_RNA_modules_for_TPS.pdf",width = 10, height = 6,device = cairo_pdf)

```

# 3C Imaging modules for TPS

## LUAD
```{r}
dat_z <- left_join(modules, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC")) %>%
  filter(LUAD_vs_LUSC == "LUAD") %>%
  pivot_longer(cols = 1:7)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B)
p_LUAD <- result$plot +
  ggtitle(paste0("Imaging modules for ", title_name," in LUAD"))+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))
p_LUAD
cur_order <- result$data$name

```

## LUSC
```{r, fig.height=5, fig.width=15}
dat_z <- left_join(modules, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC")) %>%
  filter(LUAD_vs_LUSC == "LUSC") %>%
  pivot_longer(cols = 1:7)

result <- lollipop_plot(dat_z,condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B,order = FALSE, order_vec = cur_order)
p_LUSC <- result$plot+ 
  ggtitle(paste0("Imaging modules for ", title_name," in LUSC"))+
  theme(axis.text.y = element_blank())+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))
p_LUSC

p <- plot_grid(p_LUAD, p_LUSC, nrow = 1, rel_widths = c(1, 0.6))
p

ggsave(plot = p, path = out_folder,filename = "3C_Imaging_modules_for_TPS.pdf",width = 8, height = 3.5,device = cairo_pdf)

```

# 3D inflammation for TPS 

### get gene signatures
```{r}
library(org.Hs.eg.db)
library(AnnotationDbi)
library(msigdbr)
library(singscore)

rna_mat <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/rna_mat_tpm.rds")


# ── gene signatures ────────────────────────────────────────────────────
hallmark_sets <- msigdbr(species = "Homo sapiens", collection = "H") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  split(.$gs_name) %>%
  lapply(function(x) x$gene_symbol)

# Targeted Hallmark gene sets of interest
target_sets <- c(
  "HALLMARK_INFLAMMATORY_RESPONSE",
  "HALLMARK_INTERFERON_GAMMA_RESPONSE",
  "HALLMARK_INTERFERON_ALPHA_RESPONSE",
  "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
  "HALLMARK_IL6_JAK_STAT3_SIGNALING",
  "HALLMARK_IL2_STAT5_SIGNALING"
)
gs_list_hallmark <- hallmark_sets[intersect(target_sets, names(hallmark_sets))]

# ── 2. Custom gene sets ────────────────────────────────────────────────────────
# CXCL9/10/11 chemokine axis (T cell recruitment)
gs_list_custom <- list(
  CXCL9_10_11_axis = c("CXCL9", "CXCL10", "CXCL11", "CXCR3",
                       "CXCL13", "CXCR5"),
  T_cell_activation = c("CD8A", "CD8B", "GZMB", "PRF1", "IFNG",
                        "TNF", "IL2", "PDCD1", "HAVCR2", "LAG3"),
  type_I_IFN = c("IFNB1", "IRF7", "MX1", "OAS1", "OAS2", "OAS3",
                 "ISG15", "IFIT1", "IFIT2", "IFIT3", "RSAD2", "HERC5"),
  NF_kB_target  = c("NFKB1", "RELA", "RELB", "NFKB2", "REL",
                    "BCL2L1", "IL1B", "IL6", "CXCL8", "TNF", "ICAM1"),
  TCR_signaling = c("CD3D", "CD3E", "CD3G", "ZAP70", "LCK", "LAT",
                    "ITK", "PLCG1", "NFATC1", "NFATC2"),
  cytotoxic = c("GZMB", "GZMA", "GZMK", "PRF1", "NKG7", "IFNG", "KLRG1"),
  exhaustion = c("PDCD1","LAG3","HAVCR2","TX","CTLA4", "IDO1")
  )

all_gs <- c(gs_list_hallmark, gs_list_custom)

# compute score
rna_for_score <- t(rna_mat)

# Filter gene sets to genes present in the RNA-seq matrix
all_gs <- lapply(all_gs, function(g) intersect(g, rownames(rna_for_score)))
all_gs <- all_gs[sapply(all_gs, length) >= 4]



overlap <- sapply(all_gs, function(g) sum(g %in% rownames(rna_for_score)))
total   <- sapply(all_gs, length)
print(data.frame(found = overlap, total = total, pct = round(100 * overlap / total)))

# ── Score with singscore ──────────────────────────────────────────────────────
ranked <- rankGenes(rna_for_score)

scores_list <- lapply(names(all_gs), function(nm) {
  sc <- simpleScore(ranked, upSet = all_gs[[nm]], centerScore = TRUE)
  setNames(sc$TotalScore, rownames(sc))
})


rna_scores <- as.data.frame(do.call(cbind, scores_list))
colnames(rna_scores) <- names(all_gs)
rownames(rna_scores) <- colnames(rna_for_score)
rna_scores$immucan_id <- rownames(rna_scores)
rownames(rna_scores) <- NULL

```

## LUAD 
```{r }
dat_z <- left_join(rna_scores, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC")) %>%
  filter(LUAD_vs_LUSC == "LUAD") %>%
  pivot_longer(cols = 1:11)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B, zscore_data = TRUE)

p_LUAD <- result$plot+
    ylab("Inf. signatures")+
    ggtitle(paste0("Inf. signatures for ", title_name," in LUAD"))+
  theme(axis.title.y = element_blank())+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))+
  xlab("Δ enrichment score" )

p_LUAD

cur_order <- result$data$name
```

## LUSC
```{r, fig.height=5, fig.width=15}
dat_z <- left_join(rna_scores, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC")) %>%
  filter(LUAD_vs_LUSC == "LUSC") %>%
  pivot_longer(cols = 1:11)

result <- lollipop_plot(dat_z,condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B, order = FALSE, order_vec = cur_order, zscore_data = TRUE)

p_LUSC <- result$plot+
    ylab("Inf. signatures")+
  ggtitle(paste0("Inf. signatures for ", title_name," in LUSC"))+
  theme(axis.text.y = element_blank())+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))+
  xlab("Δ enrichment score" )
p_LUSC

p <- plot_grid(p_LUAD, p_LUSC, nrow = 1, rel_widths = c(1, 0.6))

ggsave(plot = p, path = out_folder,filename = "3D_inflammation_signatures_for_TPS.pdf",width = 12, height = 4.5,device = cairo_pdf)
```


# 3F - Cox model for infiltration

```{r}
inf_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/infiltration_harmonized.rds"))

# compare densities of infiltrating, bystanding and ignored from IMC1, IMC2 and mIF
inf_dat <- inf_dat %>%
  group_by(immucan_id,tech) %>%
  mutate(total_CD8 = sum(bystanding,ignored,infiltrating,na.rm = TRUE)/im_size,
         bystanding = round(bystanding/im_size),
         ignored = round(ignored/im_size),
         infiltrating = round(infiltrating/im_size)) %>%
  ungroup()


p1 <- inf_dat %>%
  ggplot()+
  geom_density(aes(x = bystanding, col = tech), lwd = 1)+
  ggtitle("bystanding")
p2 <- inf_dat %>%
  ggplot()+
  geom_density(aes(x = ignored, col = tech), lwd = 1)+
  ggtitle("ignored")
p3 <- inf_dat %>%
  ggplot()+
  geom_density(aes(x = infiltrating, col = tech), lwd = 1)+
  ggtitle("infiltrating")

plot_grid(p1, p2,p3)

# define interactor over bystander ratio
inf_dat <- inf_dat %>%
  group_by(immucan_id, tech) %>%
  mutate(int_over_bystander = (infiltrating+bystanding)/ignored)
  
inf_dat %>%
  ggplot()+
  geom_density(aes(x = int_over_bystander, col = tech), lwd = 1)+
  ggtitle("infiltrating")

inf_dat[(is.infinite(inf_dat$int_over_bystander)),]$int_over_bystander <- NA

# we will average the ratios across technologies
inf_dat <- inf_dat %>%
  group_by(immucan_id) %>%
  mutate(mean_inf_ratio = mean(int_over_bystander,na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(mean_inf_ratio_dicho = ntile(mean_inf_ratio,2),
         mean_inf_ratio_dicho = ifelse(mean_inf_ratio_dicho == 2, "high", "low"))

res <- left_join(inf_dat %>% dplyr::select(immucan_id,mean_inf_ratio_dicho) %>% distinct(), traits,"immucan_id")

cm_LUSC <- coxph( Surv(time_to_event, censoring) ~  mean_inf_ratio_dicho ,data = as.data.frame(res) %>% filter(simple_histology == "Squamous cell carcinoma"))
cm_LUAD <- coxph( Surv(time_to_event, censoring) ~  mean_inf_ratio_dicho ,data = as.data.frame(res) %>% filter(simple_histology == "Adenocarcinoma"))

dfLUAD <- data.frame(histology = "LUAD",
                     HR = summary(cm_LUAD)$coefficients[,"exp(coef)"],
                     p.value = summary(cm_LUAD)$coefficients[, "Pr(>|z|)"],
                     lower_conf = summary(cm_LUAD)$conf.int[, "lower .95"],
                     upper_conf = summary(cm_LUAD)$conf.int[, "upper .95"]) 
dfLUSC <- data.frame(histology = "LUSC",
                     HR = summary(cm_LUSC)$coefficients[,"exp(coef)"],
                     p.value = summary(cm_LUSC)$coefficients[, "Pr(>|z|)"],
                     lower_conf = summary(cm_LUSC)$conf.int[, "lower .95"],
                     upper_conf = summary(cm_LUSC)$conf.int[, "upper .95"]) 
df1 <- rbind(dfLUAD,dfLUSC)

p1 <- df1 %>%
  ggplot(aes(x = HR,y = histology, col = histology))+
  geom_point(size = 2)+
  geom_errorbar(aes(xmin=lower_conf, xmax = upper_conf),lwd =1,width = 0.3)+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust= 1))+
  #scale_y_discrete(labels = function(x) str_split_i(x,"_",2))+
  scale_x_log10()+
  theme_bw()+
  geom_vline(xintercept = 1,linetype = "dashed")+
  geom_text(aes(label = paste0("p : ",round(p.value,digits = 3))),
            col = "black",
            vjust = -1,
            inherit.aes = TRUE)+
  scale_color_manual(values = c("LUAD" = "blue", "LUSC" = "orange"))+
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        legend.position = "none",
        axis.title.y = element_blank(),
        plot.title = element_text(hjust = 0.5))+
  ggtitle("HR for low infiltration ratio")
  p1

ggsave(filename = paste0(out_folder,"3F_cox_model_infiltration.pdf"),
       plot = p1,
       device = "pdf",
       width = 3,
       height = 2)
```

check if infiltration ratio also works for each individual modality
```{r}
res <- inf_dat %>%
  filter(tech  == "IMC1") %>%
  group_by(immucan_id) %>%
  ungroup() %>%
  mutate(mean_inf_ratio_dicho = ntile(int_over_bystander,2),
         mean_inf_ratio_dicho = ifelse(mean_inf_ratio_dicho == 2, "high", "low"))

res <- left_join(res, traits,"immucan_id")
p1 <- ggforest(coxph( Surv(time_to_event, censoring) ~  mean_inf_ratio_dicho ,data = as.data.frame(res) %>% filter(simple_histology == "Squamous cell carcinoma")))
p2 <- ggforest(coxph( Surv(time_to_event, censoring) ~ mean_inf_ratio_dicho,data = as.data.frame(res) %>% filter(simple_histology == "Adenocarcinoma")))
p <- plot_grid(p1, p2,ncol = 1)
p

# check individual techs IMC2
res <- inf_dat %>%
  filter(tech  == "IMC2") %>%
  group_by(immucan_id) %>%
  ungroup() %>%
  mutate(mean_inf_ratio_dicho = ntile(int_over_bystander,2),
         mean_inf_ratio_dicho = ifelse(mean_inf_ratio_dicho == 2, "high", "low"))

res <- left_join(res, traits,"immucan_id")
p1 <- ggforest(coxph( Surv(time_to_event, censoring) ~  mean_inf_ratio_dicho ,data = as.data.frame(res) %>% filter(simple_histology == "Squamous cell carcinoma")))
p2 <- ggforest(coxph( Surv(time_to_event, censoring) ~ mean_inf_ratio_dicho,data = as.data.frame(res) %>% filter(simple_histology == "Adenocarcinoma")))
p <- plot_grid(p1, p2,ncol = 1)
p

# check individual techs mIF
res <- inf_dat %>%
  filter(tech  == "mIF") %>%
  ungroup() %>%
  mutate(mean_inf_ratio_dicho = ntile(int_over_bystander,2),
         mean_inf_ratio_dicho = ifelse(mean_inf_ratio_dicho == 2, "high", "low"))

res <- left_join(res, traits,"immucan_id")
p1 <- ggforest(coxph( Surv(time_to_event, censoring) ~  mean_inf_ratio_dicho ,data = as.data.frame(res) %>% filter(simple_histology == "Squamous cell carcinoma")))
p2 <- ggforest(coxph( Surv(time_to_event, censoring) ~ mean_inf_ratio_dicho,data = as.data.frame(res) %>% filter(simple_histology == "Adenocarcinoma")))
p <- plot_grid(p1, p2,ncol = 1)
p

# correlation mIF CD8 and TPS
res <- left_join(inf_dat,traits,"immucan_id")
res %>%
  filter(tech == "mIF") %>%
  ggplot(aes(x = TPS, y = total_CD8))+
  geom_point()
```



# 3G tumor type enrichment TPS

```{r}
scores <- read.csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/tumor_subtype_scores.csv")

scores <- left_join(scores, traits,"immucan_id")

p <- scores %>%
  pivot_longer(cols = 1:3) %>% 
  filter(LUAD_vs_LUSC == "Adenocarcinoma",
         !is.na(TPS_score)) %>%
  ggplot(aes(x = TPS_score, y = value, fill = TPS_score))+
  geom_boxplot()+
  geom_jitter(aes(col = TPS_score),position = position_jitterdodge(jitter.width = 0.2),size= 0.5)+
  scale_fill_manual(values = alpha(c("positive" = "orange", "negative" = "darkgreen"),alpha = 0.6))+
  scale_color_manual(values = c("positive" = "orange", "negative" = "darkgreen"))+
  facet_wrap(~name)+
  stat_compare_means(method = "t.test",label.y = 1, size = 3,label = "p.format")+
  theme_bw()+
  coord_cartesian(ylim = c(-0.7, 1))+
  xlab("TPS score")+
  ylab("Tumor type score")+
  theme(axis.text.x = element_text(angle = 90, vjust = .05, hjust = 1),
        axis.title.x = element_blank())
  
p
ggsave(plot = p, path = out_folder,filename = "3G_tumor_types_for_TPS_LUAD.pdf",width = 4, height = 3.5,device = cairo_pdf)
```

# 3H tumor type enrichment TMB

```{r}
scores <- read.csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/tumor_subtype_scores.csv")

scores <- left_join(scores, traits,"immucan_id")

p <- scores %>%
  pivot_longer(cols = 1:3) %>% 
  filter(LUAD_vs_LUSC == "Adenocarcinoma",
         !is.na(TMB_score)) %>%
  ggplot(aes(x = TMB_score, y = value, fill = TMB_score))+
  geom_boxplot()+
  geom_jitter(aes(col = TMB_score),position = position_jitterdodge(jitter.width = 0.2),size= 0.5)+
  scale_fill_manual(values = alpha(c("high" = "blue", "low" = "red"),alpha = 0.6))+
  scale_color_manual(values = c("high" = "blue", "low" = "red"))+
  facet_wrap(~name)+
  stat_compare_means(method = "t.test",label.y = 1, size = 3,label = "p.format")+
  theme_bw()+
  coord_cartesian(ylim = c(-0.7, 1))+
  xlab("TPS score")+
  ylab("Tumor type score")+
  theme(axis.text.x = element_text(angle = 90, vjust = .05, hjust = 1),
        axis.title.x = element_blank())
  
p
ggsave(plot = p, path = out_folder,filename = "3H_tumor_types_for_TMB_LUAD.pdf",width = 4, height = 3.5,device = cairo_pdf)
```



# TMB stats

```{r}
unique(traits$TMB_score)
cur_var <- "TMB_score"
var_A <- "high"
var_B <-  "low"
col_A <- "blue"
col_B <- "red"
title_name <- "TMB"
```

```{r}
table(traits$TMB_score,traits$LUAD_vs_LUSC)
```

# 3I RNA for TMB

## LUAD
```{r}
dat_z <- left_join(rna, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC")) %>%
  filter(LUAD_vs_LUSC == "LUAD") %>%
  pivot_longer(cols = 1:29)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B, zscore_data = TRUE)
p_LUAD <- result$plot +
  ggtitle(paste0("RNA modules for ", title_name," in LUAD"))+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))
p_LUAD
cur_order <- result$data$name

```

### LUSC
```{r, fig.height=5, fig.width=15}
dat_z <- left_join(rna, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC")) %>%
  filter(LUAD_vs_LUSC == "LUSC") %>%
  pivot_longer(cols = 1:29)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B,order = FALSE, order_vec = cur_order,zscore_data = TRUE)
p_LUSC <- result$plot+ 
  ggtitle(paste0("RNA modules for ", title_name," in LUSC"))+
  theme(axis.text.y = element_blank())+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))
p_LUSC

p <- plot_grid(p_LUAD, p_LUSC, nrow = 1, rel_widths = c(1, 0.6))

ggsave(plot = p, path = out_folder,filename = "3I_RNA modues_for_TPS.pdf",width = 10, height = 6,device = cairo_pdf)
```

# 3J Imaging modules for TMB

## LUAD
```{r}
dat_z <- left_join(modules, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC")) %>%
  filter(LUAD_vs_LUSC == "LUAD") %>%
  pivot_longer(cols = 1:7)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B)
p_LUAD <- result$plot +
  ggtitle(paste0("Imaging modules for ", title_name," in LUAD"))+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))
p_LUAD
cur_order <- result$data$name
result$data
```

### LUSC
```{r, fig.height=5, fig.width=15}
dat_z <- left_join(modules, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC")) %>%
  filter(LUAD_vs_LUSC == "LUSC") %>%
  pivot_longer(cols = 1:7)

result <- lollipop_plot(dat_z,condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B,order = FALSE, order_vec = cur_order)
p_LUSC <- result$plot+ 
  ggtitle(paste0("Imaging modules for ", title_name," in LUSC"))+
  theme(axis.text.y = element_blank())+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))
p_LUSC

p <- plot_grid(p_LUAD, p_LUSC, nrow = 1, rel_widths = c(1, 0.6))
p

ggsave(plot = p, path = out_folder,filename = "3J_Imaging_modules_for_TMB.pdf",width = 8, height = 3.5,device = cairo_pdf)
```

# 3K inflammation  for TMB

## get gene signatures
```{r}
library(org.Hs.eg.db)
library(AnnotationDbi)
library(msigdbr)
library(singscore)

rna_mat <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/rna_mat_tpm.rds")


# ── gene signatures ────────────────────────────────────────────────────
hallmark_sets <- msigdbr(species = "Homo sapiens", collection = "H") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  split(.$gs_name) %>%
  lapply(function(x) x$gene_symbol)

# Targeted Hallmark gene sets of interest
target_sets <- c(
  "HALLMARK_INFLAMMATORY_RESPONSE",
  "HALLMARK_INTERFERON_GAMMA_RESPONSE",
  "HALLMARK_INTERFERON_ALPHA_RESPONSE",
  "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
  "HALLMARK_IL6_JAK_STAT3_SIGNALING",
  "HALLMARK_IL2_STAT5_SIGNALING"
)
gs_list_hallmark <- hallmark_sets[intersect(target_sets, names(hallmark_sets))]

# ── 2. Custom gene sets ────────────────────────────────────────────────────────
# CXCL9/10/11 chemokine axis (T cell recruitment)
gs_list_custom <- list(
  CXCL9_10_11_axis = c("CXCL9", "CXCL10", "CXCL11", "CXCR3",
                       "CXCL13", "CXCR5"),
  T_cell_activation = c("CD8A", "CD8B", "GZMB", "PRF1", "IFNG",
                        "TNF", "IL2", "PDCD1", "HAVCR2", "LAG3"),
  type_I_IFN = c("IFNB1", "IRF7", "MX1", "OAS1", "OAS2", "OAS3",
                 "ISG15", "IFIT1", "IFIT2", "IFIT3", "RSAD2", "HERC5"),
  NF_kB_target  = c("NFKB1", "RELA", "RELB", "NFKB2", "REL",
                    "BCL2L1", "IL1B", "IL6", "CXCL8", "TNF", "ICAM1"),
  TCR_signaling = c("CD3D", "CD3E", "CD3G", "ZAP70", "LCK", "LAT",
                    "ITK", "PLCG1", "NFATC1", "NFATC2"),
  cytotoxic = c("GZMB", "GZMA", "GZMK", "PRF1", "NKG7", "IFNG", "KLRG1"),
  exhaustion = c("PDCD1","LAG3","HAVCR2","TX","CTLA4", "IDO1")
  )

all_gs <- c(gs_list_hallmark, gs_list_custom)

# compute score
rna_for_score <- t(rna_mat)

# Filter gene sets to genes present in the RNA-seq matrix
all_gs <- lapply(all_gs, function(g) intersect(g, rownames(rna_for_score)))
all_gs <- all_gs[sapply(all_gs, length) >= 4]



overlap <- sapply(all_gs, function(g) sum(g %in% rownames(rna_for_score)))
total   <- sapply(all_gs, length)
print(data.frame(found = overlap, total = total, pct = round(100 * overlap / total)))

# ── Score with singscore ──────────────────────────────────────────────────────
ranked <- rankGenes(rna_for_score)

scores_list <- lapply(names(all_gs), function(nm) {
  sc <- simpleScore(ranked, upSet = all_gs[[nm]], centerScore = TRUE)
  setNames(sc$TotalScore, rownames(sc))
})


rna_scores <- as.data.frame(do.call(cbind, scores_list))
colnames(rna_scores) <- names(all_gs)
rownames(rna_scores) <- colnames(rna_for_score)
rna_scores$immucan_id <- rownames(rna_scores)
rownames(rna_scores) <- NULL

```

### LUAD 
```{r }
dat_z <- left_join(rna_scores, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC")) %>%
  filter(LUAD_vs_LUSC == "LUAD") %>%
  pivot_longer(cols = 1:11)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B, zscore_data = TRUE)

p_LUAD <- result$plot+
    ylab("Inflammation signatures")+
    ggtitle(paste0("Inflammation signatures for ", title_name," in LUAD"))+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))

p_LUAD

cur_order <- result$data$name
```

### LUSC
```{r, fig.height=5, fig.width=15}
dat_z <- left_join(rna_scores, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC")) %>%
  filter(LUAD_vs_LUSC == "LUSC") %>%
  pivot_longer(cols = 1:11)

result <- lollipop_plot(dat_z,condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B, order = FALSE, order_vec = cur_order, zscore_data = TRUE)

p_LUSC <- result$plot+
    ylab("Inflammation signatures")+
  ggtitle(paste0("Inflammation signatures for ", title_name," in LUSC"))+
  theme(axis.text.y = element_blank())+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))

p_LUSC

p <- plot_grid(p_LUAD, p_LUSC, nrow = 1, rel_widths = c(1, 0.6))
p
ggsave(plot = p, path = out_folder,filename = "3K_inflammation_signatures_for_TMB.pdf",width = 12, height = 4.5,device = cairo_pdf)

```

# TMB difference within smokers

```{r}
dat_z <- left_join(modules, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC")) %>%
  filter(LUAD_vs_LUSC == "LUAD",
         smoking_status == "Current smoker") %>%
  pivot_longer(cols = 1:7)

table(dat_z$TMB_score)
result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B)
p_LUAD <- result$plot +
  ggtitle(paste0("Imaging modules for ", title_name," in LUAD"))+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))
p_LUAD
cur_order <- result$data$name
result$data
```

