---
title: "Figure S5_revised"
author: "Daniel Schulz"
date: "2024-11-15"
output: html_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggpubr)
library(SingleCellExperiment)
library(survivalAnalysis)
library(ggsurvfit)
library(survminer)
library(survival)
library(edgeR)
library(cowplot)
library(ggrepel)
library(igraph)
library(ggraph)
library(ggrepel)
library(scales)
library(ggnewscale)
```

```{r}
mount_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/"
out_folder <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/revised_figures/Figure_5_supp/"
old_mount <- "/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/"
if (!dir.exists(out_folder)) {
  dir.create(out_folder)
}
```

# load data
we will load the data we generated at the beginning of figure 2
```{r}
#dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/abundances_dat.rds"))
dens_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/densities_dat.rds"))
#cello_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/cello_dat.rds"))
clin_dat <- read.csv("/mnt/immucan_volume/clinical_data/Lung_NSCLC2/2026_final/clinical.csv")
traits <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/all_traits.csv"))

modules <- as.data.frame(readRDS( "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/patient_imaging_meta_cluster_scores_tpm.rds"))

rna <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/rna_scores_tpm.rds")
rna$immucan_id  <- rownames(rna)
rownames(rna) <- NULL

modules <- modules  %>%
  mutate(immucan_id = rownames(modules))
rownames(modules) <- NULL

# colors
stage_col = setNames(c("palevioletred3","darkseagreen3","dodgerblue"),c("I","II","III"))
stage_col_num = setNames(c("palevioletred3","darkseagreen3","dodgerblue"),c("1","2","3"))
hist_col_2 = setNames(c("blue","orange"),c("LUAD","LUSC"))
hist_col_3 = setNames(c("blue","orange", "gray40"),c("LUAD","LUSC","Other"))
```

# calculate tissue area from IF1
```{r chunk-4, fig.height = 20, fig.width = 20}
process_one_file <- function(x) {
  suppressPackageStartupMessages({
    library(SingleCellExperiment)
    library(concaveman)
    library(sf)
    library(stringr)
  })
  
  cur_sce <- readRDS(x)
  
  if (ncol(cur_sce) < 3) {
    stop("fewer than 3 cells (", ncol(cur_sce), ") - cannot build a hull")
  }
  if (!all(c("nucleus.x", "nucleus.y") %in% colnames(SingleCellExperiment::colData(cur_sce)))) {
    stop("missing nucleus.x/nucleus.y in colData")
  }
  
  cur_mat <- cbind(cur_sce$nucleus.x, cur_sce$nucleus.y)
  
  ok <- stats::complete.cases(cur_mat)
  if (sum(ok) < 3) stop("fewer than 3 non-NA coordinate pairs")
  if (any(!ok)) cur_mat <- cur_mat[ok, , drop = FALSE]
  
  cur_mat <- unique(cur_mat)
  if (nrow(cur_mat) < 3) stop("fewer than 3 distinct points after dedup")
  
  rng_x <- diff(range(cur_mat[, 1]))
  rng_y <- diff(range(cur_mat[, 2]))
  if (rng_x == 0 || rng_y == 0) stop("points are collinear (zero-area bounding box)")
  
  tumor_hull <- concaveman::concaveman(cur_mat, concavity = 2)
  
  if (!identical(tumor_hull[1, ], tumor_hull[nrow(tumor_hull), ])) {
    tumor_hull <- rbind(tumor_hull, tumor_hull[1, ])
  }
  if (nrow(tumor_hull) < 4) stop("degenerate hull (< 4 ring points after closing)")
  
  polygon <- sf::st_polygon(list(as.matrix(tumor_hull)))
  if (!sf::st_is_valid(polygon)) polygon <- sf::st_make_valid(polygon)
  
  data.frame(
    immucan_id  = stringr::str_split_i(basename(x), pattern = "-FIXT", 1),
    tissue_size = as.numeric(sf::st_area(polygon)) / 1e6,
    n_cells     = ncol(cur_sce),
    stringsAsFactors = FALSE
  )
}

# ---- crash-isolated + error-tolerant wrapper ----
# callr::r() runs the file in its OWN process. If concaveman/GEOS segfaults,
# only that subprocess dies - this wrapper catches it as a normal R condition.
safe_process_one_file <- function(x) {
  res <- tryCatch(
    callr::r(process_one_file, args = list(x = x), error = "error"),
    error = function(e) {
      list(error = TRUE, file = x, message = conditionMessage(e))
    }
  )
  if (is.list(res) && !is.null(res$error) && isTRUE(res$error)) {
    message("FAILED: ", basename(x), " -- ", res$message)
    return(NULL)
  }
  res
}

cur_files <- list.files(
  "/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/IF/IF1/tsv",
  pattern = ".rds$", full.names = TRUE
)

if(!file.exists("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/IF1_tissue_size.rds")){
  out.list <- bplapply(
    cur_files,
    safe_process_one_file,
    BPPARAM = MulticoreParam(workers = 6)
  )
  
  ok       <- !vapply(out.list, is.null, logical(1))
  tissue_df <- do.call(rbind, out.list[ok])
  failed_files <- cur_files[!ok]
  
  message(sum(ok), " / ", length(cur_files), " files succeeded")
  if (length(failed_files)) {
    message("Failed files:\n", paste(" -", failed_files, collapse = "\n"))
  }
  
  IF1 <- left_join(IF1, tissue_df, "immucan_id")
  saveRDS(IF1,"/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/IF1_tissue_size.rds")

}

# load tumor size data
IF1 <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/IF1_tissue_size.rds")
```


# B cell patches

## caculate B cell patch summaries
NOTE: the code is not evaluated automatically as the output is saved. we will
keep these files also at their original location to avoid duplicated data

```{r, fig.height=20, fig.width=20, eval=FALSE}
library(BiocParallel)
cur_files <- list.files(paste0(old_mount"IF/IF1/tsv"), 
                        pattern = ".rds$", full.names = TRUE)

bplapply(cur_files,function(x){
  cur_sce <- readRDS(x)
  
  if (is.numeric(cur_sce$TLS.ID)) {
    dat <- as.data.frame(colData(cur_sce))
    sample_id <- unique(cur_sce$sample_id)

    dat <- dat %>%
      dplyr::select(sample_id,celltype, TLS.ID,cell.ID, in.ROI.tumor_tissue, nucleus.x,nucleus.y, dist_to_tumor)

    # here we will collect the information of the size of TLS and their
    # location. We will identify the location based on subset of cells in
    # the tumor_tissue
    dat <- dat  %>%
      filter(!is.na(TLS.ID)) %>%
      group_by(TLS.ID, in.ROI.tumor_tissue) %>%
      summarise(size = n(),
                TLS_dist_to_tumor = mean(dist_to_tumor)) %>%
      mutate(tissue_location = "in_tumor",
             tissue_location = ifelse(TLS_dist_to_tumor > -500 & TLS_dist_to_tumor < 500, "boundary",tissue_location),
             tissue_location = ifelse(TLS_dist_to_tumor > 500, "outside_tumor",tissue_location))


    cell_props <- as.data.frame(colData(cur_sce))
    cell_props <- cell_props %>%
      filter(!is.na(TLS.ID)) %>%
      dplyr::select(sample_id,celltype, TLS.ID,cell.ID, in.ROI.tumor_tissue, nucleus.x,nucleus.y) %>%
      group_by(TLS.ID) %>%
      dplyr::count(celltype) %>%
      mutate(total_cellcount = sum(n),
             frac = n/total_cellcount) %>%
      pivot_wider(id_cols = c(TLS.ID),names_from = celltype,values_from = frac, values_fill = 0)


    dat <- left_join(dat,cell_props,"TLS.ID")
    dat$location <- NULL
    dat$sample_id <- sample_id

  }
  # No TLS in sample
  else {
    dat <- data.frame(TLS.ID = NA,
                      size = NA,
                      TLS_dist_to_tumor =NA,
                      tissue_location = NA,
                      B = NA,
                      BnT = NA,
                      DC = NA,
                      `T`= NA,
                      other = NA,
                      Tumor = NA,
                      MacCD163 = NA,
                      Tumor_CD15 = NA,
                      sample_id = unique(cur_sce$sample_id))
                      
  }
  
  write_csv(dat,paste0(old_mount,"IF/Rout/IF1_features/TLS_summary/",unique(dat$sample_id),".csv"))
}, BPPARAM = MulticoreParam(workers = 12))
gc()
```

## calculate compartment based cell counts for tumor and outside only

Here we calculate the number of cells in the normal and tumor tissue

```{r, fig.height=20, fig.width=20, eval=FALSE}
cur_files <- list.files(paste0(old_mount,"IF/IF1/tsv"),
                        pattern = ".rds$", full.names = TRUE)

library(BiocParallel)
if (!file.exists(paste0(mount_path,"Figures/scores_for_sharing/dist_dat_all_cells_inTumor_outside.rds"))) {
    cur_list <- bplapply(cur_files,function(x){
    cur_sce <- readRDS(x)
    
    
    cur_dat <- as.data.frame(colData(cur_sce))
    sample_id <- unique(cur_sce$sample_id)
    rownames(cur_dat) <- NULL
    
    cur_dat <- cur_dat %>%
      dplyr::select(sample_id,celltype, dist_to_tumor) %>%
      mutate(location = ifelse(dist_to_tumor < 500, "in_tumor", "outside")) %>%
      group_by(celltype,location) %>%
      summarise(n = n(), .groups = "drop") %>%
      mutate(sample_id = sample_id)
    
    return(cur_dat)
    
  }, BPPARAM = MulticoreParam(workers = 12))
  gc()
  dist_dat <- do.call(rbind, cur_list)
  saveRDS(dist_dat,paste0(mount_path,"Figures/scores_for_sharing/dist_dat_all_cells_inTumor_outside.rds"))
}

```



# S5A B cell patch image - alternative image
```{r, fig.height=40, fig.width=40}
# old image
#cur_sce_2 <- readRDS(paste0(old_mount,"IF/IF1/tsv/LUNG-NSCLC2-0565-FIXT-01-IF1-01_#_cells_properties_#_12cf4d92f8de6204055505affdf85e90.rds"))

cur_sce_2 <- readRDS(paste0(old_mount,"IF/IF1/tsv/LUNG-NSCLC2-0708-FIXT-01-IF1-01_#_cells_properties_#_35821bbb8888909a391b6cb7ce34df49.rds"))
# we will assign all cells within 500 µm of the tumor annotation to tumor
cur_sce_2[,cur_sce_2$dist_to_tumor < 500]$in.ROI.tumor_tissue <- TRUE

cur_sce_2$TLS <- !is.na(cur_sce_2$TLS.ID)
df <- as.data.frame(colData(cur_sce_2))
df <- df %>%
  mutate(TLS = ifelse(celltype %in% c("Tumor","Tumor_CD15"),"Tumor cells","Stroma cells"),
         TLS = ifelse(in.ROI.tumor_tissue == FALSE,"Healthy",TLS),
         TLS = ifelse(!is.na(TLS.ID),"B cell patch",TLS))
colData(cur_sce_2) <- DataFrame(df)

p <- imcRtools::plotSpatial(cur_sce_2,img_id = "sample_id",coords = c("nucleus.x","nucleus.y"),node_color_by = "TLS",node_size_fix = 0.2)+
  scale_colour_manual(values = c("Healthy" = "gray", "Stroma cells" = "darkred", "Tumor cells" = "magenta", "B cell patch" = "darkorange1"))
#p2 <- imcRtools::plotSpatial(cur_sce_2,img_id = "sample_id",coords = c("nucleus.x","nucleus.y"),node_color_by = "celltype",node_size_fix = 0.2)+
#  scale_colour_manual(values = col_vec)

p <- p +
  theme(axis.text.x = element_text(size = 15),
        axis.text.y = element_text(size = 15),
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 15))+
  guides(color = guide_legend(override.aes = list(size = 5)))
#p2
p
ggsave(filename = paste0(out_folder,"S5A_TLS_example_alternative.png"),
       plot = p,
       device = "png",
       width = 4000,
       height = 4000,
       units = "px")
```


## patients with TLS 
percentage of patients at least 10 patches of 100 B cells
```{r}
cur_files <- list.files(paste0(old_mount,"IF/Rout/IF1_features/TLS_summary"),full.names = TRUE)
cur_list <- list()
for (i in cur_files) {
  cur_file <- read.csv(i, check.names = FALSE)
  cur_list[[i]] <- cur_file %>%
    pivot_longer(cols = -c(colnames(cur_file)[1:4],colnames(cur_file)[ncol(cur_file)]))

}

TLS_summary <- do.call(rbind,cur_list)


TLS_pats <- TLS_summary %>%
  as.data.frame() %>%
  mutate(immucan_id = str_split_i(sample_id,"-FIXT",1),
         B_count = size*value) %>%
  filter(name == "B") %>%
  select(immucan_id,B_count, TLS.ID) %>%
  distinct() %>%
  filter(B_count > 100) %>%
  dplyr::count(immucan_id) %>%
  filter(n >10) %>%
  ungroup() %>%
  left_join(clin_dat[,c("immucan_id","simple_histology")], by = "immucan_id")

cat("patients with at least ten B patch of 100 cells:",dim(TLS_pats)[1]/length(unique(clin_dat$immucan_id)))
df <- data.frame(hist_count = unclass(table(clin_dat$simple_histology)))

# per histology
Adeno_pats <- TLS_pats %>%
  filter(simple_histology == "Adenocarcinoma")
dim(Adeno_pats)[1]/df["Adenocarcinoma",]

LUSC_pats <- TLS_pats %>%
  filter(simple_histology == "Squamous cell carcinoma")
dim(LUSC_pats)[1]/df["Squamous cell carcinoma",]
```


# calculate B cell patch abundance in compartments

```{r}
dist_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/dist_dat_all_cells_inTumor_outside.rds"))
dist_dat <- dist_dat %>%
  mutate(immucan_id = str_split_i(sample_id,"-FIXT",1)) %>%
  select(-sample_id)

# check the fraction of B cells across compartments
dist_dat %>%
  group_by(immucan_id, location) %>%
  mutate(total_cells = sum(n),
         frac = n/total_cells) %>%
  ungroup() %>%
  filter(celltype == "B") %>%
  ggplot(aes(x = location,y = frac))+
  geom_boxplot()+
  theme(legend.position = "none")+
  stat_compare_means()

# count the number of TLS in each compartment
TLS_dat <- TLS_summary %>%
  as.data.frame() %>%
  mutate(immucan_id = str_split_i(sample_id,"-FIXT",1)) %>%
  dplyr::select(immucan_id,size, TLS.ID,TLS_dist_to_tumor) %>%
  distinct() %>%
 mutate(location = ifelse(TLS_dist_to_tumor > 500, "outside","in_tumor")) %>%
  group_by(immucan_id) %>%
  dplyr::count(location,name = "TLS_count") %>%
  filter(!is.na(location))
```

# merge all TLS and HE info

```{r}
TLS_dat

# HE data
dat <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/HE/hooknetTLS/filtered-20260817T121553Z-1-001/filtered_geojson/annotation_summary.rds")

dat <- dat %>%
  mutate(sample  = str_split_i(sample,"_#_",1)) %>%
  group_by(sample) %>%
  dplyr::count(class) %>%
  arrange(desc(n))

dat <- dat %>%
  ungroup() %>%
  pivot_wider(id_cols = sample,names_from = class, values_from = n) %>%
  mutate(immucan_id = str_split_i(sample,"-FIXT",1),
         HE_section = str_split_i(sample,"-FIXT-01-",2))

dup_entries <- dat %>%
  group_by(immucan_id) %>%
  summarise(n = n()) %>%
  filter(n > 1) %>%
  pull(immucan_id)


#THere are duplicated HE files: manual selection
keep_ids <- 
  c("LUNG-NSCLC2-0613-FIXT-01-HES-02","LUNG-NSCLC2-0614-FIXT-01-HES-02","LUNG-NSCLC2-0616-FIXT-01-HES-02","LUNG-NSCLC2-0617-FIXT-01-HES-02","LUNG-NSCLC2-0621-FIXT-01-HES-02",
  "LUNG-NSCLC2-0622-FIXT-01-HES-02","LUNG-NSCLC2-0623-FIXT-01-HES-02","LUNG-NSCLC2-0628-FIXT-01-HES-02","LUNG-NSCLC2-0629-FIXT-01-HES-02",
  "LUNG-NSCLC2-0630-FIXT-01-HES-02","LUNG-NSCLC2-0631-FIXT-01-HES-02","LUNG-NSCLC2-0632-FIXT-01-HES-02","LUNG-NSCLC2-0633-FIXT-01-HES-02",
  "LUNG-NSCLC2-0634-FIXT-01-HES-02","LUNG-NSCLC2-0635-FIXT-01-HES-01","LUNG-NSCLC2-0636-FIXT-01-HES-02","LUNG-NSCLC2-0636-FIXT-01-HES-02",
  "LUNG-NSCLC2-0639-FIXT-01-HES-02","LUNG-NSCLC2-0642-FIXT-01-HES-02","LUNG-NSCLC2-0643-FIXT-01-HES-02","LUNG-NSCLC2-0644-FIXT-01-HES-02",
  "LUNG-NSCLC2-0646-FIXT-01-HES-02","LUNG-NSCLC2-0647-FIXT-01-HES-02","LUNG-NSCLC2-0648-FIXT-01-HES-02","LUNG-NSCLC2-0650-FIXT-01-HES-02",
  "LUNG-NSCLC2-0651-FIXT-01-HES-02","LUNG-NSCLC2-0652-FIXT-01-HES-02","LUNG-NSCLC2-0657-FIXT-01-HES-02","LUNG-NSCLC2-0658-FIXT-01-HES-02",
  "LUNG-NSCLC2-0660-FIXT-01-HES-02","LUNG-NSCLC2-0662-FIXT-01-HES-01","LUNG-NSCLC2-0663-FIXT-01-HES-01","LUNG-NSCLC2-0665-FIXT-01-HES-02",
  "LUNG-NSCLC2-0667-FIXT-01-HES-02","LUNG-NSCLC2-0670-FIXT-01-HES-02","LUNG-NSCLC2-0672-FIXT-01-HES-02","LUNG-NSCLC2-0673-FIXT-01-HES-02",
  "LUNG-NSCLC2-0676-FIXT-01-HES-02")

dat <- dat %>%
  mutate( keep = ifelse(immucan_id %in% dup_entries & !sample %in% keep_ids, FALSE, TRUE ))%>%
  filter(keep == TRUE)

dat <- left_join(TLS_dat,dat, "immucan_id")

# fill NAs with 0
dat <- dat %>%
  pivot_wider(id_cols = c(immucan_id,tls, gc) ,names_from = location, values_from = TLS_count) %>%
   mutate(gc = ifelse(is.na(gc),0, gc),
         tls = ifelse(is.na(tls),0, tls),
         outside = ifelse(is.na(outside),0, outside),
         in_tumor = ifelse(is.na(in_tumor),0, in_tumor))
head(dat)

# merge with tissue size information

dat <- dat %>%
  left_join(IF1,"immucan_id") %>%
  mutate(in_tumor_density = in_tumor/tumor_size,
         outside_density = outside/(tissue_size-tumor_size))
```

# S5C, D, E B cell patch TLS core
```{r, fig.height=4, fig.width=4}
p1 <- dat %>%
  mutate(total_Bcell_patch = sum(in_tumor,outside)) %>%
  ggplot(aes(x = tls, y = total_Bcell_patch))+
  geom_point()+
  theme_bw()+
  ylab("B cell patch count (IF1)")+
  xlab("TLS count (H&E)")+
  geom_smooth(method='lm', formula = y~x) +
  stat_cor(aes(label = paste(..r.label.., ..p.label.., sep = "~`,`~")),
           label.x = 5, label.y = 600, size = 4, method = "pearson")
p1
```

```{r, fig.width=4, fig.height=4}
p2 <- dat %>%
  mutate(total_Bcell_patch = sum(in_tumor,outside)) %>%
  ggplot(aes(x = gc, y = total_Bcell_patch))+
  geom_point()+
  theme_bw()+
  ylab("B cell patch count (IF1)")+
  xlab("GC count (H&E)")+
  geom_smooth(method='lm', formula = y~x) +
  stat_cor(aes(label = paste(..r.label.., ..p.label.., sep = "~`,`~")),
           label.x = 2.5, label.y = 600, size = 4, method = "pearson")
p2
```

```{r, fig.width=4, fig.height=4}
p3 <- dat %>%
  ggplot(aes(x = tls, y = gc))+
  geom_point()+
  theme_bw()+
  xlab("TLS count (H&E)")+
  ylab("GC count (H&E)")+
  geom_smooth(method='lm', formula = y~x) +
  stat_cor(aes(label = paste(..r.label.., ..p.label.., sep = "~`,`~")),
           label.x = 5, label.y = 15, size = 4, method = "pearson")
p3
```

```{r}
p <- cowplot::plot_grid(p1, p2, p3, nrow = 1)
ggsave(plot = p,filename = "S5C_D_E.pdf",device = "pdf",path = out_folder,width = 8, height = 3)
```


# S5F size of patches by location
check if patches are larger in border
```{r}
p <- TLS_summary %>%
  as.data.frame() %>%
  mutate(immucan_id = str_split_i(sample_id,"-FIXT",1)) %>%
  select(immucan_id,size, TLS.ID,TLS_dist_to_tumor) %>%
  distinct() %>%
  mutate(location = ifelse(TLS_dist_to_tumor > 500, "outside","in_tumor")) %>%
  #group_by(immucan_id, location) %>%
  #summarise(mean_size = mean(size)) %>%
  filter(!is.na(location)) %>%
  ggplot(aes(y = (size),x = location, col= location))+
  geom_boxplot()+
  scale_y_log10()+
  stat_compare_means(label = "p.signif",method = "wilcox.test")+
   stat_summary(fun = median, geom = "text",
               aes(label = round(after_stat(y), 2), colour = location),
               vjust = -0.5, size = 6, show.legend = FALSE)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = .5, hjust = 1, size = 15),
        axis.text.y = element_text(size = 15),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 15),
        legend.position = "bottom",
        legend.text = element_text(size = 15),
        legend.title = element_text(size = 15))+
  scale_y_log10()+
  labs(x = "Location",y = "Patch size", color = "Location")+
  scale_color_manual(values = c("in_tumor" = "magenta", "outside" = "grey40"))+
  guides(color = guide_legend(nrow = 2))

p

ggsave(filename = paste0(out_folder,"S5F_patch_size.pdf"),
       plot = p,
       device = "pdf",
       width = 3,
       height = 5)
```



# S5G T and B cell proportions in Patches by location
composition of outside TLS
```{r, fig.height=6, fig.width=6}
p <- TLS_summary %>%
  as.data.frame() %>%
  filter(name %in% c("B","T")) %>%
  mutate(immucan_id = str_split_i(sample_id,"-FIXT",1)) %>%
  mutate(location = ifelse(TLS_dist_to_tumor > 500, "outside","in_tumor")) %>%
  filter(!is.na(location)) %>%
  ggplot(aes(x = location,y = value, fill=location))+
  geom_boxplot(outliers = TRUE)+
  stat_compare_means(label.y = 1, label = "p.signif")+
  facet_wrap(.~name, scales = "free",ncol = 2)+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = .5, hjust = 1, size = 14),
        axis.text.y = element_text(size = 14),
        axis.title.x = element_blank(),
        axis.title.y = element_text(size = 14),
        legend.text = element_text(size =14),
        legend.title = element_text(size =14),
        legend.position = "bottom",
        strip.text = element_text(size = 14))+
  scale_fill_manual(values = c("in_tumor" = "magenta","outside" = "gray40"))+
  labs(y = "Frequency of cell type in patch")+
  coord_cartesian(ylim = c(0,1.1))

p

ggsave(filename = paste0(out_folder,"S5G_B_and_T_cell_by_location.pdf"),
       plot = p,
       device = "pdf",
       width = 6,
       height = 5)
```


# S5H imaging modules association with B cell patches

```{r}
cur_mod <- modules %>%
  filter(immucan_id %in% dat$immucan_id)

cur_dat <- dat[match(cur_mod$immucan_id,dat$immucan_id),]

identical(cur_dat$immucan_id,cur_mod$immucan_id)

# -----------------------------------------------------------------------------
# 1. Extract numeric matrices
# -----------------------------------------------------------------------------
imc_mat   <- cur_mod %>% dplyr::select(-immucan_id) %>% as.matrix()
gc_mat <- cur_dat %>% ungroup() %>% dplyr::select(-immucan_id)  %>% as.matrix()

imc_cols   <- colnames(imc_mat)
gc_cols <- colnames(gc_mat)[1:4]

cat("IMC features: ",   length(imc_cols),   "\n")
cat("TLS and GC: ", length(gc_cols), "\n")
cat("Total tests: ",    length(imc_cols) * length(gc_cols), "\n")


# -----------------------------------------------------------------------------
# 2. Compute all pairwise Spearman correlations
# -----------------------------------------------------------------------------
cor_list <- vector("list", length(imc_cols) * length(gc_cols))
idx <- 1L

for (imc_feat in imc_cols) {
  for (gc_prot in gc_cols) {
    
    x <- imc_mat[, imc_feat]
    y <- gc_mat[, gc_prot]
    
    # Only test if both vectors have sufficient non-NA observations
    n_complete <- sum(complete.cases(x, y))
    
    if (n_complete >= 5) {
      ct <- cor.test(x, y, method = "spearman", exact = FALSE)
      cor_list[[idx]] <- tibble(
        imc_feature  = imc_feat,
        gc_feature = gc_prot,
        rho           = ct$estimate,
        pval          = ct$p.value,
        n             = n_complete
      )
    } else {
      cor_list[[idx]] <- tibble(
        imc_feature   = imc_feat,
        gc_feature = gc_prot,
        rho           = NA_real_,
        pval          = NA_real_,
        n             = n_complete
      )
    }
    idx <- idx + 1L
  }
}

cor_df <- bind_rows(cor_list)


# -----------------------------------------------------------------------------
# 3. Multiple testing correction
#    BH (FDR) correction across all tests; also Bonferroni for reference
# -----------------------------------------------------------------------------
cor_df <- cor_df %>%
  mutate(
    padj_BH          = p.adjust(pval, method = "BH"),
    padj_bonferroni  = p.adjust(pval, method = "bonferroni")
  ) %>%
  arrange(pval)

cat("\nSignificant associations (FDR < 0.05):", sum(cor_df$padj_BH < 0.05, na.rm = TRUE), "\n")
cat("Significant associations (FDR < 0.10):", sum(cor_df$padj_BH < 0.10, na.rm = TRUE), "\n")


# -----------------------------------------------------------------------------
# 4. Preview top hits
# -----------------------------------------------------------------------------
cat("\nTop 20 associations by p-value:\n")
print(cor_df %>%
        filter(!is.na(rho)) %>%
        mutate(across(c(rho, pval, padj_BH), \(x) signif(x, 3))))
```

## heatmap TLS with modules

```{r}
plot_df <- cor_df %>%
    mutate(label = case_when(pval < 0.001 & padj_BH < 0.1 ~ "***",
                           pval < 0.01  & padj_BH < 0.1 ~ "**",
                           pval < 0.05  & padj_BH < 0.1 ~ "*",
                           TRUE        ~ "")) %>%
  mutate(gc_feature = case_when(gc_feature == "outside" ~ "B patch outside tumor",
                                gc_feature == "in_tumor" ~ "B patch inside tumor",
                                gc_feature == "gc" ~ "H&E GC",
                                gc_feature == "tls" ~ "H&E TLS"))


lim <- max(abs(plot_df$rho), na.rm = TRUE)


p <- ggplot(plot_df, aes(gc_feature, imc_feature, fill = rho)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  geom_text(aes(label = label,
                colour = abs(rho) > 0.7 * lim,
                size   = grepl("^\\*+$", label))) +
  scale_fill_gradient2(low = "#2166AC", mid = "#F0F0F0", high = "#B2182B",
                       midpoint = 0, limits = c(-lim, lim),
                       name = "Spearman rho") +
  scale_colour_manual(values = c(`FALSE` = "black", `TRUE` = "white"), guide = "none") +
  scale_size_manual(values = c(`FALSE` = 2.1, `TRUE` = 4), guide = "none") +
  coord_fixed() +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 10) +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1),
        plot.caption = element_text(hjust = 0, size = 8, color = "grey30"))

p

ggsave(filename = paste0(out_folder,"S5H_image_module_TLS_correlation.pdf"),
       plot = p,
       device = "pdf",
       width = 4,
       height = 4)
```



# S5I RNA modules association with B cell patches

```{r}
cur_mod <- rna %>%
  filter(immucan_id %in% dat$immucan_id)

cur_dat <- dat[match(cur_mod$immucan_id,dat$immucan_id),]

identical(cur_dat$immucan_id,cur_mod$immucan_id)

# -----------------------------------------------------------------------------
# 1. Extract numeric matrices
# -----------------------------------------------------------------------------
imc_mat   <- cur_mod %>% dplyr::select(-immucan_id) %>% as.matrix()
gc_mat <- cur_dat %>% ungroup() %>% dplyr::select(-immucan_id)  %>% as.matrix()

imc_cols   <- colnames(imc_mat)
gc_cols <- colnames(gc_mat)[1:4]

cat("IMC features: ",   length(imc_cols),   "\n")
cat("TLS and GC: ", length(gc_cols), "\n")
cat("Total tests: ",    length(imc_cols) * length(gc_cols), "\n")


# -----------------------------------------------------------------------------
# 2. Compute all pairwise Spearman correlations
# -----------------------------------------------------------------------------
cor_list <- vector("list", length(imc_cols) * length(gc_cols))
idx <- 1L

for (imc_feat in imc_cols) {
  for (gc_prot in gc_cols) {
    
    x <- imc_mat[, imc_feat]
    y <- gc_mat[, gc_prot]
    
    # Only test if both vectors have sufficient non-NA observations
    n_complete <- sum(complete.cases(x, y))
    
    if (n_complete >= 5) {
      ct <- cor.test(x, y, method = "spearman", exact = FALSE)
      cor_list[[idx]] <- tibble(
        imc_feature  = imc_feat,
        gc_feature = gc_prot,
        rho           = ct$estimate,
        pval          = ct$p.value,
        n             = n_complete
      )
    } else {
      cor_list[[idx]] <- tibble(
        imc_feature   = imc_feat,
        gc_feature = gc_prot,
        rho           = NA_real_,
        pval          = NA_real_,
        n             = n_complete
      )
    }
    idx <- idx + 1L
  }
}

cor_df <- bind_rows(cor_list)


# -----------------------------------------------------------------------------
# 3. Multiple testing correction
#    BH (FDR) correction across all tests; also Bonferroni for reference
# -----------------------------------------------------------------------------
cor_df <- cor_df %>%
  mutate(
    padj_BH          = p.adjust(pval, method = "BH"),
    padj_bonferroni  = p.adjust(pval, method = "bonferroni")
  ) %>%
  arrange(pval)

cat("\nSignificant associations (FDR < 0.05):", sum(cor_df$padj_BH < 0.05, na.rm = TRUE), "\n")
cat("Significant associations (FDR < 0.10):", sum(cor_df$padj_BH < 0.10, na.rm = TRUE), "\n")


# -----------------------------------------------------------------------------
# 4. Preview top hits
# -----------------------------------------------------------------------------
cat("\nTop 20 associations by p-value:\n")
print(cor_df %>%
        filter(!is.na(rho)) %>%
        mutate(across(c(rho, pval, padj_BH), \(x) signif(x, 3))))
```

## heatmap TLS with modules

```{r, fig.height=10, fig.width=5}
plot_df <- cor_df %>%
    mutate(label = case_when(pval < 0.001 & padj_BH < 0.1 ~ "***",
                           pval < 0.01  & padj_BH < 0.1 ~ "**",
                           pval < 0.05  & padj_BH < 0.1 ~ "*",
                           TRUE        ~ "")) %>%
  mutate(gc_feature = case_when(gc_feature == "outside" ~ "B patch outside tumor",
                                gc_feature == "in_tumor" ~ "B patch inside tumor",
                                gc_feature == "gc" ~ "H&E GC",
                                gc_feature == "tls" ~ "H&E TLS"))


lim <- max(abs(plot_df$rho), na.rm = TRUE)

p <- ggplot(plot_df, aes(gc_feature, imc_feature, fill = rho)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  geom_text(aes(label = label,
                colour = abs(rho) > 0.7 * lim,
                size   = grepl("^\\*+$", label))) +
  scale_fill_gradient2(low = "#2166AC", mid = "#F0F0F0", high = "#B2182B",
                       midpoint = 0, limits = c(-lim, lim),
                       name = "Spearman rho") +
  scale_colour_manual(values = c(`FALSE` = "black", `TRUE` = "white"), guide = "none") +
  scale_size_manual(values = c(`FALSE` = 2.1, `TRUE` = 4), guide = "none") +
  coord_fixed() +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 10) +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
        axis.text.y = element_text(size = 14),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 14))
p
ggsave(filename = paste0(out_folder,"S5I_RNA_module_TLS_correlation.pdf"),
       plot = p,
       device = "pdf",
       width = 7,
       height = 10)
```


# S5J Enrichment plot

comparison of TMB, TPS, stage, histology, driver mutations with TLS, etc

```{r, fig.height=10, fig.width=5}
cor_cols   <- colnames(traits)[grepl(c("TPS$|TMB$"), colnames(traits))]
bin_cols <- c("TPS_score","TMB_score","age_dicho", "gender","stage_1_2","stage_1_3","stage_2_3","never_vs_current_smoker",
                "never_vs_Ex_smoker","current_vs_Ex_smoker", "N0_vs_Nx","M0_vs_Mx", "long_survivors","T1_vs_Tother")
condition <- c("positive","high","low", "Female","I","I","II","Current smoker","Non-smoker","Current smoker", "node_neg", "met_neg", "alive","T1")

y_labels <- c("TPS_positive", "TMB_highhigh","low_age","Female","stage_1_over_2","stage_1_over_3","stage_2_over_3","current_over_never-smoker","never_over_ex-smoker","Current_over_ex-smoker","node_negative","met_negative","alive","T1_tumor")
names(y_labels) <- bin_cols
names(condition) <- bin_cols

gc_cols <- c("tls","gc","in_tumor","outside")

library(dplyr)

min_n <- 5   # minimum per group; raise if you want

p.list <- list()

for (feat in names(condition)) {
  target <- condition[[feat]]          # [[ ]] not [ ]

  for (gc_prot in gc_cols) {

    cur_dat <- dat %>%
      select(immucan_id, value = all_of(gc_prot)) %>%
      left_join(traits %>% select(immucan_id, group = all_of(feat)),
                by = "immucan_id",
                relationship = "many-to-one") %>%   # errors if traits has dup ids
      filter(!is.na(value), !is.na(group))          # drop NAs BEFORE splitting

    y_test <- cur_dat$value[cur_dat$group == target]
    y_ref  <- cur_dat$value[cur_dat$group != target]

    n_test <- length(y_test)
    n_ref  <- length(y_ref)

    out <- data.frame(
      imc_feature = feat,
      gc_feature  = gc_prot,
      test_level  = as.character(target),
      n_test      = n_test,
      n_ref       = n_ref,
      median_test = if (n_test) median(y_test) else NA_real_,
      median_ref  = if (n_ref)  median(y_ref)  else NA_real_,
      mean_test   = if (n_test) mean(y_test)   else NA_real_,
      mean_ref    = if (n_ref)  mean(y_ref)    else NA_real_,
      p_value     = NA_real_,
      r_rb        = NA_real_,
      stringsAsFactors = FALSE
    )
    
    # the guard your comment described
    if (n_test >= min_n && n_ref >= min_n) {
      res <- wilcox.test(y_test, y_ref, exact = FALSE, conf.int = TRUE)
      
      out$p_value  <- res$p.value
      out$r_rb     <- 2 * unname(res$statistic) / (n_test * n_ref) - 1  # fill this
      out$auc      <- unname(res$statistic) / (n_test * n_ref)          # same info, 0–1
      out$hl_shift <- unname(res$estimate)                              # Hodges–Lehmann
      out$hl_lo    <- res$conf.int[1]
      out$hl_hi    <- res$conf.int[2]
    }
    
    p.list[[paste(gc_prot, feat, sep = "__")]] <- out
  }
}

cor_df <- bind_rows(p.list) %>%
  group_by(gc_feature) %>%
  mutate(padj_BH = p.adjust(p_value, method = "BH"))   # one FDR family, at the end
cor_df %>%
  filter(gc_feature == "in_tumor")

lim <- max(abs(cor_df$r_rb), na.rm = TRUE)

cor_df$y_label <- y_labels[cor_df$imc_feature]
p <- cor_df %>%
    mutate(label = case_when(p_value < 0.001 & padj_BH < 0.1 ~ "***",
                           p_value < 0.01  & padj_BH < 0.1 ~ "**",
                           p_value < 0.05  & padj_BH < 0.1 ~ "*",
                           TRUE        ~ "")) %>%
  
  ggplot(aes(gc_feature, y_label, fill = r_rb)) +
  geom_tile(colour = "white", linewidth = 0.6) +
  geom_text(aes(label = label)) +
  scale_fill_gradient2(low = "#2166AC", mid = "#F0F0F0", high = "#B2182B",
                       midpoint = 0, limits = c(-lim, lim),
                       name = "rank biseral \ncorrelation") +
  scale_colour_manual(values = c(`FALSE` = "black", `TRUE` = "white"), guide = "none") +
  scale_size_manual(values = c(`FALSE` = 2.1, `TRUE` = 4), guide = "none") +
  coord_fixed() +
  labs(x = NULL, y = NULL) +
  theme_minimal(base_size = 10) +
  theme(panel.grid = element_blank(),
        axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
        axis.text.y = element_text(size = 14),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 14))

p

ggsave(filename = paste0(out_folder,"S5J_TLS_enrichment_clinical_variables.pdf"),
       plot = p,
       device = "pdf",
       width = 5,
       height = 5)

```


# S5K B cell and Patch density survival

```{r}
plot_dat <- dat %>%
  ungroup() %>%
  mutate(TLS_density = tls/tissue_size,
         gc_density = gc/tissue_size,
         Patch_density_in_tumor_dicho = as.factor(ntile(in_tumor_density, 2)),
         Patch_density_outside_dicho = as.factor(ntile(outside_density, 2)),
         Patch_density_overall_dicho =as.factor(ntile((in_tumor+outside)/tissue_size,2)),
         TLS_density_dicho = as.factor(ntile(tls/tissue_size, 2)),
         GC_density_dicho = as.factor(ntile(gc/tissue_size, 2))) %>%
  left_join(traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(is.na(LUAD_vs_LUSC),"Other",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma","LUAD",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Squamous cell carcinoma","LUSC",LUAD_vs_LUSC))
plot_dat
p.list <- list()
for(i in c("Patch_density_in_tumor_dicho","Patch_density_outside_dicho", "Patch_density_overall_dicho","TLS_density_dicho","GC_density_dicho")){
  for(j in c("LUAD", "LUSC", "Other", "All")) {
  if (j == "All") {
    df <- plot_dat
  } else{
        df <- plot_dat %>%
      filter(LUAD_vs_LUSC == j)
  }
 
  surv_ob <- Surv(df$time_to_event,df$censoring)
  p.list[[paste0(j, "_", i)]] <- coxph( as.formula(paste0("surv_ob ~ ", i)),
                                                        data = df)
  }
}


df <- data.frame(HR =      unlist(lapply(p.list, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p.list, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "upper .95"]}))) 

df1 <- df %>% 
  mutate(hist = str_split_i(rownames(df),"_",1))
  
df1 %>%
  mutate(group = gsub(pattern = "LUSC_|LUAD_|Other_|All_","",rownames(df1)),
         group = gsub("TLS","H&E TLS", group))

p <- df1 %>%
  mutate(group = gsub(pattern = "LUSC_|LUAD_|Other_|All_","",rownames(df1)),
         group = gsub("TLS","H&E TLS", group),
         group = gsub("GC","H&E GC", group),
         group = gsub("_dicho","", group)) %>%
  ggplot(aes(x = HR,y = hist, col = hist))+
  geom_point(size = 2)+
  geom_errorbar(aes(xmin=lower_conf, xmax = upper_conf),lwd =1,width = 0.3)+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust= 1))+
  #scale_y_discrete(labels = function(x) str_split_i(x,"_",2))+
  facet_wrap(.~group) +
  scale_x_log10()+
  theme_bw()+
  geom_vline(xintercept = 1,linetype = "dashed")+
  geom_text(aes(label = paste0("p : ",round(p.value,digits = 3))),
            col = "black",
            vjust = -1,
            inherit.aes = TRUE)+
  scale_color_manual(values = c("LUAD" = "blue", "LUSC" = "orange", "Other" = "gray40", "All" = "black"),name = "Histology")+
    labs(x = "HR",y = "Histology")+
  theme(axis.text.x = element_text(size = 12),
        axis.text.y = element_text(size = 12),
        axis.title.x = element_text(size = 12),
        axis.title.y = element_text(size = 14),
        legend.title = element_text(size = 12),
         legend.text = element_text(size = 12))

p

ggsave(filename = paste0(out_folder,"S5L_survival.pdf"),
       plot = p,
       device = "pdf",
       width = 7,
       height = 4.75)
```

##  sanity check within LUAD

```{r}
cur_dat <- dat %>%
  left_join(traits,by= "immucan_id") %>%
  filter(LUAD_vs_LUSC == "Adenocarcinoma") %>%
  ungroup() %>%
  mutate(Patch_density= in_tumor / tumor_size,
         Patch_density_dicho = as.factor(ntile(Patch_density, 2)),
         Patch_density_group_median = ifelse(Patch_density > median(Patch_density),TRUE, FALSE),
         LUAD_vs_LUSC = ifelse(is.na(LUAD_vs_LUSC),"Other",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma","LUAD",LUAD_vs_LUSC))



p.list <- list()
for(j in c("LUAD")) {
  if (j == "All") {
    df <- cur_dat %>%
      filter(!is.na(cur_dat[,"Patch_density_dicho"]))
  } else{
        df <- cur_dat %>%
      filter(!is.na(cur_dat[,"Patch_density_dicho"]),
             LUAD_vs_LUSC == j)
  }
 
  surv_ob <- Surv(df$time_to_event,df$censoring)
  p.list[[paste0(j, "_Patch_density_dicho")]] <- coxph( as.formula(paste0("surv_ob ~ + Patch_density_dicho")),
                                                        data = df)
}

df <- data.frame(HR =      unlist(lapply(p.list, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p.list, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "upper .95"]}))) 

df1 <- df %>% 
  mutate(hist = str_split_i(rownames(df),"_",1))
  
df1

p <- df1 %>%
  ggplot(aes(x = HR,y = hist, col = hist))+
  geom_point(size = 2)+
  geom_errorbar(aes(xmin=lower_conf, xmax = upper_conf),lwd =1,width = 0.3)+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust= 1))+
  #scale_y_discrete(labels = function(x) str_split_i(x,"_",2))+
  scale_x_log10()+
  theme_bw()+
  geom_vline(xintercept = 1,linetype = "dashed")+
  geom_text(aes(label = paste0("p : ",round(p.value,digits = 3))),
            col = "black",
            vjust = -1,
            inherit.aes = TRUE)+
  scale_color_manual(values = c("LUAD" = "blue", "LUSC" = "orange", "Other" = "gray40", "All" = "black"),name = "Histology")+
  theme(axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10))+
  labs(x = "HR",y = "Histology")

p
```

dichotomizing within LUAD also splits for favorable survival

