---
title: "Figure_4"
author: "Daniel Schulz"
date: "2024-10-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```


```{r}
library(tidyverse)
library(ggpubr)
library(SingleCellExperiment)
library(scales)
library(ggnewscale)
library(edgeR)
library(cowplot)
library(igraph)
library(ggraph)
library(ggrepel)
```

```{r}
mount_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/"
out_folder <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/revised_figures/Figure_4/"

if (!dir.exists(out_folder)) {
  dir.create(out_folder)
}
source("code/lollipop_function.R")
```



# load data
we will load the data we generated at the beginning of figure 2
```{r}
dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/abundances_dat.rds"))
dens_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/densities_dat.rds"))
cello_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/cello_dat.rds"))
clin_dat <- read.csv("/mnt/immucan_volume/clinical_data/Lung_NSCLC2/2026_final/clinical.csv")
int_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/interaction_dat.rds"))
traits <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/all_traits.csv"))

modules <- as.data.frame(readRDS( "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/patient_imaging_meta_cluster_scores.rds"))

rna <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/rna_scores.rds")
rna$immucan_id  <- rownames(rna)
rownames(rna) <- NULL

modules <- modules  %>%
  mutate(immucan_id = rownames(modules))
rownames(modules) <- NULL


# variables
data_type <- unique(str_split_i(string = colnames(dat[,-1]),pattern = "__",i = 2))
```

# 4A Waterfall plot

The code for this plot is here: [here](../../WES/analysis/oncoplot.Rmd)

# 4B Mutations 

```{r}
# ============================================================
# TME vs. driver mutation status analysis - stratified by histology
#
# Assumes:
#   - One row per patient/tumor
#   - 8 TME measurement columns, already z-scored (continuous)
#   - 6 binary (0/1) mutation columns: EGFR, KRAS, STK11, KEAP1, PIK3CA, CDKN2A
#   - A histology column with values "LUAD" / "LUSC"
#
# WHY STRATIFY BY HISTOLOGY:
# Histology is a major confounder here, not just an imbalance issue. EGFR,
# KRAS, STK11, and KEAP1 are classic LUAD-enriched drivers and are expected
# to be rare/near-absent in LUSC. PIK3CA occurs only once in the LUAD
# samples in this cohort - with essentially zero variance there it cannot
# be tested in LUAD at all, but occurs at usable frequency in LUSC. Rather
# than adding histology as a simple covariate (which would not solve the
# "one mutant sample" problem for PIK3CA in LUAD, and would force one model
# structure onto two biologically different populations), the cleanest
# approach is to run separate gene panels per histology:
#   - LUAD: EGFR, KRAS, STK11, KEAP1, CDKN2A   (PIK3CA excluded - see above)
#   - LUSC: CDKN2A, PIK3CA                      (EGFR/KRAS/STK11/KEAP1 excluded
#                                                 - expected too rare in LUSC)
# CDKN2A is analysed in both strata since it's not histology-restricted here.
#
# Implements the tiered comparison strategy, run separately within each
# histology stratum where applicable:
#   Tier 1: EGFR vs KRAS vs driver-negative (mutually exclusive backbone) - LUAD only
#   Tier 2: within KRAS-mutant tumors, STK11/KEAP1 co-mutation stratification - LUAD only
#   Tier 3: multivariate model per TME outcome, all mutations in that stratum's
#           panel as covariates (LUAD panel includes KRAS:STK11 / KRAS:KEAP1
#           interaction terms; LUSC panel has no interaction terms specified)
#   Tier 4: simple univariate mut-vs-wt t-test per mutation, per TME outcome,
#           disregarding all other mutations' status. Compared side-by-side
#           against Tier 3 in a combined heatmap, per stratum.
#
# NOTE on CDKN2A: in NSCLC this is frequently inactivated by homozygous
# deletion rather than point mutation. If your CDKN2A column is derived only
# from small-variant VCF calls, it likely undercounts true CDKN2A loss unless
# you've merged in copy-number data.
# ============================================================

library(tidyverse)
library(broom)

# data
dat_z <- left_join(modules, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"),
         EGFR = ifelse(EGFR == TRUE, 1,0),
         KRAS = ifelse(KRAS == TRUE, 1,0),
         STK11 = ifelse(STK11 == TRUE, 1,0),
         KEAP1 = ifelse(KEAP1 == TRUE, 1,0),
         PIK3CA = ifelse(PIK3CA == TRUE, 1,0),
         CDKN2A = ifelse(CDKN2A == TRUE, 1, 0))

# ---- 1. Config: EDIT these to match your actual column names ----
tme_cols  <- colnames(dat_z)[1:8]
histology_col  <- "LUAD_vs_LUSC"                              # expects values "LUAD" / "LUSC"
mut_cols_luad  <- c("EGFR", "KRAS", "STK11", "KEAP1", "CDKN2A", "TP53")   # PIK3CA excluded: ~1 case in LUAD
mut_cols_lusc  <- c("CDKN2A", "PIK3CA", "TP53")                            # EGFR/KRAS/STK11/KEAP1 excluded in LUSC
min_group_n    <- 5   # minimum mutant (or wt) count required to test a gene in a stratum

df<- dat_z %>%
  dplyr::select(all_of(c(tme_cols,mut_cols_luad, mut_cols_lusc, histology_col))) %>%
  filter(!is.na(STK11)) # we will remove two patients that have missing genomic information

# sanity checks
stopifnot(all(tme_cols %in% names(df)))
stopifnot(histology_col %in% names(df))
all_mut_cols <- union(mut_cols_luad, mut_cols_lusc)
stopifnot(all(all_mut_cols %in% names(df)))
df[all_mut_cols] <- lapply(df[all_mut_cols], as.integer)  # ensure 0/1 numeric

cat("\n--- Histology counts ---\n")
print(table(df[[histology_col]]))

df_luad <- df %>% filter(.data[[histology_col]] == "LUAD")
df_lusc <- df %>% filter(.data[[histology_col]] == "LUSC")

cat(sprintf("\nLUAD n = %d, LUSC n = %d\n", nrow(df_luad), nrow(df_lusc)))

# ---- 2. Helper functions ----

# Drops a mutation from testing if it has fewer than min_group_n carriers
# (or fewer than min_group_n wt), rather than letting t.test()/lm() error out
# or silently return an unstable estimate.
check_group_size <- function(data, mut_var, min_n = min_group_n) {
  tab <- table(data[[mut_var]])
  if (length(tab) < 2 || min(tab) < min_n) {
    message(sprintf(
      "  -> dropping %s: insufficient variation/group size (counts: %s)",
      mut_var, paste(names(tab), tab, sep = "=", collapse = ", ")
    ))
    return(FALSE)
  }
  TRUE
}


run_tier4 <- function(data, tme_cols, mut_cols) {
  valid_muts <- mut_cols[sapply(mut_cols, function(m) check_group_size(data, m))]

  expand_grid(tme_var = tme_cols, mut_var = valid_muts) %>%
    pmap_dfr(function(tme_var, mut_var) {
      grp <- factor(data[[mut_var]], levels = c(0, 1), labels = c("wt", "mut"))
      tt  <- t.test(data[[tme_var]] ~ grp)
      diff_mut_minus_wt <- unname(tt$estimate["mean in group mut"] - tt$estimate["mean in group wt"])
      tibble(tme_var = tme_var, mut_var = mut_var, estimate = diff_mut_minus_wt, p_value = tt$p.value)
    }) %>%
    group_by(mut_var) %>%
    mutate(p_fdr = p.adjust(p_value, method = "BH"),
           p_fdr = p_value) %>%
    ungroup() %>%
    arrange(p_fdr)
}

# ---- 4. Tier 2 (LUAD only): within KRAS-mutant tumors, STK11/KEAP1 co-mutation ----
kras_df_luad <- df_luad %>%
  filter(KRAS == 1) %>%
  mutate(co_mut = if_else(STK11 == 1 | KEAP1 == 1,
                           "STK11_or_KEAP1_altered", "KRAS_only"))

cat("\n--- [LUAD] Tier 2 group sizes (within KRAS-mutant tumors) ---\n")
print(table(kras_df_luad$co_mut))

tier2_results_luad <- map_dfr(tme_cols, function(tme) {
  wt <- t.test(kras_df_luad[[tme]] ~ kras_df_luad$co_mut)
  tibble(tme_var = tme, wilcox_p = wt$p.value)
})
tier2_results_luad$fdr <- p.adjust(tier2_results_luad$wilcox_p, method = "BH")
cat("\n--- [LUAD] Tier 2: STK11/KEAP1 co-mutation effect within KRAS-mutant tumors ---\n")
print(tier2_results_luad)

# ---- 4b. Tier 2 visualization: KRAS-only vs STK11/KEAP1-altered, per TME score ----
# Panels ordered by FDR (strongest hits first, e.g. Cold/Neutrophils) so the
# STK11/KEAP1 "immune-cold" signature is easy to spot at a glance.
kras_long_luad <- kras_df_luad %>%
  dplyr::select(all_of(tme_cols), co_mut) %>%
  pivot_longer(cols = all_of(tme_cols), names_to = "tme_var", values_to = "z_score")

tme_order_tier2 <- tier2_results_luad %>% arrange(fdr) %>% pull(tme_var)
kras_long_luad$tme_var <- factor(kras_long_luad$tme_var, levels = tme_order_tier2)

tier2_sig_labels <- tier2_results_luad %>%
  mutate(
    tme_var = factor(tme_var, levels = tme_order_tier2),
    label = case_when(
      fdr < 0.001 ~ "***",
      fdr < 0.01  ~ "**",
      fdr < 0.05  ~ "*",
      TRUE        ~ paste0("p=", signif(wilcox_p, 2))
    )
  )

tier2_plot <- ggplot(kras_long_luad, aes(x = co_mut, y = z_score, fill = co_mut)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.5) +
  facet_wrap(~tme_var, nrow = 2) +
  geom_text(data = tier2_sig_labels, aes(x = 1.5, y = Inf, label = label),
            inherit.aes = FALSE, vjust = 1.5, size = 3.5) +
  scale_fill_manual(values = c("KRAS_only" = "grey70", "STK11_or_KEAP1_altered" = "firebrick")) +
  theme_minimal() +
  labs(
    title = "KRAS-mutant LUAD: STK11/KEAP1 co-mutation vs. KRAS-only",
    subtitle = "Labels show FDR stars where significant, else raw Wilcoxon p",
    x = NULL, y = "TME z-score", fill = "Group"
  ) +
  theme(legend.position = "bottom")

print(tier2_plot)
#ggsave("tme_kras_costk11keap1_boxplot.png", tier2_plot, width = 9, height = 6, dpi = 300)

# ---- 4c. Encode KRAS+STK11/KEAP1 co-mutation as its own predictor ----
# split KRAS-mutant tumors into two
# mutually exclusive, directly-plottable predictors:
#   - KRAS_only:          KRAS-mutant, no STK11/KEAP1 alteration
#   - KRAS_STK11KEAP1_co: KRAS-mutant WITH STK11 and/or KEAP1 alteration
# STK11 and KEAP1 are left as-is (they still capture any STK11/KEAP1
# mutation regardless of KRAS status).
#

df_luad <- df_luad %>%
  mutate(
    KRAS_only          = as.integer(KRAS == 1 & STK11 == 0 & KEAP1 == 0),
    KRAS_co = as.integer(KRAS == 1 & (STK11 == 1 | KEAP1 == 1))
  )

# Check whether STK11/KEAP1 mutations occur meaningfully outside the KRAS-
# mutant group. Tested this on synthetic data: if STK11/KEAP1 occur almost
# exclusively within KRAS-mutant tumors (as expected from the literature),
# keeping them as standalone predictors alongside KRAS_STK11KEAP1_co makes
# them strongly collinear with it (VIF ~6-10 in the test), which *dilutes*
# the co-mutation signal across three correlated terms instead of clarifying
# it - the opposite of what we want. In that case, drop the standalone
# STK11/KEAP1 terms and let the composite carry that signal on its own.
stk11_keap1_outside_kras <- sum((df_luad$STK11 == 1 | df_luad$KEAP1 == 1) & df_luad$KRAS == 0)
cat(sprintf("\n[LUAD] STK11/KEAP1-mutant tumors outside the KRAS-mutant group: %d\n",
            stk11_keap1_outside_kras))

if (stk11_keap1_outside_kras < min_group_n) {
  message("  -> STK11/KEAP1 occur almost exclusively within KRAS-mutant tumors here; ",
          "dropping standalone STK11/KEAP1 terms to avoid collinearity with KRAS_STK11KEAP1_co.")
  mut_cols_luad_v2 <- c("EGFR", "KRAS_only", "CDKN2A", "KRAS_STK11KEAP1_co")
} else {
  mut_cols_luad_v2 <- c("EGFR", "KRAS_only", "STK11", "KEAP1", "CDKN2A", "KRAS_co", "TP53")
}

# ---- 5. Tier 3 + Tier 4 (LUAD): expanded panel including the co-mutation predictor ----

tier4_luad <- run_tier4(df_luad, tme_cols, mut_cols_luad_v2)

pluad <- tier4_luad %>%
 ggplot(aes(x = mut_var, y = tme_var, fill = estimate)) +
    geom_tile() +
    geom_text(aes(label = ifelse(p_fdr < 0.05, "*", "")), color = "black") +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
    #facet_wrap(~method) +
    theme_minimal() +
    labs(
      title = "Drivers LUAD",
      subtitle = "* = P < 0.05 ",
      x = "Mutation", y = "Imaging module", fill = paste0("Δ z-score")
    )+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
pluad

# ============================================================
# LUSC stratum: CDKN2A, PIK3CA
# ============================================================

tier4_lusc <- run_tier4(df_lusc, tme_cols, mut_cols_lusc)

plusc <- tier4_lusc %>%
 ggplot(aes(x = mut_var, y = tme_var, fill = estimate)) +
    geom_tile() +
    geom_text(aes(label = ifelse(p_fdr < 0.05, "*", "")), color = "black") +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
    #facet_wrap(~method) +
    theme_minimal() +
    labs(
      title = "Drivers LUSC",
      subtitle = "* = P < 0.05 ",
      x = "Mutation", y = "Imaging module", fill = paste0("Δ z-score")
    )+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
plusc

p <- plot_grid(pluad, plusc,ncol= 1 )
p

ggsave(plot = p, path = out_folder,filename = "3B_Imaging_modules_drivers.pdf",width = 5, height = 8,device = cairo_pdf)

#ggsave("tme_mutation_heatmap_LUSC.png", heatmap_lusc, width = 8, height = 6, dpi = 300)
```
# to be deleted
## EGFR mutation

```{r}
cur_var <- "EGFR"
var_A <- "mut"
var_B <-  "wt"
col_A <- "blue"
col_B <- "red"
title_name <- "EGFR"
```

```{r}
table(traits$EGFR, traits$LUAD_vs_LUSC)
```

## regression based
```{r}
dat_z <- left_join(modules, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"),
         EGFR = ifelse(EGFR == TRUE, 1,0),
         KRAS = ifelse(KRAS == TRUE, 1,0),
         STK11 = ifelse(STK11 == TRUE, 1,0),
         KEAP1 = ifelse(KEAP1 == TRUE, 1,0),
         PIK3CA = ifelse(PIK3CA == TRUE, 1,0),
         CDKN2A = ifelse(CDKN2A == TRUE, 1, 0))

dat_z

# ============================================================
# TME vs. driver mutation status analysis
#
# Assumes:
#   - One row per patient/tumor
#   - 8 TME measurement columns, already z-scored (continuous)
#   - 6 binary (0/1) mutation columns: EGFR, KRAS, STK11, KEAP1, PIK3CA, CDKN2A
#
# Implements the tiered comparison strategy:
#   Tier 1: EGFR vs KRAS vs driver-negative (mutually exclusive backbone)
#   Tier 2: within KRAS-mutant tumors, STK11/KEAP1 co-mutation stratification
#   Tier 3: multivariate model per TME outcome, all 6 mutations as covariates,
#           with KRAS:STK11 / KRAS:KEAP1 interaction terms
#
# NOTE on CDKN2A: in NSCLC this is frequently inactivated by homozygous
# deletion rather than point mutation. If your CDKN2A column is derived only
# from small-variant VCF calls, it likely undercounts true CDKN2A loss unless
# you've merged in copy-number data.
# ============================================================

library(tidyverse)
library(broom)
library(car)      # for VIF
# install.packages(c("tidyverse", "broom", "car")) if any are missing

# ---- 1. Config: EDIT these to match your actual column names ----
data_path <- "your_data.csv"     # path to your table
tme_cols  <- colnames(dat_z)[1:8]
mut_cols  <- c("EGFR", "KRAS", "STK11", "KEAP1")

df<- dat_z %>%
  dplyr::select(all_of(c(tme_cols,mut_cols)))

# sanity checks
stopifnot(all(tme_cols %in% names(df)))
stopifnot(all(mut_cols %in% names(df)))
df[mut_cols] <- lapply(df[mut_cols], as.integer)  # ensure 0/1 numeric

# ---- 2. Tier 1: EGFR vs KRAS vs driver-negative backbone ----
df <- df %>%
  mutate(driver_group = case_when(
    EGFR == 1 & KRAS == 1 ~ "EGFR_KRAS_co",   # should be rare; inspect if non-trivial
    EGFR == 1              ~ "EGFR",
    KRAS == 1              ~ "KRAS",
    TRUE                    ~ "driver_negative"
  ))

cat("\n--- Tier 1 group sizes ---\n")
print(table(df$driver_group))

# Overall test per TME score (Kruskal-Wallis, non-parametric, robust to small groups)
tier1_results <- map_dfr(tme_cols, function(tme) {
  kw <- kruskal.test(df[[tme]] ~ df$driver_group)
  tibble(tme_var = tme, kw_p = kw$p.value)
})


tier1_results$kw_fdr <- p.adjust(tier1_results$kw_p, method = "BH")
cat("\n--- Tier 1: omnibus test per TME variable ---\n")
print(tier1_results)

# Pairwise post-hoc comparisons (EGFR vs KRAS, EGFR vs driver-negative, KRAS vs driver-negative)
tier1_pairwise <- map_dfr(tme_cols, function(tme) {
  pw <- pairwise.wilcox.test(df[[tme]], df$driver_group, p.adjust.method = "none")
  as.data.frame(as.table(pw$p.value)) %>%
    dplyr::rename(group1 = Var1, group2 = Var2, p_raw = Freq) %>%
    mutate(tme_var = tme) %>%
    filter(!is.na(p_raw))
})
tier1_pairwise$p_fdr <- p.adjust(tier1_pairwise$p_raw, method = "BH")
cat("\n--- Tier 1: pairwise comparisons (FDR-corrected) ---\n")
print(tier1_pairwise)

# ---- 3. Tier 2: within KRAS-mutant tumors, STK11/KEAP1 co-mutation ----
kras_df <- df %>%
  filter(KRAS == 1) %>%
  mutate(co_mut = if_else(STK11 == 1 | KEAP1 == 1,
                           "STK11_or_KEAP1_altered", "KRAS_only"))

cat("\n--- Tier 2 group sizes (within KRAS-mutant tumors) ---\n")
print(table(kras_df$co_mut))

tier2_results <- map_dfr(tme_cols, function(tme) {
  wt <- wilcox.test(kras_df[[tme]] ~ kras_df$co_mut)
  tibble(tme_var = tme, wilcox_p = wt$p.value)
})
tier2_results$fdr <- p.adjust(tier2_results$wilcox_p, method = "BH")
cat("\n--- Tier 2: STK11/KEAP1 co-mutation effect within KRAS-mutant tumors ---\n")
print(tier2_results)

# ---- 4. Tier 3: multivariate model, all 6 mutations, per TME outcome ----
# KRAS:STK11 and KRAS:KEAP1 interactions directly test whether co-mutation
# modifies the KRAS-associated TME effect (rather than relying only on the
# Tier 2 subgroup test).
tier3_results <- map_dfr(tme_cols, function(tme) {
  form <- as.formula(paste0(
    tme, " ~ EGFR + KRAS + STK11 + KEAP1 + PIK3CA + CDKN2A + KRAS:STK11 + KRAS:KEAP1"
  ))
  fit <- lm(form, data = df)
  tidy(fit) %>% mutate(tme_var = tme)
})

tier3_results <- tier3_results %>%
  filter(term != "(Intercept)") %>%
  group_by(term) %>%
  mutate(p_fdr = p.adjust(p.value, method = "BH")) %>%
  ungroup() %>%
  arrange(p_fdr)

cat("\n--- Tier 3: multivariate model results (FDR-corrected within each term) ---\n")
print(tier3_results, n = Inf)

# ---- 5. Diagnostics ----
cat("\n--- Mutation frequencies (check for underpowered subgroups) ---\n")
print(sapply(df[mut_cols], table))

cat("\n--- Multicollinearity check (VIF), example on first TME variable ---\n")
example_fit <- lm(
  as.formula(paste0(tme_cols[1], " ~ EGFR + KRAS + STK11 + KEAP1 + PIK3CA + CDKN2A")),
  data = df
)
print(vif(example_fit))

# ---- 6. Visualization: effect size heatmap (main effects only) ----
heatmap_plot <- tier3_results %>%
  filter(!str_detect(term, ":")) %>%
  ggplot(aes(x = term, y = tme_var, fill = estimate)) +
  geom_tile() +
  geom_text(aes(label = ifelse(p.value < 0.05, "*", "")), color = "black") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  theme_minimal() +
  labs(
    title = "Mutation-associated TME effect sizes",
    subtitle = "* = P < 0.05 (within term, across TME variables)",
    x = "Mutation", y = "TME measurement", fill = "Beta"
  )

print(heatmap_plot)
#ggsave("tme_mutation_heatmap.png", heatmap_plot, width = 7, height = 6, dpi = 300)


# ---- 7. Tier 4: univariate mut-vs-wt t-test per mutation, ignoring other mutations ----
# Uses Welch's t-test (unequal variance, the standard default) since the TME
# scores are z-scored/normally distributed. Each mutation is tested completely
# independently of the other 5 - no adjustment, no interaction terms.
tier4_results <- expand_grid(tme_var = tme_cols, mut_var = mut_cols) %>%
  pmap_dfr(function(tme_var, mut_var) {
    grp <- factor(df[[mut_var]], levels = c(0, 1), labels = c("wt", "mut"))
    tt  <- t.test(df[[tme_var]] ~ grp)
    diff_mut_minus_wt <- unname(tt$estimate["mean in group mut"] - tt$estimate["mean in group wt"])
    tibble(tme_var = tme_var, mut_var = mut_var,
           estimate = diff_mut_minus_wt, p_value = tt$p.value)
  })
 
tier4_results <- tier4_results %>%
  group_by(mut_var) %>%
  mutate(p_fdr = p.adjust(p_value, method = "BH")) %>%
  ungroup() %>%
  arrange(p_fdr)
 
cat("\n--- Tier 4: univariate mut-vs-wt t-tests (FDR-corrected within each mutation) ---\n")
print(tier4_results, n = Inf)
 
# ---- 8. Tier 3 vs Tier 4 comparison heatmap ----
# Same color scale and gene ordering across both panels so differences are
# purely due to adjustment (Tier 3, multivariate) vs. no adjustment (Tier 4,
# univariate). Large discrepancies for a gene point to confounding by its
# co-occurring mutations (e.g. KRAS appearing "significant" alone but losing
# signal once STK11/KEAP1 co-mutation is accounted for, or vice versa).
comparison_effects <- bind_rows(
  tier3_results %>%
    filter(!str_detect(term, ":")) %>%
    transmute(method = "Multivariate LM (Tier 3)", mut_var = term, tme_var, estimate, p_fdr),
  tier4_results %>%
    transmute(method = "Univariate t-test (Tier 4)", mut_var, tme_var, estimate, p_fdr)
)
 
comparison_effects$mut_var <- factor(comparison_effects$mut_var, levels = mut_cols)
comparison_effects$method  <- factor(comparison_effects$method,
                                      levels = c("Univariate t-test (Tier 4)", "Multivariate LM (Tier 3)"))
 
comparison_heatmap <- comparison_effects %>%
  ggplot(aes(x = mut_var, y = tme_var, fill = estimate)) +
  geom_tile() +
  geom_text(aes(label = ifelse(p_fdr < 0.05, "*", "")), color = "black") +
  scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
  facet_wrap(~method) +
  theme_minimal() +
  labs(
    title = "TME effect sizes: univariate t-test vs. multivariate linear model",
    subtitle = "* = FDR < 0.05 (within mutation/term, across TME variables)",
    x = "Mutation", y = "TME measurement", fill = "Effect\n(mut - wt / beta)"
  )
print(comparison_heatmap)
```


```{r}
# ============================================================
# TME vs. driver mutation status analysis - stratified by histology
#
# Assumes:
#   - One row per patient/tumor
#   - 8 TME measurement columns, already z-scored (continuous)
#   - 6 binary (0/1) mutation columns: EGFR, KRAS, STK11, KEAP1, PIK3CA, CDKN2A
#   - A histology column with values "LUAD" / "LUSC"
#
# WHY STRATIFY BY HISTOLOGY:
# Histology is a major confounder here, not just an imbalance issue. EGFR,
# KRAS, STK11, and KEAP1 are classic LUAD-enriched drivers and are expected
# to be rare/near-absent in LUSC. PIK3CA occurs only once in the LUAD
# samples in this cohort - with essentially zero variance there it cannot
# be tested in LUAD at all, but occurs at usable frequency in LUSC. Rather
# than adding histology as a simple covariate (which would not solve the
# "one mutant sample" problem for PIK3CA in LUAD, and would force one model
# structure onto two biologically different populations), the cleanest
# approach is to run separate gene panels per histology:
#   - LUAD: EGFR, KRAS, STK11, KEAP1, CDKN2A   (PIK3CA excluded - see above)
#   - LUSC: CDKN2A, PIK3CA                      (EGFR/KRAS/STK11/KEAP1 excluded
#                                                 - expected too rare in LUSC)
# CDKN2A is analysed in both strata since it's not histology-restricted here.
#
# Implements the tiered comparison strategy, run separately within each
# histology stratum where applicable:
#   Tier 1: EGFR vs KRAS vs driver-negative (mutually exclusive backbone) - LUAD only
#   Tier 2: within KRAS-mutant tumors, STK11/KEAP1 co-mutation stratification - LUAD only
#   Tier 3: multivariate model per TME outcome, all mutations in that stratum's
#           panel as covariates (LUAD panel includes KRAS:STK11 / KRAS:KEAP1
#           interaction terms; LUSC panel has no interaction terms specified)
#   Tier 4: simple univariate mut-vs-wt t-test per mutation, per TME outcome,
#           disregarding all other mutations' status. Compared side-by-side
#           against Tier 3 in a combined heatmap, per stratum.
#
# NOTE on CDKN2A: in NSCLC this is frequently inactivated by homozygous
# deletion rather than point mutation. If your CDKN2A column is derived only
# from small-variant VCF calls, it likely undercounts true CDKN2A loss unless
# you've merged in copy-number data.
# ============================================================

library(tidyverse)
library(broom)
library(car)      # for VIF
# install.packages(c("tidyverse", "broom", "car")) if any are missing

# ---- 1. Config: EDIT these to match your actual column names ----
tme_cols  <- colnames(dat_z)[1:8]
histology_col  <- "LUAD_vs_LUSC"                              # expects values "LUAD" / "LUSC"
mut_cols_luad  <- c("EGFR", "KRAS", "STK11", "KEAP1", "CDKN2A")   # PIK3CA excluded: ~1 case in LUAD
mut_cols_lusc  <- c("CDKN2A", "PIK3CA")                            # EGFR/KRAS/STK11/KEAP1 excluded in LUSC
min_group_n    <- 5   # minimum mutant (or wt) count required to test a gene in a stratum

df<- dat_z %>%
  dplyr::select(all_of(c(tme_cols,mut_cols_luad,mut_cols_lusc, histology_col)))

# sanity checks
stopifnot(all(tme_cols %in% names(df)))
stopifnot(histology_col %in% names(df))
all_mut_cols <- union(mut_cols_luad, mut_cols_lusc)
stopifnot(all(all_mut_cols %in% names(df)))
df[all_mut_cols] <- lapply(df[all_mut_cols], as.integer)  # ensure 0/1 numeric

cat("\n--- Histology counts ---\n")
print(table(df[[histology_col]]))

df_luad <- df %>% filter(.data[[histology_col]] == "LUAD")
df_lusc <- df %>% filter(.data[[histology_col]] == "LUSC")

cat(sprintf("\nLUAD n = %d, LUSC n = %d\n", nrow(df_luad), nrow(df_lusc)))

# ---- 2. Helper functions ----

# Drops a mutation from testing if it has fewer than min_group_n carriers
# (or fewer than min_group_n wt), rather than letting t.test()/lm() error out
# or silently return an unstable estimate.
check_group_size <- function(data, mut_var, min_n = min_group_n) {
  tab <- table(data[[mut_var]])
  if (length(tab) < 2 || min(tab) < min_n) {
    message(sprintf(
      "  -> dropping %s: insufficient variation/group size (counts: %s)",
      mut_var, paste(names(tab), tab, sep = "=", collapse = ", ")
    ))
    return(FALSE)
  }
  TRUE
}

run_tier3 <- function(data, tme_cols, mut_cols, interaction_terms = character(0), hist = "LUAD") {
  if(hist == "LUAD"){
    valid_muts <- mut_cols[sapply(mut_cols, function(m) check_group_size(data, m))]
    valid_interactions <- interaction_terms[sapply(interaction_terms, function(term) {
      parts <- str_split(term, ":")[[1]]
      all(parts %in% valid_muts)
    })]
    rhs <- paste(c(valid_muts, valid_interactions), collapse = " + ")} else {
      rhs <- paste(c(mut_cols_lusc), collapse = " + ")
    }


  map_dfr(tme_cols, function(tme) {
    form <- as.formula(paste0(tme, " ~ ", rhs))
    fit <- lm(form, data = data)
    tidy(fit) %>% mutate(tme_var = tme)
  }) %>%
    filter(term != "(Intercept)") %>%
    group_by(term) %>%
    mutate(p_value = p.value,
           p_fdr = p.adjust(p.value, method = "BH")) %>%
    ungroup() %>%
    arrange(p_fdr)
}

run_tier4 <- function(data, tme_cols, mut_cols) {
  valid_muts <- mut_cols[sapply(mut_cols, function(m) check_group_size(data, m))]

  expand_grid(tme_var = tme_cols, mut_var = valid_muts) %>%
    pmap_dfr(function(tme_var, mut_var) {
      grp <- factor(data[[mut_var]], levels = c(0, 1), labels = c("wt", "mut"))
      tt  <- t.test(data[[tme_var]] ~ grp)
      diff_mut_minus_wt <- unname(tt$estimate["mean in group mut"] - tt$estimate["mean in group wt"])
      tibble(tme_var = tme_var, mut_var = mut_var, estimate = diff_mut_minus_wt, p_value = tt$p.value)
    }) %>%
    group_by(mut_var) %>%
    mutate(p_fdr = p.adjust(p_value, method = "BH")) %>%
    ungroup() %>%
    arrange(p_fdr)
}

make_comparison_heatmap <- function(tier3_res, tier4_res, mut_order, title_suffix) {
  combined <- bind_rows(
    tier3_res %>%
      filter(!str_detect(term, ":")) %>%
      transmute(method = "Multivariate LM", mut_var = term, tme_var, estimate, p_fdr, p_value),
    tier4_res %>%
      transmute(method = "Univariate t-test", mut_var, tme_var, estimate, p_fdr, p_value)
  )
  #combined$mut_var <- factor(combined$mut_var, levels = mut_order)
  #combined$method  <- factor(combined$method, levels = c("Univariate t-test", "Multivariate LM"))

  combined %>%
    ggplot(aes(x = mut_var, y = tme_var, fill = estimate)) +
    geom_tile() +
    geom_text(aes(label = ifelse(p_value < 0.05, "*", "")), color = "black") +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
    facet_wrap(~method) +
    theme_minimal() +
    labs(
      title = paste0("TME effect sizes: univariate vs. multivariate (", title_suffix, ")"),
      subtitle = "* = p < 0.05 (within mutation/term, across TME variables)",
      x = "Mutation", y = "TME measurement", fill = "Effect"
    )
}

# ============================================================
# LUAD stratum: EGFR, KRAS, STK11, KEAP1, CDKN2A
# ============================================================

# ---- 3. Tier 1 (LUAD only): EGFR vs KRAS vs driver-negative backbone ----
df_luad <- df_luad %>%
  mutate(driver_group = case_when(
    EGFR == 1 & KRAS == 1 ~ "EGFR_KRAS_co",   # should be rare; inspect if non-trivial
    EGFR == 1              ~ "EGFR",
    KRAS == 1              ~ "KRAS",
    TRUE                    ~ "driver_negative"
  ))

cat("\n--- [LUAD] Tier 1 group sizes ---\n")
print(table(df_luad$driver_group))

tier1_results_luad <- map_dfr(tme_cols, function(tme) {
  kw <- kruskal.test(df_luad[[tme]] ~ df_luad$driver_group)
  tibble(tme_var = tme, kw_p = kw$p.value)
})
tier1_results_luad$kw_fdr <- p.adjust(tier1_results_luad$kw_p, method = "BH")
cat("\n--- [LUAD] Tier 1: omnibus test per TME variable ---\n")
print(tier1_results_luad)

tier1_pairwise_luad <- map_dfr(tme_cols, function(tme) {
  pw <- pairwise.wilcox.test(df_luad[[tme]], df_luad$driver_group, p.adjust.method = "none")
  as.data.frame(as.table(pw$p.value)) %>%
    dplyr::rename(group1 = Var1, group2 = Var2, p_raw = Freq) %>%
    mutate(tme_var = tme) %>%
    filter(!is.na(p_raw))
})
tier1_pairwise_luad$p_fdr <- p.adjust(tier1_pairwise_luad$p_raw, method = "BH")
cat("\n--- [LUAD] Tier 1: pairwise comparisons (FDR-corrected) ---\n")
print(tier1_pairwise_luad)

# ---- 4. Tier 2 (LUAD only): within KRAS-mutant tumors, STK11/KEAP1 co-mutation ----
kras_df_luad <- df_luad %>%
  filter(KRAS == 1) %>%
  mutate(co_mut = if_else(STK11 == 1 | KEAP1 == 1,
                           "STK11_or_KEAP1_altered", "KRAS_only"))

cat("\n--- [LUAD] Tier 2 group sizes (within KRAS-mutant tumors) ---\n")
print(table(kras_df_luad$co_mut))

tier2_results_luad <- map_dfr(tme_cols, function(tme) {
  wt <- t.test(kras_df_luad[[tme]] ~ kras_df_luad$co_mut)
  tibble(tme_var = tme, p = wt$p.value)
})
tier2_results_luad$fdr <- p.adjust(tier2_results_luad$p, method = "BH")
cat("\n--- [LUAD] Tier 2: STK11/KEAP1 co-mutation effect within KRAS-mutant tumors ---\n")
print(tier2_results_luad)

kras_long_luad <- kras_df_luad %>%
  dplyr::select(all_of(tme_cols), co_mut) %>%
  pivot_longer(cols = all_of(tme_cols), names_to = "tme_var", values_to = "z_score")
 
tme_order_tier2 <- tier2_results_luad %>% arrange(fdr) %>% pull(tme_var)
kras_long_luad$tme_var <- factor(kras_long_luad$tme_var, levels = tme_order_tier2)
 
tier2_sig_labels <- tier2_results_luad %>%
  mutate(
    tme_var = factor(tme_var, levels = tme_order_tier2),
    label = case_when(
      fdr < 0.001 ~ "***",
      fdr < 0.01  ~ "**",
      fdr < 0.05  ~ "*",
      TRUE        ~ paste0("p=", signif(p, 2))
    )
  )
 
tier2_plot <- ggplot(kras_long_luad, aes(x = co_mut, y = z_score, fill = co_mut)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.5) +
  facet_wrap(~tme_var, nrow = 2) +
  geom_text(data = tier2_sig_labels, aes(x = 1.5, y = Inf, label = label),
            inherit.aes = FALSE, vjust = 1.5, size = 3.5) +
  scale_fill_manual(values = c("KRAS_only" = "grey70", "STK11_or_KEAP1_altered" = "firebrick")) +
  theme_minimal() +
  labs(
    title = "KRAS-mutant LUAD: STK11/KEAP1 co-mutation vs. KRAS-only",
    subtitle = "Labels show FDR stars where significant, else raw Wilcoxon p",
    x = NULL, y = "TME z-score", fill = "Group"
  ) +
  theme(legend.position = "bottom")
 
print(tier2_plot)

# ---- 5. Tier 3 + Tier 4 (LUAD): all 5 LUAD-panel mutations ----
tier3_luad <- run_tier3(df_luad, tme_cols, mut_cols_luad,
                         interaction_terms = c("KRAS:STK11", "KRAS:KEAP1"))
cat("\n--- [LUAD] Tier 3: multivariate model results (FDR-corrected within each term) ---\n")
print(tier3_luad, n = Inf)

tier4_luad <- run_tier4(df_luad, tme_cols, mut_cols_luad)
cat("\n--- [LUAD] Tier 4: univariate mut-vs-wt t-tests (FDR-corrected within each mutation) ---\n")
print(tier4_luad, n = Inf)

cat("\n--- [LUAD] Multicollinearity check (VIF), example on first TME variable ---\n")
example_fit_luad <- lm(
  as.formula(paste0(tme_cols[1], " ~ ", paste(mut_cols_luad, collapse = " + "))),
  data = df_luad
)
print(vif(example_fit_luad))

heatmap_luad <- make_comparison_heatmap(tier3_luad, tier4_luad, mut_cols_luad, "LUAD")
print(heatmap_luad)
#ggsave("tme_mutation_heatmap_LUAD.png", heatmap_luad, width = 10, height = 6, dpi = 300)

# ============================================================
# LUSC stratum: CDKN2A, PIK3CA
# (No EGFR/KRAS backbone here - Tier 1/2 don't apply to LUSC)
# ============================================================

cat("\n--- [LUSC] mutation counts ---\n")
print(sapply(df_lusc[mut_cols_lusc], table))

tier3_lusc <- run_tier3(df_lusc, tme_cols, mut_cols_lusc, interaction_terms = character(0), hist = "LUSC")
cat("\n--- [LUSC] Tier 3: multivariate model results (FDR-corrected within each term) ---\n")
print(tier3_lusc, n = Inf)

tier4_lusc <- run_tier4(df_lusc, tme_cols, mut_cols_lusc)
cat("\n--- [LUSC] Tier 4: univariate mut-vs-wt t-tests (FDR-corrected within each mutation) ---\n")
print(tier4_lusc, n = Inf)

heatmap_lusc <- make_comparison_heatmap(tier3_lusc, tier4_lusc, mut_cols_lusc, "LUSC")
print(heatmap_lusc)
#ggsave("tme_mutation_heatmap_LUSC.png", heatmap_lusc, width = 8, height = 6, dpi = 300)
```


## reg based final

```{r}
# ============================================================
# TME vs. driver mutation status analysis - stratified by histology
#
# Assumes:
#   - One row per patient/tumor
#   - 8 TME measurement columns, already z-scored (continuous)
#   - 6 binary (0/1) mutation columns: EGFR, KRAS, STK11, KEAP1, PIK3CA, CDKN2A
#   - A histology column with values "LUAD" / "LUSC"
#
# WHY STRATIFY BY HISTOLOGY:
# Histology is a major confounder here, not just an imbalance issue. EGFR,
# KRAS, STK11, and KEAP1 are classic LUAD-enriched drivers and are expected
# to be rare/near-absent in LUSC. PIK3CA occurs only once in the LUAD
# samples in this cohort - with essentially zero variance there it cannot
# be tested in LUAD at all, but occurs at usable frequency in LUSC. Rather
# than adding histology as a simple covariate (which would not solve the
# "one mutant sample" problem for PIK3CA in LUAD, and would force one model
# structure onto two biologically different populations), the cleanest
# approach is to run separate gene panels per histology:
#   - LUAD: EGFR, KRAS, STK11, KEAP1, CDKN2A   (PIK3CA excluded - see above)
#   - LUSC: CDKN2A, PIK3CA                      (EGFR/KRAS/STK11/KEAP1 excluded
#                                                 - expected too rare in LUSC)
# CDKN2A is analysed in both strata since it's not histology-restricted here.
#
# Implements the tiered comparison strategy, run separately within each
# histology stratum where applicable:
#   Tier 1: EGFR vs KRAS vs driver-negative (mutually exclusive backbone) - LUAD only
#   Tier 2: within KRAS-mutant tumors, STK11/KEAP1 co-mutation stratification - LUAD only
#   Tier 3: multivariate model per TME outcome, all mutations in that stratum's
#           panel as covariates (LUAD panel includes KRAS:STK11 / KRAS:KEAP1
#           interaction terms; LUSC panel has no interaction terms specified)
#   Tier 4: simple univariate mut-vs-wt t-test per mutation, per TME outcome,
#           disregarding all other mutations' status. Compared side-by-side
#           against Tier 3 in a combined heatmap, per stratum.
#
# NOTE on CDKN2A: in NSCLC this is frequently inactivated by homozygous
# deletion rather than point mutation. If your CDKN2A column is derived only
# from small-variant VCF calls, it likely undercounts true CDKN2A loss unless
# you've merged in copy-number data.
# ============================================================

library(tidyverse)
library(broom)
library(car)      # for VIF
# install.packages(c("tidyverse", "broom", "car")) if any are missing

# ---- 1. Config: EDIT these to match your actual column names ----
tme_cols  <- colnames(dat_z)[1:8]
histology_col  <- "LUAD_vs_LUSC"                              # expects values "LUAD" / "LUSC"
mut_cols_luad  <- c("EGFR", "KRAS", "STK11", "KEAP1", "CDKN2A")   # PIK3CA excluded: ~1 case in LUAD
mut_cols_lusc  <- c("CDKN2A", "PIK3CA")                            # EGFR/KRAS/STK11/KEAP1 excluded in LUSC
min_group_n    <- 5   # minimum mutant (or wt) count required to test a gene in a stratum

df<- dat_z %>%
  dplyr::select(all_of(c(tme_cols,mut_cols_luad, mut_cols_lusc, histology_col))) %>%
  filter(!is.na(STK11)) # we will remove two patients that have missing genomic information

# sanity checks
stopifnot(all(tme_cols %in% names(df)))
stopifnot(histology_col %in% names(df))
all_mut_cols <- union(mut_cols_luad, mut_cols_lusc)
stopifnot(all(all_mut_cols %in% names(df)))
df[all_mut_cols] <- lapply(df[all_mut_cols], as.integer)  # ensure 0/1 numeric

cat("\n--- Histology counts ---\n")
print(table(df[[histology_col]]))

df_luad <- df %>% filter(.data[[histology_col]] == "LUAD")
df_lusc <- df %>% filter(.data[[histology_col]] == "LUSC")

cat(sprintf("\nLUAD n = %d, LUSC n = %d\n", nrow(df_luad), nrow(df_lusc)))

# ---- 2. Helper functions ----

# Drops a mutation from testing if it has fewer than min_group_n carriers
# (or fewer than min_group_n wt), rather than letting t.test()/lm() error out
# or silently return an unstable estimate.
check_group_size <- function(data, mut_var, min_n = min_group_n) {
  tab <- table(data[[mut_var]])
  if (length(tab) < 2 || min(tab) < min_n) {
    message(sprintf(
      "  -> dropping %s: insufficient variation/group size (counts: %s)",
      mut_var, paste(names(tab), tab, sep = "=", collapse = ", ")
    ))
    return(FALSE)
  }
  TRUE
}


run_tier3 <- function(data, tme_cols, mut_cols) {
  valid_muts <- mut_cols[sapply(mut_cols, function(m) check_group_size(data, m))]
  
  rhs <- paste(c(valid_muts), collapse = " + ")
  
  map_dfr(tme_cols, function(tme) {
    form <- as.formula(paste0(tme, " ~ ", rhs))
    fit <- lm(form, data = data)
    tidy(fit) %>% mutate(tme_var = tme)
  }) %>%
    filter(term != "(Intercept)") %>%
    group_by(term) %>%
    mutate(p_fdr = p.adjust(p.value, method = "BH"),
           p_fdr = p.value) %>%
    ungroup() %>%
    arrange(p_fdr)
}

run_tier4 <- function(data, tme_cols, mut_cols) {
  valid_muts <- mut_cols[sapply(mut_cols, function(m) check_group_size(data, m))]

  expand_grid(tme_var = tme_cols, mut_var = valid_muts) %>%
    pmap_dfr(function(tme_var, mut_var) {
      grp <- factor(data[[mut_var]], levels = c(0, 1), labels = c("wt", "mut"))
      tt  <- t.test(data[[tme_var]] ~ grp)
      diff_mut_minus_wt <- unname(tt$estimate["mean in group mut"] - tt$estimate["mean in group wt"])
      tibble(tme_var = tme_var, mut_var = mut_var, estimate = diff_mut_minus_wt, p_value = tt$p.value)
    }) %>%
    group_by(mut_var) %>%
    mutate(p_fdr = p.adjust(p_value, method = "BH"),
           p_fdr = p_value) %>%
    ungroup() %>%
    arrange(p_fdr)
}

make_comparison_heatmap <- function(tier3_res, tier4_res, mut_order, title_suffix) {
  combined <- bind_rows(
    tier3_res %>%
      filter(!str_detect(term, ":")) %>%
      transmute(method = "Multivariate LM", mut_var = term, tme_var, estimate, p_fdr),
    tier4_res %>%
      transmute(method = "Univariate t-test", mut_var, tme_var, estimate, p_fdr)
  )
  combined$mut_var <- factor(combined$mut_var, levels = mut_order)
  combined$method  <- factor(combined$method, levels = c("Univariate t-test", "Multivariate LM"))

  combined %>%
    ggplot(aes(x = mut_var, y = tme_var, fill = estimate)) +
    geom_tile() +
    geom_text(aes(label = ifelse(p_fdr < 0.05, "*", "")), color = "black") +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
    facet_wrap(~method) +
    theme_minimal() +
    labs(
      title = paste0("TME effect sizes: univariate vs. multivariate (", title_suffix, ")"),
      subtitle = "* = P < 0.05 (within mutation/term, across TME variables)",
      x = "Mutation", y = "TME measurement", fill = "Effect"
    )
}

# ============================================================
# LUAD stratum: EGFR, KRAS, STK11, KEAP1, CDKN2A
# ============================================================

# ---- 3. Tier 1 (LUAD only): EGFR vs KRAS vs driver-negative backbone ----
df_luad <- df_luad %>%
  mutate(driver_group = case_when(
    EGFR == 1 & KRAS == 1 ~ "EGFR_KRAS_co",   # should be rare; inspect if non-trivial
    EGFR == 1              ~ "EGFR",
    KRAS == 1              ~ "KRAS",
    TRUE                    ~ "driver_negative"
  ))

cat("\n--- [LUAD] Tier 1 group sizes ---\n")
print(table(df_luad$driver_group))

tier1_results_luad <- map_dfr(tme_cols, function(tme) {
  kw <- kruskal.test(df_luad[[tme]] ~ df_luad$driver_group)
  tibble(tme_var = tme, kw_p = kw$p.value)
})
tier1_results_luad$kw_fdr <- p.adjust(tier1_results_luad$kw_p, method = "BH")
cat("\n--- [LUAD] Tier 1: omnibus test per TME variable ---\n")
print(tier1_results_luad)

tier1_pairwise_luad <- map_dfr(tme_cols, function(tme) {
  pw <- pairwise.t.test(df_luad[[tme]], df_luad$driver_group, p.adjust.method = "none")
  as.data.frame(as.table(pw$p.value)) %>%
    dplyr::rename(group1 = Var1, group2 = Var2, p_raw = Freq) %>%
    mutate(tme_var = tme) %>%
    filter(!is.na(p_raw))
})
tier1_pairwise_luad$p_fdr <- p.adjust(tier1_pairwise_luad$p_raw, method = "BH")
cat("\n--- [LUAD] Tier 1: pairwise comparisons (FDR-corrected) ---\n")
print(tier1_pairwise_luad)

# ---- 4. Tier 2 (LUAD only): within KRAS-mutant tumors, STK11/KEAP1 co-mutation ----
kras_df_luad <- df_luad %>%
  filter(KRAS == 1) %>%
  mutate(co_mut = if_else(STK11 == 1 | KEAP1 == 1,
                           "STK11_or_KEAP1_altered", "KRAS_only"))

cat("\n--- [LUAD] Tier 2 group sizes (within KRAS-mutant tumors) ---\n")
print(table(kras_df_luad$co_mut))

tier2_results_luad <- map_dfr(tme_cols, function(tme) {
  wt <- t.test(kras_df_luad[[tme]] ~ kras_df_luad$co_mut)
  tibble(tme_var = tme, wilcox_p = wt$p.value)
})
tier2_results_luad$fdr <- p.adjust(tier2_results_luad$wilcox_p, method = "BH")
cat("\n--- [LUAD] Tier 2: STK11/KEAP1 co-mutation effect within KRAS-mutant tumors ---\n")
print(tier2_results_luad)

# ---- 4b. Tier 2 visualization: KRAS-only vs STK11/KEAP1-altered, per TME score ----
# Panels ordered by FDR (strongest hits first, e.g. Cold/Neutrophils) so the
# STK11/KEAP1 "immune-cold" signature is easy to spot at a glance.
kras_long_luad <- kras_df_luad %>%
  dplyr::select(all_of(tme_cols), co_mut) %>%
  pivot_longer(cols = all_of(tme_cols), names_to = "tme_var", values_to = "z_score")

tme_order_tier2 <- tier2_results_luad %>% arrange(fdr) %>% pull(tme_var)
kras_long_luad$tme_var <- factor(kras_long_luad$tme_var, levels = tme_order_tier2)

tier2_sig_labels <- tier2_results_luad %>%
  mutate(
    tme_var = factor(tme_var, levels = tme_order_tier2),
    label = case_when(
      fdr < 0.001 ~ "***",
      fdr < 0.01  ~ "**",
      fdr < 0.05  ~ "*",
      TRUE        ~ paste0("p=", signif(wilcox_p, 2))
    )
  )

tier2_plot <- ggplot(kras_long_luad, aes(x = co_mut, y = z_score, fill = co_mut)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.5) +
  facet_wrap(~tme_var, nrow = 2) +
  geom_text(data = tier2_sig_labels, aes(x = 1.5, y = Inf, label = label),
            inherit.aes = FALSE, vjust = 1.5, size = 3.5) +
  scale_fill_manual(values = c("KRAS_only" = "grey70", "STK11_or_KEAP1_altered" = "firebrick")) +
  theme_minimal() +
  labs(
    title = "KRAS-mutant LUAD: STK11/KEAP1 co-mutation vs. KRAS-only",
    subtitle = "Labels show FDR stars where significant, else raw Wilcoxon p",
    x = NULL, y = "TME z-score", fill = "Group"
  ) +
  theme(legend.position = "bottom")

print(tier2_plot)
#ggsave("tme_kras_costk11keap1_boxplot.png", tier2_plot, width = 9, height = 6, dpi = 300)

# ---- 4c. Encode KRAS+STK11/KEAP1 co-mutation as its own predictor ----
# Instead of relying on KRAS:STK11 / KRAS:KEAP1 interaction terms (which are
# underpowered in a small co-mutated subgroup and were being filtered out of
# the comparison heatmap entirely - !str_detect(term, ":") - so that signal
# was never actually visible there), split KRAS-mutant tumors into two
# mutually exclusive, directly-plottable predictors:
#   - KRAS_only:          KRAS-mutant, no STK11/KEAP1 alteration
#   - KRAS_STK11KEAP1_co: KRAS-mutant WITH STK11 and/or KEAP1 alteration
# STK11 and KEAP1 are left as-is (they still capture any STK11/KEAP1
# mutation regardless of KRAS status).
#
# NOTE on comparator groups: this changes what "wt" means for the co-mutation
# effect. Tier 2 (above) compares co-mutated vs. KRAS-only *within KRAS-mutant
# tumors only* - a matched, within-driver comparison. KRAS_STK11KEAP1_co in
# Tier 3/4 below compares co-mutated (=1) vs. everyone else (=0), i.e. EGFR-
# mutant, driver-negative, AND KRAS-only tumors combined - a broader, less
# matched comparator. Treat the two as complementary views, not interchangeable.
df_luad <- df_luad %>%
  mutate(
    KRAS_only          = as.integer(KRAS == 1 & STK11 == 0 & KEAP1 == 0),
    KRAS_STK11KEAP1_co = as.integer(KRAS == 1 & (STK11 == 1 | KEAP1 == 1))
  )

# Check whether STK11/KEAP1 mutations occur meaningfully outside the KRAS-
# mutant group. Tested this on synthetic data: if STK11/KEAP1 occur almost
# exclusively within KRAS-mutant tumors (as expected from the literature),
# keeping them as standalone predictors alongside KRAS_STK11KEAP1_co makes
# them strongly collinear with it (VIF ~6-10 in the test), which *dilutes*
# the co-mutation signal across three correlated terms instead of clarifying
# it - the opposite of what we want. In that case, drop the standalone
# STK11/KEAP1 terms and let the composite carry that signal on its own.
stk11_keap1_outside_kras <- sum((df_luad$STK11 == 1 | df_luad$KEAP1 == 1) & df_luad$KRAS == 0)
cat(sprintf("\n[LUAD] STK11/KEAP1-mutant tumors outside the KRAS-mutant group: %d\n",
            stk11_keap1_outside_kras))

if (stk11_keap1_outside_kras < min_group_n) {
  message("  -> STK11/KEAP1 occur almost exclusively within KRAS-mutant tumors here; ",
          "dropping standalone STK11/KEAP1 terms to avoid collinearity with KRAS_STK11KEAP1_co.")
  mut_cols_luad_v2 <- c("EGFR", "KRAS_only", "CDKN2A", "KRAS_STK11KEAP1_co")
} else {
  mut_cols_luad_v2 <- c("EGFR", "KRAS_only", "STK11", "KEAP1", "CDKN2A", "KRAS_STK11KEAP1_co")
}

# ---- 5. Tier 3 + Tier 4 (LUAD): expanded panel including the co-mutation predictor ----
# No interaction terms needed anymore - the co-mutation effect is now its own
# main-effect column and will appear directly in the heatmap.
tier3_luad <- run_tier3(df_luad, tme_cols, mut_cols_luad_v2)
cat("\n--- [LUAD] Tier 3: multivariate model results (FDR-corrected within each term) ---\n")
print(tier3_luad, n = Inf)

tier4_luad <- run_tier4(df_luad, tme_cols, mut_cols_luad_v2)
cat("\n--- [LUAD] Tier 4: univariate mut-vs-wt t-tests (FDR-corrected within each mutation) ---\n")
print(tier4_luad, n = Inf)

cat("\n--- [LUAD] Multicollinearity check (VIF), example on first TME variable ---\n")
example_fit_luad <- lm(
  as.formula(paste0(tme_cols[1], " ~ ", paste(mut_cols_luad_v2, collapse = " + "))),
  data = df_luad
)

heatmap_luad <- make_comparison_heatmap(tier3_luad, tier4_luad, mut_cols_luad_v2, "LUAD")
print(heatmap_luad)
#ggsave("tme_mutation_heatmap_LUAD.png", heatmap_luad, width = 11, height = 6, dpi = 300)
pluad <- tier4_luad %>%
 ggplot(aes(x = mut_var, y = tme_var, fill = estimate)) +
    geom_tile() +
    geom_text(aes(label = ifelse(p_fdr < 0.05, "*", "")), color = "black") +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
    #facet_wrap(~method) +
    theme_minimal() +
    labs(
      title = "TME effect sizes: univariate t.test",
      subtitle = "* = P < 0.05 ",
      x = "Mutation", y = "Imaging module", fill = paste0("Δ z-score")
    )+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
pluad
# ============================================================
# LUSC stratum: CDKN2A, PIK3CA
# (No EGFR/KRAS backbone here - Tier 1/2 don't apply to LUSC)
# ============================================================

cat("\n--- [LUSC] mutation counts ---\n")
print(sapply(df_lusc[mut_cols_lusc], table))

tier3_lusc <- run_tier3(df_lusc, tme_cols, mut_cols_lusc)
cat("\n--- [LUSC] Tier 3: multivariate model results (FDR-corrected within each term) ---\n")
print(tier3_lusc, n = Inf)

tier4_lusc <- run_tier4(df_lusc, tme_cols, mut_cols_lusc)
cat("\n--- [LUSC] Tier 4: univariate mut-vs-wt t-tests (FDR-corrected within each mutation) ---\n")
print(tier4_lusc, n = Inf)

heatmap_lusc <- make_comparison_heatmap(tier3_lusc, tier4_lusc, mut_cols_lusc, "LUSC")
print(heatmap_lusc)

plusc <- tier4_lusc %>%
 ggplot(aes(x = mut_var, y = tme_var, fill = estimate)) +
    geom_tile() +
    geom_text(aes(label = ifelse(p_fdr < 0.05, "*", "")), color = "black") +
    scale_fill_gradient2(low = "blue", mid = "white", high = "red", midpoint = 0) +
    #facet_wrap(~method) +
    theme_minimal() +
    labs(
      title = "TME effect sizes: univariate t.test",
      subtitle = "* = P < 0.05 ",
      x = "Mutation", y = "Imaging module", fill = paste0("Δ z-score")
    )+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
plusc
#ggsave("tme_mutation_heatmap_LUSC.png", heatmap_lusc, width = 8, height = 6, dpi = 300)
```


```{r}
v1 <- df_luad %>%
  filter(EGFR == 1) %>%
  pull(Thelper_MHCII)

v2 <- df_luad %>%
  filter(EGFR == 0) %>%
  pull(Thelper_MHCII)

t.test(v1, v2)
mean(v1)
mean(v2)
```


## imaging modules


```{r}
dat_z <- left_join(modules, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"),
         EGFR = ifelse(EGFR == TRUE, "mut", "wt")) %>%
  filter(LUAD_vs_LUSC == "LUAD")

print(table(dat_z[[cur_var]])  )
dat_z <- dat_z %>%
pivot_longer(cols = 1:8)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B, zscore_data = FALSE)
result$data
p_LUAD <- result$plot +
  ggtitle(paste0("Imaging modules for ", title_name," in LUAD"))+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))
p_LUAD
```

## RNA modules

```{r}
dat_z <- left_join(rna, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"),
         EGFR = ifelse(EGFR == TRUE, "mut", "wt")) %>%
  filter(LUAD_vs_LUSC == "LUAD")

print(table(dat_z[[cur_var]])  )
dat_z <- dat_z %>%
pivot_longer(cols = 1:29)


result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B,zscore_data = TRUE)
result$data
p_LUAD <- result$plot +
  ggtitle(paste0("RNA modules for ", title_name," in LUAD"))+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))
p_LUAD
```

## inflammation 

### get gene signatures
```{r}
library(org.Hs.eg.db)
library(AnnotationDbi)
library(msigdbr)
library(singscore)

rna_mat <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/rna_mat.rds")


# ── gene signatures ────────────────────────────────────────────────────
hallmark_sets <- msigdbr(species = "Homo sapiens", collection = "H") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  split(.$gs_name) %>%
  lapply(function(x) x$gene_symbol)

# Targeted Hallmark gene sets of interest
target_sets <- c(
  "HALLMARK_INFLAMMATORY_RESPONSE",
  "HALLMARK_INTERFERON_GAMMA_RESPONSE",
  "HALLMARK_INTERFERON_ALPHA_RESPONSE",
  "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
  "HALLMARK_IL6_JAK_STAT3_SIGNALING",
  "HALLMARK_IL2_STAT5_SIGNALING"
)
gs_list_hallmark <- hallmark_sets[intersect(target_sets, names(hallmark_sets))]

# ── 2. Custom gene sets ────────────────────────────────────────────────────────
# CXCL9/10/11 chemokine axis (T cell recruitment)
gs_list_custom <- list(
  CXCL9_10_11_axis = c("CXCL9", "CXCL10", "CXCL11", "CXCR3",
                       "CXCL13", "CXCR5"),
  T_cell_activation = c("CD8A", "CD8B", "GZMB", "PRF1", "IFNG",
                        "TNF", "IL2", "PDCD1", "HAVCR2", "LAG3"),
  type_I_IFN = c("IFNB1", "IRF7", "MX1", "OAS1", "OAS2", "OAS3",
                 "ISG15", "IFIT1", "IFIT2", "IFIT3", "RSAD2", "HERC5"),
  NF_kB_target  = c("NFKB1", "RELA", "RELB", "NFKB2", "REL",
                    "BCL2L1", "IL1B", "IL6", "CXCL8", "TNF", "ICAM1"),
  TCR_signaling = c("CD3D", "CD3E", "CD3G", "ZAP70", "LCK", "LAT",
                    "ITK", "PLCG1", "NFATC1", "NFATC2"),
  cytotoxic = c("GZMB", "GZMA", "GZMK", "PRF1", "NKG7", "IFNG", "KLRG1"),
  exhaustion = c("PDCD1","LAG3","HAVCR2","TX","CTLA4", "IDO1")
  )

all_gs <- c(gs_list_hallmark, gs_list_custom)

# compute score
rna_for_score <- t(rna_mat)

# Filter gene sets to genes present in the RNA-seq matrix
all_gs <- lapply(all_gs, function(g) intersect(g, rownames(rna_for_score)))
all_gs <- all_gs[sapply(all_gs, length) >= 5]



overlap <- sapply(all_gs, function(g) sum(g %in% rownames(rna_for_score)))
total   <- sapply(all_gs, length)
print(data.frame(found = overlap, total = total, pct = round(100 * overlap / total)))

# ── Score with singscore ──────────────────────────────────────────────────────
ranked <- rankGenes(rna_for_score)

scores_list <- lapply(names(all_gs), function(nm) {
  sc <- simpleScore(ranked, upSet = all_gs[[nm]], centerScore = TRUE)
  setNames(sc$TotalScore, rownames(sc))
})


rna_scores <- as.data.frame(do.call(cbind, scores_list))
colnames(rna_scores) <- names(all_gs)
rownames(rna_scores) <- colnames(rna_for_score)
rna_scores$immucan_id <- rownames(rna_scores)
rownames(rna_scores) <- NULL

```

### LUAD 
```{r }
dat_z <- left_join(rna_scores, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"),
         EGFR = ifelse(EGFR == TRUE, "mut", "wt")) %>%
  filter(LUAD_vs_LUSC == "LUAD")

print(table(dat_z[[cur_var]])  )
dat_z <- dat_z %>%
  pivot_longer(cols = 1:13)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B)

p_LUAD <- result$plot+
    ylab("Inf. signatures")+
    ggtitle(paste0("Inf. signatures for ", title_name," in LUAD"))+
  theme(axis.title.y = element_blank())+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))

p_LUAD
```





## sanity check

```{r, fig.width=12}
ctr <- read.csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/image_feature_module_correlation.csv")
im <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/imaging_mat.rds")
feats <- ctr %>% 
  arrange(desc(Proliferating)) %>%
  slice_head(n = 20) %>%
  pull(X)
  
im %>%
  dplyr::select(all_of(feats)) %>%
  mutate(immucan_id = rownames(im)) %>%
  pivot_longer(-immucan_id) %>%
  left_join(traits, "immucan_id") %>%
  filter(!is.na(EGFR),
         LUAD_vs_LUSC == "Adenocarcinoma",
         KRAS == FALSE,
         STK11 == FALSE,
         TP53 == FALSE) %>%
  mutate(EGFR = ifelse(EGFR == TRUE, "mut", "wt")) %>%
  ggplot(aes(x = name,y = value, fill = EGFR))+
  geom_boxplot()+
  scale_y_log10()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust= 1))+
  stat_compare_means(method = "wilcox.test",label = "p.format")

feats <- ctr %>% 
  arrange(desc(Neutrophils)) %>%
  slice_head(n = 20) %>%
  pull(X)
  
im %>%
  dplyr::select(all_of(feats)) %>%
  mutate(immucan_id = rownames(im)) %>%
  pivot_longer(-immucan_id) %>%
  left_join(traits, "immucan_id") %>%
  filter(!is.na(EGFR),
         LUAD_vs_LUSC == "Adenocarcinoma",
         KRAS == FALSE,
         STK11 == FALSE,
         TP53 == FALSE) %>%
  mutate(EGFR = ifelse(EGFR == TRUE, "mut", "wt")) %>%
  ggplot(aes(x = name,y = value, fill = EGFR))+
  geom_boxplot()+
  scale_y_log10()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust= 1))+
  stat_compare_means(method = "wilcox.test",label = "p.format")

feats <- ctr %>% 
  arrange(desc(Thelper_MHCII)) %>%
  slice_head(n = 20) %>%
  pull(X)
  
im %>%
  dplyr::select(all_of(feats)) %>%
  mutate(immucan_id = rownames(im)) %>%
  pivot_longer(-immucan_id) %>%
  left_join(traits, "immucan_id") %>%
  filter(!is.na(EGFR),
         LUAD_vs_LUSC == "Adenocarcinoma") %>%
  mutate(EGFR = ifelse(EGFR == TRUE, "mut", "wt")) %>%
  ggplot(aes(x = name,y = value, fill = EGFR))+
  geom_boxplot()+
  scale_y_log10()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust= 1))+
  stat_compare_means(method = "wilcox.test",label = "p.format")
```


## KRAS mutation

```{r}
cur_var <- "KRAS"
var_A <- "mut"
var_B <-  "wt"
col_A <- "blue"
col_B <- "red"
title_name <- "KRAS"
```

```{r}
table(traits$KRAS, traits$LUAD_vs_LUSC)
```

## imaging modules

```{r}
dat_z <- left_join(modules, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"),
         KRAS = ifelse(KRAS == TRUE, "mut", "wt")) %>%
  filter(LUAD_vs_LUSC == "LUAD")

print(table(dat_z[[cur_var]])  )
dat_z <- dat_z %>%
  pivot_longer(cols = 1:8)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B)
p_LUAD <- result$plot +
  ggtitle(paste0("Imaging modules for ", title_name," in LUAD"))+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))
p_LUAD
```

## RNA modules

```{r}
dat_z <- left_join(rna, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"),
         KRAS = ifelse(KRAS == TRUE, "mut", "wt")) %>%
  filter(LUAD_vs_LUSC == "LUAD",
         TP53 == FALSE,
         EGFR == FALSE,
         STK11 == FALSE) 


print(table(dat_z[[cur_var]]))
dat_z <- dat_z%>%
  pivot_longer(cols = 1:29)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B)
p_LUAD <- result$plot +
  ggtitle(paste0("Imaging modules for ", title_name," in LUAD"))+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))
p_LUAD
```

## inflammation 

### LUAD 
```{r }
dat_z <- left_join(rna_scores, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"),
         KRAS = ifelse(KRAS == TRUE, "mut", "wt")) %>%
  filter(LUAD_vs_LUSC == "LUAD",
         TP53 == FALSE,
         EGFR == FALSE,
         STK11 == FALSE)

print(table(dat_z[[cur_var]])  )
dat_z <- dat_z %>%
  pivot_longer(cols = 1:13)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B)

p_LUAD <- result$plot+
    ylab("Inf. signatures")+
    ggtitle(paste0("Inf. signatures for ", title_name," in LUAD"))+
  theme(axis.title.y = element_blank())+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))

p_LUAD
```





## sanity check
```{r, fig.width=12}
ctr <- read.csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/image_feature_module_correlation.csv")
im <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/imaging_mat.rds")
feats <- ctr %>% 
  arrange(desc(Thelper_MHCII)) %>%
  slice_head(n = 20) %>%
  pull(X)
  
im %>%
  dplyr::select(all_of(feats)) %>%
  mutate(immucan_id = rownames(im)) %>%
  pivot_longer(-immucan_id) %>%
  left_join(traits, "immucan_id") %>%
  filter(!is.na(KRAS),
         LUAD_vs_LUSC == "Adenocarcinoma",
         EGFR == FALSE,
         STK11 == FALSE,
         TP53 == FALSE) %>%
  mutate(KRAS = ifelse(KRAS == TRUE, "mut", "wt")) %>%
  ggplot(aes(x = name,y = value, fill = KRAS))+
  geom_boxplot()+
  scale_y_log10()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust= 1))+
  stat_compare_means(method = "wilcox.test",label = "p.format")

feats <- ctr %>% 
  arrange(desc(Neutrophils)) %>%
  slice_head(n = 20) %>%
  pull(X)
  
im %>%
  dplyr::select(all_of(feats)) %>%
  mutate(immucan_id = rownames(im)) %>%
  pivot_longer(-immucan_id) %>%
  left_join(traits, "immucan_id") %>%
  filter(!is.na(KRAS),
         LUAD_vs_LUSC == "Adenocarcinoma",
         EGFR == FALSE,
         STK11 == FALSE,
         TP53 == FALSE) %>%
  mutate(KRAS = ifelse(KRAS == TRUE, "mut", "wt")) %>%
  ggplot(aes(x = name,y = value, fill = KRAS))+
  geom_boxplot()+
  scale_y_log10()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust= 1))+
  stat_compare_means(method = "wilcox.test",label = "p.format")

feats <- ctr %>% 
  arrange(desc(Thelper_MHCII)) %>%
  slice_head(n = 20) %>%
  pull(X)
  
im %>%
  dplyr::select(all_of(feats)) %>%
  mutate(immucan_id = rownames(im)) %>%
  pivot_longer(-immucan_id) %>%
  left_join(traits, "immucan_id") %>%
  filter(!is.na(EGFR),
         LUAD_vs_LUSC == "Adenocarcinoma") %>%
  mutate(EGFR = ifelse(EGFR == TRUE, "mut", "wt")) %>%
  ggplot(aes(x = name,y = value, fill = EGFR))+
  geom_boxplot()+
  scale_y_log10()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust= 1))+
  stat_comp

# STK11 mutation

```{r}
cur_var <- "STK11"
var_A <- "mut"
var_B <-  "wt"
col_A <- "blue"
col_B <- "red"
title_name <- "STK11"
```

```{r}
table(traits$STK11, traits$LUAD_vs_LUSC)
```


## STK11 mutation

```{r}
cur_var <- "STK11"
var_A <- "mut"
var_B <-  "wt"
col_A <- "blue"
col_B <- "red"
title_name <- "STK11"
```

```{r}
table(traits$STK11, traits$LUAD_vs_LUSC)
```

## imaging modules

```{r}
dat_z <- left_join(modules, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"),
         STK11 = ifelse(STK11 == TRUE, "mut", "wt")) %>%
  filter(LUAD_vs_LUSC == "LUAD",
         EGFR == FALSE,
         KRAS == FALSE,
         TP53 == FALSE)

print(table(dat_z[[cur_var]])  )
dat_z <- dat_z %>%
  pivot_longer(cols = 1:8)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B)
p_LUAD <- result$plot +
  ggtitle(paste0("Imaging modules for ", title_name," in LUAD"))+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))
p_LUAD
```

## RNA modules

```{r}
dat_z <- left_join(rna, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"),
         STK11 = ifelse(STK11 == TRUE, "mut", "wt")) %>%
  filter(LUAD_vs_LUSC == "LUAD",
         EGFR == FALSE,
         KRAS == FALSE,
         TP53 == FALSE)


print(table(dat_z[[cur_var]])  )
dat_z <- dat_z %>%
  pivot_longer(cols = 1:29)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B)
p_LUAD <- result$plot +
  ggtitle(paste0("Imaging modules for ", title_name," in LUAD"))+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))
p_LUAD
```

## inflammation 

### LUAD 
```{r }
dat_z <- left_join(rna_scores, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"),
         STK11 = ifelse(STK11 == TRUE, "mut", "wt")) %>%
  filter(LUAD_vs_LUSC == "LUAD",
         EGFR == FALSE,
         KRAS == FALSE,
         TP53 == FALSE)


print(table(dat_z[[cur_var]])  )
dat_z <- dat_z %>%
  pivot_longer(cols = 1:13)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B)

p_LUAD <- result$plot+
    ylab("Inf. signatures")+
    ggtitle(paste0("Inf. signatures for ", title_name," in LUAD"))+
  theme(axis.title.y = element_blank())+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))

p_LUAD

cur_order <- result$data$name
```





## sanity check

```{r}
ctr <- read.csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/image_feature_module_correlation.csv")
im <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/imaging_mat.rds")
feats <- ctr %>% 
  arrange(desc(Neutrophils)) %>%
  slice_head(n = 20) %>%
  pull(X)
  
im %>%
  dplyr::select(all_of(feats)) %>%
  mutate(immucan_id = rownames(im)) %>%
  pivot_longer(-immucan_id) %>%
  left_join(traits, "immucan_id") %>%
  filter(!is.na(STK11),
         LUAD_vs_LUSC == "Adenocarcinoma",
         EGFR == FALSE,
         KRAS == FALSE,
         TP53 == FALSE) %>%
  mutate(STK11 = ifelse(STK11 == TRUE, "mut", "wt")) %>%
  ggplot(aes(x = name,y = value, fill = STK11))+
  geom_boxplot()+
  scale_y_log10()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust= 1))+
  stat_compare_means(method = "wilcox.test",label = "p.format")

```




## PIK3CA mutation

```{r}
cur_var <- "PIK3CA"
var_A <- "mut"
var_B <-  "wt"
col_A <- "blue"
col_B <- "red"
title_name <- "PIK3CA"
```

```{r}
table(traits$PIK3CA, traits$LUAD_vs_LUSC)
```

## imaging modules

```{r}
dat_z <- left_join(modules, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"),
         PIK3CA = ifelse(PIK3CA == TRUE, "mut", "wt")) %>%
  filter(LUAD_vs_LUSC == "LUSC")

print(table(dat_z[[cur_var]])  )
dat_z <- dat_z %>%
pivot_longer(cols = 1:8)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B)
p <- result$plot +
  ggtitle(paste0("Imaging modules for ", title_name," in LUSC"))+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))
p
```

## RNA modules

```{r}
dat_z <- left_join(rna, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"),
         PIK3CA = ifelse(PIK3CA == TRUE, "mut", "wt")) %>%
  filter(LUAD_vs_LUSC == "LUSC")

print(table(dat_z[[cur_var]])  )
dat_z <- dat_z %>%
pivot_longer(cols = 1:29)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B)
p <- result$plot +
  ggtitle(paste0("Imaging modules for ", title_name," in LUSC"))+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))
p
```

## inflammation 

### LUSC 
```{r }
dat_z <- left_join(rna_scores, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"),
         PIK3CA = ifelse(PIK3CA == TRUE, "mut", "wt")) %>%
  filter(LUAD_vs_LUSC == "LUSC")

print(table(dat_z[[cur_var]])  )
dat_z <- dat_z %>%
pivot_longer(cols = 1:13)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B)

p <- result$plot+
    ylab("Inf. signatures")+
    ggtitle(paste0("Inf. signatures for ", title_name," in LUSC"))+
  theme(axis.title.y = element_blank())+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))

p
```



## CDKN2A mutation

```{r}
cur_var <- "CDKN2A"
var_A <- "mut"
var_B <-  "wt"
col_A <- "blue"
col_B <- "red"
title_name <- "CDKN2A"
```

```{r}
table(traits$CDKN2A, traits$LUAD_vs_LUSC)
```

## imaging modules

```{r}
dat_z <- left_join(modules, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"),
         CDKN2A = ifelse(CDKN2A == TRUE, "mut", "wt")) %>%
  filter(LUAD_vs_LUSC == "LUSC")

print(table(dat_z[[cur_var]])  )
dat_z <- dat_z %>%
pivot_longer(cols = 1:8)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B)
result$data
p <- result$plot +
  ggtitle(paste0("Imaging modules for ", title_name," in LUSC"))+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))
p
```

## RNA modules

```{r}
dat_z <- left_join(rna, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"),
         CDKN2A = ifelse(CDKN2A == TRUE, "mut", "wt")) %>%
  filter(LUAD_vs_LUSC == "LUSC")

print(table(dat_z[[cur_var]])  )
dat_z <- dat_z %>%
pivot_longer(cols = 1:29)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B)
p <- result$plot +
  ggtitle(paste0("Imaging modules for ", title_name," in LUSC"))+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))
p
```

## inflammation 

### LUSC
```{r }
dat_z <- left_join(rna_scores, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"),
         CDKN2A = ifelse(CDKN2A == TRUE, "mut", "wt")) %>%
  filter(LUAD_vs_LUSC == "LUSC")

print(table(dat_z[[cur_var]])  )
dat_z <- dat_z %>%
pivot_longer(cols = 1:13)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B)

p <- result$plot+
    ylab("Inf. signatures")+
    ggtitle(paste0("Inf. signatures for ", title_name," in LUSC"))+
  theme(axis.title.y = element_blank())+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))

p
```












## KEAP1 mutation

```{r}
cur_var <- "KEAP1"
var_A <- "mut"
var_B <-  "wt"
col_A <- "blue"
col_B <- "red"
title_name <- "KEAP1"
```

```{r}
table(traits$KEAP1, traits$LUAD_vs_LUSC)
```

## imaging modules

```{r}
dat_z <- left_join(modules, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"),
         KEAP1 = ifelse(KEAP1 == TRUE, "mut", "wt")) %>%
  filter(LUAD_vs_LUSC == "LUAD")

print(table(dat_z[[cur_var]])  )
dat_z <- dat_z %>%
pivot_longer(cols = 1:8)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B)
result$data
p <- result$plot +
  ggtitle(paste0("Imaging modules for ", title_name," in LUAD"))+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))
p
```

## RNA modules

```{r}
dat_z <- left_join(rna, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"),
         KEAP1 = ifelse(KEAP1 == TRUE, "mut", "wt")) %>%
  filter(LUAD_vs_LUSC == "LUAD")

print(table(dat_z[[cur_var]])  )
dat_z <- dat_z %>%
pivot_longer(cols = 1:29)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B)
p <- result$plot +
  ggtitle(paste0("Imaging modules for ", title_name," in LUAD"))+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))
p
```

## inflammation 

### LUSC
```{r }
dat_z <- left_join(rna_scores, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"),
         KEAP1 = ifelse(KEAP1 == TRUE, "mut", "wt")) %>%
  filter(LUAD_vs_LUSC == "LUAD")

print(table(dat_z[[cur_var]])  )
dat_z <- dat_z %>%
pivot_longer(cols = 1:13)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B)

p <- result$plot+
    ylab("Inf. signatures")+
    ggtitle(paste0("Inf. signatures for ", title_name," in LUAD"))+
  theme(axis.title.y = element_blank())+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))

p
```











## TP53 LUSC mutation

```{r}
cur_var <- "TP53"
var_A <- "mut"
var_B <-  "wt"
col_A <- "blue"
col_B <- "red"
title_name <- "TP53"
```

```{r}
table(traits$CDKN2A, traits$LUAD_vs_LUSC)
```

## imaging modules

```{r}
dat_z <- left_join(modules, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"),
         TP53 = ifelse(TP53 == TRUE, "mut", "wt")) %>%
  filter(LUAD_vs_LUSC == "LUSC")

print(table(dat_z[[cur_var]])  )
dat_z <- dat_z %>%
pivot_longer(cols = 1:8)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B)
result$data
p <- result$plot +
  ggtitle(paste0("Imaging modules for ", title_name," in LUSC"))+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))
p
```

## RNA modules

```{r}
dat_z <- left_join(rna, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"),
         TP53 = ifelse(TP53 == TRUE, "mut", "wt")) %>%
  filter(LUAD_vs_LUSC == "LUSC")

print(table(dat_z[[cur_var]])  )
dat_z <- dat_z %>%
pivot_longer(cols = 1:29)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B)
p <- result$plot +
  ggtitle(paste0("Imaging modules for ", title_name," in LUSC"))+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))
p
```

## inflammation 

### LUSC
```{r }
dat_z <- left_join(rna_scores, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"),
         TP53 = ifelse(TP53 == TRUE, "mut", "wt")) %>%
  filter(LUAD_vs_LUSC == "LUSC")

print(table(dat_z[[cur_var]])  )
dat_z <- dat_z %>%
pivot_longer(cols = 1:13)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B)

p <- result$plot+
    ylab("Inf. signatures")+
    ggtitle(paste0("Inf. signatures for ", title_name," in LUSC"))+
  theme(axis.title.y = element_blank())+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))

p
```












## TP53 LUAD mutation

```{r}
cur_var <- "TP53"
var_A <- "mut"
var_B <-  "wt"
col_A <- "blue"
col_B <- "red"
title_name <- "TP53"
```

```{r}
table(traits$TP53, traits$LUAD_vs_LUSC)
```

## imaging modules

only in current smokers
```{r}
dat_z <- left_join(modules, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"),
         TP53 = ifelse(TP53 == TRUE, "mut", "wt")) %>%
  filter(LUAD_vs_LUSC == "LUAD",
         smoking_status == "Current smoker")

print(table(dat_z[[cur_var]]))
dat_z <- dat_z %>%
pivot_longer(cols = 1:8)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B)
result$data
p <- result$plot +
  ggtitle(paste0("Imaging modules for ", title_name," in LUAD"))+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))
p
```

## RNA modules

```{r}
dat_z <- left_join(rna, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"),
         TP53 = ifelse(TP53 == TRUE, "mut", "wt")) %>%
  filter(LUAD_vs_LUSC == "LUAD",
         smoking_status == "Current smoker")

print(table(dat_z[[cur_var]])  )
dat_z <- dat_z %>%
pivot_longer(cols = 1:29)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B)
p <- result$plot +
  ggtitle(paste0("RNA modules for ", title_name," in LUAD"))+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))
p
```

## inflammation 

### LUAD
```{r }
dat_z <- left_join(rna_scores, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"),
         TP53 = ifelse(TP53 == TRUE, "mut", "wt")) %>%
  filter(LUAD_vs_LUSC == "LUAD",
         smoking_status == "Current smoker")

print(table(dat_z[[cur_var]])  )
dat_z <- dat_z %>%
pivot_longer(cols = 1:13)

result <- lollipop_plot(dat_z, condition = cur_var, var_A = var_A, var_B = var_B, col_A = col_A, col_B = col_B)

p <- result$plot+
    ylab("Inf. signatures")+
    ggtitle(paste0("Inf. signatures for ", title_name," in LUAD"))+
  theme(axis.title.y = element_blank())+
  coord_cartesian(xlim = c(min(result$data$delta_z)-0.1,max(result$data$delta_z)+0.1))

p
```












## 4B Driver mutations

### LUAD

```{r, fig.height=15, fig.width=20}
condition <- c(FALSE,FALSE,FALSE, FALSE)
names(condition) <- c("KRAS","EGFR", "STK11", "TP53")

# loop over conditions and do two way abundance difference tests
p.list <- list()
for (i in c(data_type,"cello1","cello2")) {
  for (j in names(condition)) {
    
    if(grepl("cello",i)){
      abundances <- cello_dat %>%
        select(contains(i))
      abundances <- t(abundances)
      colnames(abundances) <- cello_dat$immucan_id
    } else {
      abundances <- dat %>%
        select(contains(i)) %>%
        filter(complete.cases(.))
      abundances <- t(abundances)
      colnames(abundances) <-dat %>%
        select(contains(i),immucan_id) %>%
        filter(complete.cases(.)) %>%
        pull(immucan_id)
    }
    anno <- traits
    
    meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
    meta$histology_cond <- paste0(meta$simple_histology,"_",meta[,j])
    Size <- factor(meta[,"histology_cond"])
    meta$Size <- Size
    
    # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
    y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
    design <- model.matrix(~0+group,data = y.ab$samples)
    colnames(design) <- levels(y.ab$samples$group)
    # we do not fit a trend 
    y.ab <- estimateDisp(y.ab, design, trend.method = "none")
    y.ab$common.dispersion
    #plotBCV(y.ab,pch = 5)
    
    # subset design matrix for Adenocarcinoma
    design <- design[,c("Adenocarcinoma_FALSE","Adenocarcinoma_TRUE")]
    
    fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
    fit_LUAD <- glmLRT(glmfit = fit.ab,contrast = c(-1,1))

    res_list <-list(fit_LUAD)
    cur_list <- lapply(as.list(res_list),function(x){
      sig_dat <- x$table
      sig_dat$celltypes <- rownames(sig_dat)
      sig_dat$cond <- j
      sig_dat$comparison <- grep("Adeno|Squamous",x$comparison,value = TRUE)
      return(sig_dat)
    })

    p.list[[paste0(i,"_",j)]] <- do.call(rbind,cur_list)
    
  }
}

# define fold change cutoff for plotting
cutoff <- 0

cur_dat <- do.call(rbind,p.list)
cur_dat <- cur_dat %>%
  mutate(tech = gsub("_abundance","",str_split_i(celltypes,"__",2)),
         cond = paste0(cond, "_mut")) %>%
  group_by(tech, cond) %>%
  mutate(FDR = p.adjust(PValue,method = "BH"),
         sig = ifelse(PValue < 0.05 & logFC > 0, "UP_trend",NA),
         sig = ifelse(FDR < 0.1 & logFC > 0, "UP",sig),
         sig = ifelse(PValue < 0.05 & logFC < 0, "DOWN_trend",sig),
         sig = ifelse(FDR < 0.1 & logFC < 0, "DOWN",sig),
         sig = ifelse(is.na(sig),"NOTSIG", sig),
         ct_name = gsub("_density","",str_split_i(celltypes,"__",1)))
  
cur_dat_LUAD <- cur_dat 

# identify histology
cur_dat_LUAD <- cur_dat_LUAD %>% 
  mutate(cond = paste0("LUAD_",cond))

to_show_LUAD <- cur_dat %>%
  filter(sig %in% c("UP", "DOWN")) %>%
  pull(celltypes)

cur_dat <- cur_dat %>%
  filter(celltypes %in% to_show_LUAD) %>%
  mutate(celltypes = gsub("_abundance","",celltypes))


# spatial cluster renaming
cur_dat <- cur_dat %>%
  mutate(celltypes = gsub(pattern = "_cello",replacement = "IMC",x = celltypes))

# get max and min fold change values
max_up <- max(cur_dat$logFC)+1
min_down <- min(cur_dat$logFC)-1

# plot the data
p <- cur_dat %>%
  ggplot(aes(x=celltypes, y = cond,col = logFC, size= -log10(PValue)))+
  #geom_tile(aes(fill = sig))+
  #theme_bw()+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(round(min_down),0,round(max_up))),
                         guide = "colorbar", limits=c(round(min_down),round(max_up)))+
  #theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  #      panel.background = element_blank(), axis.line = element_line(colour = "black"))+
  #theme_bw()+
  #scale_color_gradient(low = "white",   high = "red")+
  scale_size(range = c(1,8))+
  new_scale_colour() +
  geom_point(data = cur_dat,shape = 1,aes(size = -log10(PValue), col = sig), stroke =2.5)+
  scale_color_manual(values = c("UP" = "#9f1214","UP_trend" = "#e73032","DOWN" = "#25567d","DOWN_trend" = "#428bc7","NOTSIG" = "snow2"), name = "stats")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 20),
        xis.text.y = element_text(size = 20),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 20),
        legend.location = "plot",
        axis.title.x = element_blank(),
        axis.title.y = element_blank())+
  coord_flip()
  #theme(plot.margin = unit(c(5, 2.5, 2.5, 2.5), "cm"))
p
```

### LUAD TP53 only in smokers

```{r, fig.height=15, fig.width=20}
cur_traits <- traits %>%
  filter(smoking_status == "Current smoker")

# loop over conditions and do two way abundance difference tests
p.list <- list()
for (i in c(data_type,"cello1","cello2")) {
  for (j in "TP53") {
    
    if(grepl("cello",i)){
      abundances <- cello_dat %>%
        select(contains(i))
      abundances <- t(abundances)
      colnames(abundances) <- cello_dat$immucan_id
    } else {
      abundances <- dat %>%
        select(contains(i)) %>%
        filter(complete.cases(.))
      abundances <- t(abundances)
      colnames(abundances) <-dat %>%
        select(contains(i),immucan_id) %>%
        filter(complete.cases(.)) %>%
        pull(immucan_id)
    }
    anno <- cur_traits
    
    meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
    meta$histology_cond <- paste0(meta$simple_histology,"_",meta[,j])
    Size <- factor(meta[,"histology_cond"])
    meta$Size <- Size
    
    # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
    y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
    design <- model.matrix(~0+group,data = y.ab$samples)
    colnames(design) <- levels(y.ab$samples$group)
    # we do not fit a trend 
    y.ab <- estimateDisp(y.ab, design, trend.method = "none")
    y.ab$common.dispersion
    #plotBCV(y.ab,pch = 5)
    
    # subset design matrix for Adenocarcinoma
    design <- design[,c("Adenocarcinoma_FALSE","Adenocarcinoma_TRUE")]
    
    fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
    fit_LUAD <- glmLRT(glmfit = fit.ab,contrast = c(-1,1))

    res_list <-list(fit_LUAD)
    cur_list <- lapply(as.list(res_list),function(x){
      sig_dat <- x$table
      sig_dat$celltypes <- rownames(sig_dat)
      sig_dat$cond <- j
      sig_dat$comparison <- grep("Adeno|Squamous",x$comparison,value = TRUE)
      return(sig_dat)
    })

    p.list[[paste0(i,"_",j)]] <- do.call(rbind,cur_list)
    
  }
}

# define fold change cutoff for plotting
cutoff <- 0

cur_dat <- do.call(rbind,p.list)
cur_dat <- cur_dat %>%
  mutate(tech = gsub("_abundance","",str_split_i(celltypes,"__",2)),
         cond = paste0(cond, "_mut")) %>%
  group_by(tech, cond) %>%
  mutate(FDR = p.adjust(PValue,method = "BH"),
         sig = ifelse(PValue < 0.05 & logFC > 0, "UP_trend",NA),
         sig = ifelse(FDR < 0.1 & logFC > 0, "UP",sig),
         sig = ifelse(PValue < 0.05 & logFC < 0, "DOWN_trend",sig),
         sig = ifelse(FDR < 0.1 & logFC < 0, "DOWN",sig),
         sig = ifelse(is.na(sig),"NOTSIG", sig),
         ct_name = gsub("_density","",str_split_i(celltypes,"__",1)))
  
cur_dat_LUAD_TP53_smoker <- cur_dat 

# identify histology
cur_dat_LUAD_TP53_smoker <- cur_dat_LUAD_TP53_smoker %>% 
  mutate(cond = paste0("LUAD_smoker_",cond))

to_show_LUAD_smoker <- cur_dat %>%
  filter(sig %in% c("UP", "DOWN")) %>%
  pull(celltypes)

cur_dat <- cur_dat %>%
  filter(celltypes %in% to_show_LUAD_smoker) %>%
  mutate(celltypes = gsub("_abundance","",celltypes))


# spatial cluster renaming
cur_dat <- cur_dat %>%
  mutate(celltypes = gsub(pattern = "_cello",replacement = "IMC",x = celltypes))

# get max and min fold change values
max_up <- max(cur_dat$logFC)+1
min_down <- min(cur_dat$logFC)-1

# plot the data
p <- cur_dat %>%
  ggplot(aes(x=celltypes, y = cond,col = logFC, size= -log10(PValue)))+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(round(min_down),0,round(max_up))),
                         guide = "colorbar", limits=c(round(min_down),round(max_up)))+
  scale_size(range = c(1,8))+
  new_scale_colour() +
  geom_point(data = cur_dat,shape = 1,aes(size = -log10(PValue), col = sig), stroke =2.5)+
  scale_color_manual(values = c("UP" = "#9f1214","UP_trend" = "#e73032","DOWN" = "#25567d","DOWN_trend" = "#428bc7","NOTSIG" = "snow2"), name = "stats")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 20),
        xis.text.y = element_text(size = 20),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 20),
        legend.location = "plot",
        axis.title.x = element_blank(),
        axis.title.y = element_blank())+
  coord_flip()
  #theme(plot.margin = unit(c(5, 2.5, 2.5, 2.5), "cm"))
p
cur_dat %>%
  filter(celltypes == "Tumor_Ki67__IF2")
```

### Wilcoxon test for EGFR - not in paper
here we see that actually the differential abundance test does not capture the difference in HLADR+ tumor cells while there is a strong difference based on a Wilcoxon or t-test.
```{r}
dat %>%
  left_join(traits,"immucan_id") %>%
  filter(LUAD_vs_LUSC == "Adenocarcinoma") %>%
  ggplot(aes(x = EGFR,y = log2(Tumor_HLADR_abundance__IMC1)))+
  geom_boxplot()+
  stat_compare_means(method = "t.test")
dens_dat %>%
  left_join(traits,"immucan_id") %>%
  filter(LUAD_vs_LUSC == "Adenocarcinoma") %>%
  ggplot(aes(x = EGFR,y = log2(Tumor_HLADR_density__IMC1)))+
  geom_boxplot()+
  stat_compare_means(method = "t.test")

# we run systematic wilcoxon tests for all markers for EGFR
dat %>%
  select(contains("immucan_id"), contains("IMC1")) %>%
  pivot_longer(cols = -immucan_id) %>%
  left_join(traits,"immucan_id") %>%
  filter(LUAD_vs_LUSC == "Adenocarcinoma") %>%
  group_by(name) %>%
  nest() %>%
  mutate(test = map(data, ~ wilcox.test(value ~ EGFR, data = .x)),
         summary = map(test, broom::tidy))%>%
  select(name, summary) %>%
  unnest(summary)

w_test <- dat %>%
  select(contains("immucan_id"), contains("IMC1")) %>%
  pivot_longer(cols = -immucan_id) %>%
  left_join(traits,"immucan_id") %>%
  filter(LUAD_vs_LUSC == "Adenocarcinoma") %>%
  group_by(name) %>%
  nest() %>%
  mutate(
    # Wilcoxon test
    test = map(data, ~ wilcox.test(value ~ EGFR, data = .x)),
    summary = map(test, broom::tidy),
    
    # Compute medians for fold change
    medians = map(data, ~ .x %>%
                    group_by(EGFR) %>%
                    summarise(median = median(value, na.rm = TRUE), .groups = "drop") %>%
                    pivot_wider(names_from = EGFR, values_from = median, names_prefix = "EGFR_")
                 ),
    
    # Compute fold change: TRUE / FALSE
    fold_change = map_dbl(medians, ~ .x$EGFR_TRUE / .x$EGFR_FALSE)
  ) %>%
  select(name, summary, fold_change) %>%
  unnest(summary)

w_test <- w_test %>%
  mutate(test = "EGFR",
         sig = p.value < 0.05) 
  
w_test %>%
  ggplot(aes(x=name, y = test,col = log2(fold_change), size= -log10(p.value)))+
  #geom_tile(aes(fill = sig))+
  #theme_bw()+
  geom_point()+
  theme_bw()+
  scale_colour_gradient2(low = "blue", mid = "white",high ="red")+
  new_scale_colour() +
  geom_point(data = w_test %>% filter(sig == TRUE),shape = 1,aes(size = -log10(p.value), col = sig), stroke =2.5)+
  scale_color_manual(values = c("TRUE" = "black","FALSE" = "white"))+
coord_flip()
```


### LUSC

```{r, fig.height=10, fig.width=20}
condition <- c(FALSE,FALSE, FALSE)
names(condition) <- c("PIK3CA", "TP53", "CDKN2A")

# loop over conditions and do two way abundance difference tests
p.list <- list()
for (i in c(data_type,"cello1","cello2")) {
  for (j in names(condition)) {
    
    if(grepl("cello",i)){
      abundances <- cello_dat %>%
        select(contains(i))
      abundances <- t(abundances)
      colnames(abundances) <- cello_dat$immucan_id
    } else {
      abundances <- dat %>%
        select(contains(i)) %>%
        filter(complete.cases(.))
      abundances <- t(abundances)
      colnames(abundances) <-dat %>%
        select(contains(i),immucan_id) %>%
        filter(complete.cases(.)) %>%
        pull(immucan_id)
    }
    anno <- traits
    
    meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
    meta$histology_cond <- paste0(meta$simple_histology,"_",meta[,j])
    Size <- factor(meta[,"histology_cond"])
    meta$Size <- Size
    
    # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
    y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
    design <- model.matrix(~0+group,data = y.ab$samples)
    colnames(design) <- levels(y.ab$samples$group)
    # we do not fit a trend 
    y.ab <- estimateDisp(y.ab, design, trend.method = "none")
    y.ab$common.dispersion
    #plotBCV(y.ab,pch = 5)
    
    # subset design matrix  for Adenocarcinoma
    design <- design[,c("Squamous cell carcinoma_FALSE","Squamous cell carcinoma_TRUE")]
    # we only compare the diff abundance for Adeno vs Sqam.
    fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
    fit_LUSC <- glmLRT(glmfit = fit.ab,contrast = c(-1,1))

    res_list <- list( fit_LUSC)
    
    cur_list <- lapply(res_list,function(x){
      sig_dat <- x$table
      sig_dat$celltypes <- rownames(sig_dat)
      sig_dat$cond <- j
      sig_dat$comparison <- grep("Adeno|Squamous",x$comparison,value = TRUE)
      return(sig_dat)
    })

    p.list[[paste0(i,"_",j)]] <- do.call(rbind,cur_list)
    
  }
}

cur_dat <- do.call(rbind,p.list)
cur_dat <- cur_dat %>%
  mutate(tech = gsub("_abundance","",str_split_i(celltypes,"__",2)),
         cond = paste0(cond, "_mut")) %>%
  group_by(tech, cond) %>%
  mutate(FDR = p.adjust(PValue,method = "BH"),
         sig = ifelse(PValue < 0.05 & logFC > 0, "UP_trend",NA),
         sig = ifelse(FDR < 0.1 & logFC > 0, "UP",sig),
         sig = ifelse(PValue < 0.05 & logFC < 0, "DOWN_trend",sig),
         sig = ifelse(FDR < 0.1 & logFC < 0, "DOWN",sig),
         sig = ifelse(is.na(sig),"NOTSIG", sig),
         ct_name = gsub("_density","",str_split_i(celltypes,"__",1)))
  
cur_dat_LUSC <- cur_dat

# identify histology
cur_dat_LUSC <- cur_dat_LUSC %>% 
  mutate(cond = paste0("LUSC_",cond))

to_show_LUSC <- cur_dat %>%
  filter(sig %in% c("UP", "DOWN")) %>%
  pull(celltypes)

cur_dat <- cur_dat %>%
  filter(celltypes %in% to_show_LUSC) %>%
  mutate(celltypes = gsub("_abundance","",celltypes))


# spatial cluster renaming
cur_dat <- cur_dat %>%
  mutate(celltypes = gsub(pattern = "_cello",replacement = "IMC",x = celltypes))

# get max and min fold change values
max_up <- max(cur_dat$logFC)+1
min_down <- min(cur_dat$logFC)-1

# plot the data
p <- cur_dat %>%
  ggplot(aes(x=celltypes, y = cond,col = logFC, size= -log10(PValue)))+
  #geom_tile(aes(fill = sig))+
  #theme_bw()+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(round(min_down),0,round(max_up))),
                         guide = "colorbar", limits=c(round(min_down),round(max_up)))+
  #theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  #      panel.background = element_blank(), axis.line = element_line(colour = "black"))+
  #theme_bw()+
  #scale_color_gradient(low = "white",   high = "red")+
  scale_size(range = c(1,8))+
  new_scale_colour() +
  geom_point(data = cur_dat,shape = 1,aes(size = -log10(PValue), col = sig), stroke =2.5)+
  scale_color_manual(values = c("UP" = "#9f1214","UP_trend" = "#e73032","DOWN" = "#25567d","DOWN_trend" = "#428bc7","NOTSIG" = "snow2"), name = "stats")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 20),
        xis.text.y = element_text(size = 20),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 20),
        legend.location = "plot",
        axis.title.x = element_blank(),
        axis.title.y = element_blank())+
  coord_flip()
  #theme(plot.margin = unit(c(5, 2.5, 2.5, 2.5), "cm"))
p

```

### all
not in paper
```{r, fig.height=10, fig.width=20}
condition <- c(FALSE,FALSE)
names(condition) <- c("NF1","KEAP1")

# loop over conditions and do two way abundance difference tests
p.list <- list()
for (i in c(data_type,"cello1","cello2")) {
  for (j in names(condition)) {
    
    if(grepl("cello",i)){
      abundances <- cello_dat %>%
        select(contains(i))
      abundances <- t(abundances)
      colnames(abundances) <- cello_dat$immucan_id
    } else {
      abundances <- dat %>%
        select(contains(i)) %>%
        filter(complete.cases(.))
      abundances <- t(abundances)
      colnames(abundances) <-dat %>%
        select(contains(i),immucan_id) %>%
        filter(complete.cases(.)) %>%
        pull(immucan_id)
    }
    anno <- traits
    
    meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
    meta$histology_cond <- paste0(meta[,j])
    Size <- factor(meta[,"histology_cond"])
    meta$Size <- Size
    
    # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
    y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
    design <- model.matrix(~0+group,data = y.ab$samples)
    colnames(design) <- levels(y.ab$samples$group)
    # we do not fit a trend 
    y.ab <- estimateDisp(y.ab, design, trend.method = "none")
    y.ab$common.dispersion
    #plotBCV(y.ab,pch = 5)
    
    # subset design matrix  for Adenocarcinoma
    design <- design[,c("FALSE","TRUE")]
    # we only compare the diff abundance for Adeno vs Sqam.
    fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
    fit_LUSC <- glmLRT(glmfit = fit.ab,contrast = c(-1,1))

    res_list <- list( fit_LUSC)
    
    cur_list <- lapply(res_list,function(x){
      sig_dat <- x$table
      sig_dat$celltypes <- rownames(sig_dat)
      sig_dat$cond <- j
      sig_dat$comparison <- x$comparison
      return(sig_dat)
    })

    p.list[[paste0(i,"_",j)]] <- do.call(rbind,cur_list)
    
  }
}

cur_dat <- do.call(rbind,p.list)
cur_dat <- cur_dat %>%
  mutate(tech = gsub("_abundance","",str_split_i(celltypes,"__",2)),
         cond = paste0(cond, "_mut")) %>%
  group_by(tech, cond) %>%
  mutate(FDR = p.adjust(PValue,method = "BH"),
         sig = ifelse(PValue < 0.05 & logFC > 0, "UP_trend",NA),
         sig = ifelse(FDR < 0.1 & logFC > 0, "UP",sig),
         sig = ifelse(PValue < 0.05 & logFC < 0, "DOWN_trend",sig),
         sig = ifelse(FDR < 0.1 & logFC < 0, "DOWN",sig),
         sig = ifelse(is.na(sig),"NOTSIG", sig),
         ct_name = gsub("_density","",str_split_i(celltypes,"__",1)))
  
cur_dat_all <- cur_dat

# identify histology
cur_dat_all <- cur_dat_all %>% 
  mutate(cond = paste0("All_",cond))

to_show_all <- cur_dat %>%
  filter(sig %in% c("UP", "DOWN")) %>%
  pull(celltypes)

cur_dat <- cur_dat %>%
  filter(celltypes %in% to_show_all) %>%
  mutate(celltypes = gsub("_abundance","",celltypes))


# spatial cluster renaming
cur_dat <- cur_dat %>%
  mutate(celltypes = gsub(pattern = "_cello",replacement = "IMC",x = celltypes))

# get max and min fold change values
max_up <- max(cur_dat$logFC)+1
min_down <- min(cur_dat$logFC)-1

# plot the data
p <- cur_dat %>%
  ggplot(aes(x=celltypes, y = cond,col = logFC, size= -log10(PValue)))+
  #geom_tile(aes(fill = sig))+
  #theme_bw()+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(round(min_down),0,round(max_up))),
                         guide = "colorbar", limits=c(round(min_down),round(max_up)))+
  #theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  #      panel.background = element_blank(), axis.line = element_line(colour = "black"))+
  #theme_bw()+
  #scale_color_gradient(low = "white",   high = "red")+
  scale_size(range = c(1,8))+
  new_scale_colour() +
  geom_point(data = cur_dat,shape = 1,aes(size = -log10(PValue), col = sig), stroke =2.5)+
  scale_color_manual(values = c("UP" = "#9f1214","UP_trend" = "#e73032","DOWN" = "#25567d","DOWN_trend" = "#428bc7","NOTSIG" = "snow2"), name = "stats")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 20),
        xis.text.y = element_text(size = 20),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 20),
        legend.location = "plot",
        axis.title.x = element_blank(),
        axis.title.y = element_blank())+
  coord_flip()
  #theme(plot.margin = unit(c(5, 2.5, 2.5, 2.5), "cm"))
p

```

### plot  LUSC and LUAD together
```{r, fig.height=15, fig.width=10}
cur_dat <- rbind(cur_dat_LUAD,cur_dat_LUSC)
cur_dat <- rbind(cur_dat, cur_dat_LUAD_TP53_smoker)


cur_dat <- cur_dat %>%
  filter(celltypes %in% c(to_show_LUAD,to_show_LUSC, to_show_LUAD_smoker)) %>%
  mutate(celltypes = gsub("_abundance","",celltypes))


# spatial cluster renaming
cur_dat <- cur_dat %>%
  mutate(celltypes = gsub(pattern = "_cello",replacement = "IMC",x = celltypes))

max_up <- max(cur_dat$logFC)+1
min_down <- min(cur_dat$logFC)-1

# plot the data
p <- cur_dat %>%
  ggplot(aes(x=(celltypes), y = factor(cond, levels = c("LUAD_EGFR_mut","LUAD_KRAS_mut","LUAD_STK11_mut","LUAD_TP53_mut", "LUAD_smoker_TP53_mut","LUSC_TP53_mut", "LUSC_PIK3CA_mut", "LUSC_CDKN2A_mut","All_NF1_mut", "All_KEAP1_mut")),col = logFC, size= -log10(PValue)))+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(round(min_down),0,round(max_up))),
                         guide = "colorbar", limits=c(round(min_down),round(max_up)))+
  scale_size(range = c(1,3.5))+
  new_scale_colour() +
  geom_point(data = cur_dat,shape = 1,aes(size = -log10(PValue), col = sig), stroke =1)+
  scale_color_manual(values = c("UP" = "#9f1214","UP_trend" = "#e73032","DOWN" = "#25567d","DOWN_trend" = "#428bc7","NOTSIG" = "snow2"), name = "stats")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 7),
        axis.text.y = element_text(size = 7),
        legend.text = element_text(size = 7),
        legend.title = element_text(size = 7),
        axis.title.x = element_blank(),
        axis.title.y = element_blank())+
  coord_flip()

p

ggsave(filename = paste0(out_folder,"4B_diff_abundance_driver_mutations.pdf"),
       plot = p,
       device = "pdf",
       width = 5,
       height = 8)
```



# expression differences across cell types
```{r, fig.height=7, fig.width=15}
mean_exp1 <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/mean_marker_expr_per_celltype_p1.csv"))
mean_exp2 <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/mean_marker_expr_per_celltype_p2.csv"))

# long format
mean_exp1 <- mean_exp1 %>%
  pivot_longer(-immucan_id)
mean_exp2 <- mean_exp2 %>%
  pivot_longer(-immucan_id)

mean_exp1$tech <- "IMC1"
mean_exp2$tech <- "IMC2"
mean_exp <- rbind(mean_exp1,mean_exp2)

mean_exp <- mean_exp %>%
  mutate(celltypes = str_split_i(name,"-",1),
         name = str_split_i(name,"-",2))
# load sce object panel 1
sce1 <- readRDS(paste0(mount_path,"data/IMC/Panel_1/Rout/sce.rds"))
sce1 <- sce1[,sce1$ROI != 4]
# Panel 2
sce2 <- readRDS(paste0(mount_path,"data/IMC/Panel_2/Rout/sce.rds"))
```



```{r}

mean_exp %>%
  filter(celltypes == "CD8",
         grepl("PD1|LAG|GrzB|Ki67|CD38|TCF7|CD27|ICOS|CXCL13|GITR|Tbet|GATA3|NKG2A|Tim3|Ki67|EOMES",name)) %>%
  left_join(traits, by = "immucan_id") %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  ggplot(aes(x = EGFR, y = value,fill= EGFR))+
  geom_boxplot()+
  facet_wrap(.~name, scales= "free")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  stat_compare_means()+
  coord_cartesian(ylim =c(0,3))
```



# 4C HLADR on tumor cells in EGFR mutated

## Tumor cells HLADR in EGFR
```{r, fig.height=5, fig.width=5}
p_Tumor_HLADR_EGFR <- mean_exp %>%
  filter(celltypes == "Tumor",
         name %in% c("HLADR_p1")) %>%
  left_join(traits, by = "immucan_id") %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  ggplot(aes(x = EGFR, y = value,fill= EGFR))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(aes(color = EGFR),position = position_jitterdodge(jitter.width = 0.2), size= 0.3)+
  scale_fill_manual(values = alpha(c("red", "blue"),alpha = 0.6))+
  scale_color_manual(values = c("red", "blue"))+
  facet_wrap(.~name, scales= "free")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  stat_compare_means(size = 2,label.y = 4,label = "p.format")+
    theme_bw()+
  theme(axis.text.y = element_text(size = 6),
        axis.text.x = element_text(size = 6,angle = 90, vjust = 1, hjust = 0.5),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 6),
        strip.text = element_text(size = 6),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7))+
  theme(legend.position = "none")+
  coord_cartesian(ylim = c(0.2,4.2))
p_Tumor_HLADR_EGFR

ggsave(filename = paste0(out_folder,"4C_HLADR_on_tumor_cells_EGFR.pdf"),
       plot = p_Tumor_HLADR_EGFR,
       device = "pdf",
       width = 1,
       height = 2)
```

## generate IMC images HLADR

```{r, fig.height=20, fig.width=20}
# get the expression matrix and caculate stats
counts <- as.data.frame(t(assay(sce1,"exprs")))
head(counts)
dim(counts)

counts$celltypes <- sce1$celltypes
counts$immucan_id <- sce1$immucan_id
counts$sample_id <- sce1$sample_id
counts <- left_join(counts, traits,"immucan_id")

# select samples from the frequency of positive cells
counts %>%
  mutate(HLADRpos = ifelse(HLADR > 3, TRUE,FALSE)) %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  group_by(immucan_id,EGFR,.drop = FALSE) %>%
  dplyr::count(HLADRpos) %>%
  mutate(total_cells = sum(n),
         frac = n/total_cells) %>%
  filter(HLADRpos == TRUE) %>%
  arrange(frac)

EGFRpos <- c("LUNG-NSCLC2-0624","LUNG-NSCLC2-0689","LUNG-NSCLC2-0659")

# load images
library(cytomapper)
masks <- readRDS(paste0(mount_path,"data/IMC/Panel_1/Rout/masks.rds"))
images <- readRDS(paste0(mount_path,"data/IMC/Panel_1/Rout/images.rds"))

cur_df <- as.data.frame(colData(sce1))
rownames(cur_df) <- NULL
cur_df <- cur_df %>%
  select(immucan_id,image, sample_id) %>%
  distinct() %>%
  mutate(image = gsub(".tiff","",image))
rownames(cur_df) <- cur_df$image


# select images
ims <- cur_df %>% 
  filter(immucan_id %in% EGFRpos,
         grepl("002$", sample_id)) %>%
  pull(sample_id)

# unify metadata of masks and images
mcols(images)$immucan_id <- cur_df[names(images),"immucan_id"]
mcols(masks)$immucan_id <- cur_df[names(masks),"immucan_id"]
mcols(images)$sample_id <- cur_df[names(images),"sample_id"]
mcols(masks)$sample_id <- cur_df[names(masks),"sample_id"]


cur_sce <- sce1[,sce1$sample_id %in% ims]
channelNames(images) <- rownames(cur_sce)

cur_ims <- getImages(images,ims)
cur_masks <- getImages(masks, ims)

# read images into memory and normalize
cur_images <- CytoImageList(cur_ims, on_disk = FALSE)
cur_masks <- CytoImageList(cur_masks, on_disk = FALSE)
cur_images <- scaleImages(cur_images,value = 2^16-1)
cur_images <- normalize(cur_images, separateImages = TRUE)
cur_images <- normalize(cur_images, separateImages = TRUE, 
                        inputRange = c(0, 0.2))

# define all tumor and non-tumor cells
cur_sce$tumor <- "non-tumor"
cur_sce[,cur_sce$celltypes == "Tumor"]$tumor <- "tumor"
plot_sce <- cur_sce[,cur_sce$celltypes == "Tumor"]

# generate images
plotPixels(image = cur_images,
           object = plot_sce,
           cell_id = "ObjectNumber",
           img_id = "sample_id",
           colour_by = c("Ecad", "HLADR", "DNA1"),
           colour = list(Ecad = c("black", "green"),HLADR = c("black","red"),DNA1 = c("black","blue")),
           bcg = list(Ecad  = c(0,2,1),
                      HLADR = c(0,1.5,1),
                      DNA1  = c(0,1,1)),
            image_title = NULL,save_plot = list(filename = paste0(out_folder,"HLADR_tumor_2.png")))
```



# 4D CD8 T cells CXC13 in EGFR
```{r, fig.height=5, fig.width=5}
p_CD8_CXCL13_EGFR <- mean_exp %>%
  filter(celltypes == "CD8",
         name %in% c("CXCL13_p2", "PD1_p2","LAG3_p1", "Tim3_p2")) %>%
  left_join(traits, by = "immucan_id") %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  ggplot(aes(x = EGFR, y = value,fill= EGFR))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(color = EGFR),position = position_jitterdodge(jitter.width = 0.2), size= 0.5)+
  scale_fill_manual(values = alpha(c("red", "blue"),alpha = 0.6))+
  scale_color_manual(values = c("red", "blue"))+
  facet_wrap(.~name, scales= "free",nrow = 1)+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  stat_compare_means(size = 2.7, label = "p.format")+
    theme_bw()+
  theme(axis.text.y = element_text(size = 6),
        axis.text.x = element_text(size = 6, angle = 90, vjust = 1,hjust = 0.5),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 6),
        strip.text = element_text(size = 6))+
  theme(legend.position = "none")
p_CD8_CXCL13_EGFR

ggsave(filename = paste0(out_folder,"4D_exhaustion_on_CD8_Tcells.pdf"),
       plot = p_CD8_CXCL13_EGFR,
       device = "pdf",
       width = 3.5,
       height = 2.5)
```


# 4F Tumor cells EGFR HLADR in STK11
```{r, fig.height=5, fig.width=5}
p_Tumor_HLADR_STK11 <- mean_exp %>%
  filter(celltypes == "Tumor",
         name %in% c("EGFR_p2", "HLADR_p1")) %>%
  left_join(traits, by = "immucan_id") %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  ggplot(aes(x = STK11, y = value,fill= STK11))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(color = STK11),position = position_jitterdodge(jitter.width = 0.2), size= 0.5)+
  scale_fill_manual(values = alpha(c("red", "blue"),alpha = 0.6))+
  scale_color_manual(values = c("red", "blue"))+
  facet_wrap(.~name, scales= "free")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  stat_compare_means(size = 2.5, label = "p.format")+
    theme_bw()+
  theme(axis.text.y = element_text(size = 6),
        axis.text.x = element_text(size = 6, angle = 90, vjust = 1, hjust = 0.5),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 6),
        strip.text = element_text(size = 6),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7))+
  theme(legend.position = "none")
p_Tumor_HLADR_STK11

ggsave(filename = paste0(out_folder,"4F_HLADR_and_EGFR_on_tumor_cells_STK11.pdf"),
       plot = p_Tumor_HLADR_STK11,
       device = "pdf",
       width = 1.8,
       height = 2)
```

# 4E Ki67 and p53 on tumor cells in TP53 mutated in LUAD

```{r, fig.height=5, fig.width=5}
p_Tumor_Ki67_P53_TP53 <- mean_exp %>%
  filter(celltypes == "Tumor",
         name %in% c("p53_p2","Ki67_p1")) %>%
  left_join(traits, by = "immucan_id") %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  ggplot(aes(x = TP53, y = value,fill= TP53))+
  geom_boxplot(outlier.shape = NA)+
  geom_jitter(aes(color = TP53),position = position_jitterdodge(jitter.width = 0.2), size= 0.75)+
  scale_fill_manual(values = alpha(c("red", "blue"),alpha = 0.6))+
  scale_color_manual(values = c("red", "blue"))+
  facet_wrap(.~name, scales= "free")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  stat_compare_means(size = 2.2, label = "p.format")+
    theme_bw()+
  theme(axis.text.y = element_text(size = 6),
        axis.text.x = element_text(size = 6, angle = 90, vjust = 1, hjust = 0.5),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 6),
        strip.text = element_text(size = 6),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7))+
  theme(legend.position = "none")
p_Tumor_Ki67_P53_TP53

ggsave(filename = paste0(out_folder,"4E_Ki67_p53_on_tumor_cells_TP53.pdf"),
       plot = p_Tumor_Ki67_P53_TP53,
       device = "pdf",
       width = 1.8,
       height = 2)
```


## P53 images
```{r, fig.height=20, fig.width=20}
counts2 <- as.data.frame(t(assay(sce2,"exprs")))
head(counts2)
dim(counts2)

counts2$celltypes <- sce2$celltypes
counts2$immucan_id <- sce2$immucan_id
counts2$sample_id <- sce2$sample_id
counts2 %>%
  filter(celltypes == "Tumor") %>%
  ggplot(aes(x = Ecad,y = p53))+
  geom_point()

counts2 <- left_join(counts2, traits,"immucan_id")

counts2 %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  mutate(p53pos = ifelse(p53 > 1, TRUE,FALSE)) %>%
  group_by(sample_id,.drop = FALSE) %>%
  dplyr::count(p53pos) %>%
  mutate(total_cells = sum(n),
         frac = n/total_cells) %>%
  filter(p53pos == TRUE) %>%
  arrange(frac)

p53pos <- c("10040769","10040838")
p53pos <- traits$immucan_id[grepl(paste(p53pos,collapse = "|"),traits$ibbl_kit_ids)]

# load images
library(cytomapper)
masks <- readRDS(paste0(mount_path,"data/IMC/Panel_2/Rout/masks.rds"))
images <- readRDS(paste0(mount_path,"data/IMC/Panel_2/Rout/images.rds"))

cur_df <- as.data.frame(colData(sce2))
rownames(cur_df) <- NULL
cur_df <- cur_df %>%
  select(immucan_id,image, sample_id) %>%
  distinct() %>%
  mutate(image = gsub(".tiff","",image))
rownames(cur_df) <- cur_df$image


# select images
ims <- cur_df %>% 
  filter(immucan_id %in% p53pos,
         grepl("002$", sample_id)) %>%
  pull(sample_id)

# unify metadata of masks and images
mcols(images)$immucan_id <- cur_df[names(images),"immucan_id"]
mcols(masks)$immucan_id <- cur_df[names(masks),"immucan_id"]
mcols(images)$sample_id <- cur_df[names(images),"sample_id"]
mcols(masks)$sample_id <- cur_df[names(masks),"sample_id"]


cur_sce <- sce2[,sce2$sample_id %in% ims]


cur_ims <- getImages(images,ims)
cur_masks <- getImages(masks, ims)

# read images into memory and normalize
cur_images <- CytoImageList(cur_ims, on_disk = FALSE)
cur_masks <- CytoImageList(cur_masks, on_disk = FALSE)
cur_images <- scaleImages(cur_images,value = 2^16-1)
cur_images <- normalize(cur_images, separateImages = TRUE)
cur_images <- normalize(cur_images, separateImages = TRUE, 
                        inputRange = c(0, 0.2))

channelNames(cur_images) <- rownames(cur_sce)

# define all tumor and non-tumor cells
cur_sce$tumor <- "non-tumor"
cur_sce[,cur_sce$celltypes == "Tumor"]$tumor <- "tumor"
plot_sce <- cur_sce[,cur_sce$celltypes == "Tumor"]

# generate images
plotPixels(image = cur_images,
           object = plot_sce,
           cell_id = "ObjectNumber",
           img_id = "sample_id",
           colour_by = c("Ecad", "p53", "DNA1"),
           colour = list(Ecad = c("black", "green"),p53 = c("black","red"),DNA1 = c("black","blue")),
           bcg = list(Ecad  = c(0,2,1),
                      p53 = c(0,1.5,1),
                      DNA1  = c(0,1,1)),
            image_title = NULL,save_plot = list(filename = paste0(out_folder,"p53_tumor.png")))
```

```{r}
counts2 %>%
  filter(celltypes == "Tumor") %>%
  ggplot(aes(x = Ecad,y = p53))+
  geom_point(size = 0.1)+
  facet_wrap(LUAD_vs_LUSC~TP53)+
  theme_bw()

hist(counts2$p53)

counts2 %>%
  mutate(p53pos = ifelse(p53 > 2, TRUE,FALSE)) %>%
  group_by(immucan_id,TP53,.drop = FALSE) %>%
  dplyr::count(p53pos) %>%
  left_join(traits[,c("immucan_id","LUAD_vs_LUSC")],"immucan_id") %>%
  mutate(total_cells = sum(n),
         frac = n/total_cells) %>%
  filter(p53pos == TRUE) %>%
  ggplot(aes(x = TP53,y = log2(frac)))+
  geom_boxplot()+
  facet_wrap(.~LUAD_vs_LUSC, scales = "free")+
  theme_bw()+
  stat_compare_means(method = "wilcox.test")
```




# 4G

## Treg cells GITR PIK3CA
```{r, fig.height=5, fig.width=5}
p_Treg_GITR_PIK3CA <- mean_exp %>%
  filter(celltypes == "Treg",
         name %in% c("GITR_p2","CD137_p2", "GATA3_p2")) %>%
  left_join(traits, by = "immucan_id") %>%
  filter(simple_histology == "Squamous cell carcinoma") %>%
  ggplot(aes(x = PIK3CA, y = value,fill= PIK3CA))+
  geom_boxplot(outliers = FALSE)+
  geom_jitter(aes(color = PIK3CA),position = position_jitterdodge(jitter.width = 0.2), size= 0.5)+
  scale_fill_manual(values = alpha(c("red", "blue"),alpha = 0.6))+
  scale_color_manual(values = c("red", "blue"))+
  facet_wrap(.~name, scales= "free")+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
  stat_compare_means(size = 2, label = "p.format")+
    theme_bw()+
  theme(axis.text.y = element_text(size = 6),
        axis.text.x = element_text(size = 6, angle = 90, vjust = 1, hjust = 0.5),
        legend.text = element_text(size = 6),
        legend.title = element_text(size = 6),
        strip.text = element_text(size = 6),
        axis.title.x = element_text(size = 7),
        axis.title.y = element_text(size = 7))+
  theme(legend.position = "none")
p_Treg_GITR_PIK3CA

ggsave(filename = paste0(out_folder,"4G_activation_markers_on_Tregs.pdf"),
       plot = p_Treg_GITR_PIK3CA,
       device = "pdf",
       width = 2.5,
       height = 2)
```

test smoking and TP53 association in LUAD

```{r, fig.height=15, fig.width=20}
cur_traits <- traits %>%
  filter(smoking_status %in% c("Current smoker"))

condition <- c( FALSE)
names(condition) <- c( "TP53")

# loop over conditions and do two way abundance difference tests
p.list <- list()
for (i in c(data_type,"cello1","cello2")) {
  for (j in names(condition)) {
    
    if(grepl("cello",i)){
      abundances <- cello_dat %>%
        select(contains(i))
      abundances <- t(abundances)
      colnames(abundances) <- cello_dat$immucan_id
    } else {
      abundances <- dat %>%
        select(contains(i)) %>%
        filter(complete.cases(.))
      abundances <- t(abundances)
      colnames(abundances) <-dat %>%
        select(contains(i),immucan_id) %>%
        filter(complete.cases(.)) %>%
        pull(immucan_id)
    }
    anno <- cur_traits
    
    meta <- as.data.frame(anno)[match(colnames(abundances),anno$immucan_id),]
    meta$histology_cond <- paste0(meta$simple_histology,"_",meta[,j])
    Size <- factor(meta[,"histology_cond"])
    meta$Size <- Size
    
    # here we generate the design matrix across all histologies and then also estimate the dispersion across all histologies.
    y.ab <- DGEList(abundances, samples=meta, group = meta$Size)
    design <- model.matrix(~0+group,data = y.ab$samples)
    colnames(design) <- levels(y.ab$samples$group)
    # we do not fit a trend 
    y.ab <- estimateDisp(y.ab, design, trend.method = "none")
    y.ab$common.dispersion
    #plotBCV(y.ab,pch = 5)
    
    
    # subset design matrix for Adenocarcinoma
    design <- design[,c("Adenocarcinoma_FALSE","Adenocarcinoma_TRUE")]
    
    fit.ab <- glmFit(y.ab, design, robust=TRUE, abundance.trend=FALSE)
    fit_LUAD <- glmLRT(glmfit = fit.ab,contrast = c(-1,1))

    res_list <-list(fit_LUAD)
    cur_list <- lapply(as.list(res_list),function(x){
      sig_dat <- x$table
      sig_dat$celltypes <- rownames(sig_dat)
      sig_dat$cond <- j
      sig_dat$comparison <- grep("Adeno|Squamous",x$comparison,value = TRUE)
      return(sig_dat)
    })

    p.list[[paste0(i,"_",j)]] <- do.call(rbind,cur_list)
    
  }
}

# define fold change cutoff for plotting
cutoff <- 0

cur_dat <- do.call(rbind,p.list)
cur_dat <- cur_dat %>%
  mutate(tech = gsub("_abundance","",str_split_i(celltypes,"__",2)),
         cond = paste0(cond, "_mut")) %>%
  group_by(tech, cond) %>%
  mutate(FDR = p.adjust(PValue,method = "BH"),
         sig = ifelse(PValue < 0.05 & logFC > 0, "UP_trend",NA),
         sig = ifelse(FDR < 0.1 & logFC > 0, "UP",sig),
         sig = ifelse(PValue < 0.05 & logFC < 0, "DOWN_trend",sig),
         sig = ifelse(FDR < 0.1 & logFC < 0, "DOWN",sig),
         sig = ifelse(is.na(sig),"NOTSIG", sig),
         ct_name = gsub("_density","",str_split_i(celltypes,"__",1)))
  
cur_dat_LUAD <- cur_dat 

# identify histology
cur_dat_LUAD <- cur_dat_LUAD %>% 
  mutate(cond = paste0("LUAD_",cond))

to_show_LUAD <- cur_dat %>%
  filter(sig %in% c("UP", "DOWN")) %>%
  pull(celltypes)

cur_dat <- cur_dat %>%
  filter(celltypes %in% to_show_LUAD) %>%
  mutate(celltypes = gsub("_abundance","",celltypes))


# spatial cluster renaming
cur_dat <- cur_dat %>%
  mutate(celltypes = gsub(pattern = "_cello",replacement = "IMC",x = celltypes))

# get max and min fold change values
max_up <- max(cur_dat$logFC)+1
min_down <- min(cur_dat$logFC)-1

# plot the data
p <- cur_dat %>%
  ggplot(aes(x=celltypes, y = cond,col = logFC, size= -log10(PValue)))+
  #geom_tile(aes(fill = sig))+
  #theme_bw()+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(round(min_down),0,round(max_up))),
                         guide = "colorbar", limits=c(round(min_down),round(max_up)))+
  #theme(panel.grid.major = element_blank(), panel.grid.minor = element_blank(),
  #      panel.background = element_blank(), axis.line = element_line(colour = "black"))+
  #theme_bw()+
  #scale_color_gradient(low = "white",   high = "red")+
  scale_size(range = c(1,8))+
  new_scale_colour() +
  geom_point(data = cur_dat,shape = 1,aes(size = -log10(PValue), col = sig), stroke =2.5)+
  scale_color_manual(values = c("UP" = "#9f1214","UP_trend" = "#e73032","DOWN" = "#25567d","DOWN_trend" = "#428bc7","NOTSIG" = "snow2"), name = "stats")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 20),
        xis.text.y = element_text(size = 20),
        legend.text = element_text(size = 20),
        legend.title = element_text(size = 20),
        legend.location = "plot",
        axis.title.x = element_blank(),
        axis.title.y = element_blank())+
  coord_flip()
  #theme(plot.margin = unit(c(5, 2.5, 2.5, 2.5), "cm"))
p
```
