---
title: "Figure_4"
author: "Daniel Schulz"
date: "2024-10-04"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

options(bitmapType = "cairo")
```


```{r}
library(tidyverse)
library(ggpubr)
library(SingleCellExperiment)
library(scales)
library(ggnewscale)
library(edgeR)
library(cowplot)
library(igraph)
library(ggraph)
library(ggrepel)
library(rstatix)
```

```{r}
mount_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/"
out_folder <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/revised_figures/Figure_4/"

if (!dir.exists(out_folder)) {
  dir.create(out_folder)
}
source("code/lollipop_function.R")
source("code/bplot_functions.R")
```



# load data
we will load the data we generated at the beginning of figure 2
```{r}
dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/abundances_dat.rds"))
dens_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/densities_dat.rds"))
cello_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/cello_dat.rds"))
clin_dat <- read.csv("/mnt/immucan_volume/clinical_data/Lung_NSCLC2/2026_final/clinical.csv")
int_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/interaction_dat.rds"))
traits <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/all_traits.csv"))

traits <- traits %>%
  mutate(STK11_no_KRAS = STK11 & !KRAS, 
         KEAP1_no_KRAS = KEAP1 & !KRAS,
         KRAS_only = KRAS & !STK11 & !KEAP1,
         KRAS_co   = KRAS & (STK11 | KEAP1),
         KRAS_new = ifelse(KRAS_only, "KRAS_only", ifelse(KRAS_co, "KRAS_co", "Other")))

modules <- as.data.frame(readRDS( "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/patient_imaging_meta_cluster_scores_tpm.rds"))

rna <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/rna_scores_tpm.rds")
rna$immucan_id  <- rownames(rna)
rownames(rna) <- NULL

modules <- modules  %>%
  mutate(immucan_id = rownames(modules))
rownames(modules) <- NULL

modules_names <- colnames(modules)[1:7]
rna_module_names <- colnames(rna)[1:29]
# variables
data_type <- unique(str_split_i(string = colnames(dat[,-1]),pattern = "__",i = 2))
```

# 4A Waterfall plot

The code for this plot is here: [here](../../WES/analysis/oncoplot.Rmd)

# 4B Mutations 

```{r}
# ============================================================
# TME vs. driver mutation status analysis - stratified by histology
#
# Assumes:
#   - One row per patient/tumor
#   - 8 TME measurement columns, already z-scored (continuous)
#   - 6 binary (0/1) mutation columns: EGFR, KRAS, STK11, KEAP1, PIK3CA, CDKN2A
#   - A histology column with values "LUAD" / "LUSC"
#
# WHY STRATIFY BY HISTOLOGY:
# Histology is a major confounder here, not just an imbalance issue. EGFR,
# KRAS, STK11, and KEAP1 are classic LUAD-enriched drivers and are expected
# to be rare/near-absent in LUSC. PIK3CA occurs only once in the LUAD
# samples in this cohort - with essentially zero variance there it cannot
# be tested in LUAD at all, but occurs at usable frequency in LUSC. Rather
# than adding histology as a simple covariate (which would not solve the
# "one mutant sample" problem for PIK3CA in LUAD, and would force one model
# structure onto two biologically different populations), the cleanest
# approach is to run separate gene panels per histology:
#   - LUAD: EGFR, KRAS, STK11, KEAP1, CDKN2A   (PIK3CA excluded - see above)
#   - LUSC: CDKN2A, PIK3CA                      (EGFR/KRAS/STK11/KEAP1 excluded
#                                                 - expected too rare in LUSC)
# CDKN2A is analysed in both strata since it's not histology-restricted here.
#
# Implements the tiered comparison strategy, run separately within each
# histology stratum where applicable:
#   Tier 1: EGFR vs KRAS vs driver-negative (mutually exclusive backbone) - LUAD only
#   Tier 2: within KRAS-mutant tumors, STK11/KEAP1 co-mutation stratification - LUAD only
#   Tier 3: multivariate model per TME outcome, all mutations in that stratum's
#           panel as covariates (LUAD panel includes KRAS:STK11 / KRAS:KEAP1
#           interaction terms; LUSC panel has no interaction terms specified)
#   Tier 4: simple univariate mut-vs-wt t-test per mutation, per TME outcome,
#           disregarding all other mutations' status. Compared side-by-side
#           against Tier 3 in a combined heatmap, per stratum.
#
# NOTE on CDKN2A: in NSCLC this is frequently inactivated by homozygous
# deletion rather than point mutation. If your CDKN2A column is derived only
# from small-variant VCF calls, it likely undercounts true CDKN2A loss unless
# you've merged in copy-number data.
# ============================================================

library(tidyverse)
library(broom)

# data
dat_z <- left_join(modules, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"),
         EGFR = ifelse(EGFR == TRUE, 1,0),
         KRAS = ifelse(KRAS == TRUE, 1,0),
         STK11 = ifelse(STK11 == TRUE, 1,0),
         KEAP1 = ifelse(KEAP1 == TRUE, 1,0),
         PIK3CA = ifelse(PIK3CA == TRUE, 1,0),
         CDKN2A = ifelse(CDKN2A == TRUE, 1, 0))

# ---- 1. Config: EDIT these to match your actual column names ----
tme_cols  <- modules_names
histology_col  <- "LUAD_vs_LUSC"                              # expects values "LUAD" / "LUSC"
mut_cols_luad  <- c("EGFR", "KRAS", "STK11", "KEAP1", "CDKN2A", "TP53")   # PIK3CA excluded: ~1 case in LUAD
mut_cols_lusc  <- c("CDKN2A", "PIK3CA", "TP53")                            # EGFR/KRAS/STK11/KEAP1 excluded in LUSC
min_group_n    <- 5   # minimum mutant (or wt) count required to test a gene in a stratum

df<- dat_z %>%
  dplyr::select(all_of(c(tme_cols,mut_cols_luad, mut_cols_lusc, histology_col))) %>%
  filter(!is.na(STK11)) # we will remove two patients that have missing genomic information

# sanity checks
stopifnot(all(tme_cols %in% names(df)))
stopifnot(histology_col %in% names(df))
all_mut_cols <- union(mut_cols_luad, mut_cols_lusc)
stopifnot(all(all_mut_cols %in% names(df)))
df[all_mut_cols] <- lapply(df[all_mut_cols], as.integer)  # ensure 0/1 numeric

cat("\n--- Histology counts ---\n")
print(table(df[[histology_col]]))

df_luad <- df %>% filter(.data[[histology_col]] == "LUAD")
df_lusc <- df %>% filter(.data[[histology_col]] == "LUSC")

cat(sprintf("\nLUAD n = %d, LUSC n = %d\n", nrow(df_luad), nrow(df_lusc)))

# ---- 2. Helper functions ----

# Drops a mutation from testing if it has fewer than min_group_n carriers
# (or fewer than min_group_n wt), rather than letting t.test()/lm() error out
# or silently return an unstable estimate.
check_group_size <- function(data, mut_var, min_n = min_group_n) {
  tab <- table(data[[mut_var]])
  if (length(tab) < 2 || min(tab) < min_n) {
    message(sprintf(
      "  -> dropping %s: insufficient variation/group size (counts: %s)",
      mut_var, paste(names(tab), tab, sep = "=", collapse = ", ")
    ))
    return(FALSE)
  }
  TRUE
}


run_tier4 <- function(data, tme_cols, mut_cols) {
  valid_muts <- mut_cols[sapply(mut_cols, function(m) check_group_size(data, m))]
  expand_grid(tme_var = tme_cols, mut_var = valid_muts) %>%
    pmap_dfr(function(tme_var, mut_var) {
      grp <- factor(data[[mut_var]], levels = c(0, 1), labels = c("wt", "mut"))
      tt  <- t.test(data[[tme_var]] ~ grp)
      diff_mut_minus_wt <- unname(tt$estimate["mean in group mut"] - tt$estimate["mean in group wt"])
      tibble(tme_var = tme_var, mut_var = mut_var, estimate = diff_mut_minus_wt, p_value = tt$p.value)
    }) %>%
    group_by(mut_var) %>%
    mutate(p_fdr = p.adjust(p_value, method = "BH")) %>%
    ungroup() %>%
    arrange(p_fdr)
}

# ---- 4. Tier 2 (LUAD only): within KRAS-mutant tumors, STK11/KEAP1 co-mutation ----
kras_df_luad <- df_luad %>%
  filter(KRAS == 1) %>%
  mutate(co_mut = if_else(STK11 == 1 | KEAP1 == 1,
                           "STK11_or_KEAP1_altered", "KRAS_only"))

cat("\n--- [LUAD] Tier 2 group sizes (within KRAS-mutant tumors) ---\n")
print(table(kras_df_luad$co_mut))

tier2_results_luad <- map_dfr(tme_cols, function(tme) {
  wt <- t.test(kras_df_luad[[tme]] ~ kras_df_luad$co_mut)
  tibble(tme_var = tme, ttest_p = wt$p.value)
})
tier2_results_luad$fdr <- p.adjust(tier2_results_luad$ttest_p, method = "BH")
cat("\n--- [LUAD] Tier 2: STK11/KEAP1 co-mutation effect within KRAS-mutant tumors ---\n")
print(tier2_results_luad)

# ---- 4b. Tier 2 visualization: KRAS-only vs STK11/KEAP1-altered, per TME score ----
# Panels ordered by FDR (strongest hits first, e.g. Cold/Neutrophils) so the
# STK11/KEAP1 "immune-cold" signature is easy to spot at a glance.
kras_long_luad <- kras_df_luad %>%
  dplyr::select(all_of(tme_cols), co_mut) %>%
  pivot_longer(cols = all_of(tme_cols), names_to = "tme_var", values_to = "z_score")

tme_order_tier2 <- tier2_results_luad %>% arrange(fdr) %>% pull(tme_var)
kras_long_luad$tme_var <- factor(kras_long_luad$tme_var, levels = tme_order_tier2)

tier2_sig_labels <- tier2_results_luad %>%
  mutate(
    tme_var = factor(tme_var, levels = tme_order_tier2),
    label = case_when(
      ttest_p < 0.001 & fdr < 0.1 ~ "***",
      ttest_p < 0.01 & fdr < 0.1 ~ "**",
      ttest_p < 0.05 & fdr < 0.1 ~ "*",
      TRUE        ~ ""
    )
  )

tier2_plot <- ggplot(kras_long_luad, aes(x = co_mut, y = z_score, fill = co_mut)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.5) +
  facet_wrap(~tme_var, nrow = 2) +
  geom_text(data = tier2_sig_labels, aes(x = 1.5, y = Inf, label = label),
            inherit.aes = FALSE, vjust = 1.5, size = 3.5) +
  scale_fill_manual(values = c("KRAS_only" = "grey70", "STK11_or_KEAP1_altered" = "firebrick")) +
  theme_minimal() +
  labs(
    title = "KRAS-mutant LUAD: STK11/KEAP1 co-mutation vs. KRAS-only",
    subtitle = "Labels show FDR stars where significant, else raw Wilcoxon p",
    x = NULL, y = "TME z-score", fill = "Group"
  ) +
  theme(legend.position = "bottom")

print(tier2_plot)
#ggsave("tme_kras_costk11keap1_boxplot.png", tier2_plot, width = 9, height = 6, dpi = 300)

# ---- 4c. Encode KRAS+STK11/KEAP1 co-mutation as its own predictor ----
# split KRAS-mutant tumors into two
# mutually exclusive, directly-plottable predictors:
#   - KRAS_only:          KRAS-mutant, no STK11/KEAP1 alteration
#   - KRAS_STK11KEAP1_co: KRAS-mutant WITH STK11 and/or KEAP1 alteration
# STK11 and KEAP1 are left as-is (they still capture any STK11/KEAP1
# mutation regardless of KRAS status).
#

df_luad <- df_luad %>%
  mutate(
    KRAS_only          = as.integer(KRAS == 1 & STK11 == 0 & KEAP1 == 0),
    KRAS_co = as.integer(KRAS == 1 & (STK11 == 1 | KEAP1 == 1))
  )

# Check whether STK11/KEAP1 mutations occur meaningfully outside the KRAS-
# mutant group. Tested this on synthetic data: if STK11/KEAP1 occur almost
# exclusively within KRAS-mutant tumors (as expected from the literature),
# keeping them as standalone predictors alongside KRAS_STK11KEAP1_co makes
# them strongly collinear with it (VIF ~6-10 in the test), which *dilutes*
# the co-mutation signal across three correlated terms instead of clarifying
# it - the opposite of what we want. In that case, drop the standalone
# STK11/KEAP1 terms and let the composite carry that signal on its own.
stk11_keap1_outside_kras <- sum((df_luad$STK11 == 1 | df_luad$KEAP1 == 1) & df_luad$KRAS == 0)
cat(sprintf("\n[LUAD] STK11/KEAP1-mutant tumors outside the KRAS-mutant group: %d\n",
            stk11_keap1_outside_kras))

if (stk11_keap1_outside_kras < min_group_n) {
  message("  -> STK11/KEAP1 occur almost exclusively within KRAS-mutant tumors here; ",
          "dropping standalone STK11/KEAP1 terms to avoid collinearity with KRAS_STK11KEAP1_co.")
  mut_cols_luad_v2 <- c("EGFR", "KRAS_only", "CDKN2A", "KRAS_STK11KEAP1_co")
} else {
  mut_cols_luad_v2 <- c("EGFR", "KRAS_only", "STK11", "KEAP1", "CDKN2A", "KRAS_co", "TP53")
}

# ---- 5. Tier 3 + Tier 4 (LUAD): expanded panel including the co-mutation predictor ----

tier4_luad <- run_tier4(df_luad, tme_cols, mut_cols_luad_v2)

pluad <- tier4_luad %>%
  mutate(
    label = case_when(
      p_value < 0.001 & p_fdr < 0.1 ~ "***",
      p_value < 0.01 & p_fdr < 0.1 ~ "**",
      p_value < 0.05 & p_fdr < 0.1 ~ "*",
      TRUE        ~ ""
    )
  ) %>%
 ggplot(aes(x = mut_var, y = tme_var, fill = estimate)) +
    geom_tile() +
    geom_text(aes(label =label), color = "black") +
    scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0) +
    #facet_wrap(~method) +
    theme_minimal() +
    labs(
      title = "Drivers LUAD",
      x = "Mutation", y = "Imaging module", fill = paste0("Δ z-score")
    )+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
pluad

# ============================================================
# LUSC stratum: CDKN2A, PIK3CA
# ============================================================

tier4_lusc <- run_tier4(df_lusc, tme_cols, mut_cols_lusc)

plusc <- tier4_lusc %>%
    mutate(
    label = case_when(
      p_value < 0.001 & p_fdr < 0.1 ~ "***",
      p_value < 0.01 & p_fdr < 0.1 ~ "**",
      p_value < 0.05 & p_fdr < 0.1 ~ "*",
      TRUE        ~ ""
    )
  ) %>%
 ggplot(aes(x = mut_var, y = tme_var, fill = estimate)) +
    geom_tile() +
    geom_text(aes(label = ifelse(p_fdr < 0.1, "*", "")), color = "black") +
    scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0) +
    #facet_wrap(~method) +
    theme_minimal() +
    labs(
      title = "Drivers LUSC",
      x = "Mutation", y = "Imaging module", fill = paste0("Δ z-score")
    )+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
plusc

p <- plot_grid(pluad, plusc,ncol= 1 )
p

ggsave(plot = p, path = out_folder,filename = "4B_Imaging_modules_drivers.pdf",width = 5, height = 8,device = cairo_pdf)
```



# Expression data preparation
```{r, fig.height=7, fig.width=15}
mean_exp1 <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/mean_marker_expr_per_celltype_p1.csv"))
mean_exp2 <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/mean_marker_expr_per_celltype_p2.csv"))

# long format
mean_exp1 <- mean_exp1 %>%
  pivot_longer(-immucan_id)
mean_exp2 <- mean_exp2 %>%
  pivot_longer(-immucan_id)

mean_exp1$tech <- "IMC1"
mean_exp2$tech <- "IMC2"
mean_exp <- rbind(mean_exp1,mean_exp2)

mean_exp <- mean_exp %>%
  mutate(celltypes = str_split_i(name,"-",1),
         name = str_split_i(name,"-",2))
# load sce object panel 1
sce1 <- readRDS(paste0(mount_path,"data/IMC/Panel_1/Rout/sce.rds"))
sce1 <- sce1[,sce1$ROI != 4]
# Panel 2
sce2 <- readRDS(paste0(mount_path,"data/IMC/Panel_2/Rout/sce.rds"))
```

# 4C HLADR on tumor cells in EGFR mutated

## Tumor cells HLADR in EGFR
```{r, fig.height=5, fig.width=5}
p_Tumor_HLADR_EGFR <- bplot_cor(ct = "Tumor", markers = "HLADR_p1",mutation = "EGFR",histology = "Adenocarcinoma",stat_size = 2)

p_Tumor_HLADR_EGFR

ggsave(filename = paste0(out_folder,"4C_HLADR_on_tumor_cells_EGFR.pdf"),
       plot = p_Tumor_HLADR_EGFR,
       device = "pdf",
       width = 1,
       height = 2)
```

## generate IMC images HLADR

```{r, fig.height=20, fig.width=20}
# get the expression matrix and caculate stats
counts <- as.data.frame(t(assay(sce1,"exprs")))
head(counts)
dim(counts)

counts$celltypes <- sce1$celltypes
counts$immucan_id <- sce1$immucan_id
counts$sample_id <- sce1$sample_id
counts <- left_join(counts, traits,"immucan_id")

# select samples from the frequency of positive cells
counts %>%
  mutate(HLADRpos = ifelse(HLADR > 3, TRUE,FALSE)) %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  group_by(immucan_id,EGFR,.drop = FALSE) %>%
  dplyr::count(HLADRpos) %>%
  mutate(total_cells = sum(n),
         frac = n/total_cells) %>%
  filter(HLADRpos == TRUE) %>%
  arrange(frac)

EGFRpos <- c("LUNG-NSCLC2-0624","LUNG-NSCLC2-0689","LUNG-NSCLC2-0659")

# load images
library(cytomapper)
masks <- readRDS(paste0(mount_path,"data/IMC/Panel_1/Rout/masks.rds"))
images <- readRDS(paste0(mount_path,"data/IMC/Panel_1/Rout/images.rds"))

cur_df <- as.data.frame(colData(sce1))
rownames(cur_df) <- NULL
cur_df <- cur_df %>%
  select(immucan_id,image, sample_id) %>%
  distinct() %>%
  mutate(image = gsub(".tiff","",image))
rownames(cur_df) <- cur_df$image


# select images
ims <- cur_df %>% 
  filter(immucan_id %in% EGFRpos,
         grepl("002$", sample_id)) %>%
  pull(sample_id)

# unify metadata of masks and images
mcols(images)$immucan_id <- cur_df[names(images),"immucan_id"]
mcols(masks)$immucan_id <- cur_df[names(masks),"immucan_id"]
mcols(images)$sample_id <- cur_df[names(images),"sample_id"]
mcols(masks)$sample_id <- cur_df[names(masks),"sample_id"]


cur_sce <- sce1[,sce1$sample_id %in% ims]
channelNames(images) <- rownames(cur_sce)

cur_ims <- getImages(images,ims)
cur_masks <- getImages(masks, ims)

# read images into memory and normalize
cur_images <- CytoImageList(cur_ims, on_disk = FALSE)
cur_masks <- CytoImageList(cur_masks, on_disk = FALSE)
cur_images <- scaleImages(cur_images,value = 2^16-1)
cur_images <- normalize(cur_images, separateImages = TRUE)
cur_images <- normalize(cur_images, separateImages = TRUE, 
                        inputRange = c(0, 0.2))

# define all tumor and non-tumor cells
cur_sce$tumor <- "non-tumor"
cur_sce[,cur_sce$celltypes == "Tumor"]$tumor <- "tumor"
plot_sce <- cur_sce[,cur_sce$celltypes == "Tumor"]

# generate images
plotPixels(image = cur_images,
           object = plot_sce,
           cell_id = "ObjectNumber",
           img_id = "sample_id",
           colour_by = c("Ecad", "HLADR", "DNA1"),
           colour = list(Ecad = c("black", "green"),HLADR = c("black","red"),DNA1 = c("black","blue")),
           bcg = list(Ecad  = c(0,2,1),
                      HLADR = c(0,1.5,1),
                      DNA1  = c(0,1,1)),
            image_title = NULL,save_plot = list(filename = paste0(out_folder,"HLADR_tumor_2.png")))
```


# 4D
single cell analysis script

# 4E interactions of Tumor cells 
## proportions - not shown in paper
very similar results to "classic" interaction counting
```{r, fig.height=6, fig.width=10}
library(data.table)
cur_label <- factor(colData(sce1)[["celltypes_minor"]])
cur_table <- imcRtools:::.prepare_table(sce1, "image", cur_label, "neighborhood")
  
# here we define the function for proportional interaction counts manually as it is defined in later version of imcRtools  
int_quant <- function(dat_table, object, group_by, label,
                                             check_missing = TRUE) {
  
  . <- ct <- .N <- N <- n_tot_int <- NULL
  
  dat_temp <- dat_table[, .N, by = .(group_by, from_label, to_label)
  ][, n_tot_int := sum(N), by = .(group_by, from_label)
  ][, ct := N / n_tot_int]
  
  if (check_missing) {
    dat_temp <- dat_temp[CJ(group_by = unique(dat_table$group_by),
                            from_label = as.factor(levels(dat_table$from_label)),
                            to_label = as.factor(levels(dat_table$to_label))),
                         on = c("group_by", "from_label", "to_label")]
    ct <- from_label <- to_label <- NULL
    dat_temp[is.na(dat_temp$ct), ct := 0]
    
    # Set all cells that are not contained in specific groups to NA
    cur_dat <- unclass(table(colData(object)[[group_by]],
                             colData(object)[[label]]))
    cur_ind <- which(cur_dat == 0, arr.ind = TRUE)
    
    if (nrow(cur_ind) > 0) {
      apply(cur_ind, 1 , function(x){
        dat_temp[group_by == rownames(cur_dat)[x[1]] &
                   (from_label == colnames(cur_dat)[x[2]] |
                      to_label == colnames(cur_dat)[x[2]]),
                 ct := NA]
      })
    }
  }
  
  dat_temp <- dat_temp[, .(group_by, from_label, to_label, ct)]
  return(dat_temp)
}

cur_count <- int_quant(dat_table = cur_table,object = sce1,group_by = "image",label = "celltypes_minor")
cur_count <- setorder(cur_count, "group_by", "from_label", "to_label")

cur_count <- cur_count %>%
  rename(group_by = "image")
cur_dat <- as.data.frame(colData(sce1))

p <- cur_dat %>%
  select(immucan_id, image) %>%
  distinct() %>%
  left_join(cur_count, "image") %>%
  left_join(traits, "immucan_id") %>%
  filter(grepl("Tumor", from_label),
         grepl("CD4|CD8|Treg", to_label), 
         !is.na(EGFR)) %>%
  group_by( from_label, to_label) %>%
  summarise(Avg_int = median(ct,na.rm = TRUE )) %>%
  filter(!grepl("BnT",to_label)) %>%
  ggplot(aes(x= to_label, y = from_label,fill = Avg_int))+
  geom_tile()+
  #facet_wrap(.~EGFR)+
  scale_fill_viridis()+
  theme(axis.text.x = element_text(angle=90, vjust = .5, hjust = 1))

p <- cur_dat %>%
  select(immucan_id, image) %>%
  distinct() %>%
  left_join(cur_count, "image") %>%
  left_join(traits, "immucan_id") %>%
  filter(grepl("Tumor", from_label),
         grepl("CD4|CD8|Treg", to_label), 
         !is.na(EGFR)) %>%
  group_by( EGFR, from_label, to_label) %>%
  summarise(Avg_int = median(ct,na.rm = TRUE )) %>%
  ggplot(aes(x= to_label, y = from_label,fill = Avg_int))+
  geom_tile()+
  facet_wrap(.~EGFR)+
  scale_fill_viridis()+
  theme(axis.text.x = element_text(angle=90, vjust = .5, hjust = 1))

p

```



## classic interactions
```{r}
minor_interactions <- IMC1_int <- read.csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/cell-pair_interactions_celltypes_minor_directional_P1.csv")

p <- minor_interactions %>%
  pivot_longer(-immucan_id) %>%
  filter(grepl("^Tumor", name),
         grepl("CD4|CD8|Treg", name)) %>%
  mutate(from_label = str_split_i(name,"\\.",1),
         to_label = str_split_i(name,"\\.",2)) %>%
  left_join(traits, "immucan_id") %>%
  filter(!is.na(EGFR)) %>%
    group_by(from_label, to_label) %>%
  summarise(Avg_int = median(value,na.rm = TRUE )) %>%
  ggplot(aes(x= to_label, y = from_label,fill = Avg_int))+
  geom_tile()+
  scale_fill_viridis()+
  theme(axis.text.x = element_text(angle=90, vjust = .5, hjust = 1))
  
p <- p+ggtitle("Normalized interactions of all tumor cell clusters")
p
ggsave(filename = paste0(out_folder,"4E_interactions_of_tumor_cells.pdf"),
       plot = p,
       device = "pdf",
       width = 5.5,
       height = 3.5)
```

# 4F interaction quntification
```{r}
int_df <- int_dat %>%
  select(immucan_id,
         Tumor.CD4__IF3_int, Tumor.CD8__IF3_int, Tumor.Treg__IF3_int,
         Tumor.CD4__IMC1_int, Tumor.CD8__IMC1_int, Tumor.Treg__IMC1_int) %>%
  pivot_longer(-immucan_id) %>%
  mutate(name = gsub("_int", "", name)) %>%
  left_join(traits, "immucan_id") %>%
  filter(!is.na(EGFR))

stopifnot(nrow(int_df) > 0)
if (dplyr::n_distinct(int_df$EGFR) < 2)
  stop("EGFR has fewer than 2 levels after filtering.")

fdr_cut <- 0.1
digits  <- 3

stat.test <- int_df %>%
  group_by(name) %>%
  wilcox_test(value ~ EGFR) %>%
  ungroup() %>%                        # BH correction spans all 6 panels
  adjust_pvalue(method = "BH") %>%
  add_significance(p.col = "p.adj")

# rounded label, with a floor so tiny p-values don't print as "0.00"
floor_p <- 10^(-digits)
stat.test <- stat.test %>%
  mutate(p.label = ifelse(
    p < floor_p,
    sprintf("p < %s", format(floor_p, scientific = FALSE)),
    sprintf("p = %.*f", digits, p)
  ))

# only the comparisons that pass FDR get plotted
stat.sig <- stat.test %>%
  filter(!is.na(p.adj), p.adj < fdr_cut) %>%
  mutate(y.position = 0.75)            # fixed scale, so one height for all panels

p <- ggplot(int_df, aes(x = EGFR, y = value, fill = EGFR)) +
  geom_boxplot(outliers = FALSE) +
  geom_jitter(aes(color = EGFR),
              position = position_jitterdodge(jitter.width = 0.2), size = 1.5) +
  scale_fill_manual(values = alpha(c("red", "blue"), alpha = 0.6)) +
  scale_color_manual(values = c("red", "blue")) +
  facet_wrap(. ~ name, nrow = 1) +
  coord_cartesian(ylim = c(0, .8)) +
  theme_bw() +
  theme(legend.position = "none",
        axis.text.x  = element_text(size = 12, angle = 90, vjust = .5, hjust = 1),
        axis.text.y  = element_text(size = 12),
        axis.title.x = element_blank()) +
  ylab("Avg interactions")

if (nrow(stat.sig) > 0) {
  p <- p + stat_pvalue_manual(stat.sig, label = "{p.label}", size = 5, tip.length = 0.01)
} else {
  message("No comparison passed FDR < ", fdr_cut, " — plotting without labels.")
}

attr(p, "stat.test") <- stat.test   # full table for a supplementary table
p


ggsave(filename = paste0(out_folder,"4F_CD4_interactions_quantification.pdf"),
       plot = p,
       device = "pdf",
       width = 8,
       height = 3)
```


# expression boxplots

# 4G Ki67 and p53 on tumor cells in TP53 mutated in LUAD

```{r, fig.height=5, fig.width=5}
p_Tumor_Ki67_P53_TP53 <- bplot_cor(ct = "Tumor", c("p53_p2","Ki67_p1"), mutation = "TP53", 1, histology = c("Adenocarcinoma", "Squamous cell carcinoma"),stat_size = 2,headroom = 0.1)

p_Tumor_Ki67_P53_TP53

ggsave(filename = paste0(out_folder,"4G_Ki67_p53_on_tumor_cells_TP53.pdf"),
       plot = p_Tumor_Ki67_P53_TP53,
       device = "pdf",
       width = 1.8,
       height = 2)
```



# 4H P53 images
```{r, fig.height=20, fig.width=20}
counts2 <- as.data.frame(t(assay(sce2,"exprs")))
head(counts2)
dim(counts2)

counts2$celltypes <- sce2$celltypes
counts2$immucan_id <- sce2$immucan_id
counts2$sample_id <- sce2$sample_id
counts2 %>%
  filter(celltypes == "Tumor") %>%
  ggplot(aes(x = Ecad,y = p53))+
  geom_point()

counts2 <- left_join(counts2, traits,"immucan_id")

counts2 %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  mutate(p53pos = ifelse(p53 > 1, TRUE,FALSE)) %>%
  group_by(sample_id,.drop = FALSE) %>%
  dplyr::count(p53pos) %>%
  mutate(total_cells = sum(n),
         frac = n/total_cells) %>%
  filter(p53pos == TRUE) %>%
  arrange(frac)

p53pos <- c("10040769","10040838")
p53pos <- traits$immucan_id[grepl(paste(p53pos,collapse = "|"),traits$ibbl_kit_ids)]

# load images
library(cytomapper)
masks <- readRDS(paste0(mount_path,"data/IMC/Panel_2/Rout/masks.rds"))
images <- readRDS(paste0(mount_path,"data/IMC/Panel_2/Rout/images.rds"))

cur_df <- as.data.frame(colData(sce2))
rownames(cur_df) <- NULL
cur_df <- cur_df %>%
  select(immucan_id,image, sample_id) %>%
  distinct() %>%
  mutate(image = gsub(".tiff","",image))
rownames(cur_df) <- cur_df$image


# select images
ims <- cur_df %>% 
  filter(immucan_id %in% p53pos,
         grepl("002$", sample_id)) %>%
  pull(sample_id)

# unify metadata of masks and images
mcols(images)$immucan_id <- cur_df[names(images),"immucan_id"]
mcols(masks)$immucan_id <- cur_df[names(masks),"immucan_id"]
mcols(images)$sample_id <- cur_df[names(images),"sample_id"]
mcols(masks)$sample_id <- cur_df[names(masks),"sample_id"]


cur_sce <- sce2[,sce2$sample_id %in% ims]


cur_ims <- getImages(images,ims)
cur_masks <- getImages(masks, ims)

# read images into memory and normalize
cur_images <- CytoImageList(cur_ims, on_disk = FALSE)
cur_masks <- CytoImageList(cur_masks, on_disk = FALSE)
cur_images <- scaleImages(cur_images,value = 2^16-1)
cur_images <- normalize(cur_images, separateImages = TRUE)
cur_images <- normalize(cur_images, separateImages = TRUE, 
                        inputRange = c(0, 0.2))

channelNames(cur_images) <- rownames(cur_sce)

# define all tumor and non-tumor cells
cur_sce$tumor <- "non-tumor"
cur_sce[,cur_sce$celltypes == "Tumor"]$tumor <- "tumor"
plot_sce <- cur_sce[,cur_sce$celltypes == "Tumor"]

# generate images
plotPixels(image = cur_images,
           object = plot_sce,
           cell_id = "ObjectNumber",
           img_id = "sample_id",
           colour_by = c("Ecad", "p53", "DNA1"),
           colour = list(Ecad = c("black", "green"),p53 = c("black","red"),DNA1 = c("black","blue")),
           bcg = list(Ecad  = c(0,2,1),
                      p53 = c(0,1.5,1),
                      DNA1  = c(0,1,1)),
            image_title = NULL,save_plot = list(filename = paste0(out_folder,"p53_tumor.png")))
```

```{r}
counts2 %>%
  filter(celltypes == "Tumor") %>%
  ggplot(aes(x = Ecad,y = p53))+
  geom_point(size = 0.1)+
  facet_wrap(LUAD_vs_LUSC~TP53)+
  theme_bw()

hist(counts2$p53)

counts2 %>%
  mutate(p53pos = ifelse(p53 > 2, TRUE,FALSE)) %>%
  group_by(immucan_id,TP53,.drop = FALSE) %>%
  dplyr::count(p53pos) %>%
  left_join(traits[,c("immucan_id","LUAD_vs_LUSC")],"immucan_id") %>%
  mutate(total_cells = sum(n),
         frac = n/total_cells) %>%
  filter(p53pos == TRUE) %>%
  ggplot(aes(x = TP53,y = log2(frac)))+
  geom_boxplot()+
  facet_wrap(.~LUAD_vs_LUSC, scales = "free")+
  theme_bw()+
  stat_compare_means(method = "wilcox.test")
```
# 4I KRAS

there seem to be very little differences between individual cell type expression between KRAS_co mutation and KRAS_only mutations on the immune level. on the tumor level there are differences
```{r, fig.height=5, fig.width=10}
p <- bplot_cor(ct = "Tumor", markers = c("EGFR_p2","HLADR_p1", "CD73_p2", "p53_p2","MMP9_p2", "CD56_p2", "CD155_p2", "CXCL8_p2", "Arg1_p2", "Ki67_p1"),mutation = "KRAS_new",sig_format = FALSE,stat_size = 2)

p
ggsave(filename = paste0(out_folder,"4I_KRAS_comutations.pdf"),
       plot = p+xlab(element_blank()),
       device = "pdf",
       width = 8,
       height = 2.75)

```


# 4J Treg cells GITR PIK3CA
```{r, fig.height=5, fig.width=5}
p_Treg_GITR_PIK3CA <- bplot_cor(ct = "Treg", c("GITR_p2","CD137_p2", "GATA3_p2"), mutation = "PIK3CA", rownum = 1, histology = c("Squamous cell carcinoma"),stat_size = 2)

p_Treg_GITR_PIK3CA
ggsave(filename = paste0(out_folder,"4J_activation_markers_on_Tregs.pdf"),
       plot = p_Treg_GITR_PIK3CA,
       device = "pdf",
       width = 2.5,
       height = 2)
```
