---
title: "Figure S6"
author: "Daniel Schulz"
date: "2025-03-10"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(bitmapType = "cairo")
```

```{r}
library(tidyverse)
library(ggpubr)
library(SingleCellExperiment)
library(scales)
library(ggnewscale)
library(edgeR)
library(cowplot)
library(igraph)
library(ggraph)
library(ggrepel)
library(survminer)
library(survivalAnalysis)
library(survival)
library(rstatix)

source("../code/bplot_functions.R")
```


```{r}
mount_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/"
out_folder <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/revised_figures/Figure_6_supp/"

if (!dir.exists(out_folder)) {
  dir.create(out_folder)
}
```


# load data
we will load the data we generated at the beginning of figure 2
```{r}
dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/abundances_dat.rds"))
dens_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/densities_dat.rds"))
cello_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/cello_dat.rds"))
clin_dat <- read.csv("/mnt/immucan_volume/clinical_data/Lung_NSCLC2/2026_final/clinical.csv")
int_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/interaction_dat.rds"))
traits <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/all_traits.csv"))

rna <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/rna_scores_tpm.rds")
rna$immucan_id  <- rownames(rna)
rownames(rna) <- NULL
rna_module_names <- colnames(rna)[!grepl("immucan_id", colnames(rna))]
# variables
data_type <- unique(str_split_i(string = colnames(dat[,-1]),pattern = "__",i = 2))

# colors
hist_col_2 = setNames(c("blue","orange"),c("LUAD","LUSC"))
hist_col_3 = setNames(c("blue","orange", "gray40"),c("LUAD","LUSC","Other"))
```

# hazards for oncogenes
not in paper
```{r}
p.list <- list()
for (i in c("TP53","KRAS", "EGFR","STK11", "PIK3CA")) {
  for(j in c("Adenocarcinoma", "Squamous cell carcinoma")) {
    cur_dat <- as.data.frame(traits)
    #cur_dat[,i] <- as.factor(cur_dat[,i])
    cur_dat <- cur_dat %>%
      filter(!is.na(cur_dat[,i]),
             simple_histology == j)
    surv_ob <- Surv(cur_dat$time_to_event,cur_dat$censoring)
    p.list[[paste0(j, "_",i)]] <- coxph( as.formula(paste0("surv_ob ~ simple_stage + ",i)),
                          data = cur_dat)
  }
}

df <- data.frame(HR =      unlist(lapply(p.list, function(x){summary(x)$coefficients[,"exp(coef)"][4]})),
                 p.value = unlist(lapply(p.list, function(x){summary(x)$coefficients[, "Pr(>|z|)"][4]})),
                 lower_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "lower .95"][4]})),
                 upper_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "upper .95"][4]}))) 

df1 <- df %>% 
  mutate(hist = str_split_i(rownames(df),"_",1),
         cond = str_split_i(rownames(df),"\\.",2),
         cond = gsub("TRUE","mut", cond)) %>%
  filter(!is.na(HR),
         !(hist=="Squamous cell carcinoma" & cond == "EGFRmut"),
         !(hist=="Adenocarcinoma" & cond == "PIK3CAmut"),
         !(hist=="Adenocarcinoma" & cond == "EGFRmut"))

p1 <- df1 %>%
  mutate(cond = paste0(hist,"_",cond)) %>%
  ggplot(aes(x = HR,y = cond, col = hist))+
  geom_point(size = 2)+
  geom_errorbar(aes(xmin=lower_conf, xmax = upper_conf),lwd =1,width = 0.3)+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust= 1))+
  scale_y_discrete(labels = function(x) str_split_i(x,"_",2))+
  scale_x_log10()+
  theme_bw()+
  geom_vline(xintercept = 1,linetype = "dashed")+
  geom_text(aes(label = paste0("p : ",round(p.value,digits = 3))),
            col = "black",
            vjust = -1,
            inherit.aes = TRUE)+
  scale_color_manual(values = c("Adenocarcinoma" = "blue", "Squamous cell carcinoma" = "orange"))

# Kaplan Meier plot for EGFR since no patients die
COSMIC_EGFR<- c("LUNG-NSCLC2-0564", "LUNG-NSCLC2-0564", "LUNG-NSCLC2-0593", "LUNG-NSCLC2-0600", "LUNG-NSCLC2-0624" ,"LUNG-NSCLC2-0633" ,"LUNG-NSCLC2-0641" ,"LUNG-NSCLC2-0659" ,"LUNG-NSCLC2-0659","LUNG-NSCLC2-0689" ,"LUNG-NSCLC2-0690", "LUNG-NSCLC2-0704", "LUNG-NSCLC2-0712", "LUNG-NSCLC2-0736" ,"LUNG-NSCLC2-0743")

cur_traits <- traits %>%
  mutate(EGFR = ifelse(immucan_id %in% COSMIC_EGFR, TRUE, FALSE))
surv_fit <- survfit(Surv(time_to_event, censoring) ~ EGFR, data = cur_traits)

p2 <- ggsurvplot(surv_fit, data = traits %>% filter(simple_histology == "Adenocarcinoma"), risk.table = TRUE, pval = TRUE,palette = c("red","blue"))
p2 <- p2$plot+
  ylab("OS probability")+
  xlab("Time (days)")

p2 <- plot_grid(p2,ncol = 1, rel_heights = c(1,0.5))

p <- plot_grid(p1,p2,ncol = 1,rel_heights = c(1,1))
p

```


# S6A driver mutation enrichment in histology

```{r, fig.height=8}
cur_traits <- traits %>% select("LUAD_vs_LUSC")
condition <- c("Adenocarcinoma")


traits <- traits %>%
  mutate(KRAS_only = KRAS & !STK11 & !KEAP1,
         KRAS_co   = KRAS & (STK11 | KEAP1))
names(condition) <- colnames(cur_traits)
plots_list <- list()
for(j in c("TP53","KRAS_co", "KRAS_only","EGFR","STK11","PIK3CA","NF1","KEAP1", "CDKN2A")){
  p.list <- list()
  l.list <- list()
  for (i in names(condition)) {
    x <- factor(traits[[j]])
    x <- relevel(x, ref = "FALSE")
    
    y <- factor(unlist(traits[,i]))
    y <- relevel(y, ref = condition[i])
    
    p.list[[i]] <- fisher.test(table(x, y))
    l.list[[i]] <- c(levels(y)[1])
  }
  
  out <- data.frame(feature = names(p.list),
                    p.value = unlist(lapply(p.list, function(x){
                      x$p.value})),
                    odds_ratio = unlist(lapply(p.list, function(x){
                      x$estimate})),
                    reference = unlist(l.list))
  
  
  # generate annotation vector:
  anno_vec <- c("LUSC_vs_LUAD")
  out$annotation = anno_vec
  out[,"mutation"] <- paste0(j,"_mut")
  plots_list[[j]]<- out 
}

cur_dat <- do.call(rbind,plots_list)

cur_dat <- cur_dat %>%
    mutate(odds_ratio = ifelse(is.infinite(odds_ratio),1024,odds_ratio),
           odds_ratio = ifelse(odds_ratio == 0,2^-10,odds_ratio),
           FDR = p.adjust(p.value,method = "BH"),
           sig = ifelse(p.value < 0.05 & odds_ratio > 1, "UP_trend",NA),
           sig = ifelse(FDR < 0.1 & odds_ratio > 1, "UP",sig),
           sig = ifelse(p.value < 0.05 & odds_ratio < 1, "DOWN_trend",sig),
           sig = ifelse(FDR < 0.1 & odds_ratio < 1, "DOWN",sig),
           sig = ifelse(is.na(sig),"NOTSIG", sig),
           p.value = ifelse(is.infinite(odds_ratio),NA,p.value),
           odds_ratio = ifelse(odds_ratio == 0,1/max(odds_ratio),odds_ratio),
           log_odds = log2(odds_ratio))

min_down <- min(cur_dat$log_odds-1)
max_up <- max(cur_dat$log_odds)

p1 <- cur_dat %>%
  ggplot(aes(x=mutation, y = annotation,col = log_odds, size= -log10(p.value)))+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(round(min_down),0,round(max_up+1))),
                         guide = "colorbar", limits=c(round(min_down),round(max_up+1)),name = "log2 odds ratio")+
  scale_size(range = c(1,6))+
  new_scale_colour() +
  geom_point(data =cur_dat, shape = 1,aes(x = mutation,size = -log10(p.value), col = sig), stroke =1.75)+
  scale_color_manual(values = c("UP" = "#9f1214","UP_trend" = "#e73032","DOWN" = "#25567d","DOWN_trend" = "#6a9ec7","NOTSIG" = "snow2"), name = "stats")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 10),
        axis.text.y = element_text(size = 10),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10),
        plot.margin = unit(c(0.5, 0,0,0), "cm"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.position = "none")
p1

ggsave(filename = paste0(out_folder,"S6A_Driver_enrichment_in_histology.pdf"),
       plot = p1,
       device = "pdf",
       width = 3,
       height = 1.5)
```

# S6A enrichment of clinical in oncogenes

## LUAD
```{r, fig.height=8}
cur_traits <- traits %>% select("TPS_score", "TMB_score","age_dicho", "gender","stage_1_2","stage_1_3","stage_2_3","never_vs_current_smoker",
            "never_vs_Ex_smoker","current_vs_Ex_smoker", "N0_vs_Nx","M0_vs_Mx")
condition <- c("negative", "low","low", "Female","I","I","II","Non-smoker","Non-smoker","Ex-smoker", "node_neg", "met_neg")

names(condition) <- colnames(cur_traits)
plots_list <- list()
cur_traits <- traits %>% filter(simple_histology == "Adenocarcinoma")
for(j in c("TP53","KRAS_co", "KRAS_only", "KRAS", "KEAP1","EGFR","STK11")){
  p.list <- list()
  l.list <- list()
  for (i in names(condition)) {
    x <- factor(cur_traits[[j]])
    x <- relevel(x, ref = "FALSE")
    
    y <- factor(unlist(cur_traits[,i]))
    y <- relevel(y, ref = condition[i])
    
    p.list[[i]] <- fisher.test(table(x, y))
    l.list[[i]] <- c(levels(y)[1])
  }
  
  out <- data.frame(feature = names(p.list),
                    p.value = unlist(lapply(p.list, function(x){
                      x$p.value})),
                    odds_ratio = unlist(lapply(p.list, function(x){
                      x$estimate})),
                    reference = unlist(l.list))
  
  
  # generate annotation vector:
  anno_vec <- c("TPSpos_vs_TPSneg","TMBhigh_vs_TMBlow","HighAge_vs_lowAge","Male_vs_Female","StageII_vs_I","StageIII_vs_I","StageIII_vs_II","Smoker_vs_neverSmoker","formerSmoker_vs_neverSmoker","Smoker_vs_formerSmoker","NodePos_vs_NodeNeg","MetPos_Vs_MetNeg")
  out$annotation = anno_vec
  out[,"mutation"] <- paste0(j,"_mut")
  plots_list[[j]]<- out 
}

cur_dat <- do.call(rbind,plots_list)

cur_dat_LUAD <- cur_dat %>%
    mutate(FDR = p.adjust(p.value,method = "BH"),
           sig = ifelse(p.value < 0.05 & odds_ratio > 1, "UP_trend",NA),
           sig = ifelse(FDR < 0.1 & odds_ratio > 1, "UP",sig),
           sig = ifelse(p.value < 0.05 & odds_ratio < 1, "DOWN_trend",sig),
           sig = ifelse(FDR < 0.1 & odds_ratio < 1, "DOWN",sig),
           sig = ifelse(is.na(sig),"NOTSIG", sig),
           p.value = ifelse(is.infinite(odds_ratio),NA,p.value),
           odds_ratio = ifelse(odds_ratio == 0,1/max(odds_ratio),odds_ratio),
           log_odds = log2(odds_ratio))


p <- cur_dat_LUAD %>%
  ggplot(aes(x=mutation, y = annotation,col = log_odds, size= -log10(p.value)))+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(round(min_down),0,round(max_up+1))),
                         guide = "colorbar", limits=c(round(min_down),round(max_up+1)))+
  scale_size(range = c(1,6))+
  new_scale_colour() +
  geom_point(shape = 1,aes(size = -log10(p.value), col = sig), stroke =1.75)+
  scale_color_manual(values = c("UP" = "#9f1214","UP_trend" = "#e73032","DOWN" = "#25567d","DOWN_trend" = "#6a9ec7","NOTSIG" = "snow2"), name = "stats")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 10),
        axis.text.y = element_text(size = 10),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10),
        plot.margin = unit(c(0.5, 0,0,0), "cm"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank())
p

```

## LUSC
```{r, fig.height=8}
cur_traits <- traits %>% select("TPS_score", "TMB_score","age_dicho", "gender","stage_1_2","stage_1_3","stage_2_3","current_vs_Ex_smoker", "N0_vs_Nx","M0_vs_Mx")
condition <- c("negative", "low","low", "Female","I","I","II","Ex-smoker", "node_neg", "met_neg")

names(condition) <- colnames(cur_traits)
plots_list <- list()
cur_traits <- traits %>% filter(simple_histology == "Squamous cell carcinoma")
for(j in c("TP53","PIK3CA", "CDKN2A")){
  p.list <- list()
  l.list <- list()
  for (i in names(condition)) {
    x <- factor(cur_traits[[j]])
    x <- relevel(x, ref = "FALSE")
    
    y <- factor(unlist(cur_traits[,i]))
    y <- relevel(y, ref = condition[i])
    
    p.list[[i]] <- fisher.test(table(x, y))
    l.list[[i]] <- c(levels(y)[1])
  }
  
  out <- data.frame(feature = names(p.list),
                    p.value = unlist(lapply(p.list, function(x){
                      x$p.value})),
                    odds_ratio = unlist(lapply(p.list, function(x){
                      x$estimate})),
                    reference = unlist(l.list))
  
  
  # generate annotation vector:
  anno_vec <- c("TPSpos_vs_TPSneg","TMBhigh_vs_TMBlow","HighAge_vs_lowAge","Male_vs_Female","StageII_vs_I","StageIII_vs_I","StageIII_vs_II","Smoker_vs_formerSmoker","NodePos_vs_NodeNeg","MetPos_Vs_MetNeg")
  out$annotation = anno_vec
  out[,"mutation"] <- paste0(j,"_mut")
  plots_list[[j]]<- out
}

cur_dat <- do.call(rbind,plots_list)


cur_dat_LUSC <-  cur_dat %>%
    mutate(FDR = p.adjust(p.value,method = "BH"),
           sig = ifelse(p.value < 0.05 & odds_ratio > 1, "UP_trend",NA),
           sig = ifelse(FDR < 0.1 & odds_ratio > 1, "UP",sig),
           sig = ifelse(p.value < 0.05 & odds_ratio < 1, "DOWN_trend",sig),
           sig = ifelse(FDR < 0.1 & odds_ratio < 1, "DOWN",sig),
           sig = ifelse(is.na(sig),"NOTSIG", sig),
           p.value = ifelse(is.infinite(odds_ratio),NA,p.value),
           odds_ratio = ifelse(odds_ratio == 0,1/max(odds_ratio),odds_ratio),
           odds_ratio = ifelse(is.infinite(odds_ratio),NA,odds_ratio),
           log_odds = log2(odds_ratio))

p <- cur_dat_LUSC %>%
  ggplot(aes(x=mutation, y = annotation,col = log_odds, size= -log10(p.value)))+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(round(min_down),0,round(max_up+1))),
                         guide = "colorbar", limits=c(round(min_down),round(max_up+1)))+
  scale_size(range = c(1,6))+
  new_scale_colour() +
  geom_point(shape = 1,aes(size = -log10(p.value), col = sig), stroke =1.75)+
  scale_color_manual(values = c("UP" = "#9f1214","UP_trend" = "#e73032","DOWN" = "#25567d","DOWN_trend" = "#6a9ec7","NOTSIG" = "snow2"), name = "stats")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 10),
        axis.text.y = element_text(size = 10),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10),
        plot.margin = unit(c(0.5, 0,0,0), "cm"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank())
p

```


## plot LUSC and LUAD together

```{r}
cur_dat_LUAD$histology <- "LUAD"
cur_dat_LUSC$histology <- "LUSC"

cur_dat <- rbind(cur_dat_LUAD,cur_dat_LUSC)

cur_dat <- cur_dat %>%
  mutate(FDR = p.adjust(p.value,method = "BH"),
         sig = ifelse(p.value < 0.05 & odds_ratio > 1, "UP_trend",NA),
         sig = ifelse(FDR < 0.1 & odds_ratio > 1, "UP",sig),
         sig = ifelse(p.value < 0.05 & odds_ratio < 1, "DOWN_trend",sig),
         sig = ifelse(FDR < 0.1 & odds_ratio < 1, "DOWN",sig),
         sig = ifelse(is.na(sig),"NOTSIG", sig),
         p.value = ifelse(is.infinite(odds_ratio),NA,p.value),
         odds_ratio = ifelse(odds_ratio == 0,1/max(odds_ratio),odds_ratio),
         log_odds = log2(odds_ratio),
         mutation = paste0(histology,"_",mutation))
p <- cur_dat %>%
  ggplot(aes(x=mutation, y = annotation,col = log_odds, size= -log10(p.value)))+
  geom_point()+
  theme_bw()+
  scale_colour_gradientn(colours = c("blue","white","red"), 
                         values = rescale(c(round(min_down),0,round(max_up+1))),
                         guide = "colorbar", limits=c(round(min_down),round(max_up+1)))+
  scale_size(range = c(1,6))+
  new_scale_colour() +
  geom_point(data = cur_dat ,shape = 1,aes(size = -log10(p.value), col = sig), stroke =1.75)+
  scale_color_manual(values = c("UP" = "#9f1214","UP_trend" = "#e73032","DOWN" = "#25567d","DOWN_trend" = "#6a9ec7","NOTSIG" = "snow2"), name = "stats")+
  theme(axis.text.x = element_text(angle = 90, vjust=0.5, hjust=1,size = 10),
        axis.text.y = element_text(size = 10),
        legend.text = element_text(size = 10),
        legend.title = element_text(size = 10),
        plot.margin = unit(c(0.5, 0,0,0), "cm"),
        axis.title.x = element_blank(),
        axis.title.y = element_blank(),
        legend.location = "plot")

p
p2 <- plot_grid(p1,p,ncol = 1, rel_heights = c(1.2,4))
p2
ggsave(filename = paste0(out_folder,"S6A_driver_mutation_enrichment_in_clinical.pdf"),
       plot = p2,
       device = "pdf",
       width = 6,
       height = 7)
```


# S6B Mutations RNA

```{r, fig.height=6, fig.width=12}
# ============================================================
# TME vs. driver mutation status analysis - stratified by histology
#
# Assumes:
#   - One row per patient/tumor
#   - 8 TME measurement columns, already z-scored (continuous)
#   - 6 binary (0/1) mutation columns: EGFR, KRAS, STK11, KEAP1, PIK3CA, CDKN2A
#   - A histology column with values "LUAD" / "LUSC"
#
# WHY STRATIFY BY HISTOLOGY:
# Histology is a major confounder here, not just an imbalance issue. EGFR,
# KRAS, STK11, and KEAP1 are classic LUAD-enriched drivers and are expected
# to be rare/near-absent in LUSC. PIK3CA occurs only once in the LUAD
# samples in this cohort - with essentially zero variance there it cannot
# be tested in LUAD at all, but occurs at usable frequency in LUSC. Rather
# than adding histology as a simple covariate (which would not solve the
# "one mutant sample" problem for PIK3CA in LUAD, and would force one model
# structure onto two biologically different populations), the cleanest
# approach is to run separate gene panels per histology:
#   - LUAD: EGFR, KRAS, STK11, KEAP1, CDKN2A   (PIK3CA excluded - see above)
#   - LUSC: CDKN2A, PIK3CA                      (EGFR/KRAS/STK11/KEAP1 excluded
#                                                 - expected too rare in LUSC)
# CDKN2A is analysed in both strata since it's not histology-restricted here.
#
# Implements the tiered comparison strategy, run separately within each
# histology stratum where applicable:
#   Tier 1: EGFR vs KRAS vs driver-negative (mutually exclusive backbone) - LUAD only
#   Tier 2: within KRAS-mutant tumors, STK11/KEAP1 co-mutation stratification - LUAD only
#   Tier 3: multivariate model per TME outcome, all mutations in that stratum's
#           panel as covariates (LUAD panel includes KRAS:STK11 / KRAS:KEAP1
#           interaction terms; LUSC panel has no interaction terms specified)
#   Tier 4: simple univariate mut-vs-wt t-test per mutation, per TME outcome,
#           disregarding all other mutations' status. Compared side-by-side
#           against Tier 3 in a combined heatmap, per stratum.
#
# NOTE on CDKN2A: in NSCLC this is frequently inactivated by homozygous
# deletion rather than point mutation. If your CDKN2A column is derived only
# from small-variant VCF calls, it likely undercounts true CDKN2A loss unless
# you've merged in copy-number data.
# ============================================================

library(tidyverse)
library(broom)

# data
dat_z <- left_join(rna, traits, "immucan_id") %>%
  mutate(LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma", "LUAD", "LUSC"),
         EGFR = ifelse(EGFR == TRUE, 1,0),
         KRAS = ifelse(KRAS == TRUE, 1,0),
         STK11 = ifelse(STK11 == TRUE, 1,0),
         KEAP1 = ifelse(KEAP1 == TRUE, 1,0),
         PIK3CA = ifelse(PIK3CA == TRUE, 1,0),
         CDKN2A = ifelse(CDKN2A == TRUE, 1, 0))

# ---- 1. Config: EDIT these to match your actual column names ----
tme_cols  <- rna_module_names
histology_col  <- "LUAD_vs_LUSC"                              # expects values "LUAD" / "LUSC"
mut_cols_luad  <- c("EGFR", "KRAS", "STK11", "KEAP1", "CDKN2A", "TP53")   # PIK3CA excluded: ~1 case in LUAD
mut_cols_lusc  <- c("CDKN2A", "PIK3CA", "TP53")                            # EGFR/KRAS/STK11/KEAP1 excluded in LUSC
min_group_n    <- 5   # minimum mutant (or wt) count required to test a gene in a stratum

df<- dat_z %>%
  dplyr::select(all_of(c(tme_cols,mut_cols_luad, mut_cols_lusc, histology_col))) %>%
  filter(!is.na(STK11)) # we will remove two patients that have missing genomic information

# sanity checks
stopifnot(all(tme_cols %in% names(df)))
stopifnot(histology_col %in% names(df))
all_mut_cols <- union(mut_cols_luad, mut_cols_lusc)
stopifnot(all(all_mut_cols %in% names(df)))
df[all_mut_cols] <- lapply(df[all_mut_cols], as.integer)  # ensure 0/1 numeric

cat("\n--- Histology counts ---\n")
print(table(df[[histology_col]]))

df_luad <- df %>% filter(.data[[histology_col]] == "LUAD")
df_lusc <- df %>% filter(.data[[histology_col]] == "LUSC")

cat(sprintf("\nLUAD n = %d, LUSC n = %d\n", nrow(df_luad), nrow(df_lusc)))

# ---- 2. Helper functions ----

# Drops a mutation from testing if it has fewer than min_group_n carriers
# (or fewer than min_group_n wt), rather than letting t.test()/lm() error out
# or silently return an unstable estimate.
check_group_size <- function(data, mut_var, min_n = min_group_n) {
  tab <- table(data[[mut_var]])
  if (length(tab) < 2 || min(tab) < min_n) {
    message(sprintf(
      "  -> dropping %s: insufficient variation/group size (counts: %s)",
      mut_var, paste(names(tab), tab, sep = "=", collapse = ", ")
    ))
    return(FALSE)
  }
  TRUE
}


run_tier4 <- function(data, tme_cols, mut_cols) {
  valid_muts <- mut_cols[sapply(mut_cols, function(m) check_group_size(data, m))]
  expand_grid(tme_var = tme_cols, mut_var = valid_muts) %>%
    pmap_dfr(function(tme_var, mut_var) {
      grp <- factor(data[[mut_var]], levels = c(0, 1), labels = c("wt", "mut"))
      tt  <- t.test(data[[tme_var]] ~ grp)
      diff_mut_minus_wt <- unname(tt$estimate["mean in group mut"] - tt$estimate["mean in group wt"])
      tibble(tme_var = tme_var, mut_var = mut_var, estimate = diff_mut_minus_wt, p_value = tt$p.value)
    }) %>%
    group_by(mut_var) %>%
    mutate(p_fdr = p.adjust(p_value, method = "BH")) %>%
    ungroup() %>%
    arrange(p_fdr)
}

# ---- 4. Tier 2 (LUAD only): within KRAS-mutant tumors, STK11/KEAP1 co-mutation ----
kras_df_luad <- df_luad %>%
  filter(KRAS == 1) %>%
  mutate(co_mut = if_else(STK11 == 1 | KEAP1 == 1,
                           "STK11_or_KEAP1_altered", "KRAS_only"))

cat("\n--- [LUAD] Tier 2 group sizes (within KRAS-mutant tumors) ---\n")
print(table(kras_df_luad$co_mut))

tier2_results_luad <- map_dfr(tme_cols, function(tme) {
  wt <- t.test(kras_df_luad[[tme]] ~ kras_df_luad$co_mut)
  tibble(tme_var = tme, ttest_p = wt$p.value)
})
tier2_results_luad$fdr <- p.adjust(tier2_results_luad$ttest_p, method = "BH")
cat("\n--- [LUAD] Tier 2: STK11/KEAP1 co-mutation effect within KRAS-mutant tumors ---\n")
print(tier2_results_luad)

# ---- 4b. Tier 2 visualization: KRAS-only vs STK11/KEAP1-altered, per TME score ----
# Panels ordered by FDR (strongest hits first, e.g. Cold/Neutrophils) so the
# STK11/KEAP1 "immune-cold" signature is easy to spot at a glance.
kras_long_luad <- kras_df_luad %>%
  dplyr::select(all_of(tme_cols), co_mut) %>%
  pivot_longer(cols = all_of(tme_cols), names_to = "tme_var", values_to = "z_score")

tme_order_tier2 <- tier2_results_luad %>% arrange(fdr) %>% pull(tme_var)
kras_long_luad$tme_var <- factor(kras_long_luad$tme_var, levels = tme_order_tier2)

tier2_sig_labels <- tier2_results_luad %>%
  mutate(
    tme_var = factor(tme_var, levels = tme_order_tier2),
    label = case_when(
      ttest_p < 0.001 & fdr < 0.1 ~ "***",
      ttest_p < 0.01 & fdr < 0.1 ~ "**",
      ttest_p < 0.05 & fdr < 0.1 ~ "*",
      TRUE        ~ ""
    )
  )

tier2_plot <- ggplot(kras_long_luad, aes(x = co_mut, y = z_score, fill = co_mut)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.6) +
  geom_jitter(width = 0.15, size = 1, alpha = 0.5) +
  facet_wrap(~tme_var, nrow = 2) +
  geom_text(data = tier2_sig_labels, aes(x = 1.5, y = Inf, label = label),
            inherit.aes = FALSE, vjust = 1.5, size = 3.5) +
  scale_fill_manual(values = c("KRAS_only" = "grey70", "STK11_or_KEAP1_altered" = "firebrick")) +
  theme_minimal() +
  labs(
    title = "KRAS-mutant LUAD: STK11/KEAP1 co-mutation vs. KRAS-only",
    subtitle = "Labels show FDR stars where significant, else raw Wilcoxon p",
    x = NULL, y = "TME z-score", fill = "Group"
  ) +
  theme(legend.position = "bottom")

print(tier2_plot)
#ggsave("tme_kras_costk11keap1_boxplot.png", tier2_plot, width = 9, height = 6, dpi = 300)

# ---- 4c. Encode KRAS+STK11/KEAP1 co-mutation as its own predictor ----
# split KRAS-mutant tumors into two
# mutually exclusive, directly-plottable predictors:
#   - KRAS_only:          KRAS-mutant, no STK11/KEAP1 alteration
#   - KRAS_STK11KEAP1_co: KRAS-mutant WITH STK11 and/or KEAP1 alteration
# STK11 and KEAP1 are left as-is (they still capture any STK11/KEAP1
# mutation regardless of KRAS status).
#

df_luad <- df_luad %>%
  mutate(
    KRAS_only          = as.integer(KRAS == 1 & STK11 == 0 & KEAP1 == 0),
    KRAS_co = as.integer(KRAS == 1 & (STK11 == 1 | KEAP1 == 1))
  )

# Check whether STK11/KEAP1 mutations occur meaningfully outside the KRAS-
# mutant group. Tested this on synthetic data: if STK11/KEAP1 occur almost
# exclusively within KRAS-mutant tumors (as expected from the literature),
# keeping them as standalone predictors alongside KRAS_STK11KEAP1_co makes
# them strongly collinear with it (VIF ~6-10 in the test), which *dilutes*
# the co-mutation signal across three correlated terms instead of clarifying
# it - the opposite of what we want. In that case, drop the standalone
# STK11/KEAP1 terms and let the composite carry that signal on its own.
stk11_keap1_outside_kras <- sum((df_luad$STK11 == 1 | df_luad$KEAP1 == 1) & df_luad$KRAS == 0)
cat(sprintf("\n[LUAD] STK11/KEAP1-mutant tumors outside the KRAS-mutant group: %d\n",
            stk11_keap1_outside_kras))

if (stk11_keap1_outside_kras < min_group_n) {
  message("  -> STK11/KEAP1 occur almost exclusively within KRAS-mutant tumors here; ",
          "dropping standalone STK11/KEAP1 terms to avoid collinearity with KRAS_STK11KEAP1_co.")
  mut_cols_luad_v2 <- c("EGFR", "KRAS_only", "CDKN2A", "KRAS_STK11KEAP1_co")
} else {
  mut_cols_luad_v2 <- c("EGFR", "KRAS","KRAS_only", "STK11", "KEAP1", "CDKN2A", "KRAS_co", "TP53")
}

# ---- 5. Tier 3 + Tier 4 (LUAD): expanded panel including the co-mutation predictor ----

tier4_luad <- run_tier4(df_luad, tme_cols, mut_cols_luad_v2)

pluad <- tier4_luad %>%
    mutate(
    label = case_when(p_value < 0.001 & p_fdr < 0.1 ~ "***",
      p_value < 0.01   & p_fdr < 0.1~ "**",
      p_value < 0.05  & p_fdr < 0.1 ~ "*",
      TRUE        ~  "")
    ) %>%
 ggplot(aes(x =factor(mut_var, levels = c("EGFR","KRAS", "KRAS_only","KRAS_co", "STK11", "KEAP1", "CDKN2A", "TP53")), y = tme_var, fill = estimate)) +
    geom_tile() +
    geom_text(aes(label = label), color = "black") +
    scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0) +
    #facet_wrap(~method) +
    theme_minimal() +
    labs(
      title = "Drivers LUAD",
      x = "Mutation", y = "RNA module", fill = paste0("Δ enrichment \n score")
    )+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
pluad

# ============================================================
# LUSC stratum: CDKN2A, PIK3CA
# ============================================================

tier4_lusc <- run_tier4(df_lusc, tme_cols, mut_cols_lusc)

plusc <- tier4_lusc %>%
  mutate(
    label = case_when(
      p_value < 0.001 & p_fdr < 0.1 ~ "***",
      p_value < 0.01  & p_fdr < 0.1 ~ "**",
      p_value < 0.05  & p_fdr < 0.1 ~ "*",
      TRUE                          ~ ""
    )
  ) %>%
  ggplot(aes(x = mut_var, y = tme_var, fill = estimate)) +
  geom_tile() +
  geom_text(aes(label = label), color = "black") +
  scale_fill_gradient2(low = "#2166AC", mid = "white", high = "#B2182B", midpoint = 0) +
  #facet_wrap(~method) +
  theme_minimal() +
 labs(
  title = "Drivers LUSC",
  x = "Mutation", y = "RNA module",
  fill = "Δ enrichment \n score"
)+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))
plusc

p <- plot_grid(pluad+theme(legend.position = "bottom"), plusc+theme(axis.text.y = element_blank(), legend.position = "bottom"),nrow = 1, rel_widths = c(2.25,1))
p

ggsave(filename = paste0(out_folder,"S6B_RNA_modules_for_mutations.pdf"),
       plot = p,
       device = cairo_pdf,
       width = 8,
       height = 6)

```
# S6C mean expression of markers across cell types and driver mutations
```{r}
mean_exp1 <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/mean_marker_expr_per_celltype_p1.csv"))
mean_exp2 <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/mean_marker_expr_per_celltype_p2.csv"))

# long format
mean_exp1 <- mean_exp1 %>%
  pivot_longer(-immucan_id)
mean_exp2 <- mean_exp2 %>%
  pivot_longer(-immucan_id)

mean_exp1$tech <- "IMC1"
mean_exp2$tech <- "IMC2"
mean_exp <- rbind(mean_exp1,mean_exp2)

mean_exp <- mean_exp %>%
  mutate(celltypes = str_split_i(name,"-",1),
         name = str_split_i(name,"-",2))
```

## panel 1 markers of interest - to be removeed
```{r}
markers_1 <- list()
# we will not subcluters "uncertain" cells as the merging will be very difficult. we may inspect this at a later time point again

mean_exp1 <- mean_exp %>%
  filter(tech == "IMC1")
#functional markers_1 for T cells
markers_1[["CD4"]] <- mean_exp1 %>%
  filter(grepl("CD27|LAG3|PD1|CD7|GrzB|ICOS|Ki67|TCF7",name)) %>%
  pull(name) %>%
  unique()

markers_1[["CD8"]] <- mean_exp1 %>%
  filter(grepl(pattern = "CD27|LAG3|PD1|CD7|GrzB|ICOS|Ki67|TCF7",name)) %>%
  pull(name) %>%
  unique()

markers_1[["Treg"]] <- mean_exp1 %>%
  filter(grepl(pattern = "CD27|LAG3|ICOS|Ki67|TCF7",name)) %>%
  pull(name) %>% 
  unique()

# B and BnT cell markers_1
markers_1[["BnT"]] <- mean_exp1 %>%
  filter(grepl(pattern = "CD27|CD45RA|CD45RO|LAG3|PD1|CD7|GrzB|FOXP3|ICOS|Ki67|TCF7|VISTA|CD40|HLADR|CD38",name))%>% 
  pull(name) %>%  
  unique()

markers_1[["B"]] <- mean_exp1 %>%  
  filter(grepl(pattern = "CD27|CD40|Ki67",name)) %>% 
  pull(name) %>% 
  unique()

# myeloid markers_1
markers_1[["MacCD163"]] <- mean_exp1 %>%  
  filter(grepl(pattern = "CD68|CD11c|PDL1|VISTA|CD40|CD14|CD206|IDO1|CD16",name)) %>% 
  pull(name) %>% 
  unique()

# we will also use CD20 for DCs to detect those DCs that sit in TLS
markers_1[["DC"]] <- mean_exp1 %>%  
  filter(grepl(pattern = "PDL1|VISTA|CD40|IDO1|CD20",name)) %>% 
  pull(name) %>%  
  unique()

# we will also use E-cad for HLADR cells as there may still be tumor cells in these cells
markers_1[["HLADR"]] <- mean_exp1 %>%  
  filter(grepl(pattern = "CD11b|CD163|CD68|CD11c|PDL1|ICOS|VISTA|TIM3|CD40|CD14|CD206|CD163|HLADR|Ecad",name)) %>%  
  pull(name) %>%   
  unique()

# NK markers_1
markers_1[["NK"]] <- mean_exp1 %>%  
  filter(grepl(pattern = "CD7|GrzB|CD16|CD45RA|CD45RO",name)) %>%
  pull(name) %>% 
  unique()

# Neutrophil markers_1
markers_1[["Neutrophil"]] <- mean_exp1 %>%  
  filter(grepl(pattern = "MPO|PDL1|VISTA",name)) %>% 
  pull(name) %>% 
  unique()

# pDC markers_1
markers_1[["pDC"]] <- mean_exp1 %>%  
  filter(grepl(pattern = "CD303|CD4|GrzB|CD45RA|CD45RO|CD27",name)) %>%  
  pull(name) %>%  
  unique()

# Plasma cell markers_1
markers_1[["plasma"]] <- mean_exp1 %>%  
  filter(grepl(pattern = "CD38|TCF7|CD27|CD45RA|CD45RO",name)) %>% 
  pull(name) %>% 
  unique()

# stroma markers_1
markers_1[["Mural"]] <- mean_exp1 %>%   
  filter(grepl(pattern = "PDGFR|SMA",name)) %>% 
  pull(name) %>% 
  unique()

# tumor markers_1
markers_1[["Tumor"]] <- mean_exp1 %>%   
  filter(grepl(pattern = "HLADR|CarbonicAnhydrase|cleavedPARP|PDL1|IDO1|CD15|Ki67|Ido|B2M",name)) %>%  
  pull(name) %>%  
  unique()
```

## panel 2 markers of interest

```{r}
mean_exp2 <- mean_exp %>%
  filter(tech == "IMC2")
markers_2 <- list()
# we will not subcluters "uncertain" cells as the merging will be very difficult. we may inspect this at a later time point again

markers_2[["CD8"]] <- mean_exp2 %>% 
  filter(grepl(pattern = "CD7$|PD1|Bcl2|GATA3|PD1|NKG2A|CXCL13|EOMES|CD137|CD134|Tbet|GITR|Tim3|Ki67",name)) %>%
  pull(name) %>%  
  unique()
  
markers_2[["CD4"]] <- mean_exp2 %>% 
  filter(grepl(pattern = "CD7$|PD1|Bcl2|GATA3|PD1|NKG2A|CXCL13|EOMES|CD137|CD134|Tbet|GITR|Tim3|Ki67",name)) %>%
  pull(name) %>%  
  unique()
markers_2[["Treg"]] <- mean_exp2 %>%
  filter(grepl(pattern = "CD7$|Bcl2|GATA3|PD1|NKG2A|CXCL13|EOMES|CD137|CD134|Tbet|GITR|Tim3|Ki67",name)) %>%
  pull(name) %>%  
  unique()


# NK markers_2
markers_2[["NK"]] <- mean_exp2 %>% 
  filter(grepl(pattern = "Tbet|Bcl2|GATA3|NKG2A|EOMES|GITR",name)) %>%
  pull(name) %>%  
  unique()


# B and BnT cell markers_2
markers_2[["B"]] <- mean_exp2 %>%
  filter(grepl(pattern = "Bcl2|Ki67",name)) %>%
  pull(name) %>%  
  unique()
markers_2[["BnT"]] <- mean_exp2 %>% 
  filter(grepl(pattern = "CD7$|PD1|FOXP3|Bcl2|GATA3|CD8a|LAG3|PD1|NKG2A|CXCL13|EOMES|CD137|CD134|Tbet|GITR|Tim3|Ki67",name)) %>%
  pull(name) %>%  
  unique()



# myeloid markers_2
markers_2[["MacCD204"]] <- mean_exp2 %>% 
  filter(grepl(pattern = "CD11c|Arg1|MMP9|CD73|CD10|Tim3",name)) %>%
  pull(name) %>%  
  unique()
markers_2[["MacCD209"]] <- mean_exp2 %>% 
  filter(grepl(pattern = "CD11c|Arg1|MMP9|CD73|CD10|Tim3",name)) %>%
  pull(name) %>%  
  unique()
markers_2[["MacCD204_CD209"]] <- mean_exp2 %>%
  filter(grepl(pattern = "CD11c|Arg1|MMP9|CD73|CD10|Tim3",name)) %>%
  pull(name) %>%  
  unique()
# DCs
markers_2[["DC"]] <- mean_exp2 %>% 
  filter(grepl(pattern = "CD137|Tim3|CD20$",name)) %>%
  pull(name) %>%  
  unique()
markers_2[["mregDC"]] <- mean_exp2 %>% 
  filter(grepl(pattern = "CD11c|CD137|Tim3",name)) %>%
  pull(name) %>%  
  unique()


# MDSCs
markers_2[["MDSC"]] <- mean_exp2 %>%
  filter(grepl(pattern = "CD11b|CD10|CXCL8|Arg1|MMP9",name)) %>%
  pull(name) %>%  
  unique()
markers_2[["PMN_MDSC"]] <- mean_exp2 %>%
  filter(grepl(pattern = "CD11b|CD15$|CD66b|CD10|CXCL8|Arg1|MMP9",name)) %>%
  pull(name) %>%  
  unique()
markers_2[["Neutrophil"]] <- mean_exp2 %>%
  filter(grepl(pattern = "CD11b|CD15$|CD66b|CD10|CXCL8|Arg1|MMP9",name)) %>%
  pull(name) %>%  
  unique()


# stroma markers_2
markers_2[["Vasculature"]] <- mean_exp2 %>% filter(grepl(pattern = "CD31|CD146",name)) %>%
  pull(name) %>%  
  unique()

markers_2[["vCAF"]] <- mean_exp2 %>%
  filter(grepl(pattern = "SMA|CD146",name)) %>%
  pull(name) %>%  
  unique()

markers_2[["Fibro"]] <- mean_exp2 %>% 
  filter(grepl(pattern = "Podoplanin|CD10|MMP9|SMA",name)) %>%
  pull(name) %>%  
  unique()


# tumor markers_2
markers_2[["Tumor"]] <- mean_exp2 %>%
  filter(grepl(pattern = "panCK|Ecad|Arg1|EGFR|MMP9|p53|Lamp3|CD73|Bcl2|GATA3|CD155|CD10|CD56|CXCL8|Ki67|Podoplanin|IgG",name)) %>%
  pull(name) %>%  
  unique()

```

to do log2 foldchange comparison we will run the statistics on the arcsinh expression spase. to calculate log2 FC we will backtransform the mean and calculate the log2FC.

```{r, fig.height=25, fig.width=25}
cur_files_1 <- lapply(names(markers_1), function(x){
  cur_dat <- mean_exp %>%
    filter(celltypes == x,
           name %in% markers_1[[x]])
})
cur_files_2 <- lapply(names(markers_2), function(x){
  mean_exp %>%
    filter(celltypes == x,
           name %in% markers_2[[x]])
})

mean_exp1 <- do.call(rbind,cur_files_1)
mean_exp2 <- do.call(rbind,cur_files_2)

mean_data <- distinct(rbind(mean_exp1,mean_exp2))

mean_exp %>%
  left_join(traits,"immucan_id") %>%
  group_by(EGFR, celltypes,name) %>%
  summarise(mean(value))

library(dplyr)
library(rstatix)

p.list <- list()

# LUAD EGFR
p.list[["LUAD_EGFR"]] <- mean_data %>%
  left_join(traits,"immucan_id") %>%
  filter(!is.na(EGFR),
         simple_histology == "Adenocarcinoma") %>%
  group_by(celltypes, name) %>%
  summarise(
    wilcox_test = list(wilcox_test(value ~ EGFR, data = cur_data(), detailed = TRUE)),
    logFC = log2(sinh(mean(value[EGFR == TRUE], na.rm = TRUE)) / sinh(mean(value[EGFR == FALSE], na.rm = TRUE))),
    .groups = "drop"
  ) %>%
  unnest(wilcox_test) %>%
  mutate(condition = "LUAD_EGFR")

# LUAD KRAS only
p.list[["LUAD_KRAS_only"]] <- mean_data %>%
  left_join(traits,"immucan_id") %>%
  mutate(KRAS_only = ifelse(KRAS == TRUE & STK11 == FALSE & KEAP1 == FALSE, TRUE, FALSE)) %>%
  filter(!is.na(KRAS_only),
         simple_histology == "Adenocarcinoma") %>%
  group_by(celltypes, name) %>%
  summarise(
    wilcox_test = list(wilcox_test(value ~ KRAS_only, data = cur_data(), detailed = TRUE)),
    logFC = log2(sinh(mean(value[KRAS_only == TRUE], na.rm = TRUE)) / sinh(mean(value[KRAS_only == FALSE], na.rm = TRUE))),
    .groups = "drop"
  ) %>%
  unnest(wilcox_test) %>%
  mutate(condition = "LUAD_KRAS_only")

# LUAD KRAS_co
p.list[["LUAD_KRAS_co"]] <- mean_data %>%
  left_join(traits,"immucan_id") %>%
  filter(!is.na(KRAS_co),
         simple_histology == "Adenocarcinoma") %>%
  group_by(celltypes, name) %>%
  summarise(
    wilcox_test = list(wilcox_test(value ~ KRAS_co, data = cur_data(), detailed = TRUE)),
    logFC = log2(sinh(mean(value[KRAS_co == TRUE], na.rm = TRUE)) / sinh(mean(value[KRAS_co == FALSE], na.rm = TRUE))),
    .groups = "drop"
  ) %>%
  unnest(wilcox_test) %>%
  mutate(condition = "LUAD_KRAS_co")


#LUAD STK11
p.list[["LUAD_STK11"]] <- mean_data %>%
  left_join(traits,"immucan_id") %>%
  filter(!is.na(STK11),
         simple_histology == "Adenocarcinoma") %>%
  group_by(celltypes, name) %>%
  summarise(
    wilcox_test = list(wilcox_test(value ~ STK11, data = cur_data(), detailed = TRUE)),
    logFC = log2(sinh(mean(value[STK11 == TRUE], na.rm = TRUE)) / sinh(mean(value[STK11 == FALSE], na.rm = TRUE))),
    .groups = "drop"
  ) %>%
  unnest(wilcox_test) %>%
  mutate(condition = "LUAD_STK11")

# LUAD KEAP1
p.list[["LUAD_KEAP1"]] <- mean_data %>%
  left_join(traits,"immucan_id") %>%
  filter(!is.na(KEAP1),
         simple_histology == "Adenocarcinoma") %>%
  group_by(celltypes, name) %>%
  summarise(
    wilcox_test = list(wilcox_test(value ~ KEAP1, data = cur_data(), detailed = TRUE)),
    logFC = log2(sinh(mean(value[KEAP1 == TRUE], na.rm = TRUE)) / sinh(mean(value[KEAP1 == FALSE], na.rm = TRUE))),
    .groups = "drop"
  ) %>%
  unnest(wilcox_test) %>%
  mutate(condition = "LUAD_KEAP1")

# LUAD TP53
p.list[["LUAD_p53"]] <- mean_data %>%
  left_join(traits,"immucan_id") %>%
  filter(!is.na(TP53),
         simple_histology == "Adenocarcinoma") %>%
  group_by(celltypes, name) %>%
  summarise(
    wilcox_test = list(wilcox_test(value ~ TP53, data = cur_data(), detailed = TRUE)),
    logFC = log2(sinh(mean(value[TP53 == TRUE], na.rm = TRUE)) / sinh(mean(value[TP53 == FALSE], na.rm = TRUE))),
    .groups = "drop"
  ) %>%
  unnest(wilcox_test) %>%
  mutate(condition = "LUAD_TP53")

# LUSC PIK3CA
p.list[["LUSC_PIK3CA"]] <- mean_data %>%
  left_join(traits,"immucan_id") %>%
  filter(!is.na(PIK3CA),
         simple_histology == "Squamous cell carcinoma") %>%
  group_by(celltypes, name) %>%
  summarise(
    wilcox_test = list(wilcox_test(value ~ PIK3CA, data = cur_data(), detailed = TRUE)),
    logFC = log2(sinh(mean(value[PIK3CA == TRUE], na.rm = TRUE)) / sinh(mean(value[PIK3CA == FALSE], na.rm = TRUE))),
    .groups = "drop"
  ) %>%
  unnest(wilcox_test) %>%
  mutate(condition = "LUSC_PIK3CA")

# LUSC TP53
p.list[["LUSC_TP53"]] <- mean_data %>%
  left_join(traits,"immucan_id") %>%
  filter(!is.na(TP53),
         simple_histology == "Squamous cell carcinoma") %>%
  group_by(celltypes, name) %>%
  summarise(
    wilcox_test = list(wilcox_test(value ~ TP53, data = cur_data(), detailed = TRUE)),
    logFC = log2(sinh(mean(value[TP53 == TRUE], na.rm = TRUE)) / sinh(mean(value[TP53 == FALSE], na.rm = TRUE))),
    .groups = "drop"
  ) %>%
  unnest(wilcox_test) %>%
  mutate(condition = "LUSC_TP53")

dat <- do.call(rbind,p.list)

dat <- dat %>%
  group_by(celltypes, condition) %>%
  mutate(p_adjust =p.adjust(p = p,method = "BH"),
         FCcutoff = ifelse(abs(logFC)>0.33,TRUE,FALSE),
         show =  p_adjust < 0.1 & FCcutoff == TRUE)
dat %>%
  filter(celltypes == "Treg",
         condition == "LUSC_PIK3CA")

library(ggrepel)

p <- dat %>%
  ggplot(aes(x = logFC,y = -log10(p), col=condition))+
  geom_point(size = 4)+
  geom_text_repel(data = dat %>% filter(show == TRUE),aes(x = logFC,y = -log10(p),label= paste(name, celltypes,sep = "-")), col = "black",size = 5,box.padding = 0.5,max.overlaps = 20)+
  scale_color_manual(values = c("LUAD_EGFR" = "green", "LUAD_KRAS" = "orange", "LUAD_KRAS_co" = "darkorange4", "LUAD_KRAS_only" = "red", "LUAD_KEAP1" = "gold", "LUAD_STK11" = "magenta", 
                                "LUAD_TP53" = "cyan", "LUSC_TP53" = "blue", "LUSC_PIK3CA" = "darkorchid"))+
  theme_bw()+
  theme(legend.position = "bottom",
        legend.text = element_text(size = 12),
        axis.text.x = element_text(size =12),
        axis.text.y = element_text(size =12),
        axis.title.x = element_text(size =12),
        axis.title.y = element_text(size =12))

p

ggsave(filename = paste0(out_folder,"S6C_marker_exp_fold_changes_all_celltypes.pdf"),
       plot = p,
       device = "pdf",
       width = 16,
       height = 12)

dat %>%
  filter(condition == "LUAD_EGFR",
         celltypes == "Tumor")
```

# S6D number HLADR positive tumor cells in EGFR

```{r, fig.width=4, fig.height=8}
sce1 <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/IMC/Panel_1/Rout/sce.rds")
sce1 <- sce1[,sce1$ROI != 4]

counts <- as.data.frame(t(assay(sce1,"exprs")))
head(counts)
dim(counts)

counts$celltypes <- sce1$celltypes
counts$immucan_id <- sce1$immucan_id
counts$sample_id <- sce1$sample_id
counts <- left_join(counts, traits,"immucan_id")

p <- counts %>%
  filter(celltypes == "Tumor", 
         simple_histology == "Adenocarcinoma") %>%
  ggplot(aes(x = Ecad,y = HLADR))+
  geom_point(size = 0.1)+
  facet_wrap(.~EGFR)+
  theme_bw()

counts_EGFR <- counts %>%
  filter(celltypes == "Tumor", 
         simple_histology == "Adenocarcinoma",
         EGFR == TRUE)

counts_EGFRneg <- counts %>%
  filter(celltypes == "Tumor", 
         simple_histology == "Adenocarcinoma",
         EGFR == FALSE)

fit2 <- mixtools::normalmixEM(counts_EGFRneg$HLADR,k = 2)
# get cut point
post <- fit2$posterior[, 1]
# Approximate cut point: where posteriors are equal (0.5)
idx <- which.min(abs(post - 0.5))
cut_point_HLADR <- counts_EGFRneg$HLADR[idx]
cut_point_HLADR

Cairo::CairoPNG(filename = paste0(out_folder,"S6D_HLADR_tumor_cell_EGFRmutated.png"),width = 1000, height = 500,units = "px")
par(mfrow = c(1,2))
LSD::heatscatter(x = counts_EGFR$Ecad, y = counts_EGFR$HLADR,main = "EGFR mutated")
abline(h = cut_point_HLADR, lty = 2, lwd = 5)
LSD::heatscatter(x = counts_EGFRneg$Ecad, y = counts_EGFRneg$HLADR, main = "EGFR wild type")
abline(h = cut_point_HLADR, lty = 2, lwd = 5)
dev.off()
```

# S6E CD8 T cells CXC13 in EGFR

```{r, fig.height=5, fig.width=5}
p_CD8_CXCL13_EGFR <- bplot_cor("CD8",markers =  c("CXCL13_p2", "PD1_p2","PD1_p1","LAG3_p1", "Tim3_p2"),mutation = "EGFR", rownum = 1, histology = "Adenocarcinoma")

p_CD8_CXCL13_EGFR

ggsave(filename = paste0(out_folder,"S6E_exhaustion_on_CD8_Tcells.pdf"),
       plot = p_CD8_CXCL13_EGFR,
       device = "pdf",
       width = 4,
       height = 1.8)
```

# CD4 T cells CXC13 in EGFR

```{r, fig.height=5, fig.width=15}
p1 <- bplot_cor("CD4",markers =  c(markers_2$CD4,markers_1$CD4),mutation = "EGFR", rownum = 1, histology = "Adenocarcinoma")

p1

p2 <- bplot_cor("CD8",markers =  c(markers_2$CD8,markers_1$CD8),mutation = "EGFR", rownum = 1, histology = "Adenocarcinoma")

p2
```

# S6F Ki67 and p53 on tumor cells in LUSC

```{r}
p_Tumor_Ki67_P53_TP53_LUSC <- bplot_cor(ct = "Tumor",markers =  c("p53_p2","Ki67_p1","Ki67_p2"), mutation = "TP53",histology = "Squamous cell carcinoma",headroom = 0.1)

ggsave(filename = paste0(out_folder,"S6F_Ki67_p53_tumor_cells_LUSC.pdf"),
       plot = p_Tumor_Ki67_P53_TP53_LUSC,
       device = "pdf",
       width = 2.5,
       height = 1.8)
```

# S6G Tumor cells EGFR HLADR in STK11
```{r, fig.height=5, fig.width=5}
p_Tumor_HLADR_STK11 <- bplot_cor(ct = "Tumor", c("EGFR_p2","HLADR_p1", "CD155_p2", "CD73_p2", "Arg1_p2"), mutation = "STK11", rownum = 1, histology = c("Adenocarcinoma"),stat_size = 3, sig_format = TRUE)

ggsave(filename = paste0(out_folder,"S6G_HLADR_and_EGFR_on_tumor_cells_STK11.pdf"),
       plot = p_Tumor_HLADR_STK11,
       device = "pdf",
       width = 4,
       height = 2.5)
```

# S6H Tumor cells EGFR HLADR in KEAP1 
```{r, fig.height=5, fig.width=5}
p_Tumor_HLADR_KEAP1 <- bplot_cor(ct = "Tumor", c("EGFR_p2","HLADR_p1", "CD155_p2", "CD73_p2", "Arg1_p2"), mutation = "KEAP1", rownum = 1, histology = c("Adenocarcinoma"),stat_size = 3,sig_format = TRUE)
p_Tumor_HLADR_KEAP1
ggsave(filename = paste0(out_folder,"S6H_HLADR_and_EGFR_on_tumor_cells_KEAP1.pdf"),
       plot = p_Tumor_HLADR_KEAP1,
       device = "pdf",
       width = 4,
       height = 2.5)

table(traits$KRAS, traits$STK11)
```

# S6I GITR on Tregs

```{r, fig.width=10, fig.height=10}
sce2 <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/IMC/Panel_2/Rout/sce.rds")
counts2 <- as.data.frame(t(assay(sce2,"exprs")))
head(counts2)
dim(counts2)

counts2$celltypes <- sce2$celltypes
counts2$immucan_id <- sce2$immucan_id
counts2$sample_id <- sce2$sample_id
counts2 %>%
  filter(celltypes == "Treg") %>%
  ggplot(aes(x = FOXP3,y = GITR))+
  geom_point()

counts2 <- left_join(counts2, traits,"immucan_id")

counts2 %>%
  filter(simple_histology == "Squamous cell carcinoma",
         celltypes == "Treg") %>%
  mutate(PIK3CApos = ifelse(GITR > 1, TRUE,FALSE)) %>%
  group_by(sample_id,.drop = FALSE) %>%
  dplyr::count(PIK3CApos) %>%
  mutate(total_cells = sum(n),
         frac = n/total_cells) %>%
  filter(PIK3CApos == TRUE) %>%
  arrange(frac)

# selected images
ims <- c("IMMUcan_Batch20210921_LUNG_10041582-LUNG-VAR-TIS-01-IMC-02_003","IMMUcan_Batch20210701_LUNG_10040761-LUNG-VAR-TIS-01-IMC-02_001")


# load images
library(cytomapper)
masks <- readRDS(paste0(mount_path,"data/IMC/Panel_2/Rout/masks.rds"))
images <- readRDS(paste0(mount_path,"data/IMC/Panel_2/Rout/images.rds"))

cur_df <- as.data.frame(colData(sce2))
rownames(cur_df) <- NULL
cur_df <- cur_df %>%
  select(immucan_id,image, sample_id) %>%
  distinct() %>%
  mutate(image = gsub(".tiff","",image))
rownames(cur_df) <- cur_df$image


# unify metadata of masks and images
mcols(images)$immucan_id <- cur_df[names(images),"immucan_id"]
mcols(masks)$immucan_id <- cur_df[names(masks),"immucan_id"]
mcols(images)$sample_id <- cur_df[names(images),"sample_id"]
mcols(masks)$sample_id <- cur_df[names(masks),"sample_id"]


cur_sce <- sce2[,sce2$sample_id %in% ims]


cur_ims <- getImages(images,ims)
cur_masks <- getImages(masks, ims)

# read images into memory and normalize
cur_images <- CytoImageList(cur_ims, on_disk = FALSE)
cur_masks <- CytoImageList(cur_masks, on_disk = FALSE)
cur_images <- scaleImages(cur_images,value = 2^16-1)
cur_masks <- scaleImages(cur_masks,value = 2^16-1)
cur_images <- normalize(cur_images, separateImages = TRUE)
cur_images <- normalize(cur_images, separateImages = TRUE, 
                        inputRange = c(0, 0.2))

channelNames(cur_images) <- rownames(cur_sce)

# define all tumor and non-tumor cells
cur_sce$Treg <- "nonTreg"
cur_sce[,cur_sce$celltypes == "Treg"]$Treg <- "Treg"
plot_sce <- cur_sce[,cur_sce$celltypes == "Treg"]

test_images <- endoapply(cur_images, gblur, sigma = 1)

# make sure png device is avaialable
options(bitmapType = "cairo")

# generate images
plotPixels(image = test_images,
           mask = cur_masks,
           object = plot_sce,
           cell_id = "ObjectNumber",
           img_id = "sample_id",
           colour_by = c("CD137", "GITR", "FOXP3"),
           outline_by = "Treg",
           colour = list(CD137 = c("black", "green"),GITR = c("black","red"),FOXP3 = c("black","blue"),
                        Treg = c(Treg = "white")),
           bcg = list(CD137  = c(0,1.5,1),
                      GITR = c(0,1.5,1),
                      FOXP3  = c(0,3,1)),
           image_title = NULL,
           save_plot = list(filename = paste0(out_folder,"S6I_GITR_Tregs_blurred.png")))



plotPixels(image = cur_images,
           mask = cur_masks,
           object = plot_sce,
           cell_id = "ObjectNumber",
           img_id = "sample_id",
           colour_by = c("CD137", "GITR", "FOXP3"),
           outline_by = "Treg",
           colour = list(CD137 = c("black", "green"),GITR = c("black","red"),FOXP3 = c("black","blue"),
                        Treg = c(Treg = "white")),
           bcg = list(CD137  = c(0,1.5,1),
                      GITR = c(0,1.5,1),
                      FOXP3  = c(0,3,1)),
            image_title = NULL,save_plot = list(filename = paste0(out_folder,"S6I_GITR_Tregs.png")))

```

