---
title: "Figure 5 - MOFA"
output: html_document
date: '2023-03-15'
---

```{r}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

# Set path
if (Sys.info()["sysname"] == "Darwin"){
  mount_path <- "/Volumes/G_DQBM_BB_Central$/projects/immucan/"
} else if (Sys.info()["sysname"] == "Linux"){
  mount_path <- "/mnt/immucan_volume/"
} else if (Sys.info()["sysname"] == "Windows"){
  mount_path <- "Y:/"
} else {
  stop("Operating system is not supported")
}

options(bitmapType = "cairo")
```


load libraries
```{r}
library(ggplot2)
library(reticulate)
library(MOFA2)
library(dplyr)
library(tidyverse)

```

## load model
```{r}
mount_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/MOFA"
out_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/revised_figures/Figure_5/"


model_path <- paste0(mount_path, '/mofa_V6_minor.rds')

if (!dir.exists(out_path)){
  dir.create(out_path)
  dir.create(paste0(out_path, "/extras"))
}
model <- readRDS(model_path)
```

## flip factors for explorations
note, we are using the original rotations of the model
```{r, eval=FALSE}
# ## flip factor values
# model@expectations$Z$group1[,"Factor1"] <- -model@expectations$Z$group1[,"Factor1"]
# # flip feature weights
# model@expectations$W$Densities[,"Factor1"] <- -model@expectations$W$Densities[,"Factor1"]
# model@expectations$W$Expressions[,"Factor1"] <- -model@expectations$W$Expressions[,"Factor1"]
# model@expectations$W$RNA[,"Factor1"] <- -model@expectations$W$RNA[,"Factor1"]
# model@expectations$W$BINARY[,"Factor1"] <- -model@expectations$W$BINARY[,"Factor1"]
# model@expectations$W$Cellohood[,"Factor1"] <- -model@expectations$W$Cellohood[,"Factor1"]
# model@expectations$W$Interactions[,"Factor1"] <- -model@expectations$W$Interactions[,"Factor1"]
# 
# saveRDS(model, model_path)
```

## Overview
```{r}
plot_data_overview(model)
```


# 5B Variation explained
```{r, fig.height=6, fig.width=15}
library(MOFA2)
library(ComplexHeatmap)
library(circlize)
library(dplyr)
library(RColorBrewer)
library(grid)

# Step 1: Get variance explained per view & factor
var_df <- get_variance_explained(model, as.data.frame = TRUE)$r2_per_factor

# Step 2: Matrix with factors as rows, views as columns
var_mat <- var_df %>%
  dplyr::select(view, factor, value) %>%
  tidyr::pivot_wider(names_from = view, values_from = value, values_fill = 0) %>%
  tibble::column_to_rownames("factor") %>%
  as.matrix()

# Step 3: Compute row and column sums
row_sums <- rowSums(var_mat)
col_sums <- colSums(var_mat)
# Reorder columns (views) by total variance explained
sorted_view_order <- names(sort(col_sums, decreasing = TRUE))
var_mat <- var_mat[, sorted_view_order]
col_sums <- col_sums[sorted_view_order]
views <- sorted_view_order


# Step 4: Color definitions
green_scale <- colorRamp2(
  c(0, max(var_mat)),
  c("white", "#238b45")  # white → strong green
)
blue_scale <- colorRamp2(c(0, max(c(row_sums, col_sums))), c("white", "blue"))

# View color palette (matching other plots)
view_colors <- structure(
  RColorBrewer::brewer.pal(length(views), "Set1"),
  names = sort(views)
)

# Step 5: Annotations with text labels
row_anno <- rowAnnotation(
  `Total %` = anno_barplot(row_sums, gp = gpar(fill = blue_scale(row_sums))),
  total_label = anno_text(sprintf("%.1f", row_sums), location = 0, just = "left", gp = gpar(fontsize = 8, fontface = "bold")),
  show_annotation_name = FALSE
)

col_anno <- HeatmapAnnotation(
  `Total %` = anno_barplot(col_sums, gp = gpar(fill = blue_scale(col_sums))),
  total_label = anno_text(sprintf("%.1f", col_sums), 
                          location = 0.5, 
                          just = "top", 
                          rot = 0,                       # ✅ Rotation goes here
                          gp = gpar(fontsize = 8, fontface = "bold")),
  annotation_name_side = "left",
  show_annotation_name = TRUE,
  annotation_name_rot = 45
)


# Custom annotation grob (as a gTree)
custom_view_grob <- function(views, view_colors) {
  n <- length(views)
  grid::gTree(children = do.call(gList, lapply(seq_along(views), function(i) {
    view <- views[i]
    vp <- viewport(
      x = unit(i - 0.5, "native"),
      y = unit(0.5, "npc"),
      width = unit(1, "native"),
      height = unit(1, "npc"),
      just = c("center", "center")
    )
    gList(
      rectGrob(
        gp = gpar(fill = view_colors[view], col = NA),
        vp = vp
      ),
      textGrob(
        label = view,
        gp = gpar(col = "black", fontsize = 8, fontface = "bold"),
        rot = 90,  
        vp = vp
      )
    )
  })))
}

# HeatmapAnnotation using annotation_function
col_anno_bottom <- HeatmapAnnotation(
  View = function(index) {
    pushViewport(viewport(xscale = c(0, length(views))))
    grid.draw(custom_view_grob(views, view_colors))
    popViewport()
  },
  annotation_height = unit(6, "mm"),
  show_annotation_name = FALSE
)

# Step 6: Final heatmap
ht_opt$ROW_ANNO_PADDING <- unit(0.5, "cm")
ht_opt$COLUMN_ANNO_PADDING <- unit(0.5, "cm")  # Optional
view_order <- colnames(var_mat)

ht <- Heatmap(
  var_mat,
  name = "% Variance",
  col = green_scale,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  row_names_side = "left",
  column_names_side = "bottom",
  left_annotation = row_anno,
  top_annotation = col_anno,
  bottom_annotation = col_anno_bottom,
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(sprintf("%.1f", var_mat[i, j]), x, y, gp = gpar(fontsize = 9))
  },
  heatmap_legend_param = list(
    title = "Variance\nExplained [%]",
    title_gp = gpar(fontsize = 8, fontface = "bold")
  ),
  show_column_names = FALSE
)

# Set PDF output
pdf(paste0(out_path, "/5B_variance_explained.pdf"), width = 6, height = 4)  # adjust width/height as needed


# Draw the heatmap
draw(ht, padding = unit(c(6, 2, 2, 8), "mm"))

# Close the PDF device
dev.off()
```

# 5C Factor correlations
## calculate per patient inflammation
### get gene signatures
```{r}
library(org.Hs.eg.db)
library(AnnotationDbi)
library(msigdbr)
library(singscore)

rna_mat <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/rna_mat_tpm.rds")


# ── gene signatures ────────────────────────────────────────────────────
hallmark_sets <- msigdbr(species = "Homo sapiens", collection = "H") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  split(.$gs_name) %>%
  lapply(function(x) x$gene_symbol)

# Targeted Hallmark gene sets of interest
target_sets <- c(
  "HALLMARK_INFLAMMATORY_RESPONSE",
  "HALLMARK_INTERFERON_GAMMA_RESPONSE",
  "HALLMARK_INTERFERON_ALPHA_RESPONSE",
  "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
  "HALLMARK_IL6_JAK_STAT3_SIGNALING",
  "HALLMARK_IL2_STAT5_SIGNALING"
)
gs_list_hallmark <- hallmark_sets[intersect(target_sets, names(hallmark_sets))]

# ── 2. Custom gene sets ────────────────────────────────────────────────────────
# CXCL9/10/11 chemokine axis (T cell recruitment)
gs_list_custom <- list(
  CXCL9_10_11_axis = c("CXCL9", "CXCL10", "CXCL11", "CXCR3",
                       "CXCL13", "CXCR5"),
  T_cell_activation = c("CD8A", "CD8B", "GZMB", "PRF1", "IFNG",
                        "TNF", "IL2", "PDCD1", "HAVCR2", "LAG3"),
  type_I_IFN = c("IFNB1", "IRF7", "MX1", "OAS1", "OAS2", "OAS3",
                 "ISG15", "IFIT1", "IFIT2", "IFIT3", "RSAD2", "HERC5"),
  NF_kB_target  = c("NFKB1", "RELA", "RELB", "NFKB2", "REL",
                    "BCL2L1", "IL1B", "IL6", "CXCL8", "TNF", "ICAM1"),
  TCR_signaling = c("CD3D", "CD3E", "CD3G", "ZAP70", "LCK", "LAT",
                    "ITK", "PLCG1", "NFATC1", "NFATC2"),
  cytotoxic = c("GZMB", "GZMA", "GZMK", "PRF1", "NKG7", "IFNG", "KLRG1"),
  exhaustion = c("PDCD1","LAG3","HAVCR2","TX","CTLA4", "IDO1")
  )

all_gs <- c(gs_list_hallmark, gs_list_custom)

# compute score
rna_for_score <- t(rna_mat)

# Filter gene sets to genes present in the RNA-seq matrix
all_gs <- lapply(all_gs, function(g) intersect(g, rownames(rna_for_score)))
all_gs <- all_gs[sapply(all_gs, length) >= 4]



overlap <- sapply(all_gs, function(g) sum(g %in% rownames(rna_for_score)))
total   <- sapply(all_gs, length)
print(data.frame(found = overlap, total = total, pct = round(100 * overlap / total)))

# ── Score with singscore ──────────────────────────────────────────────────────
ranked <- rankGenes(rna_for_score)

scores_list <- lapply(names(all_gs), function(nm) {
  sc <- simpleScore(ranked, upSet = all_gs[[nm]], centerScore = TRUE)
  setNames(sc$TotalScore, rownames(sc))
})


inf_scores <- as.data.frame(do.call(cbind, scores_list))
colnames(inf_scores) <- names(all_gs)
rownames(inf_scores) <- colnames(rna_for_score)
inf_scores$immucan_id <- rownames(inf_scores)
rownames(inf_scores) <- NULL

```
## get infiltration score
```{r}
inf_dat <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/infiltration_harmonized.rds")

# compare densities of infiltrating, bystanding and ignored from IMC1, IMC2 and mIF
inf_dat <- inf_dat %>%
  group_by(immucan_id,tech) %>%
  mutate(total_CD8 = sum(bystanding,ignored,infiltrating,na.rm = TRUE)/im_size,
         bystanding = round(bystanding/im_size),
         ignored = round(ignored/im_size),
         infiltrating = round(infiltrating/im_size)) %>%
  ungroup()

# define interactor over bystander ratio
inf_dat <- inf_dat %>%
  group_by(immucan_id, tech) %>%
  mutate(int_over_bystander = (infiltrating+bystanding)/ignored)

inf_dat[(is.infinite(inf_dat$int_over_bystander)),]$int_over_bystander <- NA

# we will average the ratios across technologies
inf_dat <- inf_dat %>%
  group_by(immucan_id) %>%
  mutate(mean_inf_ratio = mean(int_over_bystander,na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(mean_inf_ratio_dicho = ntile(mean_inf_ratio,2),
         mean_inf_ratio_dicho = ifelse(mean_inf_ratio_dicho == 2, "high", "low"))%>%
  dplyr::select(immucan_id,mean_inf_ratio) %>%
  distinct() %>%
  as.data.frame()
inf_dat
```


## load other data

```{r}
mofa <- readRDS( '/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/MOFA/mofa_V6_minor.rds')
meta_scores <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/patient_imaging_meta_cluster_scores_tpm.rds")
rna <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/rna_scores_tpm.rds")
traits <- read_csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/all_traits.csv")


# preprocess mofa
lf_raw <- as.data.frame(get_factors(mofa)[["group1"]])   # or [[1]]

lf_raw <- lf_raw %>%
  mutate(immucan_id = rownames(lf_raw),
         immucan_id = gsub("_","-",immucan_id))

# preprocess imaging modules
meta_scores_df <- as.data.frame(meta_scores)
meta_scores_df$immucan_id <- rownames(meta_scores_df)
rownames(meta_scores_df) <- NULL


# preprocess RNA modules
rna <- as.data.frame(rna)
rna$immucan_id <- rownames(rna)
rownames(rna) <- NULL
colnames(rna) <- gsub(" ", "_", colnames(rna))
colnames(rna) <- gsub("-", "_", colnames(rna))

# cur_traits
cur_traits <- traits %>%
  dplyr::select(immucan_id,age_dicho, gender,stage_1_2,stage_1_3,stage_2_3,LUAD_vs_LUSC,never_vs_current_smoker,        never_vs_Ex_smoker,current_vs_Ex_smoker, N0_vs_Nx,M0_vs_Mx, long_survivors,T1_vs_Tother,EGFR,TP53,KRAS, PIK3CA,STK11,KEAP1, STK11,TPS, CPS, TMB ) %>%
  mutate(
    KRAS_only          = ifelse(KRAS == 1 & STK11 == 0 & KEAP1 == 0, "TRUE", "FALSE"),
    KRAS_co = ifelse(KRAS == 1 & (STK11 == 1 | KEAP1 == 1),"TRUE","FALSE")
  )

#IF1 T cell density
dens_dat <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/densities_dat.rds")
T1_density <- dens_dat %>%
  dplyr::select(immucan_id, T_density__IF1)

# merge
df_all <- plyr::join_all(dfs = list(lf_raw, meta_scores_df, rna, inf_scores, inf_dat, cur_traits, T1_density),by = "immucan_id",type = "left")
```

## compute correlation
```{r}
cur_traits <- c("age_dicho", "gender","stage_1_2","stage_1_3","stage_2_3","LUAD_vs_LUSC","never_vs_current_smoker",
            "never_vs_Ex_smoker","current_vs_Ex_smoker", "N0_vs_Nx","M0_vs_Mx", "long_survivors","T1_vs_Tother", "EGFR","TP53","PIK3CA","KRAS","KRAS_only","KRAS_co","KEAP1","STK11", "TPS","CPS","TMB","T_density__IF1","mean_inf_ratio")
condition <- c("low", "Female","I","I","II","Adenocarcinoma","Non-smoker","Non-smoker","Current smoker", "node_neg", "met_neg", "alive","T1","TRUE","TRUE","TRUE","TRUE","TRUE", "TRUE", "TRUE", "TRUE","numeric","numeric","numeric","numeric","numeric")
names(condition) <- cur_traits

factors <- get_factors(model, as.data.frame = TRUE) %>%
  dplyr::select(- group) %>%
  pivot_wider(names_from = "factor", values_from = value) %>%
  column_to_rownames("sample")

test_vars <- colnames(df_all)[8:ncol(df_all)]
lf_vars <- colnames(lf_raw)[1:6]

results <- lapply(test_vars, function(clin_var) {
  values_raw <- df_all[[clin_var]]
  print(clin_var)
  # Try coercion to numeric
  values_num <- suppressWarnings(as.numeric(as.character(values_raw)))
  num_valid <- sum(!is.na(values_num))

  per_feat <- lapply(lf_vars, function(fac) {
    x <- df_all[[fac]]
    # 1. Treat as numeric if coercion worked well (e.g., >80% non-NA)
    if (num_valid > 0.6 * length(values_raw)) {
      temp <- cor.test(x, values_num, method = "spearman")
      p_val <- temp$p.value
      estimate <- temp$estimate
      
      return(data.frame(estimate = estimate, p_val = p_val, type = "continuous"))
    }
    
    # 2. Treat as binary if exactly 2 unique non-NA values
    values_fac <- as.factor(values_raw)
    if (nlevels(values_fac) == 2) {
      values <- values_raw == condition[clin_var]
      if (table(values)[[1]] >= 2 & table(values)[[2]] >= 2){
        temp <- t.test(x ~ values)
        diff_mean <- diff(temp$estimate)
        return(data.frame(estimate = diff_mean, p_val = temp$p.value, type = "binary"))
      }else{
        return(data.frame(estimate = NA, p_val = NA, type = "binary"))
      }
    }
    
    # 3. Treat as categorical if ≥3 levels
    if (is.factor(values_fac) || is.character(values_raw)) {
      # Skip test if x has no variance
      if (length(unique(na.omit(values_fac))) <= 1) {
        return(data.frame(estimate = NA, p_val = NA, type = "categorical"))
      }
      
      temp <- kruskal.test(x, values_fac)
      group_means <- tapply(x, values_fac, mean, na.rm = TRUE)
      effect_estimate <- max(group_means, na.rm = TRUE) - min(group_means, na.rm = TRUE)
      
      return(data.frame(estimate = effect_estimate, p_val = temp$p.value, type = "categorical"))
    }
    
    # 4. Fallback
    return(NA)
  })
  
  temp <- do.call(rbind, per_feat)
  temp$factor <- colnames(factors)
  temp$feature <- clin_var
  
  return (temp)
})

results <- do.call(rbind, results)


## survival analysis
library(survminer)
library(survivalAnalysis)
library(survival)

factors <- get_factors(model, as.data.frame = TRUE) %>%
  dplyr::select(- group) %>%
  pivot_wider(names_from = "factor", values_from = value) %>%
  column_to_rownames("sample")

clin <- samples_metadata(model)
cur_dat <- factors %>%
  rownames_to_column("sample") %>%
  full_join(clin, by = "sample")

p.list <- list()
for (i in names(factors)) {
  # for(j in c("Adenocarcinoma", "Squamous cell carcinoma")) {
    # cur_dat <- as.data.frame(traits)
    #cur_dat[,i] <- as.factor(cur_dat[,i])
    # cur_dat <- cur_dat %>% filter(!is.na(cur_dat[,i]),
             # simple_histology == j)
    surv_ob <- Surv(cur_dat$time_to_event,cur_dat$censoring)
    p.list[[i]] <- coxph( as.formula(paste0("surv_ob ~ ",i)),
                          data = cur_dat)
  # }
}

df <- data.frame(HR = unlist(lapply(p.list, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 #p.value = unlist(lapply(p.list, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 p.value = unlist(lapply(p.list, function(x){summary(x)$logtest["pvalue"]})),
                 lower_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "upper .95"]}))) 

df

```


## plot heatmap

### wide
```{r,fig.height=10, fig.width=15}
library(ComplexHeatmap)
library(circlize)
library(grid)
library(Cairo)
library(tidyr)

## ---- 1. significance labels --------------------------
# Create significance stars
results_annotated <- results %>%
  group_by(factor) %>% 
  mutate(
    p_adj = p.adjust(p_val, method = "fdr"),
    sig_label = case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE          ~ ""
    )
  ) %>%
  ungroup()

results_annotated <- results %>%
  group_by(factor) %>% 
  mutate(
    p_adj = p.adjust(p_val, method = "fdr"),
    sig_label = case_when(
      p_val < 0.001 & p_adj < 0.1 ~ "***",
      p_val < 0.01 & p_adj < 0.1  ~ "**",
      p_val < 0.05 & p_adj < 0.1  ~ "*",
      TRUE          ~ ""
    )
  ) %>%
  ungroup()

# Create estimate matrix for heatmap values
estimate_mat <- results_annotated %>%
  dplyr::select(factor, feature, estimate) %>%
  pivot_wider(names_from = factor, values_from = estimate) %>%
  column_to_rownames("feature") %>%
  as.matrix()

# Create annotation matrix for labels
label_mat <- results_annotated %>%
  dplyr::select(factor, feature, sig_label) %>%
  pivot_wider(names_from = factor, values_from = sig_label) %>%
  column_to_rownames("feature") %>%
  as.matrix()

## ---- 2. row renaming + group assignment --------------------------------
rename_rows <- c(
  "age_dicho" = "Young vs Old",
  "gender" = "Female vs Male",
  "stage_1_3" = "T: I vs III",
  "stage_1_2" = "T: I vs II",
  "stage_2_3" = "T: II vs III",
  "never_vs_current_smoker" = "Never vs Current",
  "never_vs_Ex_smoker" = "Never vs Ex",
  "current_vs_Ex_smoker" = "Current vs Ex",
  "long_survivors" = "Long vs Short",
  "N0_vs_Nx" = "N: N0 vs Nx",
  "M0_vs_Mx" = "M: M0 vs Mx",
  "T1_vs_Tother" = "T: T1 vs T2+",
  "LUAD_vs_LUSC" = "LUAD vs LUSC",
  "EGFR" = "EGFR", "TP53" = "TP53", "KRAS" = "KRAS", "PIK3CA" = "PIK3CA",
  "STK11" = "STK11", "KEAP1" = "KEAP1",
  "KRAS_only" = "KRAS_only", "KRAS_co" = "KRAS_co",
  "TPS" = "TPS", "CPS" = "CPS", "TMB" = "TMB",
  "T_density__IF1" = "T cell density",
  "mean_inf_ratio" = "CD8 infiltration"
)

## data-driven module blocks (names come through unrenamed)
img_mods <- colnames(meta_scores_df)[1:7]
rna_mods <- colnames(rna)[1:29]
inf_mods <- colnames(inf_scores)[1:11]

## hand-curated blocks (display names)
biomarker_feats <- c("TPS", "CPS", "TMB")
immune_feats    <- c("T cell density", "CD8 infiltration")
demogr_feats    <- c("Female vs Male", "Young vs Old")
clinical_feats  <- c("LUAD vs LUSC", "T: I vs II", "T: I vs III", "T: II vs III",
                     "T: T1 vs T2+", "N: N0 vs Nx", "M: M0 vs Mx")
smoking_feats   <- c("Never vs Current", "Current vs Ex", "Never vs Ex")
os_feats        <- "Long vs Short"
mutation_feats  <- c("EGFR", "KRAS", "KRAS_only", "KRAS_co",
                     "STK11", "KEAP1", "TP53", "PIK3CA")

## one place defining both order and grouping
block_defs <- list(
  "Img modules"  = img_mods,
  "RNA modules"  = rna_mods,
  "Infl. sign."  = inf_mods,
  "Biomarkers"   = biomarker_feats,
  "Immune"       = immune_feats,
  "Demogr."      = demogr_feats,
  "Clinical"     = clinical_feats,
  "Smoking"      = smoking_feats,
  "Mutations"    = mutation_feats,
  "OS"           = os_feats
)

custom_order   <- unlist(block_defs, use.names = FALSE)
feature_groups <- setNames(
  rep(names(block_defs), lengths(block_defs)),
  custom_order
)
group_levels <- names(block_defs)

## sanity check: anything tested but not placed will be silently dropped
tested <- unique(ifelse(results_annotated$feature %in% names(rename_rows),
                        rename_rows[results_annotated$feature],
                        results_annotated$feature))
missing <- setdiff(tested, custom_order)
if (length(missing)) warning("Not in custom_order, will be dropped: ",
                             paste(missing, collapse = ", "))

## ---- 3. helper: matrices for one `type` --------------------------------
make_mats <- function(dat, which_type, feat_order) {
  d <- dplyr::filter(dat, type == which_type)

  est <- d %>%
    dplyr::select(factor, feature, estimate) %>%
    pivot_wider(names_from = factor, values_from = estimate) %>%
    column_to_rownames("feature") %>% as.matrix()

  lab <- d %>%
    dplyr::select(factor, feature, sig_label) %>%
    pivot_wider(names_from = factor, values_from = sig_label) %>%
    column_to_rownames("feature") %>% as.matrix()

  rownames(est) <- ifelse(rownames(est) %in% names(rename_rows),
                          rename_rows[rownames(est)], rownames(est))
  rownames(lab) <- rownames(est)

  ord <- feat_order[feat_order %in% rownames(est)]
  list(
    est = t(est[ord, , drop = FALSE]),   # rows = factors, cols = features
    lab = t(lab[ord, , drop = FALSE]),
    grp = factor(feature_groups[ord], levels = group_levels)
  )
}

cont <- make_mats(results_annotated, "continuous", custom_order)
bin  <- make_mats(results_annotated, "binary",     custom_order)

# drop unused group levels so no empty slices are drawn
cont$grp <- droplevels(cont$grp)
bin$grp  <- droplevels(bin$grp)

n_fac <- nrow(cont$est)

## ---- 4. HR annotation (left of the first heatmap) ----------------------
df <- df[rownames(cont$est), , drop = FALSE]   # keep factor order in sync
hr_mat <- df %>% dplyr::select(-p.value) %>% as.matrix()
hr_ticks <- c(0.75, 1, 1.3)

hr_anno <- rowAnnotation(
  HR = anno_boxplot(
    hr_mat, box_width = 0.005, add_points = TRUE, pch = "|",
    axis_param = list(side = "bottom", at = hr_ticks,
                      labels = as.character(hr_ticks),
                      gp = gpar(fontsize = 5), facing = "outside",
                      labels_rot = 0)
  )
)

## ---- 5. two colour scales ---------------------------------------------
rho_max  <- max(abs(cont$est), na.rm = TRUE)
diff_max <- max(abs(bin$est),  na.rm = TRUE)

col_rho  <- colorRamp2(c(-rho_max, 0, rho_max),
                       c("#762a83", "white", "#1b7837"))   # correlation
col_diff <- colorRamp2(c(-diff_max, 0, diff_max),
                       c("blue", "white", "red"))          # mean difference

## ---- 6. cell_fun factory ----------------------------------------------
# NOTE: ComplexHeatmap passes (j = column, i = row)
make_cell_fun <- function(lab_mat) {
  function(j, i, x, y, width, height, fill) {
    grid.text(lab_mat[i, j], x, y, gp = gpar(fontsize = 7))
    grid.rect(x = x, y = y, width = width, height = height,
              gp = gpar(col = "black", fill = NA, lwd = 0.2))
  }
}

## ---- 7. the two heatmaps ----------------------------------------------
cell_w <- 12 / (ncol(cont$est) + ncol(bin$est))   # cm per column, shared
cell_w <- 0.25
ht_cont <- Heatmap(
  cont$est,
  name = "rho",
  col = col_rho,
  cluster_rows = FALSE, cluster_columns = FALSE,
  column_split = cont$grp,
  row_names_gp = gpar(fontsize = 8, fontface = "bold"),
  row_title = "MOFA Factors",
  row_title_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 7),
  column_title_gp = gpar(fontface = "bold", fontsize = 7),
  left_annotation = hr_anno,
  cell_fun = make_cell_fun(cont$lab),
  heatmap_legend_param = list(
    title = "Spearman \u03c1",
    title_gp = gpar(fontsize = 8, fontface = "bold"),
    labels_gp = gpar(fontsize = 7)
  ),
  width  = unit(cell_w * ncol(cont$est), "cm"),
  height = unit(4, "cm"),
  border = TRUE
)

ht_bin <- Heatmap(
  bin$est,
  name = "delta",
  col = col_diff,
  cluster_rows = FALSE, cluster_columns = FALSE,
  column_split = bin$grp,
  show_row_names = FALSE,          # row labels already on the left panel
  column_names_gp = gpar(fontsize = 7),
  column_title_gp = gpar(fontface = "bold", fontsize = 7),
  cell_fun = make_cell_fun(bin$lab),
  heatmap_legend_param = list(
    title = "\u0394Mean",
    title_gp = gpar(fontsize = 8, fontface = "bold"),
    labels_gp = gpar(fontsize = 7)
  ),
  width  = unit(cell_w * ncol(bin$est), "cm"),
  height = unit(4, "cm"),
  border = TRUE
)

ht_list <- ht_cont + ht_bin

## ---- 8. draw -----------------------------------------------------------
#CairoPDF(paste0(out_path, "/5D_clinical_x_factors_new.pdf"), width = 10, height = 5)

draw(ht_list,
     padding = unit(c(2, 2, 2, 2), "mm"),
     merge_legend = TRUE,
     ht_gap = unit(4, "mm"))

decorate_annotation("HR", {
  grid.lines(c(1, 1), c(0.5, n_fac + 0.5),
             gp = gpar(lty = 1, col = "grey50"), default.units = "native")
  for (x in hr_ticks) {
    grid.lines(c(x, x), c(0.5, n_fac + 0.5),
               gp = gpar(lty = 2, col = "grey50"), default.units = "native")
  }
  grid.points(hr_mat[, 1], n_fac:1,
              pch = 19, size = unit(2, "mm"),
              gp = gpar(col = ifelse(df$p.value < 0.05, "#D55E00", "grey30")),
              default.units = "native")
})

#dev.off()
```


### long
```{r,fig.height=10, fig.width=15}
library(ComplexHeatmap)
library(circlize)
library(grid)
library(Cairo)
library(tidyr)

## ---- 1. significance labels --------------------------
# Create significance stars
results_annotated_fdr <- results %>%
  group_by(factor) %>% 
  mutate(
    p_adj = p.adjust(p_val, method = "fdr"),
    sig_label = case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE          ~ ""
    )
  ) %>%
  ungroup()

results_annotated <- results %>%
  group_by(factor) %>% 
  mutate(
    p_adj = p.adjust(p_val, method = "fdr"),
    sig_label = case_when(
      p_val < 0.001 & p_adj < 0.1 ~ "***",
      p_val < 0.01 & p_adj < 0.1  ~ "**",
      p_val < 0.05 & p_adj < 0.1  ~ "*",
      TRUE          ~ ""
    )
  ) %>%
  ungroup()

# Create estimate matrix for heatmap values
estimate_mat <- results_annotated %>%
  dplyr::select(factor, feature, estimate) %>%
  pivot_wider(names_from = factor, values_from = estimate) %>%
  column_to_rownames("feature") %>%
  as.matrix()

# Create annotation matrix for labels
label_mat <- results_annotated %>%
  dplyr::select(factor, feature, sig_label) %>%
  pivot_wider(names_from = factor, values_from = sig_label) %>%
  column_to_rownames("feature") %>%
  as.matrix()

## ---- 2. row renaming + group assignment --------------------------------
rename_rows <- c(
  "age_dicho" = "Young vs Old",
  "gender" = "Female vs Male",
  "stage_1_3" = "T: I vs III",
  "stage_1_2" = "T: I vs II",
  "stage_2_3" = "T: II vs III",
  "never_vs_current_smoker" = "Never vs Current",
  "never_vs_Ex_smoker" = "Never vs Ex",
  "current_vs_Ex_smoker" = "Current vs Ex",
  "long_survivors" = "Long vs Short",
  "N0_vs_Nx" = "N: N0 vs Nx",
  "M0_vs_Mx" = "M: M0 vs Mx",
  "T1_vs_Tother" = "T: T1 vs T2+",
  "LUAD_vs_LUSC" = "LUAD vs LUSC",
  "EGFR" = "EGFR", "TP53" = "TP53", "KRAS" = "KRAS", "PIK3CA" = "PIK3CA",
  "STK11" = "STK11", "KEAP1" = "KEAP1",
  "KRAS_only" = "KRAS_only", "KRAS_co" = "KRAS_co",
  "TPS" = "TPS", "CPS" = "CPS", "TMB" = "TMB",
  "T_density__IF1" = "T cell density",
  "mean_inf_ratio" = "CD8 infiltration"
)

## data-driven module blocks (names come through unrenamed)
img_mods <- colnames(meta_scores_df)[1:7]
rna_mods <- colnames(rna)[1:29]
inf_mods <- colnames(inf_scores)[1:11]

## hand-curated blocks (display names)
biomarker_feats <- c("TPS", "CPS", "TMB")
immune_feats    <- c("T cell density", "CD8 infiltration")
demogr_feats    <- c("Female vs Male", "Young vs Old")
clinical_feats  <- c("LUAD vs LUSC", "T: I vs II", "T: I vs III", "T: II vs III",
                     "T: T1 vs T2+", "N: N0 vs Nx", "M: M0 vs Mx")
smoking_feats   <- c("Never vs Current", "Current vs Ex", "Never vs Ex")
os_feats        <- "Long vs Short"
mutation_feats  <- c("EGFR", "KRAS", "KRAS_only", "KRAS_co",
                     "STK11", "KEAP1", "TP53", "PIK3CA")

## one place defining both order and grouping
block_defs <- list(
  "Img modules"  = img_mods,
  "RNA modules"  = rna_mods,
  "Infl. sign."  = inf_mods,
  "Biomarkers"   = biomarker_feats,
  "Immune"       = immune_feats,
  "Demogr."      = demogr_feats,
  "Clinical"     = clinical_feats,
  "Smoking"      = smoking_feats,
  "Mutations"    = mutation_feats,
  "OS"           = os_feats
)

custom_order   <- unlist(block_defs, use.names = FALSE)
feature_groups <- setNames(
  rep(names(block_defs), lengths(block_defs)),
  custom_order
)
group_levels <- names(block_defs)

## sanity check: anything tested but not placed will be silently dropped
tested <- unique(ifelse(results_annotated$feature %in% names(rename_rows),
                        rename_rows[results_annotated$feature],
                        results_annotated$feature))
missing <- setdiff(tested, custom_order)
if (length(missing)) warning("Not in custom_order, will be dropped: ",
                             paste(missing, collapse = ", "))

## ---- 3. helper: matrices for one `type` --------------------------------
make_mats <- function(dat, which_type, feat_order) {
  d <- dplyr::filter(dat, type == which_type)

  est <- d %>%
    dplyr::select(factor, feature, estimate) %>%
    pivot_wider(names_from = factor, values_from = estimate) %>%
    column_to_rownames("feature") %>% as.matrix()

  lab <- d %>%
    dplyr::select(factor, feature, sig_label) %>%
    pivot_wider(names_from = factor, values_from = sig_label) %>%
    column_to_rownames("feature") %>% as.matrix()

  rownames(est) <- ifelse(rownames(est) %in% names(rename_rows),
                          rename_rows[rownames(est)], rownames(est))
  rownames(lab) <- rownames(est)

  ord <- feat_order[feat_order %in% rownames(est)]
  list(
    est = est[ord, , drop = FALSE],      # rows = features, cols = factors
    lab = lab[ord, , drop = FALSE],
    grp = factor(feature_groups[ord], levels = group_levels)
  )
}

cont <- make_mats(results_annotated, "continuous", custom_order)
bin  <- make_mats(results_annotated, "binary",     custom_order)

cont$grp <- droplevels(cont$grp)
bin$grp  <- droplevels(bin$grp)

n_fac <- ncol(cont$est)                  # factors are columns now

## ---- 4. HR annotation (above the first heatmap) ------------------------
df <- df[colnames(cont$est), , drop = FALSE]   # factor order = heatmap columns
hr_ticks <- c(0.75, 1, 1.5, 2)

# pull columns by name so nothing depends on df's column order
hr_est <- df$HR
hr_lo  <- df$lower_conf
hr_hi  <- df$upper_conf

hr_rg <- range(c(hr_lo, hr_hi, hr_ticks), na.rm = TRUE)
hr_rg <- hr_rg + c(-1, 1) * 0.05 * diff(hr_rg)

hr_anno <- HeatmapAnnotation(
  HR = anno_empty(border = TRUE, height = unit(1.5, "cm")),
  annotation_name_gp = gpar(fontsize = 7)
)

## ---- 5. two colour scales ---------------------------------------------
rho_max  <- max(abs(cont$est), na.rm = TRUE)
diff_max <- max(abs(bin$est),  na.rm = TRUE)

col_rho  <- colorRamp2(c(-rho_max, 0, rho_max),
                       c("#762a83", "white", "#1b7837"))   # correlation
col_diff <- colorRamp2(c(-diff_max, 0, diff_max),
                       c("blue", "white", "red"))          # mean difference

## ---- 6. cell_fun factory ----------------------------------------------
# NOTE: ComplexHeatmap passes (j = column, i = row)
make_cell_fun <- function(lab_mat) {
  function(j, i, x, y, width, height, fill) {
    grid.text(lab_mat[i, j], x, y, gp = gpar(fontsize = 7))
    grid.rect(x = x, y = y, width = width, height = height,
              gp = gpar(col = "black", fill = NA, lwd = 0.2))
  }
}

## ---- 7. the two heatmaps ----------------------------------------------
cell_h <- 0.28   # cm per feature row  — main knob for total height
cell_w <- 0.45   # cm per factor column

ht_cont <- Heatmap(
  cont$est,
  name = "rho",
  col = col_rho,
  cluster_rows = FALSE, cluster_columns = FALSE,
  row_split = cont$grp,
  row_title_rot = 0,
  row_title_gp = gpar(fontface = "bold", fontsize = 7),
  row_names_gp = gpar(fontsize = 6),
  show_column_names = FALSE,        # shown on the bottom panel
  top_annotation = hr_anno,
  cell_fun = make_cell_fun(cont$lab),
  heatmap_legend_param = list(
    title = "Spearman \u03c1",
    title_gp = gpar(fontsize = 8, fontface = "bold"),
    labels_gp = gpar(fontsize = 7)
  ),
  width  = unit(cell_w * ncol(cont$est), "cm"),
  height = unit(cell_h * nrow(cont$est), "cm"),
  border = TRUE
)

ht_bin <- Heatmap(
  bin$est,
  name = "delta",
  col = col_diff,
  cluster_rows = FALSE, cluster_columns = FALSE,
  row_split = bin$grp,
  row_title_rot = 0,
  row_title_gp = gpar(fontface = "bold", fontsize = 7),
  row_names_gp = gpar(fontsize = 6),
  column_names_gp = gpar(fontsize = 8, fontface = "bold"),
  cell_fun = make_cell_fun(bin$lab),
  heatmap_legend_param = list(
    title = "\u0394Mean",
    title_gp = gpar(fontsize = 8, fontface = "bold"),
    labels_gp = gpar(fontsize = 7)
  ),
  width  = unit(cell_w * ncol(bin$est), "cm"),
  height = unit(cell_h * nrow(bin$est), "cm"),
  border = TRUE
)

ht_list <- ht_cont %v% ht_bin

pdf_h <- 1.6 + (cell_h * (nrow(cont$est) + nrow(bin$est)) + 1.5) / 2.54
CairoPDF(paste0(out_path, "/5C_clinical_x_factors_long.pdf"),width = 4.5, height = pdf_h)


draw(ht_list,
     padding = unit(c(2, 2, 2, 2), "mm"),
     merge_legend = TRUE,
     ht_gap = unit(4, "mm"),
     column_title = "MOFA Factors",
     column_title_gp = gpar(fontsize = 9, fontface = "bold"))

decorate_annotation("HR", {
  pushViewport(viewport(xscale = c(0.5, n_fac + 0.5), yscale = hr_rg))

  grid.yaxis(at = hr_ticks, label = as.character(hr_ticks), gp = gpar(fontsize = 5))

  grid.lines(c(0.5, n_fac + 0.5), c(1, 1),
             gp = gpar(lty = 1, col = "grey50"), default.units = "native")
  for (y in setdiff(hr_ticks, 1)) {
    grid.lines(c(0.5, n_fac + 0.5), c(y, y),
               gp = gpar(lty = 2, col = "grey50"), default.units = "native")
  }

  # exactly one segment per factor
  grid.segments(x0 = seq_len(n_fac), y0 = hr_lo,
                x1 = seq_len(n_fac), y1 = hr_hi,
                default.units = "native", gp = gpar(col = "grey30", lwd = 1))

  grid.points(seq_len(n_fac), hr_est,
              pch = 19, size = unit(2, "mm"),
              gp = gpar(col = ifelse(df$p.value < 0.05, "#D55E00", "grey30")),
              default.units = "native")

  popViewport()
})

dev.off()
```
# 5D
## VALIDATIONS
### Fetch TCGA datasets
```{r}
library(TCGAbiolinks)
library(SummarizedExperiment)
if(!file.exists(paste0(mount_path,"/tcga_LUAD_LUSC.rds"))){
  # Query and download data
query <- GDCquery(
  project = c("TCGA-LUAD"),
  data.category = "Transcriptome Profiling",
  data.type = "Gene Expression Quantification",
  workflow.type = "STAR - Counts"
)
GDCdownload(query)
LUAD_data <- GDCprepare(query)

# Query and download data
query <- GDCquery(
  project = c("TCGA-LUSC"),
  data.category = "Transcriptome Profiling",
  data.type = "Gene Expression Quantification",
  workflow.type = "STAR - Counts"
)
GDCdownload(query)
LUSC_data <- GDCprepare(query)


# Unify data
LUSC_data$survival <- LUSC_data$vital_status
LUSC_data$histology <- "LUSC"
LUAD_data$survival <- LUAD_data$vital_status
LUAD_data$histology <- "LUAD"

table(LUAD_data$vital_status, LUAD_data$paper_Survival)
# table(is.na(LUAD_data$days_to_death), LUAD_data$survival)
# table(is.na(LUSC_data$days_to_death), LUSC_data$survival)

inters <- intersect(colnames(colData(LUAD_data)), colnames(colData(LUSC_data)))
colData(LUAD_data) <- colData(LUAD_data)[,inters]
colData(LUSC_data) <- colData(LUSC_data)[,inters]

sce <- cbind(LUAD_data, LUSC_data)

rownames(sce) <- rowData(sce)$gene_name  

# filter out deceased with no info on days to survival
sce <- sce[,!(sce$survival == "Dead" & is.na(sce$days_to_death))]

# replace - with . in some cases
rownames(sce)[startsWith(rownames(sce),"IGHV") |  startsWith(rownames(sce),"IGKV") |  startsWith(rownames(sce),"IGLV")] <- gsub("-", "\\.",rownames(sce)[startsWith(rownames(sce),"IGHV") |  startsWith(rownames(sce),"IGKV") |  startsWith(rownames(sce),"IGLV")])
# Save as .RDS (recommended for R objects)
saveRDS(sce, file = paste0(mount_path,"/tcga_LUAD_LUSC.rds"))
}
```

## Gene set enrichment for Combined FACTOR1 (Signature enrichment/in Combined) TPM

## load TCGA data
```{r}
library(SummarizedExperiment)
library(SingleCellExperiment)
library(singscore)
library(dplyr)
library(survival)
library(survminer)
# # Load TCGA RNA data
tcga <- readRDS(paste0(mount_path,"/tcga_LUAD_LUSC.rds"))


# Convert to SCE
sce <- as(tcga, "SingleCellExperiment")
sce <-  sce[,sce$tissue_type != "Normal"]

tpm_mat <- assay(tcga, "tpm_unstrand")
tpm_mat <- as.matrix(tpm_mat)   # in case it's a DelayedArray/sparse matrix

# if your gene set is symbols rather than Ensembl IDs, map via rowData
if ("gene_name" %in% colnames(rowData(tcga))) {
  symbols <- rowData(tcga)$gene_name
  keep <- !is.na(symbols) & !duplicated(symbols)   # drop NA + duplicate symbols
  tpm_mat <- tpm_mat[keep, ]
  rownames(tpm_mat) <- symbols[keep]
}

# 3. rank
ranked_data <- rankGenes(tpm_mat)
```


## helper function for survival in TCGA
```{r}
mofa_factor_survival <- function(model,
                                 sce,
                                 ranked_data,
                                 top_n         = 20,
                                 factors       = 1:6,
                                 view          = "RNA",
                                 min_set_size  = 3,
                                 histology_col = "histology",
                                 conf_level    = 0.95,
                                 p_adjust      = "BH",
                                 verbose       = TRUE) {

  stopifnot(is.numeric(top_n), top_n >= 2)

  ## -- signature from MOFA weights -------------------------------------------
  .signature <- function(f) {
    w <- get_weights(model, view = view, factors = f)[[view]][, 1]
    names(w) <- .demangle(names(w), rownames(ranked_data))
    w <- w[!is.na(names(w)) & names(w) %in% rownames(ranked_data)]   # scoring universe
    w <- w[names(w) %in% rownames(sce)]
    if (length(w) < top_n)
      stop("Only ", length(w), " scoreable features for factor ", f)
    top <- names(sort(abs(w), decreasing = TRUE))[seq_len(top_n)]
    list(up = top[w[top] > 0], down = top[w[top] < 0], weights = w[top])
  }
  
  .demangle <- function(x, universe) {
    key <- make.names(universe)
    dup <- key[duplicated(key)]
    if (length(dup)) warning(length(dup), " ambiguous name collisions")
    lut <- setNames(universe, key)
    ifelse(x %in% universe, x, unname(lut[x]))
  }

  ## -- singscore, named by sample --------------------------------------------
  .score <- function(sig, fname) {
    n_up <- length(sig$up); n_dn <- length(sig$down)

    if (n_up < min_set_size && n_dn < min_set_size) {
      warning(fname, ": both gene sets smaller than min_set_size; skipped")
      return(NULL)
    }
    if (n_up < min_set_size) {
      if (verbose) message(fname, ": only ", n_up,
                           " up genes - scoring the down set and negating")
      miss <- setdiff(sig$down, rownames(ranked_data))
      message(fname, ": ", length(miss), "/", length(sig$down),
              " down absent | rankData ", nrow(ranked_data), " rows, ",
              "md5 ", substr(digest::digest(rownames(ranked_data)), 1, 8))
      sc   <- simpleScore(ranked_data, upSet = sig$down)
      tot  <- -sc$TotalScore
      inv  <- TRUE
    } else if (n_dn < min_set_size) {
      sc  <- simpleScore(ranked_data, upSet = sig$up)
      tot <- sc$TotalScore
      inv <- FALSE
    } else {
      sc  <- simpleScore(ranked_data, upSet = sig$up, downSet = sig$down)
      tot <- sc$TotalScore
      inv <- FALSE
    }
    list(score = setNames(tot, rownames(sc)), inverted = inv)
  }

  ## -- clinical frame, aligned to sce by sample name -------------------------
  .surv_df <- function(score_vec) {
    ids <- colnames(sce)
    miss <- setdiff(ids, names(score_vec))
    if (length(miss))
      warning(length(miss), " sce samples absent from ranked_data; dropped")

    data.frame(
      ID        = ids,
      time      = ifelse(!is.na(sce$days_to_death),
                         sce$days_to_death, sce$days_to_last_follow_up),
      status    = as.integer(tolower(as.character(sce$survival)) %in%
                               c("dead", "deceased")),
      score     = as.numeric(score_vec[ids]),
      Histology = as.character(colData(sce)[[histology_col]]),
      stringsAsFactors = FALSE
    ) %>%
      filter(!is.na(time), !is.na(status), !is.na(score),
             !is.na(Histology), time > 0) %>%
      group_by(Histology) %>%
      mutate(
        score_z     = as.numeric(scale(score)),
        score_group = factor(ifelse(score_z >= median(score_z), "High", "Low"),
                             levels = c("Low", "High"))
      ) %>%
      ungroup()
  }

  ## -- one Cox fit -> one row ------------------------------------------------
  .fit_row <- function(d, rhs, fname, hist, model_label) {
    out <- data.frame(
      Factor = fname, Histology = hist, model = model_label,
      HR = NA_real_, lower = NA_real_, upper = NA_real_,
      p_wald = NA_real_, p_lr = NA_real_,
      n = nrow(d), n_event = sum(d$status),
      n_high = sum(d$score_group == "High"),
      n_low  = sum(d$score_group == "Low"),
      stringsAsFactors = FALSE
    )
    if (sum(d$status) < 3) return(out)
    if (rhs == "score_group" && nlevels(droplevels(d$score_group)) < 2) return(out)

    fit <- tryCatch(
      coxph(as.formula(paste("Surv(time, status) ~", rhs)), data = d),
      error = function(e) NULL, warning = function(w) NULL
    )
    if (is.null(fit)) return(out)

    s <- summary(fit, conf.int = conf_level)
    out$HR     <- s$conf.int[1, "exp(coef)"]
    out$lower  <- s$conf.int[1, 3]          # positional: colname tracks conf_level
    out$upper  <- s$conf.int[1, 4]
    out$p_wald <- s$coefficients[1, "Pr(>|z|)"]
    out$p_lr   <- unname(s$logtest["pvalue"])   # univariable -> LR for this term
    out
  }

  ## -- loop ------------------------------------------------------------------
  gene_df <- list(); hr_group <- list(); hr_cont <- list(); fits <- list()

  for (f in factors) {
    fname <- paste0("Factor", f)
    sig   <- .signature(f)

    gene_df[[fname]] <- data.frame(
      Factor = fname,
      gene   = names(sig$weights),
      weight = as.numeric(sig$weights),
      direction = ifelse(sig$weights > 0, "up", "down"),
      stringsAsFactors = FALSE
    )

    sc <- .score(sig, fname)
    if (is.null(sc)) next
    sdf <- .surv_df(sc$score)
    if (sc$inverted) sdf$score_z <- sdf$score_z   # already negated upstream

    for (h in sort(unique(sdf$Histology))) {
      d <- filter(sdf, Histology == h)
      hr_group[[paste(fname, h)]] <- .fit_row(d, "score_group", fname, h, "High vs Low")
      hr_cont[[paste(fname, h)]]  <- .fit_row(d, "score_z",     fname, h, "per SD")
      fits[[paste(fname, h)]] <- tryCatch(
        coxph(Surv(time, status) ~ score_z, data = d), error = function(e) NULL)
    }

    if (verbose) message(fname, ": ", length(sig$up), " up / ",
                         length(sig$down), " down, n = ", nrow(sdf),
                         ", events = ", sum(sdf$status))
  }

  .bind <- function(l) {
    r <- bind_rows(l)
    if (!nrow(r)) return(r)
    r %>% group_by(Histology) %>%
      mutate(p_lr_adj = p.adjust(p_lr, method = p_adjust)) %>%
      ungroup() %>%
      mutate(Factor = factor(Factor, levels = paste0("Factor", factors)))
  }
  hr_group <- .bind(hr_group)
  hr_cont  <- .bind(hr_cont)

  ## -- forest plot -----------------------------------------------------------
  plot_hr <- function(tab, subtitle) {
    d <- filter(tab, !is.na(HR), is.finite(lower), is.finite(upper))
    if (!nrow(d)) return(NULL)
    ggplot(d, aes(x = HR, y = Factor, colour = Histology)) +
      geom_vline(xintercept = 1, linetype = "dashed", colour = "grey45") +
      geom_pointrange(aes(xmin = lower, xmax = upper),
                      position = position_dodge(width = 0.55),
                      size = 0.4, fatten = 2.5) +
      scale_x_log10() +
      scale_y_discrete(limits = rev) +
      labs(x = paste0("Hazard ratio (", round(conf_level * 100), "% CI, log scale)"),
           y = NULL, colour = "Histology", subtitle = subtitle) +
      theme_bw(base_size = 11) +
      theme(panel.grid.minor = element_blank())
  }

  list(
    genes         = bind_rows(gene_df),
    hr_group      = hr_group,
    hr_continuous = hr_cont,
    plot_group    = plot_hr(hr_group, "Median split: High vs Low, within histology"),
    plot_cont     = plot_hr(hr_cont,  "Continuous score: HR per SD, within histology"),
    models        = fits,
    call          = match.call()
  )
}
```

## helper function to plot survival for factors
```{r}
## ---------------------------------------------------------------------------
## plot_factor_hr_heatmap()
##
## Gene-weight heatmap for one MOFA factor with a linked HR forest panel,
## consuming the output of mofa_factor_survival(). LUAD / LUSC only.
##
## The per-histology CIs come straight from the separate Cox fits, so the
## delta-method block the old script needed for LUSC is gone.
## ---------------------------------------------------------------------------

suppressPackageStartupMessages({
  library(ComplexHeatmap)
  library(grid)
  library(dplyr)
})

plot_factor_hr_heatmap <- function(res,
                                   factor_id,
                                   hr_table        = c("continuous", "group"),
                                   p_col           = c("p_lr", "p_wald", "p_lr_adj"),
                                   histology_order = c("LUSC", "LUAD"),
                                   show_top        = NULL,
                                   sig_cutoff      = 0.05,
                                   x_labels        = NULL,
                                   gene_fontsize   = 8,
                                   heatmap_height  = NULL,
                                   draw_it         = TRUE) {

  hr_table <- match.arg(hr_table)
  p_col    <- match.arg(p_col)

  ## -- HR panel data ---------------------------------------------------------
  tab <- if (hr_table == "continuous") res$hr_continuous else res$hr_group
  if (!p_col %in% names(tab)) stop("`", p_col, "` not in res$hr_", hr_table)

  hr_df <- tab %>%
    filter(Factor == factor_id, Histology %in% histology_order) %>%
    mutate(Histology = factor(as.character(Histology), levels = histology_order)) %>%
    arrange(Histology) %>%                    # row 1 renders at the BOTTOM
    as.data.frame()

  if (!nrow(hr_df)) stop("No HR rows for ", factor_id)
  if (anyNA(hr_df$HR))
    warning("NA hazard ratio for: ",
            paste(hr_df$Histology[is.na(hr_df$HR)], collapse = ", "))
  if (any(!is.finite(c(hr_df$lower, hr_df$upper))))
    warning("Non-finite CI bound - likely separation in one arm")

  hr_df$p     <- hr_df[[p_col]]
  hr_df$label <- paste0("Score effect ", as.character(hr_df$Histology))
  temp        <- as.matrix(hr_df[, c("HR", "lower", "upper")])

  ## -- gene weight matrix ----------------------------------------------------
  plot_df <- res$genes %>%
    filter(Factor == factor_id) %>%
    as.data.frame()

  if (!is.null(show_top) && show_top < nrow(plot_df)) {
    keep    <- order(abs(plot_df$weight), decreasing = TRUE)[seq_len(show_top)]
    plot_df <- plot_df[keep, ]
  }
  plot_df <- plot_df[order(plot_df$weight), ]

  mat <- matrix(ifelse(plot_df$weight > 0, 2, -2), ncol = 1,
                dimnames = list(plot_df$gene, factor_id))

  col_fun <- c("2" = "chartreuse4", "-2" = "darkorchid")

  ## -- x axis ----------------------------------------------------------------
  if (is.null(x_labels)) {
    rng      <- range(c(temp, 1), na.rm = TRUE, finite = TRUE)
    x_labels <- pretty(rng, n = 6)
  }
  pad <- 0.05 * diff(range(x_labels))
  rg  <- c(min(x_labels) - pad, max(x_labels) + pad)

  ## -- forest panel ----------------------------------------------------------
  panel_fun <- function(index, nm) {
    pushViewport(viewport(xscale = rg, yscale = c(0.5, nrow(temp) + 0.5)))
    grid.rect()
    grid.xaxis(at = x_labels, label = x_labels, gp = gpar(fontsize = 8))
    grid.yaxis(at = seq_len(nrow(temp)), label = hr_df$label,
               main = TRUE, gp = gpar(fontsize = 8))
    grid.yaxis(at = seq_len(nrow(temp)),
               label = paste0(round(hr_df$HR, 2), "\n(",
                              round(hr_df$lower, 2), " - ",
                              round(hr_df$upper, 2), "), p=",
                              signif(hr_df$p, 2)),
               main = FALSE, gp = gpar(fontsize = 7))

    for (i in seq_len(nrow(temp)))
      grid.boxplot(temp[i, ], pch = 2, size = unit(40, "mm"), pos = i,
                   direction = "horizontal", box_width = 0.00005)

    grid.lines(c(1, 1), c(0.5, nrow(temp) + 0.5),
               gp = gpar(lty = 1, col = "grey50"), default.units = "native")
    for (x in x_labels)
      grid.lines(c(x, x), c(0.5, nrow(temp) + 0.5),
                 gp = gpar(lty = 2, col = "grey50"), default.units = "native")

    grid.points(temp[, 1], seq_len(nrow(temp)), pch = 19, size = unit(2, "mm"),
                gp = gpar(col = ifelse(!is.na(hr_df$p) & hr_df$p < sig_cutoff,
                                       "#D55E00", "grey30")),
                default.units = "native")
    popViewport()
  }

  anno <- anno_zoom(align_to = rep("a", nrow(mat)), which = "row",
                    panel_fun = panel_fun,
                    size = unit(1.2 * nrow(temp), "cm"),
                    width = unit(10, "cm"), link_width = unit(3, "cm"))

  ## -- heatmap ---------------------------------------------------------------
  if (is.null(heatmap_height))
    heatmap_height <- unit(max(4, 0.32 * nrow(mat)), "cm")

  ht <- Heatmap(
    mat,
    name = paste0(factor_id, " weight"),
    heatmap_legend_param = list(labels = c("Depleted", "Enriched"), at = c(-2, 2)),
    right_annotation = rowAnnotation(HR = anno, gap = unit(0, "points")),
    col = col_fun,
    row_names_side   = "left",
    row_names_gp     = gpar(fontsize = gene_fontsize),
    row_names_max_width = unit(8, "cm"),
    cluster_rows     = FALSE,
    cluster_columns  = FALSE,
    show_column_names = FALSE,
    width  = unit(0.4, "cm"),
    height = heatmap_height,
    border = TRUE,
    cell_fun = function(j, i, x, y, width, height, fill) {
      grid.rect(x = x, y = y, width = width, height = height,
                gp = gpar(col = "black", fill = NA, lwd = 0.5))
    }
  )

  if (draw_it)
    draw(ht, padding = unit(c(10, 2, 20, 20), "mm"), heatmap_legend_side = "bottom")

  invisible(list(heatmap = ht, hr = hr_df, mat = mat))
}

```

## plot heatmap Factor 1
```{r}

## ---------------------------------------------------------------------------
res <- mofa_factor_survival(model, sce, ranked_data, top_n = 20)
#
plot_factor_hr_heatmap(res, "Factor1", show_top = 20)

pdf(paste0(out_path, "5D_TCGA_validation_Factor1.pdf"), width = 6.5, height = 3.5)  # adjust width/height as needed
plot_factor_hr_heatmap(res, "Factor1", show_top = 20)
dev.off()
```

## plot heatmap Factor 4
```{r}

## ---------------------------------------------------------------------------
res <- mofa_factor_survival(model, sce, ranked_data, top_n = 20)
#
plot_factor_hr_heatmap(res, "Factor4", show_top = 20)

pdf("/mnt/immucan_volume/processed_data/NSCLC2_cohort/revised_figures/Figure_7_supp/S7F_TCGA_validation_Factor4.pdf", width = 6.5, height = 3.5)  # adjust width/height as needed
plot_factor_hr_heatmap(res, "Factor4", show_top = 20)
dev.off()

res$hr_continuous
```

# interaction term
```{r}
library(MOFA2); library(singscore); library(survival); library(dplyr)

f <- 1
top_n <- 20

# --- signature ---
w <- get_weights(model, view = "RNA", factors = f)$RNA[, 1]
w <- w[names(w) %in% rownames(ranked_data) & names(w) %in% rownames(sce)]
top <- names(sort(abs(w), decreasing = TRUE))[seq_len(top_n)]
up <- top[w[top] > 0]; dn <- top[w[top] < 0]

sc <- if (length(up) < 3) {
  setNames(-simpleScore(ranked_data, upSet = dn)$TotalScore,
           rownames(simpleScore(ranked_data, upSet = dn)))
} else if (length(dn) < 3) {
  s <- simpleScore(ranked_data, upSet = up); setNames(s$TotalScore, rownames(s))
} else {
  s <- simpleScore(ranked_data, upSet = up, downSet = dn); setNames(s$TotalScore, rownames(s))
}

# --- clinical frame ---
ids <- colnames(sce)
d <- data.frame(
  time      = ifelse(!is.na(sce$days_to_death), sce$days_to_death, sce$days_to_last_follow_up),
  status    = as.integer(tolower(as.character(sce$survival)) %in% c("dead", "deceased")),
  score     = as.numeric(sc[ids]),
  Histology = as.character(colData(sce)$histology)
) %>%
  filter(!is.na(time), !is.na(status), !is.na(score), !is.na(Histology), time > 0) %>%
  mutate(score_z = as.numeric(scale(score)),          # one scale, whole cohort
         Histology = relevel(factor(Histology), ref = "LUAD"))

# --- interaction test ---
fit_int <- coxph(Surv(time, status) ~ score_z * strata(Histology), data = d)
fit_add <- coxph(Surv(time, status) ~ score_z + strata(Histology), data = d)

anova(fit_add, fit_int)      # <- LR p value for the manuscript

# --- per-histology HRs and ratio of HRs ---
b <- coef(fit_int); V <- vcov(fit_int)
se_lusc <- sqrt(V[1,1] + V[2,2] + 2*V[1,2])

data.frame(
  term  = c("HR per SD, LUAD", "HR per SD, LUSC", "Ratio of HRs, LUSC vs LUAD"),
  est   = exp(c(b[1], b[1]+b[2], b[2])),
  lower = exp(c(b[1] - 1.96*sqrt(V[1,1]), b[1]+b[2] - 1.96*se_lusc, b[2] - 1.96*sqrt(V[2,2]))),
  upper = exp(c(b[1] + 1.96*sqrt(V[1,1]), b[1]+b[2] + 1.96*se_lusc, b[2] + 1.96*sqrt(V[2,2])))
)
```



