---
title: "Figure 5 - MOFA"
output: html_document
date: '2023-03-15'
---

```{r}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

# Set path
if (Sys.info()["sysname"] == "Darwin"){
  mount_path <- "/Volumes/G_DQBM_BB_Central$/projects/immucan/"
} else if (Sys.info()["sysname"] == "Linux"){
  mount_path <- "/mnt/immucan_volume/"
} else if (Sys.info()["sysname"] == "Windows"){
  mount_path <- "Y:/"
} else {
  stop("Operating system is not supported")
}

options(bitmapType = "cairo")
```


load libraries
```{r}
library(ggplot2)
library(reticulate)
library(MOFA2)
library(dplyr)
library(tidyverse)

```

## load model
```{r}
mount_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/MOFA"
out_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/revised_figures/Figure_5/"


model_path <- paste0(mount_path, '/mofa_V6_minor.rds')

if (!dir.exists(out_path)){
  dir.create(out_path)
  dir.create(paste0(out_path, "/extras"))
}
model <- readRDS(model_path)
```

## flip factors for explorations
note, we are using the original rotations of the model
```{r, eval=FALSE}
# ## flip factor values
# model@expectations$Z$group1[,"Factor1"] <- -model@expectations$Z$group1[,"Factor1"]
# # flip feature weights
# model@expectations$W$Densities[,"Factor1"] <- -model@expectations$W$Densities[,"Factor1"]
# model@expectations$W$Expressions[,"Factor1"] <- -model@expectations$W$Expressions[,"Factor1"]
# model@expectations$W$RNA[,"Factor1"] <- -model@expectations$W$RNA[,"Factor1"]
# model@expectations$W$BINARY[,"Factor1"] <- -model@expectations$W$BINARY[,"Factor1"]
# model@expectations$W$Cellohood[,"Factor1"] <- -model@expectations$W$Cellohood[,"Factor1"]
# model@expectations$W$Interactions[,"Factor1"] <- -model@expectations$W$Interactions[,"Factor1"]
# 
# saveRDS(model, model_path)
```

## Overview
```{r}
plot_data_overview(model)
```


# 5B Variation explained
```{r, fig.height=6, fig.width=15}
library(MOFA2)
library(ComplexHeatmap)
library(circlize)
library(dplyr)
library(RColorBrewer)
library(grid)

# Step 1: Get variance explained per view & factor
var_df <- get_variance_explained(model, as.data.frame = TRUE)$r2_per_factor

# Step 2: Matrix with factors as rows, views as columns
var_mat <- var_df %>%
  dplyr::select(view, factor, value) %>%
  tidyr::pivot_wider(names_from = view, values_from = value, values_fill = 0) %>%
  tibble::column_to_rownames("factor") %>%
  as.matrix()

# Step 3: Compute row and column sums
row_sums <- rowSums(var_mat)
col_sums <- colSums(var_mat)
# Reorder columns (views) by total variance explained
sorted_view_order <- names(sort(col_sums, decreasing = TRUE))
var_mat <- var_mat[, sorted_view_order]
col_sums <- col_sums[sorted_view_order]
views <- sorted_view_order


# Step 4: Color definitions
green_scale <- colorRamp2(
  c(0, max(var_mat)),
  c("white", "#238b45")  # white → strong green
)
blue_scale <- colorRamp2(c(0, max(c(row_sums, col_sums))), c("white", "blue"))

# View color palette (matching other plots)
view_colors <- structure(
  RColorBrewer::brewer.pal(length(views), "Set1"),
  names = sort(views)
)

# Step 5: Annotations with text labels
row_anno <- rowAnnotation(
  `Total %` = anno_barplot(row_sums, gp = gpar(fill = blue_scale(row_sums))),
  total_label = anno_text(sprintf("%.1f", row_sums), location = 0, just = "left", gp = gpar(fontsize = 8, fontface = "bold")),
  show_annotation_name = FALSE
)

col_anno <- HeatmapAnnotation(
  `Total %` = anno_barplot(col_sums, gp = gpar(fill = blue_scale(col_sums))),
  total_label = anno_text(sprintf("%.1f", col_sums), 
                          location = 0.5, 
                          just = "top", 
                          rot = 0,                       # ✅ Rotation goes here
                          gp = gpar(fontsize = 8, fontface = "bold")),
  annotation_name_side = "left",
  show_annotation_name = TRUE,
  annotation_name_rot = 45
)


# Custom annotation grob (as a gTree)
custom_view_grob <- function(views, view_colors) {
  n <- length(views)
  grid::gTree(children = do.call(gList, lapply(seq_along(views), function(i) {
    view <- views[i]
    vp <- viewport(
      x = unit(i - 0.5, "native"),
      y = unit(0.5, "npc"),
      width = unit(1, "native"),
      height = unit(1, "npc"),
      just = c("center", "center")
    )
    gList(
      rectGrob(
        gp = gpar(fill = view_colors[view], col = NA),
        vp = vp
      ),
      textGrob(
        label = view,
        gp = gpar(col = "black", fontsize = 8, fontface = "bold"),
        rot = 90,  
        vp = vp
      )
    )
  })))
}

# HeatmapAnnotation using annotation_function
col_anno_bottom <- HeatmapAnnotation(
  View = function(index) {
    pushViewport(viewport(xscale = c(0, length(views))))
    grid.draw(custom_view_grob(views, view_colors))
    popViewport()
  },
  annotation_height = unit(6, "mm"),
  show_annotation_name = FALSE
)

# Step 6: Final heatmap
ht_opt$ROW_ANNO_PADDING <- unit(0.5, "cm")
ht_opt$COLUMN_ANNO_PADDING <- unit(0.5, "cm")  # Optional
view_order <- colnames(var_mat)

ht <- Heatmap(
  var_mat,
  name = "% Variance",
  col = green_scale,
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  row_names_side = "left",
  column_names_side = "bottom",
  left_annotation = row_anno,
  top_annotation = col_anno,
  bottom_annotation = col_anno_bottom,
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(sprintf("%.1f", var_mat[i, j]), x, y, gp = gpar(fontsize = 9))
  },
  heatmap_legend_param = list(
    title = "Variance\nExplained [%]",
    title_gp = gpar(fontsize = 8, fontface = "bold")
  ),
  show_column_names = FALSE
)

# Set PDF output
pdf(paste0(out_path, "/5B_variance_explained.pdf"), width = 6, height = 4)  # adjust width/height as needed


# Draw the heatmap
draw(ht, padding = unit(c(6, 2, 2, 8), "mm"))

# Close the PDF device
dev.off()
```

# 5C Factor correlations
## calculate per patient inflammation
### get gene signatures
```{r}
library(org.Hs.eg.db)
library(AnnotationDbi)
library(msigdbr)
library(singscore)

rna_mat <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/rna_mat_tpm.rds")


# ── gene signatures ────────────────────────────────────────────────────
hallmark_sets <- msigdbr(species = "Homo sapiens", collection = "H") %>%
  dplyr::select(gs_name, gene_symbol) %>%
  split(.$gs_name) %>%
  lapply(function(x) x$gene_symbol)

# Targeted Hallmark gene sets of interest
target_sets <- c(
  "HALLMARK_INFLAMMATORY_RESPONSE",
  "HALLMARK_INTERFERON_GAMMA_RESPONSE",
  "HALLMARK_INTERFERON_ALPHA_RESPONSE",
  "HALLMARK_TNFA_SIGNALING_VIA_NFKB",
  "HALLMARK_IL6_JAK_STAT3_SIGNALING",
  "HALLMARK_IL2_STAT5_SIGNALING"
)
gs_list_hallmark <- hallmark_sets[intersect(target_sets, names(hallmark_sets))]

# ── 2. Custom gene sets ────────────────────────────────────────────────────────
# CXCL9/10/11 chemokine axis (T cell recruitment)
gs_list_custom <- list(
  CXCL9_10_11_axis = c("CXCL9", "CXCL10", "CXCL11", "CXCR3",
                       "CXCL13", "CXCR5"),
  T_cell_activation = c("CD8A", "CD8B", "GZMB", "PRF1", "IFNG",
                        "TNF", "IL2", "PDCD1", "HAVCR2", "LAG3"),
  type_I_IFN = c("IFNB1", "IRF7", "MX1", "OAS1", "OAS2", "OAS3",
                 "ISG15", "IFIT1", "IFIT2", "IFIT3", "RSAD2", "HERC5"),
  NF_kB_target  = c("NFKB1", "RELA", "RELB", "NFKB2", "REL",
                    "BCL2L1", "IL1B", "IL6", "CXCL8", "TNF", "ICAM1"),
  TCR_signaling = c("CD3D", "CD3E", "CD3G", "ZAP70", "LCK", "LAT",
                    "ITK", "PLCG1", "NFATC1", "NFATC2"),
  cytotoxic = c("GZMB", "GZMA", "GZMK", "PRF1", "NKG7", "IFNG", "KLRG1"),
  exhaustion = c("PDCD1","LAG3","HAVCR2","TX","CTLA4", "IDO1")
  )

all_gs <- c(gs_list_hallmark, gs_list_custom)

# compute score
rna_for_score <- t(rna_mat)

# Filter gene sets to genes present in the RNA-seq matrix
all_gs <- lapply(all_gs, function(g) intersect(g, rownames(rna_for_score)))
all_gs <- all_gs[sapply(all_gs, length) >= 4]



overlap <- sapply(all_gs, function(g) sum(g %in% rownames(rna_for_score)))
total   <- sapply(all_gs, length)
print(data.frame(found = overlap, total = total, pct = round(100 * overlap / total)))

# ── Score with singscore ──────────────────────────────────────────────────────
ranked <- rankGenes(rna_for_score)

scores_list <- lapply(names(all_gs), function(nm) {
  sc <- simpleScore(ranked, upSet = all_gs[[nm]], centerScore = TRUE)
  setNames(sc$TotalScore, rownames(sc))
})


inf_scores <- as.data.frame(do.call(cbind, scores_list))
colnames(inf_scores) <- names(all_gs)
rownames(inf_scores) <- colnames(rna_for_score)
inf_scores$immucan_id <- rownames(inf_scores)
rownames(inf_scores) <- NULL

```
## get infiltration score
```{r}
inf_dat <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/infiltration_harmonized.rds")

# compare densities of infiltrating, bystanding and ignored from IMC1, IMC2 and mIF
inf_dat <- inf_dat %>%
  group_by(immucan_id,tech) %>%
  mutate(total_CD8 = sum(bystanding,ignored,infiltrating,na.rm = TRUE)/im_size,
         bystanding = round(bystanding/im_size),
         ignored = round(ignored/im_size),
         infiltrating = round(infiltrating/im_size)) %>%
  ungroup()

# define interactor over bystander ratio
inf_dat <- inf_dat %>%
  group_by(immucan_id, tech) %>%
  mutate(int_over_bystander = (infiltrating+bystanding)/ignored)

inf_dat[(is.infinite(inf_dat$int_over_bystander)),]$int_over_bystander <- NA

# we will average the ratios across technologies
inf_dat <- inf_dat %>%
  group_by(immucan_id) %>%
  mutate(mean_inf_ratio = mean(int_over_bystander,na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(mean_inf_ratio_dicho = ntile(mean_inf_ratio,2),
         mean_inf_ratio_dicho = ifelse(mean_inf_ratio_dicho == 2, "high", "low"))%>%
  dplyr::select(immucan_id,mean_inf_ratio) %>%
  distinct() %>%
  as.data.frame()
inf_dat
```


## load other data

```{r}
mofa <- readRDS( '/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/MOFA/mofa_V6_minor.rds')
meta_scores <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/patient_imaging_meta_cluster_scores_tpm.rds")
rna <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/rna_scores_tpm.rds")
traits <- read_csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/all_traits.csv")


# preprocess mofa
lf_raw <- as.data.frame(get_factors(mofa)[["group1"]])   # or [[1]]

lf_raw <- lf_raw %>%
  mutate(immucan_id = rownames(lf_raw),
         immucan_id = gsub("_","-",immucan_id))

# preprocess imaging modules
meta_scores_df <- as.data.frame(meta_scores)
meta_scores_df$immucan_id <- rownames(meta_scores_df)
rownames(meta_scores_df) <- NULL


# preprocess RNA modules
rna <- as.data.frame(rna)
rna$immucan_id <- rownames(rna)
rownames(rna) <- NULL
colnames(rna) <- gsub(" ", "_", colnames(rna))
colnames(rna) <- gsub("-", "_", colnames(rna))

# cur_traits
cur_traits <- traits %>%
  dplyr::select(immucan_id,age_dicho, gender,stage_1_2,stage_1_3,stage_2_3,LUAD_vs_LUSC,never_vs_current_smoker,        never_vs_Ex_smoker,current_vs_Ex_smoker, N0_vs_Nx,M0_vs_Mx, long_survivors,T1_vs_Tother,EGFR,TP53,KRAS, PIK3CA,STK11,KEAP1, STK11,TPS, CPS, TMB ) %>%
  mutate(
    KRAS_only          = ifelse(KRAS == 1 & STK11 == 0 & KEAP1 == 0, "TRUE", "FALSE"),
    KRAS_co = ifelse(KRAS == 1 & (STK11 == 1 | KEAP1 == 1),"TRUE","FALSE")
  )

#IF1 T cell density
dens_dat <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/densities_dat.rds")
T1_density <- dens_dat %>%
  dplyr::select(immucan_id, T_density__IF1)

# merge
df_all <- plyr::join_all(dfs = list(lf_raw, meta_scores_df, rna, inf_scores, inf_dat, cur_traits, T1_density),by = "immucan_id",type = "left")
```

## compute correlation
```{r}
cur_traits <- c("age_dicho", "gender","stage_1_2","stage_1_3","stage_2_3","LUAD_vs_LUSC","never_vs_current_smoker",
            "never_vs_Ex_smoker","current_vs_Ex_smoker", "N0_vs_Nx","M0_vs_Mx", "long_survivors","T1_vs_Tother", "EGFR","TP53","PIK3CA","KRAS","KRAS_only","KRAS_co","KEAP1","STK11", "TPS","CPS","TMB","T_density__IF1","mean_inf_ratio")
condition <- c("low", "Female","I","I","II","Adenocarcinoma","Non-smoker","Non-smoker","Current smoker", "node_neg", "met_neg", "alive","T1","TRUE","TRUE","TRUE","TRUE","TRUE", "TRUE", "TRUE", "TRUE","numeric","numeric","numeric","numeric","numeric")
names(condition) <- cur_traits

factors <- get_factors(model, as.data.frame = TRUE) %>%
  dplyr::select(- group) %>%
  pivot_wider(names_from = "factor", values_from = value) %>%
  column_to_rownames("sample")

test_vars <- colnames(df_all)[8:ncol(df_all)]
lf_vars <- colnames(lf_raw)[1:6]

results <- lapply(test_vars, function(clin_var) {
  values_raw <- df_all[[clin_var]]
  print(clin_var)
  # Try coercion to numeric
  values_num <- suppressWarnings(as.numeric(as.character(values_raw)))
  num_valid <- sum(!is.na(values_num))

  per_feat <- lapply(lf_vars, function(fac) {
    x <- df_all[[fac]]
    # 1. Treat as numeric if coercion worked well (e.g., >80% non-NA)
    if (num_valid > 0.6 * length(values_raw)) {
      temp <- cor.test(x, values_num, method = "spearman")
      p_val <- temp$p.value
      estimate <- temp$estimate
      
      return(data.frame(estimate = estimate, p_val = p_val, type = "continuous"))
    }
    
    # 2. Treat as binary if exactly 2 unique non-NA values
    values_fac <- as.factor(values_raw)
    if (nlevels(values_fac) == 2) {
      values <- values_raw == condition[clin_var]
      if (table(values)[[1]] >= 2 & table(values)[[2]] >= 2){
        temp <- t.test(x ~ values)
        diff_mean <- diff(temp$estimate)
        return(data.frame(estimate = diff_mean, p_val = temp$p.value, type = "binary"))
      }else{
        return(data.frame(estimate = NA, p_val = NA, type = "binary"))
      }
    }
    
    # 3. Treat as categorical if ≥3 levels
    if (is.factor(values_fac) || is.character(values_raw)) {
      # Skip test if x has no variance
      if (length(unique(na.omit(values_fac))) <= 1) {
        return(data.frame(estimate = NA, p_val = NA, type = "categorical"))
      }
      
      temp <- kruskal.test(x, values_fac)
      group_means <- tapply(x, values_fac, mean, na.rm = TRUE)
      effect_estimate <- max(group_means, na.rm = TRUE) - min(group_means, na.rm = TRUE)
      
      return(data.frame(estimate = effect_estimate, p_val = temp$p.value, type = "categorical"))
    }
    
    # 4. Fallback
    return(NA)
  })
  
  temp <- do.call(rbind, per_feat)
  temp$factor <- colnames(factors)
  temp$feature <- clin_var
  
  return (temp)
})

results <- do.call(rbind, results)


## survival analysis
library(survminer)
library(survivalAnalysis)
library(survival)

factors <- get_factors(model, as.data.frame = TRUE) %>%
  dplyr::select(- group) %>%
  pivot_wider(names_from = "factor", values_from = value) %>%
  column_to_rownames("sample")

clin <- samples_metadata(model)
cur_dat <- factors %>%
  rownames_to_column("sample") %>%
  full_join(clin, by = "sample")

p.list <- list()
for (i in names(factors)) {
  # for(j in c("Adenocarcinoma", "Squamous cell carcinoma")) {
    # cur_dat <- as.data.frame(traits)
    #cur_dat[,i] <- as.factor(cur_dat[,i])
    # cur_dat <- cur_dat %>% filter(!is.na(cur_dat[,i]),
             # simple_histology == j)
    surv_ob <- Surv(cur_dat$time_to_event,cur_dat$censoring)
    p.list[[i]] <- coxph( as.formula(paste0("surv_ob ~ ",i)),
                          data = cur_dat)
  # }
}

df <- data.frame(HR = unlist(lapply(p.list, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p.list, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "upper .95"]}))) 

```


## plot heatmap

### wide
```{r,fig.height=10, fig.width=15}
library(ComplexHeatmap)
library(circlize)
library(grid)
library(Cairo)
library(tidyr)

## ---- 1. significance labels --------------------------
# Create significance stars
results_annotated <- results %>%
  group_by(factor) %>% 
  mutate(
    p_adj = p.adjust(p_val, method = "fdr"),
    sig_label = case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE          ~ ""
    )
  ) %>%
  ungroup()

results_annotated <- results %>%
  group_by(factor) %>% 
  mutate(
    p_adj = p.adjust(p_val, method = "fdr"),
    sig_label = case_when(
      p_val < 0.001 & p_adj < 0.1 ~ "***",
      p_val < 0.01 & p_adj < 0.1  ~ "**",
      p_val < 0.05 & p_adj < 0.1  ~ "*",
      TRUE          ~ ""
    )
  ) %>%
  ungroup()

# Create estimate matrix for heatmap values
estimate_mat <- results_annotated %>%
  dplyr::select(factor, feature, estimate) %>%
  pivot_wider(names_from = factor, values_from = estimate) %>%
  column_to_rownames("feature") %>%
  as.matrix()

# Create annotation matrix for labels
label_mat <- results_annotated %>%
  dplyr::select(factor, feature, sig_label) %>%
  pivot_wider(names_from = factor, values_from = sig_label) %>%
  column_to_rownames("feature") %>%
  as.matrix()

## ---- 2. row renaming + group assignment --------------------------------
rename_rows <- c(
  "age_dicho" = "Young vs Old",
  "gender" = "Female vs Male",
  "stage_1_3" = "T: I vs III",
  "stage_1_2" = "T: I vs II",
  "stage_2_3" = "T: II vs III",
  "never_vs_current_smoker" = "Never vs Current",
  "never_vs_Ex_smoker" = "Never vs Ex",
  "current_vs_Ex_smoker" = "Current vs Ex",
  "long_survivors" = "Long vs Short",
  "N0_vs_Nx" = "N: N0 vs Nx",
  "M0_vs_Mx" = "M: M0 vs Mx",
  "T1_vs_Tother" = "T: T1 vs T2+",
  "LUAD_vs_LUSC" = "LUAD vs LUSC",
  "EGFR" = "EGFR", "TP53" = "TP53", "KRAS" = "KRAS", "PIK3CA" = "PIK3CA",
  "STK11" = "STK11", "KEAP1" = "KEAP1",
  "KRAS_only" = "KRAS_only", "KRAS_co" = "KRAS_co",
  "TPS" = "TPS", "CPS" = "CPS", "TMB" = "TMB",
  "T_density__IF1" = "T cell density",
  "mean_inf_ratio" = "CD8 infiltration"
)

## data-driven module blocks (names come through unrenamed)
img_mods <- colnames(meta_scores_df)[1:7]
rna_mods <- colnames(rna)[1:29]
inf_mods <- colnames(inf_scores)[1:11]

## hand-curated blocks (display names)
biomarker_feats <- c("TPS", "CPS", "TMB")
immune_feats    <- c("T cell density", "CD8 infiltration")
demogr_feats    <- c("Female vs Male", "Young vs Old")
clinical_feats  <- c("LUAD vs LUSC", "T: I vs II", "T: I vs III", "T: II vs III",
                     "T: T1 vs T2+", "N: N0 vs Nx", "M: M0 vs Mx")
smoking_feats   <- c("Never vs Current", "Current vs Ex", "Never vs Ex")
os_feats        <- "Long vs Short"
mutation_feats  <- c("EGFR", "KRAS", "KRAS_only", "KRAS_co",
                     "STK11", "KEAP1", "TP53", "PIK3CA")

## one place defining both order and grouping
block_defs <- list(
  "Img modules"  = img_mods,
  "RNA modules"  = rna_mods,
  "Infl. sign."  = inf_mods,
  "Biomarkers"   = biomarker_feats,
  "Immune"       = immune_feats,
  "Demogr."      = demogr_feats,
  "Clinical"     = clinical_feats,
  "Smoking"      = smoking_feats,
  "Mutations"    = mutation_feats,
  "OS"           = os_feats
)

custom_order   <- unlist(block_defs, use.names = FALSE)
feature_groups <- setNames(
  rep(names(block_defs), lengths(block_defs)),
  custom_order
)
group_levels <- names(block_defs)

## sanity check: anything tested but not placed will be silently dropped
tested <- unique(ifelse(results_annotated$feature %in% names(rename_rows),
                        rename_rows[results_annotated$feature],
                        results_annotated$feature))
missing <- setdiff(tested, custom_order)
if (length(missing)) warning("Not in custom_order, will be dropped: ",
                             paste(missing, collapse = ", "))

## ---- 3. helper: matrices for one `type` --------------------------------
make_mats <- function(dat, which_type, feat_order) {
  d <- dplyr::filter(dat, type == which_type)

  est <- d %>%
    dplyr::select(factor, feature, estimate) %>%
    pivot_wider(names_from = factor, values_from = estimate) %>%
    column_to_rownames("feature") %>% as.matrix()

  lab <- d %>%
    dplyr::select(factor, feature, sig_label) %>%
    pivot_wider(names_from = factor, values_from = sig_label) %>%
    column_to_rownames("feature") %>% as.matrix()

  rownames(est) <- ifelse(rownames(est) %in% names(rename_rows),
                          rename_rows[rownames(est)], rownames(est))
  rownames(lab) <- rownames(est)

  ord <- feat_order[feat_order %in% rownames(est)]
  list(
    est = t(est[ord, , drop = FALSE]),   # rows = factors, cols = features
    lab = t(lab[ord, , drop = FALSE]),
    grp = factor(feature_groups[ord], levels = group_levels)
  )
}

cont <- make_mats(results_annotated, "continuous", custom_order)
bin  <- make_mats(results_annotated, "binary",     custom_order)

# drop unused group levels so no empty slices are drawn
cont$grp <- droplevels(cont$grp)
bin$grp  <- droplevels(bin$grp)

n_fac <- nrow(cont$est)

## ---- 4. HR annotation (left of the first heatmap) ----------------------
df <- df[rownames(cont$est), , drop = FALSE]   # keep factor order in sync
hr_mat <- df %>% dplyr::select(-p.value) %>% as.matrix()
hr_ticks <- c(0.75, 1, 1.3)

hr_anno <- rowAnnotation(
  HR = anno_boxplot(
    hr_mat, box_width = 0.005, add_points = TRUE, pch = "|",
    axis_param = list(side = "bottom", at = hr_ticks,
                      labels = as.character(hr_ticks),
                      gp = gpar(fontsize = 5), facing = "outside",
                      labels_rot = 0)
  )
)

## ---- 5. two colour scales ---------------------------------------------
rho_max  <- max(abs(cont$est), na.rm = TRUE)
diff_max <- max(abs(bin$est),  na.rm = TRUE)

col_rho  <- colorRamp2(c(-rho_max, 0, rho_max),
                       c("#762a83", "white", "#1b7837"))   # correlation
col_diff <- colorRamp2(c(-diff_max, 0, diff_max),
                       c("blue", "white", "red"))          # mean difference

## ---- 6. cell_fun factory ----------------------------------------------
# NOTE: ComplexHeatmap passes (j = column, i = row)
make_cell_fun <- function(lab_mat) {
  function(j, i, x, y, width, height, fill) {
    grid.text(lab_mat[i, j], x, y, gp = gpar(fontsize = 7))
    grid.rect(x = x, y = y, width = width, height = height,
              gp = gpar(col = "black", fill = NA, lwd = 0.2))
  }
}

## ---- 7. the two heatmaps ----------------------------------------------
cell_w <- 12 / (ncol(cont$est) + ncol(bin$est))   # cm per column, shared
cell_w <- 0.25
ht_cont <- Heatmap(
  cont$est,
  name = "rho",
  col = col_rho,
  cluster_rows = FALSE, cluster_columns = FALSE,
  column_split = cont$grp,
  row_names_gp = gpar(fontsize = 8, fontface = "bold"),
  row_title = "MOFA Factors",
  row_title_gp = gpar(fontsize = 8),
  column_names_gp = gpar(fontsize = 7),
  column_title_gp = gpar(fontface = "bold", fontsize = 7),
  left_annotation = hr_anno,
  cell_fun = make_cell_fun(cont$lab),
  heatmap_legend_param = list(
    title = "Spearman \u03c1",
    title_gp = gpar(fontsize = 8, fontface = "bold"),
    labels_gp = gpar(fontsize = 7)
  ),
  width  = unit(cell_w * ncol(cont$est), "cm"),
  height = unit(4, "cm"),
  border = TRUE
)

ht_bin <- Heatmap(
  bin$est,
  name = "delta",
  col = col_diff,
  cluster_rows = FALSE, cluster_columns = FALSE,
  column_split = bin$grp,
  show_row_names = FALSE,          # row labels already on the left panel
  column_names_gp = gpar(fontsize = 7),
  column_title_gp = gpar(fontface = "bold", fontsize = 7),
  cell_fun = make_cell_fun(bin$lab),
  heatmap_legend_param = list(
    title = "\u0394Mean",
    title_gp = gpar(fontsize = 8, fontface = "bold"),
    labels_gp = gpar(fontsize = 7)
  ),
  width  = unit(cell_w * ncol(bin$est), "cm"),
  height = unit(4, "cm"),
  border = TRUE
)

ht_list <- ht_cont + ht_bin

## ---- 8. draw -----------------------------------------------------------
#CairoPDF(paste0(out_path, "/5D_clinical_x_factors_new.pdf"), width = 10, height = 5)

draw(ht_list,
     padding = unit(c(2, 2, 2, 2), "mm"),
     merge_legend = TRUE,
     ht_gap = unit(4, "mm"))

decorate_annotation("HR", {
  grid.lines(c(1, 1), c(0.5, n_fac + 0.5),
             gp = gpar(lty = 1, col = "grey50"), default.units = "native")
  for (x in hr_ticks) {
    grid.lines(c(x, x), c(0.5, n_fac + 0.5),
               gp = gpar(lty = 2, col = "grey50"), default.units = "native")
  }
  grid.points(hr_mat[, 1], n_fac:1,
              pch = 19, size = unit(2, "mm"),
              gp = gpar(col = ifelse(df$p.value < 0.05, "#D55E00", "grey30")),
              default.units = "native")
})

#dev.off()
```

### long
```{r,fig.height=10, fig.width=15}
library(ComplexHeatmap)
library(circlize)
library(grid)
library(Cairo)
library(tidyr)

## ---- 1. significance labels --------------------------
# Create significance stars
results_annotated_fdr <- results %>%
  group_by(factor) %>% 
  mutate(
    p_adj = p.adjust(p_val, method = "fdr"),
    sig_label = case_when(
      p_adj < 0.001 ~ "***",
      p_adj < 0.01  ~ "**",
      p_adj < 0.05  ~ "*",
      TRUE          ~ ""
    )
  ) %>%
  ungroup()

results_annotated <- results %>%
  group_by(factor) %>% 
  mutate(
    p_adj = p.adjust(p_val, method = "fdr"),
    sig_label = case_when(
      p_val < 0.001 & p_adj < 0.1 ~ "***",
      p_val < 0.01 & p_adj < 0.1  ~ "**",
      p_val < 0.05 & p_adj < 0.1  ~ "*",
      TRUE          ~ ""
    )
  ) %>%
  ungroup()

# Create estimate matrix for heatmap values
estimate_mat <- results_annotated %>%
  dplyr::select(factor, feature, estimate) %>%
  pivot_wider(names_from = factor, values_from = estimate) %>%
  column_to_rownames("feature") %>%
  as.matrix()

# Create annotation matrix for labels
label_mat <- results_annotated %>%
  dplyr::select(factor, feature, sig_label) %>%
  pivot_wider(names_from = factor, values_from = sig_label) %>%
  column_to_rownames("feature") %>%
  as.matrix()

## ---- 2. row renaming + group assignment --------------------------------
rename_rows <- c(
  "age_dicho" = "Young vs Old",
  "gender" = "Female vs Male",
  "stage_1_3" = "T: I vs III",
  "stage_1_2" = "T: I vs II",
  "stage_2_3" = "T: II vs III",
  "never_vs_current_smoker" = "Never vs Current",
  "never_vs_Ex_smoker" = "Never vs Ex",
  "current_vs_Ex_smoker" = "Current vs Ex",
  "long_survivors" = "Long vs Short",
  "N0_vs_Nx" = "N: N0 vs Nx",
  "M0_vs_Mx" = "M: M0 vs Mx",
  "T1_vs_Tother" = "T: T1 vs T2+",
  "LUAD_vs_LUSC" = "LUAD vs LUSC",
  "EGFR" = "EGFR", "TP53" = "TP53", "KRAS" = "KRAS", "PIK3CA" = "PIK3CA",
  "STK11" = "STK11", "KEAP1" = "KEAP1",
  "KRAS_only" = "KRAS_only", "KRAS_co" = "KRAS_co",
  "TPS" = "TPS", "CPS" = "CPS", "TMB" = "TMB",
  "T_density__IF1" = "T cell density",
  "mean_inf_ratio" = "CD8 infiltration"
)

## data-driven module blocks (names come through unrenamed)
img_mods <- colnames(meta_scores_df)[1:7]
rna_mods <- colnames(rna)[1:29]
inf_mods <- colnames(inf_scores)[1:11]

## hand-curated blocks (display names)
biomarker_feats <- c("TPS", "CPS", "TMB")
immune_feats    <- c("T cell density", "CD8 infiltration")
demogr_feats    <- c("Female vs Male", "Young vs Old")
clinical_feats  <- c("LUAD vs LUSC", "T: I vs II", "T: I vs III", "T: II vs III",
                     "T: T1 vs T2+", "N: N0 vs Nx", "M: M0 vs Mx")
smoking_feats   <- c("Never vs Current", "Current vs Ex", "Never vs Ex")
os_feats        <- "Long vs Short"
mutation_feats  <- c("EGFR", "KRAS", "KRAS_only", "KRAS_co",
                     "STK11", "KEAP1", "TP53", "PIK3CA")

## one place defining both order and grouping
block_defs <- list(
  "Img modules"  = img_mods,
  "RNA modules"  = rna_mods,
  "Infl. sign."  = inf_mods,
  "Biomarkers"   = biomarker_feats,
  "Immune"       = immune_feats,
  "Demogr."      = demogr_feats,
  "Clinical"     = clinical_feats,
  "Smoking"      = smoking_feats,
  "Mutations"    = mutation_feats,
  "OS"           = os_feats
)

custom_order   <- unlist(block_defs, use.names = FALSE)
feature_groups <- setNames(
  rep(names(block_defs), lengths(block_defs)),
  custom_order
)
group_levels <- names(block_defs)

## sanity check: anything tested but not placed will be silently dropped
tested <- unique(ifelse(results_annotated$feature %in% names(rename_rows),
                        rename_rows[results_annotated$feature],
                        results_annotated$feature))
missing <- setdiff(tested, custom_order)
if (length(missing)) warning("Not in custom_order, will be dropped: ",
                             paste(missing, collapse = ", "))

## ---- 3. helper: matrices for one `type` --------------------------------
make_mats <- function(dat, which_type, feat_order) {
  d <- dplyr::filter(dat, type == which_type)

  est <- d %>%
    dplyr::select(factor, feature, estimate) %>%
    pivot_wider(names_from = factor, values_from = estimate) %>%
    column_to_rownames("feature") %>% as.matrix()

  lab <- d %>%
    dplyr::select(factor, feature, sig_label) %>%
    pivot_wider(names_from = factor, values_from = sig_label) %>%
    column_to_rownames("feature") %>% as.matrix()

  rownames(est) <- ifelse(rownames(est) %in% names(rename_rows),
                          rename_rows[rownames(est)], rownames(est))
  rownames(lab) <- rownames(est)

  ord <- feat_order[feat_order %in% rownames(est)]
  list(
    est = est[ord, , drop = FALSE],      # rows = features, cols = factors
    lab = lab[ord, , drop = FALSE],
    grp = factor(feature_groups[ord], levels = group_levels)
  )
}

cont <- make_mats(results_annotated, "continuous", custom_order)
bin  <- make_mats(results_annotated, "binary",     custom_order)

cont$grp <- droplevels(cont$grp)
bin$grp  <- droplevels(bin$grp)

n_fac <- ncol(cont$est)                  # factors are columns now

## ---- 4. HR annotation (above the first heatmap) ------------------------
df <- df[colnames(cont$est), , drop = FALSE]   # factor order = heatmap columns
hr_mat <- df %>% dplyr::select(-p.value) %>% as.matrix()
hr_ticks <- c(0.75, 1,1.5, 2)

hr_anno <- HeatmapAnnotation(
  HR = anno_boxplot(
    t(hr_mat),                       # column annotation -> stats by column
    box_width = 0.005, add_points = TRUE, pch = "|",
    axis_param = list(side = "left", at = hr_ticks,
                      labels = as.character(hr_ticks),
                      gp = gpar(fontsize = 5), facing = "outside",
                      labels_rot = 0)
  ),
  height = unit(1.5, "cm"),
  annotation_name_gp = gpar(fontsize = 7)
)

## ---- 5. two colour scales ---------------------------------------------
rho_max  <- max(abs(cont$est), na.rm = TRUE)
diff_max <- max(abs(bin$est),  na.rm = TRUE)

col_rho  <- colorRamp2(c(-rho_max, 0, rho_max),
                       c("#762a83", "white", "#1b7837"))   # correlation
col_diff <- colorRamp2(c(-diff_max, 0, diff_max),
                       c("blue", "white", "red"))          # mean difference

## ---- 6. cell_fun factory ----------------------------------------------
# NOTE: ComplexHeatmap passes (j = column, i = row)
make_cell_fun <- function(lab_mat) {
  function(j, i, x, y, width, height, fill) {
    grid.text(lab_mat[i, j], x, y, gp = gpar(fontsize = 7))
    grid.rect(x = x, y = y, width = width, height = height,
              gp = gpar(col = "black", fill = NA, lwd = 0.2))
  }
}

## ---- 7. the two heatmaps ----------------------------------------------
cell_h <- 0.28   # cm per feature row  — main knob for total height
cell_w <- 0.45   # cm per factor column

ht_cont <- Heatmap(
  cont$est,
  name = "rho",
  col = col_rho,
  cluster_rows = FALSE, cluster_columns = FALSE,
  row_split = cont$grp,
  row_title_rot = 0,
  row_title_gp = gpar(fontface = "bold", fontsize = 7),
  row_names_gp = gpar(fontsize = 6),
  show_column_names = FALSE,        # shown on the bottom panel
  top_annotation = hr_anno,
  cell_fun = make_cell_fun(cont$lab),
  heatmap_legend_param = list(
    title = "Spearman \u03c1",
    title_gp = gpar(fontsize = 8, fontface = "bold"),
    labels_gp = gpar(fontsize = 7)
  ),
  width  = unit(cell_w * ncol(cont$est), "cm"),
  height = unit(cell_h * nrow(cont$est), "cm"),
  border = TRUE
)

ht_bin <- Heatmap(
  bin$est,
  name = "delta",
  col = col_diff,
  cluster_rows = FALSE, cluster_columns = FALSE,
  row_split = bin$grp,
  row_title_rot = 0,
  row_title_gp = gpar(fontface = "bold", fontsize = 7),
  row_names_gp = gpar(fontsize = 6),
  column_names_gp = gpar(fontsize = 8, fontface = "bold"),
  cell_fun = make_cell_fun(bin$lab),
  heatmap_legend_param = list(
    title = "\u0394Mean",
    title_gp = gpar(fontsize = 8, fontface = "bold"),
    labels_gp = gpar(fontsize = 7)
  ),
  width  = unit(cell_w * ncol(bin$est), "cm"),
  height = unit(cell_h * nrow(bin$est), "cm"),
  border = TRUE
)

ht_list <- ht_cont %v% ht_bin

## ---- 8. draw -----------------------------------------------------------
## ---- 8. draw -----------------------------------------------------------
pdf_h <- 1.6 + (cell_h * (nrow(cont$est) + nrow(bin$est)) + 1.5) / 2.54
CairoPDF(paste0(out_path, "/5D_clinical_x_factors_long.pdf"),width = 4.5, height = pdf_h)

draw(ht_list,
     padding = unit(c(2, 2, 2, 2), "mm"),
     merge_legend = TRUE,
     ht_gap = unit(4, "mm"),
     column_title = "MOFA Factors",
     column_title_gp = gpar(fontsize = 9, fontface = "bold"))

decorate_annotation("HR", {
  # x is now the factor axis (1..n_fac), y is the HR value axis
  grid.lines(c(0.5, n_fac + 0.5), c(1, 1),
             gp = gpar(lty = 1, col = "grey50"), default.units = "native")
  for (y in hr_ticks) {
    grid.lines(c(0.5, n_fac + 0.5), c(y, y),
               gp = gpar(lty = 2, col = "grey50"), default.units = "native")
  }
  grid.points(1:n_fac, hr_mat[, 1],
              pch = 19, size = unit(2, "mm"),
              gp = gpar(col = ifelse(df$p.value < 0.05, "#D55E00", "grey30")),
              default.units = "native")
})

dev.off()
```

# 5D overall feature selection
```{r}
# load data from survival analysis

cox_luad  <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/survival/Rout/cox_luad.rds")
cox_lusc  <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/survival/Rout/cox_lusc.rds")

df <- rbind(cox_luad, cox_lusc)
p <- df %>%
  filter(p_lr < 0.05) %>%
  group_by(histology) %>%
  count(category) %>%
  ggplot(aes(x = category,y = n, fill = histology))+
  geom_bar(stat= "identity", position= "dodge")+
  scale_fill_manual(values = c("LUAD" = "blue", "LUSC"= "orange"))+
  labs(y = "Features p < 0.05", x = "Modality")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5,hjust = 1),
        legend.position = "bottom")
p

ggsave(plot = p, path = out_path,
       filename = "5D_sig_feature_count.pdf",
       width = 2.5, 
       height = 3.5,
       device = cairo_pdf)
```

# 5E overlap between LUAD and LUSC
```{r}
library(ggVennDiagram)
library(ggvenn)
luad_names <- df %>% 
  filter(p_lr < 0.05,
         histology == "LUAD",
         scale == "rank_INT") %>% 
  pull(name)

lusc_names <- df %>% 
  filter(p_lr < 0.05,
         histology == "LUSC",
         scale == "rank_INT") %>% 
  pull(name)

p <- ggvenn(data =  list(LUAD = luad_names, LUSC = lusc_names),fill_color = c("LUAD" = "blue", "LUSC" = "orange"),set_name_size = 4)+
  coord_flip()+
  ggtitle("Feature overlap")
p
ggsave(plot = p, path = out_path,
       filename = "5D_overlap_LUAD_LUSC.pdf",
       width = 2.15, 
       height = 2.5,
       device = cairo_pdf)
```

# 5E top survival features


## LUAD
```{r}
feat_number <- 10
# colors
col_vec <- c(pals::alphabet(n = 8))
names(col_vec) <- c("IMC1", "IMC2", "IF1", "IF2", "IF3", "RNA", "cello1", "cello2")
library(tidyverse)
cur_luad <- cox_luad %>%
  filter(scale == "rank_INT") %>%
   arrange(p_lr) %>%
  #filter(p_lr < 0.05) %>%
  slice_head(n = feat_number) %>%
  mutate(modality = str_split_i(string = name, pattern = "__", 2),
         modality = str_split_i(string = modality, pattern = "_", 1),
         feature = str_split_i(string = name, pattern = "__", 1))

cur_ord_luad <- cur_luad %>%
  group_by(category) %>%
  arrange(desc(hr), .by_group = TRUE) %>%
  pull(feature)

p_luad <- cur_luad %>%
  ggplot(aes(x = hr, y = factor(feature, cur_ord_luad), col = modality)) +
  geom_point(size = 2) +
  geom_errorbar(aes(xmin = lo, xmax = hi), lwd = 1, width = 0.3) +
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1)) +
  scale_x_log10(limits = c(-0.2, 6.5))+
  theme_bw() +
  scale_color_manual(values = col_vec)+
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_text(aes(label = paste0("p = ", round(p_lr, 3))),
            x = Inf, hjust = 1.05, col = "black", size = 3)+
  facet_wrap(.~category, scales = "free_y", space = "free_y",strip.position = "top") +

  theme(
    strip.text       = element_text(size = 10, face = "bold", margin = margin(3, 3, 3, 3)),
    strip.background = element_rect(fill = "grey90", color = NA),
    panel.spacing    = unit(0.8, "lines"))+
  labs(y = "Feature", x = "HR")

p_luad
ggsave(plot = p_luad, path = out_path,
       filename = "5E_top_features_LUAD.pdf",
       width = 5, 
       height = 8,
       device = cairo_pdf)
```

## LUSC
```{r}
cur_lusc <- cox_lusc %>%
  filter(scale == "rank_INT") %>%
   arrange(p_lr) %>%
  #filter(p_lr < 0.05) %>%
  slice_head(n = feat_number) %>%
  mutate(modality = str_split_i(string = name, pattern = "__", 2),
         modality = str_split_i(string = modality, pattern = "_", 1),
         feature = str_split_i(string = name, pattern = "__", 1))

cur_ord_lusc <- cur_lusc %>%
  group_by(category) %>%
  arrange(desc(hr), .by_group = TRUE) %>%
  pull(feature)

p_lusc <- cur_lusc %>%
  ggplot(aes(x = hr, y = factor(feature, cur_ord_lusc), col = modality)) +
  geom_point(size = 2) +
  geom_errorbar(aes(xmin = lo, xmax = hi), lwd = 1, width = 0.3) +
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1)) +
  scale_x_log10(limits = c(-0.2, 28))+
  scale_color_manual(values = col_vec)+
  theme_bw() +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_text(aes(label = paste0("p = ", round(p_lr, 3))),
            x = Inf, hjust = 1.05, col = "black", size = 3)+
  facet_wrap(.~category, scales = "free_y", space = "free_y",strip.position = "top") +

  theme(
    strip.text       = element_text(size = 10, face = "bold", margin = margin(3, 3, 3, 3)),
    strip.background = element_rect(fill = "grey90", color = NA),
    panel.spacing    = unit(0.8, "lines"))+
  labs(y = "Feature", x = "HR")

p_lusc

p <- cowplot::plot_grid(p_luad+theme(legend.position = "bottom"), p_lusc+theme(legend.position = "bottom"), ncol = 2)
p
ggsave(plot = p, path = out_path,
       filename = "5E_top_features.pdf",
       width = 6, 
       height = 5,
       device = cairo_pdf)
```

# directionality of interactions

```{r}
cox_luad %>%
  filter(grepl("Neutrophi|PMN", name),
         scale == "rank_INT", 
         p_lr < 0.05) %>%
  select(name, hr, p_lr)

int_dat <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/interaction_dat.rds")
traits  <-read.csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/all_traits.csv")
df <- left_join(traits, int_dat, "immucan_id")
df$Tumor.Neutrophil__IF1_int
df$Neutrophil.Tumor__IF1_int
coxph(Surv(time_to_event, censoring) ~ Tumor.Neutrophil__IF1_int, data =  df %>% filter(simple_histology == "Adenocarcinoma"))
coxph(Surv(time_to_event, censoring) ~ Neutrophil.Tumor__IF1_int, data =  df %>% filter(simple_histology == "Adenocarcinoma"))

df%>%
  ggplot(aes(x=Neutrophil.Tumor__IF1_int, y = Tumor.Neutrophil__IF1_int, color = censoring))+
  geom_point()
df %>%
  pivot_longer(cols = c(Neutrophil.Tumor__IF1_int,Tumor.Neutrophil__IF1_int)) %>%
  select(immucan_id, censoring, value, name) %>%
  ggplot(aes(x=name, y = value, fill = as.factor(censoring)))+
  geom_boxplot()+
  ggpubr::stat_compare_means()+
  theme_bw()
```


# Survival for imaging modules
```{r}


library(tidyverse)
library(survival)

traits <- readr::read_csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/all_traits.csv")
cur_dat <- left_join(as.data.frame(meta_scores_df), traits,by= "immucan_id") %>%
  ungroup() %>%
  mutate(LUAD_vs_LUSC = ifelse(is.na(LUAD_vs_LUSC),"Other",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma","LUAD",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Squamous cell carcinoma","LUSC",LUAD_vs_LUSC))



p.list <- list()
for(j in c("LUAD", "LUSC", "Other")) {
  df <- cur_dat %>%
    filter(LUAD_vs_LUSC == j)
  
  for (i in colnames(meta_scores)[1:7]) {
    cur_df <- df %>%
      filter(!is.na(.data[[i]]))
      surv_ob <- survival::Surv(cur_df$time_to_event,cur_df$censoring)
    # rank normalization
    cur_df[[i]] <- inv_norm(cur_df[[i]]) 
    # z score normalization
    #cur_df[[i]] <- scale(cur_df[[i]]) 
    p.list[[paste0(j, "_", i)]] <- survival::coxph( as.formula(paste0("surv_ob ~ ", i)),
                                                    data = cur_df) 
  }

}

df <- data.frame(HR =      unlist(lapply(p.list, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p.list, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 p_lr = unlist(lapply(p.list, function(x){summary(x)$logtest["pvalue"]})),
                 lower_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "upper .95"]}))) 

df1 <- df %>% 
  mutate(hist = stringr::str_split_i(rownames(df),"_",1),
         feature = gsub("LUAD_|LUSC_|Other_","", rownames(df)))
  
df1

p_img_mod <- df1 %>%
  ggplot(aes(x = HR,y = feature, col = hist))+
  geom_point(size = 2)+
  facet_wrap(.~hist)+
  geom_errorbar(aes(xmin=lower_conf, xmax = upper_conf),lwd =1,width = 0.3)+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust= 1))+
  #scale_y_discrete(labels = function(x) str_split_i(x,"_",2))+
  scale_x_log10()+
  theme_bw()+
  geom_vline(xintercept = 1,linetype = "dashed")+
  geom_text(aes(label = paste0("p : ",round(p_lr,digits = 3))),
            col = "black",
            vjust = -1,
            inherit.aes = TRUE)+
  scale_color_manual(values = c("LUAD" = "blue", "LUSC" = "orange", "Other" = "gray40", "All" = "black"),name = "Histology")+
  theme(axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10))+
  labs(x = "HR",y = "Histology")

p_img_mod
```

# Survival for RNA modules
```{r, fig.height=12}
traits <- readr::read_csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/all_traits.csv")
cur_dat <- left_join(as.data.frame(rna), traits,by= "immucan_id") %>%
  ungroup() %>%
  mutate(LUAD_vs_LUSC = ifelse(is.na(LUAD_vs_LUSC),"Other",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma","LUAD",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Squamous cell carcinoma","LUSC",LUAD_vs_LUSC))



p.list <- list()
for(j in c("LUAD", "LUSC", "Other")) {
  df <- cur_dat %>%
    filter(LUAD_vs_LUSC == j)
  
  surv_ob <- survival::Surv(df$time_to_event,df$censoring)
  for (i in colnames(rna)[1:29]) {
    cur_df <- df %>%
      filter(!is.na(.data[[i]]))
    surv_ob <- survival::Surv(cur_df$time_to_event,cur_df$censoring)
    # rank normalization
    cur_df[[i]] <- inv_norm(cur_df[[i]]) 
    p.list[[paste0(j, "_", i)]] <- survival::coxph( as.formula(paste0("surv_ob ~ ", i)),
                                                    data = df) 
  }

}

df <- data.frame(HR =      unlist(lapply(p.list, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p_lr = unlist(lapply(p.list, function(x){summary(x)$logtest["pvalue"]})),
                 p.value = unlist(lapply(p.list, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "upper .95"]}))) 

df1 <- df %>% 
  mutate(hist = stringr::str_split_i(rownames(df),"_",1),
         feature = gsub("LUAD_|LUSC_|Other_","", rownames(df)))
  
df1

p_RNA_mod <- df1 %>%
  ggplot(aes(x = HR,y = feature, col = hist))+
  geom_point(size = 2)+
  facet_wrap(.~hist)+
  geom_errorbar(aes(xmin=lower_conf, xmax = upper_conf),lwd =1,width = 0.3)+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust= 1))+
  #scale_y_discrete(labels = function(x) str_split_i(x,"_",2))+
  scale_x_log10()+
  theme_bw()+
  geom_vline(xintercept = 1,linetype = "dashed")+
  geom_text(aes(label = paste0("p : ",round(p_lr,digits = 3))),
            col = "black",
            vjust = -1,
            inherit.aes = TRUE)+
  scale_color_manual(values = c("LUAD" = "blue", "LUSC" = "orange", "Other" = "gray40", "All" = "black"),name = "Histology")+
  theme(axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10))+
  labs(x = "HR",y = "Histology")

p_RNA_mod
```

# Survival for cellohood
```{r, fig.height=12}
# rank normalization function:
inv_norm <- function(x) {
  n <- sum(!is.na(x)); if (n < 3) return(rep(NA_real_, length(x)))
  qnorm((rank(x, na.last = "keep", ties.method = "average") - 0.5) / n)
}

library(tidyverse)
library(survival)
cellodat <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/cello_dat.rds")
traits <- readr::read_csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/all_traits.csv")

cur_dat <- left_join(cellodat, traits,by= "immucan_id") %>%
  ungroup() %>%
  mutate(LUAD_vs_LUSC = ifelse(is.na(LUAD_vs_LUSC),"Other",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma","LUAD",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Squamous cell carcinoma","LUSC",LUAD_vs_LUSC))



p.list <- list()
for(j in c("LUAD", "LUSC", "Other")) {
  df <- cur_dat %>%
    filter(LUAD_vs_LUSC == j)
  
  for (i in colnames(cellodat)[-1]) {
    cur_df <- df %>%
      filter(!is.na(.data[[i]]))

    surv_ob <- survival::Surv(cur_df$time_to_event,cur_df$censoring)
    # rank normalization
    cur_df[[i]] <- inv_norm(cur_df[[i]]) 
    # z score normalization
    #cur_df[[i]] <- scale(cur_df[[i]]) 
    p.list[[paste0(j, "_", i)]] <- survival::coxph( as.formula(paste0("surv_ob ~ ", i)),
                                                    data = cur_df) 
  }

}

df <- data.frame(HR =      unlist(lapply(p.list, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p.list, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 p_lr = unlist(lapply(p.list, function(x){summary(x)$logtest["pvalue"]})),
                 lower_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "upper .95"]}))) 

df1 <- df %>% 
  mutate(hist = stringr::str_split_i(rownames(df),"_",1),
         feature = gsub("LUAD_|LUSC_|Other_","", rownames(df)))
  
df1

p_cello <- df1 %>%
  filter(p.value < 0.05) %>%
  ggplot(aes(x = HR,y = feature, col = hist))+
  geom_point(size = 2)+
  facet_wrap(.~hist)+
  geom_errorbar(aes(xmin=lower_conf, xmax = upper_conf),lwd =1,width = 0.3)+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust= 1))+
  #scale_y_discrete(labels = function(x) str_split_i(x,"_",2))+
  scale_x_log10()+
  theme_bw()+
  geom_vline(xintercept = 1,linetype = "dashed")+
  geom_text(aes(label = paste0("p : ",round(p_lr,digits = 4))),
            col = "black",
            vjust = -1,
            inherit.aes = TRUE)+
  scale_color_manual(values = c("LUAD" = "blue", "LUSC" = "orange", "Other" = "gray40", "All" = "black"),name = "Histology")+
  theme(axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10))+
  labs(x = "HR",y = "Histology")

p_cello

# df1 %>%
#   filter(grepl("Tumor_PDL1", feature))
# # sanity check
# cur_dens <- dens_dat %>% select(Tumor_PDL1_density__IMC1, immucan_id)
# cur_cello <- cellodat %>% 
#   select(immucan_id,CN_2__cello1)
# 
# cur_df <- left_join(dens_dat, cellodat, "immucan_id")
# plot(cur_df$Tumor_PDL1_density__IMC1, cur_df$CN_2__cello1)


```

# Survival for densities
```{r, fig.height=12}
# rank normalization function:
inv_norm <- function(x) {
  n <- sum(!is.na(x)); if (n < 3) return(rep(NA_real_, length(x)))
  qnorm((rank(x, na.last = "keep", ties.method = "average") - 0.5) / n)
}

library(tidyverse)
library(survival)
dens_dat <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/densities_dat.rds")
traits <- readr::read_csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/all_traits.csv")
cur_dat <- left_join(dens_dat, traits,by= "immucan_id") %>%
  ungroup() %>%
  mutate(LUAD_vs_LUSC = ifelse(is.na(LUAD_vs_LUSC),"Other",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma","LUAD",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Squamous cell carcinoma","LUSC",LUAD_vs_LUSC))



p.list <- list()
for(j in c("LUAD", "LUSC", "Other")) {
  df <- cur_dat %>%
    filter(LUAD_vs_LUSC == j)
  
  for (i in colnames(dens_dat)[-1]) {
    cur_df <- df %>%
      filter(!is.na(.data[[i]]))

    surv_ob <- survival::Surv(cur_df$time_to_event,cur_df$censoring)
    # rank normalization
    cur_df[[i]] <- inv_norm(cur_df[[i]]) 
    # z score normalization
    #cur_df[[i]] <- scale(cur_df[[i]]) 
    p.list[[paste0(j, "_", i)]] <- survival::coxph( as.formula(paste0("surv_ob ~ ", i)),
                                                    data = cur_df) 
  }

}

df <- data.frame(HR =      unlist(lapply(p.list, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p.list, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 p_lr = unlist(lapply(p.list, function(x){summary(x)$logtest["pvalue"]})),
                 lower_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "upper .95"]}))) 

df1 <- df %>% 
  mutate(hist = stringr::str_split_i(rownames(df),"_",1),
         feature = gsub("LUAD_|LUSC_|Other_","", rownames(df)))
  
df1

p_dens <- df1 %>%
  filter(p.value < 0.05) %>%
  ggplot(aes(x = HR,y = feature, col = hist))+
  geom_point(size = 2)+
  facet_wrap(.~hist)+
  geom_errorbar(aes(xmin=lower_conf, xmax = upper_conf),lwd =1,width = 0.3)+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust= 1))+
  #scale_y_discrete(labels = function(x) str_split_i(x,"_",2))+
  scale_x_log10()+
  theme_bw()+
  geom_vline(xintercept = 1,linetype = "dashed")+
  geom_text(aes(label = paste0("p : ",round(p_lr,digits = 4))),
            col = "black",
            vjust = -1,
            inherit.aes = TRUE)+
  scale_color_manual(values = c("LUAD" = "blue", "LUSC" = "orange", "Other" = "gray40", "All" = "black"),name = "Histology")+
  theme(axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10))+
  labs(x = "HR",y = "Histology")

p_dens

```
# survival for interactions

```{r, fig.height=20}
# rank normalization function:
inv_norm <- function(x) {
  n <- sum(!is.na(x)); if (n < 3) return(rep(NA_real_, length(x)))
  qnorm((rank(x, na.last = "keep", ties.method = "average") - 0.5) / n)
}

library(tidyverse)
library(survival)
int_dat <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/interaction_dat.rds")
traits <- readr::read_csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/all_traits.csv")

cur_dat <- left_join(int_dat, traits,by= "immucan_id") %>%
  ungroup() %>%
  mutate(LUAD_vs_LUSC = ifelse(is.na(LUAD_vs_LUSC),"Other",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Adenocarcinoma","LUAD",LUAD_vs_LUSC),
         LUAD_vs_LUSC = ifelse(LUAD_vs_LUSC == "Squamous cell carcinoma","LUSC",LUAD_vs_LUSC))



p.list <- list()
for(j in c("LUAD", "LUSC", "Other")) {
  df <- cur_dat %>%
    filter(LUAD_vs_LUSC == j)
  
  for (i in colnames(int_dat)[-1]) {
    cur_df <- df %>%
      filter(!is.na(.data[[i]]))

    surv_ob <- survival::Surv(cur_df$time_to_event,cur_df$censoring)
    # rank normalization
    cur_df[[i]] <- inv_norm(cur_df[[i]]) 
    # z score normalization
    #cur_df[[i]] <- scale(cur_df[[i]]) 
    p.list[[paste0(j, "_", i)]] <- survival::coxph( as.formula(paste0("surv_ob ~ ", i)),
                                                    data = cur_df) 
  }

}

df <- data.frame(HR =      unlist(lapply(p.list, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p.list, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 p_lr = unlist(lapply(p.list, function(x){summary(x)$logtest["pvalue"]})),
                 lower_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "upper .95"]}))) 

df1 <- df %>% 
  mutate(hist = stringr::str_split_i(rownames(df),"_",1),
         feature = gsub("LUAD_|LUSC_|Other_","", rownames(df)))
  
cur_ord_luad <- df1 %>%
  filter(hist == "LUAD") %>%
  arrange(desc(HR)) %>%
  pull(feature)

p_luad_int <- df1 %>%
  filter(hist == "LUAD") %>%
  filter(p.value < 0.05) %>%
  ggplot(aes(x = HR,y = factor(feature, levels = cur_ord_luad), col = hist))+
  geom_point(size = 2)+
  geom_errorbar(aes(xmin=lower_conf, xmax = upper_conf),lwd =1,width = 0.3)+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust= 1))+
  #scale_y_discrete(labels = function(x) str_split_i(x,"_",2))+
  scale_x_log10()+
  theme_bw()+
  geom_vline(xintercept = 1,linetype = "dashed")+
  geom_text(aes(label = paste0("p : ",round(p_lr,digits = 4))),
            col = "black",
            vjust = -1,
            inherit.aes = TRUE)+
  scale_color_manual(values = c("LUAD" = "blue", "LUSC" = "orange", "Other" = "gray40", "All" = "black"),name = "Histology")+
  theme(axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10))+
  labs(x = "HR",y = "Histology")


cur_ord_lusc <- df1 %>%
  filter(hist == "LUSC") %>%
  arrange(desc(HR)) %>%
  pull(feature)

p_lusc_int <- df1 %>%
  filter(hist == "LUSC") %>%
  filter(p.value < 0.05) %>%
  ggplot(aes(x = HR,y = factor(feature, levels = cur_ord_lusc), col = hist))+
  geom_point(size = 2)+
  geom_errorbar(aes(xmin=lower_conf, xmax = upper_conf),lwd =1,width = 0.3)+
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust= 1))+
  #scale_y_discrete(labels = function(x) str_split_i(x,"_",2))+
  scale_x_log10()+
  theme_bw()+
  geom_vline(xintercept = 1,linetype = "dashed")+
  geom_text(aes(label = paste0("p : ",round(p_lr,digits = 4))),
            col = "black",
            vjust = -1,
            inherit.aes = TRUE)+
  scale_color_manual(values = c("LUAD" = "blue", "LUSC" = "orange", "Other" = "gray40", "All" = "black"),name = "Histology")+
  theme(axis.text.x = element_text(size = 10),
        axis.text.y = element_text(size = 10),
        axis.title.x = element_text(size = 10),
        axis.title.y = element_text(size = 10))+
  labs(x = "HR",y = "Histology")

p_luad_int
p_lusc_int
df1 %>%
  filter(grepl("Tumor.Neutrophil", feature))
```

# 5E
## VALIDATIONS
### Fetch TCGA datasets
```{r}
library(TCGAbiolinks)
library(SummarizedExperiment)

# Query and download data
query <- GDCquery(
  project = c("TCGA-LUAD"),
  data.category = "Transcriptome Profiling",
  data.type = "Gene Expression Quantification",
  workflow.type = "STAR - Counts"
)
GDCdownload(query)
LUAD_data <- GDCprepare(query)

# Query and download data
query <- GDCquery(
  project = c("TCGA-LUSC"),
  data.category = "Transcriptome Profiling",
  data.type = "Gene Expression Quantification",
  workflow.type = "STAR - Counts"
)
GDCdownload(query)
LUSC_data <- GDCprepare(query)


# Unify data
LUSC_data$survival <- LUSC_data$vital_status
LUSC_data$histology <- "LUSC"
LUAD_data$survival <- LUAD_data$vital_status
LUAD_data$histology <- "LUAD"

table(LUAD_data$vital_status, LUAD_data$paper_Survival)
# table(is.na(LUAD_data$days_to_death), LUAD_data$survival)
# table(is.na(LUSC_data$days_to_death), LUSC_data$survival)

inters <- intersect(colnames(colData(LUAD_data)), colnames(colData(LUSC_data)))
colData(LUAD_data) <- colData(LUAD_data)[,inters]
colData(LUSC_data) <- colData(LUSC_data)[,inters]

sce <- cbind(LUAD_data, LUSC_data)

rownames(sce) <- rowData(sce)$gene_name  

# filter out deceased with no info on days to survival
sce <- sce[,!(sce$survival == "Dead" & is.na(sce$days_to_death))]

# replace - with . in some cases
rownames(sce)[startsWith(rownames(sce),"IGHV") |  startsWith(rownames(sce),"IGKV") |  startsWith(rownames(sce),"IGLV")] <- gsub("-", "\\.",rownames(sce)[startsWith(rownames(sce),"IGHV") |  startsWith(rownames(sce),"IGKV") |  startsWith(rownames(sce),"IGLV")])
# Save as .RDS (recommended for R objects)
saveRDS(sce, file = paste0(mount_path,"/tcga_LUAD_LUSC.rds"))
```

## Gene set enrichment for Combined FACTOR1 (Signature enrichment/in Combined)
```{r}
library(SummarizedExperiment)
library(SingleCellExperiment)
library(singscore)
library(dplyr)
library(survival)
library(survminer)

# # Load TCGA RNA data
tcga <- readRDS(paste0(mount_path,"/tcga_LUAD_LUSC.rds"))


# Convert to SCE
sce <- as(tcga, "SingleCellExperiment")



# Rank genes and score
ranked_data <- rankGenes(sce)

top_n <- 10
p.list <- list()
gene_df <- data.frame()
for (f in seq(1,6,1)){
  
  # Extract weights for RNA view and Factor 1
  rna_weights <- get_weights(model, view = "RNA", factors = f)$RNA[,1]
  # filter for the ones present in the tcga data
  rna_weights <- rna_weights[names(rna_weights) %in% rownames(sce)]
  top_abs_features <- names(sort(abs(rna_weights), decreasing = TRUE))[1:top_n]
  
  signature_mofa_f1 <- list(
    up = top_abs_features[rna_weights[top_abs_features] > 0],
    down = top_abs_features[rna_weights[top_abs_features] < 0]
  )
  temp <- data.frame(Factor = rep(paste0("Factor",f),length(top_abs_features)),
                     gene = top_abs_features,
                     weight = rna_weights[top_abs_features])
  gene_df <- rbind(gene_df, temp)
  if (length(signature_mofa_f1$up) <= 2){
    print("WARNING: Score is computed inverted and is negated")
    score <- simpleScore(ranked_data, upSet = signature_mofa_f1$down)
    # Merge with clinical data
    survival_df <- data.frame(
      ID = rownames(colData(sce)),
      survival = sce$survival,
      days_to_death = sce$days_to_death,
      days_to_last_follow_up = sce$days_to_last_follow_up,
      score = -score$TotalScore,
      Histology = sce$histology
    )
  }else if(length(signature_mofa_f1$down) <= 2){
    score <- simpleScore(ranked_data, upSet = signature_mofa_f1$up)
    # Merge with clinical data
    survival_df <- data.frame(
      ID = rownames(colData(sce)),
      survival = sce$survival,
      days_to_death = sce$days_to_death,
      days_to_last_follow_up = sce$days_to_last_follow_up,
      score = score$TotalScore,
      Histology = sce$histology
    )
  }else{
    score <- simpleScore(ranked_data, downSet = signature_mofa_f1$down, upSet = signature_mofa_f1$up)
    # Merge with clinical data
    survival_df <- data.frame(
      ID = rownames(colData(sce)),
      survival = sce$survival,
      days_to_death = sce$days_to_death,
      days_to_last_follow_up = sce$days_to_last_follow_up,
      score = score$TotalScore,
      score_up = score$UpScore,
      score_down = score$DownScore,
      Histology = sce$histology
    )
  }
  
  # normalize values before fitting model
  survival_df[survival_df$Histology == "LUAD",]$score <- scale(survival_df[survival_df$Histology == "LUAD",]$score)
  survival_df[survival_df$Histology == "LUSC",]$score <- scale(survival_df[survival_df$Histology == "LUSC",]$score)
  
  
  # Compute survival object
  survival_df <- survival_df %>%
    mutate(
      time = ifelse(!is.na(days_to_death), days_to_death, days_to_last_follow_up),
      status = ifelse(tolower(survival) %in% c("dead", "deceased"), 1, 0)
    ) %>%
    filter(!is.na(time) & !is.na(status) & !is.na(score))
  
  # Cox regression
  cox_model <- coxph(Surv(time, status) ~ score * Histology, data = survival_df)
  print(paste0("Factor",f))
  print(summary(cox_model))
  p.list[[paste0("Factor",f)]] <- cox_model
  
  # Stratify into high/low groups using median
  luad_mean <- median(survival_df %>% filter(Histology == "LUAD") %>% pull(score))
  lusc_mean <- median(survival_df %>% filter(Histology == "LUAD") %>% pull(score))
  survival_df$score_group <- ifelse(survival_df$Histology =="LUAD",
                                    ifelse(survival_df$score >= luad_mean, "LUAD HIGH", "LUAD LOW"),
                                    ifelse(survival_df$score >= lusc_mean, "LUSC HIGH", "LUSC LOW"))
  
  # hist(survival_df %>% filter(Histology == "LUAD") %>% pull(score))
  # hist(survival_df %>% filter(Histology == "LUSC") %>% pull(score))
  
  # KM plot
  fit <- survfit(Surv(time, status) ~ score_group, data = survival_df)
  cox_model <- coxph(Surv(time, status) ~ score_group, data = survival_df)
  
  print(ggsurvplot(fit,
             data = survival_df,
             pval = TRUE,
             risk.table = TRUE,
             ggtheme = theme_minimal(),
              title = paste0("Survival by ssGSEA Score Group (Factor ",f,")")))
  
}


df <- data.frame(HR = unlist(lapply(p.list, function(x){summary(x)$coefficients[,"coef"]})),
                 p.value = unlist(lapply(p.list, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "upper .95"]}))) %>%
  rownames_to_column("term")
```

## plot heatmap
```{r}
custom_order <- c("Factor6", "Factor5", "Factor4", "Factor3", "Factor2", "Factor1")


library(ComplexHeatmap)
library(circlize)
library(dplyr)

# Prepare data
plot_df <- gene_df %>%
  filter(Factor == "Factor1") %>%
  arrange(weight)  # to sort genes by weight

# Create a named vector of weights
weights <- plot_df$weight
names(weights) <- plot_df$gene

# Convert to matrix (one column)
mat <- matrix(weights, ncol = 1)
rownames(mat) <- names(weights)
colnames(mat) <- "Factor1"
mat <- round(mat)

# Define color scale
col_fun <- c("1" = "chartreuse4", "-1" = "darkorchid")


# compute CI for LUSC
coef_vec <- coef(p.list[[1]])
vcov_mat <- vcov(p.list[[1]])
coef_luad <- coef_vec["score"]
coef_interaction <- coef_vec["score:HistologyLUSC"]
coef_lusc <- coef_luad + coef_interaction
hr_lusc <- exp(coef_lusc)
var_lusc <- vcov_mat["score", "score"] +
            vcov_mat["score:HistologyLUSC", "score:HistologyLUSC"] +
            2 * vcov_mat["score", "score:HistologyLUSC"]
se_lusc <- sqrt(var_lusc)
ci_lower <- exp(coef_lusc - 1.96 * se_lusc)
ci_upper <- exp(coef_lusc + 1.96 * se_lusc)

temp_df <- df %>%
  filter(startsWith(term, "Factor1")) %>%
  column_to_rownames("term")
temp_df <- temp_df %>% mutate(HR = exp(HR))
temp_df["Factor1.score:HistologyLUSC","HR"] <- hr_lusc

temp_df["Factor1.score:HistologyLUSC","lower_conf"] <- ci_lower
temp_df["Factor1.score:HistologyLUSC","upper_conf"] <- ci_upper

rownames(temp_df) <- c("Score effect LUAD", "Baseline risk\n(LUSC vs LUAD)", "Score effect LUSC")
# order data frame
temp_df <- temp_df[c("Baseline risk\n(LUSC vs LUAD)", "Score effect LUSC", "Score effect LUAD"),]

temp <- temp_df %>%
  dplyr::select(-p.value) %>%
  as.matrix() 


x_labels <- c(0.7,0.8,0.9,1,1.1,1.2,1.3,1.4)
rg <- c(min(x_labels)-0.1, max(x_labels)+0.1)
panel_fun = function(index, nm) {
    pushViewport(viewport(xscale = rg, yscale = c(0.5, nrow(temp) + 0.5)))
    grid.rect()
    grid.xaxis(at = x_labels, label = x_labels, gp = gpar(fontsize = 8))
    grid.yaxis(at = seq(nrow(temp)), label = rownames(temp), main = TRUE, gp = gpar(fontsize = 8))
    grid.yaxis(at = seq(nrow(temp)), label = paste0(round(temp_df$HR, 2) , "\n",paste0(paste0("(",round(temp_df$lower_conf,2), " - ",round(temp_df$upper_conf,2), ")")), ", p=", signif(temp_df$p.value,2)), main = FALSE, gp = gpar(fontsize = 7))
    for (i in seq_len(nrow(temp))){
      grid.boxplot(temp[i, ], pch = 2, size = unit(40, "mm"), pos = i, direction = "horizontal", box_width = 0.00005)
    }
    # Solid line at HR = 1
    grid.lines(c(1, 1), c(0.5, nrow(temp) + 0.5),
             gp = gpar(lty = 1, col = "grey50"), default.units = "native")
    # Example: Add dashed lines at HR = 0.6, 0.8, 1.2, 1.4
    for (x in x_labels){
      grid.lines(c(x, x), c(0.5, nrow(temp) + 0.5),
                 gp = gpar(lty = 2, col = "grey50"), default.units = "native")
    }
    grid.points(
      temp[, 1], 1:nrow(temp),
      pch = 19, size = unit(2, "mm"),
      gp = gpar(col = ifelse(temp_df$p.value < 0.05, "#D55E00", "grey30")), 
      default.units = "native"
    )
    popViewport()
}
anno = anno_link(align_to = rep("a", nrow(mat)), which = "row", panel_fun = panel_fun, 
    size = unit(2.5, "cm"), width = unit(10, "cm"), link_width = unit(3, "cm"))



# Plot heatmap
ht <- Heatmap(
  mat,
  name = "Factor1 Weight",
  heatmap_legend_param = list(labels = c("Depleted", "Enriched"), at = c(-1,1)),
  right_annotation = rowAnnotation(HR = anno, gap = unit(0, "points")),
  col = col_fun,
  row_names_side = "left",
  row_names_gp = gpar(fontsize = 8),
  row_names_max_width = unit(8, "cm"),  # ← Increase space for row names here
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  show_column_names = FALSE,
  # show_heatmap_legend = FALSE,
  width = unit(0.4, "cm"),
  height = unit(7, "cm"),
  border = TRUE,
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.rect(x = x, y = y, width = width, height = height,
              gp = gpar(col = "black", fill = NA, lwd = 0.5))
  }
)

draw(ht, padding = unit(c(10, 2, 20, 20), "mm"), heatmap_legend_side = "bottom")
# Set PDF output
CairoPDF(paste0(out_path, "/5E_Combined_Factor_survival_TCGA.pdf"), width = 7, height = 4)  # adjust width/height as needed

# Draw the heatmap
draw(ht, padding = unit(c(0,0,0,30), "mm"), heatmap_legend_side = "bottom")

# Close the PDF device
dev.off() 
```



## Gene set enrichment for Combined FACTOR1 (Signature enrichment/in Combined) TPM
```{r}
library(SummarizedExperiment)
library(SingleCellExperiment)
library(singscore)
library(dplyr)
library(survival)

# # Load TCGA RNA data
tcga <- readRDS(paste0(mount_path,"/tcga_LUAD_LUSC.rds"))


# Convert to SCE
sce <- as(tcga, "SingleCellExperiment")



tpm_mat <- assay(tcga, "tpm_unstrand")
tpm_mat <- as.matrix(tpm_mat)   # in case it's a DelayedArray/sparse matrix

# if your gene set is symbols rather than Ensembl IDs, map via rowData
if ("gene_name" %in% colnames(rowData(tcga))) {
  symbols <- rowData(tcga)$gene_name
  keep <- !is.na(symbols) & !duplicated(symbols)   # drop NA + duplicate symbols
  tpm_mat <- tpm_mat[keep, ]
  rownames(tpm_mat) <- symbols[keep]
}

# 3. rank
ranked_data <- rankGenes(tpm_mat)

top_n <- 20
p.list <- list()
gene_df <- data.frame()
for (f in seq(1,6,1)){
  
  # Extract weights for RNA view and Factor 1
  rna_weights <- get_weights(model, view = "RNA", factors = f)$RNA[,1]
  # filter for the ones present in the tcga data
  rna_weights <- rna_weights[names(rna_weights) %in% rownames(sce)]
  top_abs_features <- names(sort(abs(rna_weights), decreasing = TRUE))[1:top_n]
  
  signature_mofa_f1 <- list(
    up = top_abs_features[rna_weights[top_abs_features] > 0],
    down = top_abs_features[rna_weights[top_abs_features] < 0]
  )
  temp <- data.frame(Factor = rep(paste0("Factor",f),length(top_abs_features)),
                     gene = top_abs_features,
                     weight = rna_weights[top_abs_features])
  gene_df <- rbind(gene_df, temp)
  if (length(signature_mofa_f1$up) <= 2){
    print("WARNING: Score is computed inverted and is negated")
    score <- simpleScore(ranked_data, upSet = signature_mofa_f1$down)
    # Merge with clinical data
    survival_df <- data.frame(
      ID = rownames(colData(sce)),
      survival = sce$survival,
      days_to_death = sce$days_to_death,
      days_to_last_follow_up = sce$days_to_last_follow_up,
      score = -score$TotalScore,
      Histology = sce$histology
    )
  }else if(length(signature_mofa_f1$down) <= 2){
    score <- simpleScore(ranked_data, upSet = signature_mofa_f1$up)
    # Merge with clinical data
    survival_df <- data.frame(
      ID = rownames(colData(sce)),
      survival = sce$survival,
      days_to_death = sce$days_to_death,
      days_to_last_follow_up = sce$days_to_last_follow_up,
      score = score$TotalScore,
      Histology = sce$histology
    )
  }else{
    score <- simpleScore(ranked_data, downSet = signature_mofa_f1$down, upSet = signature_mofa_f1$up)
    # Merge with clinical data
    survival_df <- data.frame(
      ID = rownames(colData(sce)),
      survival = sce$survival,
      days_to_death = sce$days_to_death,
      days_to_last_follow_up = sce$days_to_last_follow_up,
      score = score$TotalScore,
      score_up = score$UpScore,
      score_down = score$DownScore,
      Histology = sce$histology
    )
  }
  
  # normalize values before fitting model
  survival_df[survival_df$Histology == "LUAD",]$score <- scale(survival_df[survival_df$Histology == "LUAD",]$score)
  survival_df[survival_df$Histology == "LUSC",]$score <- scale(survival_df[survival_df$Histology == "LUSC",]$score)
  
  
  # Compute survival object
  survival_df <- survival_df %>%
    mutate(
      time = ifelse(!is.na(days_to_death), days_to_death, days_to_last_follow_up),
      status = ifelse(tolower(survival) %in% c("dead", "deceased"), 1, 0)
    ) %>%
    filter(!is.na(time) & !is.na(status) & !is.na(score))
  
  # Cox regression
  cox_model <- coxph(Surv(time, status) ~ score * Histology, data = survival_df)
  print(paste0("Factor",f))
  print(summary(cox_model))
  p.list[[paste0("Factor",f)]] <- cox_model
  
  # Stratify into high/low groups using median
  luad_mean <- median(survival_df %>% filter(Histology == "LUAD") %>% pull(score))
  lusc_mean <- median(survival_df %>% filter(Histology == "LUAD") %>% pull(score))
  survival_df$score_group <- ifelse(survival_df$Histology =="LUAD",
                                    ifelse(survival_df$score >= luad_mean, "LUAD HIGH", "LUAD LOW"),
                                    ifelse(survival_df$score >= lusc_mean, "LUSC HIGH", "LUSC LOW"))
  
  # hist(survival_df %>% filter(Histology == "LUAD") %>% pull(score))
  # hist(survival_df %>% filter(Histology == "LUSC") %>% pull(score))
  
  # KM plot
  fit <- survfit(Surv(time, status) ~ score_group, data = survival_df)
  cox_model <- coxph(Surv(time, status) ~ score_group, data = survival_df)
  
  print(ggsurvplot(fit,
             data = survival_df,
             pval = TRUE,
             risk.table = TRUE,
             ggtheme = theme_minimal(),
              title = paste0("Survival by ssGSEA Score Group (Factor ",f,")")))
  
}

gene_df
df <- data.frame(HR = unlist(lapply(p.list, function(x){summary(x)$coefficients[,"coef"]})),
                 p.value = unlist(lapply(p.list, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "upper .95"]}))) %>%
  rownames_to_column("term")
```

## plot heatmap Factor 1
```{r}
custom_order <- c("Factor6", "Factor5", "Factor4", "Factor3", "Factor2", "Factor1")


library(ComplexHeatmap)
library(circlize)
library(dplyr)

# Prepare data
plot_df <- gene_df %>%
  filter(Factor == "Factor1") %>%
  arrange(weight)  # to sort genes by weight

# Create a named vector of weights
weights <- plot_df$weight
names(weights) <- plot_df$gene

# Convert to matrix (one column)
mat <- matrix(weights, ncol = 1)
rownames(mat) <- names(weights)
colnames(mat) <- "Factor1"
mat <- round(mat)

# Define color scale
col_fun <- c("1" = "chartreuse4", "-1" = "darkorchid")


# compute CI for LUSC
coef_vec <- coef(p.list[[1]])
vcov_mat <- vcov(p.list[[1]])
coef_luad <- coef_vec["score"]
coef_interaction <- coef_vec["score:HistologyLUSC"]
coef_lusc <- coef_luad + coef_interaction
hr_lusc <- exp(coef_lusc)
var_lusc <- vcov_mat["score", "score"] +
            vcov_mat["score:HistologyLUSC", "score:HistologyLUSC"] +
            2 * vcov_mat["score", "score:HistologyLUSC"]
se_lusc <- sqrt(var_lusc)
ci_lower <- exp(coef_lusc - 1.96 * se_lusc)
ci_upper <- exp(coef_lusc + 1.96 * se_lusc)

temp_df <- df %>%
  filter(startsWith(term, "Factor1")) %>%
  column_to_rownames("term")
temp_df <- temp_df %>% mutate(HR = exp(HR))
temp_df["Factor1.score:HistologyLUSC","HR"] <- hr_lusc

temp_df["Factor1.score:HistologyLUSC","lower_conf"] <- ci_lower
temp_df["Factor1.score:HistologyLUSC","upper_conf"] <- ci_upper

rownames(temp_df) <- c("Score effect LUAD", "Baseline risk\n(LUSC vs LUAD)", "Score effect LUSC")
# order data frame
temp_df <- temp_df[c("Baseline risk\n(LUSC vs LUAD)", "Score effect LUSC", "Score effect LUAD"),]

temp <- temp_df %>%
  dplyr::select(-p.value) %>%
  as.matrix() 


x_labels <- c(0.7,0.8,0.9,1,1.1,1.2,1.3,1.4)
rg <- c(min(x_labels)-0.1, max(x_labels)+0.1)
panel_fun = function(index, nm) {
    pushViewport(viewport(xscale = rg, yscale = c(0.5, nrow(temp) + 0.5)))
    grid.rect()
    grid.xaxis(at = x_labels, label = x_labels, gp = gpar(fontsize = 8))
    grid.yaxis(at = seq(nrow(temp)), label = rownames(temp), main = TRUE, gp = gpar(fontsize = 8))
    grid.yaxis(at = seq(nrow(temp)), label = paste0(round(temp_df$HR, 2) , "\n",paste0(paste0("(",round(temp_df$lower_conf,2), " - ",round(temp_df$upper_conf,2), ")")), ", p=", signif(temp_df$p.value,2)), main = FALSE, gp = gpar(fontsize = 7))
    for (i in seq_len(nrow(temp))){
      grid.boxplot(temp[i, ], pch = 2, size = unit(40, "mm"), pos = i, direction = "horizontal", box_width = 0.00005)
    }
    # Solid line at HR = 1
    grid.lines(c(1, 1), c(0.5, nrow(temp) + 0.5),
             gp = gpar(lty = 1, col = "grey50"), default.units = "native")
    # Example: Add dashed lines at HR = 0.6, 0.8, 1.2, 1.4
    for (x in x_labels){
      grid.lines(c(x, x), c(0.5, nrow(temp) + 0.5),
                 gp = gpar(lty = 2, col = "grey50"), default.units = "native")
    }
    grid.points(
      temp[, 1], 1:nrow(temp),
      pch = 19, size = unit(2, "mm"),
      gp = gpar(col = ifelse(temp_df$p.value < 0.05, "#D55E00", "grey30")), 
      default.units = "native"
    )
    popViewport()
}
anno = anno_link(align_to = rep("a", nrow(mat)), which = "row", panel_fun = panel_fun, 
    size = unit(2.5, "cm"), width = unit(10, "cm"), link_width = unit(3, "cm"))



# Plot heatmap
ht <- Heatmap(
  mat,
  name = "Factor1 Weight",
  heatmap_legend_param = list(labels = c("Depleted", "Enriched"), at = c(-1,1)),
  right_annotation = rowAnnotation(HR = anno, gap = unit(0, "points")),
  col = col_fun,
  row_names_side = "left",
  row_names_gp = gpar(fontsize = 8),
  row_names_max_width = unit(8, "cm"),  # ← Increase space for row names here
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  show_column_names = FALSE,
  # show_heatmap_legend = FALSE,
  width = unit(0.4, "cm"),
  height = unit(7, "cm"),
  border = TRUE,
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.rect(x = x, y = y, width = width, height = height,
              gp = gpar(col = "black", fill = NA, lwd = 0.5))
  }
)

draw(ht, padding = unit(c(10, 2, 20, 20), "mm"), heatmap_legend_side = "bottom")
# Set PDF output
#CairoPDF(paste0(out_path, "/5E_Combined_Factor_survival_TCGA.pdf"), width = 7, height = 4)  # adjust width/height as needed

# Draw the heatmap
draw(ht, padding = unit(c(0,0,0,30), "mm"), heatmap_legend_side = "bottom")

# Close the PDF device
#dev.off() 
```


## plot heatmap Factor 5
```{r}
custom_order <- c("Factor6", "Factor5", "Factor4", "Factor3", "Factor2", "Factor1")


library(ComplexHeatmap)
library(circlize)
library(dplyr)

# Prepare data
plot_df <- gene_df %>%
  filter(Factor == "Factor5") %>%
  arrange(weight)  # to sort genes by weight

# Create a named vector of weights
weights <- plot_df$weight
names(weights) <- plot_df$gene

# Convert to matrix (one column)
mat <- matrix(weights, ncol = 1)
rownames(mat) <- names(weights)
colnames(mat) <- "Factor5"
mat <- round(mat)

# Define color scale
col_fun <- c("2" = "chartreuse4", "-2" = "darkorchid")


# compute CI for LUSC
coef_vec <- coef(p.list[[1]])
vcov_mat <- vcov(p.list[[1]])
coef_luad <- coef_vec["score"]
coef_interaction <- coef_vec["score:HistologyLUSC"]
coef_lusc <- coef_luad + coef_interaction
hr_lusc <- exp(coef_lusc)
var_lusc <- vcov_mat["score", "score"] +
            vcov_mat["score:HistologyLUSC", "score:HistologyLUSC"] +
            2 * vcov_mat["score", "score:HistologyLUSC"]
se_lusc <- sqrt(var_lusc)
ci_lower <- exp(coef_lusc - 1.96 * se_lusc)
ci_upper <- exp(coef_lusc + 1.96 * se_lusc)

temp_df <- df %>%
  filter(startsWith(term, "Factor5")) %>%
  column_to_rownames("term")
temp_df <- temp_df %>% mutate(HR = exp(HR))
temp_df["Factor5.score:HistologyLUSC","HR"] <- hr_lusc

temp_df["Factor5.score:HistologyLUSC","lower_conf"] <- ci_lower
temp_df["Factor5.score:HistologyLUSC","upper_conf"] <- ci_upper

rownames(temp_df) <- c("Score effect LUAD", "Baseline risk\n(LUSC vs LUAD)", "Score effect LUSC")
# order data frame
temp_df <- temp_df[c("Baseline risk\n(LUSC vs LUAD)", "Score effect LUSC", "Score effect LUAD"),]

temp <- temp_df %>%
  dplyr::select(-p.value) %>%
  as.matrix() 


x_labels <- c(0.7,0.8,0.9,1,1.1,1.2,1.3,1.4)
rg <- c(min(x_labels)-0.1, max(x_labels)+0.1)
panel_fun = function(index, nm) {
    pushViewport(viewport(xscale = rg, yscale = c(0.5, nrow(temp) + 0.5)))
    grid.rect()
    grid.xaxis(at = x_labels, label = x_labels, gp = gpar(fontsize = 8))
    grid.yaxis(at = seq(nrow(temp)), label = rownames(temp), main = TRUE, gp = gpar(fontsize = 8))
    grid.yaxis(at = seq(nrow(temp)), label = paste0(round(temp_df$HR, 2) , "\n",paste0(paste0("(",round(temp_df$lower_conf,2), " - ",round(temp_df$upper_conf,2), ")")), ", p=", signif(temp_df$p.value,2)), main = FALSE, gp = gpar(fontsize = 7))
    for (i in seq_len(nrow(temp))){
      grid.boxplot(temp[i, ], pch = 2, size = unit(40, "mm"), pos = i, direction = "horizontal", box_width = 0.00005)
    }
    # Solid line at HR = 1
    grid.lines(c(1, 1), c(0.5, nrow(temp) + 0.5),
             gp = gpar(lty = 1, col = "grey50"), default.units = "native")
    # Example: Add dashed lines at HR = 0.6, 0.8, 1.2, 1.4
    for (x in x_labels){
      grid.lines(c(x, x), c(0.5, nrow(temp) + 0.5),
                 gp = gpar(lty = 2, col = "grey50"), default.units = "native")
    }
    grid.points(
      temp[, 1], 1:nrow(temp),
      pch = 19, size = unit(2, "mm"),
      gp = gpar(col = ifelse(temp_df$p.value < 0.05, "#D55E00", "grey30")), 
      default.units = "native"
    )
    popViewport()
}
anno = anno_link(align_to = rep("a", nrow(mat)), which = "row", panel_fun = panel_fun, 
    size = unit(2.5, "cm"), width = unit(10, "cm"), link_width = unit(3, "cm"))



# Plot heatmap
ht <- Heatmap(
  mat,
  name = "Factor5 Weight",
  heatmap_legend_param = list(labels = c("Depleted", "Enriched"), at = c(-2,2)),
  right_annotation = rowAnnotation(HR = anno, gap = unit(0, "points")),
  col = col_fun,
  row_names_side = "left",
  row_names_gp = gpar(fontsize = 8),
  row_names_max_width = unit(8, "cm"),  # ← Increase space for row names here
  cluster_rows = FALSE,
  cluster_columns = FALSE,
  show_column_names = FALSE,
  # show_heatmap_legend = FALSE,
  width = unit(0.4, "cm"),
  height = unit(7, "cm"),
  border = TRUE,
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.rect(x = x, y = y, width = width, height = height,
              gp = gpar(col = "black", fill = NA, lwd = 0.5))
  }
)

draw(ht, padding = unit(c(10, 2, 20, 20), "mm"), heatmap_legend_side = "bottom")
# Set PDF output
#CairoPDF(paste0(out_path, "/5E_Combined_Factor_survival_TCGA.pdf"), width = 7, height = 4)  # adjust width/height as needed

# Draw the heatmap
draw(ht, padding = unit(c(0,0,0,30), "mm"), heatmap_legend_side = "bottom")

# Close the PDF device
#dev.off() 
```


# feature selection for survival

