---
title: "Supplementary Figure 6"
output: html_document
date: '2023-03-15'
---

```{r}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

# Set path
if (Sys.info()["sysname"] == "Darwin"){
  mount_path <- "/Volumes/G_DQBM_BB_Central$/projects/immucan/"
} else if (Sys.info()["sysname"] == "Linux"){
  mount_path <- "/mnt/immucan_volume/"
} else if (Sys.info()["sysname"] == "Windows"){
  mount_path <- "Y:/"
} else {
  stop("Operating system is not supported")
}
```


load libraries
```{r}
library(ggplot2)
library(reticulate)
library(MOFA2)
library(dplyr)
library(tidyverse)
```

## load model
```{r}
mount_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/MOFA"
out_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/revised_figures/Figure_7_supp/"

model_path <- paste0(mount_path, '/mofa_V6_minor.rds')

if (!dir.exists(out_path)){
  dir.create(out_path)
  dir.create(paste0(out_path, "/extras"))
}
model <- readRDS(model_path)
```

# S7A bootstrapping

code here: [bootstrapping](../../MOFA/Analysis/analysis/03_bootstrapping_mofa.Rmd)

# S7B - features per factor

## plotting function
```{r}
library(purrr)
library(dplyr)
library(tidyr)
library(ComplexHeatmap)
library(circlize)
library(data.table)

get_clusterd_features <- function(view, features, version){
  files <- list.files(mount_path, full.names = TRUE, pattern = "clusters")
  files <- files[grepl(version, files)]
  file <- files[grepl(tolower(view), tolower(files))]
  if (length(file) == 0){
    return (NULL)
  }
  df <- read.csv(file) 
  feats <- df %>%
    filter(feature %in% as.character(features))
  clusters <- feats$cluster_id
  
  df <- df %>%
    filter(cluster_id %in% clusters) %>%
    mutate(top_feature = feature %in% features,
           view = view)
  
  return (df)
}

get_top_features <- function(model, view, factor, top_n = top_n) {
  weights_df <- get_weights(model, views = view, factors = factor, as.data.frame = TRUE)
  
  top_df <- weights_df %>%
    arrange(desc(abs(value))) %>%
    slice_head(n = top_n) %>%
    dplyr::rename(weight = value) %>%
    mutate(view = view, factor = factor) %>%
    dplyr::select(view, factor, feature, weight)
  
  return(top_df)
}

plot_top_feature_heatmap <- function(model, views, view_order, view_colors, model_version, n_factors = 10, top_n = 5) {

  factors <- paste0("Factor", 1:n_factors)

  top_feats <- map_dfr(factors, function(fac) {
    map_dfr(views, function(v) {
      tryCatch({
        get_top_features(model, view = v, factor = fac, top_n = top_n)
      }, error = function(e) NULL)
    })
  })

  top_features <- top_feats %>%
    unite("feature_id", view, feature, sep = " | ", remove = FALSE)
  
  related_features <- do.call(rbind,lapply(unique(top_feats$view), function(x){
    get_clusterd_features(x, top_feats %>%
                            filter(view == x) %>%
                            pull(feature),
                          model_version)
    
  })) %>%
      mutate(cluster_id = paste0(view, "_", cluster_id))

  all_weights <- get_weights(model, views = unique(top_features$view), 
                             factors = unique(top_features$factor), as.data.frame = TRUE) %>%
    unite("feature_id", view, feature, sep = " | ", remove = FALSE) %>%
    filter(feature_id %in% top_features$feature_id)

  feature_matrix <- all_weights %>%
    dplyr::select(factor, feature_id, value) %>%
    pivot_wider(names_from = feature_id, values_from = value, values_fill = NA)

  feature_mat <- as.matrix(feature_matrix[, -1])
  rownames(feature_mat) <- feature_matrix$factor

  highlight_matrix <- outer(rownames(feature_mat), colnames(feature_mat),
                            Vectorize(function(fac, feat) {
                              any(top_features$factor == fac & top_features$feature_id == feat)
                            }))
  dimnames(highlight_matrix) <- dimnames(feature_mat)

  feature_ids <- colnames(feature_mat)
  view_labels <- sapply(strsplit(feature_ids, " \\| "), `[`, 1)
  feature_names <- sapply(strsplit(feature_ids, " \\| "), `[`, 2)
  # test to change names
  feature_names <- gsub(pattern = "_P1|_p1|_p1_p1", replacement = "_IMC1",feature_names)
  feature_names <- gsub(pattern = "_P2|_p2|_p2_p2", replacement = "_IMC2",feature_names)
  feature_names <- gsub(pattern = "_density", replacement = "",feature_names)
  feature_names <- gsub(pattern = "_minor", replacement = "",feature_names)
  
  feature_priority <- top_features %>%
    group_by(feature_id) %>%
    summarise(first_factor = factor[1], .groups = "drop") %>%
    mutate(priority = match(first_factor, sort(unique(first_factor)))) %>%
    dplyr::select(feature_id, priority)

  feature_order_df <- data.frame(
    feature_id = colnames(feature_mat),
    view = view_labels,
    name = feature_names,
    stringsAsFactors = FALSE
  ) %>%
    left_join(feature_priority, by = "feature_id") %>%
    arrange(view, priority)
  
  # sort views
  feature_order_df$view <- factor(feature_order_df$view, levels = view_order)
  feature_order_df <- feature_order_df[order(feature_order_df$view),]
  feature_order_df$view <- as.character(feature_order_df$view)

  feature_mat <- feature_mat[, feature_order_df$feature_id]
  highlight_matrix <- highlight_matrix[, feature_order_df$feature_id]
  feature_names <- feature_order_df$name
  view_labels <- feature_order_df$view

  column_split <- factor(view_labels, levels = unique(view_labels))
  
  # Convert to data.table
  related_features_dt <- as.data.table(related_features)
  
  # For each cluster_id, create a label with all its features pasted together
  cluster_labels <- related_features_dt[, .(
    cluster_label = paste(unique(feature), collapse = " | ")
  ), by = cluster_id]
  
  # Map from feature -> cluster_id
  feature_to_cluster <- related_features_dt[, .(feature, cluster_id)]
  # Merge to get feature -> cluster_label mapping
  feature_labels <- merge(feature_to_cluster, cluster_labels, by = "cluster_id")
  # Create a named vector from feature to cluster_label
  feature_label_map <- setNames(feature_labels$cluster_label, feature_labels$feature)
  # Extend feature_names: replace each with its cluster label
  extended_feature_names <- sapply(feature_names, function(f) {
    if (!is.na(feature_label_map[f])) feature_label_map[f] else f
  })

  ht <- Heatmap(
    feature_mat,
    name = "Factor Weight",
    column_split = column_split,
    top_annotation = HeatmapAnnotation(
      View = anno_block(
        gp = gpar(fill = view_colors[levels(column_split)]),
        labels = levels(column_split),
        labels_gp = gpar(fontsize = 9, fontface = "bold")
      )
    ),
    column_labels = feature_names,
    show_column_names = TRUE,
    show_row_names = TRUE,
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    show_column_dend = FALSE,
    column_title = "Top Features by View and Factor",
    row_title = "MOFA Factors",
    heatmap_legend_param = list(title = "Weight", direction = "horizontal"),
    column_names_gp = gpar(fontsize = 7),
    cell_fun = function(j, i, x, y, width, height, fill) {
      if (highlight_matrix[i, j]) {
        grid.rect(x, y, width, height, gp = gpar(col = "black", fill = NA, lwd = 1.5))
      }
    }
  )
  
  ht_extended <- Heatmap(
    feature_mat,
    name = "Factor Weight",
    column_split = column_split,
    top_annotation = HeatmapAnnotation(
      View = anno_block(
        gp = gpar(fill = view_colors[levels(column_split)]),
        labels = levels(column_split),
        labels_gp = gpar(fontsize = 9, fontface = "bold")
      )
    ),
    column_labels = extended_feature_names,
    show_column_names = TRUE,
    show_row_names = TRUE,
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    show_column_dend = FALSE,
    show_heatmap_legend = FALSE,
    column_title = "Top Features by View and Factor",
    row_title = "MOFA Factors",
    heatmap_legend_param = list(title = "Weight", direction = "horizontal"),
    column_names_gp = gpar(fontsize = 2),
    cell_fun = function(j, i, x, y, width, height, fill) {
      if (highlight_matrix[i, j]) {
        grid.rect(x, y, width, height, gp = gpar(col = "black", fill = NA, lwd = 1))
      }
    }
  )

  return(list(ht = ht, ht_extended = ht_extended, related_features = related_features))
}
```

## medium
```{r} 
num_fac <- 6
num_features <- 5
VIEWS <- MOFA2::views_names(model)
view_order <- c("Densities","RNA","Cellohood","Expressions","Interactions","BINARY") 
# View color palette (matching other plots)
view_colors <- structure(
  RColorBrewer::brewer.pal(length(VIEWS), "Set1"),
  names = sort(VIEWS)
)

res <- plot_top_feature_heatmap(model, views = VIEWS, view_order, view_colors, "V6_minor", n_factors = num_fac, top_n = num_features)


# Set PDF output
pdf(paste0(out_path, "7B_",num_fac,"factors_x_",num_features,"features.pdf"), width = 14, height = 4) 
draw(res$ht, heatmap_legend_side = "bottom", padding = unit(c(1, 2, 2, 10), "mm"))

dev.off()


# EXTRA
#pdf(paste0(out_path, "/extras/7B_",num_fac,"factors_x_",num_features,"features_ext.pdf"), width = 14, height = 4) 
draw(res$ht_extended, heatmap_legend_side = "bottom", padding = unit(c(1, 2, 2, 100), "mm"))
#dev.off()
```

## big (extra)
```{r, fig.width=25, fig.height=10}
num_fac <- 6
num_features <- 10
VIEWS <- MOFA2::views_names(model)
res <- plot_top_feature_heatmap(model, views = c("Cellohood", "Densities", "Interactions", "Expressions", "RNA", "BINARY"), view_order,view_colors, "V6_minor",  n_factors = num_fac, top_n = num_features)

# Set PDF output
#pdf(paste0(out_path, "/extras/5C_",num_fac,"factors_x_",num_features,"features.pdf"), width = 17, height = 4) 
draw(res$ht, padding = unit(c(15, 2, 0, 2), "mm"))
#dev.off()

# EXTRA
#pdf(paste0(out_path, "/extras/5C_",num_fac,"factors_x_",num_features,"features_ext.pdf"), width = 14, height = 4) 
draw(res$ht_extended, heatmap_legend_side = "bottom", padding = unit(c(1, 2, 2, 100), "mm"))
#dev.off()
```


## plotting function rotated
```{r}
plot_top_feature_heatmap <- function(model, views, view_order, view_colors, model_version, n_factors = 10, top_n = 5) {

  factors <- paste0("Factor", 1:n_factors)

  top_feats <- map_dfr(factors, function(fac) {
    map_dfr(views, function(v) {
      tryCatch({
        get_top_features(model, view = v, factor = fac, top_n = top_n)
      }, error = function(e) NULL)
    })
  })

  top_features <- top_feats %>%
    unite("feature_id", view, feature, sep = " | ", remove = FALSE)

  related_features <- do.call(rbind,lapply(unique(top_feats$view), function(x){
    get_clusterd_features(x, top_feats %>%
                            filter(view == x) %>%
                            pull(feature),
                          model_version)

  })) %>%
      mutate(cluster_id = paste0(view, "_", cluster_id))

  all_weights <- get_weights(model, views = unique(top_features$view),
                             factors = unique(top_features$factor), as.data.frame = TRUE) %>%
    unite("feature_id", view, feature, sep = " | ", remove = FALSE) %>%
    filter(feature_id %in% top_features$feature_id)

  feature_matrix <- all_weights %>%
    dplyr::select(factor, feature_id, value) %>%
    pivot_wider(names_from = feature_id, values_from = value, values_fill = NA)

  feature_mat <- as.matrix(feature_matrix[, -1])
  rownames(feature_mat) <- feature_matrix$factor

  highlight_matrix <- outer(rownames(feature_mat), colnames(feature_mat),
                            Vectorize(function(fac, feat) {
                              any(top_features$factor == fac & top_features$feature_id == feat)
                            }))
  dimnames(highlight_matrix) <- dimnames(feature_mat)

  feature_ids <- colnames(feature_mat)
  view_labels <- sapply(strsplit(feature_ids, " \\| "), `[`, 1)
  feature_names <- sapply(strsplit(feature_ids, " \\| "), `[`, 2)
  feature_names <- gsub(pattern = "_P1|_p1|_p1_p1", replacement = "_IMC1",feature_names)
  feature_names <- gsub(pattern = "_P2|_p2|_p2_p2", replacement = "_IMC2",feature_names)
  feature_names <- gsub(pattern = "_density", replacement = "",feature_names)
  feature_names <- gsub(pattern = "_minor", replacement = "",feature_names)

  feature_priority <- top_features %>%
    group_by(feature_id) %>%
    summarise(first_factor = factor[1], .groups = "drop") %>%
    mutate(priority = match(first_factor, sort(unique(first_factor)))) %>%
    dplyr::select(feature_id, priority)

  feature_order_df <- data.frame(
    feature_id = colnames(feature_mat),
    view = view_labels,
    name = feature_names,
    stringsAsFactors = FALSE
  ) %>%
    left_join(feature_priority, by = "feature_id") %>%
    arrange(view, priority)

  feature_order_df$view <- factor(feature_order_df$view, levels = view_order)
  feature_order_df <- feature_order_df[order(feature_order_df$view),]
  feature_order_df$view <- as.character(feature_order_df$view)

  feature_mat <- feature_mat[, feature_order_df$feature_id]
  highlight_matrix <- highlight_matrix[, feature_order_df$feature_id]
  feature_names <- feature_order_df$name
  view_labels <- feature_order_df$view

  # was column_split - now splits ROWS since features move to the row axis after rotation
  row_split <- factor(view_labels, levels = unique(view_labels))

  related_features_dt <- as.data.table(related_features)

  cluster_labels <- related_features_dt[, .(
    cluster_label = paste(unique(feature), collapse = " | ")
  ), by = cluster_id]

  feature_to_cluster <- related_features_dt[, .(feature, cluster_id)]
  feature_labels <- merge(feature_to_cluster, cluster_labels, by = "cluster_id")
  feature_label_map <- setNames(feature_labels$cluster_label, feature_labels$feature)
  extended_feature_names <- sapply(feature_names, function(f) {
    if (!is.na(feature_label_map[f])) feature_label_map[f] else f
  })

  # ── Rotate 90°: transpose the matrix and the highlight mask ──────────────
  feature_mat_t      <- t(feature_mat)       # now: features (rows) x Factors (columns)
  highlight_matrix_t <- t(highlight_matrix)

  ht <- Heatmap(
    feature_mat_t,
    name = "Factor Weight",
    row_split = row_split,
    left_annotation = rowAnnotation(
      View = anno_block(
        gp = gpar(fill = view_colors[levels(row_split)]),
        labels = levels(row_split),
        labels_gp = gpar(fontsize = 9, fontface = "bold")
      )
    ),
    row_labels = feature_names,
    show_column_names = TRUE,
    show_row_names = TRUE,
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    show_column_dend = FALSE,
    row_title = "Top Features by View and Factor",
    column_title = "MOFA Factors",
    heatmap_legend_param = list(title = "Weight", direction = "horizontal"),
    row_names_gp = gpar(fontsize = 7),
    column_names_gp = gpar(fontsize = 9),
    cell_fun = function(j, i, x, y, width, height, fill) {
      if (highlight_matrix_t[i, j]) {
        grid.rect(x, y, width, height, gp = gpar(col = "black", fill = NA, lwd = 1.5))
      }
    }
  )

  ht_extended <- Heatmap(
    feature_mat_t,
    name = "Factor Weight",
    row_split = row_split,
    left_annotation = rowAnnotation(
      View = anno_block(
        gp = gpar(fill = view_colors[levels(row_split)]),
        labels = levels(row_split),
        labels_gp = gpar(fontsize = 9, fontface = "bold")
      )
    ),
    row_labels = extended_feature_names,
    show_column_names = TRUE,
    show_row_names = TRUE,
    cluster_rows = FALSE,
    cluster_columns = FALSE,
    show_column_dend = FALSE,
    show_heatmap_legend = FALSE,
    row_title = "Top Features by View and Factor",
    column_title = "MOFA Factors",
    heatmap_legend_param = list(title = "Weight", direction = "horizontal"),
    row_names_gp = gpar(fontsize = 2),
    column_names_gp = gpar(fontsize = 9),
    cell_fun = function(j, i, x, y, width, height, fill) {
      if (highlight_matrix_t[i, j]) {
        grid.rect(x, y, width, height, gp = gpar(col = "black", fill = NA, lwd = 1))
      }
    }
  )

  return(list(ht = ht, ht_extended = ht_extended, related_features = related_features))
}
```



```{r}
views <- MOFA2::views_names(model)
view_colors <- structure(
  RColorBrewer::brewer.pal(length(views), "Set1"),
  names = sort(views)
)

num_fac <- 6

view_order <- c("Densities","RNA","Cellohood","Expressions","Interactions","BINARY") 
res <- plot_top_feature_heatmap(model, views = VIEWS, view_order, view_colors, "V6_minor", n_factors = num_fac, top_n = num_features)


# Set PDF output
#pdf(paste0(out_path, "7B_",num_fac,"factors_x_",num_features,"features_rotated.pdf"), width = 5, height = 15) 
draw(res$ht, heatmap_legend_side = "bottom", padding = unit(c(1, 2, 2, 10), "mm"))

#dev.off()
```


# S7C supp: GSEA per Factor
```{r}
# Load required libraries
library(MOFA2)
library(dplyr)
library(clusterProfiler)
library(msigdbr)
library(tibble)
library(readr)

# === Define your MOFA object and output path ===
mofa_model <- model  # replace with your MOFA object
output_dir <- mount_path
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

# === Step 1: Load GO Biological Process gene sets ===
msig_gobp <- msigdbr(species = "Homo sapiens", category = "C5", subcategory = "GO:BP")
term2gene <- msig_gobp[, c("gs_name", "gene_symbol")]
term2name <- msig_gobp[, c("gs_name", "gs_description")]

# === Step 2: Extract RNA weights from MOFA model ===
weights <- get_weights(mofa_model, views = "RNA", as.data.frame = TRUE)

# === Step 3: Loop over factors and run GSEA ===
for (f in unique(weights$factor)) {
  # Filter and rank genes by factor weight
  factor_weights <- weights %>%
    filter(factor == f) %>%
    arrange(desc(value))  # or use abs(value) if needed

  ranked_genes <- factor_weights$value
  names(ranked_genes) <- factor_weights$feature

  # Clean ranking list
  ranked_genes <- ranked_genes[!is.na(ranked_genes)]
  ranked_genes <- ranked_genes[!duplicated(names(ranked_genes))]

  # Run GSEA
  set.seed(123)
  gsea_res <- GSEA(
    geneList = ranked_genes,
    TERM2GENE = term2gene,
    TERM2NAME = term2name,
    pvalueCutoff = 1,
    pAdjustMethod = "BH",
    minGSSize = 10,
    maxGSSize = 500,
    scoreType = "std"
  )

  # Save results
  save(gsea_res, file = file.path(output_dir, paste0("GSEA_GOBP_Factor", f, ".RData")))
  write.table(gsea_res@result,
              file = file.path(output_dir, paste0("GSEA_GOBP_Factor", f, ".txt")),
              sep = "\t", row.names = FALSE)
}
```



## plot heatmap

```{r, fig.height=8, fig.width=8}

library(ComplexHeatmap)
library(circlize)
# === Define parameters ===
pathway <- "GOBP"
input_dir <- output_dir  # defined earlier as your GSEA output directory
factors <- unique(weights$factor)
comparison_dir <- file.path(input_dir, "GOBP/comparison")
if (!file.exists(comparison_dir)) dir.create(comparison_dir, recursive = TRUE)

# === Retrieve top 10 enriched pathways per factor ===
all_SignP <- lapply(factors, function(f) {
  res_GSEA_tb <- read.delim(file.path(input_dir, paste0("GSEA_GOBP_Factor", f, ".txt")))
  res_GSEA_tb$ID[1:min(10, nrow(res_GSEA_tb))]
})
all_SignP_uq <- unique(unlist(all_SignP))
# === Filter to retain only pathways with significant FDR in at least one factor ===
signif_pathways <- unique(unlist(lapply(factors, function(f) {
  res <- read.delim(file.path(input_dir, paste0("GSEA_GOBP_Factor", f, ".txt")))
  sig <- res[res$p.adjust < 0.05, "ID"]
  return(sig)
})))

# Keep only those pathways that are both in the top 10 and significant somewhere
all_SignP_uq <- intersect(all_SignP_uq, signif_pathways)

# === Collect enrichment results for selected pathways ===
comp_tables <- lapply(factors, function(f) {
  res_GSEA_tb <- read.delim(file.path(input_dir, paste0("GSEA_GOBP_Factor", f, ".txt")))
  # rownames(res_GSEA_tb) <- res_GSEA_tb$Description
  rownames(res_GSEA_tb) <- res_GSEA_tb$ID
  res_GSEA_sel <- res_GSEA_tb[intersect(all_SignP_uq, rownames(res_GSEA_tb)), ]
  return(res_GSEA_sel)
})
names(comp_tables) <- paste0("Factor", factors)

# === Prepare NES table ===
summary_tb_NES <- do.call("cbind", lapply(1:length(comp_tables), function(i) {
  tb <- comp_tables[[i]]
  res_df <- data.frame(NES = tb$NES, row.names = gsub("GOBP_", "", rownames(tb)))
  colnames(res_df) <- names(comp_tables)[i]
  return(res_df)
}))
hr_rows <- hclust(dist(summary_tb_NES))
hr_cols <- hclust(dist(t(summary_tb_NES)))

# === Prepare FDR table and stars ===
summary_tb_padj <- do.call("cbind", lapply(1:length(comp_tables), function(i) {
  tb <- comp_tables[[i]]
  res_df <- data.frame(padj = tb$p.adjust, row.names = gsub("GOBP_", "", rownames(tb)))
  colnames(res_df) <- names(comp_tables)[i]
  return(res_df)
}))
pval_to_stars <- function(p) {
  sapply(p, function(x) {
    if (is.na(x)) return("")
    else if (x < 0.001) return("***")
    else if (x < 0.01) return("**")
    else if (x < 0.05) return("*")
    else return("")
  })
}

# Apply across summary_tb_padj to create cellnotes
cellnotes <- apply(summary_tb_padj, 2, pval_to_stars)
# === Plot heatmap ===
library(gplots)
library(RColorBrewer)

Reds_palette <- brewer.pal(n = 7, name = "Reds")
myPalette <- c("lightgray", Reds_palette[c(1,3,5,7)])
colnames(summary_tb_NES) <- c(1,2,3,4,5,6)

cell_fun = function(j, i, x, y, width, height, fill) {
  grid.text(cellnotes[i, j], x, y, gp = gpar(fontsize = 6, col = "black"))
  
  # Get corresponding padj value
  pval <- summary_tb_padj[i, j]
  
  # Draw black box if significant
  if (!is.na(pval) && pval < 0.05) {
    grid.rect(x, y, width, height, gp = gpar(col = "black", fill = NA, lwd = 1))
  }
}

ht <- Heatmap(
  as.matrix(summary_tb_NES),
  heatmap_legend_param = list(),
  name = "NES",
  column_title = "Factor",
  column_title_gp = gpar(fontsize = 13.2, fontface = "bold"),
  col = colorRamp2(
    c(min(summary_tb_NES, na.rm=TRUE), 0, max(summary_tb_NES, na.rm=TRUE)), 
    c("blue", "white", "red")
  ),
  cell_fun = function(j, i, x, y, width, height, fill) {
    grid.text(cellnotes[i, j], x, y, gp = gpar(fontsize = 6, col = "black"))
    pval <- summary_tb_padj[i, j]
    if (!is.na(pval) && pval < 0.05) {
      grid.rect(x, y, width, height, gp = gpar(col = "black", fill = NA, lwd = 1))
    }
  },
  cluster_rows = TRUE,
  cluster_columns = FALSE,
  width = unit(4, "cm"),
  row_names_gp = gpar(fontsize = 10),
  column_names_gp = gpar(fontsize = 10, fontface = "bold", just = "left"),
  column_names_side = "top",
  column_names_rot = 0
)
pdf(paste0(out_path,"/S7C_GSEA_GOBP_C5_heatmap.pdf"), height = 14, width = 8.5)
draw(ht, heatmap_legend_side = "right", padding = unit(c(1, 1,1, 11), "cm"))
dev.off()
```

# S7D supp: Association LF5 and LCNEC
```{r}
library(ggpubr)
factors <- get_factors(model, as.data.frame = TRUE) %>%
  dplyr::select(- group) %>%
  pivot_wider(names_from = "factor", values_from = value) %>%
  column_to_rownames("sample")

clin <- samples_metadata(model) %>%
  dplyr::select(-group)

factors <- factors[clin$sample,]

temp <- factors %>%
  rownames_to_column("Immucan_id") %>%
  left_join(clin %>%
              rownames_to_column("Immucan_id"), by = "Immucan_id")

ggplot(temp %>%
         mutate(is_LCNEC = simple_histology == "LCNEC"), aes(x = is_LCNEC, y = Factor4)) +
  geom_boxplot() +
  stat_compare_means(method = "t.test")  # adds p-value from t-test

ggplot(temp, aes(x = simple_histology, y = Factor4)) +
  geom_boxplot() +
  stat_compare_means(method = "t.test")

library(ggplot2)
library(dplyr)
library(ggpubr)

highlight_groups <- c("LCNEC", "Other")
highlight_groups <- c("LCNEC")

temp <- temp %>%
  mutate(
    hist_group = ifelse(simple_histology %in% highlight_groups, "Highlighted", "Other"),
    shape_group = case_when(
      simple_histology == "LCNEC" ~ "LCNEC",
      simple_histology == "Other" ~ "Other",
      TRUE ~ "LUSC/LUAD/Adenosquamous/Unknown"
    ),
    shape_group = factor(shape_group, levels = c("LCNEC", "Other", "LUSC/LUAD/Adenosquamous/Unknown"))
  )

temp <- temp %>%
  mutate(hist_group = case_when(
    simple_histology == "LCNEC" ~ "LCNEC",
    simple_histology == "Other" ~ "Other",
    TRUE ~ "LUSC/LUAD/Adenosquamous/Unknown"
  ),
  shape_group = factor(hist_group, levels = c("LCNEC", "Other", "LUSC/LUAD/Adenosquamous/Unknown")))

  #   hist_group = ifelse(simple_histology %in% highlight_groups, "Highlighted", "Other"),
  #   shape_group = case_when(
  #     simple_histology == "LCNEC" ~ "LCNEC",
  #     simple_histology == "Other" ~ "Other",
  #     TRUE ~ "LUSC/LUAD/Adenosquamous/Unknown"
  #   ),
  #   shape_group = factor(shape_group, levels = c("LCNEC", "Other", "LUSC/LUAD/Adenosquamous/Unknown"))
  # )

# Define shapes: triangle (17), square (15), circle (16)
shape_values <- c("LCNEC" = 8, "Other" = 0, "LUSC/LUAD/Adenosquamous/Unknown" = 16)

hist_col = structure(
  c("blue", "orange", "firebrick", "gray30", "gray30", "gray30"),
  names = c("Adenocarcinoma", "Squamous cell carcinoma", "LCNEC","Other", "unknown", "Adenosquamous carcinoma")
)

gg <- ggplot(temp, aes(x = hist_group, y = Factor5)) +
  geom_boxplot(outlier.shape = NA, alpha = 0.8) +
  geom_jitter(aes(color = simple_histology), width = 0.2, alpha = 0.6, size = 2) +
  #scale_shape_manual(values = shape_values, name = "Histology") +
  scale_color_manual(values =hist_col, guide = "none") +
  #scale_x_discrete(labels = c("Highlighted" = "LCNEC + Other", "Other" = "Rest")) +  # relabel x-axis
  stat_compare_means(method = "wilcox.test",size = 5,label.y = c(4.5, 6,7), comparisons = list(c("LCNEC", "Other"),
                                                                         c("LCNEC","LUSC/LUAD/Adenosquamous/Unknown"),
                                                                         c("Other","LUSC/LUAD/Adenosquamous/Unknown")
                                                                         ), label = "p.signif") +
  theme_minimal() +
  theme(
    plot.title = element_blank(),
    axis.title.x = element_blank(),
    axis.title.y = element_text(size = 14),
    axis.text.x = element_text(size = 14, angle = 45, hjust = 1),
    axis.text.y = element_text(size = 14),
    axis.ticks.x = element_blank(),
    legend.text = element_text(size = 14),
    
    legend.position = "bottom"
  ) +
  labs(y = "Factor5")+
  coord_cartesian(ylim = c(-4,8))
gg

ggsave(paste0(out_path,"S7D_Factor5_LCNEC_OTHER.pdf"), plot = gg,device = "pdf",width = 5,height = 4.5)
```

# S7E supp: Factor1 Survival analysis

## overall
```{r}
## survival analysis
library(survminer)
library(survivalAnalysis)
library(survival)

factors <- get_factors(model, as.data.frame = TRUE) %>%
  dplyr::select(- group) %>%
  pivot_wider(names_from = "factor", values_from = value) %>%
  column_to_rownames("sample")

traits <- read.csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/scores_for_sharing/all_traits.csv")
clin <- samples_metadata(model)
cur_dat <- factors %>%
  rownames_to_column("sample") %>%
  full_join(clin, by = "sample")

p.list <- list()
for (i in names(factors)) {
    surv_ob <- Surv(cur_dat$time_to_event,cur_dat$censoring)
    p.list[[i]] <- coxph( as.formula(paste0("surv_ob ~ ",i)),
                          data = cur_dat)
  # }
}

df <- data.frame(HR = unlist(lapply(p.list, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p.list, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 p_lr = unlist(lapply(p.list, function(x){summary(x)$logtest["pvalue"]})),
                 lower_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "upper .95"]}))) 


custom_order <- c("Factor10", "Factor9", "Factor8", "Factor7", "Factor6", "Factor5", "Factor4", "Factor3", "Factor2", "Factor1")


library(dplyr)
library(ggplot2)

# ------------------------------------------------------------------
# 1.  Pull out the Factor 1 row from each data-frame
# ------------------------------------------------------------------
plot_df <- df %>% 
  select( HR, lower_conf, upper_conf, p_lr) %>%
  mutate(Cohort = rownames(df))

# ------------------------------------------------------------------
# 2.  Forest plot on a log scale
# ------------------------------------------------------------------
p1 <- ggplot(plot_df, aes(y = Cohort, x = HR, col = Cohort)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = lower_conf, xmax = upper_conf), height = 0.25) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_x_log10() +
  labs(
    x = "HR",
    y = NULL,
    title = "Factor 1",
    subtitle = sprintf(
      "LUAD p = %.3g   |   LUSC p = %.3g",
      plot_df$p_lr[plot_df$Cohort == "LUAD"],
      plot_df$p_lr[plot_df$Cohort == "LUSC"]
    )
  ) +
  theme_minimal(base_size = 11)+
  scale_color_manual(values = c("LUAD" = "blue", "LUSC" = "orange"))

print(plot_df)

p1

```


## LUAD/LUSC
```{r}
## survival analysis
library(survminer)
library(survivalAnalysis)
library(survival)

factors <- get_factors(model, as.data.frame = TRUE) %>%
  dplyr::select(- group) %>%
  pivot_wider(names_from = "factor", values_from = value) %>%
  column_to_rownames("sample")

clin <- samples_metadata(model)
cur_dat <- factors %>%
  rownames_to_column("sample") %>%
  full_join(clin, by = "sample") %>%
  filter(simple_histology == "Adenocarcinoma")

p.list <- list()
for (i in names(factors)) {
    surv_ob <- Surv(cur_dat$time_to_event,cur_dat$censoring)
    p.list[[i]] <- coxph( as.formula(paste0("surv_ob ~ ",i)),
                          data = cur_dat)
  # }
}

df_luad <- data.frame(HR = unlist(lapply(p.list, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p_lr = unlist(lapply(p.list, function(x){summary(x)$logtest["pvalue"]})),
                 p.value = unlist(lapply(p.list, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 lower_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "upper .95"]}))) 


custom_order <- c("Factor10", "Factor9", "Factor8", "Factor7", "Factor6", "Factor5", "Factor4", "Factor3", "Factor2", "Factor1")

# LUSC

library(survminer)
library(survivalAnalysis)
library(survival)

factors <- get_factors(model, as.data.frame = TRUE) %>%
  dplyr::select(- group) %>%
  pivot_wider(names_from = "factor", values_from = value) %>%
  column_to_rownames("sample")

clin <- samples_metadata(model)
cur_dat <- factors %>%
  rownames_to_column("sample") %>%
  full_join(clin, by = "sample") %>%
  filter(simple_histology == "Squamous cell carcinoma")

p.list <- list()
for (i in names(factors)) {
    surv_ob <- Surv(cur_dat$time_to_event,cur_dat$censoring)
    p.list[[i]] <- coxph( as.formula(paste0("surv_ob ~ ",i)),
                          data = cur_dat)
  # }
}

df_lusc <- data.frame(HR = unlist(lapply(p.list, function(x){summary(x)$coefficients[,"exp(coef)"]})),
                 p.value = unlist(lapply(p.list, function(x){summary(x)$coefficients[, "Pr(>|z|)"]})),
                 p_lr = unlist(lapply(p.list, function(x){summary(x)$logtest["pvalue"]})),
                 lower_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "lower .95"]})),
                 upper_conf = unlist(lapply(p.list, function(x){summary(x)$conf.int[, "upper .95"]}))) 

custom_order <- c("Factor10", "Factor9", "Factor8", "Factor7", "Factor6", "Factor5", "Factor4", "Factor3", "Factor2", "Factor1")


library(dplyr)
library(ggplot2)

# ------------------------------------------------------------------
# 1.  Pull out the Factor 1 row from each data-frame
# ------------------------------------------------------------------
plot_df_LF1 <- bind_rows(
  df_luad %>%                        # LUAD results
    dplyr::slice(match("Factor1", rownames(.))) %>% 
    mutate(Cohort = "LUAD"),
  df_lusc %>%                        # LUSC results
    dplyr::slice(match("Factor1", rownames(.))) %>% 
    mutate(Cohort = "LUSC")
) %>% 
  select(Cohort, HR, lower_conf, upper_conf, p_lr, p.value)

# ------------------------------------------------------------------
# 2.  Forest plot on a log scale
# ------------------------------------------------------------------
p_LF1 <- ggplot(plot_df_LF1, aes(y = Cohort, x = HR, col = Cohort)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = lower_conf, xmax = upper_conf), height = 0.25) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_x_log10() +
  labs(
    x = "HR",
    y = NULL,
    title = "Factor 1",
    subtitle = sprintf(
      "LUAD p = %.3g   |   LUSC p = %.3g",
      plot_df_LF1$p_lr[plot_df_LF1$Cohort == "LUAD"],
      plot_df_LF1$p_lr[plot_df_LF1$Cohort == "LUSC"]
    )
  ) +
  theme_minimal(base_size = 11)+
  scale_color_manual(values = c("LUAD" = "blue", "LUSC" = "orange"))

print(plot_df_LF1)

p_LF1
#ggsave(paste0(out_path,"/S6C_Factor1_LUAD_LUSC.pdf"), plot = p1, device = "pdf", width = 4, height = 3)

# ------------------------------------------------------------------
# 1.  Pull out the Factor 5 row from each data-frame
# ------------------------------------------------------------------
plot_df_LF4 <- bind_rows(
  df_luad %>%                        # LUAD results
    dplyr::slice(match("Factor4", rownames(.))) %>% 
    mutate(Cohort = "LUAD"),
  df_lusc %>%                        # LUSC results
    dplyr::slice(match("Factor4", rownames(.))) %>% 
    mutate(Cohort = "LUSC")
) %>% 
  select(Cohort, HR, lower_conf, upper_conf, p_lr, p.value)

# ------------------------------------------------------------------
# 2.  Forest plot on a log scale
# ------------------------------------------------------------------
p_LF4 <- ggplot(plot_df_LF4, aes(y = Cohort, x = HR, col = Cohort)) +
  geom_point(size = 3) +
  geom_errorbarh(aes(xmin = lower_conf, xmax = upper_conf), height = 0.25) +
  geom_vline(xintercept = 1, linetype = "dashed") +
  scale_x_log10() +
  labs(
    x = "HR",
    y = NULL,
    title = "Factor 4",
    subtitle = sprintf(
      "LUAD p = %.3g   |   LUSC p = %.3g",
      plot_df_LF4$p_lr[plot_df_LF4$Cohort == "LUAD"],
      plot_df_LF4$p_lr[plot_df_LF4$Cohort == "LUSC"]
    )
  ) +
  theme_minimal(base_size = 11)+
  scale_color_manual(values = c("LUAD" = "blue", "LUSC" = "orange"))

print(plot_df_LF4)

p_LF1
p_LF4

p <- cowplot::plot_grid(p_LF1,p_LF4, ncol =1)
p
ggsave(paste0(out_path,"S7E_Factor1_LUAD_LUSC.pdf"), plot = p, device = "pdf", width = 3.5, height = 3.5)

plot_df_LF1
plot_df_LF4
```

# S7F
code is [here](analysis/13_Figure_5.Rmd) under "plot heatmap Factor 5"

# S7G Gene set size stability
```{r}
## ---------------------------------------------------------------------------
## mofa_factor_survival()
##
## For each MOFA factor: build a top-N weight signature, score samples with
## singscore, z-score and median-split the score *within* histology, then fit
## a Cox model per histology.
##
## Returns HR tables (grouped + continuous) and a forest plot. No KM plots.
## ---------------------------------------------------------------------------

library(SummarizedExperiment)
library(SingleCellExperiment)
library(singscore)
library(dplyr)
library(survival)
library(survminer)
library(MOFA2)
# # Load TCGA RNA data
tcga <- readRDS(paste0(mount_path,"/tcga_LUAD_LUSC.rds"))


# Convert to SCE
sce <- as(tcga, "SingleCellExperiment")
sce <-  sce[,sce$tissue_type != "Normal"]


tpm_mat <- assay(tcga, "tpm_unstrand")
tpm_mat <- as.matrix(tpm_mat)   # in case it's a DelayedArray/sparse matrix

# if your gene set is symbols rather than Ensembl IDs, map via rowData
if ("gene_name" %in% colnames(rowData(tcga))) {
  symbols <- rowData(tcga)$gene_name
  keep <- !is.na(symbols) & !duplicated(symbols)   # drop NA + duplicate symbols
  tpm_mat <- tpm_mat[keep, ]
  rownames(tpm_mat) <- symbols[keep]
}

# 3. rank
ranked_data <- rankGenes(tpm_mat)


mofa_factor_survival <- function(model,
                                 sce,
                                 ranked_data,
                                 top_n         = 20,
                                 factors       = 1:6,
                                 view          = "RNA",
                                 min_set_size  = 3,
                                 histology_col = "histology",
                                 conf_level    = 0.95,
                                 p_adjust      = "BH",
                                 verbose       = TRUE) {

  stopifnot(is.numeric(top_n), top_n >= 2)

  ## -- signature from MOFA weights -------------------------------------------
  .signature <- function(f) {
    w <- get_weights(model, view = view, factors = f)[[view]][, 1]
    names(w) <- .demangle(names(w), rownames(ranked_data))
    w <- w[!is.na(names(w)) & names(w) %in% rownames(ranked_data)]   # scoring universe
    w <- w[names(w) %in% rownames(sce)]
    if (length(w) < top_n)
      stop("Only ", length(w), " scoreable features for factor ", f)
    top <- names(sort(abs(w), decreasing = TRUE))[seq_len(top_n)]
    list(up = top[w[top] > 0], down = top[w[top] < 0], weights = w[top])
  }
  
  .demangle <- function(x, universe) {
    key <- make.names(universe)
    dup <- key[duplicated(key)]
    if (length(dup)) warning(length(dup), " ambiguous name collisions")
    lut <- setNames(universe, key)
    ifelse(x %in% universe, x, unname(lut[x]))
  }

  ## -- singscore, named by sample --------------------------------------------
  .score <- function(sig, fname) {
    n_up <- length(sig$up); n_dn <- length(sig$down)

    if (n_up < min_set_size && n_dn < min_set_size) {
      warning(fname, ": both gene sets smaller than min_set_size; skipped")
      return(NULL)
    }
    if (n_up < min_set_size) {
      if (verbose) message(fname, ": only ", n_up,
                           " up genes - scoring the down set and negating")
      miss <- setdiff(sig$down, rownames(ranked_data))
      message(fname, ": ", length(miss), "/", length(sig$down),
              " down absent | rankData ", nrow(ranked_data), " rows, ",
              "md5 ", substr(digest::digest(rownames(ranked_data)), 1, 8))
      sc   <- simpleScore(ranked_data, upSet = sig$down)
      tot  <- -sc$TotalScore
      inv  <- TRUE
    } else if (n_dn < min_set_size) {
      sc  <- simpleScore(ranked_data, upSet = sig$up)
      tot <- sc$TotalScore
      inv <- FALSE
    } else {
      sc  <- simpleScore(ranked_data, upSet = sig$up, downSet = sig$down)
      tot <- sc$TotalScore
      inv <- FALSE
    }
    list(score = setNames(tot, rownames(sc)), inverted = inv)
  }

  ## -- clinical frame, aligned to sce by sample name -------------------------
  .surv_df <- function(score_vec) {
    ids <- colnames(sce)
    miss <- setdiff(ids, names(score_vec))
    if (length(miss))
      warning(length(miss), " sce samples absent from ranked_data; dropped")

    data.frame(
      ID        = ids,
      time      = ifelse(!is.na(sce$days_to_death),
                         sce$days_to_death, sce$days_to_last_follow_up),
      status    = as.integer(tolower(as.character(sce$survival)) %in%
                               c("dead", "deceased")),
      score     = as.numeric(score_vec[ids]),
      Histology = as.character(colData(sce)[[histology_col]]),
      stringsAsFactors = FALSE
    ) %>%
      filter(!is.na(time), !is.na(status), !is.na(score),
             !is.na(Histology), time > 0) %>%
      group_by(Histology) %>%
      mutate(
        score_z     = as.numeric(scale(score)),
        score_group = factor(ifelse(score_z >= median(score_z), "High", "Low"),
                             levels = c("Low", "High"))
      ) %>%
      ungroup()
  }

  ## -- one Cox fit -> one row ------------------------------------------------
  .fit_row <- function(d, rhs, fname, hist, model_label) {
    out <- data.frame(
      Factor = fname, Histology = hist, model = model_label,
      HR = NA_real_, lower = NA_real_, upper = NA_real_,
      p_wald = NA_real_, p_lr = NA_real_,
      n = nrow(d), n_event = sum(d$status),
      n_high = sum(d$score_group == "High"),
      n_low  = sum(d$score_group == "Low"),
      stringsAsFactors = FALSE
    )
    if (sum(d$status) < 3) return(out)
    if (rhs == "score_group" && nlevels(droplevels(d$score_group)) < 2) return(out)

    fit <- tryCatch(
      coxph(as.formula(paste("Surv(time, status) ~", rhs)), data = d),
      error = function(e) NULL, warning = function(w) NULL
    )
    if (is.null(fit)) return(out)

    s <- summary(fit, conf.int = conf_level)
    out$HR     <- s$conf.int[1, "exp(coef)"]
    out$lower  <- s$conf.int[1, 3]          # positional: colname tracks conf_level
    out$upper  <- s$conf.int[1, 4]
    out$p_wald <- s$coefficients[1, "Pr(>|z|)"]
    out$p_lr   <- unname(s$logtest["pvalue"])   # univariable -> LR for this term
    out
  }

  ## -- loop ------------------------------------------------------------------
  gene_df <- list(); hr_group <- list(); hr_cont <- list(); fits <- list()

  for (f in factors) {
    fname <- paste0("Factor", f)
    sig   <- .signature(f)

    gene_df[[fname]] <- data.frame(
      Factor = fname,
      gene   = names(sig$weights),
      weight = as.numeric(sig$weights),
      direction = ifelse(sig$weights > 0, "up", "down"),
      stringsAsFactors = FALSE
    )

    sc <- .score(sig, fname)
    if (is.null(sc)) next
    sdf <- .surv_df(sc$score)
    if (sc$inverted) sdf$score_z <- sdf$score_z   # already negated upstream

    for (h in sort(unique(sdf$Histology))) {
      d <- filter(sdf, Histology == h)
      hr_group[[paste(fname, h)]] <- .fit_row(d, "score_group", fname, h, "High vs Low")
      hr_cont[[paste(fname, h)]]  <- .fit_row(d, "score_z",     fname, h, "per SD")
      fits[[paste(fname, h)]] <- tryCatch(
        coxph(Surv(time, status) ~ score_z, data = d), error = function(e) NULL)
    }

    if (verbose) message(fname, ": ", length(sig$up), " up / ",
                         length(sig$down), " down, n = ", nrow(sdf),
                         ", events = ", sum(sdf$status))
  }

  .bind <- function(l) {
    r <- bind_rows(l)
    if (!nrow(r)) return(r)
    r %>% group_by(Histology) %>%
      mutate(p_lr_adj = p.adjust(p_lr, method = p_adjust)) %>%
      ungroup() %>%
      mutate(Factor = factor(Factor, levels = paste0("Factor", factors)))
  }
  hr_group <- .bind(hr_group)
  hr_cont  <- .bind(hr_cont)

  ## -- forest plot -----------------------------------------------------------
  plot_hr <- function(tab, subtitle) {
    d <- filter(tab, !is.na(HR), is.finite(lower), is.finite(upper))
    if (!nrow(d)) return(NULL)
    ggplot(d, aes(x = HR, y = Factor, colour = Histology)) +
      geom_vline(xintercept = 1, linetype = "dashed", colour = "grey45") +
      geom_pointrange(aes(xmin = lower, xmax = upper),
                      position = position_dodge(width = 0.55),
                      size = 0.4, fatten = 2.5) +
      scale_x_log10() +
      scale_y_discrete(limits = rev) +
      labs(x = paste0("Hazard ratio (", round(conf_level * 100), "% CI, log scale)"),
           y = NULL, colour = "Histology", subtitle = subtitle) +
      theme_bw(base_size = 11) +
      theme(panel.grid.minor = element_blank())
  }

  list(
    genes         = bind_rows(gene_df),
    hr_group      = hr_group,
    hr_continuous = hr_cont,
    plot_group    = plot_hr(hr_group, "Median split: High vs Low, within histology"),
    plot_cont     = plot_hr(hr_cont,  "Continuous score: HR per SD, within histology"),
    models        = fits,
    call          = match.call()
  )
}

## ---------------------------------------------------------------------------
## usage
## ---------------------------------------------------------------------------

# # sensitivity to signature size
sizes <- lapply(c(5,10, 20, 50, 100), function(k) {
    res <- mofa_factor_survival(model, sce, ranked_data, top_n = k, verbose = FALSE)
  cat("genes used: ",res$genes %>%
  group_by(Factor) %>%
  dplyr::count(direction) %>%
    summarise(total_genes = sum(n)) %>%
    pull(total_genes), "\n")
  res$hr_continuous %>%
    mutate(top_n = k)
})

df <- do.call(rbind, sizes)
df %>%
  filter(Factor %in% c("Factor1", "Factor5"))%>%
  group_by(top_n) %>%
  select(Factor, Histology,HR, p_lr) %>%
  ggplot(aes(x = as.factor(top_n), y = HR, fill = Histology))+
  geom_bar(stat ="identity", position = "dodge")+
  facet_wrap(.~Factor)+
  scale_fill_manual(values = c("LUAD" = "blue", "LUSC" = "orange"))

p <- df %>%
  #filter(Factor %in% c("Factor1", "Factor5")) %>%
  group_by(top_n) %>%
  select(Factor, Histology, HR, p_lr) %>%
  mutate(sig = p_lr < 0.05) %>%
  ggplot(aes(x = as.factor(top_n), y = HR, fill = Histology)) +
  geom_bar(aes(colour = sig), stat = "identity", position = "dodge", linewidth = 0.8) +
  facet_wrap(. ~ Factor, ncol = 2) +
  scale_fill_manual(values = c("LUAD" = "blue", "LUSC" = "orange")) +
  scale_colour_manual(name = "p < 0.05", values = c(`TRUE` = "black", `FALSE` = NA))+
  theme_bw()+
  labs(x = "Number of genes in gene set", y = "HR for gene set in TCGA")
p

ggsave(filename = "S7G_gene_set_stability.pdf",plot = p,device = cairo_pdf,path = out_path,width = 4, height = 6)
```
