---
title: "Feature selection"
author: "Daniel Schulz"
date: "2026-09-21"
output: html_document
---

```{r}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE, message = FALSE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

# Set path
if (Sys.info()["sysname"] == "Darwin"){
  mount_path <- "/Volumes/G_DQBM_BB_Central$/projects/immucan/"
} else if (Sys.info()["sysname"] == "Linux"){
  mount_path <- "/mnt/immucan_volume/"
} else if (Sys.info()["sysname"] == "Windows"){
  mount_path <- "Y:/"
} else {
  stop("Operating system is not supported")
}
```


load libraries
```{r}
library(ggplot2)
library(reticulate)
library(MOFA2)
library(dplyr)
library(tidyverse)
```

## load model
```{r}
out_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/revised_figures/Figure_8_supp/"

if (!dir.exists(out_path)){
  dir.create(out_path)
  dir.create(paste0(out_path, "/extras"))
}

RESULTS_DIR <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/survival/Rout/"

cox_luad  <- readRDS(file.path(RESULTS_DIR,"cox_luad.rds"))
cox_lusc  <- readRDS(file.path(RESULTS_DIR,"cox_lusc.rds"))
```

# S8A overall feature selection
```{r}
# load data from survival analysis



df <- rbind(cox_luad, cox_lusc)
p <- df %>%
  filter(p_lr < 0.05,
         scale == "rank_INT") %>%
  group_by(histology) %>%
  dplyr::count(category) %>%
  ggplot(aes(x = category,y = n, fill = histology))+
  geom_bar(stat= "identity", position= "dodge")+
  scale_fill_manual(values = c("LUAD" = "blue", "LUSC"= "orange"))+
  labs(y = "Features p < 0.05", x = "Modality")+
  theme_bw()+
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5,hjust = 1),
        legend.position = "bottom")
p

ggsave(plot = p, path = out_path,
       filename = "S8A_sig_feature_count.pdf",
       width = 2, 
       height = 3.5,
       device = cairo_pdf)
```


## feature FDR < 0.1
```{r}
cox_luad %>%
  filter(fdr_within_category < 0.1)
cox_luad %>%
  filter(fdr_global < 0.1)
cox_luad %>%
  filter(name == "Neutrophil_density__IF1")
cox_luad %>%
```

# S8B top survival features
## LUAD
```{r}
feat_number <- 20

dat1 <- readRDS(file.path(RESULTS_DIR,"hit_dedup_full.rds"))

## top features after redundancy pruning: within each correlation cluster
## (per histology) density features are preferred, otherwise the representative
select_top_names <- function(dedup, histology_label, n_top) {
  dedup %>%
    group_by(histology, cluster_id) %>%
    mutate(final_selection = if (any(category == "density")) category == "density" else is_representative) %>%
    ungroup() %>%
    filter(final_selection, histology == histology_label) %>%
    arrange(p_lr) %>%
    slice_head(n = n_top) %>%
    pull(name)
}

cur_names_luad <- select_top_names(dat1, "LUAD", feat_number)
cur_names_lusc <- select_top_names(dat1, "LUSC", feat_number)


# colors
col_vec <- c(pals::alphabet(n = 8))
names(col_vec) <- c("IMC1", "IMC2", "IF1", "IF2", "IF3", "RNA", "cello1", "cello2")
library(tidyverse)
cur_luad <- cox_luad %>%
  filter(name %in% cur_names_luad,
         scale == "rank_INT") %>%
  mutate(modality = str_split_i(string = name, pattern = "__", 2),
         modality = str_split_i(string = modality, pattern = "_", 1),
         feature = str_split_i(string = name, pattern = "__", 1))

cur_ord_luad <- cur_luad %>%
  group_by(category) %>%
  arrange(desc(hr), .by_group = TRUE) %>%
  pull(feature)

p_luad <- cur_luad %>%
  ggplot(aes(x = hr, y = factor(feature, cur_ord_luad), col = modality)) +
  geom_point(size = 2) +
  geom_errorbar(aes(xmin = lo, xmax = hi), lwd = 1, width = 0.3) +
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1)) +
  scale_x_log10(limits = c(0.1, 10))+
  theme_bw() +
  scale_color_manual(values = col_vec)+
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_text(aes( x = ifelse(hr > 1,0.8, 8),label = paste0("p = ", round(p_lr, 3))),
            hjust = 1.05, col = "black", size = 3)+
    geom_text(aes(label = ifelse(fdr_within_category < 0.1, "*", ""),
            x = hr, vjust = 0.3, col = "black"), size = 5)+
  facet_wrap(.~category, scales = "free_y", space = "free_y",strip.position = "top") +

  theme(
    strip.text       = element_text(size = 10, face = "bold", margin = margin(3, 3, 3, 3)),
    strip.background = element_rect(fill = "grey90", color = NA),
    panel.spacing    = unit(0.8, "lines"))+
  labs(y = "Feature", x = "HR")

p_luad
#ggsave(plot = p_luad, path = out_path,
#       filename = "5F_top_features_LUAD.pdf",
#       width = 5, 
#       height = 8,
#       device = cairo_pdf)


```

## LUSC
```{r}
cur_lusc <- cox_lusc %>%
  filter(name %in% cur_names_lusc,
         scale == "rank_INT") %>%
  mutate(modality = str_split_i(string = name, pattern = "__", 2),
         modality = str_split_i(string = modality, pattern = "_", 1),
         feature = str_split_i(string = name, pattern = "__", 1))

cur_ord_lusc <- cur_lusc %>%
  group_by(category) %>%
  arrange(desc(hr), .by_group = TRUE) %>%
  pull(feature)

p_lusc <- cur_lusc %>%
  ggplot(aes(x = hr, y = factor(feature, cur_ord_lusc), col = modality)) +
  geom_point(size = 2) +
  geom_errorbar(aes(xmin = lo, xmax = hi), lwd = 1, width = 0.3) +
  theme(axis.text.x = element_text(angle = 90, vjust = 1, hjust = 1)) +
  scale_x_log10(limits = c(0.01, 100))+
  scale_color_manual(values = col_vec)+
  theme_bw() +
  geom_vline(xintercept = 1, linetype = "dashed") +
  geom_text(aes( x = ifelse(hr > 1,0.8, 80),label = paste0("p = ", round(p_lr, 3))),
            hjust = 1.05, col = "black", size = 3)+
  facet_wrap(.~category, scales = "free_y", space = "free_y",strip.position = "top") +

  theme(
    strip.text       = element_text(size = 10, face = "bold", margin = margin(3, 3, 3, 3)),
    strip.background = element_rect(fill = "grey90", color = NA),
    panel.spacing    = unit(0.8, "lines"))+
  labs(y = "Feature", x = "HR")

p_lusc <- p_lusc+theme(legend.position = "bottom")+guides(color = guide_legend(nrow = 2))
p_luad <- p_luad+theme(legend.position = "bottom")+guides(color = guide_legend(nrow = 2))
p <- cowplot::plot_grid(p_luad, p_lusc, ncol = 2)

p_lusc
p_luad
ggsave(plot = p, path = out_path,
       filename = "S8B_top_features.pdf",
       width = 8, 
       height = 6.5,
       device = cairo_pdf)

cox_lusc %>%
  filter(grepl("CD8_PD1_CXCL13.CD8_BCL2|CD8_BCL2.CD8_PD1_CXCL13", name))
cox_luad %>%
  filter(grepl("CD8_PD1_CXCL13.CD8_BCL2|CD8_BCL2.CD8_PD1_CXCL13", name))

```


# S8C null model overall

## compact sig plot
NEW compact plot (round 6): same per-category permutation null as above,
but as one space-efficient panel instead of 16 separate histograms --
boxplot + jittered null draws per category (dodged LUAD/LUSC), hit RATE
(%) on a shared axis (not raw count, since categories differ ~1000x in
size), observed value as a large marker, empirical p printed per box.
biomarker/mutation excluded (both at floor -- 0 observed hits, near-0%
null, in both histologies -- not informative to plot; DS decision).

```{r}
perm <- readRDS(file.path(RESULTS_DIR, "spatial_enrichment_permutation.rds"))

EXCLUDE_CATS <- c("biomarker", "mutation")
categories   <- setdiff(perm$categories, EXCLUDE_CATS)
hist_col     <- c(LUAD = "blue", LUSC = "orange")

null_df <- do.call(rbind, lapply(categories, function(cc) {
  do.call(rbind, lapply(names(hist_col), function(h) {
    n_tested <- perm$category_summary$n_tested[perm$category_summary$histology == h &
                                                   perm$category_summary$category == cc]
    rate <- 100 * perm[[paste0("hits_null_", tolower(h))]][, cc] / n_tested
    data.frame(category = cc, histology = h, rate = rate)
  }))
}))
null_df$category  <- factor(null_df$category, levels = categories)
null_df$histology <- factor(null_df$histology, levels = names(hist_col))

## --- one row per (category, histology): observed hit rate + empirical p ---
obs_df <- perm$category_summary[perm$category_summary$category %in% categories, ]
obs_df$category  <- factor(obs_df$category, levels = categories)
obs_df$histology <- factor(obs_df$histology, levels = names(hist_col))
obs_df$p_label   <- sprintf("p=%.2f", obs_df$emp_p_hits_gt)

dodge <- position_dodge(width = 0.75)

p <- ggplot(null_df, aes(x = category, y = rate, fill = histology, color = histology)) +
  geom_hline(yintercept = 5, linetype = "dashed", color = "grey50") +
  annotate("text", x = categories[1], y = Inf, vjust = 1.5, hjust = -0.05,
           label = "- - - 5% (expected by chance)", size = 2.6, color = "grey40") +
  geom_boxplot(position = dodge, width = 0.65, alpha = 0.25, outlier.shape = NA) +
  geom_point(position = position_jitterdodge(dodge.width = 0.75, jitter.width = 0.15, jitter.height = 0),
             size = 0.5, alpha = 0.35, show.legend = FALSE) +
  geom_point(data = obs_df, aes(x = category, y = hit_rate_obs_pct, fill = histology),
             position = dodge, shape = 23, size = 3.2, color = "red", inherit.aes = FALSE) +
  geom_text(data = obs_df, aes(x = category, y = 40, label = p_label, group = histology),
            position = dodge, vjust = -1, size = 3.5, fontface = "italic", inherit.aes = FALSE) +
  scale_fill_manual(values = hist_col, name = "Histology") +
  scale_color_manual(values = hist_col, guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0.02, 0.15))) +
  labs(title = "Per-category permutation null vs. observed hit rate",
       x = NULL,size= 4, y = "Nominal (p<0.05) hit rate (%)") +
  theme_bw(base_size = 11) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1),
        legend.position = "right",
        plot.title = element_text(face = "bold"))
p
ggsave(plot = p, path = out_path,
       filename = "S8C_category_hitrate_permutation_compact.pdf",
       width = 10, 
       height = 3,
       device = cairo_pdf)
```


# S8D sub-smapled survival relevance
```{r, fig.width=12, fig.height=5}

library(survival)
full_wide <- readRDS(file.path(RESULTS_DIR, "full_wide.rds"))

## features: same as S8B (cur_names_luad / cur_names_lusc)
top_overall <- function(res, n_top = 15, name_sel) {
  sub <- res[res$scale == "rank_INT" & res$name %in% name_sel, ]
  sub <- sub[order(sub$p_lr), ]
  head(sub, n_top)
}


top_luad <- top_overall(cox_luad, feat_number, cur_names_luad)
top_lusc <- top_overall(cox_lusc, feat_number, cur_names_lusc)

inv_norm <- function(x) {
  n <- sum(!is.na(x)); if (n < 3) return(rep(NA_real_, length(x)))
  qnorm((rank(x, na.last = "keep", ties.method = "average") - 0.5) / n)
}

subsample_idx <- function(status, frac = 0.7) {
  idx <- seq_along(status)
  idx0 <- idx[status == 0]; idx1 <- idx[status == 1]
  c(sample(idx0, round(frac * length(idx0))), sample(idx1, round(frac * length(idx1))))
}

subsample_feature <- function(nm, d_base, orig_coef, frac = 0.7, B = 200) {
  d <- d_base[!is.na(d_base[[nm]]), c("time_to_event", "status", nm)]
  colnames(d)[3] <- "x_raw"
  n_sig <- 0L; n_same_dir <- 0L; n_estimable <- 0L
  for (b in seq_len(B)) {
    idx <- subsample_idx(d$status, frac = frac)
    db <- d[idx, ]
    if (length(unique(db$x_raw)) < 3) next
    db$x_rank <- inv_norm(db$x_raw)
    fit <- try(coxph(Surv(time_to_event, status) ~ x_rank, data = db, ties = "efron"), silent = TRUE)
    if (inherits(fit, "try-error") || !is.finite(stats::coef(fit))) next
    n_estimable <- n_estimable + 1L
    cf <- summary(fit)$coefficients
    pw <- cf["x_rank", "Pr(>|z|)"]
    co <- cf["x_rank", "coef"]
    if (!is.na(pw) && pw < 0.05) n_sig <- n_sig + 1L
    if (sign(co) == sign(orig_coef)) n_same_dir <- n_same_dir + 1L
  }
  data.frame(name = nm, B = B, frac = frac, n_estimable = n_estimable,
             pct_significant = 100 * n_sig / n_estimable,
             pct_same_direction = 100 * n_same_dir / n_estimable,
             stringsAsFactors = FALSE)
}

run_subsample <- function(top_df, full_wide, histology_str, n_features = 8, frac = 0.7, B = 200) {
  d_base <- full_wide[full_wide$simple_histology == histology_str, ]
  d_base <- d_base[!is.na(d_base$time_to_event) & d_base$time_to_event > 0 & !is.na(d_base$status), ]
  feats <- head(top_df$name, n_features)
  orig_coefs <- top_df$coef[match(feats, top_df$name)]
  do.call(rbind, Map(function(nm, cf) subsample_feature(nm, d_base, cf, frac = frac, B = B), feats, orig_coefs))
}

set.seed(1235)
subsample_luad <- run_subsample(top_luad, full_wide, "Adenocarcinoma", n_features = feat_number, frac = 0.7, B = 200)

set.seed(1235)
subsample_lusc <- run_subsample(top_lusc, full_wide, "Squamous cell carcinoma", n_features = feat_number, frac = 0.7, B = 200)

subsample_luad$histology <- "LUAD"
subsample_lusc$histology <- "LUSC"


p1 <- subsample_luad %>%
  ggplot(aes(y = factor(name, levels = subsample_luad %>% arrange(pct_significant) %>% pull(name)), x = histology, fill = pct_significant))+
  geom_tile()+
  geom_text(aes(label = round(pct_significant)))+
  scale_fill_viridis_c(limits = c(0,100))+
  labs(y = "Top 20 features", x = element_blank())+
  theme_bw()

p2 <- subsample_lusc %>%
  ggplot(aes(y = factor(name, levels = subsample_lusc %>% arrange(pct_significant) %>% pull(name)), x = histology, fill = pct_significant))+
  geom_tile()+  
  geom_text(aes(label = round(pct_significant)))+
  scale_fill_viridis_c(limits = c(0,100))+
  labs(y = "Top 20 features", x = element_blank())+
  theme_bw()

p <- cowplot::plot_grid(p1+theme(legend.position = "none"), p2, ncol = 2, rel_widths = c(1,1.4))

p

ggsave(plot = p, path = out_path,
       filename = "S8D_subsampled_feature_relevance.pdf",
       width = 8, 
       height = 5,
       device = cairo_pdf)
```

# KS test results spatial vs non-spatial
```{r}
SPATIAL_CATS    <- c("interaction", "cellohood")
NONSPATIAL_CATS <- c("density", "RNA_gene", "RNA_module", "mutation", "biomarker")

sp_df <- bind_rows(cox_luad, cox_lusc) %>%
  filter(scale == "rank_INT",
         category %in% c(SPATIAL_CATS, NONSPATIAL_CATS),
         !is.na(p_lr)) %>%
  mutate(class      = factor(ifelse(category %in% SPATIAL_CATS, "Spatial", "Non-spatial"),
                             levels = c("Non-spatial", "Spatial")),
         neglog10_p = -log10(p_lr),
         abs_coef   = ifelse(category == "mutation", NA, abs(coef)))   # mutations: different scale

## boxplot spatial vs non-spatial with KS p-value per histology
plot_ks <- function(df, var, ylab) {
  df <- df[!is.na(df[[var]]), ]
  ks <- df %>%
    group_by(histology) %>%
    summarise(p = suppressWarnings(ks.test(.data[[var]][class == "Spatial"],
                                           .data[[var]][class == "Non-spatial"])$p.value),
              y = 1.05 * max(.data[[var]]),
              .groups = "drop")

  ggplot(df, aes(x = class, y = .data[[var]], fill = class)) +
    geom_boxplot(outlier.size = 0.3, width = 0.6) +
    geom_text(data = ks, aes(x = 1.5, y = y, label = paste0("KS p = ", signif(p, 2))),
              inherit.aes = FALSE, size = 3) +
    facet_wrap(~ histology) +
    scale_fill_manual(values = c("Non-spatial" = "grey70", "Spatial" = "#2c7fb8")) +
    labs(x = NULL, y = ylab) +
    theme_bw() +
    theme(legend.position = "none",
          axis.text.x = element_text(angle = 45, hjust = 1))
}

p_F <- plot_ks(sp_df, "neglog10_p", expression(-log[10]~italic(p)))
p_G <- plot_ks(sp_df, "abs_coef",   "|log HR| per SD")
p_F; p_G
```

# S8E Spatial vs non-spatial

```{r}
perm <- readRDS(file.path(RESULTS_DIR, "spatial_enrichment_permutation.rds"))
hist_col <- c(LUAD = "blue", LUSC = "orange")

stat_labels <- c(
  D    = "KS D  (LR-test p)",
  diff = "Δ mean (-log10(p)"
)

## --- null distributions, long format: one row per (histology, stat, permutation draw) ---
null_df <- rbind(
  data.frame(histology = "LUAD", stat = stat_labels[["D"]],    value = perm$D_null_luad),
  data.frame(histology = "LUSC", stat = stat_labels[["D"]],    value = perm$D_null_lusc),
  data.frame(histology = "LUAD", stat = stat_labels[["diff"]], value = perm$diff_null_luad),
  data.frame(histology = "LUSC", stat = stat_labels[["diff"]], value = perm$diff_null_lusc)
)
null_df$histology <- factor(null_df$histology, levels = names(hist_col))
null_df$stat      <- factor(null_df$stat, levels = stat_labels)

## --- observed value + empirical p per (histology, stat) ---
obs_long <- rbind(
  data.frame(histology = perm$summary$histology, stat = stat_labels[["D"]],
             obs = perm$summary$D_obs, p = perm$summary$emp_p_ks),
  data.frame(histology = perm$summary$histology, stat = stat_labels[["diff"]],
             obs = perm$summary$diff_obs, p = perm$summary$emp_p_diff_one_sided_spatial_gt)
)
obs_long$histology <- factor(obs_long$histology, levels = names(hist_col))
obs_long$stat      <- factor(obs_long$stat, levels = stat_labels)
obs_long$p_label   <- sprintf("p=%.2f", obs_long$p)

p <- ggplot(null_df, aes(x = histology, y = value, fill = histology, color = histology)) +
  geom_boxplot(width = 0.5, alpha = 0.25, outlier.shape = NA) +
  geom_jitter(width = 0.12, height = 0, size = 0.6, alpha = 0.35, show.legend = FALSE) +
  geom_point(data = obs_long, aes(x = histology, y = obs, fill = histology),
             shape = 23, size = 3.2, color = "red", inherit.aes = FALSE) +
  geom_text(data = obs_long, aes(x = histology, y = 0.25, label = p_label),
            vjust = -1, size = 2.8, fontface = "italic", inherit.aes = FALSE) +
  facet_wrap(~ stat, scales = "free_y") +
  scale_fill_manual(values = hist_col, name = "Histology") +
  scale_color_manual(values = hist_col, guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0.08, 0.18))) +
  labs(title = "Spatial vs non-spatial (p-value)",
       x = NULL, y = NULL) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", strip.text = element_text(size = 9),
        plot.title = element_text(face = "bold", size = 9),
        strip.text.x = element_text(size = 6))
p
ggsave(file.path(out_path, "S8E_spatial_nonspatial_pvalue_permutation_ggplot.pdf"), p, width = 2.5, height = 3.25, device = cairo_pdf)
```

# S8F spatial vs non spatial effect size
```{r}
perm <- readRDS(file.path(RESULTS_DIR, "spatial_enrichment_permutation.rds"))

stat_labels <- c(
    D    = "KS D (|coef|)",
  diff = "Δ mean (|coef|)"
)

## --- null distributions, long format: one row per (histology, stat, permutation draw) ---
null_df <- rbind(
  data.frame(histology = "LUAD", stat = stat_labels[["D"]],    value = perm$D_null_hr_luad),
  data.frame(histology = "LUSC", stat = stat_labels[["D"]],    value = perm$D_null_hr_lusc),
  data.frame(histology = "LUAD", stat = stat_labels[["diff"]], value = perm$diff_null_hr_luad),
  data.frame(histology = "LUSC", stat = stat_labels[["diff"]], value = perm$diff_null_hr_lusc)
)
null_df$histology <- factor(null_df$histology, levels = names(hist_col))
null_df$stat      <- factor(null_df$stat, levels = stat_labels)

## --- observed value + empirical p per (histology, stat) ---
obs_long <- rbind(
  data.frame(histology = perm$summary$histology, stat = stat_labels[["D"]],
             obs = perm$summary$D_obs_hr, p = perm$summary$emp_p_ks_hr),
  data.frame(histology = perm$summary$histology, stat = stat_labels[["diff"]],
             obs = perm$summary$diff_obs_hr, p = perm$summary$emp_p_diff_hr_one_sided_spatial_gt)
)
obs_long$histology <- factor(obs_long$histology, levels = names(hist_col))
obs_long$stat      <- factor(obs_long$stat, levels = stat_labels)
obs_long$p_label   <- sprintf("p=%.2f", obs_long$p)

p <- ggplot(null_df, aes(x = histology, y = value, fill = histology, color = histology)) +
  geom_boxplot(width = 0.5, alpha = 0.25, outlier.shape = NA) +
  geom_jitter(width = 0.12, height = 0, size = 0.6, alpha = 0.35, show.legend = FALSE) +
  geom_point(data = obs_long, aes(x = histology, y = obs, fill = histology),
             shape = 23, size = 3.2, color = "red", inherit.aes = FALSE) +
  geom_text(data = obs_long, aes(x = histology, y = 0.4, label = p_label),
            vjust = -1, size = 2.8, fontface = "italic", inherit.aes = FALSE) +
  facet_wrap(~ stat, scales = "free_y") +
  scale_fill_manual(values = hist_col, name = "Histology") +
  scale_color_manual(values = hist_col, guide = "none") +
  scale_y_continuous(expand = expansion(mult = c(0.08, 0.18))) +
  labs(title = "Spatial vs non-spatial (effect-size)",
       x = NULL, y = NULL) +
  theme_bw(base_size = 11) +
  theme(legend.position = "bottom", strip.text = element_text(size = 9),
        plot.title = element_text(face = "bold", size = 9),
        strip.text.x = element_text(size = 6))
p
ggsave(file.path(out_path, "S8F_spatial_nonspatial_effectsize_permutation_ggplot.pdf"), p, width = 2.5, height = 3.25, device = cairo_pdf)
```




```{r}
hist(sp_df %>%
  filter(class == "Non-spatial",
         histology == "LUAD") %>%
  pull(p_lr))
sp_df %>%
  filter(class == "Spatial",
         histology == "LUAD",
         p_lr < 0.05)
```

