---
title: "general stats for NSCLC2"
author: "Daniel Schulz"
date: "2024-08-27"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(tidyverse)
library(ggplot2)
library(survivalAnalysis)
library(ggsurvfit)
library(survminer)
library(survival)
```

```{r}
mount_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/"
out_folder <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/Figures/Figure_1_supp/"

if (!dir.exists(out_folder)) {
  dir.create(out_folder)
}
```


```{r}
traits <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/all_traits.csv"))
```

# median follow up
```{r}
median(traits$time_to_event)/365*12
```

# clinical table
```{r}
table(traits$gender)/nrow(traits)
table(traits$simple_stage)/nrow(traits)

table(traits$curative_surgery)/nrow(traits)
table(traits$smoking_status)/nrow(traits)
table(traits$death)/nrow(traits)

```


```{r}
dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/cello_dat.rds"))

cur_dat <- dat %>%
  mutate(CN15_high = as.factor(ntile(CN_15__cello1,2))) %>%
  left_join(traits, "immucan_id")

ggforest(coxph( Surv(time_to_event, censoring) ~  CN15_high,data = cur_dat))
x<- coxph( Surv(time_to_event, censoring) ~  neo_adjuvant_treatment,data = traits)

cur_dat %>%
  group_by(LUAD_vs_LUSC) %>%
  summarise(mean(CN_15__cello1))
```


# number of cells in mIF
```{r}
dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/abundances_dat.rds"))
dat <- dat[,grepl("IF",colnames(dat))]
x <- sum(rowSums(dat,na.rm = TRUE),na.rm = TRUE)/10^6
x
```

in mIF we have `r round(x)` million cells

# number of cells in IMC
```{r}
dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/abundances_dat.rds"))
dat <- dat[,grepl("IMC1",colnames(dat))]
(x <- sum(rowSums(dat),na.rm = TRUE)/10^6)

dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/abundances_dat.rds"))
dat <- dat[,grepl("IMC2",colnames(dat))]
(y <- sum(rowSums(dat),na.rm = TRUE)/10^6)
```

in IMC panel 1 we have `r round(x)` and in panel 2 `r round(y)` million cells


```{r, eval=FALSE}
sce1 <- readRDS(paste0(mount_path,"/data/IMC/Panel_1/Rout/sce.rds"))
cur_sce <- sce1[,sce1$ROI != 4]
dat <- as.data.frame(sort(table(cur_sce$celltypes)/ncol(cur_sce)))

dat %>%
  filter(Var1 %in% c("Treg","CD4","CD8")) %>%
  summarise(sum(Freq))

dat %>%
  filter(Var1 %in% c("DC","MacCD163", "HLADR","Neutrophil")) %>%
  summarise(sum(Freq))
```

# smoking and TP53 in LUAD

```{r}
table(traits$smoking_status, traits$TP53)
cur_traits <- traits %>%
  filter(!is.na(smoking_status)) %>%
  mutate(smoker = ifelse(smoking_status %in% c("Current smoker","Ex-smoker"),"smoker", "non-smoker")) %>%
  filter(simple_histology == "Adenocarcinoma")
res <- fisher.test(table(cur_traits$smoker,cur_traits$TP53), or = TRUE,conf.int = TRUE)
print(res)
```


# survival for neo-adjuvant and adjuvant treatment
```{r}
ggforest(coxph( Surv(time_to_event, censoring) ~  neo_adjuvant_treatment,data = traits))
x<- coxph( Surv(time_to_event, censoring) ~  neo_adjuvant_treatment,data = traits)

(p <- summary(x)$coefficients[, "Pr(>|z|)"])
(HR <- summary(x)$coefficients[, "exp(coef)"])
```

Neo-adjuvant treatment has a HR of `r round(HR, digits = 2)` and a p value of `r round(p, digits = 2)`

# survival for adjuvant and adjuvant treatment
```{r}
ggforest(coxph( Surv(time_to_event, censoring) ~  adjuvant_treatment,data = traits))
x<- coxph( Surv(time_to_event, censoring) ~  adjuvant_treatment,data = traits)

(p <- summary(x)$coefficients[, "Pr(>|z|)"])
(HR <- summary(x)$coefficients[, "exp(coef)"])
```

Adjuvant treatment has a HR of `r round(HR, digits = 2)` and a p value of `r round(p, digits = 2)`

# survival for TPS and CPS
```{r}
ggforest(coxph( Surv(time_to_event, censoring) ~  TPS_score,data = traits))
x<- coxph( Surv(time_to_event, censoring) ~  TPS_score,data = traits)
summary(x)

table(traits$TPS_score)/nrow(traits)

(p_TPS <- summary(x)$coefficients[, "Pr(>|z|)"])
(HR_TPS <- summary(x)$coefficients[, "exp(coef)"])

ggforest(coxph( Surv(time_to_event, censoring) ~  CPS_score,data = traits))
x<- coxph( Surv(time_to_event, censoring) ~  CPS_score,data = traits)
summary(x)

(p_CPS <- summary(x)$coefficients[, "Pr(>|z|)"])
(HR_CPS <- summary(x)$coefficients[, "exp(coef)"])
```

TPS > 1% has a HR of `r round(HR_TPS, digits = 2)` and a p value of `r round(p_TPS, digits = 2)` and CPS has a HR of `r round(HR_CPS, digits = 2)` and a p value of `r round(p_CPS, digits = 2)`

# survival for TMB
```{r}
ggforest(coxph( Surv(time_to_event, censoring) ~  TMB_score ,data = traits))
x<- coxph( Surv(time_to_event, censoring) ~  TMB_score* LUAD_vs_LUSC,data = traits)
summary(x)

(p <- summary(x)$coefficients[, "Pr(>|z|)"])
(HR <- summary(x)$coefficients[, "exp(coef)"])
```

TMB > 10 mut/MB has a HR of `r round(HR, digits = 2)` and a p value of `r round(p, digits = 2)` 

# survival for smoking
```{r}
cur_traits <- traits
cur_traits$smoking_status <- factor(cur_traits$smoking_status,levels = c("Non-smoker" ,"Ex-smoker", "Current smoker"))
ggforest(coxph( Surv(time_to_event, censoring) ~  smoking_status,data = cur_traits))
x<- coxph( Surv(time_to_event, censoring) ~  smoking_status,data = cur_traits)
summary(x)

(p <- summary(x)$coefficients[, "Pr(>|z|)"])
(HR <- summary(x)$coefficients[, "exp(coef)"])
```

# association of smoking with TP53 mutation
```{r}
table(traits$smoking_status,traits$TP53)
cur_traits <- traits
cur_traits <- cur_traits %>%
  filter(simple_histology == "Adenocarcinoma") %>%
  mutate(smoking_status = ifelse(smoking_status == "Ex-smoker", "smoker",smoking_status),
         smoking_status = ifelse(smoking_status == "Current smoker", "smoker",smoking_status))
table(cur_traits$smoking_status, cur_traits$TP53)
fisher.test(cur_traits$smoking_status, cur_traits$TP53)

cur_traits <- traits[which(traits$smoking_status %in% c("Current smoker", "Non-smoker")),]
fisher.test(cur_traits$smoking_status,cur_traits$TP53)

cur_traits <- traits[which(traits$smoking_status %in% c("Ex-smoker", "Non-smoker")),]
fisher.test(cur_traits$smoking_status,cur_traits$TP53)
```


# TMB and TPS
```{r}
res <- fisher.test(traits$TMB_score,traits$TPS_score)
res$p.value
```

TMB and TPS are not asociated in fisher test `r round(res$p.value, digits = 2)`

# p53 mutaion
```{r}
ggforest(coxph( Surv(time_to_event, censoring) ~  TP53,data = traits %>% filter(simple_histology == "Adenocarcinoma")))
ggforest(coxph( Surv(time_to_event, censoring) ~  TP53,data = traits))
ggforest(coxph( Surv(time_to_event, censoring) ~  TP53,data = traits %>% filter(simple_histology == "Squamous cell carcinoma")))
x<- coxph( Surv(time_to_event, censoring) ~  TP53,data =  traits %>% filter(simple_histology == "Adenocarcinoma"))

(p_LUAD <- summary(x)$coefficients[, "Pr(>|z|)"])
(HR_LUAD <- summary(x)$coefficients[, "exp(coef)"])

y <- coxph( Surv(time_to_event, censoring) ~  TP53,data =  traits %>% filter(simple_histology == "Squamous cell carcinoma"))
(p_LUSC <- summary(y)$coefficients[, "Pr(>|z|)"])
(HR_LUSC <- summary(y)$coefficients[, "exp(coef)"])
```



TP53 has a HR of `r round(HR_LUAD, digits = 2)` and a p value of `r round(p_LUAD, digits = 2)` and CPS has a HR of `r round(HR_LUSC, digits = 2)` and a p value of `r round(p_LUSC, digits = 2)`

# percent of TPS > 1 % patients
```{r}
table(traits$TPS_score)/nrow(traits)
```

```{r}
cur_traits <- traits %>%
  filter(simple_histology == "Adenocarcinoma")
table(cur_traits$smoking_status, cur_traits$TPS_score)
```


# Reporting n

```{r}
dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/abundances_dat.rds"))
dens_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/densities_dat.rds"))
cello_dat <- readRDS(paste0(mount_path,"Figures/scores_for_sharing/cello_dat.rds"))
traits <- read_csv(paste0(mount_path,"Figures/scores_for_sharing/all_traits.csv"))
```

# densities & counts
```{r}
dens_dat %>%
  mutate(across(-immucan_id,~sum(!is.na(.))))
  pivot_longer(-immucan_id) %>%
  mutate(tech = str_split_i(name, "__",2)) %>%
  select(immucan_id,tech) %>%
  distinct() %>%
  group_by(tech) %>%
  dplyr::count(immucan_id) %>%
  summarise(total = sum(n))
```
# histology
```{r}
traits %>%
  dplyr::count(simple_histology)
```

# Smoking

```{r}
# imaging
traits %>%
  dplyr::count(smoking_status)

# RNA
RNA <- read.csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/RNASeq/Rout/tables/RNAseq/log2normcounts_adj.txt", sep = "\t")
RNA_pats <- str_split_i(colnames(RNA),".FIXT",1)
RNA_pats <- gsub("\\.","-", RNA_pats)
traits %>%
  filter(immucan_id %in% RNA_pats)%>%
  dplyr::count(smoking_status)
```


# stage

```{r}
# imaging
table(traits$simple_stage,traits$simple_histology)

# RNA
RNA <- read.csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/RNASeq/Rout/tables/RNAseq/log2normcounts_adj.txt", sep = "\t")
RNA_pats <- str_split_i(colnames(RNA),".FIXT",1)
RNA_pats <- gsub("\\.","-", RNA_pats)
traits %>%
  filter(immucan_id %in% RNA_pats)%>%
  dplyr::count(smoking_status)
```

# TPS

```{r}
# imaging
table(traits$simple_histology,traits$TPS_score)

# RNA
RNA <- read.csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/RNASeq/Rout/tables/RNAseq/log2normcounts_adj.txt", sep = "\t")
RNA_pats <- str_split_i(colnames(RNA),".FIXT",1)
RNA_pats <- gsub("\\.","-", RNA_pats)
cur_traits <- traits %>%
  filter(immucan_id %in% RNA_pats)
table(cur_traits$simple_histology,cur_traits$TPS_score)
```
# TMB
```{r}
# imaging
table(traits$simple_histology,traits$TMB_score)

# RNA
RNA <- read.csv("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/RNASeq/Rout/tables/RNAseq/log2normcounts_adj.txt", sep = "\t")
RNA_pats <- str_split_i(colnames(RNA),".FIXT",1)
RNA_pats <- gsub("\\.","-", RNA_pats)
cur_traits <- traits %>%
  filter(immucan_id %in% RNA_pats)
table(cur_traits$simple_histology,cur_traits$TMB_score)
```
# TMB-TPS
```{r}
cur_traits <- traits %>%
  mutate(TMB_TPS = ifelse(TMB_score == "low" & TPS_score == "positive", "TMBlow_TPShigh",ifelse(TMB_score == "high" & TPS_score == "negative", "TMBhigh_TPSlow", NA))) 
table(cur_traits$simple_histology,cur_traits$TMB_TPS)
```


# LUAD mutations
## TP53
```{r}
table(traits$EGFR,traits$simple_histology)
```


## EGFR
```{r}
table(traits$TP53,traits$simple_histology)
```

## KRAS
```{r}
table(traits$KRAS,traits$simple_histology)
```

## STK11 
```{r}
table(traits$STK11,traits$simple_histology)
```

## STK11 
```{r}
cur_traits <- traits %>%
  filter(smoking_status == "Current smoker")
table(cur_traits$TP53,cur_traits$simple_histology)

```

# LUSC
## TP53
```{r}
table(traits$PIK3CA,traits$simple_histology)
```

## CDKN2A
```{r}
table(traits$CDKN2A,traits$simple_histology)
```

