---
title: "Lung_NSCLC2_RNAseq"
author: "Flavia Marzetta"
date: "12/09/2023"
output: html
---

Analysis description:
 Here we want to explore the lung NSCLC2 RNAseq dataset (data were found in the custom cohort "Lung_2022" directory)
 Old Specta Lung (retrospective cohort, a.k.a. NSCLC2)
 the most updated clinical table was prepared by Daniel on 25.08.2023

1) Prepare gene counts (retrieve HTSeq count files of selected RNA-seq samples using the clinical data table provided by Daniel)
2) Prepare metadata
3) Filtering expressed genes
4) Library normalisation
5) (as alternative to 4) lib.kit batch correction and lib.size normalisation


NOTE: this script was taken from Flavia Marzetta and modified to work with our directories

```{r setup}
knitr::opts_chunk$set(echo = TRUE)
```

# Libraries
```{r}
library(edgeR)
library(limma)
library(sva)
library(ggplot2)
```

```{r}
mnt_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/"
```

# Directories
```{r}
if (!file.exists(paste0(mnt_path,"RNAseq/Rout/tables/RNAseq"))) dir.create(paste0(mnt_path,"RNAseq/Rout/tables/RNAseq"), recursive=T)
```



# 1. Prepare gene counts

The HTSeq count files were selected using the clinical data table provided by Daniel

## Select clinical data
```{r}
clinical_data <- read.delim("/mnt/immucan_volume/clinical_data/Lung_NSCLC2/20230825/NSCLC2_for_R_update_2024.csv", sep=",") #192  31
stopifnot(sum(duplicated(clinical_data$immucan_id))==0)
```



## Select NSCLC2 samples with available counts
```{r}
path_to_files_all <- dir(paste0(mnt_path,"RNAseq/STAR_counts/"), pattern = "counts.txt")
path_to_files <- path_to_files_all[grepl("^LUNG-.*", path_to_files_all)] #125
sampleID <- sapply(path_to_files, function(x) paste(strsplit(x, split = "-|_")[[1]][1:7], collapse="-"))
patientID <- sapply(path_to_files, function(x) paste(strsplit(x, split = "-")[[1]][1:3], collapse="-"))
sel.patient.idx <- patientID %in% clinical_data$immucan_id #125
path_to_files_sel <- path_to_files[sel.patient.idx]
```


## Read count tables
```{r}
htseq_files <- lapply(path_to_files_sel, function(x) {
  sample_counts <- read.table(paste0(mnt_path,"RNAseq/STAR_counts/", x), sep="\t")
  sample_name <- paste(strsplit(x, split = "-|_")[[1]][1:7], collapse="-")
  colnames(sample_counts) <- c("ensemblid", sample_name)
  sample_counts })
```



## Verify the table size (it has to be the same for all samples)
```{r}
unique(unlist(lapply(htseq_files,dim)))
```
58681     2

## Merge count tables and remove htseq summary rows
```{r}
rawData_wHeader <- Reduce(function(x,y) merge(x,y,by="ensemblid",all=TRUE, sort=F), htseq_files)
rownames(rawData_wHeader) <- rawData_wHeader$ensemblid
rawcounts_woHeader <- rawData_wHeader[-grep("__", rownames(rawData_wHeader)), -1]
```

58676   125

# 2. Prepare metadata

## Prepare clinical info
```{r}
sampleID.sel <- sampleID[sel.patient.idx]
patientID.sel <- patientID[sel.patient.idx]
ID.sel <- data.frame(immucan_id=patientID.sel, sample_id=sampleID.sel, fileID=names(sampleID.sel), row.names = NULL)
clinical_data_sel <- merge(ID.sel, clinical_data, by="immucan_id", sort=F)
clinical_data_sel <- clinical_data_sel[,c(1:2,4:ncol(clinical_data_sel), 3)]
rownames(clinical_data_sel) <- clinical_data_sel$sample_id
```


## Remove any duplicated patient from rawcounts and clinical data table
```{r}
clinical_data_sel$immucan_patient_id[duplicated(clinical_data_sel$immucan_patient_id)] #none
```

## remove metastatic sample (we want to keep only primary tumors)

```{r}
sample_to_exclude = "LUNG-NSCLC2-0677"
clinical_data_clean <- clinical_data_sel[!grepl(sample_to_exclude, clinical_data_sel$sample_id),]
rawcounts_unfiltered <- rawcounts_woHeader[,!grepl(sample_to_exclude, colnames(rawcounts_woHeader))]
stopifnot(all(rownames(clinical_data_clean) == colnames(rawcounts_unfiltered)))

short_id <- sapply(clinical_data_clean$immucan_id, function(x) gsub("^.*-", "", x))
metadata <- data.frame(short_id=short_id, clinical_data_clean[,1:ncol(clinical_data_clean)], row.names = rownames(clinical_data_clean))
```



## Retrieve other library and useful sequencing info
```{r}
# 1) qc_matrix.tsv
qc_matrix <- read.delim(paste0(mnt_path,"RNAseq/STAR_counts/qc_matrix.tsv"), sep="\t", header=T)
rownames(qc_matrix) <- qc_matrix$ID
qc_matrix_sel <- qc_matrix[rownames(metadata),c("qc_status"), drop=F]

# 1) sequencing_qc.csv
seq_matrix <- read.delim(paste0(mnt_path,"RNAseq/STAR_counts/sequencing_qc.csv"), sep=",", header = T)
rownames(seq_matrix) <- seq_matrix$sample_id
seq_matrix$lib.kit <- factor(ifelse(grepl("Takara",seq_matrix$assay_library), "takara", "kapa"))
seq_matrix_sel <- seq_matrix[rownames(metadata),c("sample_id","date_of_extraction", "number_of_reads", "percent_unique","lib.kit")]

# merge
qc_seq_matrix <- merge(seq_matrix_sel, qc_matrix_sel, by=0, sort=F)
qc_seq_matrix <- qc_seq_matrix[,-which(colnames(qc_seq_matrix)=="Row.names")]

# merge with clinical data
metadata_qc <- merge(metadata, qc_seq_matrix , by="sample_id", sort=F)
rownames(metadata_qc) <- metadata_qc$sample_id

## Save
write.table(rawcounts_unfiltered, file=paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/rawcounts_unfiltered.txt"), col.names = T, row.names = T, sep="\t", quote = F)
write.table(path_to_files_sel, file=paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/files_list.txt"), quote=F, col.names = FALSE, row.names = F)
write.table(metadata_qc, file=paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/metadata.txt"), sep="\t", col.names = T, row.names = T)

```


# 3. Filtering expressed genes

## Read in data and metadata

```{r}
rawcounts_unfiltered <- read.delim(paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/rawcounts_unfiltered.txt"), check.names = FALSE)
metadata <- read.delim(paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/metadata.txt"))
stopifnot(all(colnames(rawcounts_unfiltered)==rownames(metadata)))
```

## Remove genes not expressed in one batch
```{r}
ind_Takara = grep("takara", metadata$lib.kit)
ind_Kapa = grep("kapa", metadata$lib.kit)
keep_Takara <- rowSums(rawcounts_unfiltered[,ind_Takara]) > 0 # at least 1 count in at least 1 takara sample
rawcounts_filt1 <- rawcounts_unfiltered[keep_Takara,]
keep_Kapa <- rowSums(rawcounts_filt1[,ind_Kapa]) > 0 # at least 1 count in at least 1 kapa sample
rawcounts_filt2 <- rawcounts_filt1[keep_Kapa,]
```



## Keep genes with at least 1 cpm in at least 10% samples
here we apply a filtering that is more stringent than the minimal pre-filtering proposed by the DESeq2 workflow,
i.e. keep only rows that have at least 10 reads total (http://bioconductor.org/packages/devel/bioc/vignettes/DESeq2/inst/doc/DESeq2.html#pre-filtering)
```{r}
cpm_counts <- edgeR::cpm(rawcounts_filt2, normalized.lib.sizes=TRUE)
cpm.val <- 1 # Define the cutoff value
min.samplenum <- round(ncol(cpm_counts)/100*10, 0)
isexpr <- rowSums(cpm_counts>=cpm.val) >= min.samplenum
rawcounts_filt=rawcounts_filt2[isexpr,] #24376 (46% of the features)
write.table(rawcounts_filt, file=paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/rawcounts_filtered.txt"), col.names = T, row.names = T, sep="\t", quote = F)
```


# 4. Libraries normalization
```{r}
# i. Read in raw filtered counts
rawcounts_filt <- read.delim(paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/rawcounts_filtered.txt"), check.names = FALSE)

# ii. TMM normalization and voom transformation
dge <- edgeR::DGEList(counts=rawcounts_filt)
dge <- edgeR::calcNormFactors(dge, method="TMM")
log2.norm.counts <- cpm(dge, log=TRUE) #TMM normalized expression values on the log2 scale
write.table(log2.norm.counts, file=paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/log2normcounts.txt"), col.names = T, row.names = T, sep="\t", quote = F)

# iii. Vst transformation (~roughly log2(normalized counts) but with variance stabilization)
library(DESeq2)
vst_counts <- data.frame(vst(as.matrix(rawcounts_filt)))
colnames(vst_counts) <- gsub("\\.", "-", colnames(vst_counts))
write.table(vst_counts, file=paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/vstcounts.txt"), col.names = T, row.names = T, sep="\t", quote = F)
```


# 5. Lib.kit batch correction and normalisation
```{r}
# i. Read in raw filtered counts and metadata
rawcounts_filt <- read.delim(paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/rawcounts_filtered.txt"), check.names = FALSE)
metadata <- read.delim(paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/metadata.txt"), stringsAsFactors = T)
stopifnot(all(colnames(rawcounts_filt)==rownames(metadata)))

# ii. Correct for lib.size using ComBat_seq
# run combatseq on raw, filtered counts (since we didn't provide any 'group', full_mode T or F doesn't not matter)
adjusted_data <- sva::ComBat_seq(counts=as.matrix(rawcounts_filt), batch = metadata$lib.kit)
write.table(adjusted_data, file=paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/rawcounts_filtered_adj.txt"), col.names = T, row.names = T, sep="\t", quote = F)

# iii. TMM normalization and voom transformation
dge.adj <- edgeR::DGEList(counts=adjusted_data)
dge.adj <- edgeR::calcNormFactors(dge.adj, method="TMM")
log2.norm.counts.adj <- cpm(dge.adj, log=TRUE) #TMM normalized expression values on the log2 scale
write.table(log2.norm.counts.adj, file=paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/log2normcounts_adj.txt"), col.names = T, row.names = T, sep="\t", quote = F)

# # this correspond to Jane's table  "NSCLC2_combatseq_24220_genes_normalized_counts_N=124_2023-10-07.csv"
# norm.counts.adj <- cpm(dge.adj, log=FALSE)

# iv. Vst transformation (~roughly log2(normalized counts) but with variance stabilization)
vst_counts_adj <- data.frame(vst(as.matrix(adjusted_data)))
colnames(vst_counts_adj) <- gsub("\\.", "-", colnames(vst_counts_adj))
write.table(vst_counts_adj, file=paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/vstcounts_adj.txt"), col.names = T, row.names = T, sep="\t", quote = F)

```



# 6. Filtering protein-coding genes
```{r}
# genecode annotations
annot <- read.csv(paste0(mnt_path,"RNAseq/annotations/gencode.v29.annotation.genes.txt"), sep = "\t")
#colnames(annot) <- c("gene_type","gene_id")
annot_type <- annot[,c("gene_id", "gene_type")]

# 6.1 filter unadjusted counts
#################################
vst_counts <- read.delim(paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/vstcounts.txt"), check.names = F)
vst_counts_annot <- merge(vst_counts, annot_type, by.x=0, by.y="gene_id", all.x=T, all.y=F, sort=F)
vst_counts_prot <- vst_counts_annot[vst_counts_annot$gene_type=="protein_coding",]
rownames(vst_counts_prot) <- vst_counts_prot$Row.names
vst_counts_prot <- vst_counts_prot[,-which(colnames(vst_counts_prot) %in% c("Row.names", "gene_type"))]
write.table(vst_counts_prot, file=paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/vstcounts_proteins.txt"), sep="\t")

# raw counts
rawcounts_filt <- read.delim(paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/rawcounts_filtered.txt"), check.names = FALSE)
counts_annot <- merge(rawcounts_filt, annot_type, by.x=0, by.y="gene_id", all.x=T, all.y=F, sort=F)
counts_prot <- counts_annot[counts_annot$gene_type=="protein_coding",]
rownames(counts_prot) <- counts_prot$Row.names
counts_prot <- counts_prot[,-which(colnames(counts_prot) %in% c("Row.names", "gene_type"))]
write.table(counts_prot, file=paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/rawcounts_proteins.txt"), sep="\t")

# log2 normalised counts
log2normcounts <- read.delim(paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/log2normcounts.txt"), check.names = F)
log2normcounts_annot <- merge(log2normcounts, annot_type, by.x=0, by.y="gene_id", all.x=T, all.y=F, sort=F)
log2normcounts_prot <- log2normcounts_annot[log2normcounts_annot$gene_type=="protein_coding",]
rownames(log2normcounts_prot) <- log2normcounts_prot$Row.names
log2normcounts_prot <- log2normcounts_prot[,-which(colnames(log2normcounts_prot) %in% c("Row.names", "gene_type"))]
write.table(log2normcounts_prot, file=paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/log2normcounts_proteins.txt"), sep="\t")

# 6.2 filter batch-adjusted counts
#################################
vst_counts_adj <- read.delim(paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/vstcounts_adj.txt"), check.names = F)
vst_counts_adj_annot <- merge(vst_counts_adj, annot_type, by.x=0, by.y="gene_id", all.x=T, all.y=F, sort=F)
vst_counts_adj_prot <- vst_counts_adj_annot[vst_counts_adj_annot$gene_type=="protein_coding",]
rownames(vst_counts_adj_prot) <- vst_counts_adj_prot$Row.names
vst_counts_adj_prot <- vst_counts_adj_prot[,-which(colnames(vst_counts_adj_prot) %in% c("Row.names", "gene_type"))]
write.table(vst_counts_adj_prot, file=paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/vstcounts_adj_proteins.txt"), sep="\t")
```


