---
title: "RNA QC"
author: "Flavia Marzettat"
date: "2024-06-14"
output: html_document
---

Analysis description:
Here we want to explore the lung NSCLC2 RNAseq dataset (data were found in the custom cohort "IMMUcan_NSCLC2" directory)
Old Specta Lung (retrospective cohort, a.k.a. NSCLC2)
the most updated clinical table was prepared by Daniel on 25.08.2023

1) Plotting counts (distribution before/after TMM norm, barplot total raw counts, boxplot counts before/after TMM norm)
2) PCA (biplots, hclust) using: 
vst counts (lib.norm + log2 + variance stabilization + scaling) ("/PCA")
log2CPM (lib.norm + log2 + scaling) ("/PCA_normcounts")
vst counts UNSCALED (lib.norm + log2 + variance stabilization)  ("/PCA_unscaled")

## Set wkdir ##
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())
```

## Libraries ##
```{r}
library(edgeR)
library(limma)
library(PCAtools)
library(scales)
library(ComplexHeatmap)
library(data.table)
library(gridExtra)
library(RColorBrewer)
library(gplots)
library(GGally)
```


```{r}
mnt_path <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/"
out_folder <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/RNASeq/Rout/plots/QC/"

if (!dir.exists(out_folder)) {
  dir.create(path = out_folder, recursive = TRUE)
}
```


# 1. Plotting counts

## Read in raw filtered counts, metadata and qc
```{r}
rawcounts_unfiltered <- read.delim(paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/rawcounts_unfiltered.txt"), check.names = FALSE)
rawcounts_filt <- read.delim(paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/rawcounts_filtered.txt"), check.names = FALSE)
norm.counts <- read.delim(paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/log2normcounts.txt"), check.names = FALSE)
metadata <- read.delim(paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/metadata.txt"), stringsAsFactors = T)
stopifnot(rownames(metadata)==colnames(rawcounts_unfiltered))

lib.kit.tb <- metadata[,"lib.kit",drop=F]
myCol <- hue_pal()(2)
```


## Raw unfiltered
```{r}
log2.raw.unfilt <- log2(rawcounts_unfiltered+1)
log2.raw.unfilt.long <- melt(t(log2.raw.unfilt))
names(log2.raw.unfilt.long) <- c("sample_id", "ensemblID", "counts")
lib.kit.tb <- metadata[,"lib.kit",drop=F]
lib.kit <- lib.kit.tb$lib.kit[match(log2.raw.unfilt.long$sample_id, rownames(lib.kit.tb))]
pdf(paste0(out_folder,"density_plot_raw_unfilt_counts.pdf"), width=10, height = 5)
ggplot(log2.raw.unfilt.long, aes(x=counts, col=sample_id), size=2) + geom_density() + xlab("Log2 Counts") + theme_light() +
  scale_color_manual(name="lib.kit", values = myCol[lib.kit]) +
  theme(legend.position="none")+
  ggtitle("Raw unfiltered counts")
dev.off()
```

## Raw filtered
```{r}
log2.raw.filt <- log2(rawcounts_filt+1)
log2.raw.filt.long <- melt(t(log2.raw.filt))
names(log2.raw.filt.long) <- c("sample_id", "ensemblID", "counts")
lib.kit <- lib.kit.tb$lib.kit[match(log2.raw.filt.long$sample_id, rownames(lib.kit.tb))]
pdf(paste0(out_folder,"density_plot_raw_filt_counts.pdf"), width=10, height = 5)
ggplot(log2.raw.filt.long, aes(x=counts, col=sample_id)) + geom_density() + xlab("Log2 Counts") + theme_light() +
  scale_color_manual(name="lib.kit", values = myCol[lib.kit]) +
  theme(legend.position="none") + ggtitle("Raw filtered counts")
dev.off()
```

## Norm + filtered
```{r}

norm.long <- melt(t(norm.counts)) #TMM normalized expression values on the log2 scale
names(norm.long) <- c("sample_id", "ensemblID", "counts")
lib.kit <- lib.kit.tb$lib.kit[match(norm.long$sample_id, rownames(lib.kit.tb))]
pdf(paste0(out_folder,"density_plot_norm_counts.pdf"), width=10, height = 5)
ggplot(norm.long, aes(x=counts, col=sample_id)) + geom_density() + xlab("Log2 Counts") + theme_light() +
  scale_color_manual(name="lib.kit", values = myCol[lib.kit]) +
  theme(legend.position="none") + ggtitle("Normalised filtered counts")
dev.off()
```

## Plotting density of counts split by sample
```{r}
norm.long.by.sample <- split(norm.long, norm.long$sample_id)
gg_list <- lapply(1:length(norm.long.by.sample), function(i) {
  longID=names(norm.long.by.sample)[i]
  ID <- strsplit(longID, split = "-")[[1]][3]
  ggplot(norm.long.by.sample[[longID]], aes(x=counts)) +
    geom_density() + xlab("Log2 Counts") +
    theme_light() +
    ggtitle(paste0("Sample_", ID)) +
    theme(legend.key.size = unit(0.5,"line"),
          axis.text = element_text(size=6),
          axis.title = element_text(size=8),
          plot.title = element_text(size=10))
})
ml <- marrangeGrob(grobs=gg_list, nrow=6, ncol=6, top=FALSE)
ml
ggsave(ml, filename = paste0(out_folder,"density_plots_by_sample.pdf"), width=15, height = 15)
```

## Barplot total counts
```{r}
# before filtering
tot_counts <- data.frame(tot_counts_Million=apply(rawcounts_unfiltered, 2, function(x) sum(x)/10^6))
tot_counts$sample_id <- rownames(tot_counts)
data_to_plot <- Reduce(function(x,y) merge(x,y,by="sample_id",sort=F), list(tot_counts,metadata))
data_to_plot$short_id <- as.character(data_to_plot$short_id)
# data_to_plot$lib_kit <- factor(data_to_plot$lib_kit, levels=c("kapa", "takara"))
pdf(paste0(out_folder,"barplot_HTSeq_totcounts.pdf"), height = 7, width = 15)
ggplot(data_to_plot, aes(fill=lib.kit, y=tot_counts_Million, x=short_id)) + #x=reorder(short_id, -cohort))) +
  geom_bar(stat = "identity") +
  facet_grid(~ lib.kit, scales = "free", space = "free") + #, space = "free_x") +
  ylab("# Counts on genes (Millions)") +
  xlab("") +
  theme(axis.text.x = element_text(angle = 90)) +
  ggtitle(paste("Total HTSeq-counts"))
dev.off()

```



## Boxplot of read counts
```{r}
pdf(paste0(out_folder,"boxplots_counts.pdf"), width = 20)
par(mfrow=c(2, 1))
par(mar=c(2, 4.1, 4.1, 2.1))
boxplot(log2.raw.filt, outline=F, las=2, cex.axis=.3, ylab='Log2 counts', names=metadata$short_id, main='Filtered counts before normalisation')
boxplot(norm.counts, outline=F, las=2, cex.axis=.3, ylab='Log2 counts', names=metadata$short_id, main='Filtered counts after TMM normalisation')
dev.off()

```


# 2. PCA

```{r}
out_folder <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/RNAseq/Rout/plots/QC/PCA/"

if (!dir.exists(out_folder)) {
  dir.create(path = out_folder)
}
```

## Read in vst-counts and metadata
```{r}
vst_counts <- read.delim(paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/vstcounts.txt"), check.names = F)
metadata <- read.delim(paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/metadata.txt"), stringsAsFactors = T)
stopifnot(rownames(metadata)==colnames(vst_counts))

```


## clean metadata and run PCA
```{r}
metadata_clean <- data.frame(metadata[,c("sample_id", "short_id", "immucan_id",
                                         "lib.kit", 
                                         "gender", "simple_histology", "simple_stage",
                                         "metastatic_treatment","neo_adjuvant_treatment", "adjuvant_treatment",
                                         "age_at_diagnosis", 
                                         "side_localization", "primary_localization", "smoking_status", "environmental_exposure")])
metadata_clean$date_of_extraction_num <- as.numeric(metadata$date_of_extraction)

uq_values <- apply(vst_counts,1,function(x) length(unique(x)))==1
p <- pca(vst_counts[!uq_values,], metadata=metadata_clean, scale = TRUE, transposed = FALSE)
pca_df <- p$rotated
```



## screeplot
```{r}
elbow <- findElbowPoint(p$variance)
pdf(paste0(out_folder,"pca_screeplot.pdf"))
screeplot(p, axisLabSize = 18, titleLabSize = 22, components=getComponents(p)[1:15], vline = elbow)+
  geom_label(aes(x = elbow - 1, y = 50,
                 label = 'Elbow point', vjust = -1, size = 8))
dev.off()
```



## pairsplot
```{r, message=FALSE}
clin_var <- colnames(metadata_clean)[grepl("simple|lib.kit|age|smoking|adjuvant_treatment", colnames(metadata_clean))]
pdf(paste0(out_folder,"pca_pairsplot.pdf"), width=6, height=6)
lapply(clin_var, function(trait) {
  pp <- pairsplot(p,
                  components = getComponents(p, c(1:10)),
                  triangle = TRUE, trianglelabSize = 12,
                  hline = 0, vline = 0,
                  pointSize = 0.4,
                  gridlines.major = FALSE, gridlines.minor = FALSE,
                  colby = trait,
                  title = trait,
                  plotaxes = FALSE,
                  margingaps = unit(c(-0.01, -0.01, -0.01, -0.01), 'cm'))
  pp
})
dev.off()
pdf(paste0(out_folder,"pca_pairsplot_PC1toPC4.pdf"), width=8, height=8)
lapply(clin_var, function(trait) {
  pp <- pairsplot(p,
                  components = getComponents(p, c(1:4)),
                  triangle = TRUE, trianglelabSize = 12,
                  hline = 0, vline = 0,
                  pointSize = 1,
                  gridlines.major = FALSE, gridlines.minor = FALSE,
                  colby = trait,
                  title = trait,
                  plotaxes = FALSE,legendPosition = "none",
                  margingaps = unit(c(-0.01, -0.01, -0.01, -0.01), 'cm'))
  pp
})
dev.off()
```




## biplots 

### PC1 vs PC2
```{r}
pdf(paste0(out_folder,"pca_PC1vsPC2_labeled.pdf"), width=6, height=4)
lapply(colnames(metadata_clean)[-(1:3)], function(trait) {
  bp <- biplot(p,
               x="PC1",y="PC2",
               lab = p$metadata$short_id,
               colby = trait,
               hline = 0, vline = 0,
               pointSize=2,
               legendPosition = 'right',
               legendTitleSize = 8,
               legendLabSize = 6,
               legendIconSize = 2) + ggtitle(trait)
  bp
})
dev.off()
```



### PC3 vs PC4
```{r}
pdf(paste0(out_folder,"pca_PC3vsPC4_labeled.pdf"), width=6, height=4)
lapply(colnames(metadata_clean)[-(1:3)], function(trait) {
  bp <- biplot(p,
               x="PC3",y="PC4",
               lab = p$metadata$short_id,
               colby = trait,
               hline = 0, vline = 0,
               pointSize=2,
               legendPosition = 'right',
               legendTitleSize = 8,
               legendLabSize = 6,
               legendIconSize = 2) + ggtitle(trait)
  bp
})
dev.off()
```





## biplots unlabeled

### PC1 vs PC2
```{r}

pdf(paste0(out_folder,"pca_PC1vsPC2.pdf"), width=6, height=4)
lapply(colnames(metadata_clean)[-(1:3)], function(trait) {
  bp <- biplot(p,
               x="PC1",y="PC2",
               lab = NULL,
               colby = trait,
               hline = 0, vline = 0,
               pointSize=2,
               legendPosition = 'right',
               legendTitleSize = 8,
               legendLabSize = 6,
               legendIconSize = 2) + ggtitle(trait)
  bp
})
dev.off()
```


### PC3 vs PC4
```{r}
pdf(paste0(out_folder,"pca_PC3vsPC4.pdf"), width=6, height=4)
lapply(colnames(metadata_clean)[-(1:3)], function(trait) {
  bp <- biplot(p,
               x="PC3",y="PC4",
               lab = NULL,
               colby = trait,
               hline = 0, vline = 0,
               pointSize=2,
               legendPosition = 'right',
               legendTitleSize = 8,
               legendLabSize = 6,
               legendIconSize = 2) +
    ggtitle(trait)
  bp
})
dev.off()
```





## pca hclust
```{r}
# Retrieve PCA scores
scores <- p$rotated
rownames(scores) <- p$metadata$short_id
# Prepare colors for side bar annotation
ggplot_colors2 <- hue_pal()(2)
ggplot_colors <- hue_pal()(10)
rb_colors <- rainbow(4)
col.anno <- HeatmapAnnotation(stage = p$metadata$simple_stage,
                              adjuvant_treatment=p$metadata$adjuvant_treatment,
                              histology = p$metadata$simple_histology,
                              gender = p$metadata$gender,
                              smoking_status = p$metadata$smoking_status,
                              lib.kit = p$metadata$lib.kit,
                              col = list(stage=c("I"="yellow",
                                                 "II"="orange",
                                                 "III"="red",
                                                 "IV"="brown"),
                                         adjuvant_treatment=c("No"=ggplot_colors2[1], "Yes"=ggplot_colors2[2]),
                                         histology=c("Adenocarcinoma"=ggplot_colors[2],
                                                     "Squamous cell carcinoma"=ggplot_colors[8],
                                                     "Adenosquamous carcinoma"=ggplot_colors[3],
                                                     "LCNEC" = ggplot_colors[5],
                                                     "Other"=ggplot_colors[4],
                                                     "unknown" = "gray"),
                                         gender = c("Female" = "pink", "Male"="cyan"),
                                         smoking_status = c("Non-smoker"="yellow", "Ex-smoker"="orange", "Current smoker"="red"),
                                         lib.kit = c("kapa"=ggplot_colors2[1], "takara"=ggplot_colors2[2])))
```


## pca hclust using top 10 PCs
```{r}
pdf(paste0(out_folder,"pca_hclust_top10PCs_avgClust.pdf"), width=15, height=7)
sel_scores <- as.matrix(scores[,1:10])
Heatmap(t((sel_scores)),
        name = "PCA score",
        row_title = "PCs", row_title_side = "left",
        column_title = "Hierarchical clustering of samples based on top 10 PCs",
        column_title_gp = gpar(fontsize = 11),
        top_annotation = col.anno,
        cluster_rows = FALSE,
        clustering_method_columns = "average",
        show_row_names = T,
        row_names_gp = gpar(fontsize = 8),
        column_names_gp = gpar(fontsize = 5))
dev.off()

pdf(paste0(out_folder,"pca_hclust_top8PCs_complClust.pdf"), width=15, height=7)
sel_scores <- as.matrix(scores[,1:10])
Heatmap(t((sel_scores)),
        name = "PCA score",
        row_title = "PCs", row_title_side = "left",
        column_title = "Hierarchical clustering of samples based on top 10 PCs",
        column_title_gp = gpar(fontsize = 11),
        top_annotation = col.anno,
        cluster_rows = FALSE,
        clustering_method_columns = "complete",
        show_row_names = T,
        row_names_gp = gpar(fontsize = 8),
        column_names_gp = gpar(fontsize = 5))
dev.off()

```


## pca hclust using PCs that explains 90% variance (sum(p$variance[1:74]))
```{r}
pdf(paste0(out_folder,"pca_hclust_PCs90percVar_avglClust.pdf"), width=15, height=9)
sel_scores <- as.matrix(scores[,1:74])
Heatmap(t((sel_scores)),
        name = "PCA score",
        row_title = "PCs", row_title_side = "left",
        column_title = "Hierarchical clustering of samples based on PCs that explain 90% variance",
        column_title_gp = gpar(fontsize = 11),
        top_annotation = col.anno,
        cluster_rows = FALSE,
        clustering_method_columns = "average",
        show_row_names = T,
        row_names_gp = gpar(fontsize = 7),
        column_names_gp = gpar(fontsize = 5))
dev.off()

pdf(paste0(out_folder,"pca_hclust_PCs90percVar_complClust.pdf"), width=15, height=9)
sel_scores <- as.matrix(scores[,1:74])
Heatmap(t((sel_scores)),
        name = "PCA score",
        row_title = "PCs", row_title_side = "left",
        column_title = "Hierarchical clustering of samples based on PCs that explain 90% variance",
        column_title_gp = gpar(fontsize = 11),
        top_annotation = col.anno,
        cluster_rows = FALSE,
        clustering_method_columns = "complete",
        show_row_names = T,
        row_names_gp = gpar(fontsize = 7),
        column_names_gp = gpar(fontsize = 5))
dev.off()
```


## Run F.test
table of stat (ANOVA/t.test)
```{r}
sel_traits <- c("simple_stage", "side_localization", "primary_localization",
                "age_at_diagnosis", "gender", "smoking_status",
                "neo_adjuvant_treatment",  "adjuvant_treatment", "metastatic_treatment", "simple_histology",
                "lib.kit")
allTraits_allPCs <- lapply(sel_traits, function(trait) {
  allPCs <- lapply(colnames(p$rotated)[1:6], function(pc) {
    res.aov <- summary(aov(p$rotated[,pc] ~ p$metadata[,trait]))[[1]][["Pr(>F)"]][1]
    return(res.aov)
  })
  names(allPCs) <- colnames(p$rotated)[1:6]
  return(allPCs)
})
names(allTraits_allPCs) <- sel_traits
aov_traits_PCs <- do.call(rbind.data.frame, allTraits_allPCs)

pdf(paste0(out_folder,"Ftest_PC_traits.pdf"), width=7, height=5)
cellnotes <- ifelse(aov_traits_PCs<0.001, "***",
                    ifelse(aov_traits_PCs<0.01, "**",
                           ifelse(aov_traits_PCs<0.05, "*", "")))
myCol <- brewer.pal(n=9, name = "Reds")
heatmap.2(as.matrix(-log10(aov_traits_PCs)),
          cellnote = cellnotes,
          notecol="gray1",
          notecex=1.5,
          margin=c(5,15), 
          density.info="none", trace="none",
          Colv = FALSE, Rowv = TRUE, dendrogram="row",
          key=T, key.xlab="-log10 pvalue (F-test)",
          col=myCol,
          sepcolor="white", colsep=c(1:ncol(aov_traits_PCs)),
          rowsep=c(1:nrow(aov_traits_PCs)), sepwidth=c(0.01,0.01))
dev.off()    
```

    
# 3. PCA using logCPM (TMM-normalisation)

```{r}
out_folder <- "/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/RNASeq/Rout/plots/QC/PCA_normcounts/"
if (!dir.exists(out_folder)) {
  dir.create(path = out_folder)
}
```

## Read in raw filtered counts and metadata
```{r}
normcounts <- read.delim(paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/log2normcounts.txt"), check.names = FALSE)
metadata <- read.delim(paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/metadata.txt"), stringsAsFactors = T)
stopifnot(rownames(metadata)==colnames(normcounts))
```



# clean metadata and run PCA
```{r}
metadata_clean <- data.frame(metadata[,c("sample_id", "short_id", "immucan_id",
                                         "lib.kit", 
                                         "gender", "simple_histology", "simple_stage",
                                         "metastatic_treatment","neo_adjuvant_treatment", "adjuvant_treatment",
                                         "age_at_diagnosis",
                                         "side_localization", "primary_localization", "smoking_status", "environmental_exposure")])
metadata_clean$date_of_extraction_num <- as.numeric(metadata$date_of_extraction)

uq_values <- apply(normcounts,1,function(x) length(unique(x)))==1
p <- pca(normcounts[!uq_values,], metadata=metadata_clean, scale = TRUE, transposed = FALSE)
pca_df <- p$rotated
```

## screeplot
```{r}
elbow <- findElbowPoint(p$variance)
pdf(paste0(out_folder,"pca_screeplot.pdf"))
screeplot(p, axisLabSize = 18, titleLabSize = 22, components=getComponents(p)[1:15], vline = elbow)+
  geom_label(aes(x = elbow - 1, y = 50,
                 label = 'Elbow point', vjust = -1, size = 8))
dev.off()
```



## pairsplot
```{r, message=FALSE}
clin_var <- colnames(metadata_clean)[grepl("simple|lib.kit|age|smoking|adjuvant_treatment", colnames(metadata_clean))]
pdf(paste0(out_folder,"pca_pairsplot.pdf"), width=6, height=6)
lapply(clin_var, function(trait) {
  pp <- pairsplot(p,
                  components = getComponents(p, c(1:10)),
                  triangle = TRUE, trianglelabSize = 12,
                  hline = 0, vline = 0,
                  pointSize = 0.4,
                  gridlines.major = FALSE, gridlines.minor = FALSE,
                  colby = trait,
                  title = trait,
                  plotaxes = FALSE,
                  margingaps = unit(c(-0.01, -0.01, -0.01, -0.01), 'cm'))
  pp
})
dev.off()
pdf(paste0(out_folder,"pca_pairsplot_PC1toPC4.pdf"), width=8, height=8)
lapply(clin_var, function(trait) {
  pp <- pairsplot(p,
                  components = getComponents(p, c(1:4)),
                  triangle = TRUE, trianglelabSize = 12,
                  hline = 0, vline = 0,
                  pointSize = 1,
                  gridlines.major = FALSE, gridlines.minor = FALSE,
                  colby = trait,
                  title = trait,
                  plotaxes = FALSE,legendPosition = "none",
                  margingaps = unit(c(-0.01, -0.01, -0.01, -0.01), 'cm'))
  pp
})
dev.off()
```




## biplots 

### PC1 vs PC2
```{r}
pdf(paste0(out_folder,"pca_PC1vsPC2_labeled.pdf"), width=6, height=4)
lapply(colnames(metadata_clean)[-(1:3)], function(trait) {
  bp <- biplot(p,
               x="PC1",y="PC2",
               lab = p$metadata$short_id,
               colby = trait,
               hline = 0, vline = 0,
               pointSize=2,
               legendPosition = 'right',
               legendTitleSize = 8,
               legendLabSize = 6,
               legendIconSize = 2) + ggtitle(trait)
  bp
})
dev.off()
```



### PC3 vs PC4
```{r}
pdf(paste0(out_folder,"pca_PC3vsPC4_labeled.pdf"), width=6, height=4)
lapply(colnames(metadata_clean)[-(1:3)], function(trait) {
  bp <- biplot(p,
               x="PC3",y="PC4",
               lab = p$metadata$short_id,
               colby = trait,
               hline = 0, vline = 0,
               pointSize=2,
               legendPosition = 'right',
               legendTitleSize = 8,
               legendLabSize = 6,
               legendIconSize = 2) + ggtitle(trait)
  bp
})
dev.off()
```


## biplots unlabeled

### PC1 vs PC2
```{r}

pdf(paste0(out_folder,"pca_PC1vsPC2.pdf"), width=6, height=4)
lapply(colnames(metadata_clean)[-(1:3)], function(trait) {
  bp <- biplot(p,
               x="PC1",y="PC2",
               lab = NULL,
               colby = trait,
               hline = 0, vline = 0,
               pointSize=2,
               legendPosition = 'right',
               legendTitleSize = 8,
               legendLabSize = 6,
               legendIconSize = 2) + ggtitle(trait)
  bp
})
dev.off()
```


### PC3 vs PC4
```{r}
pdf(paste0(out_folder,"pca_PC3vsPC4.pdf"), width=6, height=4)
lapply(colnames(metadata_clean)[-(1:3)], function(trait) {
  bp <- biplot(p,
               x="PC3",y="PC4",
               lab = NULL,
               colby = trait,
               hline = 0, vline = 0,
               pointSize=2,
               legendPosition = 'right',
               legendTitleSize = 8,
               legendLabSize = 6,
               legendIconSize = 2) +
    ggtitle(trait)
  bp
})
dev.off()
```

## pca hclust
```{r}
# Retrieve PCA scores
scores <- p$rotated
rownames(scores) <- p$metadata$short_id
# Prepare colors for side bar annotation
ggplot_colors2 <- hue_pal()(2)
ggplot_colors <- hue_pal()(10)
rb_colors <- rainbow(4)
col.anno <- HeatmapAnnotation(stage = p$metadata$simple_stage,
                              adjuvant_treatment=p$metadata$adjuvant_treatment,
                              histology = p$metadata$simple_histology,
                              gender = p$metadata$gender,
                              smoking_status = p$metadata$smoking_status,
                              lib.kit = p$metadata$lib.kit,
                              col = list(stage=c("I"="yellow",
                                                 "II"="orange",
                                                 "III"="red",
                                                 "IV"="brown"),
                                         adjuvant_treatment=c("No"=ggplot_colors2[1], "Yes"=ggplot_colors2[2]),
                                         histology=c("Adenocarcinoma"=ggplot_colors[2],
                                                     "Squamous cell carcinoma"=ggplot_colors[8],
                                                     "Adenosquamous carcinoma"=ggplot_colors[3],
                                                     "LCNEC" = ggplot_colors[5],
                                                     "Other"=ggplot_colors[4],
                                                     "unknown" = "gray"),
                                         gender = c("Female" = "pink", "Male"="cyan"),
                                         smoking_status = c("Non-smoker"="yellow", "Ex-smoker"="orange", "Current smoker"="red"),
                                         lib.kit = c("kapa"=ggplot_colors2[1], "takara"=ggplot_colors2[2])))
```

## pca hclust using top 10 PCs
```{r}
pdf(paste0(out_folder,"pca_hclust_top10PCs_avgClust.pdf"), width=15, height=7)
sel_scores <- as.matrix(scores[,1:10])
Heatmap(t((sel_scores)),
        name = "PCA score",
        row_title = "PCs", row_title_side = "left",
        column_title = "Hierarchical clustering of samples based on top 10 PCs",
        column_title_gp = gpar(fontsize = 11),
        top_annotation = col.anno,
        cluster_rows = FALSE,
        clustering_method_columns = "average",
        show_row_names = T,
        row_names_gp = gpar(fontsize = 8),
        column_names_gp = gpar(fontsize = 5))
dev.off()

pdf(paste0(out_folder,"pca_hclust_top8PCs_complClust.pdf"), width=15, height=7)
sel_scores <- as.matrix(scores[,1:10])
Heatmap(t((sel_scores)),
        name = "PCA score",
        row_title = "PCs", row_title_side = "left",
        column_title = "Hierarchical clustering of samples based on top 10 PCs",
        column_title_gp = gpar(fontsize = 11),
        top_annotation = col.anno,
        cluster_rows = FALSE,
        clustering_method_columns = "complete",
        show_row_names = T,
        row_names_gp = gpar(fontsize = 8),
        column_names_gp = gpar(fontsize = 5))
dev.off()

```

## pca hclust using PCs that explains 90% variance (sum(p$variance[1:74]))
```{r}
pdf(paste0(out_folder,"pca_hclust_PCs90percVar_avglClust.pdf"), width=15, height=9)
sel_scores <- as.matrix(scores[,1:74])
Heatmap(t((sel_scores)),
        name = "PCA score",
        row_title = "PCs", row_title_side = "left",
        column_title = "Hierarchical clustering of samples based on PCs that explain 90% variance",
        column_title_gp = gpar(fontsize = 11),
        top_annotation = col.anno,
        cluster_rows = FALSE,
        clustering_method_columns = "average",
        show_row_names = T,
        row_names_gp = gpar(fontsize = 7),
        column_names_gp = gpar(fontsize = 5))
dev.off()

pdf(paste0(out_folder,"pca_hclust_PCs90percVar_complClust.pdf"), width=15, height=9)
sel_scores <- as.matrix(scores[,1:74])
Heatmap(t((sel_scores)),
        name = "PCA score",
        row_title = "PCs", row_title_side = "left",
        column_title = "Hierarchical clustering of samples based on PCs that explain 90% variance",
        column_title_gp = gpar(fontsize = 11),
        top_annotation = col.anno,
        cluster_rows = FALSE,
        clustering_method_columns = "complete",
        show_row_names = T,
        row_names_gp = gpar(fontsize = 7),
        column_names_gp = gpar(fontsize = 5))
dev.off()
```

## Run F.test
table of stat (ANOVA/t.test)
```{r}
sel_traits <- c("simple_stage", "side_localization", "primary_localization",
                "age_at_diagnosis", "gender", "smoking_status",
                "neo_adjuvant_treatment",  "adjuvant_treatment", "metastatic_treatment", "simple_histology",
                "lib.kit")
allTraits_allPCs <- lapply(sel_traits, function(trait) {
  allPCs <- lapply(colnames(p$rotated)[1:6], function(pc) {
    res.aov <- summary(aov(p$rotated[,pc] ~ p$metadata[,trait]))[[1]][["Pr(>F)"]][1]
    return(res.aov)
  })
  names(allPCs) <- colnames(p$rotated)[1:6]
  return(allPCs)
})
names(allTraits_allPCs) <- sel_traits
aov_traits_PCs <- do.call(rbind.data.frame, allTraits_allPCs)

pdf(paste0(out_folder,"Ftest_PC_traits.pdf"), width=7, height=5)
cellnotes <- ifelse(aov_traits_PCs<0.001, "***",
                    ifelse(aov_traits_PCs<0.01, "**",
                           ifelse(aov_traits_PCs<0.05, "*", "")))
myCol <- brewer.pal(n=9, name = "Reds")
heatmap.2(as.matrix(-log10(aov_traits_PCs)),
          cellnote = cellnotes,
          notecol="gray1",
          notecex=1.5,
          margin=c(5,15), 
          density.info="none", trace="none",
          Colv = FALSE, Rowv = TRUE, dendrogram="row",
          key=T, key.xlab="-log10 pvalue (F-test)",
          col=myCol,
          sepcolor="white", colsep=c(1:ncol(aov_traits_PCs)),
          rowsep=c(1:nrow(aov_traits_PCs)), sepwidth=c(0.01,0.01))
dev.off()    
```



# 4. PCA ( scale=F)

```{r}
out_folder <- "/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/RNAseq/Rout/plots/QC/PCA_unscaled/"
if (!dir.exists(out_folder)) {
  dir.create(path = out_folder)
}

```


## Read in vst-counts and metadata
```{r}
vst_counts <- read.delim(paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/vstcounts.txt"), check.names = F)
metadata <- read.delim(paste0(mnt_path,"RNAseq/Rout/tables/RNAseq/metadata.txt"), stringsAsFactors = T)
stopifnot(rownames(metadata)==colnames(vst_counts))
```




# clean metadata and run PCA
```{r}
metadata_clean <- data.frame(metadata[,c("sample_id", "short_id", "immucan_id",
                                         "lib.kit", 
                                         "gender", "simple_histology", "simple_stage",
                                         "metastatic_treatment","neo_adjuvant_treatment", "adjuvant_treatment",
                                         "age_at_diagnosis",
                                         "side_localization", "primary_localization", "smoking_status", "environmental_exposure")])
metadata_clean$date_of_extraction_num <- as.numeric(metadata$date_of_extraction)


# run PCA
uq_values <- apply(vst_counts,1,function(x) length(unique(x)))==1
p <- pca(vst_counts[!uq_values,], metadata=metadata_clean, scale = FALSE, transposed = FALSE)
pca_df <- p$rotated
```



## screeplot
```{r}
elbow <- findElbowPoint(p$variance)
pdf(paste0(out_folder,"pca_screeplot.pdf"))
screeplot(p, axisLabSize = 18, titleLabSize = 22, components=getComponents(p)[1:15], vline = elbow)+
  geom_label(aes(x = elbow - 1, y = 50,
                 label = 'Elbow point', vjust = -1, size = 8))
dev.off()
```



## pairsplot
```{r, message=FALSE}
pdf(paste0(out_folder,"pca_pairsplot.pdf"), width=6, height=6)
lapply(clin_var, function(trait) {
  pp <- pairsplot(p,
                  components = getComponents(p, c(1:10)),
                  triangle = TRUE, trianglelabSize = 12,
                  hline = 0, vline = 0,
                  pointSize = 0.4,
                  gridlines.major = FALSE, gridlines.minor = FALSE,
                  colby = trait,
                  title = trait,
                  plotaxes = FALSE,
                  margingaps = unit(c(-0.01, -0.01, -0.01, -0.01), 'cm'))
  pp
})
dev.off()
pdf(paste0(out_folder,"pca_pairsplot_PC1toPC4.pdf"), width=8, height=8)
lapply(colnames(metadata_clean)[-(1:3)], function(trait) {
  pp <- pairsplot(p,
                  components = getComponents(p, c(1:4)),
                  triangle = TRUE, trianglelabSize = 12,
                  hline = 0, vline = 0,
                  pointSize = 1,
                  gridlines.major = FALSE, gridlines.minor = FALSE,
                  colby = trait,
                  title = trait,
                  plotaxes = FALSE,legendPosition = "none",
                  margingaps = unit(c(-0.01, -0.01, -0.01, -0.01), 'cm'))
  pp
})
dev.off()
```




## biplots 

### PC1 vs PC2
```{r}
pdf(paste0(out_folder,"pca_PC1vsPC2_labeled.pdf"), width=6, height=4)
lapply(colnames(metadata_clean)[-(1:3)], function(trait) {
  bp <- biplot(p,
               x="PC1",y="PC2",
               lab = p$metadata$short_id,
               colby = trait,
               hline = 0, vline = 0,
               pointSize=2,
               legendPosition = 'right',
               legendTitleSize = 8,
               legendLabSize = 6,
               legendIconSize = 2) + ggtitle(trait)
  bp
})
dev.off()
```



### PC3 vs PC4
```{r}
pdf(paste0(out_folder,"pca_PC3vsPC4_labeled.pdf"), width=6, height=4)
lapply(colnames(metadata_clean)[-(1:3)], function(trait) {
  bp <- biplot(p,
               x="PC3",y="PC4",
               lab = p$metadata$short_id,
               colby = trait,
               hline = 0, vline = 0,
               pointSize=2,
               legendPosition = 'right',
               legendTitleSize = 8,
               legendLabSize = 6,
               legendIconSize = 2) + ggtitle(trait)
  bp
})
dev.off()
```





## biplots unlabeled

### PC1 vs PC2
```{r}

pdf(paste0(out_folder,"pca_PC1vsPC2.pdf"), width=6, height=4)
lapply(colnames(metadata_clean)[-(1:3)], function(trait) {
  bp <- biplot(p,
               x="PC1",y="PC2",
               lab = NULL,
               colby = trait,
               hline = 0, vline = 0,
               pointSize=2,
               legendPosition = 'right',
               legendTitleSize = 8,
               legendLabSize = 6,
               legendIconSize = 2) + ggtitle(trait)
  bp
})
dev.off()
```


### PC3 vs PC4
```{r}
pdf(paste0(out_folder,"pca_PC3vsPC4.pdf"), width=6, height=4)
lapply(colnames(metadata_clean)[-(1:3)], function(trait) {
  bp <- biplot(p,
               x="PC3",y="PC4",
               lab = NULL,
               colby = trait,
               hline = 0, vline = 0,
               pointSize=2,
               legendPosition = 'right',
               legendTitleSize = 8,
               legendLabSize = 6,
               legendIconSize = 2) +
    ggtitle(trait)
  bp
})
dev.off()
```





## pca hclust
```{r}
# Retrieve PCA scores
scores <- p$rotated
rownames(scores) <- p$metadata$short_id
# Prepare colors for side bar annotation
ggplot_colors2 <- hue_pal()(2)
ggplot_colors <- hue_pal()(10)
rb_colors <- rainbow(4)
col.anno <- HeatmapAnnotation(stage = p$metadata$simple_stage,
                              adjuvant_treatment=p$metadata$adjuvant_treatment,
                              histology = p$metadata$simple_histology,
                              gender = p$metadata$gender,
                              smoking_status = p$metadata$smoking_status,
                              lib.kit = p$metadata$lib.kit,
                              col = list(stage=c("I"="yellow",
                                                 "II"="orange",
                                                 "III"="red",
                                                 "IV"="brown"),
                                         adjuvant_treatment=c("No"=ggplot_colors2[1], "Yes"=ggplot_colors2[2]),
                                         histology=c("Adenocarcinoma"=ggplot_colors[2],
                                                     "Squamous cell carcinoma"=ggplot_colors[8],
                                                     "Adenosquamous carcinoma"=ggplot_colors[3],
                                                     "LCNEC" = ggplot_colors[5],
                                                     "Other"=ggplot_colors[4],
                                                     "unknown" = "gray"),
                                         gender = c("Female" = "pink", "Male"="cyan"),
                                         smoking_status = c("Non-smoker"="yellow", "Ex-smoker"="orange", "Current smoker"="red"),
                                         lib.kit = c("kapa"=ggplot_colors2[1], "takara"=ggplot_colors2[2])))
```


## pca hclust using top 10 PCs
```{r}
pdf(paste0(out_folder,"pca_hclust_top10PCs_avgClust.pdf"), width=15, height=7)
sel_scores <- as.matrix(scores[,1:10])
Heatmap(t((sel_scores)),
        name = "PCA score",
        row_title = "PCs", row_title_side = "left",
        column_title = "Hierarchical clustering of samples based on top 10 PCs",
        column_title_gp = gpar(fontsize = 11),
        top_annotation = col.anno,
        cluster_rows = FALSE,
        clustering_method_columns = "average",
        show_row_names = T,
        row_names_gp = gpar(fontsize = 8),
        column_names_gp = gpar(fontsize = 5))
dev.off()

pdf(paste0(out_folder,"pca_hclust_top8PCs_complClust.pdf"), width=15, height=7)
sel_scores <- as.matrix(scores[,1:10])
Heatmap(t((sel_scores)),
        name = "PCA score",
        row_title = "PCs", row_title_side = "left",
        column_title = "Hierarchical clustering of samples based on top 10 PCs",
        column_title_gp = gpar(fontsize = 11),
        top_annotation = col.anno,
        cluster_rows = FALSE,
        clustering_method_columns = "complete",
        show_row_names = T,
        row_names_gp = gpar(fontsize = 8),
        column_names_gp = gpar(fontsize = 5))
dev.off()

```


## pca hclust using PCs that explains 90% variance (sum(p$variance[1:74]))
```{r}
pdf(paste0(out_folder,"pca_hclust_PCs90percVar_avglClust.pdf"), width=15, height=9)
sel_scores <- as.matrix(scores[,1:74])
Heatmap(t((sel_scores)),
        name = "PCA score",
        row_title = "PCs", row_title_side = "left",
        column_title = "Hierarchical clustering of samples based on PCs that explain 90% variance",
        column_title_gp = gpar(fontsize = 11),
        top_annotation = col.anno,
        cluster_rows = FALSE,
        clustering_method_columns = "average",
        show_row_names = T,
        row_names_gp = gpar(fontsize = 7),
        column_names_gp = gpar(fontsize = 5))
dev.off()

pdf(paste0(out_folder,"pca_hclust_PCs90percVar_complClust.pdf"), width=15, height=9)
sel_scores <- as.matrix(scores[,1:74])
Heatmap(t((sel_scores)),
        name = "PCA score",
        row_title = "PCs", row_title_side = "left",
        column_title = "Hierarchical clustering of samples based on PCs that explain 90% variance",
        column_title_gp = gpar(fontsize = 11),
        top_annotation = col.anno,
        cluster_rows = FALSE,
        clustering_method_columns = "complete",
        show_row_names = T,
        row_names_gp = gpar(fontsize = 7),
        column_names_gp = gpar(fontsize = 5))
dev.off()
```


## Run F.test
table of stat (ANOVA/t.test)
```{r}
sel_traits <- c("simple_stage", "side_localization", "primary_localization",
                "age_at_diagnosis", "gender", "smoking_status",
                "neo_adjuvant_treatment",  "adjuvant_treatment", "metastatic_treatment", "simple_histology",
                "lib.kit")
allTraits_allPCs <- lapply(sel_traits, function(trait) {
  allPCs <- lapply(colnames(p$rotated)[1:6], function(pc) {
    res.aov <- summary(aov(p$rotated[,pc] ~ p$metadata[,trait]))[[1]][["Pr(>F)"]][1]
    return(res.aov)
  })
  names(allPCs) <- colnames(p$rotated)[1:6]
  return(allPCs)
})
names(allTraits_allPCs) <- sel_traits
aov_traits_PCs <- do.call(rbind.data.frame, allTraits_allPCs)

pdf(paste0(out_folder,"Ftest_PC_traits.pdf"), width=7, height=5)
cellnotes <- ifelse(aov_traits_PCs<0.001, "***",
                    ifelse(aov_traits_PCs<0.01, "**",
                           ifelse(aov_traits_PCs<0.05, "*", "")))
myCol <- brewer.pal(n=9, name = "Reds")
heatmap.2(as.matrix(-log10(aov_traits_PCs)),
          cellnote = cellnotes,
          notecol="gray1",
          notecex=1.5,
          margin=c(5,15), 
          density.info="none", trace="none",
          Colv = FALSE, Rowv = TRUE, dendrogram="row",
          key=T, key.xlab="-log10 pvalue (F-test)",
          col=myCol,
          sepcolor="white", colsep=c(1:ncol(aov_traits_PCs)),
          rowsep=c(1:nrow(aov_traits_PCs)), sepwidth=c(0.01,0.01))
dev.off()    
```

