---
title: "Reading in the data, from single patient folders"
site: workflowr::wflow_site
output:
  workflowr::wflow_html:
    toc: true
editor_options:
  chunk_output_type: inline
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

# Set path
if (Sys.info()["sysname"] == "Darwin"){
  mount_path <- "/Volumes/"
} else if (Sys.info()["sysname"] == "Linux"){
  mount_path <- "/mnt/"
} else {
  stop("Operating system is not supported")
}
```


# paths
```{r}
mount_path <- paste0(mount_path,"immucan_volume/processed_data/NSCLC2_cohort/data/IMC/Panel_1/")
```

# Load libraries

First, we will load the libraries needed

```{r load-libraries, message=FALSE}
library(S4Vectors)
library(SingleCellExperiment)
library(tidyverse)
library(imcRtools)
library(dplyr)
library(stringr)
```

# Read in the single-cell data and concatenate into one SCE

```{r read-data}

cur_paths <- list.files(paste0(mount_path,"data"), full.names = TRUE)


cur_sces <- lapply(cur_paths, function(x){
  
  # read in data
  counts <- read.csv(unz(x, filename = "raw_counts.csv"), row.names = 1)
  cell_meta <- read.csv(unz(x, filename = "cell_metadata.csv"), row.names = 1)
  row_data <- read.csv(unz(x, filename = "panel_metadata.csv"), row.names = 1)
  
  # create SCE
  sce <- SingleCellExperiment(assays = list(counts = t(counts)))
  colData(sce) <- as(cell_meta, "DataFrame")
  rowData(sce) <- as(row_data, "DataFrame")
  
  # Read in graph
  cur_graph <- read.csv(unz(x, filename = "cell_relationships.csv"), row.names = 1)
  cur_graph <- SelfHits(from = cur_graph$from, to = cur_graph$to, nnode = ncol(sce))
  colPair(sce, "neighborhood") <- cur_graph
  
  return(sce)

})

```

```{r unify-colnames}
# clean up colnames
colData_names <- Reduce(intersect, lapply(cur_sces, function(x) colnames(colData(x))))
excluded_colData_names <- unique(unlist(lapply(cur_sces, function(x) colnames(colData(x))[!colnames(colData(x)) %in% colData_names])))
excluded_colData_names # "Sample.Type"   "Pooled"        "Specimen.Type" "Is.Pooled"    

# clean up rownames
rowData_names <- Reduce(intersect, lapply(cur_sces, function(x) colnames(rowData(x))))
excluded_rowData_names <- unique(unlist(lapply(cur_sces, function(x) colnames(colData(x))[!colnames(rowData(x)) %in% rowData_names])))
excluded_rowData_names # none
```


```{r concatenate-sces}
# clean up assays
assay_names <- c("counts")

cur_sces <- lapply(cur_sces, function(x) {
  cur_sce <- x

  # unify colnames in excluded_colData_names
  if (is.null(cur_sce$Pooled)){
    cur_sce$Pooled <- cur_sce$Is.Pooled
    cur_sce$Is.Pooled <- NULL
  }
  
  if (is.null(cur_sce$Sample.Type)){
    cur_sce$Sample.Type <- cur_sce$Specimen.Type
    cur_sce$Specimen.Type <- NULL
  }
  
  colData(cur_sce) <- colData(cur_sce)[,c(colData_names, "Pooled", "Sample.Type")]
  rowData(cur_sce) <- rowData(cur_sce)[,rowData_names, drop=FALSE]
  
  
  reducedDims(cur_sce) <- NULL
  
  assays(cur_sce) <- assays(cur_sce)[assay_names]
  
  return(cur_sce)
})

sce <- do.call(cbind, cur_sces)

sce$image <- paste0(sce$sample_id, ".tiff")
assay(sce, "exprs") <- asinh(counts(sce))

```



Check for unexpected duplicates
```{r, check-duplicates}
cur_duplicates <- duplicated(paste(sce$SlideId, sce$Position, sce$ROI, sce$ObjectNumber))
unique(paste(sce$SlideId, sce$Position, sce$ROI)[cur_duplicates])

stopifnot(sum(cur_duplicates) == 0)
```

```{r}
# we will get the short IBBL IDS to get the mapping to immucan ids 
IBBL_ids <- unique(sce$SampleId)
IBBL_ids <- sapply(IBBL_ids, function(x) str_split_i(string = x,pattern = "-",1))
write_csv(as.data.frame(IBBL_ids),file = "data/IMMUcanIDMapping.csv")

# get info of imumucan ids that we actually have data of
ids <- sapply(list.files(paste0(mount_path,"data")), function(x) str_split_i(string = x,pattern = "_#_",1))
ids <- sapply(ids, function(x) str_split_i(pattern = "-FIXT",string = x,i = 1))
# from the KMS we will retrieve the patient IDs for each IBBL ID

# we will get the precise tissue ids that we have data from
tissue_ids <- unique(sce$SampleId)
tissue_ids <- sapply(tissue_ids, function(x) gsub(pattern = "-PB","",x = x))

# load mapping data retrieved from the KMS with the IBBL_ids
mapping <- read.csv("data/IMMUcanIDMapping_returned.csv") 

# clean and filter mapping according to tissue blocks we used
mapping <- mapping %>%
  separate_longer_delim(cols = ibbl_kit_ids,delim=",") %>%
  separate_longer_delim(cols = ibbl_ids,delim=",") %>%
  mutate(ibbl_ids_check = str_split_i(ibbl_ids,"-",1)) %>%
  filter(ibbl_ids_check == ibbl_kit_ids) %>%
  filter(ibbl_ids %in% tissue_ids) %>%
  distinct()

mapping %>%
  filter(duplicated(ibbl_kit_ids))

# create ibbl_kit_ids for matching
sce$ibbl_kit_ids <- str_split(sce$SampleId,"-", simplify = TRUE)[,1]

unique(sce$ibbl_kit_ids) %in% mapping$ibbl_kit_ids

# add clinical information to colData
coldata <- as.data.frame(colData(sce)) %>%
  left_join(mapping, by = "ibbl_kit_ids")

dim(coldata)
ncol(sce)

# check if all immucan IDs are available
sum(is.na(coldata$immucan_id))

colData(sce) <- as(coldata, "DataFrame")
```



```{r set-colnames}
colnames(sce) <- paste(sce$SampleId, sce$ROI, sce$ObjectNumber, sep = "_")
```


## Generate colour vectors

At this point we want to generate colour vectors that we can reproducibly and consistently use throughout the analysis.
Primarily, we will define colours for 1. the samples, 2. the slides, 3. the ROI, 4. the Indication, 5. the tumour mask.

```{r generate-colour-vectors}
library(dittoSeq)
library(RColorBrewer)
colour_vectors <- list()

col_vec_sampleIds <- dittoColors(reps = 3)[seq_len(length(unique(sce$SampleId)))]
names(col_vec_sampleIds) <- unique(sce$SampleId)

col_vec_slideIds <- dittoColors(reps = 3)[seq_len(length(unique(sce$SlideId)))]
names(col_vec_slideIds) <- unique(sce$Slide.ID)

col_vec_ROIs <- colorRampPalette(c("dark green", "white", "dark red"))(length(unique(sce$ROI)))
names(col_vec_ROIs) <- unique(sce$ROI)

col_vec_subBatchIds <- brewer.pal(name = "Set3", n = length(unique(sce$SubBatchId)))
names(col_vec_subBatchIds) <- unique(sce$SubBatchId)

# Here we can use these five colours for the five indications:
# Breast = "#3F85A7"
# Kidney = "#5B1C55"
# Gastrointestinal = "#C81F43"
# Lung = "#F79C1D"
# Head and Neck = "#39BEB4"

col_vec_indication <- c(GU = "#5B1C55", HN = "#39BEB4", THOR = "#F79C1D", 
                        BREAS = "#3F85A7", GI = "#C81F43")

col_vec_tumour <- c(Tumour = "#8c510a", Non_Tumour = "#01665e", 
                    Ambigous = "black", Error = "dark red")

colour_vectors$SampleIds <- col_vec_sampleIds
colour_vectors$SlideIds <- col_vec_slideIds
colour_vectors$ROIs <- col_vec_ROIs
colour_vectors$Indication <- col_vec_indication
colour_vectors$Tumour <- col_vec_tumour
colour_vectors$SubBatchIds <- col_vec_subBatchIds

# Save in metadata slot
metadata(sce)$colour_vectors <- colour_vectors
```

# Filter samples

sample "10022145" is a distant metastasis and will be excluded

sample 10040831 is a duplicated sample from patient LUNG-NSCLC2-0598. we will store this under the same immucan patient id
```{r}
sce <- sce[,sce$ibbl_kit_ids != "10022145"]
```


## Save SCE object

Finally, for easy accessability, we will write the SCE object out.
The object will be stored on the server so that everyone can work with it.

```{r save-RDS}
# create necessary directory first
if (!dir.exists(paste0(mount_path,"Rout/"))) {
  dir.create(paste0(mount_path,"Rout/"))
}


saveRDS(sce, paste0(mount_path,"Rout/sce.rds"))
```

# Read in image data

Here, we read in the images and masks and save them on disk.

```{r read-images, message = FALSE}
library(cytomapper)
# create necessary directory first
if (!dir.exists(paste0(mount_path, "Rout/h5/"))) {
  dir.create(paste0(mount_path, "Rout/h5/"))
}
if (!dir.exists(paste0(mount_path, "Rout/h5/images"))) {
  dir.create(paste0(mount_path, "Rout/h5/images"))  
}
if (!dir.exists(paste0(mount_path, "Rout/h5/masks"))) {
  dir.create(paste0(mount_path, "Rout/h5/masks"))  
}


# load masks to disk
masks_path <- list.files(paste0(mount_path,"data"), full.names = TRUE)

# Loop through zip files
for (zip_file in masks_path) {
  zip_name <- tools::file_path_sans_ext(basename(zip_file))

  # Create temp dir for unzipping
  temp_dir <- tempfile(pattern = "unzip_")
  dir.create(temp_dir)

  # List contents of zip
  contents <- unzip(zip_file, list = TRUE)
  mask_files <- contents$Name[grepl("^masks/|/masks/", contents$Name)]

  # Unzip only masks
  unzip(zip_file, files = mask_files, exdir = temp_dir)

  # Find the masks subfolder
  cur_mask_path <- list.dirs(temp_dir, recursive = TRUE, full.names = TRUE)
  cur_mask_path <- cur_mask_path[grepl("masks$", cur_mask_path)]

  # Load images directly (onDisk = TRUE avoids loading all into RAM)
  mask_imgs <- cytomapper::loadImages(x = cur_mask_path, on_disk = TRUE,h5FilesPath = paste0(mount_path,"Rout/h5/masks"))

  # Optional: clean up temp_dir if needed
  unlink(temp_dir, recursive = TRUE)
}


# load images to disk
img_path <- list.files(paste0(mount_path,"data"), full.names = TRUE)

# Loop through zip files
for (zip_file in img_path) {
  zip_name <- tools::file_path_sans_ext(basename(zip_file))

  # Create temp dir for unzipping
  temp_dir <- tempfile(pattern = "unzip_")
  dir.create(temp_dir)

  # List contents of zip
  contents <- unzip(zip_file, list = TRUE)
  img_files <- contents$Name[grepl("^img/|/img/", contents$Name)]

  # Unzip only images
  unzip(zip_file, files = img_files, exdir = temp_dir)

  # Find the masks subfolder
  cur_img_path <- list.dirs(temp_dir, recursive = TRUE, full.names = TRUE)
  cur_img_path <- cur_img_path[grepl("img$", cur_img_path)]

  # Load images directly (onDisk = TRUE avoids loading all into RAM)
  img_images <- cytomapper::loadImages(x = cur_img_path, on_disk = TRUE,h5FilesPath = paste0(mount_path,"Rout/h5/images"))

  # Optional: clean up temp_dir if needed
  unlink(temp_dir, recursive = TRUE)
}


  
# get path to images and masks per patient
imgs_path <- paste0(mount_path, "Rout/h5/images")
masks_path <- paste0(mount_path, "Rout/h5/masks/")



images <- loadImages(imgs_path, on_disk = TRUE, h5FilesPath = paste0(mount_path, "Rout/h5/images"))
masks <- loadImages(masks_path, on_disk = TRUE, h5FilesPath = paste0(mount_path, "Rout/h5/masks"))



# Remove duplicated images
sum(!(names(images) %in% gsub("\\.tiff", "", sce$image)))
sum(!(names(masks) %in% gsub("\\.tiff", "", sce$image)))

images <- images[names(images) %in% gsub("\\.tiff", "", sce$image)]
masks <- masks[names(masks) %in% gsub("\\.tiff", "", sce$image)]

mcols(images) <- mcols(masks) <- DataFrame(image = names(images))

saveRDS(images, paste0(mount_path, "Rout/images.rds"))
saveRDS(masks, paste0(mount_path, "Rout/masks.rds"))
```
