---
title: "Untitled"
author: "Daniel Schulz"
date: "2026-01-23"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# paths
```{r}
mount_path <- paste0("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/IMC/Panel_1/")
```

```{r}
library(SingleCellExperiment)
library(imcRtools)
library(cytomapper)
library(tidyverse)
library(dittoSeq)
library(viridis)
library(doParallel)
library(BiocParallel)
library(BiocNeighbors)
library(scran)
library(bluster)
library(scater)
library(cowplot)
```

```{r}
sce <- readRDS(paste0(mount_path,"Rout/sce.rds"))
sce$cell_id <- colnames(sce)
```

```{r determine markers for clustering}
markers <- list()
# we will not subcluters "uncertain" cells as the merging will be very difficult. we may inspect this at a later time point again

#functional markers for T cells
markers[["lymphocytes"]] <- rownames(sce)[grepl(pattern = "CD3|CD20$|CD27|CD45RA|CD45RO|LAG3|PD1|CD7|GrzB|FOXP3|ICOS|Ki67|TCF7|VISTA|CD40|HLADR|CD38",rownames(sce))]

markers[["Tcells"]] <- rownames(sce)[grepl(pattern = "CD27|LAG3|PD1|CD7|GrzB|ICOS|Ki67|TCF7|CD38",rownames(sce))]


good_markers <- rownames(sce)[!grepl("DNA|Histone",rownames(sce))]
```

# update positivity
we will select thresholds below
```{r}
sce$PD1_pos <- assay(sce, "exprs")["PD1",] > 1
sce$GrzB_pos <- assay(sce, "exprs")["GrzB",] > 2
sce$Ki67_pos <- assay(sce, "exprs")["Ki67",] > 1
sce$TCF7_pos <- assay(sce, "exprs")["TCF7",] >1.5
sce$PDL1_pos <- assay(sce, "exprs")["PDL1",] >1.3
```

# merge BnT and CD8/CD4
all BnT cells that have CD8 expression will be merged with CD8 T cells for subclsutering
all BnT cells with no CD8 expression and CD4 expression will be merged with CD8 T cells for subclsutering
```{r}
cur_sce <- sce[,sce$celltypes == "BnT"]
BnT_CD8_ids <- cur_sce[,which(assay(cur_sce,"exprs")["CD8a",]>2)]$cell_id
BnT_CD4_ids <- cur_sce[,which(assay(cur_sce,"exprs")["CD8a",]<2 & assay(cur_sce,"exprs")["CD4",] > 1)]$cell_id
```

# UMAP
```{r, fig.height=15, fig.width=20}
sce <- runUMAP(sce,
               exprs_values = "exprs", 
               subset_row = rownames(sce)[!grepl("DNA|Histone",rownames(sce))],
               name = "UMAP",
               external_neighbors = TRUE, BNPARAM = AnnoyParam(),
               BPPARAM = MulticoreParam(workers = 12))

dittoDimPlot(sce,var = "celltypes",reduction.use = "UMAP",do.label = TRUE)+
  scale_color_manual(values = metadata(sce)$colour_vectors$celltypes)
dittoDimPlot(sce,var = "celltypes_minor",reduction.use = "UMAP",do.label = TRUE)+
  scale_color_manual(values = metadata(sce)$colour_vectors$celltypes_minor)
```

# cluster sweeps

```{r, include=FALSE, eval=FALSE}

# cluster sweeps with Rphenoannoy
library(Rphenoannoy)
for (x in c("CD4","CD8")) {
    cur_sce <- sce[,sce$celltypes %in% x]
#this code potentially add all BnT_CD8 cells. however we have many markers then that are not relevant for clustering CD8 T cells and will therefore not do it
  # if(x == "CD4"){
  #   cur_sce <- sce[,c(ids,BnT_CD4_ids)]
  # } else if (x == "CD8"){
  #   cur_sce <- sce[,c(ids,BnT_CD8_ids)]
  # }
 
mat <- t(assay(cur_sce, "exprs")[markers$Tcells,])
set.seed(12345)
out_30 <- Rphenoannoy(threshold = ,mat, k = 30)
set.seed(12345)
out_50 <- Rphenoannoy(threshold = ,mat, k = 50)
set.seed(12345)
out_70 <- Rphenoannoy(threshold = ,mat, k = 70)
set.seed(12345)
out_90 <- Rphenoannoy(threshold = ,mat, k = 90)
set.seed(12345)
out_110 <- Rphenoannoy(threshold = ,mat, k = 110)
set.seed(12345)
out_130 <- Rphenoannoy(threshold = ,mat, k = 130)

cur_sce$out_30 <- factor(membership(out_30[[2]]))
cur_sce$out_50 <- factor(membership(out_50[[2]]))
cur_sce$out_70 <- factor(membership(out_70[[2]]))
cur_sce$out_90 <- factor(membership(out_90[[2]]))
cur_sce$out_110 <- factor(membership(out_110[[2]]))
cur_sce$out_130 <- factor(membership(out_130[[2]]))

saveRDS(cur_sce, paste0(mount_path,"Rout/subclustering/",x,".rds"))
}

```

# cluster evaluation
here we will use 
- silhouette width -> larger values indicate cells being cluster to cells from the same cluster
- RMSD -> dispersion of cells within each cluster -> large values indicate candidates for subclustering. RMSD corresponds to the root mean of the squared distances (RMSD = sqrt(WCSS/n)) where n is the number of points. WCSS is not scaled to the size of the clusters, RMSD basically returns the average distance to centroid for a cluster 
- number of clusters

currently not using systematically
- adjusted rand index -> higher values indicate overlap between clustering


```{r, fig.height=8, fig.width=6}
# create necessary directory first
if (!dir.exists(paste0(mount_path, "Rout/subclustering/plots"))) {
  dir.create(paste0(mount_path, "Rout/subclustering/plots"))
}


files <- list.files(paste0(mount_path,"Rout/subclustering"), full.names = TRUE, pattern = "*.rds")

for (x in files){

  cur_sce <- readRDS(x)
  cts <- gsub(".rds","",basename(x))
  
  combinations <- list(out_30 = cur_sce$out_30,
                       out_50 = cur_sce$out_50,
                       out_70 = cur_sce$out_70,
                       out_90 = cur_sce$out_90,
                       out_110 = cur_sce$out_110,
                       out_130 = cur_sce$out_130)

  mat <- t(assay(cur_sce, "exprs")[markers[["Tcells"]],])  
  
  # silhouette width
  sil <- vapply(combinations, 
                function(x) mean(approxSilhouette(mat, x)$width), 
                0)
  
  names(sil) <- gsub("out_","",names(sil))
  
  # RMSD  
  rmsd <- lapply(combinations, 
                 function(x) clusterRMSD(mat,x))
  rmsd <- lapply(rmsd,sum)
  unlist(rmsd)
  
  # number of clusters per k
  cluster_numbers <- lapply(combinations, 
                            function(x) length(unique(x)))
  
  # generate output table
  cur_dat <- data.frame(silhouette = sil,
                        k = names(sil),
                        rmsd = unlist(rmsd),
                        n = unlist(cluster_numbers))
  rownames(cur_dat) <- NULL
  
  # plots
  sil_plot <- ggplot(cur_dat,aes(x = factor(k, levels = c(30, 50, 70, 90, 110, 130)),y = silhouette)) +
    geom_point() +geom_line()+
    theme_classic(base_size = 15) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab("k") +
    ylab("Average silhouette width")+
    ggtitle(cts)

  rmsd_plot <- cur_dat %>%
    ggplot(aes(x = factor(k, levels = c("30", "50", "70","90", "110", "130")),y = rmsd))+
    geom_point()+geom_line()+
    theme_classic(base_size = 15) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab("k") +
    ylab("RMSD")+
    ggtitle(cts)
  
  cluster_n_plot <- cur_dat %>%
    ggplot(aes(x = factor(k, levels = c(30, 50, 70, 90, 110, 130)),y = n))+
    geom_point()+ geom_line()+
    theme_classic(base_size = 15) +
    theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
    xlab("k") +
    ylab("cluster number")+
    ggtitle(cts)
  
  p <- cowplot::plot_grid(sil_plot,rmsd_plot, cluster_n_plot, ncol = 1)
  p
  
   ggsave(filename = paste0(mount_path,"Rout/subclustering/plots/",gsub(".rds","",basename(x)),"_cluster_evaluation",".png"),p)
  
}
```

# meta clustering

we will select k=110 for CD4 and CD8 T cells

```{r}
total_cells <- as.data.frame((table(sce$immucan_id)))
colnames(total_cells) <- c("immucan_id", "total_cell_number")
library(factoextra)
library(NbClust)
sces <- list.files(paste0(mount_path,"Rout/subclustering"), full.names = TRUE, pattern = "*.rds")

p.list <- list()
h_list <- list()
abund_list <- list()
for (i in sces) {
  cur_sce <- readRDS(i)
  cur_name <- gsub(".rds","",basename(i))
  # total cells per cluster and fraction of patients with at least 1 % of cells belonging to that cluster
  abund <- as.data.frame(unclass(table(cur_sce$out_110, cur_sce$immucan_id))) 
  abund <- abund %>%
    mutate(cluster = paste0(cur_name,"_",rownames(abund))) %>%
    pivot_longer(-cluster, names_to =  "immucan_id") %>%
    left_join(total_cells,"immucan_id") %>%
    group_by(cluster) %>%
    mutate(cluster_sum = sum(value)) %>%
    group_by(immucan_id) %>%
    mutate(frac = round(value/total_cell_number,digits = 3)) %>%
    group_by(cluster) %>%
    mutate(cut_off = sum(frac > 0.01)) %>%
    ungroup() 
  
  abund_list[[paste0(cur_name)]] <- abund %>%
    ggplot(aes(x = cluster,y = frac)) +
    geom_boxplot()+
    theme_bw()+
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))+
    geom_hline(yintercept = 0.01, col = "red")+
    geom_text(aes(label = cut_off,y = max(frac)), col = "red")
  

  mean_sce <- aggregateAcrossCells(cur_sce,ids=cur_sce$cluster,statistics = "mean")
  assay(mean_sce,"exprs") <- asinh(counts(mean_sce))
  meta_mat <- t(assay(mean_sce[markers[[cur_name]],],"exprs"))
  #meta_mat <- scale(meta_mat)

  # meta clustering
  set.seed(12356)
  hc_sil <- fviz_nbclust(scale(meta_mat), hcut, diss = dist(meta_mat,method = "euclidian"),hc_method = "ward.D2", method = "silhouette",k.max = nrow(meta_mat)-1, print.summary = TRUE)$data
  set.seed(12356)
  km_sil <- fviz_nbclust(scale(meta_mat), kmeans, method = "silhouette",k.max = nrow(meta_mat)-1, print.summary = TRUE)$data
    set.seed(12356)
  hc_wss <- fviz_nbclust(scale(meta_mat), hcut, diss = dist(meta_mat,method = "euclidian"),hc_method = "ward.D2", method = "wss",k.max = nrow(meta_mat)-1, print.summary = TRUE)$data
  set.seed(12356)
  km_wss <- fviz_nbclust(scale(meta_mat), kmeans, method = "wss",k.max = nrow(meta_mat)-1, print.summary = TRUE)$data
  
  p.list[[paste0(cur_name)]] <- data.frame(cluster = hc_sil$clusters,
                                           hc_sil = hc_sil$y,
                                           km_sil = km_sil$y,
                                           hc_wss = hc_wss$y,
                                           km_wss = km_wss$y)
  
  # plot heatmap
  set.seed(12356)
  h_list[[paste0(cur_name)]] <- dittoSeq::dittoHeatmap(mean_sce,genes =markers[[cur_name]], assay = "exprs",
                                                       annot.by = c("cluster"),
                                                       cluster_cols = TRUE,
                                                       cluster_rows = TRUE, 
                                                       clustering_distance_rows = "euclidean",
                                                       clustering_distance_cols = "euclidean",
                                                       clustering_method = "ward.D2",
                                                       show_colnames = TRUE,
                                                       heatmap.colors = colorRampPalette(c("dark blue", "white", "dark red"))(100),
                                                       breaks = seq(-3, 3, length.out = 101))
  
}

```



























# cluster evalution CD8

for CD8 we will use 90 nearest neighbors
```{r, fig.height=10, fig.width=15}
CD8_sce <- readRDS("/mnt/immucan_volume/processed_data/NSCLC2_cohort/data/IMC/Panel_1/Rout/subclustering/CD8.rds")
set.seed(1235)
CD8_sce <- runUMAP(CD8_sce,
                   exprs_values = "exprs", 
                   subset_row = markers$lymphocytes,
                   name = "UMAP",
                   external_neighbors = TRUE, BNPARAM = AnnoyParam(),
                   BPPARAM = MulticoreParam(workers = 12))

p1 <- dittoDimPlot(CD8_sce,var = "out_90",reduction.use = "UMAP",do.label = TRUE)
p2 <- dittoDimPlot(CD8_sce,var = "celltypes_minor",reduction.use = "UMAP",do.label = TRUE)+
  scale_color_manual(values = metadata(CD8_sce)$colour_vectors$celltypes_minor)
cowplot::plot_grid(p1,p2, ncol=2)


# heatmap
mean_sce <- aggregateAcrossCells(CD8_sce,ids=CD8_sce$out_90,statistics = "mean")
assay(mean_sce,"exprs") <- asinh(counts(mean_sce))
meta_mat <- t(assay(mean_sce[markers$lymphocytes,],"exprs"))
#meta_mat <- scale(meta_mat)

dittoSeq::dittoHeatmap(mean_sce,
                       genes = markers$,
                       assay = "exprs",
                       annot.by = c("out_90"),
                       cluster_cols = TRUE,
                       cluster_rows = TRUE,
                       show_colnames = TRUE,
                       heatmap.colors = colorRampPalette(c("dark blue", "white", "dark red"))(100),
                       breaks = seq(-3, 3, length.out = 101))



  set.seed(12356)
  hc_sil <- fviz_nbclust(scale(meta_mat), hcut, method = "silhouette",k.max = nrow(meta_mat)-1, print.summary = TRUE)$data
  set.seed(12356)
  km_sil <- fviz_nbclust(scale(meta_mat), kmeans, method = "silhouette",k.max = nrow(meta_mat)-1, print.summary = TRUE)$data
  set.seed(12356)
  hc_wss <- fviz_nbclust(scale(meta_mat), hcut, method = "wss",k.max = nrow(meta_mat)-1, print.summary = TRUE)$data
  set.seed(12356)
  km_wss <- fviz_nbclust(scale(meta_mat), kmeans, method = "wss",k.max = nrow(meta_mat)-1, print.summary = TRUE)$data
  
  p.list[[paste0(cur_name)]] <- data.frame(cluster = hc_sil$clusters,
                                           hc_sil = hc_sil$y,
                                           km_sil = km_sil$y,
                                           hc_wss = hc_wss$y,
                                           km_wss = km_wss$y)

# overlap with celltypes_minor

CD8_sce_samp <- CD8_sce[,sample(colnames(CD8_sce),size = 20000, replace = FALSE)]
set.seed(12356)
dittoSeq::dittoHeatmap(CD8_sce_samp,
                       genes = markers$lymphocytes,
                       assay = "exprs",
                       annot.by = c( "out_90", "celltypes_minor"),
                       cluster_cols = FALSE,
                       cluster_rows = TRUE,
                       show_colnames = FALSE,
                       heatmap.colors = colorRampPalette(c("dark blue", "white", "dark red"))(100),
                       breaks = seq(-3, 3, length.out = 101))

```


```{r}
mean_sce <- aggregateAcrossCells(CD8_sce,ids=CD8_sce$celltypes_minor,statistics = "mean")
assay(mean_sce,"exprs") <- asinh(counts(mean_sce))
meta_mat <- t(assay(mean_sce[markers$lymphocytes,],"exprs"))
#meta_mat <- scale(meta_mat)

dittoSeq::dittoHeatmap(mean_sce,
                       genes = markers$lymphocytes,
                       assay = "exprs",
                       annot.by = c("celltypes_minor"),
                       cluster_cols = TRUE,
                       cluster_rows = TRUE,
                       show_colnames = TRUE,
                       heatmap.colors = colorRampPalette(c("dark blue", "white", "dark red"))(100),
                       breaks = seq(-3, 3, length.out = 101))
```

