---
title: "Sub-cell-typing"
author: "Daniel Schulz"
date: "7/06/2022"
output: html_document
---

in this script we will use either gating or gaussian mixture models to define the following sub-cell types:
- BnT_CD4 -> we will also classify BnT_CD84_PD1 positive cells as CD4_PD1 since if they are T cell and PD1 it seems likely that they are T cells with spill over B cells signal
- BnT_CD8 -> we will also classify BnT_CD8_PD1 positive cells as CD8_PD1
- CD8_PD1: PD1+
- CD4_PD1: PD1+
- CD8_cycling: Ki67+PD1-
- CD8_cyt: GrzB+PD1-

From images inspection with cytomapper the following seem like reasonable cut-offs:
PD1: 1.2
GrzB: 2.2
Ki67: 2.5
CD8a: 2.2


```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE, eval = TRUE)
knitr::opts_knit$set(root.dir = rprojroot::find_rstudio_root_file())

# Set path
if (Sys.info()["sysname"] == "Darwin"){
  mount_path <- "/Volumes/"
} else if (Sys.info()["sysname"] == "Linux"){
  mount_path <- "/mnt/"
} else {
  stop("Operating system is not supported")
}
```


# paths
```{r}
mount_path <- paste0(mount_path,"immucan_volume/processed_data/NSCLC2_cohort/data/IMC/Panel_1/")
```

```{r}
library(SingleCellExperiment)
library(imcRtools)
library(cytomapper)
library(tidyverse)
library(dittoSeq)
library(viridis)
library(doParallel)
library(BiocParallel)
library(BiocNeighbors)
library(scran)
library(bluster)
library(scater)
library(cowplot)
```

```{r}
sce <- readRDS(paste0(mount_path,"Rout/sce.rds"))
sce$cell_id <- colnames(sce)
```

```{r}
cur_sce <- sce[,sce$celltypes == "CD8"]
dittoScatterPlot(object = cur_sce,x.var = "PD1",y.var = "TCF7",assay.x = "exprs",assay.y = "exprs", color.var = "PD1_pos", legend.show = FALSE)
library(LSD)
cur_dat <- as.data.frame(t(assay(cur_sce,"exprs")))
LSD::heatscatter(cur_dat$PD1,cur_dat$TCF7)+abline(v = 1, h = 1.5)
tumor_sce <- sce[,sce$celltypes == "Tumor"]
tumor_sce <-tumor_sce[,sample(colnames(tumor_sce),50000,replace=FALSE)]
cur_dat_tumor <- as.data.frame(t(assay(tumor_sce,"exprs")))
LSD::heatscatter(cur_dat_tumor$PDL1,cur_dat_tumor$Ki67)
myel_sce <- sce[,sce$celltypes %in% c("MacCD163","DC")]
cur_dat_myel <- as.data.frame(t(assay(myel_sce,"exprs")))
LSD::heatscatter(cur_dat_tumor$PDL1,cur_dat_tumor$CD11c)+abline(v=1.3)
```

# update positivity
we will select thresholds below
```{r}
sce$PD1_pos <- assay(sce, "exprs")["PD1",] > 1
sce$GrzB_pos <- assay(sce, "exprs")["GrzB",] > 2
sce$Ki67_pos <- assay(sce, "exprs")["Ki67",] > 1
sce$TCF7_pos <- assay(sce, "exprs")["TCF7",] >1.5
sce$PDL1_pos <- assay(sce, "exprs")["PDL1",] >1.3
```


# minor cell types

```{r}
# define BnT_CD8 cells
cur_sce <- sce[,sce$celltypes == "BnT"]
BnT_CD8_ids <- cur_sce[,which(assay(cur_sce,"exprs")["CD8a",]>2)]$cell_id
BnT_CD4_ids <- cur_sce[,which(assay(cur_sce,"exprs")["CD8a",]<2)]$cell_id


dat <- as.data.frame(colData(sce))
dat$celltypes_minor <- dat$celltypes

dat <- dat %>%
  mutate(celltypes_minor = ifelse(cell_id %in% BnT_CD8_ids,"BnT_CD8",celltypes_minor),
         celltypes_minor = ifelse(cell_id %in% BnT_CD4_ids,"BnT_CD4",celltypes_minor),
         celltypes_minor = ifelse(celltypes_minor %in% c("CD8","BnT_CD8") & PD1_pos == TRUE & Ki67_pos == TRUE & GrzB_pos == TRUE,"CD8_PD1_Ki67_GRZB", celltypes_minor),
         celltypes_minor = ifelse(celltypes_minor %in% c("CD8","BnT_CD8") & PD1_pos == TRUE & Ki67_pos == TRUE & GrzB_pos == FALSE & TCF7_pos == FALSE,"CD8_PD1_Ki67", celltypes_minor),
         celltypes_minor = ifelse(celltypes_minor %in% c("CD8","BnT_CD8") & PD1_pos == TRUE & Ki67_pos == FALSE & GrzB_pos == FALSE & TCF7_pos == FALSE,"CD8_PD1", celltypes_minor),
         celltypes_minor = ifelse(celltypes_minor %in% c("CD8","BnT_CD8") & PD1_pos == FALSE & Ki67_pos == TRUE & GrzB_pos == TRUE,"CD8_Ki67_GRZB", celltypes_minor),
         celltypes_minor = ifelse(celltypes_minor %in% c("CD8","BnT_CD8") & PD1_pos == FALSE & Ki67_pos == TRUE & GrzB_pos == FALSE & TCF7_pos == FALSE,"CD8_Ki67", celltypes_minor),
         celltypes_minor = ifelse(celltypes_minor %in% c("CD8","BnT_CD8") & PD1_pos == FALSE & Ki67_pos == FALSE & GrzB_pos == TRUE & TCF7_pos == TRUE,"CD8_GRZB_TCF7", celltypes_minor),
         celltypes_minor = ifelse(celltypes_minor %in% c("CD8","BnT_CD8") & PD1_pos == FALSE & Ki67_pos == FALSE & GrzB_pos == FALSE & TCF7_pos == TRUE,"CD8_TCF7", celltypes_minor),
         celltypes_minor = ifelse(celltypes_minor %in% c("CD8","BnT_CD8") & PD1_pos == FALSE & Ki67_pos == FALSE & GrzB_pos == TRUE & TCF7_pos == FALSE,"CD8_GRZB", celltypes_minor),
         celltypes_minor = ifelse(celltypes_minor %in% c("CD8","BnT_CD8") & PD1_pos == FALSE & Ki67_pos == TRUE & GrzB_pos == FALSE & TCF7_pos == FALSE,"CD8_Ki67", celltypes_minor),
         celltypes_minor = ifelse(celltypes_minor %in% c("CD8","BnT_CD8") & PD1_pos == TRUE & Ki67_pos == TRUE & GrzB_pos == TRUE & TCF7_pos == FALSE,"CD8_PD1_GRZB", celltypes_minor),
         celltypes_minor = ifelse(celltypes_minor %in% c("CD8","BnT_CD8") & PD1_pos == TRUE & Ki67_pos == FALSE & GrzB_pos == FALSE & TCF7_pos == TRUE,"CD8_PD1_TCF7", celltypes_minor),
         celltypes_minor = ifelse(celltypes_minor %in% c("CD4","BnT_CD4") & PD1_pos == TRUE & TCF7_pos == TRUE,"CD4_PD1_TCF7", celltypes_minor),
         celltypes_minor = ifelse(celltypes_minor %in% c("CD4","BnT_CD4") & PD1_pos == TRUE & TCF7_pos == FALSE,"CD4_PD1", celltypes_minor),
         celltypes_minor = ifelse(celltypes_minor %in% c("CD4","BnT_CD4") & PD1_pos == FALSE & TCF7_pos == TRUE,"CD4_TCF7", celltypes_minor),
         celltypes_minor = ifelse(celltypes_minor == "MacCD163" & PDL1_pos == TRUE ,"MacCD163_PDL1", celltypes_minor),
         celltypes_minor = ifelse(celltypes_minor == "DC" & PDL1_pos == TRUE ,"DC_PDL1", celltypes_minor))

table(dat$celltypes_minor)
# define minor cell types
colData(sce) <- DataFrame(dat)
```


# color cell types
```{r}
library(pals)
cell_types <- unique(sce$celltypes)

names(cell_types) <- cell_types
cell_types["B"] <- "red3"
cell_types["BnT"] <- "orangered"
cell_types["plasma"] <- "goldenrod2"
cell_types["CD4"] <- "deeppink"
cell_types["Treg"] <- "hotpink"
cell_types["CD8"] <- "darkorchid"
cell_types["DC"] <- "green4"
cell_types["HLADR"] <- "darkgreen"
cell_types["MacCD163"] <- "green"
cell_types["Neutrophil"] <- "blue1"
cell_types["NK"] <- "yellow"
cell_types["pDC"] <- "cyan4"
cell_types["Mural"] <- "deepskyblue"
cell_types["Tumor"] <- "sienna4"
cell_types["undefined"] <- "gray"

pie(c(rep(1,length(cell_types))),col = cell_types,labels = names(cell_types))

metadata(sce)$colour_vectors$celltypes <- cell_types
```

```{r determine markers for clustering}
markers <- list()
# we will not subcluters "uncertain" cells as the merging will be very difficult. we may inspect this at a later time point again

#functional markers for T cells
markers[["CD4"]] <- rownames(sce)[grepl(pattern = "CD27|LAG3|PD1|CD7|GrzB|ICOS|Ki67|TCF7",rownames(sce))]

markers[["CD8"]] <- rownames(sce)[grepl(pattern = "CD27|LAG3|PD1|CD7|GrzB|ICOS|Ki67|TCF7",rownames(sce))]

markers[["Treg"]] <- rownames(sce)[grepl(pattern = "CD27|LAG3|ICOS|Ki67|TCF7",rownames(sce))]

# B and BnT cell markers
markers[["BnT"]] <- rownames(sce)[grepl(pattern = "CD27|CD45RA|CD45RO|LAG3|PD1|CD7|GrzB|FOXP3|ICOS|Ki67|TCF7|VISTA|CD40|HLADR|CD38",rownames(sce))]

markers[["B"]] <- rownames(sce)[grepl(pattern = "CD27|CD40|Ki67",rownames(sce))]

# myeloid markers
markers[["MacCD163"]] <- rownames(sce)[grepl(pattern = "CD68|CD11c|PDL1|VISTA|CD40|CD14|CD206|IDO1|CD16",rownames(sce))]

# we will also use CD20 for DCs to detect those DCs that sit in TLS
markers[["DC"]] <- rownames(sce)[grepl(pattern = "PDL1|VISTA|CD40|IDO1|CD20",rownames(sce))]

# we will also use E-cad for HLADR cells as there may still be tumor cells in these cells
markers[["HLADR"]] <- rownames(sce)[grepl(pattern = "CD11b|CD163|CD68|CD11c|PDL1|ICOS|VISTA|TIM3|CD40|CD14|CD206|CD163|HLADR|Ecad",rownames(sce))]

# NK markers
markers[["NK"]] <- rownames(sce)[grepl(pattern = "CD7|GrzB|CD16|CD45RA|CD45RO",rownames(sce))]

# Neutrophil markers
markers[["Neutrophil"]] <- rownames(sce)[grepl(pattern = "MPO|PDL1|VISTA",rownames(sce))]

# pDC markers
markers[["pDC"]] <- rownames(sce)[grepl(pattern = "CD303|CD4|GrzB|CD45RA|CD45RO|CD27",rownames(sce))]

# Plasma cell markers
markers[["plasma"]] <- rownames(sce)[grepl(pattern = "CD38|TCF7|CD27|CD45RA|CD45RO|",rownames(sce))]

# stroma markers
markers[["Mural"]] <- rownames(sce)[grepl(pattern = "PDGFR|SMA",rownames(sce))]

# tumor markers
markers[["Tumor"]] <- rownames(sce)[grepl(pattern = "HLADR|CarbonicAnhydrase|cleavedPARP|PDL1|IDO1|CD15|Ki67|Ido|B2M",rownames(sce))]

good_markers <- rownames(sce)[!grepl("DNA|Histone",rownames(sce))]
```

# tumor subclustering
This does not work using graph based clustering from the bluster package. we will stick to the Rphenograp implementation.
We will then group or meta cluster the phenograph clusters


```{r, fig.height=15,fig.width=15}

library(Rphenograph)
library(Rphenoannoy)
tumor_sce <- sce[,sce$celltypes %in% c("Tumor")]

set.seed(12345)
tumor_sce <- runUMAP(tumor_sce, exprs_values = "exprs", 
               subset_row = markers$Tumor,
               name = "tumor_UMAP_exprs",
               external_neighbors = TRUE, BNPARAM = AnnoyParam(),
               BPPARAM = MulticoreParam(workers = 12))

plotReducedDim(tumor_sce,dimred = "tumor_UMAP_exprs",colour_by = "immucan_id")+theme(legend.position = "none")

# time tests
test_sce <- tumor_sce[,sample(colnames(tumor_sce),100000, replace = FALSE)]
test_mat <-  t(assay(test_sce, "exprs")[markers$Tumor,])
graph_time_start <- Sys.time()
out <- Rphenograph(test_mat, k = 70)
graph_time_end <- Sys.time()
graph_time <- graph_time_end-graph_time_start

annoy_time_start <- Sys.time()
out <- Rphenoannoy(test_mat, k = 70)
annoy_time_end <- Sys.time()
annoy_time <- annoy_time_end-annoy_time_start


mat <- t(assay(tumor_sce, "exprs")[markers$Tumor,])


set.seed(12345)
out_40 <- Rphenoannoy(threshold = ,mat, k = 40)

set.seed(12345)
out_50 <- Rphenoannoy(threshold = ,mat, k = 50)


set.seed(12345)
out_60 <- Rphenoannoy(threshold = ,mat, k = 60)

set.seed(12345)
out_70 <- Rphenoannoy(threshold = ,mat, k = 70)

set.seed(12345)
out_80 <- Rphenoannoy(threshold = ,mat, k = 80)

clusters_40 <- factor(membership(out_40[[2]]))
clusters_50 <- factor(membership(out_50[[2]]))
clusters_60 <- factor(membership(out_60[[2]]))
clusters_70 <- factor(membership(out_70[[2]]))
clusters_80 <- factor(membership(out_80[[2]]))

tumor_sce$tumor_clust_40 <- clusters_40
tumor_sce$tumor_clust_50 <- clusters_50
tumor_sce$tumor_clust_60 <- clusters_60
tumor_sce$tumor_clust_70 <- clusters_70
tumor_sce$tumor_clust_80 <- clusters_80



p1 <- dittoDimPlot(tumor_sce, var = "tumor_clust_50", 
             reduction.use = "tumor_UMAP_exprs", size = 0.2,
             do.label = TRUE) +
    ggtitle("PG clusters on UMAP")
p2 <- dittoDimPlot(tumor_sce, var = "tumor_clust_60", 
             reduction.use = "tumor_UMAP_exprs", size = 0.2,
             do.label = TRUE) +
    ggtitle("PG clusters on UMAP")
p3 <- dittoDimPlot(tumor_sce, var = "tumor_clust_70", 
             reduction.use = "tumor_UMAP_exprs", size = 0.2,
             do.label = TRUE) +
    ggtitle("PG clusters on UMAP")
p4 <- dittoDimPlot(tumor_sce, var = "tumor_clust_80", 
             reduction.use = "tumor_UMAP_exprs", size = 0.2,
             do.label = TRUE) +
    ggtitle("PG clusters on UMAP")
p5 <- dittoDimPlot(tumor_sce, var = "tumor_clust_40", 
             reduction.use = "tumor_UMAP_exprs", size = 0.2,
             do.label = TRUE) +
    ggtitle("PG clusters on UMAP")

unique(tumor_sce$tumor_clust)

plot_grid(p1, p2, p3, p4, p5)
```

__there is not a big difference in cluster numbers between 50 and 80 nearest neighbors for phenograph clustering__

## Single cell tumor cluster heatmap
```{r, fig.height=8, fig.width=20}
set.seed(12345)
dittoHeatmap(tumor_sce[,sample(colnames(tumor_sce),20000,replace = FALSE)],genes = markers$Tumor, assay = "exprs",
            annot.by = c("tumor_clust_40", "tumor_clust_70", "tumor_clust_60", "tumor_clust_50", "tumor_clust_80", "celltypes_minor"),
            cluster_rows = TRUE,
            scale = "none",
            show_colnames = FALSE,
            heatmap.colors = viridis(100))
```

## merge tumor clusters
there are two small clusters that are either Caspase/Parp high or Ido1 high. we will keep those and metacluster the remaining clusters
we will metacluster on the z-scaled mean intensity of the clusters
```{r, fig.height=8, fig.width=10}
tumor_mean <- aggregateAcrossCells(tumor_sce,ids=tumor_sce$tumor_clust_40,statistics = "mean")
assay(tumor_mean,"exprs") <- asinh(assay(tumor_mean,"counts"))


dittoHeatmap(tumor_mean,genes =markers$Tumor, assay = "exprs",
             annot.by = c("tumor_clust_40"),
             cluster_rows = TRUE,
             show_colnames = TRUE,
             heatmap.colors = colorRampPalette(c("dark blue", "white", "dark red"))(100),
             breaks = seq(-3, 3, length.out = 101))

dittoHeatmap(tumor_mean,genes =c("CD15","B2M","HLADR","PDL1","CarbonicAnhydrase","Ki67"), assay = "exprs",
             annot.by = c("tumor_clust_40"),
             cluster_rows = TRUE,
             show_colnames = TRUE,
             heatmap.colors = colorRampPalette(c("dark blue", "white", "dark red"))(100),
             breaks = seq(-3, 3, length.out = 101))

#sub_tumor_sce <- tumor_sce[,!tumor_sce$tumor_clust_40 %in% c(7,12)]

#sub_tumor_mean <- aggregateAcrossCells(sub_tumor_sce,ids=sub_tumor_sce$tumor_clust_40,statistics = "mean")
#assay(sub_tumor_mean,"exprs") <- asinh(assay(tumor_mean,"counts"))

# we will meta cluster on the z-scaled mean cluster intensity
meta_mat <- t(assay(tumor_mean[markers$Tumor,],"exprs"))
meta_mat <- scale(meta_mat)

library(factoextra)
set.seed(12356)
fviz_nbclust(meta_mat, kmeans, method = "wss",k.max = 20)
fviz_nbclust(meta_mat, kmeans, method = "silhouette",k.max = 20)
fviz_nbclust(meta_mat, kmeans, method = "gap_stat",k.max = 20)

# it seems that 11 clusters are reasonable
set.seed(12356)
meta_clusters <- kmeans(meta_mat,11)

tumor_mean$meta_cluster <- as.factor(meta_clusters$cluster)


dittoHeatmap(tumor_mean,genes = markers$Tumor, assay = "exprs",
             annot.by = c( "meta_cluster", "tumor_clust_40"),
             cluster_rows = TRUE,
             show_colnames = TRUE,
             heatmap.colors = colorRampPalette(c("dark blue", "white", "dark red"))(100),
             breaks = seq(-3, 3, length.out = 101))

table(tumor_sce$tumor_clust_40)
tumor_meta_cluster <- tumor_mean$meta_cluster
names(tumor_meta_cluster) <- tumor_mean$tumor_clust_40

tumor_sce$tumor_meta_clusters <- as.factor(tumor_meta_cluster[tumor_sce$tumor_clust_40])
```


## investigate meta clusters on single cells
```{r, fig.height=15, fig.width=20}
set.seed(12345)
dittoHeatmap(tumor_sce[,sample(colnames(tumor_sce),20000,replace = FALSE)],genes = markers$Tumor, assay = "exprs",
            annot.by = c("tumor_meta_clusters","tumor_clust_40", "celltypes_minor"),
            cluster_rows = TRUE,
            scale = "none",
            show_colnames = FALSE,
            heatmap.colors = viridis(100))

plot_mean <- aggregateAcrossCells(tumor_mean,ids=tumor_mean$meta_cluster,statistics = "mean")
assay(plot_mean,"exprs") <- asinh(assay(plot_mean,"counts"))
dittoHeatmap(plot_mean,genes = markers$Tumor, assay = "exprs",
             annot.by = c( "meta_cluster"),
             cluster_rows = TRUE,
             scale =  "none",
             show_colnames = TRUE,
             heatmap.colors = viridis(100))
dittoHeatmap(plot_mean,genes = markers$Tumor, assay = "exprs",
             annot.by = c( "meta_cluster"),
             cluster_rows = TRUE,
             heatmap.colors = colorRampPalette(c("dark blue", "white", "dark red"))(100),
             breaks = seq(-3, 3, length.out = 101))
```

# assign names and save tumor_sce
```{r}
tumor_sce$tumor_clust_50 <- NULL
tumor_sce$tumor_clust_60 <- NULL
tumor_sce$tumor_clust_70 <- NULL
tumor_sce$tumor_clust_80 <- NULL

tumor_clust_names <- as.character(c(1:11))
names(tumor_clust_names) <- c("Tumor_apoptotic","Tumor_CD15_Ki67","Tumor_CD15","Tumor_B2M_HLADR_IDO1","Tumor_HLADR","Tumor_Ki67_CD15_HLADR","Tumor_all_low", "Tumor_CD15_HLADR_B2M_CAIX", "Tumor_PDL1","Tumor_CAIX_Ki67","Tumor_B2Mlow_HLADRlow_Ki67")

tumor_sce$tumor_meta_clusters <- names(tumor_clust_names[tumor_sce$tumor_meta_clusters])

saveRDS(tumor_sce,paste0(mount_path,"Rout/tumor_sce.rds"))
```

# merge with sce object to cell types minor
```{r}
tumor_cells <- tumor_sce$tumor_meta_clusters
names(tumor_cells) <- tumor_sce$cell_id

sce[,names(tumor_cells)]$celltypes_minor <- tumor_cells
table(sce$celltypes_minor)
```

# color for minor cell types and cell types
```{r, fig.height=10, fig.width=10, eval=FALSE}
library(pals)
cell_types <- sort(unique(sce$celltypes_minor))

names(cell_types) <- cell_types
cell_types[grepl("CD8",cell_types)] <- c("darkorchid1","darkorchid2","darkorchid3","darkorchid4","darkmagenta")
cell_types[grepl("CD4",cell_types)] <- c("deeppink1","deeppink2","deeppink3")
cell_types[grepl("Mac",cell_types)] <- c("green","yellowgreen")
cell_types[grepl("^DC",cell_types)] <- c("green3","forestgreen")
cell_types[grepl("Tumor",cell_types)] <- c("sienna", "sienna2","sienna3","sienna4","lightsalmon3","lightsalmon4","tan4","salmon4","peru","lightsalmon4","chocolate4")
cell_types["B"] <- "red3"
cell_types["BnT"] <- "orangered"
cell_types["plasma"] <- "goldenrod2"
cell_types["Treg"] <- "hotpink"
cell_types["HLADR"] <- "darkgreen"
cell_types["Neutrophil"] <- "blue1"
cell_types["NK"] <- "yellow"
cell_types["pDC"] <- "cyan4"
cell_types["Mural"] <- "deepskyblue"
cell_types["undefined"] <- "gray"

pie(c(rep(1,length(cell_types))),col = cell_types,labels = names(cell_types))

metadata(sce)$colour_vectors$celltypes_minor <- cell_types
```

# Heatmap all cell types minor 
```{r, fig.height=10, fig.width=15}
cur_dat <- colData(sce)
ids <- as_tibble(cur_dat) %>%
  group_by(celltypes_minor) %>%
  slice_sample(n = 150,replace = FALSE) %>%
  pull(cell_id)

cur_sce <- sce[,sce$cell_id %in% ids]

dittoHeatmap(cur_sce,
             genes = !grepl("DNA|Histone",rownames(cur_sce)),
             annot.by = c("celltypes_minor","celltypes"),
             assay = "exprs",
             annotation_colors = list(celltypes = metadata(sce)$colour_vectors$celltypes,
                                      celltypes_minor = metadata(sce)$colour_vectors$celltypes_minor),
             heatmap.colors = colorRampPalette(c("dark blue", "white", "dark red"))(100),breaks = seq(-3, 3, length.out = 101))
```


# save
```{r}
saveRDS(sce,paste0(mount_path,"Rout/sce.rds"))
```

```{r, fig.height=10, fig.width=10}
sce <- readRDS(paste0(mount_path,"Rout/sce.rds"))
old_sce <- readRDS("/mnt/immucan_volume/processed_data/Panel_1/IMMUcan_NSCLC2/IMC/Rout/data/sce.rds")

table(sce$celltypes_minor)
table(old_sce$celltypes_minor)
table(sce$celltypes)
table(old_sce$celltypes)

cur_dat <- data.frame(ID = sce$cell_id,
                      new_minor = sce$celltypes_minor)
cur_dat_old <- data.frame(ID = old_sce$cell_id,
                      old_minor = old_sce$celltypes_minor)

cur_dat <- left_join(cur_dat,cur_dat_old,"ID")
cur_dat <- cur_dat %>%
  mutate(new_minor = paste0(new_minor,"_new"))
caret::confusionMatrix(data= as.factor(cur_dat$new_minor),reference = as.factor(cur_dat$old_minor))
plot_dat <- as.data.frame(unclass(table(cur_dat$new_minor,cur_dat$old_minor)))
pheatmap::pheatmap(as.matrix(plot_dat/rowSums(plot_dat)))
unique(cur_dat$old_minor)
```







